### 2024-11-19 inst/manuscript/manuscript.qmd
initial commit

---
title: "Spatial Generalized Linear Models in [R]{.proglang} Using [spmodel]{.pkg}"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Michael Dumelle
    affiliations:
      - name: |
          | United States
          | Environmental Protection Agency
        department: Pacific Ecological Systems Division
        address: 200 SW 35th St
        city: Corvallis
        country: OR
        postal-code: 97333
    orcid: 0000-0002-3393-5529
    email: dumelle.michael@epa.gov
  - name: Jay M. Ver Hoef
    affiliations:
      - name: |
          | National Oceanic and
          | Atmospheric Administration
        department: Alaska Fisheries Science Center
    orcid: 0000-0003-4302-6895
  - name: Matt Higham
    affiliations:
      - name: St. Lawrence University
        department: Department of Math, Computer Science, and Statistics
    orcid: 0009-0006-4217-625X
abstract: |
  Generalized linear models (GLMs) describe a non-normal response variable that may be binary, count, skewed, or a proportion. Typically, observations in a GLM are assumed independent of one another. For spatial data, this independence assumption is impractical, as nearby locations tend to be more similar than locations far apart. The [spmodel]{.pkg} [R]{.proglang} package provides tools to fit GLMs that incorporate spatial autocorrelation (i.e., spatial generalized linear models, or SPGLMs). SPGLMs are fit in [spmodel]{.pkg} using a novel application of the Laplace approximation via [spglm]{.fct} for point-referenced data or [spgautor]{.fct} for areal (i.e., lattice), data. [spglm]{.fct} and [spgautor]{.fct} closely resemble [glm]{.fct} from base [R]{.proglang} but include arguments that control the spatial autocorrelation structure. [spmodel]{.pkg} has many helper functions for model inspection and diagnostics, some of which leverage other [R]{.proglang} packages like [broom]{.pkg} and [emmeans]{.pkg}. [spmodel]{.pkg} has tools to make predictions of the latent spatial-mean process at unobserved locations. [spmodel]{.pkg} also provides many advanced features like accommodating geometric anisotropy and nonspatial random effects, simulating spatially autocorrelated data, and more. Here we use [spmodel]{.pkg} to illustrate the modeling of binary, count, skewed and proportion response variables from several point-referenced and areal data sets. 
keywords: [Autoregressive Model, Geostatistics, Spatial Covariance, Spatially-Explicit Model, Statistical Model]
keywords-formatted: [Autoregressive Model, Geostatistics, Spatial Covariance, Spatially-Explicit Model, Statistical Model]
bibliography: references.bib  
preamble: |
  \usepackage{amsmath,amsfonts,amssymb}
  \usepackage{bm, bbm}
  \usepackage{lineno}
  \usepackage{caption, subcaption}
editor_options: 
  chunk_output_type: console
---

```{R}
#| include: false
library(spmodel)
library(ggplot2)
library(dplyr)
library(emmeans)
library(here)
fig_path <- here("inst", "manuscript", "figures")
```

\newpage

## Introduction {#sec-intro}

Binary, count, proportion, and skewed data are ubiquitous in practice. These data types are naturally modeled using a generalized linear model (GLM) framework [@nelder1972generalized; @mccullagh1989generalized; @myers2012generalized; @faraway2016extending]. In a GLM, a response variable $\text{y}$ belongs to a particular statistical distribution with mean $\mu$. For example, $\text{y}$ may be distributed as a Poisson random variable with some mean $\mu$. Based on the response distribution assumed for $\text{y}$, GLMs link a function of $\mu$ to explanatory variables via a link function:
\begin{equation}\label{eq-glm}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
\end{equation}
where for a sample size $n$, $\boldsymbol{\mu}$ is the mean of an $n \times 1$ response vector $\mathbf{y}$, $f(\boldsymbol{\mu})$ is the link function that connects $\boldsymbol{\mu}$ to $\mathbf{w}$, $\mathbf{X}$ is the $n \times p$ design matrix of explanatory variables, and $\boldsymbol{\beta}$ is the $p \times 1$ vector of fixed effects. The mean, $\boldsymbol{\mu}$, is usually constrained in some way (e.g., positive) that depends on the distribution assumed for $\mathbf{y}$, but $\mathbf{w}$ is unconstrained. The parameters in Equation$~$\eqref{eq-glm} are typically estimated via maximum likelihood (e.g., iteratively reweighted least squares) [@chambers1992S]. The [glm]{.fct} function is commonly used to fit GLMs in the [R]{.proglang} programming language [@rcore2024].

The GLM framework in Equation$~$\eqref{eq-glm} assumes the elements of $\mathbf{y}$ are independent of one another. This assumption is impractical for spatial data, where nearby observations tend to be more similar than distant observations [@tobler1970computer]. Ignoring this spatial dependence can give rise to misleading inference and poor prediction [@zimmerman2024spatial]. Spatial GLMs (SPGLMs) formally incorporate spatial autocorrelation into a GLM by adding to Equation$~$\eqref{eq-glm} two random effects that elucidate spatial structure:
\begin{equation}\label{eq-spglm}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
\end{equation}
where $\boldsymbol{\tau}$ is an $n \times 1$ column vector of spatially dependent random errors, and $\boldsymbol{\epsilon}$ is an $n \times 1$ column vector of spatially independent random errors. We make a few assumptions about $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$: first, that $\text{E}(\boldsymbol{\tau}) = \text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}$, where $\text{E}(\cdot)$ denotes expectation; second, that $\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}$, where $\mathbf{R}$ is an $n \times n$ matrix that determines the spatial dependence structure in $\mathbf{w}$ (which is linked to $\boldsymbol{\mu}$ and hence $\mathbf{y}$) and depends on a range parameter, $\phi$; third, that $\text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_\epsilon \mathbf{I}$, where $\mathbf{I}$ is an $n \times n$ identity matrix; and fourth, that $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$ are independent of one another. The parameter $\sigma^2_{\tau}$ is called the spatially dependent random error variance or partial sill. The parameter $\sigma^2_\epsilon$ is called the spatially independent random error variance or nugget. These two variance parameters are henceforth more intuitively written as $\sigma^2_{de}$ and $\sigma^2_{ie}$, respectively.  The covariance of $\mathbf{w}$ is denoted $\boldsymbol{\Sigma}$ and is given by
\begin{equation}\label{eq-spcov}
 \boldsymbol{\Sigma} = \sigma^2_{de}\mathbf{R} + \sigma^2_{ie} \mathbf{I}.
\end{equation}
The parameters $\sigma^2_{de}$, $\phi$, and $\sigma^2_{ie}$ are elements of $\boldsymbol{\theta}$, the covariance parameter vector.

The [spmodel]{.pkg} [R]{.proglang} package provides tools for fitting spatial statistical models and making predictions at unobserved locations [@dumelle2023spmodel]. A recent major update (v0.4.0) to [spmodel]{.pkg} added SPGLM (Equation$~$\ref{eq-spglm}) support for binary, count, skewed, and proportion response variables (Table$~$\ref{table-links}), greatly expanding the class of models that [spmodel]{.pkg} makes accessible to practitioners.
\begin{table}
\centering
\begin{tabular}{||c|ccc||} 
 \hline
 Family & Link Function & Link Name & Data Type  \\ [0.5ex] 
 \hline \hline
 Binomial & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Binary; Binary Count \\
 Poisson & $f(\mu) = \log(\mu)$ & Log & Count \\
 Negative Binomial & $f(\mu) = \log(\mu)$ & Log & Count \\
 Beta & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Proportion \\
 Gamma & $f(\mu) = \log(\mu)$ & Log & Skewed \\
 Inverse Gaussian & $f(\mu) = \log(\mu)$ & Log & Skewed \\ 
 \hline\hline
\end{tabular}
\caption{SPGLM response distributions and their link functions and data types.}
\label{table-links}
\end{table}
SPGLMs for point-referenced data are called geostatistical GLMs and are fit using [spglm]{.fct}. Data are point-referenced when the elements in $\mathbf{y}$ are observed at point-locations indexed by x-coordinates and y-coordinates on a spatially continuous surface with an infinite number of locations (e.g., point locations in a field). SPGLMs for areal data are called spatial autoregresive models and are fit using [spgautor]{.fct}. Data are areal (i.e., lattice) when they are part of a finite network of polygons whose connections are indexed by a neighborhood structure (e.g., states in a country).

Several other [R]{.proglang} packages exist for analyzing SPGLMs. The [brms]{.pkg} [@burkner2017brms], [carBayes]{.pkg} [@lee2013carbayes], [ngspatial]{.pkg} [@hughes2020ngspatial], [R-INLA]{.pkg} [@lindgren2015bayesian], [spBayes]{.pkg} [@finley2007spbayes], and [spNNGP]{.pkg} [@finley2002spnngp] packages take a Bayesian approach, either directly sampling from posterior distributions of parameters (e.g., using MCMC) or approximating them. A benefit of Bayesian approaches is that prior information can be incorporated and uncertainty quantification of parameter estimates is straightforward. However, Bayesian approaches, especially those using MCMC, tend to be computationally expensive. In order to reduce computation time, many of these packages work with the precision matrix instead of the covariance matrix so that inverting matrices (a computational bottleneck) is not required. For example, [R-INLA]{.pkg} uses the precision matrix (often directly modeled as with autoregressive models) and tends to be very fast. Working with precision matrices, however, can be more restrictive and less intuitive than working directly with the covariance matrix. The [FRK]{.pkg} [@sainsbury2024modeling], [glmmTMB]{.pkg} [@brooks2017glmmtmb], [hglm]{.pkg} [@ronnegard2010hglm], [mgcv]{.pkg} [@wood2017generalized], and [spaMM]{.pkg} [@rousset2014spamm] packages directly use Laplace, quasi-likelihood, or reduced-rank approaches to estimate parameters. These direct approaches tend to be computationally efficient, as they don't rely on MCMC sampling. In contrast to the Bayesian approach, a drawback of these direct approaches is that prior information cannot be formally incorporated and covariance parameter uncertainty is challenging to quantify. SPGLMs in [spmodel]{.pkg} are most similar to those in [glmmTMB]{.pkg} -- [spmodel]{.pkg} uses analytical solutions for maximizing the likelihood (which we describe in the next section), while [glmmTMB]{.pkg} uses automatic differentiation for maximizing the likelihood (which tends to be much slower than analytical solutions).

Missing from the aforementioned [R]{.proglang} packages is the complete set of tools for SPGLMs that [spmodel]{.pkg} provides. Importantly, the [spglm]{.fct} and [spautor]{.fct} functions act as a spatial analogue to the familiar [glm]{.fct} from base [R]{.proglang}, making the transition from GLMs to SPGLMs relatively seamless. [spmodel]{.pkg} leverages many commonly used [R]{.proglang} generics like [summary]{.fct} to better understand fitted models. Six GLM families (Table$~$\ref{table-links}) and 20 spatial covariance (or precision) functions are supported. Also available are functions for data visualization, model fitting, model summaries, model diagnostics, model comparison, and prediction, all crucial components of a data analysis. Through extra function arguments, [spmodel]{.pkg} supports advanced features like geometric anisotropy, non-spatial random effects, methods for large data sets, and more. Importantly, [spmodel]{.pkg} extends popular packages like [broom]{.pkg} [@robinson2021broom; @kuhn2022tidy] and [emmeans]{.pkg} [@lenth2024emmeans]. [spmodel]{.pkg} also provides function for simulating SPGLM data (e.g., [sprbinom]{.fct}, [sprpois]{.fct}).

The rest of this article is organized as follows. In @sec-spglm, we provide some background for the SPGLM fitting and prediction routines in [spmodel]{.pkg}. In @sec-applications, we provide several applications of [spmodel]{.pkg} to spatial binary, count, skewed and proportion data with both point-referenced and areal supports. And in @sec-discussion, we end with a discussion synthesizing \pkg{spmodel}'s contributions to the analysis of SPGLMs in [R]{.proglang}.

## Spatial generalized linear models using the Laplace approximation {#sec-spglm}

GLMs with random effects are often written hierarchically [@lee1996hierarchical; @bolker2009generalized; @wood2017generalized]. [spmodel]{.pkg} leverages this hierarchical structure and uses a novel application of the Laplace approximation [@ver2024marginal] to fit models. This marginal approach is quite flexible, accommodating a wide range of possible dependence structures and formally maximizing a likelihood. Maximizing a likelihood yields convenient likelihood-based statistics like AIC [@akaike1974new], AICc [@hoeting2006model], BIC [@schwarz1978estimating], deviance [@mccullagh1989generalized], and likelihood ratio tests for model comparison, a benefit compared to quasi-likelihood [@wedderburn1974quasi; @breslow1993approximate] or pseudo-likelihood approaches [@wolfinger1993generalized], which only specify the first two moments of a distribution. @ver2024marginal provide thorough context for the marginal approach and its associated details, but next we provide an short overview of the methodology.

Our goal is to marginalize over the latent mean $\mathbf{w}$ and fixed effects $\boldsymbol{\beta}$ in Equation$~$\eqref{eq-spglm} to obtain a response distribution for $\mathbf{y}$ that depends only on the explanatory variables, $\mathbf{X}$, the response distribution's dispersion parameter, $\varphi$, and the covariance parameters, $\boldsymbol{\theta}$. We can represent this marginal distribution hierarchically as
\begin{equation}\label{eq-marginal}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w} .
\end{equation}
In Equation$~$\eqref{eq-marginal}, $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ is the density for $\mathbf{y}$ (e.g., binomial, Poisson) given the latent mean and dispersion parameter and $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ is the Gaussian density for $\mathbf{w}$ given the explanatory variables, fixed effects, and covariance parameters. Integrating $\boldsymbol{\beta}$ out of Equation$~$\eqref{eq-spglm} yields
\begin{equation}\label{eq-marginal2}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\mathbf{w},
\end{equation}
where $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ is the restricted (i.e., residual) Gaussian density [@patterson1971recovery, @harville1977maximum, @wolfinger1994computing] for $\mathbf{w}$ given the explanatory variables and covariance parameters. This restricted Gaussian density is given by
\begin{equation}\label{eq-reml-def}
[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] = \frac{\exp(-\frac{1}{2}(\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}}) \boldsymbol{\Sigma}^{-1} (\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}})^T)}{(2 \pi)^{(n - p)/2} |\boldsymbol{\Sigma}|^{1/2}|\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X}|^{1/2}},
\end{equation}
where $\tilde{\boldsymbol{\beta}} = (\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{w}$ and $|\cdot|$ denotes the determinant.

Next, let $\ell_\mathbf{w} = \log([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}])$ and then notice that Equation$~$\eqref{eq-marginal2} can be written as 
\begin{equation}\label{eq-marginal3}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \exp(\ell_\mathbf{w}) d\mathbf{w}.
\end{equation}
A Taylor series expansion of $\ell_\mathbf{w}$ around a point $\mathbf{a}$ yields
\begin{equation}\label{eq-marginal4}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_\mathbf{a} + \mathbf{g}^T(\mathbf{w} - \mathbf{a}) + \frac{1}{2}(\mathbf{w} - \mathbf{a})^T \mathbf{G} (\mathbf{w} - \mathbf{a}))d\mathbf{w},
\end{equation}
where $\mathbf{g}$ and $\mathbf{G}$ are the gradient and Hessian, respectively, of $\ell_\mathbf{w}$ with respect to $\mathbf{w}$. If $\mathbf{a}$ is a value for which $\mathbf{g} = \mathbf{0}$,
\begin{equation}\label{eq-marginal5}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_\mathbf{a}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \mathbf{a})^T (-\mathbf{G}) (\mathbf{w} - \mathbf{a}))d\mathbf{w}.
\end{equation}
Notice that the integral in Equation$~$\ref{eq-marginal5} can be solved in the same manner as the normalizing constant in a multivariate Gaussian distribution, and rewriting $\exp(\ell_\mathbf{a})$ yields
\begin{equation}\label{eq-marginal6}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{a}), \varphi] [\mathbf{a} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{a}}|^{-1/2}.
\end{equation}
@ver2024marginal show how to evaluate $\mathbf{g}$ and $\mathbf{G}$ to obtain \textbf{a} and maximize (the natural logarithm of) Equation$~$\eqref{eq-marginal6} for the six response distributions in Table$~$\ref{table-links}. Maximizing Equation$~$\eqref{eq-marginal6} using a Newton-Rhapson approach yields the marginal restricted maximum likelihood estimators $\hat{\varphi}$ and $\hat{\boldsymbol{\theta}}$. @ver2024marginal show that the fixed effect estimator is $\hat{\boldsymbol{\beta}} = (\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{X})^{-1}\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{w}$, where $\hat{\boldsymbol{\Sigma}}$ is the covariance matrix $\boldsymbol{\Sigma}$ evaluated at $\hat{\boldsymbol{\theta}}$. @ver2024marginal also derive the covariance matrix of $\hat{\boldsymbol{\beta}}$ and show how to predict $\mathbf{w}$ at unobserved locations and quantify their uncertainties.

## Application {#sec-applications}

The [spglm]{.fct} (for point-referenced data) and [spgautor]{.fct} (for areal data) functions in [spmodel]{.pkg} fit SPGLMs using the Laplace approximation outlined in @sec-spglm. Both [spglm]{.fct} and [spgautor]{.fct} generally require the following four arguments: `formula`, a formula that describes the relationship between the response variable and explanatory variables; `family`, the response distribution (which can be `binomial`, `poisson`, `nbinomial`, `Gamma`, `inverse.gaussian`, or `beta`); `data`, the data frame that holds the variables in `formula` as well as spatial locations; and `spcov_type`, the spatial covariance type. The first three arguments are shared by [glm]{.fct}; thus, the transition from GLMs to SPGLMs requires only one additional argument: `spcov_type`. The [spglm]{.fct} spatial covariance types measure dependence as a function of Euclidean distance among observations; an example is the exponential spatial covariance:
\begin{equation}\label{eq-spcov-exp}
  \boldsymbol{\Sigma} = \sigma^2_{de} \exp(-\mathbf{H}/\phi) + \sigma^2_{ie}\mathbf{I},
\end{equation}
where $\mathbf{H}$ is a matrix of pairwise distances among all observations. [spglm]{.fct} currently supports 18 distinct spatial covariance functions.

The [spgautor]{.fct} spatial covariance types measure dependence as a function of neighborhood distance among observations; an example is the simultaneous autoregressive covariance matrix:
\begin{equation}\label{eq-spcov-sar}
  \boldsymbol{\Sigma} = \sigma^2_{de} [(\mathbf{I} - \phi \mathbf{W})(\mathbf{I} - \phi \mathbf{W})^T]^{-1} + \sigma^2_{ie}\mathbf{I},
\end{equation}
where $\mathbf{W}$ is a matrix that represents the neighborhood structure among all observations. [spgautor]{.fct} currently supports two distinct spatial covariance functions.

In the rest of this section, we use [spmodel]{.pkg} to study binary, count, skewed, and proportion response variables that are either point-referenced or areal. We use [spmodel]{.pkg} for all parts of the data analysis, from estimation to inference to model diagnostics to prediction. We first describe core [spmodel]{.pkg} functionality in an application to binary data, while additional analyses highlight count, skewed, and proportion data as well as some additional [spmodel]{.pkg} features. Before proceeding, load [spmodel]{.pkg} into the current [R]{.proglang} session:
```{R}
#| prompt: true
#| eval: false
library("spmodel")
```

### Binary data

The `moose` data in [spmodel]{.pkg} contain information on moose presence in Alaska. They are an `sf` object, a special data frame that is supplemented with spatial information using the [sf]{.pkg} package [@pebesma2018sf]. The first few rows look like:

```{R}
#| prompt: true
head(moose)
```

There are five columns: `elev`, the numeric site elevation (meters); `strat` a stratification variable for sampling with two levels, `"L"` and `"M"`, which are categorized by landscape metrics at each site; `count`, the number of moose at each site; `presence`, a factor that indicates whether at least one moose was observed at each site (`0` implies no moose; `1` implies at least one moose); and `geometry`, the NAD83 projected coordinate of each site. The `moose_preds` data in [spmodel]{.pkg} contain spatial locations at which predictions of moose presence are desired. They are also an `sf` object with the same projection and measurements for `elev` and `strat`. @fig-moose-data shows the `presence` variable in `moose` as well as the spatial locations of both `moose` and `moose_preds`. Moose are most common in the southwestern and eastern parts of the domain and least common in the northwest.
```{R}
#| prompt: false
#| echo: false
#| fig-align: center
#| fig-cap: "Moose presence in Alaska. Circles represent moose presence or absence (based on color) and triangles represent locations at which moose presence probability predictions are desired."
#| out-width: "70%"
#| out-height: "70%"
#| label: fig-moose-data
knitr::include_graphics(paste0(fig_path, "/figure-1.png"))
```

To study the effect of elevation, stratum, and their interaction on moose presence while accounting for spatial autocorrelation, we fit a SPGLM for binary data (i.e., a spatial logistic regression model) using [spglm]{.fct}:

```{R}
#| prompt: true
spbin <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = binomial,
  data = moose,
  spcov_type = "spherical"
)
```

Summarizing the model object yields a summary similar to that provided by the familiar [glm]{.fct}:
```{R}
#| prompt: true
summary(spbin)
```

The summary contains the original function call, a summary of residuals, the fixed effects coefficients table, the spatial covariance parameter estimates, and additional model information like the pseudo R-squared, which quantifies the variability in the model attributable to the fixed effects. While useful, this summary information is hard to work with, as it is printed directly to the [R]{.proglang} console. The [broom]{.pkg} package from the tidymodels [@kuhn2022tidy] ecosystem has functions to provide helpful model output in the form of tibbles (i.e., data frames) that are easily manipulated. [spmodel]{.pkg} has methods for the [tidy]{.fct}, [glance]{.fct}, and [augment]{.fct} functions from [broom]{.pkg}. The first [broom]{.pkg} function is [tidy]{.fct}, which tidies the model output:
```{R}
#| prompt: true
tidy(spbin)
```

The estimates and standard errors returned are on the log odds link (Table$~$\ref{table-links}) scale ([coef]{.fct} and [vcov]{.fct} may also be used). The output provides evidence that elevation is positively associated with moose presence in the `"L"` stratum ($p-$value $<0.05$) and, at zero elevation, moose are more likely in the `"M"` stratum than the `"L"` stratum ($p-$value $<0.01$). This output provides marginal evidence that the effect of elevation on moose presence varies across strata ($p-$value $\approx 0.1$).  The model effectively quantifies the impact of elevation on moose presence for moose in the `"L"` strata, but an analogous statement for moose in the `"M"` strata requires more context. We could refit the model treating `"M"` as the reference group instead of `"L"`:
```{R}
#| prompt: true
moose$strat2 <- factor(moose$strat, levels = c("M", "L"))
update(spbin, formula = presence ~ elev + strat2 + elev:strat2) |> 
  summary()
```

A simpler solution, especially if there categorical variables with many levels, is to leverage [emmeans]{.pkg}. [emmeans]{.pkg} is an [R]{.proglang} package for estimating marginal means of model objects. The [emtrends]{.fct} function in [emmeans]{.pkg} characterizes the effect of a continuous variable (here, `elev`) for each level of a categorical variable (here, `strat`):

```{r}
#| prompt: true
#| eval: false
library("emmeans")
```

```{R}
#| prompt: true
emtrends(spbin, "strat", "elev")
```

The asymptotic confidence intervals show that there is more evidence of an association between elevation and moose presence in the `"L"` stratum than in the `"M"` stratum. Notice that `elev.trend` for the `"L"` stratum matches the `elev` effect when `"L"` is the reference group, and similarly for `elev.trend` when `"M"` is the reference group.

The second [broom]{.pkg} function is [glance]{.fct}, which glances at the model fit:
```{R}
#| prompt: true
glance(spbin)
```

[glance]{.fct} returns several useful statistics like the sample size (`n`), number of fixed effects (`p`), number of covariance parameters (`npar`), several likelihood-based statistics (e.g., `AIC`, `AICc`, `BIC`), and pseudo R-squared.

```{R}
#| prompt: false
#| echo: false
#| label: fig-sp-diagnostic
#| fig-cap: "Spatial logistic regression model diagnostics from [augment]{.fct}. The leverage (i.e., hat) values (left) and standardized residuals (right)."
#| out-width: "100%"
#| out-height: "100%"
knitr::include_graphics(paste0(fig_path, "/figure-2.png"))
```

The third [broom]{.pkg} function is [augment]{.fct}, which augments the model data with diagnostics:
```{R}
#| prompt: true
head(augment(spbin))
```

The [augment]{.fct} function returns the fitted $\mathbf{w}$ values (`.fitted`), deviance residuals (`.resid`), leverage (i.e., hat) values (`.hat`), Cook's distance (`.cooksd`), and standardized residuals (`.std.resid`). When the data are an `sf` object, [augment]{.fct} returns another `sf` object, helpful for visualizing model diagnostics spatially as in @fig-sp-diagnostic. Leverage measures the unusualness of an observation's set of explanatory variables, while Cook's distances measures how influential an observation is on the resulting model fit [@montgomery2021introduction]. Model diagnostics are also accessible as vectors using the appropriate generic function (e.g., [fitted]{.fct}, [residuals]{.fct}).

Similar to [glm]{.fct} model objects, [plot]{.fct} can be used to visualize diagnostics (@fig-sp-diagnostic2):
```{R}
#| prompt: true
#| eval: false
plot(spbin, which = c(4, 7))
```

```{R}
#| prompt: true
#| echo: false
#| label: fig-sp-diagnostic2
#| fig-cap: "Spatial logistic regression model diagnostics from [plot]{.fct}. The Cook's distance values (left) and the fitted spatial covariance as a function of distance (right)."
#| out-width: "100%"
#| out-height: "100%"
knitr::include_graphics(paste0(fig_path, "/figure-3.png"))
```

Components of model variation are partitioned using [varcomp]{.fct}:
```{R}
#| prompt: true
varcomp(spbin)
```
The fixed effects explain roughly 9% of model variation, while the spatially dependent variance explains most of the remaining variability.

```{R}
#| prompt: false
#| echo: false
#| fig-align: center
#| fig-cap: "Moose presence probability fitted values and predictions. Fitted values are represeneted by circles and predictions by triangles."
#| out-width: "70%"
#| out-height: "70%"
#| label: fig-moose-fit
knitr::include_graphics(paste0(fig_path, "/figure-4.png"))
```

We make predictions of the log odds of moose probability presence at each site in `moose_preds` using [predict]{.fct}:
```{R}
#| prompt: true
head(predict(spbin, newdata = moose_preds))
```

[augment]{.fct} may also be used to augment the prediction data with predictions:
```{R}
head(augment(
  spbin,
  newdata = moose_preds,
  type.predict = "response",
  interval = "prediction"
))
```

Here, we requested predictions on the probability (i.e., response) scale (@fig-moose-fit) alongside lower and upper bounds of a 95% (see `level`) prediction interval (@fig-moose-int).

```{R}
#| prompt: false
#| echo: false
#| label: fig-moose-int
#| fig-cap: "Moose presence probability prediction intervals. 95% prediction interval lower bound (left) and 95% prediction interval upper bound (right)."
#| out-width: "100%"
#| out-height: "100%"
knitr::include_graphics(paste0(fig_path, "/figure-5.png"))
```

Thus far we have heuristically argued, based on first principles, that there are benefits to incorporating spatial autocorrelation for GLMs applied to spatial data. Now we provide some empirical evidence to support this claim by comparing the fits of the SPGLM and a GLM. If `spcov_type = "none"`, the resulting model fit is nearly identical to that from [glm]{.fct}:

```{R}
#| prompt: true
# fit nonspatial model using spglm()
bin <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = binomial,
  data = moose,
  spcov_type = "none"
)
# fit nonspatial model using glm()
bin_glm <- glm(
  formula = presence ~ elev + strat + elev:strat,
  family = binomial,
  data = moose
)
# compare fixed effect coefficients and standard errors up to four decimals
data.frame(
  est_none = coef(bin),
  est_glm = coef(bin_glm),
  se_none = sqrt(diag(vcov(bin))),
  se_glm = sqrt(diag(vcov(bin_glm)))
) |>
  apply(2, round, digits = 4)
```

The advantage of using [spglm]{.fct} to fit a model with `spcov_type = "none"` is that it provides access to other [spmodel]{.pkg} functions for model objects (e.g., [glances]{.fct} below) and accounts for the additional terms in the likelihood from Equation$~$\eqref{eq-marginal6}. These additional terms make the likelihood for [spglm]{.fct} and [glm]{.fct} different, though the models convey the same information. A glance at the spatial and nonspatial models reveals:
```{R}
#| prompt: true
glances(spbin, bin)
```

The spatial model has a notably lower AIC, AICc, BIC, and deviance, suggesting it is the superior model. Another model comparison approach is leave-one-out cross validation. In leave-one-out cross validation, separately each observation is held out, a model is fit to the remaining data, and a prediction is made for the mean of the held out observation on the response scale. Then, statistics like leave-one-out bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) may be computed:
```{R}
#| prompt: true
loocv(spbin) |>
  apply(2, round, digits = 4)
loocv(bin) |>
  apply(2, round, digits = 4)
```

Both models are nearly unbiased, but the spatial model has an approximately 39% lower MSPE, suggesting the probability predictions tend to be much closer to the observed presence values.

A third model comparison tool is area under the receiver operating characteristic (AUROC) curve. The AUROC curve ranges from zero to one and conveys a model's classification performance over all possible probability thresholds [@james2013introduction]. Larger values of AUROC indicate a more accurate model:
```{R}
#| prompt: true
AUROC(spbin)
AUROC(bin)
```

All three performance metrics (likelihood-based statistics, leave-one-out statistics, and AUROC) prefer the spatial model.

### Count data

```{R}
#| prompt: false
#| echo: false
#| fig-align: center
#| fig-cap: "Moose counts in Alaska. Circles represent moose counts (based on color) and triangles represent locations at which mean count predictions are desired."
#| label: fig-moose-data-count
#| out-width: "70%"
#| out-height: "70%"
knitr::include_graphics(paste0(fig_path, "/figure-6.png"))
```

The `count` variable in `moose` contains the number of moose observed at a site (@fig-moose-data-count). Count data are often modeled using Poisson or negative binomial regression with the log link function. The Poisson regression model assumes each datum's underlying latent mean equals its variance, while the negative binomial accommodates overdispersion (where the variance is greater than the mean) at the cost of estimating an extra parameter.

So far our spatial models have made an implicit assumption of geometric isotropy. A spatial covariance is geometrically isotropic if its dependence decays with distance equally in all directions. A spatial covariance is geometrically anisotropic if its dependence decays differently in different directions. The geometric anisotropy's directionality and strength are controlled by rotation and scale parameters that are applied to the original coordinates, creating a transformed set of coordinates whose spatial covariance is geometrically isotropic. Geometrically anisotropic models are fit by specifying `anisotropy`:

```{r}
#| prompt: true
sppois <- spglm(
  formula = count ~ elev + strat + elev:strat,
  family = poisson,
  data = moose,
  spcov_type = "gaussian",
  anisotropy = TRUE
)

spnbin <- update(sppois, family = nbinomial)
```

Because the models have the same support (i.e., both non-negative count models), we can use likelihood-based statistics to compare them:
```{r}
#| prompt: true
glances(sppois, spnbin, sort_by = "AIC") |>
  subset(select = c(model, npar, AIC, AICc, BIC))
```

The negative binomial model has a slightly lower AIC and AICc, while the Poisson model has a slightly lower BIC. This is reasonable given the BIC penalizes additional parameters (here, an overdispersion parameter) more heavily than AIC and AICc. The leave-out-out MSPE prefers the negative binomial model:
```{R}
#| prompt: true
loocv(sppois) |>
  apply(2, round, digits = 4)
loocv(spnbin) |>
  apply(2, round, digits = 4)
```

A likelihood-based comparison between the negative binomial anisotropic model and the negative binomial isotropic model suggests that the anisotropic model is preferred (lower AIC, AICc, and BIC):
```{R}
#| prompt: true
spnbin_iso <- update(spnbin, anisotropy = FALSE)
glances(spnbin_iso, spnbin, sort_by = "AIC") |>
  subset(select = c(model, npar, AIC, AICc, BIC))
```

[plot]{.fct} returns the spatial covariance as a function of direction (@fig-tropy):
```{R}
#| prompt: true
#| eval: false
plot(spnbin_iso, which = 8)
plot(spnbin, which = 8)
```

```{R}
#| prompt: true
#| echo: false
#| label: fig-tropy
#| fig-cap: "Level curves of equal autocorrelation for the negative binomial moose count models. The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of autocorrelation.  In the isotropic level curve (left), spatial covariance decays equally in all directions. In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts)."
#| out-width: "100%"
#| out-height: "100%"
#| fig-asp: 1
knitr::include_graphics(paste0(fig_path, "/figure-7.png"))
```

Earlier we used [tidy]{.fct} to tidy the model's fixed effects, but we can also use tidy to tidy the spatial covariance parameters:
```{R}
#| prompt: true
tidy(spnbin, effects = "spcov")
```

The `rotate` parameter is the number of radians in $[0, \pi]$ the ellipse is rotated and the `scale` parameter is the ratio in $(0, 1]$ of the minor axis length to the major axis length. The `is_known` column indicates whether the parameter was assumed known during optimization, controlled by specifying the `spcov_initial` argument.

### Skewed data

The `seal` data in [spmodel]{.pkg} is an `sf` object with data on harbor seal trends in Alaska. The `log_trend` variable is the logarithm of a seal abundance temporal trend measure at the site (based on historical data), and the `stock` variable is a factor with two levels, `8` and `10`, where each level represents one of twelve seal stocks (i.e., breeds) in Alaska. The `seal` geometry is an areal polygon geometry and hence, spatial autoregressive models based on neighborhood distance are appropriate.

SPGLMs for areal data are are fit in [spmodel]{.pkg} using [spgautor]{.fct}, which has similar syntax as [spglm]{.fct} but contains arguments to control the weight matrix ($\mathbf{W}$ in Equation$~$\ref{eq-spcov-sar}) and whether or not row-standardization [@ver2018spatial] is applied.  By default, polygons are neighbors if they share a boundary (i.e., Queen's contiguity; see @pebesma2023spatial) and row standardization is assumed. Weight matrices may be provided via the `W` argument and row standardization may be ignored via the `row_st` arguent. Following @ver2018spatial, polygons without a neighbor are given their own (independent) variance parameter called `extra`.

The trend data were originally logged to remove skew [@ver2018spatial], but we will exponentiate `log_trend` (@fig-seal) and model this skew directly using a SPGLM:
```{R}
#| prompt: true
seal$trend <- exp(seal$log_trend)
```

The `trend` variable has several missing (`NA`) values, which represent polygons at which predictions of `trend` are desired. To make predictions using spatial autoregressive models, the prediction locations must be known prior to model fitting because these locations affect the neighborhood structure of the observed data [@ver2018spatial]. This restriction is notably different than SPGLMs for point-referenced data (i.e., geostatistical models), which completely separates the estimation and prediction steps.

```{r}
#| prompt: true
#| echo: false
#| label: fig-seal
#| fig-cap: "Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed seal trends (right)."
#| out-width: "100%"
#| out-height: "100%"
knitr::include_graphics(paste0(fig_path, "/figure-8.png"))
```

[spmodel]{.pkg} supports the gamma and inverse Gaussian families for modeling skewed, positive response variables. [spmodel]{.pkg} also supports nonspatial random effects specified via the `random` argument, which uses a similar formula syntax as [nlme]{.pkg} [@pinheiro2006mixed] and [lme4]{.pkg} [@bates2015lme4]. Using likelihood-based statistics, we compare two models fit using the simultaneous autoregressive covariance and the Gamma and inverse Gaussian families. Both models have a random effect for seal stock, which builds additional correlation into the model for two polygons from the same stock:
```{r}
#| prompt: true
spgam <- spgautor(
  formula = trend ~ 1,
  family = Gamma,
  data = seal,
  spcov_type = "sar",
  random = ~ stock
)
spinvg <- update(spgam, family = inverse.gaussian)
glances(spgam, spinvg, sort_by = "AIC") |>
  subset(select = c(model, npar, AIC, AICc, BIC))
```

The inverse Gaussian model has a lower AIC, AICc, and BIC, which indicates it is a better fit than the gamma model.

We may [tidy]{.fct} the estimated stock random effect variance:
```{R}
#| prompt: true
tidy(spinvg, effects = "randcov")
```

The locations to predict `trend` (`NA` values) are stored in the `newdata` element of `spinvg` and used for prediction:
```{R}
#| prompt: true
#| results: "hide"
# output omitted
predict(spinvg, type = "response", interval = "prediction")
```

If using [augment]{.fct} for prediction, `newdata` must be specified:
```{r}
#| prompt: true
head(augment(
  spinvg,
  newdata = spinvg$newdata,
  type.predict = "response",
  interval = "prediction"
))
```

### Proportion data

We end with two examples of beta regression for proportion data [@ferrari2004beta]. First, we model the nitrogen percentage in a caribou foraging experiment. Second, we model the proportion of voter turnout by Texas county in the United States (US) 1980 presidential election.

```{R}
#| prompt: false
#| echo: false
#| label: fig-caribou
#| fig-align: center
#| fig-cap: "Caribou data from a spatial experimental design measuring the percentage of nitrogen (`z`) in soil and testing two factors: tarp type and water presence."
#| out-width: "70%"
#| out-height: "70%"
knitr::include_graphics(paste0(fig_path, "/figure-9.png"))
```

The `caribou` data in [spmodel]{.pkg} are a data frame from a caribou foraging experiment in Alaska meant to study the impact of water and tarp cover on the percentage of nitrogen in surrounding plants (@fig-caribou). @lenart2002climate studied these data treating nitrogen percentage as a continuous variable, but here we treat nitrogen percentage (`z`) as a proportion:

```{R}
#| prompt: true
spbeta <- spglm(
    formula = z/100 ~ water + tarp + water:tarp,
    family = "beta",
    data = caribou,
    spcov_type = "matern",
    xcoord = x,
    ycoord = y
)
```

The nitrogen percentage is dynamically scaled in `formula` from (0, 100) to (0, 1) so that it is a proportion. Nitrogen percentage is modeled as a function of `water` (two levels: `"Y"` for water and `"N"` for no water), `tarp` (three levels: `"clear"` for a clear tarp, `"none"` for no tarp, and `"shade"` for a shaded tarp), and their interaction, which lets the effect of water presence vary across tarp type. `caribou` is a data frame (not an `sf` object), so we supply the x-coordinate and y-coordinate directly via `xcoord` and `ycoord`, and, consistent with the tidyverse approach [@wickham2019welcome], column names in `data` do not need to be quoted when referenced (but can be quoted).

A summary of `spbeta` returns a coefficients table that provides parameter estimates relative to a reference group. When factors have more than two levels, it is not straightforward to use these contrasts to determine overall significance of the factor. The `anova()` function tests marginal (i.e., Type III sums of squares) significance of factors using the general linear hypothesis test for spatial (i.e., correlated) data [@schabenberger2017statistical]:

```{r}
tidy(anova(spbeta))
```

These results suggest there is some evidence that the effect of water on nitrogen percentage differ depending on the type of tarp used (0.01 < $p~$value < 0.1).

Sometimes averages or contrasts between factor levels that are not in the reference group are of interest. We again leverage [spmodel]{.pkg}'s built-in support for [emmeans]{.pkg} and use it to obtain the averages of each factor combination on the link (here, logit) scale:
```{r}
spemm <- emmeans(spbeta, ~ water + tarp)
spemm
```

Delta method [@ver2012invented] standard errors are used when averages on the response (here, proportion) scale are desired:
```{r}
update(spemm, type = "response")
```

Pairwise contrasts use a Tukey $p~$value adjustment [@tukey1949comparing] by default. Here, we request no $p~$value adjustment:
```{r}
head(pairs(spemm, adjust = "none"))
```

```{r}
#| prompt: true
#| echo: false
#| label: fig-texas
#| fig-cap: "Proportion of voter turnout in Texas for the 1980 presidential election. Circles represent voter turnout (based on color) and triangles represent locations at which voter turnout predictions are desired."
#| out-width: "70%"
#| out-height: "70%"
knitr::include_graphics(paste0(fig_path, "/figure-10.png"))
```

We now model the `elect80` data in [spData]{.pkg} [@bivand2024spdata], which contains voter turnout data by county in the 1980 US Presidential election [@pace1997quick]. The `texas` data in [spmodel]{.pkg} contains a subset of these data in the state of Texas. These data are point-referenced, but we may still use autoregressive models if neighborhood distance is determined using county centroids (i.e., counties whose centroid distance is less than some cutoff are defined as neighbors). The response variable of interest, `turnout`, is the proportion of registered voters in the county who voted in the election (@fig-texas).

```{R}
#| prompt: true
spgautor_mods <- spgautor(
  formula = turnout ~ log_income,
  family = beta,
  data = texas,
  spcov_type = c("car", "sar"),
  cutoff = 2e5,
  estmethod = "ml"
) 
```

We model voter turnout as a function of log income using both the conditional and simultaneous autoregressive models with a neighbor distance cutoff of 200 kilometers and the maximum likelihood estimation method. When a vector is provided to `spcov_type` in [spgautor]{.fct} (or [spglm]{.fct}), a model is fit for each spatial covariance type and stored in a list with name equal to the respective type. Then it is simple to glance at each model fit:
```{r}
#| prompt: true
glances(spgautor_mods) |>
  subset(select = c(model, npar, AIC, AICc, BIC))
```

The conditional autoregressive model has the best fit (in terms of AIC, AICc, and BIC). In this model, there is significant evidence `log_income` is positively related to average voter turnout ($p~$value < 0.001):
```{R}
tidy(spgautor_mods$car)
```

Another way to assess the impact of `log_income` on `turnout` is a likelihood ratio test:
```{R}
reduced_car <- update(spgautor_mods$car, formula = turnout ~ 1)
tidy(anova(reduced_car, spgautor_mods$car))
```

Likelihood ratio tests compare the fit of a "full" model compared to a "reduced" model that is completely nested within the full model. If there is evidence the full model explains significantly more information than the reduced model, the likelihood ratio test $p~$value will be small. Similar to the summary output from the general linear hypothesis test, the likelihood ratio suggests `log_income` is related to average voter turnout ($p~$value < 0.001).

The default estimation method in [spmodel]{.pkg} is REML, but note that these models used maximum likelihood (ML). ML is very similar to REML -- the difference is that for ML, the fixed effects $\boldsymbol{\beta}$ are not integrated out of Equation$~$\eqref{eq-marginal} but are rather back-substituted. While REML typically performs better for fixed effect estimation and prediction [@zimmerman2024spatial], ML allows likelihood-based comparisons (e.g., AIC) for models with simultaneously varying fixed effect and covariance structures, while REML likelihood-based comparisons are only valid for models sharing the same fixed effect structure (though @gurka2006selecting provides some evidence that this restriction may be unnecessary).

The point-referenced `texas` and `caribou` data may be analyzed using [spgautor]{.fct} or [spglm]{.fct} and comparisons across these structures can be made using likelihood-based statistics (as long as the supports of the response distribution are the same). Put another way, likelihood-based statistics can be used to determine whether geostatistical (distance-based) or autoregressive (neighbor-based) structures perform best when the data are point-referenced.

## Discussion {#sec-discussion}

SPGLMs are fit in [spmodel]{.pkg} using a novel application of the Laplace approximation that marginalizes over the latent (i.e., unobserved) mean, $\mathbf{w}$, and the fixed effects, $\boldsymbol{\beta}$. The approach is quite flexible and accommodates any general response distribution and covariance structure.  @ver2024marginal show that the approach, as implemented in [spmodel]{.pkg}, generally yields unbiased estimators with proper interval coverage and often outperforms the Bayesian approach from [spBayes]{.pkg}, the INLA approach from [R-INLA]{.pkg}, and the automatic differentiation approach from [glmmTMB]{.fct}.

[spmodel]{.pkg}'s [spglm]{.fct} and [spgautor]{.fct} functions are similar in structure and syntax as the base-[R]{.proglang} [glm]{.fct} function, easing the transition from GLMs to SPGLMs. These functions support six response distributions (Table$~$\ref{table-links}) and 20 spatial covariance functions. [spmodel]{.pkg} provides several additional features that accommodate geometric anisotropy, nonspatial random effects, fixing spatial covariance parameters at known values, data having thousands of observations (following @ver2023indexing), incorporating spatial dependence in machine learning (e.g., random forests; @breiman2001random), simulating spatially dependent data (e.g., [sprbinom]{.fct}, [sprpois]{.fct}), and several others. Learn more at [https://CRAN.R-project.org/package=spmodel](https://CRAN.R-project.org/package=spmodel) and links therein.

## Computational details {.unnumbered}

The results in this paper were obtained using [R]{.proglang} 4.4.0 with the
[spmodel]{.pkg} 0.9.0 package. Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.

## Data and code availability {.unnumbered}

All writing and code associated with this manuscript is available for viewing and download on GitHub at [https://github.com/USEPA/spmodel.glm.manuscript](https://github.com/USEPA/spmodel.glm.manuscript). All data used are part of the [spmodel]{.pkg} [R]{.proglang} package available for download from CRAN at [https://CRAN.R-project.org/package=spmodel](https://CRAN.R-project.org/package=spmodel).

## Acknowledgments {.unnumbered}

We would like to thank initial reviewers and editors for feedback that has greatly improved the manuscript.

The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of the U.S. Environmental Protection Agency or the National Oceanic and Atmospheric Administration. Any mention of trade names, products, or services does not imply
an endorsement by the U.S. government, the U.S. Environmental Protection Agency, or the National Oceanic and Atmospheric
Administration. The U.S. Environmental Protection Agency and the National Oceanic and Atmospheric Administration do not endorse
any commercial products, services or enterprises.

## References {.unnumbered}

:::{#refs}

:::

### 2024-11-19 inst/manuscript/manuscript.tex
initial commit

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  article]{jss}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{orcidlink,thumbpdf,lmodern}

\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}{}
\makeatother
\makeatletter
\makeatother
\makeatletter
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, breakable]}{\end{tcolorbox}}\fi
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Spatial Generalized Linear Models in  Using },
  pdfauthor={Michael Dumelle; Jay M. Ver Hoef; Matt Higham},
  pdfkeywords={Autoregressive Model, Geostatistics, Spatial
Covariance, Spatially-Explicit Model, Statistical Model},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

%% -- Article metainformation (author, title, ...) -----------------------------

%% Author information
\author{Michael Dumelle~\orcidlink{0000-0002-3393-5529}\\United States\\
Environmental Protection Agency \And Jay M. Ver
Hoef~\orcidlink{0000-0003-4302-6895}\\National Oceanic and\\
Atmospheric Administration \AND Matt
Higham~\orcidlink{0009-0006-4217-625X}\\St.~Lawrence University}
\Plainauthor{Michael Dumelle, Jay M. Ver Hoef, Matt
Higham} %% comma-separated

\title{Spatial Generalized Linear Models in \proglang{R} Using
\pkg{spmodel}}
\Plaintitle{Spatial Generalized Linear Models in
Using} %% without formatting

%% an abstract and keywords
\Abstract{Generalized linear models (GLMs) describe a non-normal
response variable that may be binary, count, skewed, or a proportion.
Typically, observations in a GLM are assumed independent of one another.
For spatial data, this independence assumption is impractical, as nearby
locations tend to be more similar than locations far apart. The
\pkg{spmodel} \proglang{R} package provides tools to fit GLMs that
incorporate spatial autocorrelation (i.e., spatial generalized linear
models, or SPGLMs). SPGLMs are fit in \pkg{spmodel} using a novel
application of the Laplace approximation via \fct{spglm} for
point-referenced data or \fct{spgautor} for areal (i.e., lattice), data.
\fct{spglm} and \fct{spgautor} closely resemble \fct{glm} from base
\proglang{R} but include arguments that control the spatial
autocorrelation structure. \pkg{spmodel} has many helper functions for
model inspection and diagnostics, some of which leverage other
\proglang{R} packages like \pkg{broom} and \pkg{emmeans}. \pkg{spmodel}
has tools to make predictions of the latent spatial-mean process at
unobserved locations. \pkg{spmodel} also provides many advanced features
like accommodating geometric anisotropy and nonspatial random effects,
simulating spatially autocorrelated data, and more. Here we use
\pkg{spmodel} to illustrate the modeling of binary, count, skewed and
proportion response variables from several point-referenced and areal
data sets.}

%% at least one keyword must be supplied
\Keywords{Autoregressive Model, Geostatistics, Spatial
Covariance, Spatially-Explicit Model, Statistical Model}
\Plainkeywords{Autoregressive Model, Geostatistics, Spatial
Covariance, Spatially-Explicit Model, Statistical Model}

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}
%% \setcounter{page}{1}
%% \Pages{1--xx}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Michael Dumelle\\
Pacific Ecological Systems Division\\
200 SW 35th St\\
Corvallis OR\\
E-mail: \email{dumelle.michael@epa.gov}\\
\\~
Jay M. Ver Hoef\\
Alaska Fisheries Science Center\\
\\~
Matt Higham\\
Department of Math, Computer Science, and Statistics\\
\\~

}

\begin{document}
\maketitle

\newpage

\section{Introduction}\label{sec-intro}

Binary, count, proportion, and skewed data are ubiquitous in practice.
These data types are naturally modeled using a generalized linear model
(GLM) framework
\citep{nelder1972generalized, mccullagh1989generalized, myers2012generalized, faraway2016extending}.
In a GLM, a response variable \(\text{y}\) belongs to a particular
statistical distribution with mean \(\mu\). For example, \(\text{y}\)
may be distributed as a Poisson random variable with some mean \(\mu\).
Based on the response distribution assumed for \(\text{y}\), GLMs link a
function of \(\mu\) to explanatory variables via a link function:
\begin{equation}\label{eq-glm}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
\end{equation} where for a sample size \(n\), \(\boldsymbol{\mu}\) is
the mean of an \(n \times 1\) response vector \(\mathbf{y}\),
\(f(\boldsymbol{\mu})\) is the link function that connects
\(\boldsymbol{\mu}\) to \(\mathbf{w}\), \(\mathbf{X}\) is the
\(n \times p\) design matrix of explanatory variables, and
\(\boldsymbol{\beta}\) is the \(p \times 1\) vector of fixed effects.
The mean, \(\boldsymbol{\mu}\), is usually constrained in some way
(e.g., positive) that depends on the distribution assumed for
\(\mathbf{y}\), but \(\mathbf{w}\) is unconstrained. The parameters in
Equation\(~\)\eqref{eq-glm} are typically estimated via maximum
likelihood (e.g., iteratively reweighted least squares)
\citep{chambers1992S}. The \fct{glm} function is commonly used to fit
GLMs in the \proglang{R} programming language \citep{rcore2024}.

The GLM framework in Equation\(~\)\eqref{eq-glm} assumes the elements of
\(\mathbf{y}\) are independent of one another. This assumption is
impractical for spatial data, where nearby observations tend to be more
similar than distant observations \citep{tobler1970computer}. Ignoring
this spatial dependence can give rise to misleading inference and poor
prediction \citep{zimmerman2024spatial}. Spatial GLMs (SPGLMs) formally
incorporate spatial autocorrelation into a GLM by adding to
Equation\(~\)\eqref{eq-glm} two random effects that elucidate spatial
structure: \begin{equation}\label{eq-spglm}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
\end{equation} where \(\boldsymbol{\tau}\) is an \(n \times 1\) column
vector of spatially dependent random errors, and
\(\boldsymbol{\epsilon}\) is an \(n \times 1\) column vector of
spatially independent random errors. We make a few assumptions about
\(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\): first, that
\(\text{E}(\boldsymbol{\tau}) = \text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}\),
where \(\text{E}(\cdot)\) denotes expectation; second, that
\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}\), where
\(\mathbf{R}\) is an \(n \times n\) matrix that determines the spatial
dependence structure in \(\mathbf{w}\) (which is linked to
\(\boldsymbol{\mu}\) and hence \(\mathbf{y}\)) and depends on a range
parameter, \(\phi\); third, that
\(\text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_\epsilon \mathbf{I}\),
where \(\mathbf{I}\) is an \(n \times n\) identity matrix; and fourth,
that \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\) are independent
of one another. The parameter \(\sigma^2_{\tau}\) is called the
spatially dependent random error variance or partial sill. The parameter
\(\sigma^2_\epsilon\) is called the spatially independent random error
variance or nugget. These two variance parameters are henceforth more
intuitively written as \(\sigma^2_{de}\) and \(\sigma^2_{ie}\),
respectively. The covariance of \(\mathbf{w}\) is denoted
\(\boldsymbol{\Sigma}\) and is given by \begin{equation}\label{eq-spcov}
 \boldsymbol{\Sigma} = \sigma^2_{de}\mathbf{R} + \sigma^2_{ie} \mathbf{I}.
\end{equation} The parameters \(\sigma^2_{de}\), \(\phi\), and
\(\sigma^2_{ie}\) are elements of \(\boldsymbol{\theta}\), the
covariance parameter vector.

The \pkg{spmodel} \proglang{R} package provides tools for fitting
spatial statistical models and making predictions at unobserved
locations \citep{dumelle2023spmodel}. A recent major update (v0.4.0) to
\pkg{spmodel} added SPGLM (Equation\(~\)\ref{eq-spglm}) support for
binary, count, skewed, and proportion response variables
(Table\(~\)\ref{table-links}), greatly expanding the class of models
that \pkg{spmodel} makes accessible to practitioners.

\begin{table}

\centering{

\centering
\begin{tabular}{||c|ccc||}
 \hline
 Family & Link Function & Link Name & Data Type  \\ [0.5ex]
 \hline \hline
 Binomial & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Binary; Binary Count \\
 Poisson & $f(\mu) = \log(\mu)$ & Log & Count \\
 Negative Binomial & $f(\mu) = \log(\mu)$ & Log & Count \\
 Beta & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Proportion \\
 Gamma & $f(\mu) = \log(\mu)$ & Log & Skewed \\
 Inverse Gaussian & $f(\mu) = \log(\mu)$ & Log & Skewed \\
 \hline\hline
\end{tabular}

}

\caption{\label{table-links}SPGLM response distributions and their link functions and data types.}

\end{table}%

SPGLMs for point-referenced data are called geostatistical GLMs and are
fit using \fct{spglm}. Data are point-referenced when the elements in
\(\mathbf{y}\) are observed at point-locations indexed by x-coordinates
and y-coordinates on a spatially continuous surface with an infinite
number of locations (e.g., point locations in a field). SPGLMs for areal
data are called spatial autoregresive models and are fit using
\fct{spgautor}. Data are areal (i.e., lattice) when they are part of a
finite network of polygons whose connections are indexed by a
neighborhood structure (e.g., states in a country).

Several other \proglang{R} packages exist for analyzing SPGLMs. The
\pkg{brms} \citep{burkner2017brms}, \pkg{carBayes}
\citep{lee2013carbayes}, \pkg{ngspatial} \citep{hughes2020ngspatial},
\pkg{R-INLA} \citep{lindgren2015bayesian}, \pkg{spBayes}
\citep{finley2007spbayes}, and \pkg{spNNGP} \citep{finley2002spnngp}
packages take a Bayesian approach, either directly sampling from
posterior distributions of parameters (e.g., using MCMC) or
approximating them. A benefit of Bayesian approaches is that prior
information can be incorporated and uncertainty quantification of
parameter estimates is straightforward. However, Bayesian approaches,
especially those using MCMC, tend to be computationally expensive. In
order to reduce computation time, many of these packages work with the
precision matrix instead of the covariance matrix so that inverting
matrices (a computational bottleneck) is not required. For example,
\pkg{R-INLA} uses the precision matrix (often directly modeled as with
autoregressive models) and tends to be very fast. Working with precision
matrices, however, can be more restrictive and less intuitive than
working directly with the covariance matrix. The \pkg{FRK}
\citep{sainsbury2024modeling}, \pkg{glmmTMB} \citep{brooks2017glmmtmb},
\pkg{hglm} \citep{ronnegard2010hglm}, \pkg{mgcv}
\citep{wood2017generalized}, and \pkg{spaMM} \citep{rousset2014spamm}
packages directly use Laplace, quasi-likelihood, or reduced-rank
approaches to estimate parameters. These direct approaches tend to be
computationally efficient, as they don't rely on MCMC sampling. In
contrast to the Bayesian approach, a drawback of these direct approaches
is that prior information cannot be formally incorporated and covariance
parameter uncertainty is challenging to quantify. SPGLMs in
\pkg{spmodel} are most similar to those in \pkg{glmmTMB} --
\pkg{spmodel} uses analytical solutions for maximizing the likelihood
(which we describe in the next section), while \pkg{glmmTMB} uses
automatic differentiation for maximizing the likelihood (which tends to
be much slower than analytical solutions).

Missing from the aforementioned \proglang{R} packages is the complete
set of tools for SPGLMs that \pkg{spmodel} provides. Importantly, the
\fct{spglm} and \fct{spautor} functions act as a spatial analogue to the
familiar \fct{glm} from base \proglang{R}, making the transition from
GLMs to SPGLMs relatively seamless. \pkg{spmodel} leverages many
commonly used \proglang{R} generics like \fct{summary} to better
understand fitted models. Six GLM families (Table\(~\)\ref{table-links})
and 20 spatial covariance (or precision) functions are supported. Also
available are functions for data visualization, model fitting, model
summaries, model diagnostics, model comparison, and prediction, all
crucial components of a data analysis. Through extra function arguments,
\pkg{spmodel} supports advanced features like geometric anisotropy,
non-spatial random effects, methods for large data sets, and more.
Importantly, \pkg{spmodel} extends popular packages like \pkg{broom}
\citep{robinson2021broom, kuhn2022tidy} and \pkg{emmeans}
\citep{lenth2024emmeans}. \pkg{spmodel} also provides function for
simulating SPGLM data (e.g., \fct{sprbinom}, \fct{sprpois}).

The rest of this article is organized as follows. In
Section~\ref{sec-spglm}, we provide some background for the SPGLM
fitting and prediction routines in \pkg{spmodel}. In
Section~\ref{sec-applications}, we provide several applications of
\pkg{spmodel} to spatial binary, count, skewed and proportion data with
both point-referenced and areal supports. And in
Section~\ref{sec-discussion}, we end with a discussion synthesizing
\pkg{spmodel}'s contributions to the analysis of SPGLMs in \proglang{R}.

\section{Spatial generalized linear models using the Laplace
approximation}\label{sec-spglm}

GLMs with random effects are often written hierarchically
\citep{lee1996hierarchical, bolker2009generalized, wood2017generalized}.
\pkg{spmodel} leverages this hierarchical structure and uses a novel
application of the Laplace approximation \citep{ver2024marginal} to fit
models. This marginal approach is quite flexible, accommodating a wide
range of possible dependence structures and formally maximizing a
likelihood. Maximizing a likelihood yields convenient likelihood-based
statistics like AIC \citep{akaike1974new}, AICc
\citep{hoeting2006model}, BIC \citep{schwarz1978estimating}, deviance
\citep{mccullagh1989generalized}, and likelihood ratio tests for model
comparison, a benefit compared to quasi-likelihood
\citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
approaches \citep{wolfinger1993generalized}, which only specify the
first two moments of a distribution. \citet{ver2024marginal} provide
thorough context for the marginal approach and its associated details,
but next we provide an short overview of the methodology.

Our goal is to marginalize over the latent mean \(\mathbf{w}\) and fixed
effects \(\boldsymbol{\beta}\) in Equation\(~\)\eqref{eq-spglm} to
obtain a response distribution for \(\mathbf{y}\) that depends only on
the explanatory variables, \(\mathbf{X}\), the response distribution's
dispersion parameter, \(\varphi\), and the covariance parameters,
\(\boldsymbol{\theta}\). We can represent this marginal distribution
hierarchically as \begin{equation}\label{eq-marginal}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w} .
\end{equation} In Equation\(~\)\eqref{eq-marginal},
\([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) is the density for
\(\mathbf{y}\) (e.g., binomial, Poisson) given the latent mean and
dispersion parameter and
\([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\) is the Gaussian
density for \(\mathbf{w}\) given the explanatory variables, fixed
effects, and covariance parameters. Integrating \(\boldsymbol{\beta}\)
out of Equation\(~\)\eqref{eq-spglm} yields
\begin{equation}\label{eq-marginal2}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\mathbf{w},
\end{equation} where \([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\)
is the restricted (i.e., residual) Gaussian density
\citep[\citet{harville1977maximum},
\citet{wolfinger1994computing}]{patterson1971recovery} for
\(\mathbf{w}\) given the explanatory variables and covariance
parameters. This restricted Gaussian density is given by
\begin{equation}\label{eq-reml-def}
[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] = \frac{\exp(-\frac{1}{2}(\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}}) \boldsymbol{\Sigma}^{-1} (\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}})^T)}{(2 \pi)^{(n - p)/2} |\boldsymbol{\Sigma}|^{1/2}|\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X}|^{1/2}},
\end{equation} where
\(\tilde{\boldsymbol{\beta}} = (\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{w}\)
and \(|\cdot|\) denotes the determinant.

Next, let
\(\ell_\mathbf{w} = \log([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}])\)
and then notice that Equation\(~\)\eqref{eq-marginal2} can be written as
\begin{equation}\label{eq-marginal3}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \exp(\ell_\mathbf{w}) d\mathbf{w}.
\end{equation} A Taylor series expansion of \(\ell_\mathbf{w}\) around a
point \(\mathbf{a}\) yields \begin{equation}\label{eq-marginal4}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_\mathbf{a} + \mathbf{g}^T(\mathbf{w} - \mathbf{a}) + \frac{1}{2}(\mathbf{w} - \mathbf{a})^T \mathbf{G} (\mathbf{w} - \mathbf{a}))d\mathbf{w},
\end{equation} where \(\mathbf{g}\) and \(\mathbf{G}\) are the gradient
and Hessian, respectively, of \(\ell_\mathbf{w}\) with respect to
\(\mathbf{w}\). If \(\mathbf{a}\) is a value for which
\(\mathbf{g} = \mathbf{0}\), \begin{equation}\label{eq-marginal5}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_\mathbf{a}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \mathbf{a})^T (-\mathbf{G}) (\mathbf{w} - \mathbf{a}))d\mathbf{w}.
\end{equation} Notice that the integral in
Equation\(~\)\ref{eq-marginal5} can be solved in the same manner as the
normalizing constant in a multivariate Gaussian distribution, and
rewriting \(\exp(\ell_\mathbf{a})\) yields
\begin{equation}\label{eq-marginal6}
  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{a}), \varphi] [\mathbf{a} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{a}}|^{-1/2}.
\end{equation} \citet{ver2024marginal} show how to evaluate
\(\mathbf{g}\) and \(\mathbf{G}\) to obtain \textbf{a} and maximize (the
natural logarithm of) Equation\(~\)\eqref{eq-marginal6} for the six
response distributions in Table\(~\)\ref{table-links}. Maximizing
Equation\(~\)\eqref{eq-marginal6} using a Newton-Rhapson approach yields
the marginal restricted maximum likelihood estimators \(\hat{\varphi}\)
and \(\hat{\boldsymbol{\theta}}\). \citet{ver2024marginal} show that the
fixed effect estimator is
\(\hat{\boldsymbol{\beta}} = (\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{X})^{-1}\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{w}\),
where \(\hat{\boldsymbol{\Sigma}}\) is the covariance matrix
\(\boldsymbol{\Sigma}\) evaluated at \(\hat{\boldsymbol{\theta}}\).
\citet{ver2024marginal} also derive the covariance matrix of
\(\hat{\boldsymbol{\beta}}\) and show how to predict \(\mathbf{w}\) at
unobserved locations and quantify their uncertainties.

\section{Application}\label{sec-applications}

The \fct{spglm} (for point-referenced data) and \fct{spgautor} (for
areal data) functions in \pkg{spmodel} fit SPGLMs using the Laplace
approximation outlined in Section~\ref{sec-spglm}. Both \fct{spglm} and
\fct{spgautor} generally require the following four arguments:
\texttt{formula}, a formula that describes the relationship between the
response variable and explanatory variables; \texttt{family}, the
response distribution (which can be \texttt{binomial}, \texttt{poisson},
\texttt{nbinomial}, \texttt{Gamma}, \texttt{inverse.gaussian}, or
\texttt{beta}); \texttt{data}, the data frame that holds the variables
in \texttt{formula} as well as spatial locations; and
\texttt{spcov\_type}, the spatial covariance type. The first three
arguments are shared by \fct{glm}; thus, the transition from GLMs to
SPGLMs requires only one additional argument: \texttt{spcov\_type}. The
\fct{spglm} spatial covariance types measure dependence as a function of
Euclidean distance among observations; an example is the exponential
spatial covariance: \begin{equation}\label{eq-spcov-exp}
  \boldsymbol{\Sigma} = \sigma^2_{de} \exp(-\mathbf{H}/\phi) + \sigma^2_{ie}\mathbf{I},
\end{equation} where \(\mathbf{H}\) is a matrix of pairwise distances
among all observations. \fct{spglm} currently supports 18 distinct
spatial covariance functions.

The \fct{spgautor} spatial covariance types measure dependence as a
function of neighborhood distance among observations; an example is the
simultaneous autoregressive covariance matrix:
\begin{equation}\label{eq-spcov-sar}
  \boldsymbol{\Sigma} = \sigma^2_{de} [(\mathbf{I} - \phi \mathbf{W})(\mathbf{I} - \phi \mathbf{W})^T]^{-1} + \sigma^2_{ie}\mathbf{I},
\end{equation} where \(\mathbf{W}\) is a matrix that represents the
neighborhood structure among all observations. \fct{spgautor} currently
supports two distinct spatial covariance functions.

In the rest of this section, we use \pkg{spmodel} to study binary,
count, skewed, and proportion response variables that are either
point-referenced or areal. We use \pkg{spmodel} for all parts of the
data analysis, from estimation to inference to model diagnostics to
prediction. We first describe core \pkg{spmodel} functionality in an
application to binary data, while additional analyses highlight count,
skewed, and proportion data as well as some additional \pkg{spmodel}
features. Before proceeding, load \pkg{spmodel} into the current
\proglang{R} session:

\begin{verbatim}
R> library("spmodel")
\end{verbatim}

\subsection{Binary data}\label{binary-data}

The \texttt{moose} data in \pkg{spmodel} contain information on moose
presence in Alaska. They are an \texttt{sf} object, a special data frame
that is supplemented with spatial information using the \pkg{sf} package
\citep{pebesma2018sf}. The first few rows look like:

\begin{verbatim}
R> head(moose)
\end{verbatim}

\begin{verbatim}
Simple feature collection with 6 features and 4 fields
Geometry type: POINT
Dimension:     XY
Bounding box:  xmin: 281896.4 ymin: 1518398 xmax: 311325.3 ymax: 1541016
Projected CRS: NAD83 / Alaska Albers
# A tibble: 6 x 5
   elev strat count presence           geometry
  <dbl> <chr> <dbl> <fct>           <POINT [m]>
1  469. L         0 0        (293542.6 1541016)
2  362. L         0 0        (298313.1 1533972)
3  173. M         0 0        (281896.4 1532516)
4  280. L         0 0        (298651.3 1530264)
5  620. L         0 0        (311325.3 1527705)
6  164. M         0 0        (291421.5 1518398)
\end{verbatim}

There are five columns: \texttt{elev}, the numeric site elevation
(meters); \texttt{strat} a stratification variable for sampling with two
levels, \texttt{"L"} and \texttt{"M"}, which are categorized by
landscape metrics at each site; \texttt{count}, the number of moose at
each site; \texttt{presence}, a factor that indicates whether at least
one moose was observed at each site (\texttt{0} implies no moose;
\texttt{1} implies at least one moose); and \texttt{geometry}, the NAD83
projected coordinate of each site. The \texttt{moose\_preds} data in
\pkg{spmodel} contain spatial locations at which predictions of moose
presence are desired. They are also an \texttt{sf} object with the same
projection and measurements for \texttt{elev} and \texttt{strat}.
Figure~\ref{fig-moose-data} shows the \texttt{presence} variable in
\texttt{moose} as well as the spatial locations of both \texttt{moose}
and \texttt{moose\_preds}. Moose are most common in the southwestern and
eastern parts of the domain and least common in the northwest.

\begin{figure}

\centering{

\includegraphics[width=0.7\textwidth,height=0.7\textheight]{figures/figure-1.png}

}

\caption{\label{fig-moose-data}Moose presence in Alaska. Circles
represent moose presence or absence (based on color) and triangles
represent locations at which moose presence probability predictions are
desired.}

\end{figure}%

To study the effect of elevation, stratum, and their interaction on
moose presence while accounting for spatial autocorrelation, we fit a
SPGLM for binary data (i.e., a spatial logistic regression model) using
\fct{spglm}:

\begin{verbatim}
R> spbin <- spglm(
+  formula = presence ~ elev + strat + elev:strat,
+  family = binomial,
+  data = moose,
+  spcov_type = "spherical"
+)
\end{verbatim}

Summarizing the model object yields a summary similar to that provided
by the familiar \fct{glm}:

\begin{verbatim}
R> summary(spbin)
\end{verbatim}

\begin{verbatim}

Call:
spglm(formula = presence ~ elev + strat + elev:strat, family = binomial,
    data = moose, spcov_type = "spherical")

Deviance Residuals:
    Min      1Q  Median      3Q     Max
-1.8423 -0.7538  0.3883  0.7604  1.6018

Coefficients (fixed):
             Estimate Std. Error z value Pr(>|z|)
(Intercept) -3.039992   1.205695  -2.521  0.01169 *
elev         0.009133   0.004126   2.213  0.02687 *
stratM       3.276511   1.162603   2.818  0.00483 **
elev:stratM -0.010882   0.006697  -1.625  0.10418
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Pseudo R-squared: 0.08845

Coefficients (spherical spatial covariance):
       de        ie     range
5.083e+00 2.580e-03 5.158e+04

Coefficients (Dispersion for binomial family):
dispersion
         1
\end{verbatim}

The summary contains the original function call, a summary of residuals,
the fixed effects coefficients table, the spatial covariance parameter
estimates, and additional model information like the pseudo R-squared,
which quantifies the variability in the model attributable to the fixed
effects. While useful, this summary information is hard to work with, as
it is printed directly to the \proglang{R} console. The \pkg{broom}
package from the tidymodels \citep{kuhn2022tidy} ecosystem has functions
to provide helpful model output in the form of tibbles (i.e., data
frames) that are easily manipulated. \pkg{spmodel} has methods for the
\fct{tidy}, \fct{glance}, and \fct{augment} functions from \pkg{broom}.
The first \pkg{broom} function is \fct{tidy}, which tidies the model
output:

\begin{verbatim}
R> tidy(spbin)
\end{verbatim}

\begin{verbatim}
# A tibble: 4 x 5
  term        estimate std.error statistic p.value
  <chr>          <dbl>     <dbl>     <dbl>   <dbl>
1 (Intercept) -3.04      1.21        -2.52 0.0117
2 elev         0.00913   0.00413      2.21 0.0269
3 stratM       3.28      1.16         2.82 0.00483
4 elev:stratM -0.0109    0.00670     -1.62 0.104
\end{verbatim}

The estimates and standard errors returned are on the log odds link
(Table\(~\)\ref{table-links}) scale (\fct{coef} and \fct{vcov} may also
be used). The output provides evidence that elevation is positively
associated with moose presence in the \texttt{"L"} stratum (\(p-\)value
\(<0.05\)) and, at zero elevation, moose are more likely in the
\texttt{"M"} stratum than the \texttt{"L"} stratum (\(p-\)value
\(<0.01\)). This output provides marginal evidence that the effect of
elevation on moose presence varies across strata (\(p-\)value
\(\approx 0.1\)). The model effectively quantifies the impact of
elevation on moose presence for moose in the \texttt{"L"} strata, but an
analogous statement for moose in the \texttt{"M"} strata requires more
context. We could refit the model treating \texttt{"M"} as the reference
group instead of \texttt{"L"}:

\begin{verbatim}
R> moose$strat2 <- factor(moose$strat, levels = c("M", "L"))
R> update(spbin, formula = presence ~ elev + strat2 + elev:strat2) |>
+  summary()
\end{verbatim}

\begin{verbatim}

Call:
spglm(formula = presence ~ elev + strat2 + elev:strat2, family = binomial,
    data = moose, spcov_type = "spherical")

Deviance Residuals:
    Min      1Q  Median      3Q     Max
-1.8423 -0.7538  0.3883  0.7604  1.6018

Coefficients (fixed):
              Estimate Std. Error z value Pr(>|z|)
(Intercept)   0.236519   1.305198   0.181  0.85620
elev         -0.001750   0.006090  -0.287  0.77385
strat2L      -3.276511   1.162603  -2.818  0.00483 **
elev:strat2L  0.010882   0.006697   1.625  0.10418
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Pseudo R-squared: 0.08845

Coefficients (spherical spatial covariance):
       de        ie     range
5.083e+00 2.580e-03 5.158e+04

Coefficients (Dispersion for binomial family):
dispersion
         1
\end{verbatim}

A simpler solution, especially if there categorical variables with many
levels, is to leverage \pkg{emmeans}. \pkg{emmeans} is an \proglang{R}
package for estimating marginal means of model objects. The
\fct{emtrends} function in \pkg{emmeans} characterizes the effect of a
continuous variable (here, \texttt{elev}) for each level of a
categorical variable (here, \texttt{strat}):

\begin{verbatim}
R> library("emmeans")
\end{verbatim}

\begin{verbatim}
R> emtrends(spbin, "strat", "elev")
\end{verbatim}

\begin{verbatim}
 strat elev.trend      SE  df asymp.LCL asymp.UCL
 L        0.00913 0.00413 Inf   0.00105    0.0172
 M       -0.00175 0.00609 Inf  -0.01369    0.0102

Degrees-of-freedom method: asymptotic
Confidence level used: 0.95
\end{verbatim}

The asymptotic confidence intervals show that there is more evidence of
an association between elevation and moose presence in the \texttt{"L"}
stratum than in the \texttt{"M"} stratum. Notice that
\texttt{elev.trend} for the \texttt{"L"} stratum matches the
\texttt{elev} effect when \texttt{"L"} is the reference group, and
similarly for \texttt{elev.trend} when \texttt{"M"} is the reference
group.

The second \pkg{broom} function is \fct{glance}, which glances at the
model fit:

\begin{verbatim}
R> glance(spbin)
\end{verbatim}

\begin{verbatim}
# A tibble: 1 x 10
      n     p  npar value   AIC  AICc   BIC logLik deviance pseudo.r.squared
  <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>            <dbl>
1   218     4     3  681.  687.  687.  697.  -340.     161.           0.0885
\end{verbatim}

\fct{glance} returns several useful statistics like the sample size
(\texttt{n}), number of fixed effects (\texttt{p}), number of covariance
parameters (\texttt{npar}), several likelihood-based statistics (e.g.,
\texttt{AIC}, \texttt{AICc}, \texttt{BIC}), and pseudo R-squared.

\begin{figure}

\centering{

\includegraphics[width=1\textwidth,height=1\textheight]{figures/figure-2.png}

}

\caption{\label{fig-sp-diagnostic}Spatial logistic regression model
diagnostics from \fct{augment}. The leverage (i.e., hat) values (left)
and standardized residuals (right).}

\end{figure}%

The third \pkg{broom} function is \fct{augment}, which augments the
model data with diagnostics:

\begin{verbatim}
R> head(augment(spbin))
\end{verbatim}

\begin{verbatim}
Simple feature collection with 6 features and 8 fields
Geometry type: POINT
Dimension:     XY
Bounding box:  xmin: 281896.4 ymin: 1518398 xmax: 311325.3 ymax: 1541016
Projected CRS: NAD83 / Alaska Albers
# A tibble: 6 x 9
  presence  elev strat .fitted .resid    .hat   .cooksd .std.resid
  <fct>    <dbl> <chr>   <dbl>  <dbl>   <dbl>     <dbl>      <dbl>
1 0         469. L      -1.47  -0.644 0.101   0.0130        -0.679
2 0         362. L      -2.77  -0.349 0.0166  0.000523      -0.352
3 0         173. M      -2.23  -0.451 0.00390 0.000200      -0.452
4 0         280. L      -3.59  -0.234 0.00343 0.0000472     -0.234
5 0         620. L      -0.774 -0.871 0.319   0.130         -1.06
6 0         164. M      -2.01  -0.502 0.00459 0.000292      -0.503
# i 1 more variable: geometry <POINT [m]>
\end{verbatim}

The \fct{augment} function returns the fitted \(\mathbf{w}\) values
(\texttt{.fitted}), deviance residuals (\texttt{.resid}), leverage
(i.e., hat) values (\texttt{.hat}), Cook's distance (\texttt{.cooksd}),
and standardized residuals (\texttt{.std.resid}). When the data are an
\texttt{sf} object, \fct{augment} returns another \texttt{sf} object,
helpful for visualizing model diagnostics spatially as in
Figure~\ref{fig-sp-diagnostic}. Leverage measures the unusualness of an
observation's set of explanatory variables, while Cook's distances
measures how influential an observation is on the resulting model fit
\citep{montgomery2021introduction}. Model diagnostics are also
accessible as vectors using the appropriate generic function (e.g.,
\fct{fitted}, \fct{residuals}).

Similar to \fct{glm} model objects, \fct{plot} can be used to visualize
diagnostics (Figure~\ref{fig-sp-diagnostic2}):

\begin{verbatim}
R> plot(spbin, which = c(4, 7))
\end{verbatim}

\begin{figure}

\centering{

\includegraphics[width=1\textwidth,height=1\textheight]{figures/figure-3.png}

}

\caption{\label{fig-sp-diagnostic2}Spatial logistic regression model
diagnostics from \fct{plot}. The Cook's distance values (left) and the
fitted spatial covariance as a function of distance (right).}

\end{figure}%

Components of model variation are partitioned using \fct{varcomp}:

\begin{verbatim}
R> varcomp(spbin)
\end{verbatim}

\begin{verbatim}
# A tibble: 3 x 2
  varcomp            proportion
  <chr>                   <dbl>
1 Covariates (PR-sq)   0.0885
2 de                   0.911
3 ie                   0.000462
\end{verbatim}

The fixed effects explain roughly 9\% of model variation, while the
spatially dependent variance explains most of the remaining variability.

\begin{figure}

\centering{

\includegraphics[width=0.7\textwidth,height=0.7\textheight]{figures/figure-4.png}

}

\caption{\label{fig-moose-fit}Moose presence probability fitted values
and predictions. Fitted values are represeneted by circles and
predictions by triangles.}

\end{figure}%

We make predictions of the log odds of moose probability presence at
each site in \texttt{moose\_preds} using \fct{predict}:

\begin{verbatim}
R> head(predict(spbin, newdata = moose_preds))
\end{verbatim}

\begin{verbatim}
          1           2           3           4           5           6
 0.08588581 -0.40762380 -1.87510889 -1.14172781  1.45701519 -2.74275553
\end{verbatim}

\fct{augment} may also be used to augment the prediction data with
predictions:

\begin{verbatim}
head(augment(
  spbin,
  newdata = moose_preds,
  type.predict = "response",
  interval = "prediction"
))
\end{verbatim}

\begin{verbatim}
Simple feature collection with 6 features and 5 fields
Geometry type: POINT
Dimension:     XY
Bounding box:  xmin: 291839.8 ymin: 1436192 xmax: 401239.6 ymax: 1512103
Projected CRS: NAD83 / Alaska Albers
# A tibble: 6 x 6
   elev strat .fitted  .lower .upper           geometry
  <dbl> <chr>   <dbl>   <dbl>  <dbl>        <POINT [m]>
1  143. L      0.521  0.0983   0.916 (401239.6 1436192)
2  324. L      0.399  0.0331   0.928 (352640.6 1490695)
3  158. L      0.133  0.00957  0.709 (360954.9 1491590)
4  221. M      0.242  0.0261   0.792 (291839.8 1466091)
5  209. M      0.811  0.289    0.978 (310991.9 1441630)
6  218. L      0.0605 0.00360  0.534 (304473.8 1512103)
\end{verbatim}

Here, we requested predictions on the probability (i.e., response) scale
(Figure~\ref{fig-moose-fit}) alongside lower and upper bounds of a 95\%
(see \texttt{level}) prediction interval (Figure~\ref{fig-moose-int}).

\begin{figure}

\centering{

\includegraphics[width=1\textwidth,height=1\textheight]{figures/figure-5.png}

}

\caption{\label{fig-moose-int}Moose presence probability prediction
intervals. 95\% prediction interval lower bound (left) and 95\%
prediction interval upper bound (right).}

\end{figure}%

Thus far we have heuristically argued, based on first principles, that
there are benefits to incorporating spatial autocorrelation for GLMs
applied to spatial data. Now we provide some empirical evidence to
support this claim by comparing the fits of the SPGLM and a GLM. If
\texttt{spcov\_type\ =\ "none"}, the resulting model fit is nearly
identical to that from \fct{glm}:

\begin{verbatim}
R> # fit nonspatial model using spglm()
R> bin <- spglm(
+  formula = presence ~ elev + strat + elev:strat,
+  family = binomial,
+  data = moose,
+  spcov_type = "none"
+)
R> # fit nonspatial model using glm()
R> bin_glm <- glm(
+  formula = presence ~ elev + strat + elev:strat,
+  family = binomial,
+  data = moose
+)
R> # compare fixed effect coefficients and standard errors up to four decimals
R> data.frame(
+  est_none = coef(bin),
+  est_glm = coef(bin_glm),
+  se_none = sqrt(diag(vcov(bin))),
+  se_glm = sqrt(diag(vcov(bin_glm)))
+) |>
+  apply(2, round, digits = 4)
\end{verbatim}

\begin{verbatim}
            est_none est_glm se_none se_glm
(Intercept)  -0.5219 -0.5219  0.5082 0.5082
elev          0.0002  0.0002  0.0024 0.0024
stratM        1.0274  1.0274  0.7150 0.7150
elev:stratM  -0.0013 -0.0013  0.0038 0.0038
\end{verbatim}

The advantage of using \fct{spglm} to fit a model with
\texttt{spcov\_type\ =\ "none"} is that it provides access to other
\pkg{spmodel} functions for model objects (e.g., \fct{glances} below)
and accounts for the additional terms in the likelihood from
Equation\(~\)\eqref{eq-marginal6}. These additional terms make the
likelihood for \fct{spglm} and \fct{glm} different, though the models
convey the same information. A glance at the spatial and nonspatial
models reveals:

\begin{verbatim}
R> glances(spbin, bin)
\end{verbatim}

\begin{verbatim}
# A tibble: 2 x 11
  model     n     p  npar value   AIC  AICc   BIC logLik deviance
  <chr> <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>
1 spbin   218     4     3  681.  687.  687.  697.  -340.     161.
2 bin     218     4     0  717.  717.  717.  717.  -359.     294.
# i 1 more variable: pseudo.r.squared <dbl>
\end{verbatim}

The spatial model has a notably lower AIC, AICc, BIC, and deviance,
suggesting it is the superior model. Another model comparison approach
is leave-one-out cross validation. In leave-one-out cross validation,
separately each observation is held out, a model is fit to the remaining
data, and a prediction is made for the mean of the held out observation
on the response scale. Then, statistics like leave-one-out bias,
mean-squared-prediction error (MSPE), and the square root of MSPE
(RMSPE) may be computed:

\begin{verbatim}
R> loocv(spbin) |>
+  apply(2, round, digits = 4)
\end{verbatim}

\begin{verbatim}
   bias    MSPE   RMSPE
-0.0006  0.1458  0.3818
\end{verbatim}

\begin{verbatim}
R> loocv(bin) |>
+  apply(2, round, digits = 4)
\end{verbatim}

\begin{verbatim}
  bias   MSPE  RMSPE
0.0000 0.2403 0.4902
\end{verbatim}

Both models are nearly unbiased, but the spatial model has an
approximately 39\% lower MSPE, suggesting the probability predictions
tend to be much closer to the observed presence values.

A third model comparison tool is area under the receiver operating
characteristic (AUROC) curve. The AUROC curve ranges from zero to one
and conveys a model's classification performance over all possible
probability thresholds \citep{james2013introduction}. Larger values of
AUROC indicate a more accurate model:

\begin{verbatim}
R> AUROC(spbin)
\end{verbatim}

\begin{verbatim}
[1] 0.9490741
\end{verbatim}

\begin{verbatim}
R> AUROC(bin)
\end{verbatim}

\begin{verbatim}
[1] 0.647138
\end{verbatim}

All three performance metrics (likelihood-based statistics,
leave-one-out statistics, and AUROC) prefer the spatial model.

\subsection{Count data}\label{count-data}

\begin{figure}

\centering{

\includegraphics[width=0.7\textwidth,height=0.7\textheight]{figures/figure-6.png}

}

\caption{\label{fig-moose-data-count}Moose counts in Alaska. Circles
represent moose counts (based on color) and triangles represent
locations at which mean count predictions are desired.}

\end{figure}%

The \texttt{count} variable in \texttt{moose} contains the number of
moose observed at a site (Figure~\ref{fig-moose-data-count}). Count data
are often modeled using Poisson or negative binomial regression with the
log link function. The Poisson regression model assumes each datum's
underlying latent mean equals its variance, while the negative binomial
accommodates overdispersion (where the variance is greater than the
mean) at the cost of estimating an extra parameter.

So far our spatial models have made an implicit assumption of geometric
isotropy. A spatial covariance is geometrically isotropic if its
dependence decays with distance equally in all directions. A spatial
covariance is geometrically anisotropic if its dependence decays
differently in different directions. The geometric anisotropy's
directionality and strength are controlled by rotation and scale
parameters that are applied to the original coordinates, creating a
transformed set of coordinates whose spatial covariance is geometrically
isotropic. Geometrically anisotropic models are fit by specifying
\texttt{anisotropy}:

\begin{verbatim}
R> sppois <- spglm(
+  formula = count ~ elev + strat + elev:strat,
+  family = poisson,
+  data = moose,
+  spcov_type = "gaussian",
+  anisotropy = TRUE
+)
R>
R> spnbin <- update(sppois, family = nbinomial)
\end{verbatim}

Because the models have the same support (i.e., both non-negative count
models), we can use likelihood-based statistics to compare them:

\begin{verbatim}
R> glances(sppois, spnbin, sort_by = "AIC") |>
+  subset(select = c(model, npar, AIC, AICc, BIC))
\end{verbatim}

\begin{verbatim}
# A tibble: 2 x 5
  model   npar   AIC  AICc   BIC
  <chr>  <int> <dbl> <dbl> <dbl>
1 spnbin     6 1318. 1319. 1339.
2 sppois     5 1320. 1321. 1337.
\end{verbatim}

The negative binomial model has a slightly lower AIC and AICc, while the
Poisson model has a slightly lower BIC. This is reasonable given the BIC
penalizes additional parameters (here, an overdispersion parameter) more
heavily than AIC and AICc. The leave-out-out MSPE prefers the negative
binomial model:

\begin{verbatim}
R> loocv(sppois) |>
+  apply(2, round, digits = 4)
\end{verbatim}

\begin{verbatim}
   bias    MSPE   RMSPE
 1.2882 31.7882  5.6381
\end{verbatim}

\begin{verbatim}
R> loocv(spnbin) |>
+  apply(2, round, digits = 4)
\end{verbatim}

\begin{verbatim}
   bias    MSPE   RMSPE
 0.3485 28.3760  5.3269
\end{verbatim}

A likelihood-based comparison between the negative binomial anisotropic
model and the negative binomial isotropic model suggests that the
anisotropic model is preferred (lower AIC, AICc, and BIC):

\begin{verbatim}
R> spnbin_iso <- update(spnbin, anisotropy = FALSE)
R> glances(spnbin_iso, spnbin, sort_by = "AIC") |>
+  subset(select = c(model, npar, AIC, AICc, BIC))
\end{verbatim}

\begin{verbatim}
# A tibble: 2 x 5
  model       npar   AIC  AICc   BIC
  <chr>      <int> <dbl> <dbl> <dbl>
1 spnbin         6 1318. 1319. 1339.
2 spnbin_iso     4 1333. 1333. 1346.
\end{verbatim}

\fct{plot} returns the spatial covariance as a function of direction
(Figure~\ref{fig-tropy}):

\begin{verbatim}
R> plot(spnbin_iso, which = 8)
R> plot(spnbin, which = 8)
\end{verbatim}

\begin{figure}

\centering{

\includegraphics[width=1\textwidth,height=1\textheight]{figures/figure-7.png}

}

\caption{\label{fig-tropy}Level curves of equal autocorrelation for the
negative binomial moose count models. The ellipse is centered at zero
distance in the x-direction and y-direction, and points along the
ellipse have equal levels of autocorrelation. In the isotropic level
curve (left), spatial covariance decays equally in all directions. In
the anistropic level curve (right), spatial covariance decays fastest in
the northeast-southwest direction and slowest in the northwest-southeast
direction (this pattern can be seen in the observed counts).}

\end{figure}%

Earlier we used \fct{tidy} to tidy the model's fixed effects, but we can
also use tidy to tidy the spatial covariance parameters:

\begin{verbatim}
R> tidy(spnbin, effects = "spcov")
\end{verbatim}

\begin{verbatim}
# A tibble: 5 x 3
  term    estimate is_known
  <chr>      <dbl> <lgl>
1 de         4.65  FALSE
2 ie         0.148 FALSE
3 range  84013.    FALSE
4 rotate     2.74  FALSE
5 scale      0.259 FALSE
\end{verbatim}

The \texttt{rotate} parameter is the number of radians in \([0, \pi]\)
the ellipse is rotated and the \texttt{scale} parameter is the ratio in
\((0, 1]\) of the minor axis length to the major axis length. The
\texttt{is\_known} column indicates whether the parameter was assumed
known during optimization, controlled by specifying the
\texttt{spcov\_initial} argument.

\subsection{Skewed data}\label{skewed-data}

The \texttt{seal} data in \pkg{spmodel} is an \texttt{sf} object with
data on harbor seal trends in Alaska. The \texttt{log\_trend} variable
is the logarithm of a seal abundance temporal trend measure at the site
(based on historical data), and the \texttt{stock} variable is a factor
with two levels, \texttt{8} and \texttt{10}, where each level represents
one of twelve seal stocks (i.e., breeds) in Alaska. The \texttt{seal}
geometry is an areal polygon geometry and hence, spatial autoregressive
models based on neighborhood distance are appropriate.

SPGLMs for areal data are are fit in \pkg{spmodel} using \fct{spgautor},
which has similar syntax as \fct{spglm} but contains arguments to
control the weight matrix (\(\mathbf{W}\) in
Equation\(~\)\ref{eq-spcov-sar}) and whether or not row-standardization
\citep{ver2018spatial} is applied. By default, polygons are neighbors if
they share a boundary (i.e., Queen's contiguity; see
\citet{pebesma2023spatial}) and row standardization is assumed. Weight
matrices may be provided via the \texttt{W} argument and row
standardization may be ignored via the \texttt{row\_st} arguent.
Following \citet{ver2018spatial}, polygons without a neighbor are given
their own (independent) variance parameter called \texttt{extra}.

The trend data were originally logged to remove skew
\citep{ver2018spatial}, but we will exponentiate \texttt{log\_trend}
(Figure~\ref{fig-seal}) and model this skew directly using a SPGLM:

\begin{verbatim}
R> seal$trend <- exp(seal$log_trend)
\end{verbatim}

The \texttt{trend} variable has several missing (\texttt{NA}) values,
which represent polygons at which predictions of \texttt{trend} are
desired. To make predictions using spatial autoregressive models, the
prediction locations must be known prior to model fitting because these
locations affect the neighborhood structure of the observed data
\citep{ver2018spatial}. This restriction is notably different than
SPGLMs for point-referenced data (i.e., geostatistical models), which
completely separates the estimation and prediction steps.

\begin{figure}

\centering{

\includegraphics[width=1\textwidth,height=1\textheight]{figures/figure-8.png}

}

\caption{\label{fig-seal}Seal trend distribution in Alaska. Observed and
missing seal polygons by stock (left) and observed seal trends (right).}

\end{figure}%

\pkg{spmodel} supports the gamma and inverse Gaussian families for
modeling skewed, positive response variables. \pkg{spmodel} also
supports nonspatial random effects specified via the \texttt{random}
argument, which uses a similar formula syntax as \pkg{nlme}
\citep{pinheiro2006mixed} and \pkg{lme4} \citep{bates2015lme4}. Using
likelihood-based statistics, we compare two models fit using the
simultaneous autoregressive covariance and the Gamma and inverse
Gaussian families. Both models have a random effect for seal stock,
which builds additional correlation into the model for two polygons from
the same stock:

\begin{verbatim}
R> spgam <- spgautor(
+  formula = trend ~ 1,
+  family = Gamma,
+  data = seal,
+  spcov_type = "sar",
+  random = ~ stock
+)
R> spinvg <- update(spgam, family = inverse.gaussian)
R> glances(spgam, spinvg, sort_by = "AIC") |>
+  subset(select = c(model, npar, AIC, AICc, BIC))
\end{verbatim}

\begin{verbatim}
# A tibble: 2 x 5
  model   npar   AIC  AICc   BIC
  <chr>  <int> <dbl> <dbl> <dbl>
1 spinvg     5  108.  109.  121.
2 spgam      5  114.  115.  127.
\end{verbatim}

The inverse Gaussian model has a lower AIC, AICc, and BIC, which
indicates it is a better fit than the gamma model.

We may \fct{tidy} the estimated stock random effect variance:

\begin{verbatim}
R> tidy(spinvg, effects = "randcov")
\end{verbatim}

\begin{verbatim}
# A tibble: 1 x 3
  term      estimate is_known
  <chr>        <dbl> <lgl>
1 1 | stock  0.00411 FALSE
\end{verbatim}

The locations to predict \texttt{trend} (\texttt{NA} values) are stored
in the \texttt{newdata} element of \texttt{spinvg} and used for
prediction:

\begin{verbatim}
R> # output omitted
R> predict(spinvg, type = "response", interval = "prediction")
\end{verbatim}

If using \fct{augment} for prediction, \texttt{newdata} must be
specified:

\begin{verbatim}
R> head(augment(
+  spinvg,
+  newdata = spinvg$newdata,
+  type.predict = "response",
+  interval = "prediction"
+))
\end{verbatim}

\begin{verbatim}
Simple feature collection with 6 features and 6 fields
Geometry type: POLYGON
Dimension:     XY
Bounding box:  xmin: 1030504 ymin: 1012786 xmax: 1115097 ymax: 1057579
Projected CRS: NAD83 / Alaska Albers
# A tibble: 6 x 7
  log_trend stock trend .fitted .lower .upper                           geometry
      <dbl> <fct> <dbl>   <dbl>  <dbl>  <dbl>                      <POLYGON [m]>
1        NA 8        NA   0.942  0.893  0.993 ((1035002 1054710, 1035002 105454~
2        NA 8        NA   0.942  0.893  0.993 ((1043093 1020553, 1043097 102055~
3        NA 8        NA   0.942  0.893  0.993 ((1099737 1054310, 1099752 105426~
4        NA 8        NA   0.942  0.893  0.993 ((1099002 1036542, 1099134 103646~
5        NA 8        NA   0.942  0.893  0.993 ((1076902 1053189, 1076912 105317~
6        NA 8        NA   0.942  0.893  0.993 ((1070501 1046969, 1070317 104659~
\end{verbatim}

\subsection{Proportion data}\label{proportion-data}

We end with two examples of beta regression for proportion data
\citep{ferrari2004beta}. First, we model the nitrogen percentage in a
caribou foraging experiment. Second, we model the proportion of voter
turnout by Texas county in the United States (US) 1980 presidential
election.

\begin{figure}

\centering{

\includegraphics[width=0.7\textwidth,height=0.7\textheight]{figures/figure-9.png}

}

\caption{\label{fig-caribou}Caribou data from a spatial experimental
design measuring the percentage of nitrogen (\texttt{z}) in soil and
testing two factors: tarp type and water presence.}

\end{figure}%

The \texttt{caribou} data in \pkg{spmodel} are a data frame from a
caribou foraging experiment in Alaska meant to study the impact of water
and tarp cover on the percentage of nitrogen in surrounding plants
(Figure~\ref{fig-caribou}). \citet{lenart2002climate} studied these data
treating nitrogen percentage as a continuous variable, but here we treat
nitrogen percentage (\texttt{z}) as a proportion:

\begin{verbatim}
R> spbeta <- spglm(
+    formula = z/100 ~ water + tarp + water:tarp,
+    family = "beta",
+    data = caribou,
+    spcov_type = "matern",
+    xcoord = x,
+    ycoord = y
+)
\end{verbatim}

The nitrogen percentage is dynamically scaled in \texttt{formula} from
(0, 100) to (0, 1) so that it is a proportion. Nitrogen percentage is
modeled as a function of \texttt{water} (two levels: \texttt{"Y"} for
water and \texttt{"N"} for no water), \texttt{tarp} (three levels:
\texttt{"clear"} for a clear tarp, \texttt{"none"} for no tarp, and
\texttt{"shade"} for a shaded tarp), and their interaction, which lets
the effect of water presence vary across tarp type. \texttt{caribou} is
a data frame (not an \texttt{sf} object), so we supply the x-coordinate
and y-coordinate directly via \texttt{xcoord} and \texttt{ycoord}, and,
consistent with the tidyverse approach \citep{wickham2019welcome},
column names in \texttt{data} do not need to be quoted when referenced
(but can be quoted).

A summary of \texttt{spbeta} returns a coefficients table that provides
parameter estimates relative to a reference group. When factors have
more than two levels, it is not straightforward to use these contrasts
to determine overall significance of the factor. The \texttt{anova()}
function tests marginal (i.e., Type III sums of squares) significance of
factors using the general linear hypothesis test for spatial (i.e.,
correlated) data \citep{schabenberger2017statistical}:

\begin{verbatim}
tidy(anova(spbeta))
\end{verbatim}

\begin{verbatim}
# A tibble: 4 x 4
  effects        df statistic       p.value
  <chr>       <int>     <dbl>         <dbl>
1 (Intercept)     1  35783.   0
2 water           1      1.52 0.218
3 tarp            2     38.4  0.00000000468
4 water:tarp      2      5.74 0.0566
\end{verbatim}

These results suggest there is some evidence that the effect of water on
nitrogen percentage differ depending on the type of tarp used (0.01
\textless{} \(p~\)value \textless{} 0.1).

Sometimes averages or contrasts between factor levels that are not in
the reference group are of interest. We again leverage \pkg{spmodel}'s
built-in support for \pkg{emmeans} and use it to obtain the averages of
each factor combination on the link (here, logit) scale:

\begin{verbatim}
spemm <- emmeans(spbeta, ~ water + tarp)
spemm
\end{verbatim}

\begin{verbatim}
 water tarp  emmean     SE  df asymp.LCL asymp.UCL
 N     clear  -3.94 0.0208 Inf     -3.98     -3.90
 Y     clear  -3.90 0.0206 Inf     -3.94     -3.86
 N     none   -3.88 0.0204 Inf     -3.92     -3.84
 Y     none   -3.91 0.0206 Inf     -3.95     -3.87
 N     shade  -3.77 0.0196 Inf     -3.81     -3.73
 Y     shade  -3.83 0.0200 Inf     -3.87     -3.79

Degrees-of-freedom method: asymptotic
Results are given on the logit (not the response) scale.
Confidence level used: 0.95
\end{verbatim}

Delta method \citep{ver2012invented} standard errors are used when
averages on the response (here, proportion) scale are desired:

\begin{verbatim}
update(spemm, type = "response")
\end{verbatim}

\begin{verbatim}
 water tarp  response       SE  df asymp.LCL asymp.UCL
 N     clear   0.0191 0.000390 Inf    0.0183    0.0198
 Y     clear   0.0198 0.000398 Inf    0.0190    0.0205
 N     none    0.0202 0.000404 Inf    0.0194    0.0210
 Y     none    0.0197 0.000398 Inf    0.0189    0.0205
 N     shade   0.0226 0.000434 Inf    0.0218    0.0235
 Y     shade   0.0213 0.000418 Inf    0.0205    0.0221

Degrees-of-freedom method: asymptotic
Confidence level used: 0.95
Intervals are back-transformed from the logit scale
\end{verbatim}

Pairwise contrasts use a Tukey \(p~\)value adjustment
\citep{tukey1949comparing} by default. Here, we request no \(p~\)value
adjustment:

\begin{verbatim}
head(pairs(spemm, adjust = "none"))
\end{verbatim}

\begin{verbatim}
 contrast          estimate     SE  df z.ratio p.value
 N clear - Y clear  -0.0361 0.0293 Inf  -1.233  0.2177
 N clear - N none   -0.0601 0.0292 Inf  -2.063  0.0391
 N clear - Y none   -0.0327 0.0293 Inf  -1.118  0.2637
 N clear - N shade  -0.1735 0.0286 Inf  -6.061  <.0001
 N clear - Y shade  -0.1136 0.0289 Inf  -3.932  0.0001
 Y clear - N none   -0.0241 0.0290 Inf  -0.830  0.4064

Degrees-of-freedom method: asymptotic
Results are given on the log odds ratio (not the response) scale.
\end{verbatim}

\begin{figure}

\centering{

\includegraphics[width=0.7\textwidth,height=0.7\textheight]{figures/figure-10.png}

}

\caption{\label{fig-texas}Proportion of voter turnout in Texas for the
1980 presidential election. Circles represent voter turnout (based on
color) and triangles represent locations at which voter turnout
predictions are desired.}

\end{figure}%

We now model the \texttt{elect80} data in \pkg{spData}
\citep{bivand2024spdata}, which contains voter turnout data by county in
the 1980 US Presidential election \citep{pace1997quick}. The
\texttt{texas} data in \pkg{spmodel} contains a subset of these data in
the state of Texas. These data are point-referenced, but we may still
use autoregressive models if neighborhood distance is determined using
county centroids (i.e., counties whose centroid distance is less than
some cutoff are defined as neighbors). The response variable of
interest, \texttt{turnout}, is the proportion of registered voters in
the county who voted in the election (Figure~\ref{fig-texas}).

\begin{verbatim}
R> spgautor_mods <- spgautor(
+  formula = turnout ~ log_income,
+  family = beta,
+  data = texas,
+  spcov_type = c("car", "sar"),
+  cutoff = 2e5,
+  estmethod = "ml"
+)
\end{verbatim}

We model voter turnout as a function of log income using both the
conditional and simultaneous autoregressive models with a neighbor
distance cutoff of 200 kilometers and the maximum likelihood estimation
method. When a vector is provided to \texttt{spcov\_type} in
\fct{spgautor} (or \fct{spglm}), a model is fit for each spatial
covariance type and stored in a list with name equal to the respective
type. Then it is simple to glance at each model fit:

\begin{verbatim}
R> glances(spgautor_mods) |>
+  subset(select = c(model, npar, AIC, AICc, BIC))
\end{verbatim}

\begin{verbatim}
# A tibble: 2 x 5
  model  npar   AIC  AICc   BIC
  <chr> <int> <dbl> <dbl> <dbl>
1 car       3 -38.3 -38.1 -21.2
2 sar       3 -35.7 -35.5 -18.6
\end{verbatim}

The conditional autoregressive model has the best fit (in terms of AIC,
AICc, and BIC). In this model, there is significant evidence
\texttt{log\_income} is positively related to average voter turnout
(\(p~\)value \textless{} 0.001):

\begin{verbatim}
tidy(spgautor_mods$car)
\end{verbatim}

\begin{verbatim}
# A tibble: 2 x 5
  term        estimate std.error statistic    p.value
  <chr>          <dbl>     <dbl>     <dbl>      <dbl>
1 (Intercept)   -4.74      1.06      -4.47 0.00000783
2 log_income     0.532     0.117      4.55 0.00000531
\end{verbatim}

Another way to assess the impact of \texttt{log\_income} on
\texttt{turnout} is a likelihood ratio test:

\begin{verbatim}
reduced_car <- update(spgautor_mods$car, formula = turnout ~ 1)
tidy(anova(reduced_car, spgautor_mods$car))
\end{verbatim}

\begin{verbatim}
# A tibble: 1 x 5
  full              reduced        df statistic     p.value
  <chr>             <chr>       <dbl>     <dbl>       <dbl>
1 spgautor_mods$car reduced_car     1      24.1 0.000000930
\end{verbatim}

Likelihood ratio tests compare the fit of a ``full'' model compared to a
``reduced'' model that is completely nested within the full model. If
there is evidence the full model explains significantly more information
than the reduced model, the likelihood ratio test \(p~\)value will be
small. Similar to the summary output from the general linear hypothesis
test, the likelihood ratio suggests \texttt{log\_income} is related to
average voter turnout (\(p~\)value \textless{} 0.001).

The default estimation method in \pkg{spmodel} is REML, but note that
these models used maximum likelihood (ML). ML is very similar to REML --
the difference is that for ML, the fixed effects \(\boldsymbol{\beta}\)
are not integrated out of Equation\(~\)\eqref{eq-marginal} but are
rather back-substituted. While REML typically performs better for fixed
effect estimation and prediction \citep{zimmerman2024spatial}, ML allows
likelihood-based comparisons (e.g., AIC) for models with simultaneously
varying fixed effect and covariance structures, while REML
likelihood-based comparisons are only valid for models sharing the same
fixed effect structure (though \citet{gurka2006selecting} provides some
evidence that this restriction may be unnecessary).

The point-referenced \texttt{texas} and \texttt{caribou} data may be
analyzed using \fct{spgautor} or \fct{spglm} and comparisons across
these structures can be made using likelihood-based statistics (as long
as the supports of the response distribution are the same). Put another
way, likelihood-based statistics can be used to determine whether
geostatistical (distance-based) or autoregressive (neighbor-based)
structures perform best when the data are point-referenced.

\section{Discussion}\label{sec-discussion}

SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace
approximation that marginalizes over the latent (i.e., unobserved) mean,
\(\mathbf{w}\), and the fixed effects, \(\boldsymbol{\beta}\). The
approach is quite flexible and accommodates any general response
distribution and covariance structure. \citet{ver2024marginal} show that
the approach, as implemented in \pkg{spmodel}, generally yields unbiased
estimators with proper interval coverage and often outperforms the
Bayesian approach from \pkg{spBayes}, the INLA approach from
\pkg{R-INLA}, and the automatic differentiation approach from
\fct{glmmTMB}.

\pkg{spmodel}'s \fct{spglm} and \fct{spgautor} functions are similar in
structure and syntax as the base-\proglang{R} \fct{glm} function, easing
the transition from GLMs to SPGLMs. These functions support six response
distributions (Table\(~\)\ref{table-links}) and 20 spatial covariance
functions. \pkg{spmodel} provides several additional features that
accommodate geometric anisotropy, nonspatial random effects, fixing
spatial covariance parameters at known values, data having thousands of
observations (following \citet{ver2023indexing}), incorporating spatial
dependence in machine learning (e.g., random forests;
\citet{breiman2001random}), simulating spatially dependent data (e.g.,
\fct{sprbinom}, \fct{sprpois}), and several others. Learn more at
\url{https://CRAN.R-project.org/package=spmodel} and links therein.

\section*{Computational details}\label{computational-details}
\addcontentsline{toc}{section}{Computational details}

The results in this paper were obtained using \proglang{R} 4.4.0 with
the \pkg{spmodel} 0.9.0 package. Figures were created using the
\pkg{ggplot2} 3.5.1 package \citep{wickham2016ggplot2} and base
\proglang{R}.

\section*{Data and code availability}\label{data-and-code-availability}
\addcontentsline{toc}{section}{Data and code availability}

All writing and code associated with this manuscript is available for
viewing and download on GitHub at
\url{https://github.com/USEPA/spmodel.glm.manuscript}. All data used are
part of the \pkg{spmodel} \proglang{R} package available for download
from CRAN at \url{https://CRAN.R-project.org/package=spmodel}.

\section*{Acknowledgments}\label{acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}

We would like to thank initial reviewers and editors for feedback that
has greatly improved the manuscript.

The views expressed in this manuscript are those of the authors and do
not necessarily represent the views or policies of the U.S.
Environmental Protection Agency or the National Oceanic and Atmospheric
Administration. Any mention of trade names, products, or services does
not imply an endorsement by the U.S. government, the U.S. Environmental
Protection Agency, or the National Oceanic and Atmospheric
Administration. The U.S. Environmental Protection Agency and the
National Oceanic and Atmospheric Administration do not endorse any
commercial products, services or enterprises.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\renewcommand{\bibsection}{}
\bibliography{references.bib}

\end{document}

### 2024-11-20 inst/manuscript/manuscript.qmd
some udpates

@@
 The approach is quite flexible and accommodates any general response distribution and covariance structure.
-@ver2024marginal show that the approach, as implemented in [spmodel]{.pkg}, generally yields unbiased estimators with proper interval coverage and often outperforms the Bayesian approach from [spBayes]{.pkg}, the INLA approach from [R-INLA]{.pkg}, and the automatic differentiation approach from [glmmTMB]{.fct}.
+@ver2024marginal show that the approach, as implemented in [spmodel]{.pkg}, generally yields unbiased estimators with proper interval coverage and often outperforms the Bayesian approach from [spBayes]{.pkg}, the INLA approach from [R-INLA]{.pkg}, and the automatic differentiation approach from [glmmTMB]{.pkg}.
 

### 2024-11-20 inst/manuscript/manuscript.tex
some udpates

@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
 \makeatother
@@
 \pkg{R-INLA}, and the automatic differentiation approach from
-\fct{glmmTMB}.
+\pkg{glmmTMB}.
 
@@
 
-\renewcommand{\bibsection}{}
+%bib-loc-124C8010
+
   \bibliography{references.bib}

### 2024-11-27 inst/manuscript/manuscript.tex
a few editorial updates

--- 2024-11-19 inst/manuscript/manuscript.tex
@@
 \makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi
 \makeatother
@@
 
-\section{Spatial generalized linear models using the Laplace
-approximation}\label{sec-spglm}
+\section{Marginal SPGLMs and the Laplace approximation}\label{sec-spglm}
 

### 2024-11-27 inst/manuscript/figures/figures.R

@@ -3,7 +3,7 @@ library(tidyverse)
 library(spmodel)
 library(patchwork)
 library(here)
-fig_path <- here("inst", "jss", "manuscript", "figures")
+fig_path <- here("inst", "manuscript", "figures")
 
 #######################
 ######## Figure 1

### 2025-08-21 inst/manuscript/manuscript.Rmd
update to rmd from qmd

--- 2024-11-20 inst/manuscript/manuscript.qmd
@@
 ---
-title: "Spatial Generalized Linear Models in [R]{.proglang} Using [spmodel]{.pkg}"
-format:
-    jss-pdf:
-        keep-tex: true
-    jss-html: default
+documentclass: jss
 author:
+    # use this syntax to add text on several lines
+    # To add another line, use \AND at the end of the previous one as above
+    # use a different affiliation in adress field (differently formated here)
   - name: Michael Dumelle
-    affiliations:
-      - name: |
+    orcid: 0000-0002-3393-5529
+    affiliation: |
       | United States
       | Environmental Protection Agency
-        department: Pacific Ecological Systems Division
-        address: 200 SW 35th St
-        city: Corvallis
-        country: OR
-        postal-code: 97333
-    orcid: 0000-0002-3393-5529
-    email: dumelle.michael@epa.gov
+    address: |
+      | 200 SW 35th St
+      | Corvallis, OR, 97330
+    email: \email{Dumelle.Michael@epa.gov}
   - name: Jay M. Ver Hoef
-    affiliations:
-      - name: |
-          | National Oceanic and
-          | Atmospheric Administration
-        department: Alaska Fisheries Science Center
     orcid: 0000-0003-4302-6895
+    affiliation: |
+      | Alaska Fisheries
+      | Science Center
   - name: Matt Higham
-    affiliations:
-      - name: St. Lawrence University
-        department: Department of Math, Computer Science, and Statistics
     orcid: 0009-0006-4217-625X
+    affiliation: |
+      | St. Lawrence University
+title:
+  # If you use tex in the formatted title, also supply version without
+  # For running headers, if needed
+  formatted: "Spatial Generalized Linear Models in [R]{.proglang} Using [spmodel]{.pkg}"
+  plain:     "Spatial Generalized Linear Models in R Using spmodel"
+  short:     "SPGLMs in [R]{.proglang} Using [spmodel]{.pkg}"
 abstract: |
@@
 Here we use [spmodel]{.pkg} to illustrate the modeling of binary, count, skewed and proportion response variables from several point-referenced and areal data sets.
-keywords: [Autoregressive Model, Geostatistics, Spatial Covariance, Spatially-Explicit Model, Statistical Model]
-keywords-formatted: [Autoregressive Model, Geostatistics, Spatial Covariance, Spatially-Explicit Model, Statistical Model]
-bibliography: references.bib  
+keywords:
+  # at least one keyword must be supplied
+  formatted: [autoregressive model, geostatistical model, spatial autocovariance, spatial autocorrelation]
+  plain:     [autoregressive model, geostatistical model, spatial autocovariance, spatial autocorrelation]
 preamble: |
@@
   \usepackage{caption, subcaption}
+output: rticles::jss_article
 editor_options: 
@@
 
-```{R}
-#| include: false
+```{r, setup, include=FALSE}
+options(prompt = 'R> ', continue = '+ ')
 library(spmodel)
@@
 
-## Introduction {#sec-intro}
+# Introduction {#sec-intro}
 
@@
 The rest of this article is organized as follows.
-In @sec-spglm, we provide some background for the SPGLM fitting and prediction routines in [spmodel]{.pkg}.
-In @sec-applications, we provide several applications of [spmodel]{.pkg} to spatial binary, count, skewed and proportion data with both point-referenced and areal supports.
-And in @sec-discussion, we end with a discussion synthesizing \pkg{spmodel}'s contributions to the analysis of SPGLMs in [R]{.proglang}.
+In Section$~$\ref{sec-spglm}, we provide some background for the SPGLM fitting and prediction routines in [spmodel]{.pkg}.
+In Section$~$\ref{sec-applications}, we provide several applications of [spmodel]{.pkg} to spatial binary, count, skewed and proportion data with both point-referenced and areal supports.
+And in Section$~$\ref{sec-discussion}, we end with a discussion synthesizing \pkg{spmodel}'s contributions to the analysis of SPGLMs in [R]{.proglang}.
 
-## Spatial generalized linear models using the Laplace approximation {#sec-spglm}
+# Spatial generalized linear models using the Laplace approximation {#sec-spglm}
 
@@
 
-## Application {#sec-applications}
+# Application {#sec-applications}
 
-The [spglm]{.fct} (for point-referenced data) and [spgautor]{.fct} (for areal data) functions in [spmodel]{.pkg} fit SPGLMs using the Laplace approximation outlined in @sec-spglm.
+The [spglm]{.fct} (for point-referenced data) and [spgautor]{.fct} (for areal data) functions in [spmodel]{.pkg} fit SPGLMs using the Laplace approximation outlined in Section$~$\ref{sec-spglm}.
 Both [spglm]{.fct} and [spgautor]{.fct} generally require the following four arguments: `formula`, a formula that describes the relationship between the response variable and explanatory variables; `family`, the response distribution (which can be `binomial`, `poisson`, `nbinomial`, `Gamma`, `inverse.gaussian`, or `beta`); `data`, the data frame that holds the variables in `formula` as well as spatial locations; and `spcov_type`, the spatial covariance type.
@@
 Before proceeding, load [spmodel]{.pkg} into the current [R]{.proglang} session:
-```{R}
-#| prompt: true
-#| eval: false
+```{r, eval = FALSE}
 library("spmodel")
@@
 
-### Binary data
+## Binary data
 
@@
 
-```{R}
-#| prompt: true
+```{r}
 head(moose)
@@
 They are also an `sf` object with the same projection and measurements for `elev` and `strat`.
-@fig-moose-data shows the `presence` variable in `moose` as well as the spatial locations of both `moose` and `moose_preds`.
+Figure$~$\ref{fig-moose-data} shows the `presence` variable in `moose` as well as the spatial locations of both `moose` and `moose_preds`.
 Moose are most common in the southwestern and eastern parts of the domain and least common in the northwest.
-```{R}
-#| prompt: false
-#| echo: false
-#| fig-align: center
-#| fig-cap: "Moose presence in Alaska. Circles represent moose presence or absence (based on color) and triangles represent locations at which moose presence probability predictions are desired."
-#| out-width: "70%"
-#| out-height: "70%"
-#| label: fig-moose-data
-knitr::include_graphics(paste0(fig_path, "/figure-1.png"))
-```
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-1.png}
+\caption{Moose presence in Alaska. Circles represent moose presence or absence (based on color) and triangles represent locations at which moose presence probability predictions are desired.}
+\label{fig-moose-data}
+\end{figure}
 
@@
 
-```{R}
-#| prompt: true
+```{r}
 spbin <- spglm(
@@
 Summarizing the model object yields a summary similar to that provided by the familiar [glm]{.fct}:
-```{R}
-#| prompt: true
+```{r}
 summary(spbin)
@@
 The first [broom]{.pkg} function is [tidy]{.fct}, which tidies the model output:
-```{R}
-#| prompt: true
+```{r}
 tidy(spbin)
@@
 We could refit the model treating `"M"` as the reference group instead of `"L"`:
-```{R}
-#| prompt: true
+```{r}
 moose$strat2 <- factor(moose$strat, levels = c("M", "L"))
@@
 
-```{r}
-#| prompt: true
-#| eval: false
+```{r, eval = FALSE}
 library("emmeans")
@@
 
-```{R}
-#| prompt: true
+```{r}
 emtrends(spbin, "strat", "elev")
@@
 The second [broom]{.pkg} function is [glance]{.fct}, which glances at the model fit:
-```{R}
-#| prompt: true
+```{r}
 glance(spbin)
@@
 
-```{R}
-#| prompt: false
-#| echo: false
-#| label: fig-sp-diagnostic
-#| fig-cap: "Spatial logistic regression model diagnostics from [augment]{.fct}. The leverage (i.e., hat) values (left) and standardized residuals (right)."
-#| out-width: "100%"
-#| out-height: "100%"
-knitr::include_graphics(paste0(fig_path, "/figure-2.png"))
-```
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-2.png}
+\caption{Spatial logistic regression model diagnostics from [augment]{.fct}. The leverage (i.e., hat) values (left) and standardized residuals (right).}
+\label{fig-sp-diagnostic}
+\end{figure}
 
@@
 The [augment]{.fct} function returns the fitted $\mathbf{w}$ values (`.fitted`), deviance residuals (`.resid`), leverage (i.e., hat) values (`.hat`), Cook's distance (`.cooksd`), and standardized residuals (`.std.resid`).
-When the data are an `sf` object, [augment]{.fct} returns another `sf` object, helpful for visualizing model diagnostics spatially as in @fig-sp-diagnostic.
+When the data are an `sf` object, [augment]{.fct} returns another `sf` object, helpful for visualizing model diagnostics spatially as in Figure$~$\ref{fig-sp-diagnostic}.
 Leverage measures the unusualness of an observation's set of explanatory variables, while Cook's distances measures how influential an observation is on the resulting model fit [@montgomery2021introduction].
@@
 
-Similar to [glm]{.fct} model objects, [plot]{.fct} can be used to visualize diagnostics (@fig-sp-diagnostic2):
-```{R}
-#| prompt: true
-#| eval: false
+Similar to [glm]{.fct} model objects, [plot]{.fct} can be used to visualize diagnostics (Figure$~$\ref{fig-sp-diagnostic2}):
+```{r, eval = FALSE}
 plot(spbin, which = c(4, 7))
@@
 
-```{R}
-#| prompt: true
-#| echo: false
-#| label: fig-sp-diagnostic2
-#| fig-cap: "Spatial logistic regression model diagnostics from [plot]{.fct}. The Cook's distance values (left) and the fitted spatial covariance as a function of distance (right)."
-#| out-width: "100%"
-#| out-height: "100%"
-knitr::include_graphics(paste0(fig_path, "/figure-3.png"))
-```
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-3.png}
+\caption{Spatial logistic regression model diagnostics from [plot]{.fct}. The Cook's distance values (left) and the fitted spatial covariance as a function of distance (right).}
+\label{fig-sp-diagnostic2}
+\end{figure}
 
 Components of model variation are partitioned using [varcomp]{.fct}:
-```{R}
-#| prompt: true
+```{r}
 varcomp(spbin)
@@
 
-```{R}
-#| prompt: false
-#| echo: false
-#| fig-align: center
-#| fig-cap: "Moose presence probability fitted values and predictions. Fitted values are represeneted by circles and predictions by triangles."
-#| out-width: "70%"
-#| out-height: "70%"
-#| label: fig-moose-fit
-knitr::include_graphics(paste0(fig_path, "/figure-4.png"))
-```
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-4.png}
+\caption{Moose presence probability fitted values and predictions. Fitted values are represeneted by circles and predictions by triangles.}
+\label{fig-moose-fit}
+\end{figure}
 
 We make predictions of the log odds of moose probability presence at each site in `moose_preds` using [predict]{.fct}:
-```{R}
-#| prompt: true
+```{r}
 head(predict(spbin, newdata = moose_preds))
@@
 [augment]{.fct} may also be used to augment the prediction data with predictions:
-```{R}
+```{r}
 head(augment(
@@
 
-Here, we requested predictions on the probability (i.e., response) scale (@fig-moose-fit) alongside lower and upper bounds of a 95% (see `level`) prediction interval (@fig-moose-int).
+Here, we requested predictions on the probability (i.e., response) scale (Figure$~$\ref{fig-moose-fit}) alongside lower and upper bounds of a 95% (see `level`) prediction interval (Figure$~$\ref{fig-moose-int}).
 
-```{R}
-#| prompt: false
-#| echo: false
-#| label: fig-moose-int
-#| fig-cap: "Moose presence probability prediction intervals. 95% prediction interval lower bound (left) and 95% prediction interval upper bound (right)."
-#| out-width: "100%"
-#| out-height: "100%"
-knitr::include_graphics(paste0(fig_path, "/figure-5.png"))
-```
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-5.png}
+\caption{Moose presence probability prediction intervals. 95\% prediction interval lower bound (left) and 95\% prediction interval upper bound (right).}
+\label{fig-moose-int}
+\end{figure}
 
@@
 
-```{R}
-#| prompt: true
-# fit nonspatial model using spglm()
+```{r}
 bin <- spglm(
@@
 )
-# fit nonspatial model using glm()
 bin_glm <- glm(
@@
 )
-# compare fixed effect coefficients and standard errors up to four decimals
 data.frame(
@@
 A glance at the spatial and nonspatial models reveals:
-```{R}
-#| prompt: true
+```{r}
 glances(spbin, bin)
@@
 Then, statistics like leave-one-out bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) may be computed:
-```{R}
-#| prompt: true
+```{r}
 loocv(spbin) |>
@@
 Larger values of AUROC indicate a more accurate model:
-```{R}
-#| prompt: true
+```{r}
 AUROC(spbin)
@@
 
-### Count data
+## Count data
 
-```{R}
-#| prompt: false
-#| echo: false
-#| fig-align: center
-#| fig-cap: "Moose counts in Alaska. Circles represent moose counts (based on color) and triangles represent locations at which mean count predictions are desired."
-#| label: fig-moose-data-count
-#| out-width: "70%"
-#| out-height: "70%"
-knitr::include_graphics(paste0(fig_path, "/figure-6.png"))
-```
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-6.png}
+\caption{Moose counts in Alaska. Circles represent moose counts (based on color) and triangles represent locations at which mean count predictions are desired.}
+\label{fig-moose-data-count}
+\end{figure}
 
-The `count` variable in `moose` contains the number of moose observed at a site (@fig-moose-data-count).
+The `count` variable in `moose` contains the number of moose observed at a site (Figure$~$\ref{fig-moose-data-count}).
 Count data are often modeled using Poisson or negative binomial regression with the log link function.
@@
 ```{r}
-#| prompt: true
 sppois <- spglm(
@@
 ```{r}
-#| prompt: true
 glances(sppois, spnbin, sort_by = "AIC") |>
@@
 The leave-out-out MSPE prefers the negative binomial model:
-```{R}
-#| prompt: true
+```{r}
 loocv(sppois) |>
@@
 A likelihood-based comparison between the negative binomial anisotropic model and the negative binomial isotropic model suggests that the anisotropic model is preferred (lower AIC, AICc, and BIC):
-```{R}
-#| prompt: true
+```{r}
 spnbin_iso <- update(spnbin, anisotropy = FALSE)
@@
 
-[plot]{.fct} returns the spatial covariance as a function of direction (@fig-tropy):
-```{R}
-#| prompt: true
-#| eval: false
+[plot]{.fct} returns the spatial covariance as a function of direction (Figure$~$\ref{fig-tropy}):
+```{r}
 plot(spnbin_iso, which = 8)
@@
 
-```{R}
-#| prompt: true
-#| echo: false
-#| label: fig-tropy
-#| fig-cap: "Level curves of equal autocorrelation for the negative binomial moose count models.
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-7.png}
+\caption{Level curves of equal autocorrelation for the negative binomial moose count models.
 The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of autocorrelation.
 In the isotropic level curve (left), spatial covariance decays equally in all directions.
-In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts)."
-#| out-width: "100%"
-#| out-height: "100%"
-#| fig-asp: 1
-knitr::include_graphics(paste0(fig_path, "/figure-7.png"))
-```
+In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts).}
+\label{fig-tropy}
+\end{figure}
 
 Earlier we used [tidy]{.fct} to tidy the model's fixed effects, but we can also use tidy to tidy the spatial covariance parameters:
-```{R}
-#| prompt: true
+```{r}
 tidy(spnbin, effects = "spcov")
@@
 
-### Skewed data
+## Skewed data
 
@@
 
-The trend data were originally logged to remove skew [@ver2018spatial], but we will exponentiate `log_trend` (@fig-seal) and model this skew directly using a SPGLM:
-```{R}
-#| prompt: true
+The trend data were originally logged to remove skew [@ver2018spatial], but we will exponentiate `log_trend` (Figure$~$\ref{fig-seal}) and model this skew directly using a SPGLM:
+```{r}
 seal$trend <- exp(seal$log_trend)
@@
 
-```{r}
-#| prompt: true
-#| echo: false
-#| label: fig-seal
-#| fig-cap: "Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed seal trends (right)."
-#| out-width: "100%"
-#| out-height: "100%"
-knitr::include_graphics(paste0(fig_path, "/figure-8.png"))
-```
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-8.png}
+\caption{Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed seal trends (right).}
+\label{fig-seal}
+\end{figure}
 
@@
 ```{r}
-#| prompt: true
 spgam <- spgautor(
@@
 We may [tidy]{.fct} the estimated stock random effect variance:
-```{R}
-#| prompt: true
+```{r}
 tidy(spinvg, effects = "randcov")
@@
 The locations to predict `trend` (`NA` values) are stored in the `newdata` element of `spinvg` and used for prediction:
-```{R}
-#| prompt: true
-#| results: "hide"
-# output omitted
+```{r, results="hide"}
 predict(spinvg, type = "response", interval = "prediction")
@@
 ```{r}
-#| prompt: true
 head(augment(
@@
 
-### Proportion data
+## Proportion data
 
@@
 
-```{R}
-#| prompt: false
-#| echo: false
-#| label: fig-caribou
-#| fig-align: center
-#| fig-cap: "Caribou data from a spatial experimental design measuring the percentage of nitrogen (`z`) in soil and testing two factors: tarp type and water presence."
-#| out-width: "70%"
-#| out-height: "70%"
-knitr::include_graphics(paste0(fig_path, "/figure-9.png"))
-```
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-9.png}
+\caption{Caribou data from a spatial experimental design measuring the percentage of nitrogen (z) in soil and testing two factors: tarp type and water presence.}
+\label{fig-caribou}
+\end{figure}
 
-The `caribou` data in [spmodel]{.pkg} are a data frame from a caribou foraging experiment in Alaska meant to study the impact of water and tarp cover on the percentage of nitrogen in surrounding plants (@fig-caribou).
+The `caribou` data in [spmodel]{.pkg} are a data frame from a caribou foraging experiment in Alaska meant to study the impact of water and tarp cover on the percentage of nitrogen in surrounding plants (Figure$~$\ref{fig-caribou}).
 @lenart2002climate studied these data treating nitrogen percentage as a continuous variable, but here we treat nitrogen percentage (`z`) as a proportion:
 
-```{R}
-#| prompt: true
+```{r}
 spbeta <- spglm(
@@
 
-```{r}
-#| prompt: true
-#| echo: false
-#| label: fig-texas
-#| fig-cap: "Proportion of voter turnout in Texas for the 1980 presidential election.
-Circles represent voter turnout (based on color) and triangles represent locations at which voter turnout predictions are desired."
-#| out-width: "70%"
-#| out-height: "70%"
-knitr::include_graphics(paste0(fig_path, "/figure-10.png"))
-```
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-10.png}
+\caption{Proportion of voter turnout in Texas for the 1980 presidential election.
+Circles represent voter turnout (based on color) and triangles represent locations at which voter turnout predictions are desired.}
+\label{fig-texas}
+\end{figure}
 
@@
 These data are point-referenced, but we may still use autoregressive models if neighborhood distance is determined using county centroids (i.e., counties whose centroid distance is less than some cutoff are defined as neighbors).
-The response variable of interest, `turnout`, is the proportion of registered voters in the county who voted in the election (@fig-texas).
+The response variable of interest, `turnout`, is the proportion of registered voters in the county who voted in the election (Figure$~$\ref{fig-texas}).
 
-```{R}
-#| prompt: true
+```{r}
 spgautor_mods <- spgautor(
@@
 ```{r}
-#| prompt: true
 glances(spgautor_mods) |>
@@
 In this model, there is significant evidence `log_income` is positively related to average voter turnout ($p~$value < 0.001):
-```{R}
+```{r}
 tidy(spgautor_mods$car)
@@
 Another way to assess the impact of `log_income` on `turnout` is a likelihood ratio test:
-```{R}
+```{r}
 reduced_car <- update(spgautor_mods$car, formula = turnout ~ 1)
@@
 
-## Discussion {#sec-discussion}
+# Discussion {#sec-discussion}
 
@@
 
-## Computational details {.unnumbered}
+# Computational details {.unnumbered}
 
@@
 
-## Data and code availability {.unnumbered}
+# Data and code availability {.unnumbered}
 
@@
 
-## Acknowledgments {.unnumbered}
+# Acknowledgments {.unnumbered}
 
@@
 
-## References {.unnumbered}
-
-:::{#refs}
-
-:::
+\bibliography{references.bib}

### 2025-08-21 inst/manuscript_old/code/code.R

@@ -1,0 +1,318 @@
+#=========================================================#
+# Load Software ####
+#=========================================================#
+
+library(spmodel)
+library(ggplot2)
+library(dplyr)
+library(emmeans)
+
+#=========================================================#
+# Section 3.1 Binary Data ####
+#=========================================================#
+
+head(moose)
+
+spbin <- spglm(
+  formula = presence ~ elev + strat + elev:strat,
+  family = binomial,
+  data = moose,
+  spcov_type = "spherical"
+)
+
+summary(spbin)
+
+tidy(spbin)
+
+moose$strat2 <- factor(moose$strat, levels = c("M", "L"))
+update(spbin, formula = presence ~ elev + strat2 + elev:strat2) |>
+  summary()
+
+emtrends(spbin, "strat", "elev")
+
+glance(spbin)
+
+head(augment(spbin))
+
+varcomp(spbin)
+
+head(predict(spbin, newdata = moose_preds))
+
+head(augment(
+  spbin,
+  newdata = moose_preds,
+  type.predict = "response",
+  interval = "prediction"
+))
+
+# fit nonspatial model using spglm()
+bin <- spglm(
+  formula = presence ~ elev + strat + elev:strat,
+  family = binomial,
+  data = moose,
+  spcov_type = "none"
+)
+# fit nonspatial model using glm()
+bin_glm <- glm(
+  formula = presence ~ elev + strat + elev:strat,
+  family = binomial,
+  data = moose
+)
+# compare fixed effect coefficients and standard errors up to four decimals
+data.frame(
+  est_none = coef(bin),
+  est_glm = coef(bin_glm),
+  se_none = sqrt(diag(vcov(bin))),
+  se_glm = sqrt(diag(vcov(bin_glm)))
+) |>
+  apply(2, round, digits = 4)
+
+glances(spbin, bin)
+
+loocv(spbin) |>
+  apply(2, round, digits = 4)
+loocv(bin) |>
+  apply(2, round, digits = 4)
+
+AUROC(spbin)
+AUROC(bin)
+
+## Figures ####
+
+### Figure 1 ####
+okabe <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+moose_comb <- bind_rows(moose |> mutate(samp = "yes"), moose_preds |> mutate(samp = "no"))
[…]

### 2025-08-21 inst/manuscript_old/figures/figures.R

@@ -1,0 +1,253 @@
+# Preliminaries
+library(tidyverse)
+library(spmodel)
+library(patchwork)
+library(here)
+fig_path <- here("inst", "jss", "manuscript", "figures")
+
+#######################
+######## Figure 1
+#######################
+
+okabe <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+moose_comb <- bind_rows(moose |> mutate(samp = "yes"), moose_preds |> mutate(samp = "no"))
+obs_pred_presence <- ggplot(moose_comb, aes(color = presence, shape = samp)) +
+  geom_sf(size = 2) +
+  scale_color_manual(name = "moose", values = okabe[1:2], breaks = c(1, 0), labels = c("yes", "no")) +
+  scale_shape_manual(name = "sampled", values = c(19, 17), breaks = c("yes", "no"), labels = c("yes", "no")) +
+  theme_bw(base_size = 16) +
+  scale_y_continuous(breaks = seq(62.6, 63.6, length.out = 5)) +
+  scale_x_continuous(breaks = seq(-148.5, -146, length.out = 3)) +
+  guides(
+    color = guide_legend(order = 1),
+    shape = guide_legend(order = 2)
+  ) +
+  geom_sf(size = 2, data = moose_comb %>% filter(samp == "no"), color = "red", shape = 17)
+obs_pred_presence
+ggsave(
+  filename = paste0(fig_path, "/figure-1.png"),
+  plot = obs_pred_presence,
+  dpi = 300,
+  height = 5,
+  width = 7.68
+)
+
+#######################
+######## Figure 2
+#######################
+
+
+spbin <- spglm(
+  formula = presence ~ elev + strat + elev:strat,
+  family = binomial,
+  data = moose,
+  spcov_type = "spherical"
+)
+spbin_aug <- augment(spbin)
+
+p1 <- ggplot(spbin_aug, aes(color = .hat)) +
+  geom_sf(size = 1) +
+  scale_color_viridis_c(option = "E", name = ".hat") +
+  scale_y_continuous(breaks = seq(62.6, 63.6, length.out = 5)) +
+  scale_x_continuous(breaks = seq(-148.5, -146, length.out = 3)) +
+  theme_bw(base_size = 14)
+
+p2 <- ggplot(spbin_aug, aes(color = .std.resid)) +
+  geom_sf(size = 1) +
[…]

### 2025-08-21 inst/manuscript/manuscript.tex
some updates

--- 2024-11-20 inst/manuscript/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
 \documentclass[
-  article]{jss}
-
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
+]{jss}
+
+%% recommended packages
+\usepackage{orcidlink,thumbpdf,lmodern}
+
 \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else
-    % xetex/luatex font selection
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
-\usepackage{orcidlink,thumbpdf,lmodern}
-
-\newcommand{\class}[1]{`\code{#1}'}
-\newcommand{\fct}[1]{\code{#1()}}
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Figure}
-\else
-  \newcommand\figurename{Figure}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
+
+\author{
+Michael Dumelle~\orcidlink{0000-0002-3393-5529}\\United States\\
+Environmental Protection Agency \And Jay M. Ver
+Hoef~\orcidlink{0000-0003-4302-6895}\\Alaska Fisheries\\
+Science Center \And Matt
+Higham~\orcidlink{0009-0006-4217-625X}\\St.~Lawrence University
 }
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}{}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
-\makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\usepackage{bookmark}
-
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdftitle={Spatial Generalized Linear Models in  Using },
-  pdfauthor={Michael Dumelle; Jay M. Ver Hoef; Matt Higham},
-  pdfkeywords={Autoregressive Model, Geostatistics, Spatial
-Covariance, Spatially-Explicit Model, Statistical Model},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-
-%% -- Article metainformation (author, title, ...) -----------------------------
-
-%% Author information
-\author{Michael Dumelle~\orcidlink{0000-0002-3393-5529}\\United States\\
-Environmental Protection Agency \And Jay M. Ver
-Hoef~\orcidlink{0000-0003-4302-6895}\\National Oceanic and\\
-Atmospheric Administration \AND Matt
-Higham~\orcidlink{0009-0006-4217-625X}\\St.~Lawrence University}
-\Plainauthor{Michael Dumelle, Jay M. Ver Hoef, Matt
-Higham} %% comma-separated
-
-\title{Spatial Generalized Linear Models in \proglang{R} Using
-\pkg{spmodel}}
-\Plaintitle{Spatial Generalized Linear Models in
-Using} %% without formatting
-
-%% an abstract and keywords
-\Abstract{Generalized linear models (GLMs) describe a non-normal
-response variable that may be binary, count, skewed, or a proportion.
-Typically, observations in a GLM are assumed independent of one another.
-For spatial data, this independence assumption is impractical, as nearby
+\title{Spatial Generalized Linear Models in {R} Using {spmodel}}
+
+\Plainauthor{Michael Dumelle, Jay M. Ver Hoef, Matt Higham}
+\Plaintitle{Spatial Generalized Linear Models in R Using spmodel}
+\Shorttitle{SPGLMs in {R} Using {spmodel}}
+
+\Abstract{
+Generalized linear models (GLMs) describe a non-normal response variable
+that may be binary, count, skewed, or a proportion. Typically,
+observations in a GLM are assumed independent of one another. For
+spatial data, this independence assumption is impractical, as nearby
 locations tend to be more similar than locations far apart. The
-\pkg{spmodel} \proglang{R} package provides tools to fit GLMs that
-incorporate spatial autocorrelation (i.e., spatial generalized linear
-models, or SPGLMs). SPGLMs are fit in \pkg{spmodel} using a novel
-application of the Laplace approximation via \fct{spglm} for
-point-referenced data or \fct{spgautor} for areal (i.e., lattice), data.
-\fct{spglm} and \fct{spgautor} closely resemble \fct{glm} from base
-\proglang{R} but include arguments that control the spatial
-autocorrelation structure. \pkg{spmodel} has many helper functions for
-model inspection and diagnostics, some of which leverage other
-\proglang{R} packages like \pkg{broom} and \pkg{emmeans}. \pkg{spmodel}
-has tools to make predictions of the latent spatial-mean process at
-unobserved locations. \pkg{spmodel} also provides many advanced features
-like accommodating geometric anisotropy and nonspatial random effects,
+{spmodel} {R} package provides tools to fit GLMs that incorporate
+spatial autocorrelation (i.e., spatial generalized linear models, or
+SPGLMs). SPGLMs are fit in {spmodel} using a novel application of the
+Laplace approximation via {spglm} for point-referenced data or
+{spgautor} for areal (i.e., lattice), data. {spglm} and {spgautor}
+closely resemble {glm} from base {R} but include arguments that control
+the spatial autocorrelation structure. {spmodel} has many helper
+functions for model inspection and diagnostics, some of which leverage
+other {R} packages like {broom} and {emmeans}. {spmodel} has tools to
+make predictions of the latent spatial-mean process at unobserved
+locations. {spmodel} also provides many advanced features like
+accommodating geometric anisotropy and nonspatial random effects,
 simulating spatially autocorrelated data, and more. Here we use
-\pkg{spmodel} to illustrate the modeling of binary, count, skewed and
+{spmodel} to illustrate the modeling of binary, count, skewed and
 proportion response variables from several point-referenced and areal
-data sets.}
-
-%% at least one keyword must be supplied
-\Keywords{Autoregressive Model, Geostatistics, Spatial
-Covariance, Spatially-Explicit Model, Statistical Model}
-\Plainkeywords{Autoregressive Model, Geostatistics, Spatial
-Covariance, Spatially-Explicit Model, Statistical Model}
+data sets.
+}
+
+\Keywords{autoregressive model, geostatistical model, spatial
+autocovariance, spatial autocorrelation}
+\Plainkeywords{autoregressive model, geostatistical model, spatial
+autocovariance, spatial autocorrelation}
 
 %% publication information
-%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
 %% \Volume{50}
@@
 %% \Year{2012}
-%% \Submitdate{2012-06-04}
+%% \Submitdate{}
 %% \Acceptdate{2012-06-04}
-%% \setcounter{page}{1}
-%% \Pages{1--xx}
-
-%% The address of (at least) one author should be given
-%% in the following format:
+
 \Address{
     Michael Dumelle\\
-Pacific Ecological Systems Division\\
+    United States\\
+Environmental Protection Agency\\
     200 SW 35th St\\
-Corvallis OR\\
-E-mail: \email{dumelle.michael@epa.gov}\\
-\\~
-Jay M. Ver Hoef\\
-Alaska Fisheries Science Center\\
-\\~
-Matt Higham\\
-Department of Math, Computer Science, and Statistics\\
-\\~
+Corvallis, OR, 97330\\
+  E-mail: \email{Dumelle.Michael@epa.gov}\\
 
@@
 
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{amsmath,amsfonts,amssymb}
+\usepackage{bm, bbm}
+\usepackage{lineno}
+\usepackage{caption, subcaption}
+
 \begin{document}
-\maketitle
 
@@
 likelihood (e.g., iteratively reweighted least squares)
-\citep{chambers1992S}. The \fct{glm} function is commonly used to fit
-GLMs in the \proglang{R} programming language \citep{rcore2024}.
+\citep{chambers1992S}. The {glm} function is commonly used to fit GLMs
+in the {R} programming language \citep{rcore2024}.
 
@@
 
-The \pkg{spmodel} \proglang{R} package provides tools for fitting
-spatial statistical models and making predictions at unobserved
-locations \citep{dumelle2023spmodel}. A recent major update (v0.4.0) to
-\pkg{spmodel} added SPGLM (Equation\(~\)\ref{eq-spglm}) support for
-binary, count, skewed, and proportion response variables
-(Table\(~\)\ref{table-links}), greatly expanding the class of models
-that \pkg{spmodel} makes accessible to practitioners.
+The {spmodel} {R} package provides tools for fitting spatial statistical
+models and making predictions at unobserved locations
+\citep{dumelle2023spmodel}. A recent major update (v0.4.0) to {spmodel}
+added SPGLM (Equation\(~\)\ref{eq-spglm}) support for binary, count,
+skewed, and proportion response variables (Table\(~\)\ref{table-links}),
+greatly expanding the class of models that {spmodel} makes accessible to
+practitioners.
 
 \begin{table}
-
-\centering{
-
 \centering
@@
 \end{tabular}
-
-}
-
-\caption{\label{table-links}SPGLM response distributions and their link functions and data types.}
-
-\end{table}%
+\caption{SPGLM response distributions and their link functions and data types.}
+\label{table-links}
+\end{table}
 
 SPGLMs for point-referenced data are called geostatistical GLMs and are
-fit using \fct{spglm}. Data are point-referenced when the elements in
+fit using {spglm}. Data are point-referenced when the elements in
 \(\mathbf{y}\) are observed at point-locations indexed by x-coordinates
@@
 data are called spatial autoregresive models and are fit using
-\fct{spgautor}. Data are areal (i.e., lattice) when they are part of a
+{spgautor}. Data are areal (i.e., lattice) when they are part of a
 finite network of polygons whose connections are indexed by a
@@
 
-Several other \proglang{R} packages exist for analyzing SPGLMs. The
-\pkg{brms} \citep{burkner2017brms}, \pkg{carBayes}
-\citep{lee2013carbayes}, \pkg{ngspatial} \citep{hughes2020ngspatial},
-\pkg{R-INLA} \citep{lindgren2015bayesian}, \pkg{spBayes}
-\citep{finley2007spbayes}, and \pkg{spNNGP} \citep{finley2002spnngp}
-packages take a Bayesian approach, either directly sampling from
-posterior distributions of parameters (e.g., using MCMC) or
-approximating them. A benefit of Bayesian approaches is that prior
-information can be incorporated and uncertainty quantification of
-parameter estimates is straightforward. However, Bayesian approaches,
-especially those using MCMC, tend to be computationally expensive. In
-order to reduce computation time, many of these packages work with the
-precision matrix instead of the covariance matrix so that inverting
-matrices (a computational bottleneck) is not required. For example,
-\pkg{R-INLA} uses the precision matrix (often directly modeled as with
-autoregressive models) and tends to be very fast. Working with precision
-matrices, however, can be more restrictive and less intuitive than
-working directly with the covariance matrix. The \pkg{FRK}
-\citep{sainsbury2024modeling}, \pkg{glmmTMB} \citep{brooks2017glmmtmb},
-\pkg{hglm} \citep{ronnegard2010hglm}, \pkg{mgcv}
-\citep{wood2017generalized}, and \pkg{spaMM} \citep{rousset2014spamm}
+Several other {R} packages exist for analyzing SPGLMs. The {brms}
+\citep{burkner2017brms}, {carBayes} \citep{lee2013carbayes}, {ngspatial}
+\citep{hughes2020ngspatial}, {R-INLA} \citep{lindgren2015bayesian},
+{spBayes} \citep{finley2007spbayes}, and {spNNGP}
+\citep{finley2002spnngp} packages take a Bayesian approach, either
+directly sampling from posterior distributions of parameters (e.g.,
+using MCMC) or approximating them. A benefit of Bayesian approaches is
+that prior information can be incorporated and uncertainty
+quantification of parameter estimates is straightforward. However,
+Bayesian approaches, especially those using MCMC, tend to be
+computationally expensive. In order to reduce computation time, many of
+these packages work with the precision matrix instead of the covariance
+matrix so that inverting matrices (a computational bottleneck) is not
+required. For example, {R-INLA} uses the precision matrix (often
+directly modeled as with autoregressive models) and tends to be very
+fast. Working with precision matrices, however, can be more restrictive
+and less intuitive than working directly with the covariance matrix. The
+{FRK} \citep{sainsbury2024modeling}, {glmmTMB}
+\citep{brooks2017glmmtmb}, {hglm} \citep{ronnegard2010hglm}, {mgcv}
+\citep{wood2017generalized}, and {spaMM} \citep{rousset2014spamm}
 packages directly use Laplace, quasi-likelihood, or reduced-rank
@@
 is that prior information cannot be formally incorporated and covariance
-parameter uncertainty is challenging to quantify. SPGLMs in
-\pkg{spmodel} are most similar to those in \pkg{glmmTMB} --
-\pkg{spmodel} uses analytical solutions for maximizing the likelihood
-(which we describe in the next section), while \pkg{glmmTMB} uses
-automatic differentiation for maximizing the likelihood (which tends to
-be much slower than analytical solutions).
-
-Missing from the aforementioned \proglang{R} packages is the complete
-set of tools for SPGLMs that \pkg{spmodel} provides. Importantly, the
-\fct{spglm} and \fct{spautor} functions act as a spatial analogue to the
-familiar \fct{glm} from base \proglang{R}, making the transition from
-GLMs to SPGLMs relatively seamless. \pkg{spmodel} leverages many
-commonly used \proglang{R} generics like \fct{summary} to better
-understand fitted models. Six GLM families (Table\(~\)\ref{table-links})
-and 20 spatial covariance (or precision) functions are supported. Also
-available are functions for data visualization, model fitting, model
-summaries, model diagnostics, model comparison, and prediction, all
-crucial components of a data analysis. Through extra function arguments,
-\pkg{spmodel} supports advanced features like geometric anisotropy,
-non-spatial random effects, methods for large data sets, and more.
-Importantly, \pkg{spmodel} extends popular packages like \pkg{broom}
-\citep{robinson2021broom, kuhn2022tidy} and \pkg{emmeans}
-\citep{lenth2024emmeans}. \pkg{spmodel} also provides function for
-simulating SPGLM data (e.g., \fct{sprbinom}, \fct{sprpois}).
+parameter uncertainty is challenging to quantify. SPGLMs in {spmodel}
+are most similar to those in {glmmTMB} -- {spmodel} uses analytical
+solutions for maximizing the likelihood (which we describe in the next
+section), while {glmmTMB} uses automatic differentiation for maximizing
+the likelihood (which tends to be much slower than analytical
+solutions).
+
+Missing from the aforementioned {R} packages is the complete set of
+tools for SPGLMs that {spmodel} provides. Importantly, the {spglm} and
+{spautor} functions act as a spatial analogue to the familiar {glm} from
+base {R}, making the transition from GLMs to SPGLMs relatively seamless.
+{spmodel} leverages many commonly used {R} generics like {summary} to
+better understand fitted models. Six GLM families
+(Table\(~\)\ref{table-links}) and 20 spatial covariance (or precision)
+functions are supported. Also available are functions for data
+visualization, model fitting, model summaries, model diagnostics, model
+comparison, and prediction, all crucial components of a data analysis.
+Through extra function arguments, {spmodel} supports advanced features
+like geometric anisotropy, non-spatial random effects, methods for large
+data sets, and more. Importantly, {spmodel} extends popular packages
+like {broom} \citep{robinson2021broom, kuhn2022tidy} and {emmeans}
+\citep{lenth2024emmeans}. {spmodel} also provides function for
+simulating SPGLM data (e.g., {sprbinom}, {sprpois}).
 
 The rest of this article is organized as follows. In
-Section~\ref{sec-spglm}, we provide some background for the SPGLM
-fitting and prediction routines in \pkg{spmodel}. In
-Section~\ref{sec-applications}, we provide several applications of
-\pkg{spmodel} to spatial binary, count, skewed and proportion data with
-both point-referenced and areal supports. And in
-Section~\ref{sec-discussion}, we end with a discussion synthesizing
-\pkg{spmodel}'s contributions to the analysis of SPGLMs in \proglang{R}.
+Section\(~\)\ref{sec-spglm}, we provide some background for the SPGLM
+fitting and prediction routines in {spmodel}. In
+Section\(~\)\ref{sec-applications}, we provide several applications of
+{spmodel} to spatial binary, count, skewed and proportion data with both
+point-referenced and areal supports. And in
+Section\(~\)\ref{sec-discussion}, we end with a discussion synthesizing
+\pkg{spmodel}'s contributions to the analysis of SPGLMs in {R}.
 
@@
 \citep{lee1996hierarchical, bolker2009generalized, wood2017generalized}.
-\pkg{spmodel} leverages this hierarchical structure and uses a novel
+{spmodel} leverages this hierarchical structure and uses a novel
 application of the Laplace approximation \citep{ver2024marginal} to fit
@@
 
-The \fct{spglm} (for point-referenced data) and \fct{spgautor} (for
-areal data) functions in \pkg{spmodel} fit SPGLMs using the Laplace
-approximation outlined in Section~\ref{sec-spglm}. Both \fct{spglm} and
-\fct{spgautor} generally require the following four arguments:
-\texttt{formula}, a formula that describes the relationship between the
-response variable and explanatory variables; \texttt{family}, the
-response distribution (which can be \texttt{binomial}, \texttt{poisson},
-\texttt{nbinomial}, \texttt{Gamma}, \texttt{inverse.gaussian}, or
-\texttt{beta}); \texttt{data}, the data frame that holds the variables
-in \texttt{formula} as well as spatial locations; and
-\texttt{spcov\_type}, the spatial covariance type. The first three
-arguments are shared by \fct{glm}; thus, the transition from GLMs to
-SPGLMs requires only one additional argument: \texttt{spcov\_type}. The
-\fct{spglm} spatial covariance types measure dependence as a function of
-Euclidean distance among observations; an example is the exponential
-spatial covariance: \begin{equation}\label{eq-spcov-exp}
+The {spglm} (for point-referenced data) and {spgautor} (for areal data)
+functions in {spmodel} fit SPGLMs using the Laplace approximation
+outlined in Section\(~\)\ref{sec-spglm}. Both {spglm} and {spgautor}
+generally require the following four arguments: \texttt{formula}, a
+formula that describes the relationship between the response variable
+and explanatory variables; \texttt{family}, the response distribution
+(which can be \texttt{binomial}, \texttt{poisson}, \texttt{nbinomial},
+\texttt{Gamma}, \texttt{inverse.gaussian}, or \texttt{beta});
+\texttt{data}, the data frame that holds the variables in
+\texttt{formula} as well as spatial locations; and \texttt{spcov\_type},
+the spatial covariance type. The first three arguments are shared by
+{glm}; thus, the transition from GLMs to SPGLMs requires only one
+additional argument: \texttt{spcov\_type}. The {spglm} spatial
+covariance types measure dependence as a function of Euclidean distance
+among observations; an example is the exponential spatial covariance:
+\begin{equation}\label{eq-spcov-exp}
   \boldsymbol{\Sigma} = \sigma^2_{de} \exp(-\mathbf{H}/\phi) + \sigma^2_{ie}\mathbf{I},
 \end{equation} where \(\mathbf{H}\) is a matrix of pairwise distances
-among all observations. \fct{spglm} currently supports 18 distinct
-spatial covariance functions.
-
-The \fct{spgautor} spatial covariance types measure dependence as a
-function of neighborhood distance among observations; an example is the
+among all observations. {spglm} currently supports 18 distinct spatial
+covariance functions.
+
+The {spgautor} spatial covariance types measure dependence as a function
+of neighborhood distance among observations; an example is the
 simultaneous autoregressive covariance matrix:
@@
 \end{equation} where \(\mathbf{W}\) is a matrix that represents the
-neighborhood structure among all observations. \fct{spgautor} currently
+neighborhood structure among all observations. {spgautor} currently
 supports two distinct spatial covariance functions.
 
-In the rest of this section, we use \pkg{spmodel} to study binary,
-count, skewed, and proportion response variables that are either
-point-referenced or areal. We use \pkg{spmodel} for all parts of the
-data analysis, from estimation to inference to model diagnostics to
-prediction. We first describe core \pkg{spmodel} functionality in an
+In the rest of this section, we use {spmodel} to study binary, count,
+skewed, and proportion response variables that are either
+point-referenced or areal. We use {spmodel} for all parts of the data
+analysis, from estimation to inference to model diagnostics to
+prediction. We first describe core {spmodel} functionality in an
 application to binary data, while additional analyses highlight count,
-skewed, and proportion data as well as some additional \pkg{spmodel}
-features. Before proceeding, load \pkg{spmodel} into the current
-\proglang{R} session:
-
-\begin{verbatim}
+skewed, and proportion data as well as some additional {spmodel}
+features. Before proceeding, load {spmodel} into the current {R}
+session:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> library("spmodel")
-\end{verbatim}
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
-The \texttt{moose} data in \pkg{spmodel} contain information on moose
+The \texttt{moose} data in {spmodel} contain information on moose
 presence in Alaska. They are an \texttt{sf} object, a special data frame
-that is supplemented with spatial information using the \pkg{sf} package
+that is supplemented with spatial information using the {sf} package
 \citep{pebesma2018sf}. The first few rows look like:
 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 R> head(moose)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 Simple feature collection with 6 features and 4 fields
@@
 6  164. M         0 0        (291421.5 1518398)
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 projected coordinate of each site. The \texttt{moose\_preds} data in
-\pkg{spmodel} contain spatial locations at which predictions of moose
+{spmodel} contain spatial locations at which predictions of moose
 presence are desired. They are also an \texttt{sf} object with the same
 projection and measurements for \texttt{elev} and \texttt{strat}.
-Figure~\ref{fig-moose-data} shows the \texttt{presence} variable in
+Figure\(~\)\ref{fig-moose-data} shows the \texttt{presence} variable in
 \texttt{moose} as well as the spatial locations of both \texttt{moose}
@@
 \begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=0.7\textheight]{figures/figure-1.png}
-
-}
-
-\caption{\label{fig-moose-data}Moose presence in Alaska. Circles
-represent moose presence or absence (based on color) and triangles
-represent locations at which moose presence probability predictions are
-desired.}
-
-\end{figure}%
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-1.png}
+\caption{Moose presence in Alaska. Circles represent moose presence or absence (based on color) and triangles represent locations at which moose presence probability predictions are desired.}
+\label{fig-moose-data}
+\end{figure}
 
@@
 SPGLM for binary data (i.e., a spatial logistic regression model) using
-\fct{spglm}:
-
-\begin{verbatim}
+{spglm}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> spbin <- spglm(
@@
 +   spcov_type = "spherical"
-+)
-\end{verbatim}
++ )
+\end{CodeInput}
+\end{CodeChunk}
 
 Summarizing the model object yields a summary similar to that provided
-by the familiar \fct{glm}:
-
-\begin{verbatim}
+by the familiar {glm}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> summary(spbin)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 
@@
          1 
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 effects. While useful, this summary information is hard to work with, as
-it is printed directly to the \proglang{R} console. The \pkg{broom}
-package from the tidymodels \citep{kuhn2022tidy} ecosystem has functions
-to provide helpful model output in the form of tibbles (i.e., data
-frames) that are easily manipulated. \pkg{spmodel} has methods for the
-\fct{tidy}, \fct{glance}, and \fct{augment} functions from \pkg{broom}.
-The first \pkg{broom} function is \fct{tidy}, which tidies the model
-output:
-
-\begin{verbatim}
+it is printed directly to the {R} console. The {broom} package from the
+tidymodels \citep{kuhn2022tidy} ecosystem has functions to provide
+helpful model output in the form of tibbles (i.e., data frames) that are
+easily manipulated. {spmodel} has methods for the {tidy}, {glance}, and
+{augment} functions from {broom}. The first {broom} function is {tidy},
+which tidies the model output:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> tidy(spbin)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 4 x 5
@@
 4 elev:stratM -0.0109    0.00670     -1.62 0.104  
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
 The estimates and standard errors returned are on the log odds link
-(Table\(~\)\ref{table-links}) scale (\fct{coef} and \fct{vcov} may also
-be used). The output provides evidence that elevation is positively
+(Table\(~\)\ref{table-links}) scale ({coef} and {vcov} may also be
+used). The output provides evidence that elevation is positively
 associated with moose presence in the \texttt{"L"} stratum (\(p-\)value
@@
 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 R> moose$strat2 <- factor(moose$strat, levels = c("M", "L"))
@@
 +   summary()
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 
@@
          1 
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
 A simpler solution, especially if there categorical variables with many
-levels, is to leverage \pkg{emmeans}. \pkg{emmeans} is an \proglang{R}
-package for estimating marginal means of model objects. The
-\fct{emtrends} function in \pkg{emmeans} characterizes the effect of a
-continuous variable (here, \texttt{elev}) for each level of a
-categorical variable (here, \texttt{strat}):
-
-\begin{verbatim}
+levels, is to leverage {emmeans}. {emmeans} is an {R} package for
+estimating marginal means of model objects. The {emtrends} function in
+{emmeans} characterizes the effect of a continuous variable (here,
+\texttt{elev}) for each level of a categorical variable (here,
+\texttt{strat}):
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> library("emmeans")
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> emtrends(spbin, "strat", "elev")
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
  strat elev.trend      SE  df asymp.LCL asymp.UCL
@@
 Confidence level used: 0.95 
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-The second \pkg{broom} function is \fct{glance}, which glances at the
-model fit:
-
-\begin{verbatim}
+The second {broom} function is {glance}, which glances at the model fit:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> glance(spbin)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 1 x 10
@@
 1   218     4     3  681.  687.  687.  697.  -340.     161.           0.0885
-\end{verbatim}
-
-\fct{glance} returns several useful statistics like the sample size
+\end{CodeOutput}
+\end{CodeChunk}
+
+{glance} returns several useful statistics like the sample size
 (\texttt{n}), number of fixed effects (\texttt{p}), number of covariance
@@
 \begin{figure}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=1\textheight]{figures/figure-2.png}
-
-}
-
-\caption{\label{fig-sp-diagnostic}Spatial logistic regression model
-diagnostics from \fct{augment}. The leverage (i.e., hat) values (left)
-and standardized residuals (right).}
-
-\end{figure}%
-
-The third \pkg{broom} function is \fct{augment}, which augments the
-model data with diagnostics:
-
-\begin{verbatim}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-2.png}
+\caption{Spatial logistic regression model diagnostics from [augment]{.fct}. The leverage (i.e., hat) values (left) and standardized residuals (right).}
+\label{fig-sp-diagnostic}
+\end{figure}
+
+The third {broom} function is {augment}, which augments the model data
+with diagnostics:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> head(augment(spbin))
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 Simple feature collection with 6 features and 8 fields
@@
 # i 1 more variable: geometry <POINT [m]>
-\end{verbatim}
-
-The \fct{augment} function returns the fitted \(\mathbf{w}\) values
+\end{CodeOutput}
+\end{CodeChunk}
+
+The {augment} function returns the fitted \(\mathbf{w}\) values
 (\texttt{.fitted}), deviance residuals (\texttt{.resid}), leverage
@@
 and standardized residuals (\texttt{.std.resid}). When the data are an
-\texttt{sf} object, \fct{augment} returns another \texttt{sf} object,
+\texttt{sf} object, {augment} returns another \texttt{sf} object,
 helpful for visualizing model diagnostics spatially as in
-Figure~\ref{fig-sp-diagnostic}. Leverage measures the unusualness of an
-observation's set of explanatory variables, while Cook's distances
+Figure\(~\)\ref{fig-sp-diagnostic}. Leverage measures the unusualness of
+an observation's set of explanatory variables, while Cook's distances
 measures how influential an observation is on the resulting model fit
@@
 accessible as vectors using the appropriate generic function (e.g.,
-\fct{fitted}, \fct{residuals}).
-
-Similar to \fct{glm} model objects, \fct{plot} can be used to visualize
-diagnostics (Figure~\ref{fig-sp-diagnostic2}):
-
-\begin{verbatim}
+{fitted}, {residuals}).
+
+Similar to {glm} model objects, {plot} can be used to visualize
+diagnostics (Figure\(~\)\ref{fig-sp-diagnostic2}):
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> plot(spbin, which = c(4, 7))
-\end{verbatim}
+\end{CodeInput}
+\end{CodeChunk}
 
 \begin{figure}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=1\textheight]{figures/figure-3.png}
-
-}
-
-\caption{\label{fig-sp-diagnostic2}Spatial logistic regression model
-diagnostics from \fct{plot}. The Cook's distance values (left) and the
-fitted spatial covariance as a function of distance (right).}
-
-\end{figure}%
-
-Components of model variation are partitioned using \fct{varcomp}:
-
-\begin{verbatim}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-3.png}
+\caption{Spatial logistic regression model diagnostics from [plot]{.fct}. The Cook's distance values (left) and the fitted spatial covariance as a function of distance (right).}
+\label{fig-sp-diagnostic2}
+\end{figure}
+
+Components of model variation are partitioned using {varcomp}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> varcomp(spbin)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 3 x 2
@@
 3 ie                   0.000462
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 \begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=0.7\textheight]{figures/figure-4.png}
-
-}
-
-\caption{\label{fig-moose-fit}Moose presence probability fitted values
-and predictions. Fitted values are represeneted by circles and
-predictions by triangles.}
-
-\end{figure}%
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-4.png}
+\caption{Moose presence probability fitted values and predictions. Fitted values are represeneted by circles and predictions by triangles.}
+\label{fig-moose-fit}
+\end{figure}
 
 We make predictions of the log odds of moose probability presence at
-each site in \texttt{moose\_preds} using \fct{predict}:
-
-\begin{verbatim}
+each site in \texttt{moose\_preds} using {predict}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> head(predict(spbin, newdata = moose_preds))
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
           1           2           3           4           5           6 
  0.08588581 -0.40762380 -1.87510889 -1.14172781  1.45701519 -2.74275553 
-\end{verbatim}
-
-\fct{augment} may also be used to augment the prediction data with
+\end{CodeOutput}
+\end{CodeChunk}
+
+{augment} may also be used to augment the prediction data with
 predictions:
 
-\begin{verbatim}
-head(augment(
-  spbin,
-  newdata = moose_preds,
-  type.predict = "response",
-  interval = "prediction"
-))
-\end{verbatim}
-
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+R> head(augment(
++   spbin,
++   newdata = moose_preds,
++   type.predict = "response",
++   interval = "prediction"
++ ))
+\end{CodeInput}
+\begin{CodeOutput}
 Simple feature collection with 6 features and 5 fields
@@
 6  218. L      0.0605 0.00360  0.534 (304473.8 1512103)
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
 Here, we requested predictions on the probability (i.e., response) scale
-(Figure~\ref{fig-moose-fit}) alongside lower and upper bounds of a 95\%
-(see \texttt{level}) prediction interval (Figure~\ref{fig-moose-int}).
+(Figure\(~\)\ref{fig-moose-fit}) alongside lower and upper bounds of a
+95\% (see \texttt{level}) prediction interval
+(Figure\(~\)\ref{fig-moose-int}).
 
 \begin{figure}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=1\textheight]{figures/figure-5.png}
-
-}
-
-\caption{\label{fig-moose-int}Moose presence probability prediction
-intervals. 95\% prediction interval lower bound (left) and 95\%
-prediction interval upper bound (right).}
-
-\end{figure}%
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-5.png}
+\caption{Moose presence probability prediction intervals. 95\% prediction interval lower bound (left) and 95\% prediction interval upper bound (right).}
+\label{fig-moose-int}
+\end{figure}
 
@@
 \texttt{spcov\_type\ =\ "none"}, the resulting model fit is nearly
-identical to that from \fct{glm}:
-
-\begin{verbatim}
-R> # fit nonspatial model using spglm()
+identical to that from {glm}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> bin <- spglm(
@@
 +   spcov_type = "none"
-+)
-R> # fit nonspatial model using glm()
++ )
 R> bin_glm <- glm(
@@
 +   data = moose
-+)
-R> # compare fixed effect coefficients and standard errors up to four decimals
++ )
 R> data.frame(
@@
 +   se_glm = sqrt(diag(vcov(bin_glm)))
-+) |>
++ ) |>
 +   apply(2, round, digits = 4)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
             est_none est_glm se_none se_glm
@@
 elev:stratM  -0.0013 -0.0013  0.0038 0.0038
-\end{verbatim}
-
-The advantage of using \fct{spglm} to fit a model with
+\end{CodeOutput}
+\end{CodeChunk}
+
+The advantage of using {spglm} to fit a model with
 \texttt{spcov\_type\ =\ "none"} is that it provides access to other
-\pkg{spmodel} functions for model objects (e.g., \fct{glances} below)
-and accounts for the additional terms in the likelihood from
+{spmodel} functions for model objects (e.g., {glances} below) and
+accounts for the additional terms in the likelihood from
 Equation\(~\)\eqref{eq-marginal6}. These additional terms make the
-likelihood for \fct{spglm} and \fct{glm} different, though the models
-convey the same information. A glance at the spatial and nonspatial
-models reveals:
-
-\begin{verbatim}
+likelihood for {spglm} and {glm} different, though the models convey the
+same information. A glance at the spatial and nonspatial models reveals:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> glances(spbin, bin)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 2 x 11
@@
 # i 1 more variable: pseudo.r.squared <dbl>
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 R> loocv(spbin) |>
 +   apply(2, round, digits = 4)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
    bias    MSPE   RMSPE 
 -0.0006  0.1458  0.3818 
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeOutput}
+\begin{CodeInput}
 R> loocv(bin) |>
 +   apply(2, round, digits = 4)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
   bias   MSPE  RMSPE 
 0.0000 0.2403 0.4902 
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 R> AUROC(spbin)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 [1] 0.9490741
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeOutput}
+\begin{CodeInput}
 R> AUROC(bin)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 [1] 0.647138
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 \begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=0.7\textheight]{figures/figure-6.png}
-
-}
-
-\caption{\label{fig-moose-data-count}Moose counts in Alaska. Circles
-represent moose counts (based on color) and triangles represent
-locations at which mean count predictions are desired.}
-
-\end{figure}%
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-6.png}
+\caption{Moose counts in Alaska. Circles represent moose counts (based on color) and triangles represent locations at which mean count predictions are desired.}
+\label{fig-moose-data-count}
+\end{figure}
 
 The \texttt{count} variable in \texttt{moose} contains the number of
-moose observed at a site (Figure~\ref{fig-moose-data-count}). Count data
-are often modeled using Poisson or negative binomial regression with the
-log link function. The Poisson regression model assumes each datum's
-underlying latent mean equals its variance, while the negative binomial
-accommodates overdispersion (where the variance is greater than the
-mean) at the cost of estimating an extra parameter.
+moose observed at a site (Figure\(~\)\ref{fig-moose-data-count}). Count
+data are often modeled using Poisson or negative binomial regression
+with the log link function. The Poisson regression model assumes each
+datum's underlying latent mean equals its variance, while the negative
+binomial accommodates overdispersion (where the variance is greater than
+the mean) at the cost of estimating an extra parameter.
 
@@
 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 R> sppois <- spglm(
@@
 +   anisotropy = TRUE
-+)
++ )
 R> 
 R> spnbin <- update(sppois, family = nbinomial)
-\end{verbatim}
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 R> glances(sppois, spnbin, sort_by = "AIC") |>
 +   subset(select = c(model, npar, AIC, AICc, BIC))
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 2 x 5
@@
 2 sppois     5 1320. 1321. 1337.
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 R> loocv(sppois) |>
 +   apply(2, round, digits = 4)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
    bias    MSPE   RMSPE 
  1.2882 31.7882  5.6381 
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeOutput}
+\begin{CodeInput}
 R> loocv(spnbin) |>
 +   apply(2, round, digits = 4)
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
    bias    MSPE   RMSPE 
  0.3485 28.3760  5.3269 
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 R> spnbin_iso <- update(spnbin, anisotropy = FALSE)
@@
 +   subset(select = c(model, npar, AIC, AICc, BIC))
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 2 x 5
@@
 2 spnbin_iso     4 1333. 1333. 1346.
-\end{verbatim}
-
-\fct{plot} returns the spatial covariance as a function of direction
-(Figure~\ref{fig-tropy}):
-
-\begin{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
+
+{plot} returns the spatial covariance as a function of direction
+(Figure\(~\)\ref{fig-tropy}):
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> plot(spnbin_iso, which = 8)
+\end{CodeInput}
+
+\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}
+
+\begin{CodeInput}
 R> plot(spnbin, which = 8)
-\end{verbatim}
+\end{CodeInput}
+
+\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-23-2} \end{center}
+
+\end{CodeChunk}
 
 \begin{figure}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=1\textheight]{figures/figure-7.png}
-
-}
-
-\caption{\label{fig-tropy}Level curves of equal autocorrelation for the
-negative binomial moose count models. The ellipse is centered at zero
-distance in the x-direction and y-direction, and points along the
-ellipse have equal levels of autocorrelation. In the isotropic level
-curve (left), spatial covariance decays equally in all directions. In
-the anistropic level curve (right), spatial covariance decays fastest in
-the northeast-southwest direction and slowest in the northwest-southeast
-direction (this pattern can be seen in the observed counts).}
-
-\end{figure}%
-
-Earlier we used \fct{tidy} to tidy the model's fixed effects, but we can
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-7.png}
+\caption{Level curves of equal autocorrelation for the negative binomial moose count models.
+The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of autocorrelation.
+In the isotropic level curve (left), spatial covariance decays equally in all directions.
+In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts).}
+\label{fig-tropy}
+\end{figure}
+
+Earlier we used {tidy} to tidy the model's fixed effects, but we can
 also use tidy to tidy the spatial covariance parameters:
 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 R> tidy(spnbin, effects = "spcov")
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 5 x 3
@@
 5 scale      0.259 FALSE   
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-The \texttt{seal} data in \pkg{spmodel} is an \texttt{sf} object with
-data on harbor seal trends in Alaska. The \texttt{log\_trend} variable
-is the logarithm of a seal abundance temporal trend measure at the site
-(based on historical data), and the \texttt{stock} variable is a factor
-with two levels, \texttt{8} and \texttt{10}, where each level represents
-one of twelve seal stocks (i.e., breeds) in Alaska. The \texttt{seal}
+The \texttt{seal} data in {spmodel} is an \texttt{sf} object with data
+on harbor seal trends in Alaska. The \texttt{log\_trend} variable is the
+logarithm of a seal abundance temporal trend measure at the site (based
+on historical data), and the \texttt{stock} variable is a factor with
+two levels, \texttt{8} and \texttt{10}, where each level represents one
+of twelve seal stocks (i.e., breeds) in Alaska. The \texttt{seal}
 geometry is an areal polygon geometry and hence, spatial autoregressive
@@
 
-SPGLMs for areal data are are fit in \pkg{spmodel} using \fct{spgautor},
-which has similar syntax as \fct{spglm} but contains arguments to
-control the weight matrix (\(\mathbf{W}\) in
-Equation\(~\)\ref{eq-spcov-sar}) and whether or not row-standardization
-\citep{ver2018spatial} is applied. By default, polygons are neighbors if
-they share a boundary (i.e., Queen's contiguity; see
-\citet{pebesma2023spatial}) and row standardization is assumed. Weight
-matrices may be provided via the \texttt{W} argument and row
-standardization may be ignored via the \texttt{row\_st} arguent.
+SPGLMs for areal data are are fit in {spmodel} using {spgautor}, which
+has similar syntax as {spglm} but contains arguments to control the
+weight matrix (\(\mathbf{W}\) in Equation\(~\)\ref{eq-spcov-sar}) and
+whether or not row-standardization \citep{ver2018spatial} is applied. By
+default, polygons are neighbors if they share a boundary (i.e., Queen's
+contiguity; see \citet{pebesma2023spatial}) and row standardization is
+assumed. Weight matrices may be provided via the \texttt{W} argument and
+row standardization may be ignored via the \texttt{row\_st} arguent.
 Following \citet{ver2018spatial}, polygons without a neighbor are given
@@
 \citep{ver2018spatial}, but we will exponentiate \texttt{log\_trend}
-(Figure~\ref{fig-seal}) and model this skew directly using a SPGLM:
-
-\begin{verbatim}
+(Figure\(~\)\ref{fig-seal}) and model this skew directly using a SPGLM:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> seal$trend <- exp(seal$log_trend)
-\end{verbatim}
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 \begin{figure}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=1\textheight]{figures/figure-8.png}
-
-}
-
-\caption{\label{fig-seal}Seal trend distribution in Alaska. Observed and
-missing seal polygons by stock (left) and observed seal trends (right).}
-
-\end{figure}%
-
-\pkg{spmodel} supports the gamma and inverse Gaussian families for
-modeling skewed, positive response variables. \pkg{spmodel} also
-supports nonspatial random effects specified via the \texttt{random}
-argument, which uses a similar formula syntax as \pkg{nlme}
-\citep{pinheiro2006mixed} and \pkg{lme4} \citep{bates2015lme4}. Using
-likelihood-based statistics, we compare two models fit using the
-simultaneous autoregressive covariance and the Gamma and inverse
-Gaussian families. Both models have a random effect for seal stock,
-which builds additional correlation into the model for two polygons from
-the same stock:
-
-\begin{verbatim}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-8.png}
+\caption{Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed seal trends (right).}
+\label{fig-seal}
+\end{figure}
+
+{spmodel} supports the gamma and inverse Gaussian families for modeling
+skewed, positive response variables. {spmodel} also supports nonspatial
+random effects specified via the \texttt{random} argument, which uses a
+similar formula syntax as {nlme} \citep{pinheiro2006mixed} and {lme4}
+\citep{bates2015lme4}. Using likelihood-based statistics, we compare two
+models fit using the simultaneous autoregressive covariance and the
+Gamma and inverse Gaussian families. Both models have a random effect
+for seal stock, which builds additional correlation into the model for
+two polygons from the same stock:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> spgam <- spgautor(
@@
 +   random = ~ stock
-+)
++ )
 R> spinvg <- update(spgam, family = inverse.gaussian)
@@
 2 spgam      5  114.  115.  127.
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-We may \fct{tidy} the estimated stock random effect variance:
-
-\begin{verbatim}
+We may {tidy} the estimated stock random effect variance:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> tidy(spinvg, effects = "randcov")
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 1 x 3
@@
 1 1 | stock  0.00411 FALSE   
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
-R> # output omitted
+\begin{CodeChunk}
+\begin{CodeInput}
 R> predict(spinvg, type = "response", interval = "prediction")
-\end{verbatim}
-
-If using \fct{augment} for prediction, \texttt{newdata} must be
-specified:
-
-\begin{verbatim}
+\end{CodeInput}
+\end{CodeChunk}
+
+If using {augment} for prediction, \texttt{newdata} must be specified:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> head(augment(
@@
 +   interval = "prediction"
-+))
-\end{verbatim}
-
-\begin{verbatim}
++ ))
+\end{CodeInput}
+\begin{CodeOutput}
 Simple feature collection with 6 features and 6 fields
@@
 6        NA 8        NA   0.942  0.893  0.993 ((1070501 1046969, 1070317 104659~
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 \begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=0.7\textheight]{figures/figure-9.png}
-
-}
-
-\caption{\label{fig-caribou}Caribou data from a spatial experimental
-design measuring the percentage of nitrogen (\texttt{z}) in soil and
-testing two factors: tarp type and water presence.}
-
-\end{figure}%
-
-The \texttt{caribou} data in \pkg{spmodel} are a data frame from a
-caribou foraging experiment in Alaska meant to study the impact of water
-and tarp cover on the percentage of nitrogen in surrounding plants
-(Figure~\ref{fig-caribou}). \citet{lenart2002climate} studied these data
-treating nitrogen percentage as a continuous variable, but here we treat
-nitrogen percentage (\texttt{z}) as a proportion:
-
-\begin{verbatim}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-9.png}
+\caption{Caribou data from a spatial experimental design measuring the percentage of nitrogen (z) in soil and testing two factors: tarp type and water presence.}
+\label{fig-caribou}
+\end{figure}
+
+The \texttt{caribou} data in {spmodel} are a data frame from a caribou
+foraging experiment in Alaska meant to study the impact of water and
+tarp cover on the percentage of nitrogen in surrounding plants
+(Figure\(~\)\ref{fig-caribou}). \citet{lenart2002climate} studied these
+data treating nitrogen percentage as a continuous variable, but here we
+treat nitrogen percentage (\texttt{z}) as a proportion:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> spbeta <- spglm(
@@
 +     ycoord = y
-+)
-\end{verbatim}
++ )
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
-tidy(anova(spbeta))
-\end{verbatim}
-
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+R> tidy(anova(spbeta))
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 4 x 4
@@
 4 water:tarp      2      5.74 0.0566       
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 Sometimes averages or contrasts between factor levels that are not in
-the reference group are of interest. We again leverage \pkg{spmodel}'s
-built-in support for \pkg{emmeans} and use it to obtain the averages of
-each factor combination on the link (here, logit) scale:
-
-\begin{verbatim}
-spemm <- emmeans(spbeta, ~ water + tarp)
-spemm
-\end{verbatim}
-
-\begin{verbatim}
+the reference group are of interest. We again leverage {spmodel}'s
+built-in support for {emmeans} and use it to obtain the averages of each
+factor combination on the link (here, logit) scale:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> spemm <- emmeans(spbeta, ~ water + tarp)
+R> spemm
+\end{CodeInput}
+\begin{CodeOutput}
  water tarp  emmean     SE  df asymp.LCL asymp.UCL
@@
 
-\begin{verbatim}
-update(spemm, type = "response")
-\end{verbatim}
-
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+R> update(spemm, type = "response")
+\end{CodeInput}
+\begin{CodeOutput}
  water tarp  response       SE  df asymp.LCL asymp.UCL
@@
 Intervals are back-transformed from the logit scale 
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
-head(pairs(spemm, adjust = "none"))
-\end{verbatim}
-
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+R> head(pairs(spemm, adjust = "none"))
+\end{CodeInput}
+\begin{CodeOutput}
  contrast          estimate     SE  df z.ratio p.value
@@
 Results are given on the log odds ratio (not the response) scale. 
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
 \begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=0.7\textheight]{figures/figure-10.png}
-
-}
-
-\caption{\label{fig-texas}Proportion of voter turnout in Texas for the
-1980 presidential election. Circles represent voter turnout (based on
-color) and triangles represent locations at which voter turnout
-predictions are desired.}
-
-\end{figure}%
-
-We now model the \texttt{elect80} data in \pkg{spData}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-10.png}
+\caption{Proportion of voter turnout in Texas for the 1980 presidential election.
+Circles represent voter turnout (based on color) and triangles represent locations at which voter turnout predictions are desired.}
+\label{fig-texas}
+\end{figure}
+
+We now model the \texttt{elect80} data in {spData}
 \citep{bivand2024spdata}, which contains voter turnout data by county in
 the 1980 US Presidential election \citep{pace1997quick}. The
-\texttt{texas} data in \pkg{spmodel} contains a subset of these data in
-the state of Texas. These data are point-referenced, but we may still
-use autoregressive models if neighborhood distance is determined using
+\texttt{texas} data in {spmodel} contains a subset of these data in the
+state of Texas. These data are point-referenced, but we may still use
+autoregressive models if neighborhood distance is determined using
 county centroids (i.e., counties whose centroid distance is less than
@@
 interest, \texttt{turnout}, is the proportion of registered voters in
-the county who voted in the election (Figure~\ref{fig-texas}).
-
-\begin{verbatim}
+the county who voted in the election (Figure\(~\)\ref{fig-texas}).
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> spgautor_mods <- spgautor(
@@
 +   estmethod = "ml"
-+)
-\end{verbatim}
++ ) 
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 distance cutoff of 200 kilometers and the maximum likelihood estimation
-method. When a vector is provided to \texttt{spcov\_type} in
-\fct{spgautor} (or \fct{spglm}), a model is fit for each spatial
-covariance type and stored in a list with name equal to the respective
-type. Then it is simple to glance at each model fit:
-
-\begin{verbatim}
+method. When a vector is provided to \texttt{spcov\_type} in {spgautor}
+(or {spglm}), a model is fit for each spatial covariance type and stored
+in a list with name equal to the respective type. Then it is simple to
+glance at each model fit:
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> glances(spgautor_mods) |>
 +   subset(select = c(model, npar, AIC, AICc, BIC))
-\end{verbatim}
-
-\begin{verbatim}
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 2 x 5
@@
 2 sar       3 -35.7 -35.5 -18.6
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
-tidy(spgautor_mods$car)
-\end{verbatim}
-
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+R> tidy(spgautor_mods$car)
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 2 x 5
@@
 2 log_income     0.532     0.117      4.55 0.00000531
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
-reduced_car <- update(spgautor_mods$car, formula = turnout ~ 1)
-tidy(anova(reduced_car, spgautor_mods$car))
-\end{verbatim}
-
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+R> reduced_car <- update(spgautor_mods$car, formula = turnout ~ 1)
+R> tidy(anova(reduced_car, spgautor_mods$car))
+\end{CodeInput}
+\begin{CodeOutput}
 # A tibble: 1 x 5
@@
 1 spgautor_mods$car reduced_car     1      24.1 0.000000930
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-The default estimation method in \pkg{spmodel} is REML, but note that
-these models used maximum likelihood (ML). ML is very similar to REML --
-the difference is that for ML, the fixed effects \(\boldsymbol{\beta}\)
-are not integrated out of Equation\(~\)\eqref{eq-marginal} but are
-rather back-substituted. While REML typically performs better for fixed
-effect estimation and prediction \citep{zimmerman2024spatial}, ML allows
+The default estimation method in {spmodel} is REML, but note that these
+models used maximum likelihood (ML). ML is very similar to REML -- the
+difference is that for ML, the fixed effects \(\boldsymbol{\beta}\) are
+not integrated out of Equation\(~\)\eqref{eq-marginal} but are rather
+back-substituted. While REML typically performs better for fixed effect
+estimation and prediction \citep{zimmerman2024spatial}, ML allows
 likelihood-based comparisons (e.g., AIC) for models with simultaneously
@@
 The point-referenced \texttt{texas} and \texttt{caribou} data may be
-analyzed using \fct{spgautor} or \fct{spglm} and comparisons across
-these structures can be made using likelihood-based statistics (as long
-as the supports of the response distribution are the same). Put another
-way, likelihood-based statistics can be used to determine whether
+analyzed using {spgautor} or {spglm} and comparisons across these
+structures can be made using likelihood-based statistics (as long as the
+supports of the response distribution are the same). Put another way,
+likelihood-based statistics can be used to determine whether
 geostatistical (distance-based) or autoregressive (neighbor-based)
@@
 
-SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace
+SPGLMs are fit in {spmodel} using a novel application of the Laplace
 approximation that marginalizes over the latent (i.e., unobserved) mean,
@@
 distribution and covariance structure. \citet{ver2024marginal} show that
-the approach, as implemented in \pkg{spmodel}, generally yields unbiased
+the approach, as implemented in {spmodel}, generally yields unbiased
 estimators with proper interval coverage and often outperforms the
-Bayesian approach from \pkg{spBayes}, the INLA approach from
-\pkg{R-INLA}, and the automatic differentiation approach from
-\pkg{glmmTMB}.
-
-\pkg{spmodel}'s \fct{spglm} and \fct{spgautor} functions are similar in
-structure and syntax as the base-\proglang{R} \fct{glm} function, easing
-the transition from GLMs to SPGLMs. These functions support six response
-distributions (Table\(~\)\ref{table-links}) and 20 spatial covariance
-functions. \pkg{spmodel} provides several additional features that
-accommodate geometric anisotropy, nonspatial random effects, fixing
-spatial covariance parameters at known values, data having thousands of
+Bayesian approach from {spBayes}, the INLA approach from {R-INLA}, and
+the automatic differentiation approach from {glmmTMB}.
+
+{spmodel}'s {spglm} and {spgautor} functions are similar in structure
+and syntax as the base-{R} {glm} function, easing the transition from
+GLMs to SPGLMs. These functions support six response distributions
+(Table\(~\)\ref{table-links}) and 20 spatial covariance functions.
+{spmodel} provides several additional features that accommodate
+geometric anisotropy, nonspatial random effects, fixing spatial
+covariance parameters at known values, data having thousands of
 observations (following \citet{ver2023indexing}), incorporating spatial
@@
 \citet{breiman2001random}), simulating spatially dependent data (e.g.,
-\fct{sprbinom}, \fct{sprpois}), and several others. Learn more at
+{sprbinom}, {sprpois}), and several others. Learn more at
 \url{https://CRAN.R-project.org/package=spmodel} and links therein.
@@
 
-The results in this paper were obtained using \proglang{R} 4.4.0 with
-the \pkg{spmodel} 0.9.0 package. Figures were created using the
-\pkg{ggplot2} 3.5.1 package \citep{wickham2016ggplot2} and base
-\proglang{R}.
+The results in this paper were obtained using {R} 4.4.0 with the
+{spmodel} 0.9.0 package. Figures were created using the {ggplot2} 3.5.1
+package \citep{wickham2016ggplot2} and base {R}.
 
@@
 \url{https://github.com/USEPA/spmodel.glm.manuscript}. All data used are
-part of the \pkg{spmodel} \proglang{R} package available for download
-from CRAN at \url{https://CRAN.R-project.org/package=spmodel}.
+part of the {spmodel} {R} package available for download from CRAN at
+\url{https://CRAN.R-project.org/package=spmodel}.
 
@@
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-%bib-loc-124C8010
-
 \bibliography{references.bib}

### 2025-08-22 inst/manuscript/manuscript.Rmd
some updates

@@
   # For running headers, if needed
-  formatted: "Spatial Generalized Linear Models in [R]{.proglang} Using [spmodel]{.pkg}"
+  formatted: "Spatial Generalized Linear Models in \\proglang{R} Using \\pkg{spmodel}"
   plain:     "Spatial Generalized Linear Models in R Using spmodel"
-  short:     "SPGLMs in [R]{.proglang} Using [spmodel]{.pkg}"
+  short:     "Spatial Generalized Linear Models in \\proglang{R} Using \\pkg{spmodel}"
 abstract: |
@@
 For spatial data, this independence assumption is impractical, as nearby locations tend to be more similar than locations far apart.
-The [spmodel]{.pkg} [R]{.proglang} package provides tools to fit GLMs that incorporate spatial autocorrelation (i.e., spatial generalized linear models, or SPGLMs).
-SPGLMs are fit in [spmodel]{.pkg} using a novel application of the Laplace approximation via [spglm]{.fct} for point-referenced data or [spgautor]{.fct} for areal (i.e., lattice), data.
-[spglm]{.fct} and [spgautor]{.fct} closely resemble [glm]{.fct} from base [R]{.proglang} but include arguments that control the spatial autocorrelation structure.
-[spmodel]{.pkg} has many helper functions for model inspection and diagnostics, some of which leverage other [R]{.proglang} packages like [broom]{.pkg} and [emmeans]{.pkg}.
-[spmodel]{.pkg} has tools to make predictions of the latent spatial-mean process at unobserved locations.
-[spmodel]{.pkg} also provides many advanced features like accommodating geometric anisotropy and nonspatial random effects, simulating spatially autocorrelated data, and more.
-Here we use [spmodel]{.pkg} to illustrate the modeling of binary, count, skewed and proportion response variables from several point-referenced and areal data sets.
+The \pkg{spmodel} \proglang{R} package provides tools to fit GLMs that incorporate spatial correlation (i.e., spatial generalized linear models, or SPGLMs).
+SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace approximation via \code{spglm()} for point-referenced data or \code{spgautor()} for areal (i.e., lattice), data.
+\code{spglm()} and \code{spgautor()} closely resemble [glm]{.fct} from base \proglang{R} but include arguments that control the spatial correlation structure.
+\pkg{spmodel} has many helper functions for model inspection and diagnostics, some of which leverage other \proglang{R} packages like [broom]{.pkg} and [emmeans]{.pkg}.
+\pkg{spmodel} has tools to make predictions of the latent spatial-mean process at unobserved locations.
+\pkg{spmodel} also provides many advanced features like accommodating geometric anisotropy and nonspatial random effects, simulating spatially autocorrelated data, and more.
+Here we use \pkg{spmodel} to illustrate the modeling of binary, count, skewed and proportion response variables from several point-referenced and areal data sets.
 keywords:
   # at least one keyword must be supplied
-  formatted: [autoregressive model, geostatistical model, spatial autocovariance, spatial autocorrelation]
-  plain:     [autoregressive model, geostatistical model, spatial autocovariance, spatial autocorrelation]
+  formatted: [autoregressive model, geostatistical model, spatial covariance, spatial correlation]
+  plain:     [autoregressive model, geostatistical model, spatial covariance, spatial correlation]
 preamble: |
@@
 
-Binary, count, proportion, and skewed data are ubiquitous in practice.
-These data types are naturally modeled using a generalized linear model (GLM) framework [@nelder1972generalized; @mccullagh1989generalized; @myers2012generalized; @faraway2016extending].
-In a GLM, a response variable $\text{y}$ belongs to a particular statistical distribution with mean $\mu$.
-For example, $\text{y}$ may be distributed as a Poisson random variable with some mean $\mu$.
-Based on the response distribution assumed for $\text{y}$, GLMs link a function of $\mu$ to explanatory variables via a link function:
+In practice, non-Gaussian data are ubiquitous.
+Non-Gaussian that belong to an exponential family data can be naturally modeled using a generalized linear model (GLM) regression framework [@nelder1972generalized; @mccullagh1989generalized; @myers2012generalized; @faraway2016extending].
+In a GLM, an $n \times 1$ response variable $\mathbf{y}$ belongs to a a statistical distribution (e.g., Poisson, Binomial) with some mean and variance.
+Often, the analysis goal is to study the impact of a linear function of several explanatory variables on $\text{y}$ through a GLM.
+In this context, the latent (i.e., unobserved) mean of $\mathbf{y}$, $\boldsymbol{\mu}$, is linked to these explanatory variables via a link function:
 \begin{equation}\label{eq-glm}
-f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
+f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
 \end{equation}
-where for a sample size $n$, $\boldsymbol{\mu}$ is the mean of an $n \times 1$ response vector $\mathbf{y}$, $f(\boldsymbol{\mu})$ is the link function that connects $\boldsymbol{\mu}$ to $\mathbf{w}$, $\mathbf{X}$ is the $n \times p$ design matrix of explanatory variables, and $\boldsymbol{\beta}$ is the $p \times 1$ vector of fixed effects.
-The mean, $\boldsymbol{\mu}$, is usually constrained in some way (e.g., positive) that depends on the distribution assumed for $\mathbf{y}$, but $\mathbf{w}$ is unconstrained.
-The parameters in Equation$~$\eqref{eq-glm} are typically estimated via maximum likelihood (e.g., iteratively reweighted least squares) [@chambers1992S].
-The [glm]{.fct} function is commonly used to fit GLMs in the [R]{.proglang} programming language [@rcore2024].
+where for a sample size $n$, $f(\cdot)$ is a link function that connects $\boldsymbol{\mu}$ to $\mathbf{w}$, $\mathbf{X}$ is the $n \times p$ design matrix of explanatory variables, and $\boldsymbol{\beta}$ is the $p \times 1$ vector of fixed effects.
+While the mean is typically constrained in some way (e.g., between zero and one if a probability), the link function generally makes $\mathbf{w}$ is unconstrained.
+Common link functions inlude the log odds (i.e., logit) link for binary and proprtion data and the log link count and skewed data.
+Equation$~$\ref{eq-glm} can also be written in terms of the inverse link function, $f^{-1}(\cdot)$:
+\begin{equation}\label{eq-glm2}
+\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}),
+\end{equation}
 
-The GLM framework in Equation$~$\eqref{eq-glm} assumes the elements of $\mathbf{y}$ are independent of one another.
-This assumption is impractical for spatial data, where nearby observations tend to be more similar than distant observations [@tobler1970computer].
-Ignoring this spatial dependence can give rise to misleading inference and poor prediction [@zimmerman2024spatial].
-Spatial GLMs (SPGLMs) formally incorporate spatial autocorrelation into a GLM by adding to Equation$~$\eqref{eq-glm} two random effects that elucidate spatial structure:
+The GLM fixed effects ($\boldsymbol{\beta}$) are typically estimated via maximum likelihood [@chambers1992S].
+It is often convenient to compute the maximum likelihood estimates using the iteratively reweighted least squares (IRWLS) algorithm [@wood2017generalized], which is the approach used by the  \code{glm()} function in the \proglang{R} programming language [@rcore2024].
+GLMs add an additional layer of complexity compared to linear regression models, as the left-hand size of Equation$~$\ref{eq-glm} is a function of the mean of $\mathbf{y}$ rather than $\mathbf{y}$ itself (as in linear regression models).
+
+The standard GLM assumes the elements of $\mathbf{y}$ are independent.
+This independence assumption is typically impractical for spatial data.
+In spatial data, nearby observations tend to be more similar than distant observations [@tobler1970computer], leading to positive spatial covariance among observations.
+The consequences of ignoring spatial covariance in statistical models for spatial data can be severe and include imprecise parameter estimates as well as misleading standard errors that inflate Type-I error rates and decrease power [@zimmerman2024spatial].
+
+An approach for handling spatial data using a GLM is to assume $\mathbf{w}$ has spatial covariance.
+This is achieved by adding to Equation$~$\ref{eq-glm} two random effects, $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$.
+The random effect $\boldsymbol{\tau}$ is an $n \times 1$ column vector of spatially dependent random errors.
+We assume that $\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}$ and $\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}$, where $\text{E}(\cdot)$ and $\text{Cov}(\cdot)$ denote expectation and covariance, respectively.
+The variance parameter $\sigma^2_\tau$ controls the magnitude of spatial covariance and is often called a partial sill, while the matrix $\mathbf{R}$ is an $n \times n$ spatial correlation matrix that depends on a range parameter controls the distance-decay rate of the spatial correlation.
+One example of a spatial covariance matrix is the "exponential", which is given by
+\begin{equation}\label{eq-spcov-exp}
+  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de} \exp(-\mathbf{H}/\phi),
+\end{equation}
+where $\mathbf{H}$ is a matrix of pairwise distances among the elements of $\mathbf{y}$ and $\phi$ is a range parameter.
+From Equation$~$\ref{eq-spcov-exp}, as the distance between two elements of $\mathbf{y}$ increases, the spatial covariance decreases, which reflects intuition.
+Moreover, as the range parameter, $\phi$, increases, the strength of spatial dependence increases (Figure$~$\ref{fig-range}).
+The random effect $\boldsymbol{\epsilon}$ is an $n \times 1$ column vector of independent random errors.
+We assume that $\text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}$ and $\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\epsilon \mathbf{I}$, where $\mathbf{I}$ is an $n \times n$ identity matrix.
+The variance parameter $\sigma^2_\epsilon$ controls the magnitude of nonspatial variability (i.e., fine-scale variation) and is often called a nugget.
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.7\linewidth]{figures/figure-01.png}
+\caption{An exponential spatial correlation function with varying range parameters.}
+\label{fig-range}
+\end{figure}
+
+Through inclusion of $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$, the spatial GLM (SPGLM) can be written as
 \begin{equation}\label{eq-spglm}
-f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
+f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
 \end{equation}
-where $\boldsymbol{\tau}$ is an $n \times 1$ column vector of spatially dependent random errors, and $\boldsymbol{\epsilon}$ is an $n \times 1$ column vector of spatially independent random errors.
-We make a few assumptions about $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$: first, that $\text{E}(\boldsymbol{\tau}) = \text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}$, where $\text{E}(\cdot)$ denotes expectation; second, that $\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}$, where $\mathbf{R}$ is an $n \times n$ matrix that determines the spatial dependence structure in $\mathbf{w}$ (which is linked to $\boldsymbol{\mu}$ and hence $\mathbf{y}$) and depends on a range parameter, $\phi$; third, that $\text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_\epsilon \mathbf{I}$, where $\mathbf{I}$ is an $n \times n$ identity matrix; and fourth, that $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$ are independent of one another.
-The parameter $\sigma^2_{\tau}$ is called the spatially dependent random error variance or partial sill.
-The parameter $\sigma^2_\epsilon$ is called the spatially independent random error variance or nugget.
-These two variance parameters are henceforth more intuitively written as $\sigma^2_{de}$ and $\sigma^2_{ie}$, respectively.
-The covariance of $\mathbf{w}$ is denoted $\boldsymbol{\Sigma}$ and is given by
+Often in spatial statistics, quantities are explicitly referenced with respect to $\mathbf{s}$, a vector of coordinates indexing the observation [@cressie1993statistics].
+For example, $\mathbf{y}$ and $\mathbf{X}$ may instead be written $\mathbf{y}(\mathbf{s})$ and $\mathbf{X}$, respectively.
+We acknowledge the utility of this nomenclature but drop the explicit dependence on $\mathbf{s}$ for simplicity of notation.
+Assuming independence among $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$, it follows that 
 \begin{equation}\label{eq-spcov}
- \boldsymbol{\Sigma} = \sigma^2_{de}\mathbf{R} + \sigma^2_{ie} \mathbf{I}.
+ \text{Cov}(\boldsymbol{\tau} + \boldsymbol{\epsilon}) = \text{Cov}(\boldsymbol{\tau}) + \text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{\tau}\mathbf{R} + \sigma^2_{\epsilon} \mathbf{I}.
 \end{equation}
-The parameters $\sigma^2_{de}$, $\phi$, and $\sigma^2_{ie}$ are elements of $\boldsymbol{\theta}$, the covariance parameter vector.
+To better align with intuition, we henceforth $\sigma^2_{\tau}$ as $\sigma^2_{de}$ (for spatial error variance) and $\sigma^2_{\epsilon}$ as $\sigma^2_{ie}$ (for independent error variance).
+The parameters $\sigma^2_{de}$, $\sigma^2_{ie}$, the range parameter $\phi$ in $\mathbf{R}$, and any other parameters in $\mathbf{R}$ compose $\boldsymbol{\theta}$, the covariance parameter vector.
 
-The [spmodel]{.pkg} [R]{.proglang} package provides tools for fitting spatial statistical models and making predictions at unobserved locations [@dumelle2023spmodel].
-A recent major update (v0.4.0) to [spmodel]{.pkg} added SPGLM (Equation$~$\ref{eq-spglm}) support for binary, count, skewed, and proportion response variables (Table$~$\ref{table-links}), greatly expanding the class of models that [spmodel]{.pkg} makes accessible to practitioners.
-\begin{table}
+Fitting and using SPGLMs is challenging both conceptually and computationally [@bolker2009generalized].
+Recently, however, there have been numerous, significant advances in \proglang{R} software that have made these models more accessible to practitioners.
+The \pkg{brms} [@burkner2017brms], \pkg{carBayes} [@lee2013carbayes], \pkg{ngspatial} [@hughes2020ngspatial], \pkg{R-INLA} [@lindgren2015bayesian] and \pkg{inlabru} [@bachl2019inlabru], \pkg{spBayes} [@finley2007spbayes], \pkg{spOccupancy} [@doser2022spoccupancy], \pkg{spAbundance} [@doser2024spabundance], and \pkg{spNNGP} [@finley2002spnngp] packages take a Bayesian approach, either directly sampling from posterior distributions of parameters (e.g., using MCMC) or approximating them.
+A benefit of Bayesian approaches is that prior information can be incorporated and uncertainty quantification of parameter estimates is straightforward.
+However, Bayesian approaches, especially those using MCMC, can be computationally expensive.
+In order to reduce computation time, many of these packages work with the precision matrix instead of the covariance matrix so that computationally expensive matrix inversion is not required.
+For example, \pkg{R-INLA} uses the precision matrix and tends to be very fast.
+Working with precision matrices, however, can be more restrictive and less intuitive than working directly with the covariance matrix.
+The [FRK]{.pkg} [@sainsbury2024modeling], \pkg{glmmTMB} [@brooks2017glmmtmb], \pkg{hglm} [@ronnegard2010hglm], \pkg{mgcv} [@wood2017generalized], and \pkg{spaMM} [@rousset2014spamm] packages directly use Laplace, quasi-likelihood, or reduced-rank approaches to estimate parameters.
+These direct approaches tend to be computationally efficient, as they don't rely on MCMC sampling.
+In contrast to the Bayesian approach, a drawback of these direct approaches is that prior information cannot be formally incorporated and covariance parameter uncertainty is more challenging to quantify.
+The \pkg{sdmTMB} [@anderson2024sdmtmb] package combines elements of \pkg{R-INLA}, \pkg{glmmTMB}, and properties of Gaussian Markov random fields to fit a wide variety of SPGLMs, and \pkg{tinyVAST} [@thorson2025tinyVAST] extends some of these models to multivariate or (dynamic) structural equation models.
+
+@ver2024marginal proposed a novel approach to fitting SPGLMs that leverages the Laplace approximation while marginalizing over both the latent $\mathbf{w}$ and the fixed effects ($\boldsymbol{\beta}$) and accommodating spatial covariance.
+@ver2024marginal showed that this approach performed efficiently in a variety of simulation settings, generally having appropriate confidence interval coverage for the fixed effects and prediction interval coverage for new $\mathbf{w}$.
+The approach performed similarly to the Bayesian SPGLM approach in \pkg{spBayes} and the automatic differentiation SPGLM approach in \pkg{glmmTMB} but was much faster.
+At small sample sizes, the approach outperformed the approximate Bayesian SPGLM approach in \pkg{R-INLA} and had similar computational times.
+For moderate sample sizes, it performed similarly to \pkg{R-INLA}, though \pkg{R-INLA} was faster.
+This novel approach is particularly attractive for two reasons.
+First, it is general enough that can be applied to any covariance structure (not just spatial).
+Second, after estimating the covariance parameters, analytical solutions exist for the fixed effects (and their standard errors) as well as predictions of the latent $\mathbf{w}$ at new locations (and their standard errors).
+The \pkg{spmodel} \proglang{R} package [@dumelle2023spmodel] recently provided full support for the methods in @ver2024marginal applied to binary, count, skewed, and proportion data for over 20 different spatial covariance types.
+
+The \pkg{spmodel} \proglang{R} package [@dumelle2023spmodel] recently provided a full set of modeling tools for SPGLMs fit using the methods described in @ver2024marginal.
+These modeling tools are approachable and mirror the familiar \code{glm()} syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs relatively seamless.
+The \code{spglm()} function fits SPGLMs for point-referenced data (e.g., x-coordinates and y-coordinates representing point locations in a field), while the \code{spgautor()} function fits SPGLMs for areal data (e.g., polygon boundaries representing geographic subsets of a region).
+\pkg{spmodel} supports the binomial distribution for binary data, Poisson and negative binomial distributions for count data, Gamma and inverse Gaussian distributions for skewed data, and the beta distribution for proportion data.
+There are 20 different spatial covariance structures available including the exponential, Gaussian, and spherical for point-referenced data (Figure$~$\ref{fig-type}) and the conditional autoregressive, and simultaneous autoregressive structures for areal data.
+\pkg{spmodel} provides tools for commonly used model summaries, visualizations, and diagnostics (e.g., Cook's distance) using standard \proglang{R} helper functions like \code{summary()}, \code{plot()}, and \code{cooks.distance()}.
+\pkg{spmodel} also provides tools to predict $\mathbf{w}$ at new locations and quantify uncertainty in those prediction using \code{predict()}.
+This core functionality, combined with several advanced features we describe throughout the manuscript, enable \pkg{spmodel} to provide some novel and important capabilities previously missing from the existing SPGLM ecosystem in \proglang{R}.
+
+\begin{figure}
 \centering
-\begin{tabular}{||c|ccc||} 
- \hline
- Family & Link Function & Link Name & Data Type  \\ [0.5ex] 
- \hline \hline
- Binomial & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Binary; Binary Count \\
- Poisson & $f(\mu) = \log(\mu)$ & Log & Count \\
- Negative Binomial & $f(\mu) = \log(\mu)$ & Log & Count \\
- Beta & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Proportion \\
- Gamma & $f(\mu) = \log(\mu)$ & Log & Skewed \\
- Inverse Gaussian & $f(\mu) = \log(\mu)$ & Log & Skewed \\ 
- \hline\hline
-\end{tabular}
-\caption{SPGLM response distributions and their link functions and data types.}
-\label{table-links}
-\end{table}
-SPGLMs for point-referenced data are called geostatistical GLMs and are fit using [spglm]{.fct}.
-Data are point-referenced when the elements in $\mathbf{y}$ are observed at point-locations indexed by x-coordinates and y-coordinates on a spatially continuous surface with an infinite number of locations (e.g., point locations in a field).
-SPGLMs for areal data are called spatial autoregresive models and are fit using [spgautor]{.fct}.
-Data are areal (i.e., lattice) when they are part of a finite network of polygons whose connections are indexed by a neighborhood structure (e.g., states in a country).
+\includegraphics[width = 0.7\linewidth]{figures/figure-02.png}
+\caption{Exponential, Gaussian, and spherical spatial correlation functions all with range parameters equal to 0.5.}
+\label{fig-type}
+\end{figure}
 
-Several other [R]{.proglang} packages exist for analyzing SPGLMs.
-The [brms]{.pkg} [@burkner2017brms], [carBayes]{.pkg} [@lee2013carbayes], [ngspatial]{.pkg} [@hughes2020ngspatial], [R-INLA]{.pkg} [@lindgren2015bayesian], [spBayes]{.pkg} [@finley2007spbayes], and [spNNGP]{.pkg} [@finley2002spnngp] packages take a Bayesian approach, either directly sampling from posterior distributions of parameters (e.g., using MCMC) or approximating them.
-A benefit of Bayesian approaches is that prior information can be incorporated and uncertainty quantification of parameter estimates is straightforward.
-However, Bayesian approaches, especially those using MCMC, tend to be computationally expensive.
-In order to reduce computation time, many of these packages work with the precision matrix instead of the covariance matrix so that inverting matrices (a computational bottleneck) is not required.
-For example, [R-INLA]{.pkg} uses the precision matrix (often directly modeled as with autoregressive models) and tends to be very fast.
-Working with precision matrices, however, can be more restrictive and less intuitive than working directly with the covariance matrix.
-The [FRK]{.pkg} [@sainsbury2024modeling], [glmmTMB]{.pkg} [@brooks2017glmmtmb], [hglm]{.pkg} [@ronnegard2010hglm], [mgcv]{.pkg} [@wood2017generalized], and [spaMM]{.pkg} [@rousset2014spamm] packages directly use Laplace, quasi-likelihood, or reduced-rank approaches to estimate parameters.
-These direct approaches tend to be computationally efficient, as they don't rely on MCMC sampling.
-In contrast to the Bayesian approach, a drawback of these direct approaches is that prior information cannot be formally incorporated and covariance parameter uncertainty is challenging to quantify.
-SPGLMs in [spmodel]{.pkg} are most similar to those in [glmmTMB]{.pkg} -- [spmodel]{.pkg} uses analytical solutions for maximizing the likelihood (which we describe in the next section), while [glmmTMB]{.pkg} uses automatic differentiation for maximizing the likelihood (which tends to be much slower than analytical solutions).
-
-Missing from the aforementioned [R]{.proglang} packages is the complete set of tools for SPGLMs that [spmodel]{.pkg} provides.
-Importantly, the [spglm]{.fct} and [spautor]{.fct} functions act as a spatial analogue to the familiar [glm]{.fct} from base [R]{.proglang}, making the transition from GLMs to SPGLMs relatively seamless.
-[spmodel]{.pkg} leverages many commonly used [R]{.proglang} generics like [summary]{.fct} to better understand fitted models.
-Six GLM families (Table$~$\ref{table-links}) and 20 spatial covariance (or precision) functions are supported.
-Also available are functions for data visualization, model fitting, model summaries, model diagnostics, model comparison, and prediction, all crucial components of a data analysis.
-Through extra function arguments, [spmodel]{.pkg} supports advanced features like geometric anisotropy, non-spatial random effects, methods for large data sets, and more.
-Importantly, [spmodel]{.pkg} extends popular packages like [broom]{.pkg} [@robinson2021broom; @kuhn2022tidy] and [emmeans]{.pkg} [@lenth2024emmeans].
-[spmodel]{.pkg} also provides function for simulating SPGLM data (e.g., [sprbinom]{.fct}, [sprpois]{.fct}).
+\pkg{spmodel} (version 0.11.0) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in terms of scope and feel.
+Both packages use similar syntax as \code{glm()}, accommodate flexible \code{formula} arguments (e.g., offsets, splines), handle spatial covariance that decays at different rates in different rates (i.e., geometric anisotropy), incorporate nonspatial random effects, support other \proglang{R} packages for modeling like \pkg{broom} [@robinson2021broom; @kuhn2022tidy] and \pkg{emmeans} [@lenth2024emmeans], and have tools for model summaries, prediction, and simulating data.
+There are some notable differences between the two packages, however.
+\pkg{sdmTMB} supports several additional GLM distributions like the Tweedie, supports Hurdle models, and can incorporate prior information through Bayesian applications.
+\pkg{sdmTMB} also provides tools for working with temporal data and enhanced visualizations of marginal effects.
+\pkg{sdmTMB} does require a preprocessing step of constructing a mesh for the stochastic partial differential equation approach, and the density of the mesh can affect model results and computational complexity.
+\pkg{spmodel} does not require the construction of a mesh prior to modeling.
+\pkg{spmodel} supports 20 different spatial covariances and models them directly, rather than using a precision matrix approximation to the Matérn spatial covariance as in \pkg{sdmTMB}.
+\pkg{spmodel} also provides experimental design tools (e.g., analysis of variance, contrasts), supports \pkg{sf} objects in modeling and prediction functions [@pebesma2018sf], has several specialized model diagnostics like leverage values and Cook's distances, and has analytic solutions for prediction standard errors.
+Other similarities and differences do exist between \pkg{sdmTMB} and \pkg{spmodel}, and both packages continue to evolve.
+Overall, we believe that these packages are complementary and enhance the suite of SPGLM tools accessible to practitioners.
 
 The rest of this article is organized as follows.
-In Section$~$\ref{sec-spglm}, we provide some background for the SPGLM fitting and prediction routines in [spmodel]{.pkg}.
-In Section$~$\ref{sec-applications}, we provide several applications of [spmodel]{.pkg} to spatial binary, count, skewed and proportion data with both point-referenced and areal supports.
-And in Section$~$\ref{sec-discussion}, we end with a discussion synthesizing \pkg{spmodel}'s contributions to the analysis of SPGLMs in [R]{.proglang}.
+In Section$~$\ref{sec-spglm}, we provide some background for the SPGLM fitting and prediction routines in \pkg{spmodel}.
+In Section$~$\ref{sec-applications}, we provide several applications of \pkg{spmodel} to spatial binary, count, skewed and proportion data with both point-referenced and areal supports.
+And in Section$~$\ref{sec-discussion}, we end with a discussion synthesizing \pkg{spmodel}'s contributions to the analysis of SPGLMs in \proglang{R}.
 
-# Spatial generalized linear models using the Laplace approximation {#sec-spglm}
+# The spatial generalized linear model and marginalizatoin {#sec-spglm}
 
-GLMs with random effects are often written hierarchically [@lee1996hierarchical; @bolker2009generalized; @wood2017generalized].
-[spmodel]{.pkg} leverages this hierarchical structure and uses a novel application of the Laplace approximation [@ver2024marginal] to fit models.
-This marginal approach is quite flexible, accommodating a wide range of possible dependence structures and formally maximizing a likelihood.
-Maximizing a likelihood yields convenient likelihood-based statistics like AIC [@akaike1974new], AICc [@hoeting2006model], BIC [@schwarz1978estimating], deviance [@mccullagh1989generalized], and likelihood ratio tests for model comparison, a benefit compared to quasi-likelihood [@wedderburn1974quasi; @breslow1993approximate] or pseudo-likelihood approaches [@wolfinger1993generalized], which only specify the first two moments of a distribution.
-@ver2024marginal provide thorough context for the marginal approach and its associated details, but next we provide an short overview of the methodology.
+\pkg{spmodel} implements the novel methods described in @ver2024marginal to fit SPGLMs, which leverages the Laplace approximation and marginalizes over both the latent $\mathbf{w}$ and the fixed effects while accommodating spatial covariance.
+A beneficial aspect of this approach is that it formally maximizes a hierarchical GLM likelihood [@lee1996hierarchical; @wood2017generalized].
+This makes likelihood-based statistics for model comparison like AIC [@akaike1974new], AICc [@hoeting2006model], BIC [@schwarz1978estimating], deviance [@mccullagh1989generalized], and likelihood ratio tests available.
+These types of statistics are not available for quasi-likelihood [@wedderburn1974quasi; @breslow1993approximate] or pseudo-likelihood approaches [@wolfinger1993generalized], which only specify the first two moments of a distribution.
+@ver2024marginal provides thorough details regarding the method and contextualizes its development which built upon similar methods [@evangelou2011estimation, @bonat2016practical].
+Next, we describe a brief overview of the approach and how it can be used for parameter estimation, inference, and prediction.
 
-Our goal is to marginalize over the latent mean $\mathbf{w}$ and fixed effects $\boldsymbol{\beta}$ in Equation$~$\eqref{eq-spglm} to obtain a response distribution for $\mathbf{y}$ that depends only on the explanatory variables, $\mathbf{X}$, the response distribution's dispersion parameter, $\varphi$, and the covariance parameters, $\boldsymbol{\theta}$.
-We can represent this marginal distribution hierarchically as
+## Formulating the hierarchical likelihood
+
+We can write the SPGLM likelihood hierarchically as
 \begin{equation}\label{eq-marginal}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w} .
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w},
 \end{equation}
-In Equation$~$\eqref{eq-marginal}, $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ is the density for $\mathbf{y}$ (e.g., binomial, Poisson) given the latent mean and dispersion parameter and $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ is the Gaussian density for $\mathbf{w}$ given the explanatory variables, fixed effects, and covariance parameters.
-Integrating $\boldsymbol{\beta}$ out of Equation$~$\eqref{eq-spglm} yields
+where $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ is the density for the appropriate response distribution of $\mathbf{y}$ (e.g., binomial, Poisson) given the latent $\mathbf{w}$ and dispersion parameter ($\varphi$), and $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ is the multivariate Gaussian density for $\mathbf{w}$ given the explanatory variables ($\mathbf{X}$), fixed effects ($\boldsymbol{\beta}$), and spatial covariance parameters ($\boldsymbol{\theta}$).
+Following @@harville1977maximum, we can integrate $\boldsymbol{\beta}$ out of Equation$~$\ref{eq-spglm}, which yields
 \begin{equation}\label{eq-marginal2}
@@
 \end{equation}
-where $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ is the restricted (i.e., residual) Gaussian density [@patterson1971recovery, @harville1977maximum, @wolfinger1994computing] for $\mathbf{w}$ given the explanatory variables and covariance parameters.
-This restricted Gaussian density is given by
+where $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ is the restricted (i.e., residual) multivariate Gaussian density [@patterson1971recovery] for $\mathbf{w}$ given the explanatory variables and covariance parameters.
+Equation$~$\ref{eq-marginal2} can be synonymous written after profiling the overall variance out of $\boldsymbol{\Sigma}$, which reduces the dimension of $\boldsymbol{\theta}$ by one for optimization [@wolfinger1994computing].
+The restricted multivariate Gaussian density is given by
 \begin{equation}\label{eq-reml-def}
@@
 \end{equation}
-where $\tilde{\boldsymbol{\beta}} = (\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{w}$ and $|\cdot|$ denotes the determinant.
-
-Next, let $\ell_\mathbf{w} = \log([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}])$ and then notice that Equation$~$\eqref{eq-marginal2} can be written as 
+where $\tilde{\boldsymbol{\beta}} = (\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{w}$ and $|\cdot|$ denotes the determinant. Next, let 
+\begin{equation}\label{eq-marginal03}
+  \ell_\mathbf{w} = \log([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}])
+\end{equation}
+and rewrite Equation$~$\ref{eq-marginal2} as 
 \begin{equation}\label{eq-marginal3}
@@
 \end{equation}
-A Taylor series expansion of $\ell_\mathbf{w}$ around a point $\mathbf{a}$ yields
+A second-order Taylor series expansion of $\ell_\mathbf{w}$ around $\hat{\mathbf{w}}$ yields
 \begin{equation}\label{eq-marginal4}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_\mathbf{a} + \mathbf{g}^T(\mathbf{w} - \mathbf{a}) + \frac{1}{2}(\mathbf{w} - \mathbf{a})^T \mathbf{G} (\mathbf{w} - \mathbf{a}))d\mathbf{w},
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\hat{\mathbf{w}}} + \mathbf{g}^T(\mathbf{w} - \hat{\mathbf{w}}) + \frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T \mathbf{G} (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w},
 \end{equation}
-where $\mathbf{g}$ and $\mathbf{G}$ are the gradient and Hessian, respectively, of $\ell_\mathbf{w}$ with respect to $\mathbf{w}$. If $\mathbf{a}$ is a value for which $\mathbf{g} = \mathbf{0}$,
+where $\mathbf{g}$ and $\mathbf{G}$ are the gradient and Hessian, respectively, of $\ell_\mathbf{w}$ with respect to $\mathbf{w}$. If $\hat{\mathbf{w}}$ is a value for which $\mathbf{g} = \mathbf{0}$,
 \begin{equation}\label{eq-marginal5}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_\mathbf{a}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \mathbf{a})^T (-\mathbf{G}) (\mathbf{w} - \mathbf{a}))d\mathbf{w}.
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\hat{\mathbf{w}}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T (-\mathbf{G}) (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w}.
 \end{equation}
-Notice that the integral in Equation$~$\ref{eq-marginal5} can be solved in the same manner as the normalizing constant in a multivariate Gaussian distribution, and rewriting $\exp(\ell_\mathbf{a})$ yields
+The integral in Equation$~$\ref{eq-marginal5} can be solved by leveraging properties of the normalizing constant of a multivariate Gaussian distribution.
+Thus, rewriting $\exp(\ell_{\hat{\mathbf{w}}})$ yields
 \begin{equation}\label{eq-marginal6}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{a}), \varphi] [\mathbf{a} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{a}}|^{-1/2}.
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\hat{\mathbf{w}}), \varphi] [\hat{\mathbf{w}} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\hat{\mathbf{w}}}|^{-1/2}.
 \end{equation}
-@ver2024marginal show how to evaluate $\mathbf{g}$ and $\mathbf{G}$ to obtain \textbf{a} and maximize (the natural logarithm of) Equation$~$\eqref{eq-marginal6} for the six response distributions in Table$~$\ref{table-links}.
-Maximizing Equation$~$\eqref{eq-marginal6} using a Newton-Rhapson approach yields the marginal restricted maximum likelihood estimators $\hat{\varphi}$ and $\hat{\boldsymbol{\theta}}$.
+
+Maximizing the natural logarithm of Equation$~$\ref{eq-marginal6} requires a doubly iterative process over $\boldsymbol{\theta}$ and $\varphi$ as well as $\mathbf{w}$, eventually yielding the the marginal restricted maximum likelihood estimators $\hat{\varphi}$ and $\hat{\boldsymbol{\theta}}$ and their corresponding values of $\hat{\mathbf{w}}$.
+Maximizing this log likelihood is a computationally expensive operation that involves repeatedly evaluating $\boldsymbol{\Sigma}^{-1}$, $\mathbf{g}$, and $\mathbf{G}$; see @ver2024marginal for more details and forms of $\mathbf{g}$ and $\mathbf{G}$ for various response distributions.
+
+## Estimating fixed effects
+
 @ver2024marginal show that the fixed effect estimator is $\hat{\boldsymbol{\beta}} = (\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{X})^{-1}\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{w}$, where $\hat{\boldsymbol{\Sigma}}$ is the covariance matrix $\boldsymbol{\Sigma}$ evaluated at $\hat{\boldsymbol{\theta}}$.
+
+## Predicting at new locations
+
 @ver2024marginal also derive the covariance matrix of $\hat{\boldsymbol{\beta}}$ and show how to predict $\mathbf{w}$ at unobserved locations and quantify their uncertainties.
 
-# Application {#sec-applications}
+# Modeling moose presence in Alaska, USA {#sec-applications}
 
-The [spglm]{.fct} (for point-referenced data) and [spgautor]{.fct} (for areal data) functions in [spmodel]{.pkg} fit SPGLMs using the Laplace approximation outlined in Section$~$\ref{sec-spglm}.
-Both [spglm]{.fct} and [spgautor]{.fct} generally require the following four arguments: `formula`, a formula that describes the relationship between the response variable and explanatory variables; `family`, the response distribution (which can be `binomial`, `poisson`, `nbinomial`, `Gamma`, `inverse.gaussian`, or `beta`); `data`, the data frame that holds the variables in `formula` as well as spatial locations; and `spcov_type`, the spatial covariance type.
-The first three arguments are shared by [glm]{.fct}; thus, the transition from GLMs to SPGLMs requires only one additional argument: `spcov_type`.
-The [spglm]{.fct} spatial covariance types measure dependence as a function of Euclidean distance among observations; an example is the exponential spatial covariance:
-\begin{equation}\label{eq-spcov-exp}
-  \boldsymbol{\Sigma} = \sigma^2_{de} \exp(-\mathbf{H}/\phi) + \sigma^2_{ie}\mathbf{I},
-\end{equation}
-where $\mathbf{H}$ is a matrix of pairwise distances among all observations. [spglm]{.fct} currently supports 18 distinct spatial covariance functions.
+## Additional applications {#sec-applications2}
 
-The [spgautor]{.fct} spatial covariance types measure dependence as a function of neighborhood distance among observations; an example is the simultaneous autoregressive covariance matrix:
-\begin{equation}\label{eq-spcov-sar}
-  \boldsymbol{\Sigma} = \sigma^2_{de} [(\mathbf{I} - \phi \mathbf{W})(\mathbf{I} - \phi \mathbf{W})^T]^{-1} + \sigma^2_{ie}\mathbf{I},
-\end{equation}
-where $\mathbf{W}$ is a matrix that represents the neighborhood structure among all observations. [spgautor]{.fct} currently supports two distinct spatial covariance functions.
+## Modeling moose counts in Alaska, USA
 
-In the rest of this section, we use [spmodel]{.pkg} to study binary, count, skewed, and proportion response variables that are either point-referenced or areal.
-We use [spmodel]{.pkg} for all parts of the data analysis, from estimation to inference to model diagnostics to prediction.
-We first describe core [spmodel]{.pkg} functionality in an application to binary data, while additional analyses highlight count, skewed, and proportion data as well as some additional [spmodel]{.pkg} features.
-Before proceeding, load [spmodel]{.pkg} into the current [R]{.proglang} session:
-```{r, eval = FALSE}
-library("spmodel")
-```
+## Modeling harbor seal trends in Alaska, USA
 
-## Binary data
+## Modeling voter turnout in Texas, USA
 
-The `moose` data in [spmodel]{.pkg} contain information on moose presence in Alaska.
-They are an `sf` object, a special data frame that is supplemented with spatial information using the [sf]{.pkg} package [@pebesma2018sf].
-The first few rows look like:
-
-```{r}
-head(moose)
-```
-
-There are five columns: `elev`, the numeric site elevation (meters); `strat` a stratification variable for sampling with two levels, `"L"` and `"M"`, which are categorized by landscape metrics at each site; `count`, the number of moose at each site; `presence`, a factor that indicates whether at least one moose was observed at each site (`0` implies no moose; `1` implies at least one moose); and `geometry`, the NAD83 projected coordinate of each site.
-The `moose_preds` data in [spmodel]{.pkg} contain spatial locations at which predictions of moose presence are desired.
-They are also an `sf` object with the same projection and measurements for `elev` and `strat`.
-Figure$~$\ref{fig-moose-data} shows the `presence` variable in `moose` as well as the spatial locations of both `moose` and `moose_preds`.
-Moose are most common in the southwestern and eastern parts of the domain and least common in the northwest.
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-1.png}
-\caption{Moose presence in Alaska. Circles represent moose presence or absence (based on color) and triangles represent locations at which moose presence probability predictions are desired.}
-\label{fig-moose-data}
-\end{figure}
-
-To study the effect of elevation, stratum, and their interaction on moose presence while accounting for spatial autocorrelation, we fit a SPGLM for binary data (i.e., a spatial logistic regression model) using [spglm]{.fct}:
-
-```{r}
-spbin <- spglm(
-  formula = presence ~ elev + strat + elev:strat,
-  family = binomial,
-  data = moose,
-  spcov_type = "spherical"
-)
-```
-
-Summarizing the model object yields a summary similar to that provided by the familiar [glm]{.fct}:
-```{r}
-summary(spbin)
-```
-
-The summary contains the original function call, a summary of residuals, the fixed effects coefficients table, the spatial covariance parameter estimates, and additional model information like the pseudo R-squared, which quantifies the variability in the model attributable to the fixed effects.
-While useful, this summary information is hard to work with, as it is printed directly to the [R]{.proglang} console.
-The [broom]{.pkg} package from the tidymodels [@kuhn2022tidy] ecosystem has functions to provide helpful model output in the form of tibbles (i.e., data frames) that are easily manipulated.
-[spmodel]{.pkg} has methods for the [tidy]{.fct}, [glance]{.fct}, and [augment]{.fct} functions from [broom]{.pkg}.
-The first [broom]{.pkg} function is [tidy]{.fct}, which tidies the model output:
-```{r}
-tidy(spbin)
-```
-
-The estimates and standard errors returned are on the log odds link (Table$~$\ref{table-links}) scale ([coef]{.fct} and [vcov]{.fct} may also be used).
-The output provides evidence that elevation is positively associated with moose presence in the `"L"` stratum ($p-$value $<0.05$) and, at zero elevation, moose are more likely in the `"M"` stratum than the `"L"` stratum ($p-$value $<0.01$).
-This output provides marginal evidence that the effect of elevation on moose presence varies across strata ($p-$value $\approx 0.1$).
-The model effectively quantifies the impact of elevation on moose presence for moose in the `"L"` strata, but an analogous statement for moose in the `"M"` strata requires more context.
-We could refit the model treating `"M"` as the reference group instead of `"L"`:
-```{r}
-moose$strat2 <- factor(moose$strat, levels = c("M", "L"))
-update(spbin, formula = presence ~ elev + strat2 + elev:strat2) |> 
-  summary()
-```
-
-A simpler solution, especially if there categorical variables with many levels, is to leverage [emmeans]{.pkg}.
-[emmeans]{.pkg} is an [R]{.proglang} package for estimating marginal means of model objects.
-The [emtrends]{.fct} function in [emmeans]{.pkg} characterizes the effect of a continuous variable (here, `elev`) for each level of a categorical variable (here, `strat`):
-
-```{r, eval = FALSE}
-library("emmeans")
-```
-
-```{r}
-emtrends(spbin, "strat", "elev")
-```
-
-The asymptotic confidence intervals show that there is more evidence of an association between elevation and moose presence in the `"L"` stratum than in the `"M"` stratum.
-Notice that `elev.trend` for the `"L"` stratum matches the `elev` effect when `"L"` is the reference group, and similarly for `elev.trend` when `"M"` is the reference group.
-
-The second [broom]{.pkg} function is [glance]{.fct}, which glances at the model fit:
-```{r}
-glance(spbin)
-```
-
-[glance]{.fct} returns several useful statistics like the sample size (`n`), number of fixed effects (`p`), number of covariance parameters (`npar`), several likelihood-based statistics (e.g., `AIC`, `AICc`, `BIC`), and pseudo R-squared.
-
-\begin{figure}
-\centering
-\includegraphics[width = 1\linewidth]{figures/figure-2.png}
-\caption{Spatial logistic regression model diagnostics from [augment]{.fct}. The leverage (i.e., hat) values (left) and standardized residuals (right).}
-\label{fig-sp-diagnostic}
-\end{figure}
-
-The third [broom]{.pkg} function is [augment]{.fct}, which augments the model data with diagnostics:
-```{R}
-#| prompt: true
-head(augment(spbin))
-```
-
-The [augment]{.fct} function returns the fitted $\mathbf{w}$ values (`.fitted`), deviance residuals (`.resid`), leverage (i.e., hat) values (`.hat`), Cook's distance (`.cooksd`), and standardized residuals (`.std.resid`).
-When the data are an `sf` object, [augment]{.fct} returns another `sf` object, helpful for visualizing model diagnostics spatially as in Figure$~$\ref{fig-sp-diagnostic}.
-Leverage measures the unusualness of an observation's set of explanatory variables, while Cook's distances measures how influential an observation is on the resulting model fit [@montgomery2021introduction].
-Model diagnostics are also accessible as vectors using the appropriate generic function (e.g., [fitted]{.fct}, [residuals]{.fct}).
-
-Similar to [glm]{.fct} model objects, [plot]{.fct} can be used to visualize diagnostics (Figure$~$\ref{fig-sp-diagnostic2}):
-```{r, eval = FALSE}
-plot(spbin, which = c(4, 7))
-```
-
-\begin{figure}
-\centering
-\includegraphics[width = 1\linewidth]{figures/figure-3.png}
-\caption{Spatial logistic regression model diagnostics from [plot]{.fct}. The Cook's distance values (left) and the fitted spatial covariance as a function of distance (right).}
-\label{fig-sp-diagnostic2}
-\end{figure}
-
-Components of model variation are partitioned using [varcomp]{.fct}:
-```{r}
-varcomp(spbin)
-```
-The fixed effects explain roughly 9% of model variation, while the spatially dependent variance explains most of the remaining variability.
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-4.png}
-\caption{Moose presence probability fitted values and predictions. Fitted values are represeneted by circles and predictions by triangles.}
-\label{fig-moose-fit}
-\end{figure}
-
-We make predictions of the log odds of moose probability presence at each site in `moose_preds` using [predict]{.fct}:
-```{r}
-head(predict(spbin, newdata = moose_preds))
-```
-
-[augment]{.fct} may also be used to augment the prediction data with predictions:
-```{r}
-head(augment(
-  spbin,
-  newdata = moose_preds,
-  type.predict = "response",
-  interval = "prediction"
-))
-```
-
-Here, we requested predictions on the probability (i.e., response) scale (Figure$~$\ref{fig-moose-fit}) alongside lower and upper bounds of a 95% (see `level`) prediction interval (Figure$~$\ref{fig-moose-int}).
-
-\begin{figure}
-\centering
-\includegraphics[width = 1\linewidth]{figures/figure-5.png}
-\caption{Moose presence probability prediction intervals. 95\% prediction interval lower bound (left) and 95\% prediction interval upper bound (right).}
-\label{fig-moose-int}
-\end{figure}
-
-Thus far we have heuristically argued, based on first principles, that there are benefits to incorporating spatial autocorrelation for GLMs applied to spatial data.
-Now we provide some empirical evidence to support this claim by comparing the fits of the SPGLM and a GLM.
-If `spcov_type = "none"`, the resulting model fit is nearly identical to that from [glm]{.fct}:
-
-```{r}
-bin <- spglm(
-  formula = presence ~ elev + strat + elev:strat,
-  family = binomial,
-  data = moose,
-  spcov_type = "none"
-)
-bin_glm <- glm(
-  formula = presence ~ elev + strat + elev:strat,
-  family = binomial,
-  data = moose
-)
-data.frame(
-  est_none = coef(bin),
-  est_glm = coef(bin_glm),
-  se_none = sqrt(diag(vcov(bin))),
-  se_glm = sqrt(diag(vcov(bin_glm)))
-) |>
-  apply(2, round, digits = 4)
-```
-
-The advantage of using [spglm]{.fct} to fit a model with `spcov_type = "none"` is that it provides access to other [spmodel]{.pkg} functions for model objects (e.g., [glances]{.fct} below) and accounts for the additional terms in the likelihood from Equation$~$\eqref{eq-marginal6}.
-These additional terms make the likelihood for [spglm]{.fct} and [glm]{.fct} different, though the models convey the same information.
-A glance at the spatial and nonspatial models reveals:
-```{r}
-glances(spbin, bin)
-```
-
-The spatial model has a notably lower AIC, AICc, BIC, and deviance, suggesting it is the superior model.
-Another model comparison approach is leave-one-out cross validation.
-In leave-one-out cross validation, separately each observation is held out, a model is fit to the remaining data, and a prediction is made for the mean of the held out observation on the response scale.
-Then, statistics like leave-one-out bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) may be computed:
-```{r}
-loocv(spbin) |>
-  apply(2, round, digits = 4)
-loocv(bin) |>
-  apply(2, round, digits = 4)
-```
-
-Both models are nearly unbiased, but the spatial model has an approximately 39% lower MSPE, suggesting the probability predictions tend to be much closer to the observed presence values.
-
-A third model comparison tool is area under the receiver operating characteristic (AUROC) curve.
-The AUROC curve ranges from zero to one and conveys a model's classification performance over all possible probability thresholds [@james2013introduction].
-Larger values of AUROC indicate a more accurate model:
-```{r}
-AUROC(spbin)
-AUROC(bin)
-```
-
-All three performance metrics (likelihood-based statistics, leave-one-out statistics, and AUROC) prefer the spatial model.
-
-## Count data
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-6.png}
-\caption{Moose counts in Alaska. Circles represent moose counts (based on color) and triangles represent locations at which mean count predictions are desired.}
-\label{fig-moose-data-count}
-\end{figure}
-
-The `count` variable in `moose` contains the number of moose observed at a site (Figure$~$\ref{fig-moose-data-count}).
-Count data are often modeled using Poisson or negative binomial regression with the log link function.
-The Poisson regression model assumes each datum's underlying latent mean equals its variance, while the negative binomial accommodates overdispersion (where the variance is greater than the mean) at the cost of estimating an extra parameter.
-
-So far our spatial models have made an implicit assumption of geometric isotropy.
-A spatial covariance is geometrically isotropic if its dependence decays with distance equally in all directions.
-A spatial covariance is geometrically anisotropic if its dependence decays differently in different directions.
-The geometric anisotropy's directionality and strength are controlled by rotation and scale parameters that are applied to the original coordinates, creating a transformed set of coordinates whose spatial covariance is geometrically isotropic.
-Geometrically anisotropic models are fit by specifying `anisotropy`:
-
-```{r}
-sppois <- spglm(
-  formula = count ~ elev + strat + elev:strat,
-  family = poisson,
-  data = moose,
-  spcov_type = "gaussian",
-  anisotropy = TRUE
-)
-
-spnbin <- update(sppois, family = nbinomial)
-```
-
-Because the models have the same support (i.e., both non-negative count models), we can use likelihood-based statistics to compare them:
-```{r}
-glances(sppois, spnbin, sort_by = "AIC") |>
-  subset(select = c(model, npar, AIC, AICc, BIC))
-```
-
-The negative binomial model has a slightly lower AIC and AICc, while the Poisson model has a slightly lower BIC.
-This is reasonable given the BIC penalizes additional parameters (here, an overdispersion parameter) more heavily than AIC and AICc.
-The leave-out-out MSPE prefers the negative binomial model:
-```{r}
-loocv(sppois) |>
-  apply(2, round, digits = 4)
-loocv(spnbin) |>
-  apply(2, round, digits = 4)
-```
-
-A likelihood-based comparison between the negative binomial anisotropic model and the negative binomial isotropic model suggests that the anisotropic model is preferred (lower AIC, AICc, and BIC):
-```{r}
-spnbin_iso <- update(spnbin, anisotropy = FALSE)
-glances(spnbin_iso, spnbin, sort_by = "AIC") |>
-  subset(select = c(model, npar, AIC, AICc, BIC))
-```
-
-[plot]{.fct} returns the spatial covariance as a function of direction (Figure$~$\ref{fig-tropy}):
-```{r}
-plot(spnbin_iso, which = 8)
-plot(spnbin, which = 8)
-```
-
-\begin{figure}
-\centering
-\includegraphics[width = 1\linewidth]{figures/figure-7.png}
-\caption{Level curves of equal autocorrelation for the negative binomial moose count models.
-The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of autocorrelation.
-In the isotropic level curve (left), spatial covariance decays equally in all directions.
-In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts).}
-\label{fig-tropy}
-\end{figure}
-
-Earlier we used [tidy]{.fct} to tidy the model's fixed effects, but we can also use tidy to tidy the spatial covariance parameters:
-```{r}
-tidy(spnbin, effects = "spcov")
-```
-
-The `rotate` parameter is the number of radians in $[0, \pi]$ the ellipse is rotated and the `scale` parameter is the ratio in $(0, 1]$ of the minor axis length to the major axis length.
-The `is_known` column indicates whether the parameter was assumed known during optimization, controlled by specifying the `spcov_initial` argument.
-
-## Skewed data
-
-The `seal` data in [spmodel]{.pkg} is an `sf` object with data on harbor seal trends in Alaska.
-The `log_trend` variable is the logarithm of a seal abundance temporal trend measure at the site (based on historical data), and the `stock` variable is a factor with two levels, `8` and `10`, where each level represents one of twelve seal stocks (i.e., breeds) in Alaska.
-The `seal` geometry is an areal polygon geometry and hence, spatial autoregressive models based on neighborhood distance are appropriate.
-
-SPGLMs for areal data are are fit in [spmodel]{.pkg} using [spgautor]{.fct}, which has similar syntax as [spglm]{.fct} but contains arguments to control the weight matrix ($\mathbf{W}$ in Equation$~$\ref{eq-spcov-sar}) and whether or not row-standardization [@ver2018spatial] is applied.
-By default, polygons are neighbors if they share a boundary (i.e., Queen's contiguity; see @pebesma2023spatial) and row standardization is assumed.
-Weight matrices may be provided via the `W` argument and row standardization may be ignored via the `row_st` arguent.
-Following @ver2018spatial, polygons without a neighbor are given their own (independent) variance parameter called `extra`.
-
-The trend data were originally logged to remove skew [@ver2018spatial], but we will exponentiate `log_trend` (Figure$~$\ref{fig-seal}) and model this skew directly using a SPGLM:
-```{r}
-seal$trend <- exp(seal$log_trend)
-```
-
-The `trend` variable has several missing (`NA`) values, which represent polygons at which predictions of `trend` are desired.
-To make predictions using spatial autoregressive models, the prediction locations must be known prior to model fitting because these locations affect the neighborhood structure of the observed data [@ver2018spatial].
-This restriction is notably different than SPGLMs for point-referenced data (i.e., geostatistical models), which completely separates the estimation and prediction steps.
-
-\begin{figure}
-\centering
-\includegraphics[width = 1\linewidth]{figures/figure-8.png}
-\caption{Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed seal trends (right).}
-\label{fig-seal}
-\end{figure}
-
-[spmodel]{.pkg} supports the gamma and inverse Gaussian families for modeling skewed, positive response variables.
-[spmodel]{.pkg} also supports nonspatial random effects specified via the `random` argument, which uses a similar formula syntax as [nlme]{.pkg} [@pinheiro2006mixed] and [lme4]{.pkg} [@bates2015lme4].
-Using likelihood-based statistics, we compare two models fit using the simultaneous autoregressive covariance and the Gamma and inverse Gaussian families.
-Both models have a random effect for seal stock, which builds additional correlation into the model for two polygons from the same stock:
-```{r}
-spgam <- spgautor(
-  formula = trend ~ 1,
-  family = Gamma,
-  data = seal,
-  spcov_type = "sar",
-  random = ~ stock
-)
-spinvg <- update(spgam, family = inverse.gaussian)
-glances(spgam, spinvg, sort_by = "AIC") |>
-  subset(select = c(model, npar, AIC, AICc, BIC))
-```
-
-The inverse Gaussian model has a lower AIC, AICc, and BIC, which indicates it is a better fit than the gamma model.
-
-We may [tidy]{.fct} the estimated stock random effect variance:
-```{r}
-tidy(spinvg, effects = "randcov")
-```
-
-The locations to predict `trend` (`NA` values) are stored in the `newdata` element of `spinvg` and used for prediction:
-```{r, results="hide"}
-predict(spinvg, type = "response", interval = "prediction")
-```
-
-If using [augment]{.fct} for prediction, `newdata` must be specified:
-```{r}
-head(augment(
-  spinvg,
-  newdata = spinvg$newdata,
-  type.predict = "response",
-  interval = "prediction"
-))
-```
-
-## Proportion data
-
-We end with two examples of beta regression for proportion data [@ferrari2004beta].
-First, we model the nitrogen percentage in a caribou foraging experiment.
-Second, we model the proportion of voter turnout by Texas county in the United States (US) 1980 presidential election.
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-9.png}
-\caption{Caribou data from a spatial experimental design measuring the percentage of nitrogen (z) in soil and testing two factors: tarp type and water presence.}
-\label{fig-caribou}
-\end{figure}
-
-The `caribou` data in [spmodel]{.pkg} are a data frame from a caribou foraging experiment in Alaska meant to study the impact of water and tarp cover on the percentage of nitrogen in surrounding plants (Figure$~$\ref{fig-caribou}).
-@lenart2002climate studied these data treating nitrogen percentage as a continuous variable, but here we treat nitrogen percentage (`z`) as a proportion:
-
-```{r}
-spbeta <- spglm(
-    formula = z/100 ~ water + tarp + water:tarp,
-    family = "beta",
-    data = caribou,
-    spcov_type = "matern",
-    xcoord = x,
-    ycoord = y
-)
-```
-
-The nitrogen percentage is dynamically scaled in `formula` from (0, 100) to (0, 1) so that it is a proportion.
-Nitrogen percentage is modeled as a function of `water` (two levels: `"Y"` for water and `"N"` for no water), `tarp` (three levels: `"clear"` for a clear tarp, `"none"` for no tarp, and `"shade"` for a shaded tarp), and their interaction, which lets the effect of water presence vary across tarp type.
-`caribou` is a data frame (not an `sf` object), so we supply the x-coordinate and y-coordinate directly via `xcoord` and `ycoord`, and, consistent with the tidyverse approach [@wickham2019welcome], column names in `data` do not need to be quoted when referenced (but can be quoted).
-
-A summary of `spbeta` returns a coefficients table that provides parameter estimates relative to a reference group.
-When factors have more than two levels, it is not straightforward to use these contrasts to determine overall significance of the factor.
-The `anova()` function tests marginal (i.e., Type III sums of squares) significance of factors using the general linear hypothesis test for spatial (i.e., correlated) data [@schabenberger2017statistical]:
-
-```{r}
-tidy(anova(spbeta))
-```
-
-These results suggest there is some evidence that the effect of water on nitrogen percentage differ depending on the type of tarp used (0.01 < $p~$value < 0.1).
-
-Sometimes averages or contrasts between factor levels that are not in the reference group are of interest.
-We again leverage [spmodel]{.pkg}'s built-in support for [emmeans]{.pkg} and use it to obtain the averages of each factor combination on the link (here, logit) scale:
-```{r}
-spemm <- emmeans(spbeta, ~ water + tarp)
-spemm
-```
-
-Delta method [@ver2012invented] standard errors are used when averages on the response (here, proportion) scale are desired:
-```{r}
-update(spemm, type = "response")
-```
-
-Pairwise contrasts use a Tukey $p~$value adjustment [@tukey1949comparing] by default. Here, we request no $p~$value adjustment:
-```{r}
-head(pairs(spemm, adjust = "none"))
-```
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-10.png}
-\caption{Proportion of voter turnout in Texas for the 1980 presidential election.
-Circles represent voter turnout (based on color) and triangles represent locations at which voter turnout predictions are desired.}
-\label{fig-texas}
-\end{figure}
-
-We now model the `elect80` data in [spData]{.pkg} [@bivand2024spdata], which contains voter turnout data by county in the 1980 US Presidential election [@pace1997quick].
-The `texas` data in [spmodel]{.pkg} contains a subset of these data in the state of Texas.
-These data are point-referenced, but we may still use autoregressive models if neighborhood distance is determined using county centroids (i.e., counties whose centroid distance is less than some cutoff are defined as neighbors).
-The response variable of interest, `turnout`, is the proportion of registered voters in the county who voted in the election (Figure$~$\ref{fig-texas}).
-
-```{r}
-spgautor_mods <- spgautor(
-  formula = turnout ~ log_income,
-  family = beta,
-  data = texas,
-  spcov_type = c("car", "sar"),
-  cutoff = 2e5,
-  estmethod = "ml"
-) 
-```
-
-We model voter turnout as a function of log income using both the conditional and simultaneous autoregressive models with a neighbor distance cutoff of 200 kilometers and the maximum likelihood estimation method.
-When a vector is provided to `spcov_type` in [spgautor]{.fct} (or [spglm]{.fct}), a model is fit for each spatial covariance type and stored in a list with name equal to the respective type.
-Then it is simple to glance at each model fit:
-```{r}
-glances(spgautor_mods) |>
-  subset(select = c(model, npar, AIC, AICc, BIC))
-```
-
-The conditional autoregressive model has the best fit (in terms of AIC, AICc, and BIC).
-In this model, there is significant evidence `log_income` is positively related to average voter turnout ($p~$value < 0.001):
-```{r}
-tidy(spgautor_mods$car)
-```
-
-Another way to assess the impact of `log_income` on `turnout` is a likelihood ratio test:
-```{r}
-reduced_car <- update(spgautor_mods$car, formula = turnout ~ 1)
-tidy(anova(reduced_car, spgautor_mods$car))
-```
-
-Likelihood ratio tests compare the fit of a "full" model compared to a "reduced" model that is completely nested within the full model.
-If there is evidence the full model explains significantly more information than the reduced model, the likelihood ratio test $p~$value will be small.
-Similar to the summary output from the general linear hypothesis test, the likelihood ratio suggests `log_income` is related to average voter turnout ($p~$value < 0.001).
-
-The default estimation method in [spmodel]{.pkg} is REML, but note that these models used maximum likelihood (ML).
-ML is very similar to REML -- the difference is that for ML, the fixed effects $\boldsymbol{\beta}$ are not integrated out of Equation$~$\eqref{eq-marginal} but are rather back-substituted.
-While REML typically performs better for fixed effect estimation and prediction [@zimmerman2024spatial], ML allows likelihood-based comparisons (e.g., AIC) for models with simultaneously varying fixed effect and covariance structures, while REML likelihood-based comparisons are only valid for models sharing the same fixed effect structure (though @gurka2006selecting provides some evidence that this restriction may be unnecessary).
-
-The point-referenced `texas` and `caribou` data may be analyzed using [spgautor]{.fct} or [spglm]{.fct} and comparisons across these structures can be made using likelihood-based statistics (as long as the supports of the response distribution are the same).
-Put another way, likelihood-based statistics can be used to determine whether geostatistical (distance-based) or autoregressive (neighbor-based) structures perform best when the data are point-referenced.
-
-# Discussion {#sec-discussion}
-
-SPGLMs are fit in [spmodel]{.pkg} using a novel application of the Laplace approximation that marginalizes over the latent (i.e., unobserved) mean, $\mathbf{w}$, and the fixed effects, $\boldsymbol{\beta}$.
-The approach is quite flexible and accommodates any general response distribution and covariance structure.
-@ver2024marginal show that the approach, as implemented in [spmodel]{.pkg}, generally yields unbiased estimators with proper interval coverage and often outperforms the Bayesian approach from [spBayes]{.pkg}, the INLA approach from [R-INLA]{.pkg}, and the automatic differentiation approach from [glmmTMB]{.pkg}.
-
-[spmodel]{.pkg}'s [spglm]{.fct} and [spgautor]{.fct} functions are similar in structure and syntax as the base-[R]{.proglang} [glm]{.fct} function, easing the transition from GLMs to SPGLMs.
-These functions support six response distributions (Table$~$\ref{table-links}) and 20 spatial covariance functions.
-[spmodel]{.pkg} provides several additional features that accommodate geometric anisotropy, nonspatial random effects, fixing spatial covariance parameters at known values, data having thousands of observations (following @ver2023indexing), incorporating spatial dependence in machine learning (e.g., random forests; @breiman2001random), simulating spatially dependent data (e.g., [sprbinom]{.fct}, [sprpois]{.fct}), and several others.
-Learn more at [https://CRAN.R-project.org/package=spmodel](https://CRAN.R-project.org/package=spmodel) and links therein.
-
-# Computational details {.unnumbered}
-
-The results in this paper were obtained using [R]{.proglang} 4.4.0 with the
-[spmodel]{.pkg} 0.9.0 package. Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.
-
-# Data and code availability {.unnumbered}
-
-All writing and code associated with this manuscript is available for viewing and download on GitHub at [https://github.com/USEPA/spmodel.glm.manuscript](https://github.com/USEPA/spmodel.glm.manuscript).
-All data used are part of the [spmodel]{.pkg} [R]{.proglang} package available for download from CRAN at [https://CRAN.R-project.org/package=spmodel](https://CRAN.R-project.org/package=spmodel).
-
-# Acknowledgments {.unnumbered}
-
-We would like to thank initial reviewers and editors for feedback that has greatly improved the manuscript.
-
-The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of the U.S.
-Environmental Protection Agency or the National Oceanic and Atmospheric Administration.
-Any mention of trade names, products, or services does not imply
-an endorsement by the U.S. government, the U.S. Environmental Protection Agency, or the National Oceanic and Atmospheric
-Administration. The U.S. Environmental Protection Agency and the National Oceanic and Atmospheric Administration do not endorse
-any commercial products, services or enterprises.
+## Modeling lake conductivity in Southwest, USA
 

### 2025-08-22 inst/manuscript/manuscript.tex
some updates

@@
 }
-\title{Spatial Generalized Linear Models in {R} Using {spmodel}}
+\title{Spatial Generalized Linear Models in \proglang{R} Using
+\pkg{spmodel}}
 
@@
 \Plaintitle{Spatial Generalized Linear Models in R Using spmodel}
-\Shorttitle{SPGLMs in {R} Using {spmodel}}
+\Shorttitle{Spatial Generalized Linear Models in \proglang{R} Using
+\pkg{spmodel}}
 
@@
 locations tend to be more similar than locations far apart. The
-{spmodel} {R} package provides tools to fit GLMs that incorporate
-spatial autocorrelation (i.e., spatial generalized linear models, or
-SPGLMs). SPGLMs are fit in {spmodel} using a novel application of the
-Laplace approximation via {spglm} for point-referenced data or
-{spgautor} for areal (i.e., lattice), data. {spglm} and {spgautor}
-closely resemble {glm} from base {R} but include arguments that control
-the spatial autocorrelation structure. {spmodel} has many helper
-functions for model inspection and diagnostics, some of which leverage
-other {R} packages like {broom} and {emmeans}. {spmodel} has tools to
-make predictions of the latent spatial-mean process at unobserved
-locations. {spmodel} also provides many advanced features like
-accommodating geometric anisotropy and nonspatial random effects,
-simulating spatially autocorrelated data, and more. Here we use
-{spmodel} to illustrate the modeling of binary, count, skewed and
-proportion response variables from several point-referenced and areal
-data sets.
+\pkg{spmodel} \proglang{R} package provides tools to fit GLMs that
+incorporate spatial correlation (i.e., spatial generalized linear
+models, or SPGLMs). SPGLMs are fit in \pkg{spmodel} using a novel
+application of the Laplace approximation via \code{spglm()} for
+point-referenced data or \code{spgautor()} for areal (i.e., lattice),
+data. \code{spglm()} and \code{spgautor()} closely resemble {glm} from
+base \proglang{R} but include arguments that control the spatial
+correlation structure. \pkg{spmodel} has many helper functions for model
+inspection and diagnostics, some of which leverage other \proglang{R}
+packages like {broom} and {emmeans}. \pkg{spmodel} has tools to make
+predictions of the latent spatial-mean process at unobserved locations.
+\pkg{spmodel} also provides many advanced features like accommodating
+geometric anisotropy and nonspatial random effects, simulating spatially
+autocorrelated data, and more. Here we use \pkg{spmodel} to illustrate
+the modeling of binary, count, skewed and proportion response variables
+from several point-referenced and areal data sets.
 }
@@
 \Keywords{autoregressive model, geostatistical model, spatial
-autocovariance, spatial autocorrelation}
+covariance, spatial correlation}
 \Plainkeywords{autoregressive model, geostatistical model, spatial
-autocovariance, spatial autocorrelation}
+covariance, spatial correlation}
 
@@
 
-Binary, count, proportion, and skewed data are ubiquitous in practice.
-These data types are naturally modeled using a generalized linear model
-(GLM) framework
+In practice, non-Gaussian data are ubiquitous. Non-Gaussian that belong
+to an exponential family data can be naturally modeled using a
+generalized linear model (GLM) regression framework
 \citep{nelder1972generalized, mccullagh1989generalized, myers2012generalized, faraway2016extending}.
-In a GLM, a response variable \(\text{y}\) belongs to a particular
-statistical distribution with mean \(\mu\). For example, \(\text{y}\)
-may be distributed as a Poisson random variable with some mean \(\mu\).
-Based on the response distribution assumed for \(\text{y}\), GLMs link a
-function of \(\mu\) to explanatory variables via a link function:
-\begin{equation}\label{eq-glm}
-f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
-\end{equation} where for a sample size \(n\), \(\boldsymbol{\mu}\) is
-the mean of an \(n \times 1\) response vector \(\mathbf{y}\),
-\(f(\boldsymbol{\mu})\) is the link function that connects
-\(\boldsymbol{\mu}\) to \(\mathbf{w}\), \(\mathbf{X}\) is the
-\(n \times p\) design matrix of explanatory variables, and
-\(\boldsymbol{\beta}\) is the \(p \times 1\) vector of fixed effects.
-The mean, \(\boldsymbol{\mu}\), is usually constrained in some way
-(e.g., positive) that depends on the distribution assumed for
-\(\mathbf{y}\), but \(\mathbf{w}\) is unconstrained. The parameters in
-Equation\(~\)\eqref{eq-glm} are typically estimated via maximum
-likelihood (e.g., iteratively reweighted least squares)
-\citep{chambers1992S}. The {glm} function is commonly used to fit GLMs
-in the {R} programming language \citep{rcore2024}.
+In a GLM, an \(n \times 1\) response variable \(\mathbf{y}\) belongs to
+a a statistical distribution (e.g., Poisson, Binomial) with some mean
+and variance. Often, the analysis goal is to study the impact of a
+linear function of several explanatory variables on \(\text{y}\) through
+a GLM. In this context, the latent (i.e., unobserved) mean of
+\(\mathbf{y}\), \(\boldsymbol{\mu}\), is linked to these explanatory
+variables via a link function: \begin{equation}\label{eq-glm}
+f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
+\end{equation} where for a sample size \(n\), \(f(\cdot)\) is a link
+function that connects \(\boldsymbol{\mu}\) to \(\mathbf{w}\),
+\(\mathbf{X}\) is the \(n \times p\) design matrix of explanatory
+variables, and \(\boldsymbol{\beta}\) is the \(p \times 1\) vector of
+fixed effects. While the mean is typically constrained in some way
+(e.g., between zero and one if a probability), the link function
+generally makes \(\mathbf{w}\) is unconstrained. Common link functions
+inlude the log odds (i.e., logit) link for binary and proprtion data and
+the log link count and skewed data. Equation\(~\)\ref{eq-glm} can also
+be written in terms of the inverse link function, \(f^{-1}(\cdot)\):
+\begin{equation}\label{eq-glm2}
+\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}),
+\end{equation}
 
-The GLM framework in Equation\(~\)\eqref{eq-glm} assumes the elements of
-\(\mathbf{y}\) are independent of one another. This assumption is
-impractical for spatial data, where nearby observations tend to be more
-similar than distant observations \citep{tobler1970computer}. Ignoring
-this spatial dependence can give rise to misleading inference and poor
-prediction \citep{zimmerman2024spatial}. Spatial GLMs (SPGLMs) formally
-incorporate spatial autocorrelation into a GLM by adding to
-Equation\(~\)\eqref{eq-glm} two random effects that elucidate spatial
-structure: \begin{equation}\label{eq-spglm}
-f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon},
-\end{equation} where \(\boldsymbol{\tau}\) is an \(n \times 1\) column
-vector of spatially dependent random errors, and
+The GLM fixed effects (\(\boldsymbol{\beta}\)) are typically estimated
+via maximum likelihood \citep{chambers1992S}. It is often convenient to
+compute the maximum likelihood estimates using the iteratively
+reweighted least squares (IRWLS) algorithm \citep{wood2017generalized},
+which is the approach used by the \code{glm()} function in the
+\proglang{R} programming language \citep{rcore2024}. GLMs add an
+additional layer of complexity compared to linear regression models, as
+the left-hand size of Equation\(~\)\ref{eq-glm} is a function of the
+mean of \(\mathbf{y}\) rather than \(\mathbf{y}\) itself (as in linear
+regression models).
+
+The standard GLM assumes the elements of \(\mathbf{y}\) are independent.
+This independence assumption is typically impractical for spatial data.
+In spatial data, nearby observations tend to be more similar than
+distant observations \citep{tobler1970computer}, leading to positive
+spatial covariance among observations. The consequences of ignoring
+spatial covariance in statistical models for spatial data can be severe
+and include imprecise parameter estimates as well as misleading standard
+errors that inflate Type-I error rates and decrease power
+\citep{zimmerman2024spatial}.
+
+An approach for handling spatial data using a GLM is to assume
+\(\mathbf{w}\) has spatial covariance. This is achieved by adding to
+Equation\(~\)\ref{eq-glm} two random effects, \(\boldsymbol{\tau}\) and
+\(\boldsymbol{\epsilon}\). The random effect \(\boldsymbol{\tau}\) is an
+\(n \times 1\) column vector of spatially dependent random errors. We
+assume that \(\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}\) and
+\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}\), where
+\(\text{E}(\cdot)\) and \(\text{Cov}(\cdot)\) denote expectation and
+covariance, respectively. The variance parameter \(\sigma^2_\tau\)
+controls the magnitude of spatial covariance and is often called a
+partial sill, while the matrix \(\mathbf{R}\) is an \(n \times n\)
+spatial correlation matrix that depends on a range parameter controls
+the distance-decay rate of the spatial correlation. One example of a
+spatial covariance matrix is the ``exponential'', which is given by
+\begin{equation}\label{eq-spcov-exp}
+  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de} \exp(-\mathbf{H}/\phi),
+\end{equation} where \(\mathbf{H}\) is a matrix of pairwise distances
+among the elements of \(\mathbf{y}\) and \(\phi\) is a range parameter.
+From Equation\(~\)\ref{eq-spcov-exp}, as the distance between two
+elements of \(\mathbf{y}\) increases, the spatial covariance decreases,
+which reflects intuition. Moreover, as the range parameter, \(\phi\),
+increases, the strength of spatial dependence increases
+(Figure\(~\)\ref{fig-range}). The random effect
 \(\boldsymbol{\epsilon}\) is an \(n \times 1\) column vector of
-spatially independent random errors. We make a few assumptions about
-\(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\): first, that
-\(\text{E}(\boldsymbol{\tau}) = \text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}\),
-where \(\text{E}(\cdot)\) denotes expectation; second, that
-\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}\), where
-\(\mathbf{R}\) is an \(n \times n\) matrix that determines the spatial
-dependence structure in \(\mathbf{w}\) (which is linked to
-\(\boldsymbol{\mu}\) and hence \(\mathbf{y}\)) and depends on a range
-parameter, \(\phi\); third, that
-\(\text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_\epsilon \mathbf{I}\),
-where \(\mathbf{I}\) is an \(n \times n\) identity matrix; and fourth,
-that \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\) are independent
-of one another. The parameter \(\sigma^2_{\tau}\) is called the
-spatially dependent random error variance or partial sill. The parameter
-\(\sigma^2_\epsilon\) is called the spatially independent random error
-variance or nugget. These two variance parameters are henceforth more
-intuitively written as \(\sigma^2_{de}\) and \(\sigma^2_{ie}\),
-respectively. The covariance of \(\mathbf{w}\) is denoted
-\(\boldsymbol{\Sigma}\) and is given by \begin{equation}\label{eq-spcov}
- \boldsymbol{\Sigma} = \sigma^2_{de}\mathbf{R} + \sigma^2_{ie} \mathbf{I}.
-\end{equation} The parameters \(\sigma^2_{de}\), \(\phi\), and
-\(\sigma^2_{ie}\) are elements of \(\boldsymbol{\theta}\), the
-covariance parameter vector.
+independent random errors. We assume that
+\(\text{E}(\boldsymbol{\epsilon}) = \boldsymbol{0}\) and
+\(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\epsilon \mathbf{I}\), where
+\(\mathbf{I}\) is an \(n \times n\) identity matrix. The variance
+parameter \(\sigma^2_\epsilon\) controls the magnitude of nonspatial
+variability (i.e., fine-scale variation) and is often called a nugget.
 
-The {spmodel} {R} package provides tools for fitting spatial statistical
-models and making predictions at unobserved locations
-\citep{dumelle2023spmodel}. A recent major update (v0.4.0) to {spmodel}
-added SPGLM (Equation\(~\)\ref{eq-spglm}) support for binary, count,
-skewed, and proportion response variables (Table\(~\)\ref{table-links}),
-greatly expanding the class of models that {spmodel} makes accessible to
-practitioners.
+\begin{figure}
+\centering
+\includegraphics[width = 0.7\linewidth]{figures/figure-01.png}
+\caption{An exponential spatial correlation function with varying range parameters.}
+\label{fig-range}
+\end{figure}
 
-\begin{table}
+Through inclusion of \(\boldsymbol{\tau}\) and
+\(\boldsymbol{\epsilon}\), the spatial GLM (SPGLM) can be written as
+\begin{equation}\label{eq-spglm}
+f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
+\end{equation} Often in spatial statistics, quantities are explicitly
+referenced with respect to \(\mathbf{s}\), a vector of coordinates
+indexing the observation \citep{cressie1993statistics}. For example,
+\(\mathbf{y}\) and \(\mathbf{X}\) may instead be written
+\(\mathbf{y}(\mathbf{s})\) and \(\mathbf{X}\), respectively. We
+acknowledge the utility of this nomenclature but drop the explicit
+dependence on \(\mathbf{s}\) for simplicity of notation. Assuming
+independence among \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\),
+it follows that \begin{equation}\label{eq-spcov}
+ \text{Cov}(\boldsymbol{\tau} + \boldsymbol{\epsilon}) = \text{Cov}(\boldsymbol{\tau}) + \text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{\tau}\mathbf{R} + \sigma^2_{\epsilon} \mathbf{I}.
+\end{equation} To better align with intuition, we henceforth
+\(\sigma^2_{\tau}\) as \(\sigma^2_{de}\) (for spatial error variance)
+and \(\sigma^2_{\epsilon}\) as \(\sigma^2_{ie}\) (for independent error
+variance). The parameters \(\sigma^2_{de}\), \(\sigma^2_{ie}\), the
+range parameter \(\phi\) in \(\mathbf{R}\), and any other parameters in
+\(\mathbf{R}\) compose \(\boldsymbol{\theta}\), the covariance parameter
+vector.
+
+Fitting and using SPGLMs is challenging both conceptually and
+computationally \citep{bolker2009generalized}. Recently, however, there
+have been numerous, significant advances in \proglang{R} software that
+have made these models more accessible to practitioners. The \pkg{brms}
+\citep{burkner2017brms}, \pkg{carBayes} \citep{lee2013carbayes},
+\pkg{ngspatial} \citep{hughes2020ngspatial}, \pkg{R-INLA}
+\citep{lindgren2015bayesian} and \pkg{inlabru} \citep{bachl2019inlabru},
+\pkg{spBayes} \citep{finley2007spbayes}, \pkg{spOccupancy}
+\citep{doser2022spoccupancy}, \pkg{spAbundance}
+\citep{doser2024spabundance}, and \pkg{spNNGP} \citep{finley2002spnngp}
+packages take a Bayesian approach, either directly sampling from
+posterior distributions of parameters (e.g., using MCMC) or
+approximating them. A benefit of Bayesian approaches is that prior
+information can be incorporated and uncertainty quantification of
+parameter estimates is straightforward. However, Bayesian approaches,
+especially those using MCMC, can be computationally expensive. In order
+to reduce computation time, many of these packages work with the
+precision matrix instead of the covariance matrix so that
+computationally expensive matrix inversion is not required. For example,
+\pkg{R-INLA} uses the precision matrix and tends to be very fast.
+Working with precision matrices, however, can be more restrictive and
+less intuitive than working directly with the covariance matrix. The
+{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
+\citep{brooks2017glmmtmb}, \pkg{hglm} \citep{ronnegard2010hglm},
+\pkg{mgcv} \citep{wood2017generalized}, and \pkg{spaMM}
+\citep{rousset2014spamm} packages directly use Laplace,
+quasi-likelihood, or reduced-rank approaches to estimate parameters.
+These direct approaches tend to be computationally efficient, as they
+don't rely on MCMC sampling. In contrast to the Bayesian approach, a
+drawback of these direct approaches is that prior information cannot be
+formally incorporated and covariance parameter uncertainty is more
+challenging to quantify. The \pkg{sdmTMB} \citep{anderson2024sdmtmb}
+package combines elements of \pkg{R-INLA}, \pkg{glmmTMB}, and properties
+of Gaussian Markov random fields to fit a wide variety of SPGLMs, and
+\pkg{tinyVAST} \citep{thorson2025tinyVAST} extends some of these models
+to multivariate or (dynamic) structural equation models.
+
+\citet{ver2024marginal} proposed a novel approach to fitting SPGLMs that
+leverages the Laplace approximation while marginalizing over both the
+latent \(\mathbf{w}\) and the fixed effects (\(\boldsymbol{\beta}\)) and
+accommodating spatial covariance. \citet{ver2024marginal} showed that
+this approach performed efficiently in a variety of simulation settings,
+generally having appropriate confidence interval coverage for the fixed
+effects and prediction interval coverage for new \(\mathbf{w}\). The
+approach performed similarly to the Bayesian SPGLM approach in
+\pkg{spBayes} and the automatic differentiation SPGLM approach in
+\pkg{glmmTMB} but was much faster. At small sample sizes, the approach
+outperformed the approximate Bayesian SPGLM approach in \pkg{R-INLA} and
+had similar computational times. For moderate sample sizes, it performed
+similarly to \pkg{R-INLA}, though \pkg{R-INLA} was faster. This novel
+approach is particularly attractive for two reasons. First, it is
+general enough that can be applied to any covariance structure (not just
+spatial). Second, after estimating the covariance parameters, analytical
+solutions exist for the fixed effects (and their standard errors) as
+well as predictions of the latent \(\mathbf{w}\) at new locations (and
+their standard errors). The \pkg{spmodel} \proglang{R} package
+\citep{dumelle2023spmodel} recently provided full support for the
+methods in \citet{ver2024marginal} applied to binary, count, skewed, and
+proportion data for over 20 different spatial covariance types.
+
+The \pkg{spmodel} \proglang{R} package \citep{dumelle2023spmodel}
+recently provided a full set of modeling tools for SPGLMs fit using the
+methods described in \citet{ver2024marginal}. These modeling tools are
+approachable and mirror the familiar \code{glm()} syntax from
+base-\proglang{R}, making the transition from GLMs to SPGLMs relatively
+seamless. The \code{spglm()} function fits SPGLMs for point-referenced
+data (e.g., x-coordinates and y-coordinates representing point locations
+in a field), while the \code{spgautor()} function fits SPGLMs for areal
+data (e.g., polygon boundaries representing geographic subsets of a
+region). \pkg{spmodel} supports the binomial distribution for binary
+data, Poisson and negative binomial distributions for count data, Gamma
+and inverse Gaussian distributions for skewed data, and the beta
+distribution for proportion data. There are 20 different spatial
+covariance structures available including the exponential, Gaussian, and
+spherical for point-referenced data (Figure\(~\)\ref{fig-type}) and the
+conditional autoregressive, and simultaneous autoregressive structures
+for areal data. \pkg{spmodel} provides tools for commonly used model
+summaries, visualizations, and diagnostics (e.g., Cook's distance) using
+standard \proglang{R} helper functions like \code{summary()},
+\code{plot()}, and \code{cooks.distance()}. \pkg{spmodel} also provides
+tools to predict \(\mathbf{w}\) at new locations and quantify
+uncertainty in those prediction using \code{predict()}. This core
+functionality, combined with several advanced features we describe
+throughout the manuscript, enable \pkg{spmodel} to provide some novel
+and important capabilities previously missing from the existing SPGLM
+ecosystem in \proglang{R}.
+
+\begin{figure}
 \centering
-\begin{tabular}{||c|ccc||} 
- \hline
- Family & Link Function & Link Name & Data Type  \\ [0.5ex] 
- \hline \hline
- Binomial & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Binary; Binary Count \\
- Poisson & $f(\mu) = \log(\mu)$ & Log & Count \\
- Negative Binomial & $f(\mu) = \log(\mu)$ & Log & Count \\
- Beta & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Proportion \\
- Gamma & $f(\mu) = \log(\mu)$ & Log & Skewed \\
- Inverse Gaussian & $f(\mu) = \log(\mu)$ & Log & Skewed \\ 
- \hline\hline
-\end{tabular}
-\caption{SPGLM response distributions and their link functions and data types.}
-\label{table-links}
-\end{table}
+\includegraphics[width = 0.7\linewidth]{figures/figure-02.png}
+\caption{Exponential, Gaussian, and spherical spatial correlation functions all with range parameters equal to 0.5.}
+\label{fig-type}
+\end{figure}
 
-SPGLMs for point-referenced data are called geostatistical GLMs and are
-fit using {spglm}. Data are point-referenced when the elements in
-\(\mathbf{y}\) are observed at point-locations indexed by x-coordinates
-and y-coordinates on a spatially continuous surface with an infinite
-number of locations (e.g., point locations in a field). SPGLMs for areal
-data are called spatial autoregresive models and are fit using
-{spgautor}. Data are areal (i.e., lattice) when they are part of a
-finite network of polygons whose connections are indexed by a
-neighborhood structure (e.g., states in a country).
-
-Several other {R} packages exist for analyzing SPGLMs. The {brms}
-\citep{burkner2017brms}, {carBayes} \citep{lee2013carbayes}, {ngspatial}
-\citep{hughes2020ngspatial}, {R-INLA} \citep{lindgren2015bayesian},
-{spBayes} \citep{finley2007spbayes}, and {spNNGP}
-\citep{finley2002spnngp} packages take a Bayesian approach, either
-directly sampling from posterior distributions of parameters (e.g.,
-using MCMC) or approximating them. A benefit of Bayesian approaches is
-that prior information can be incorporated and uncertainty
-quantification of parameter estimates is straightforward. However,
-Bayesian approaches, especially those using MCMC, tend to be
-computationally expensive. In order to reduce computation time, many of
-these packages work with the precision matrix instead of the covariance
-matrix so that inverting matrices (a computational bottleneck) is not
-required. For example, {R-INLA} uses the precision matrix (often
-directly modeled as with autoregressive models) and tends to be very
-fast. Working with precision matrices, however, can be more restrictive
-and less intuitive than working directly with the covariance matrix. The
-{FRK} \citep{sainsbury2024modeling}, {glmmTMB}
-\citep{brooks2017glmmtmb}, {hglm} \citep{ronnegard2010hglm}, {mgcv}
-\citep{wood2017generalized}, and {spaMM} \citep{rousset2014spamm}
-packages directly use Laplace, quasi-likelihood, or reduced-rank
-approaches to estimate parameters. These direct approaches tend to be
-computationally efficient, as they don't rely on MCMC sampling. In
-contrast to the Bayesian approach, a drawback of these direct approaches
-is that prior information cannot be formally incorporated and covariance
-parameter uncertainty is challenging to quantify. SPGLMs in {spmodel}
-are most similar to those in {glmmTMB} -- {spmodel} uses analytical
-solutions for maximizing the likelihood (which we describe in the next
-section), while {glmmTMB} uses automatic differentiation for maximizing
-the likelihood (which tends to be much slower than analytical
-solutions).
-
-Missing from the aforementioned {R} packages is the complete set of
-tools for SPGLMs that {spmodel} provides. Importantly, the {spglm} and
-{spautor} functions act as a spatial analogue to the familiar {glm} from
-base {R}, making the transition from GLMs to SPGLMs relatively seamless.
-{spmodel} leverages many commonly used {R} generics like {summary} to
-better understand fitted models. Six GLM families
-(Table\(~\)\ref{table-links}) and 20 spatial covariance (or precision)
-functions are supported. Also available are functions for data
-visualization, model fitting, model summaries, model diagnostics, model
-comparison, and prediction, all crucial components of a data analysis.
-Through extra function arguments, {spmodel} supports advanced features
-like geometric anisotropy, non-spatial random effects, methods for large
-data sets, and more. Importantly, {spmodel} extends popular packages
-like {broom} \citep{robinson2021broom, kuhn2022tidy} and {emmeans}
-\citep{lenth2024emmeans}. {spmodel} also provides function for
-simulating SPGLM data (e.g., {sprbinom}, {sprpois}).
+\pkg{spmodel} (version 0.11.0) is arguably most similar to \pkg{sdmTMB}
+(version 0.7.4) in terms of scope and feel. Both packages use similar
+syntax as \code{glm()}, accommodate flexible \code{formula} arguments
+(e.g., offsets, splines), handle spatial covariance that decays at
+different rates in different rates (i.e., geometric anisotropy),
+incorporate nonspatial random effects, support other \proglang{R}
+packages for modeling like \pkg{broom}
+\citep{robinson2021broom, kuhn2022tidy} and \pkg{emmeans}
+\citep{lenth2024emmeans}, and have tools for model summaries,
+prediction, and simulating data. There are some notable differences
+between the two packages, however. \pkg{sdmTMB} supports several
+additional GLM distributions like the Tweedie, supports Hurdle models,
+and can incorporate prior information through Bayesian applications.
+\pkg{sdmTMB} also provides tools for working with temporal data and
+enhanced visualizations of marginal effects. \pkg{sdmTMB} does require a
+preprocessing step of constructing a mesh for the stochastic partial
+differential equation approach, and the density of the mesh can affect
+model results and computational complexity. \pkg{spmodel} does not
+require the construction of a mesh prior to modeling. \pkg{spmodel}
+supports 20 different spatial covariances and models them directly,
+rather than using a precision matrix approximation to the Matérn spatial
+covariance as in \pkg{sdmTMB}. \pkg{spmodel} also provides experimental
+design tools (e.g., analysis of variance, contrasts), supports \pkg{sf}
+objects in modeling and prediction functions \citep{pebesma2018sf}, has
+several specialized model diagnostics like leverage values and Cook's
+distances, and has analytic solutions for prediction standard errors.
+Other similarities and differences do exist between \pkg{sdmTMB} and
+\pkg{spmodel}, and both packages continue to evolve. Overall, we believe
+that these packages are complementary and enhance the suite of SPGLM
+tools accessible to practitioners.
 
@@
 Section\(~\)\ref{sec-spglm}, we provide some background for the SPGLM
-fitting and prediction routines in {spmodel}. In
+fitting and prediction routines in \pkg{spmodel}. In
 Section\(~\)\ref{sec-applications}, we provide several applications of
-{spmodel} to spatial binary, count, skewed and proportion data with both
-point-referenced and areal supports. And in
+\pkg{spmodel} to spatial binary, count, skewed and proportion data with
+both point-referenced and areal supports. And in
 Section\(~\)\ref{sec-discussion}, we end with a discussion synthesizing
-\pkg{spmodel}'s contributions to the analysis of SPGLMs in {R}.
+\pkg{spmodel}'s contributions to the analysis of SPGLMs in \proglang{R}.
 
-\section{Spatial generalized linear models using the Laplace
-approximation}\label{sec-spglm}
+\section{The spatial generalized linear model and
+marginalizatoin}\label{sec-spglm}
 
-GLMs with random effects are often written hierarchically
-\citep{lee1996hierarchical, bolker2009generalized, wood2017generalized}.
-{spmodel} leverages this hierarchical structure and uses a novel
-application of the Laplace approximation \citep{ver2024marginal} to fit
-models. This marginal approach is quite flexible, accommodating a wide
-range of possible dependence structures and formally maximizing a
-likelihood. Maximizing a likelihood yields convenient likelihood-based
-statistics like AIC \citep{akaike1974new}, AICc
-\citep{hoeting2006model}, BIC \citep{schwarz1978estimating}, deviance
-\citep{mccullagh1989generalized}, and likelihood ratio tests for model
-comparison, a benefit compared to quasi-likelihood
+\pkg{spmodel} implements the novel methods described in
+\citet{ver2024marginal} to fit SPGLMs, which leverages the Laplace
+approximation and marginalizes over both the latent \(\mathbf{w}\) and
+the fixed effects while accommodating spatial covariance. A beneficial
+aspect of this approach is that it formally maximizes a hierarchical GLM
+likelihood \citep{lee1996hierarchical, wood2017generalized}. This makes
+likelihood-based statistics for model comparison like AIC
+\citep{akaike1974new}, AICc \citep{hoeting2006model}, BIC
+\citep{schwarz1978estimating}, deviance
+\citep{mccullagh1989generalized}, and likelihood ratio tests available.
+These types of statistics are not available for quasi-likelihood
 \citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
 approaches \citep{wolfinger1993generalized}, which only specify the
-first two moments of a distribution. \citet{ver2024marginal} provide
-thorough context for the marginal approach and its associated details,
-but next we provide an short overview of the methodology.
+first two moments of a distribution. \citet{ver2024marginal} provides
+thorough details regarding the method and contextualizes its development
+which built upon similar methods
+\citep[\citet{bonat2016practical}]{evangelou2011estimation}. Next, we
+describe a brief overview of the approach and how it can be used for
+parameter estimation, inference, and prediction.
 
-Our goal is to marginalize over the latent mean \(\mathbf{w}\) and fixed
-effects \(\boldsymbol{\beta}\) in Equation\(~\)\eqref{eq-spglm} to
-obtain a response distribution for \(\mathbf{y}\) that depends only on
-the explanatory variables, \(\mathbf{X}\), the response distribution's
-dispersion parameter, \(\varphi\), and the covariance parameters,
-\(\boldsymbol{\theta}\). We can represent this marginal distribution
-hierarchically as \begin{equation}\label{eq-marginal}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w} .
-\end{equation} In Equation\(~\)\eqref{eq-marginal},
-\([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) is the density for
-\(\mathbf{y}\) (e.g., binomial, Poisson) given the latent mean and
-dispersion parameter and
-\([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\) is the Gaussian
-density for \(\mathbf{w}\) given the explanatory variables, fixed
-effects, and covariance parameters. Integrating \(\boldsymbol{\beta}\)
-out of Equation\(~\)\eqref{eq-spglm} yields
+\subsection{Formulating the hierarchical
+likelihood}\label{formulating-the-hierarchical-likelihood}
+
+We can write the SPGLM likelihood hierarchically as
+\begin{equation}\label{eq-marginal}
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w},
+\end{equation} where \([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) is
+the density for the appropriate response distribution of \(\mathbf{y}\)
+(e.g., binomial, Poisson) given the latent \(\mathbf{w}\) and dispersion
+parameter (\(\varphi\)), and
+\([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\) is the multivariate
+Gaussian density for \(\mathbf{w}\) given the explanatory variables
+(\(\mathbf{X}\)), fixed effects (\(\boldsymbol{\beta}\)), and spatial
+covariance parameters (\(\boldsymbol{\theta}\)). Following
+@\citet{harville1977maximum}, we can integrate \(\boldsymbol{\beta}\)
+out of Equation\(~\)\ref{eq-spglm}, which yields
 \begin{equation}\label{eq-marginal2}
@@
 \end{equation} where \([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\)
-is the restricted (i.e., residual) Gaussian density
-\citep[\citet{harville1977maximum},
-\citet{wolfinger1994computing}]{patterson1971recovery} for
-\(\mathbf{w}\) given the explanatory variables and covariance
-parameters. This restricted Gaussian density is given by
-\begin{equation}\label{eq-reml-def}
+is the restricted (i.e., residual) multivariate Gaussian density
+\citep{patterson1971recovery} for \(\mathbf{w}\) given the explanatory
+variables and covariance parameters. Equation\(~\)\ref{eq-marginal2} can
+be synonymous written after profiling the overall variance out of
+\(\boldsymbol{\Sigma}\), which reduces the dimension of
+\(\boldsymbol{\theta}\) by one for optimization
+\citep{wolfinger1994computing}. The restricted multivariate Gaussian
+density is given by \begin{equation}\label{eq-reml-def}
 [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] = \frac{\exp(-\frac{1}{2}(\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}}) \boldsymbol{\Sigma}^{-1} (\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}})^T)}{(2 \pi)^{(n - p)/2} |\boldsymbol{\Sigma}|^{1/2}|\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X}|^{1/2}},
@@
 \(\tilde{\boldsymbol{\beta}} = (\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{w}\)
-and \(|\cdot|\) denotes the determinant.
-
-Next, let
-\(\ell_\mathbf{w} = \log([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}])\)
-and then notice that Equation\(~\)\eqref{eq-marginal2} can be written as
+and \(|\cdot|\) denotes the determinant. Next, let
+\begin{equation}\label{eq-marginal03}
+  \ell_\mathbf{w} = \log([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}])
+\end{equation} and rewrite Equation\(~\)\ref{eq-marginal2} as
 \begin{equation}\label{eq-marginal3}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \exp(\ell_\mathbf{w}) d\mathbf{w}.
-\end{equation} A Taylor series expansion of \(\ell_\mathbf{w}\) around a
-point \(\mathbf{a}\) yields \begin{equation}\label{eq-marginal4}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_\mathbf{a} + \mathbf{g}^T(\mathbf{w} - \mathbf{a}) + \frac{1}{2}(\mathbf{w} - \mathbf{a})^T \mathbf{G} (\mathbf{w} - \mathbf{a}))d\mathbf{w},
+\end{equation} A second-order Taylor series expansion of
+\(\ell_\mathbf{w}\) around \(\hat{\mathbf{w}}\) yields
+\begin{equation}\label{eq-marginal4}
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\hat{\mathbf{w}}} + \mathbf{g}^T(\mathbf{w} - \hat{\mathbf{w}}) + \frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T \mathbf{G} (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w},
 \end{equation} where \(\mathbf{g}\) and \(\mathbf{G}\) are the gradient
 and Hessian, respectively, of \(\ell_\mathbf{w}\) with respect to
-\(\mathbf{w}\). If \(\mathbf{a}\) is a value for which
+\(\mathbf{w}\). If \(\hat{\mathbf{w}}\) is a value for which
 \(\mathbf{g} = \mathbf{0}\), \begin{equation}\label{eq-marginal5}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_\mathbf{a}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \mathbf{a})^T (-\mathbf{G}) (\mathbf{w} - \mathbf{a}))d\mathbf{w}.
-\end{equation} Notice that the integral in
-Equation\(~\)\ref{eq-marginal5} can be solved in the same manner as the
-normalizing constant in a multivariate Gaussian distribution, and
-rewriting \(\exp(\ell_\mathbf{a})\) yields
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\hat{\mathbf{w}}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T (-\mathbf{G}) (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w}.
+\end{equation} The integral in Equation\(~\)\ref{eq-marginal5} can be
+solved by leveraging properties of the normalizing constant of a
+multivariate Gaussian distribution. Thus, rewriting
+\(\exp(\ell_{\hat{\mathbf{w}}})\) yields
 \begin{equation}\label{eq-marginal6}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{a}), \varphi] [\mathbf{a} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{a}}|^{-1/2}.
-\end{equation} \citet{ver2024marginal} show how to evaluate
-\(\mathbf{g}\) and \(\mathbf{G}\) to obtain \textbf{a} and maximize (the
-natural logarithm of) Equation\(~\)\eqref{eq-marginal6} for the six
-response distributions in Table\(~\)\ref{table-links}. Maximizing
-Equation\(~\)\eqref{eq-marginal6} using a Newton-Rhapson approach yields
-the marginal restricted maximum likelihood estimators \(\hat{\varphi}\)
-and \(\hat{\boldsymbol{\theta}}\). \citet{ver2024marginal} show that the
-fixed effect estimator is
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\hat{\mathbf{w}}), \varphi] [\hat{\mathbf{w}} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\hat{\mathbf{w}}}|^{-1/2}.
+\end{equation}
+
+Maximizing the natural logarithm of Equation\(~\)\ref{eq-marginal6}
+requires a doubly iterative process over \(\boldsymbol{\theta}\) and
+\(\varphi\) as well as \(\mathbf{w}\), eventually yielding the the
+marginal restricted maximum likelihood estimators \(\hat{\varphi}\) and
+\(\hat{\boldsymbol{\theta}}\) and their corresponding values of
+\(\hat{\mathbf{w}}\). Maximizing this log likelihood is a
+computationally expensive operation that involves repeatedly evaluating
+\(\boldsymbol{\Sigma}^{-1}\), \(\mathbf{g}\), and \(\mathbf{G}\); see
+\citet{ver2024marginal} for more details and forms of \(\mathbf{g}\) and
+\(\mathbf{G}\) for various response distributions.
+
+\subsection{Estimating fixed effects}\label{estimating-fixed-effects}
+
+\citet{ver2024marginal} show that the fixed effect estimator is
 \(\hat{\boldsymbol{\beta}} = (\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{X})^{-1}\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{w}\),
@@
 \(\boldsymbol{\Sigma}\) evaluated at \(\hat{\boldsymbol{\theta}}\).
+
+\subsection{Predicting at new
+locations}\label{predicting-at-new-locations}
+
 \citet{ver2024marginal} also derive the covariance matrix of
@@
 
-\section{Application}\label{sec-applications}
+\section{Modeling moose presence in Alaska, USA}\label{sec-applications}
 
-The {spglm} (for point-referenced data) and {spgautor} (for areal data)
-functions in {spmodel} fit SPGLMs using the Laplace approximation
-outlined in Section\(~\)\ref{sec-spglm}. Both {spglm} and {spgautor}
-generally require the following four arguments: \texttt{formula}, a
-formula that describes the relationship between the response variable
-and explanatory variables; \texttt{family}, the response distribution
-(which can be \texttt{binomial}, \texttt{poisson}, \texttt{nbinomial},
-\texttt{Gamma}, \texttt{inverse.gaussian}, or \texttt{beta});
-\texttt{data}, the data frame that holds the variables in
-\texttt{formula} as well as spatial locations; and \texttt{spcov\_type},
-the spatial covariance type. The first three arguments are shared by
-{glm}; thus, the transition from GLMs to SPGLMs requires only one
-additional argument: \texttt{spcov\_type}. The {spglm} spatial
-covariance types measure dependence as a function of Euclidean distance
-among observations; an example is the exponential spatial covariance:
-\begin{equation}\label{eq-spcov-exp}
-  \boldsymbol{\Sigma} = \sigma^2_{de} \exp(-\mathbf{H}/\phi) + \sigma^2_{ie}\mathbf{I},
-\end{equation} where \(\mathbf{H}\) is a matrix of pairwise distances
-among all observations. {spglm} currently supports 18 distinct spatial
-covariance functions.
+\subsection{Additional applications}\label{sec-applications2}
 
-The {spgautor} spatial covariance types measure dependence as a function
-of neighborhood distance among observations; an example is the
-simultaneous autoregressive covariance matrix:
-\begin{equation}\label{eq-spcov-sar}
-  \boldsymbol{\Sigma} = \sigma^2_{de} [(\mathbf{I} - \phi \mathbf{W})(\mathbf{I} - \phi \mathbf{W})^T]^{-1} + \sigma^2_{ie}\mathbf{I},
-\end{equation} where \(\mathbf{W}\) is a matrix that represents the
-neighborhood structure among all observations. {spgautor} currently
-supports two distinct spatial covariance functions.
+\subsection{Modeling moose counts in Alaska,
+USA}\label{modeling-moose-counts-in-alaska-usa}
 
-In the rest of this section, we use {spmodel} to study binary, count,
-skewed, and proportion response variables that are either
-point-referenced or areal. We use {spmodel} for all parts of the data
-analysis, from estimation to inference to model diagnostics to
-prediction. We first describe core {spmodel} functionality in an
-application to binary data, while additional analyses highlight count,
-skewed, and proportion data as well as some additional {spmodel}
-features. Before proceeding, load {spmodel} into the current {R}
-session:
+\subsection{Modeling harbor seal trends in Alaska,
+USA}\label{modeling-harbor-seal-trends-in-alaska-usa}
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("spmodel")
-\end{CodeInput}
-\end{CodeChunk}
+\subsection{Modeling voter turnout in Texas,
+USA}\label{modeling-voter-turnout-in-texas-usa}
 
-\subsection{Binary data}\label{binary-data}
-
-The \texttt{moose} data in {spmodel} contain information on moose
-presence in Alaska. They are an \texttt{sf} object, a special data frame
-that is supplemented with spatial information using the {sf} package
-\citep{pebesma2018sf}. The first few rows look like:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> head(moose)
-\end{CodeInput}
-\begin{CodeOutput}
-Simple feature collection with 6 features and 4 fields
-Geometry type: POINT
-Dimension:     XY
-Bounding box:  xmin: 281896.4 ymin: 1518398 xmax: 311325.3 ymax: 1541016
-Projected CRS: NAD83 / Alaska Albers
-# A tibble: 6 x 5
-   elev strat count presence           geometry
-  <dbl> <chr> <dbl> <fct>           <POINT [m]>
-1  469. L         0 0        (293542.6 1541016)
-2  362. L         0 0        (298313.1 1533972)
-3  173. M         0 0        (281896.4 1532516)
-4  280. L         0 0        (298651.3 1530264)
-5  620. L         0 0        (311325.3 1527705)
-6  164. M         0 0        (291421.5 1518398)
-\end{CodeOutput}
-\end{CodeChunk}
-
-There are five columns: \texttt{elev}, the numeric site elevation
-(meters); \texttt{strat} a stratification variable for sampling with two
-levels, \texttt{"L"} and \texttt{"M"}, which are categorized by
-landscape metrics at each site; \texttt{count}, the number of moose at
-each site; \texttt{presence}, a factor that indicates whether at least
-one moose was observed at each site (\texttt{0} implies no moose;
-\texttt{1} implies at least one moose); and \texttt{geometry}, the NAD83
-projected coordinate of each site. The \texttt{moose\_preds} data in
-{spmodel} contain spatial locations at which predictions of moose
-presence are desired. They are also an \texttt{sf} object with the same
-projection and measurements for \texttt{elev} and \texttt{strat}.
-Figure\(~\)\ref{fig-moose-data} shows the \texttt{presence} variable in
-\texttt{moose} as well as the spatial locations of both \texttt{moose}
-and \texttt{moose\_preds}. Moose are most common in the southwestern and
-eastern parts of the domain and least common in the northwest.
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-1.png}
-\caption{Moose presence in Alaska. Circles represent moose presence or absence (based on color) and triangles represent locations at which moose presence probability predictions are desired.}
-\label{fig-moose-data}
-\end{figure}
-
-To study the effect of elevation, stratum, and their interaction on
-moose presence while accounting for spatial autocorrelation, we fit a
-SPGLM for binary data (i.e., a spatial logistic regression model) using
-{spglm}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> spbin <- spglm(
-+   formula = presence ~ elev + strat + elev:strat,
-+   family = binomial,
-+   data = moose,
-+   spcov_type = "spherical"
-+ )
-\end{CodeInput}
-\end{CodeChunk}
-
-Summarizing the model object yields a summary similar to that provided
-by the familiar {glm}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> summary(spbin)
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-spglm(formula = presence ~ elev + strat + elev:strat, family = binomial, 
-    data = moose, spcov_type = "spherical")
-
-Deviance Residuals:
-    Min      1Q  Median      3Q     Max 
--1.8423 -0.7538  0.3883  0.7604  1.6018
-
-Coefficients (fixed):
-             Estimate Std. Error z value Pr(>|z|)   
-(Intercept) -3.039992   1.205695  -2.521  0.01169 * 
-elev         0.009133   0.004126   2.213  0.02687 * 
-stratM       3.276511   1.162603   2.818  0.00483 **
-elev:stratM -0.010882   0.006697  -1.625  0.10418   
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-
-Pseudo R-squared: 0.08845
-
-Coefficients (spherical spatial covariance):
-       de        ie     range 
-5.083e+00 2.580e-03 5.158e+04
-
-Coefficients (Dispersion for binomial family):
-dispersion 
-         1 
-\end{CodeOutput}
-\end{CodeChunk}
-
-The summary contains the original function call, a summary of residuals,
-the fixed effects coefficients table, the spatial covariance parameter
-estimates, and additional model information like the pseudo R-squared,
-which quantifies the variability in the model attributable to the fixed
-effects. While useful, this summary information is hard to work with, as
-it is printed directly to the {R} console. The {broom} package from the
-tidymodels \citep{kuhn2022tidy} ecosystem has functions to provide
-helpful model output in the form of tibbles (i.e., data frames) that are
-easily manipulated. {spmodel} has methods for the {tidy}, {glance}, and
-{augment} functions from {broom}. The first {broom} function is {tidy},
-which tidies the model output:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> tidy(spbin)
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 4 x 5
-  term        estimate std.error statistic p.value
-  <chr>          <dbl>     <dbl>     <dbl>   <dbl>
-1 (Intercept) -3.04      1.21        -2.52 0.0117 
-2 elev         0.00913   0.00413      2.21 0.0269 
-3 stratM       3.28      1.16         2.82 0.00483
-4 elev:stratM -0.0109    0.00670     -1.62 0.104  
-\end{CodeOutput}
-\end{CodeChunk}
-
-The estimates and standard errors returned are on the log odds link
-(Table\(~\)\ref{table-links}) scale ({coef} and {vcov} may also be
-used). The output provides evidence that elevation is positively
-associated with moose presence in the \texttt{"L"} stratum (\(p-\)value
-\(<0.05\)) and, at zero elevation, moose are more likely in the
-\texttt{"M"} stratum than the \texttt{"L"} stratum (\(p-\)value
-\(<0.01\)). This output provides marginal evidence that the effect of
-elevation on moose presence varies across strata (\(p-\)value
-\(\approx 0.1\)). The model effectively quantifies the impact of
-elevation on moose presence for moose in the \texttt{"L"} strata, but an
-analogous statement for moose in the \texttt{"M"} strata requires more
-context. We could refit the model treating \texttt{"M"} as the reference
-group instead of \texttt{"L"}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> moose$strat2 <- factor(moose$strat, levels = c("M", "L"))
-R> update(spbin, formula = presence ~ elev + strat2 + elev:strat2) |> 
-+   summary()
-\end{CodeInput}
-\begin{CodeOutput}
-
-Call:
-spglm(formula = presence ~ elev + strat2 + elev:strat2, family = binomial, 
-    data = moose, spcov_type = "spherical")
-
-Deviance Residuals:
-    Min      1Q  Median      3Q     Max 
--1.8423 -0.7538  0.3883  0.7604  1.6018
-
-Coefficients (fixed):
-              Estimate Std. Error z value Pr(>|z|)   
-(Intercept)   0.236519   1.305198   0.181  0.85620   
-elev         -0.001750   0.006090  -0.287  0.77385   
-strat2L      -3.276511   1.162603  -2.818  0.00483 **
-elev:strat2L  0.010882   0.006697   1.625  0.10418   
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-
-Pseudo R-squared: 0.08845
-
-Coefficients (spherical spatial covariance):
-       de        ie     range 
-5.083e+00 2.580e-03 5.158e+04
-
-Coefficients (Dispersion for binomial family):
-dispersion 
-         1 
-\end{CodeOutput}
-\end{CodeChunk}
-
-A simpler solution, especially if there categorical variables with many
-levels, is to leverage {emmeans}. {emmeans} is an {R} package for
-estimating marginal means of model objects. The {emtrends} function in
-{emmeans} characterizes the effect of a continuous variable (here,
-\texttt{elev}) for each level of a categorical variable (here,
-\texttt{strat}):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> library("emmeans")
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> emtrends(spbin, "strat", "elev")
-\end{CodeInput}
-\begin{CodeOutput}
- strat elev.trend      SE  df asymp.LCL asymp.UCL
- L        0.00913 0.00413 Inf   0.00105    0.0172
- M       -0.00175 0.00609 Inf  -0.01369    0.0102
-
-Degrees-of-freedom method: asymptotic 
-Confidence level used: 0.95 
-\end{CodeOutput}
-\end{CodeChunk}
-
-The asymptotic confidence intervals show that there is more evidence of
-an association between elevation and moose presence in the \texttt{"L"}
-stratum than in the \texttt{"M"} stratum. Notice that
-\texttt{elev.trend} for the \texttt{"L"} stratum matches the
-\texttt{elev} effect when \texttt{"L"} is the reference group, and
-similarly for \texttt{elev.trend} when \texttt{"M"} is the reference
-group.
-
-The second {broom} function is {glance}, which glances at the model fit:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> glance(spbin)
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 1 x 10
-      n     p  npar value   AIC  AICc   BIC logLik deviance pseudo.r.squared
-  <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>            <dbl>
-1   218     4     3  681.  687.  687.  697.  -340.     161.           0.0885
-\end{CodeOutput}
-\end{CodeChunk}
-
-{glance} returns several useful statistics like the sample size
-(\texttt{n}), number of fixed effects (\texttt{p}), number of covariance
-parameters (\texttt{npar}), several likelihood-based statistics (e.g.,
-\texttt{AIC}, \texttt{AICc}, \texttt{BIC}), and pseudo R-squared.
-
-\begin{figure}
-\centering
-\includegraphics[width = 1\linewidth]{figures/figure-2.png}
-\caption{Spatial logistic regression model diagnostics from [augment]{.fct}. The leverage (i.e., hat) values (left) and standardized residuals (right).}
-\label{fig-sp-diagnostic}
-\end{figure}
-
-The third {broom} function is {augment}, which augments the model data
-with diagnostics:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> head(augment(spbin))
-\end{CodeInput}
-\begin{CodeOutput}
-Simple feature collection with 6 features and 8 fields
-Geometry type: POINT
-Dimension:     XY
-Bounding box:  xmin: 281896.4 ymin: 1518398 xmax: 311325.3 ymax: 1541016
-Projected CRS: NAD83 / Alaska Albers
-# A tibble: 6 x 9
-  presence  elev strat .fitted .resid    .hat   .cooksd .std.resid
-  <fct>    <dbl> <chr>   <dbl>  <dbl>   <dbl>     <dbl>      <dbl>
-1 0         469. L      -1.47  -0.644 0.101   0.0130        -0.679
-2 0         362. L      -2.77  -0.349 0.0166  0.000523      -0.352
-3 0         173. M      -2.23  -0.451 0.00390 0.000200      -0.452
-4 0         280. L      -3.59  -0.234 0.00343 0.0000472     -0.234
-5 0         620. L      -0.774 -0.871 0.319   0.130         -1.06 
-6 0         164. M      -2.01  -0.502 0.00459 0.000292      -0.503
-# i 1 more variable: geometry <POINT [m]>
-\end{CodeOutput}
-\end{CodeChunk}
-
-The {augment} function returns the fitted \(\mathbf{w}\) values
-(\texttt{.fitted}), deviance residuals (\texttt{.resid}), leverage
-(i.e., hat) values (\texttt{.hat}), Cook's distance (\texttt{.cooksd}),
-and standardized residuals (\texttt{.std.resid}). When the data are an
-\texttt{sf} object, {augment} returns another \texttt{sf} object,
-helpful for visualizing model diagnostics spatially as in
-Figure\(~\)\ref{fig-sp-diagnostic}. Leverage measures the unusualness of
-an observation's set of explanatory variables, while Cook's distances
-measures how influential an observation is on the resulting model fit
-\citep{montgomery2021introduction}. Model diagnostics are also
-accessible as vectors using the appropriate generic function (e.g.,
-{fitted}, {residuals}).
-
-Similar to {glm} model objects, {plot} can be used to visualize
-diagnostics (Figure\(~\)\ref{fig-sp-diagnostic2}):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot(spbin, which = c(4, 7))
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{figure}
-\centering
-\includegraphics[width = 1\linewidth]{figures/figure-3.png}
-\caption{Spatial logistic regression model diagnostics from [plot]{.fct}. The Cook's distance values (left) and the fitted spatial covariance as a function of distance (right).}
-\label{fig-sp-diagnostic2}
-\end{figure}
-
-Components of model variation are partitioned using {varcomp}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> varcomp(spbin)
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 3 x 2
-  varcomp            proportion
-  <chr>                   <dbl>
-1 Covariates (PR-sq)   0.0885  
-2 de                   0.911   
-3 ie                   0.000462
-\end{CodeOutput}
-\end{CodeChunk}
-
-The fixed effects explain roughly 9\% of model variation, while the
-spatially dependent variance explains most of the remaining variability.
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-4.png}
-\caption{Moose presence probability fitted values and predictions. Fitted values are represeneted by circles and predictions by triangles.}
-\label{fig-moose-fit}
-\end{figure}
-
-We make predictions of the log odds of moose probability presence at
-each site in \texttt{moose\_preds} using {predict}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> head(predict(spbin, newdata = moose_preds))
-\end{CodeInput}
-\begin{CodeOutput}
-          1           2           3           4           5           6 
- 0.08588581 -0.40762380 -1.87510889 -1.14172781  1.45701519 -2.74275553 
-\end{CodeOutput}
-\end{CodeChunk}
-
-{augment} may also be used to augment the prediction data with
-predictions:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> head(augment(
-+   spbin,
-+   newdata = moose_preds,
-+   type.predict = "response",
-+   interval = "prediction"
-+ ))
-\end{CodeInput}
-\begin{CodeOutput}
-Simple feature collection with 6 features and 5 fields
-Geometry type: POINT
-Dimension:     XY
-Bounding box:  xmin: 291839.8 ymin: 1436192 xmax: 401239.6 ymax: 1512103
-Projected CRS: NAD83 / Alaska Albers
-# A tibble: 6 x 6
-   elev strat .fitted  .lower .upper           geometry
-  <dbl> <chr>   <dbl>   <dbl>  <dbl>        <POINT [m]>
-1  143. L      0.521  0.0983   0.916 (401239.6 1436192)
-2  324. L      0.399  0.0331   0.928 (352640.6 1490695)
-3  158. L      0.133  0.00957  0.709 (360954.9 1491590)
-4  221. M      0.242  0.0261   0.792 (291839.8 1466091)
-5  209. M      0.811  0.289    0.978 (310991.9 1441630)
-6  218. L      0.0605 0.00360  0.534 (304473.8 1512103)
-\end{CodeOutput}
-\end{CodeChunk}
-
-Here, we requested predictions on the probability (i.e., response) scale
-(Figure\(~\)\ref{fig-moose-fit}) alongside lower and upper bounds of a
-95\% (see \texttt{level}) prediction interval
-(Figure\(~\)\ref{fig-moose-int}).
-
-\begin{figure}
-\centering
-\includegraphics[width = 1\linewidth]{figures/figure-5.png}
-\caption{Moose presence probability prediction intervals. 95\% prediction interval lower bound (left) and 95\% prediction interval upper bound (right).}
-\label{fig-moose-int}
-\end{figure}
-
-Thus far we have heuristically argued, based on first principles, that
-there are benefits to incorporating spatial autocorrelation for GLMs
-applied to spatial data. Now we provide some empirical evidence to
-support this claim by comparing the fits of the SPGLM and a GLM. If
-\texttt{spcov\_type\ =\ "none"}, the resulting model fit is nearly
-identical to that from {glm}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> bin <- spglm(
-+   formula = presence ~ elev + strat + elev:strat,
-+   family = binomial,
-+   data = moose,
-+   spcov_type = "none"
-+ )
-R> bin_glm <- glm(
-+   formula = presence ~ elev + strat + elev:strat,
-+   family = binomial,
-+   data = moose
-+ )
-R> data.frame(
-+   est_none = coef(bin),
-+   est_glm = coef(bin_glm),
-+   se_none = sqrt(diag(vcov(bin))),
-+   se_glm = sqrt(diag(vcov(bin_glm)))
-+ ) |>
-+   apply(2, round, digits = 4)
-\end{CodeInput}
-\begin{CodeOutput}
-            est_none est_glm se_none se_glm
-(Intercept)  -0.5219 -0.5219  0.5082 0.5082
-elev          0.0002  0.0002  0.0024 0.0024
-stratM        1.0274  1.0274  0.7150 0.7150
-elev:stratM  -0.0013 -0.0013  0.0038 0.0038
-\end{CodeOutput}
-\end{CodeChunk}
-
-The advantage of using {spglm} to fit a model with
-\texttt{spcov\_type\ =\ "none"} is that it provides access to other
-{spmodel} functions for model objects (e.g., {glances} below) and
-accounts for the additional terms in the likelihood from
-Equation\(~\)\eqref{eq-marginal6}. These additional terms make the
-likelihood for {spglm} and {glm} different, though the models convey the
-same information. A glance at the spatial and nonspatial models reveals:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> glances(spbin, bin)
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 2 x 11
-  model     n     p  npar value   AIC  AICc   BIC logLik deviance
-  <chr> <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>
-1 spbin   218     4     3  681.  687.  687.  697.  -340.     161.
-2 bin     218     4     0  717.  717.  717.  717.  -359.     294.
-# i 1 more variable: pseudo.r.squared <dbl>
-\end{CodeOutput}
-\end{CodeChunk}
-
-The spatial model has a notably lower AIC, AICc, BIC, and deviance,
-suggesting it is the superior model. Another model comparison approach
-is leave-one-out cross validation. In leave-one-out cross validation,
-separately each observation is held out, a model is fit to the remaining
-data, and a prediction is made for the mean of the held out observation
-on the response scale. Then, statistics like leave-one-out bias,
-mean-squared-prediction error (MSPE), and the square root of MSPE
-(RMSPE) may be computed:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> loocv(spbin) |>
-+   apply(2, round, digits = 4)
-\end{CodeInput}
-\begin{CodeOutput}
-   bias    MSPE   RMSPE 
--0.0006  0.1458  0.3818 
-\end{CodeOutput}
-\begin{CodeInput}
-R> loocv(bin) |>
-+   apply(2, round, digits = 4)
-\end{CodeInput}
-\begin{CodeOutput}
-  bias   MSPE  RMSPE 
-0.0000 0.2403 0.4902 
-\end{CodeOutput}
-\end{CodeChunk}
-
-Both models are nearly unbiased, but the spatial model has an
-approximately 39\% lower MSPE, suggesting the probability predictions
-tend to be much closer to the observed presence values.
-
-A third model comparison tool is area under the receiver operating
-characteristic (AUROC) curve. The AUROC curve ranges from zero to one
-and conveys a model's classification performance over all possible
-probability thresholds \citep{james2013introduction}. Larger values of
-AUROC indicate a more accurate model:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> AUROC(spbin)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0.9490741
-\end{CodeOutput}
-\begin{CodeInput}
-R> AUROC(bin)
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0.647138
-\end{CodeOutput}
-\end{CodeChunk}
-
-All three performance metrics (likelihood-based statistics,
-leave-one-out statistics, and AUROC) prefer the spatial model.
-
-\subsection{Count data}\label{count-data}
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-6.png}
-\caption{Moose counts in Alaska. Circles represent moose counts (based on color) and triangles represent locations at which mean count predictions are desired.}
-\label{fig-moose-data-count}
-\end{figure}
-
-The \texttt{count} variable in \texttt{moose} contains the number of
-moose observed at a site (Figure\(~\)\ref{fig-moose-data-count}). Count
-data are often modeled using Poisson or negative binomial regression
-with the log link function. The Poisson regression model assumes each
-datum's underlying latent mean equals its variance, while the negative
-binomial accommodates overdispersion (where the variance is greater than
-the mean) at the cost of estimating an extra parameter.
-
-So far our spatial models have made an implicit assumption of geometric
-isotropy. A spatial covariance is geometrically isotropic if its
-dependence decays with distance equally in all directions. A spatial
-covariance is geometrically anisotropic if its dependence decays
-differently in different directions. The geometric anisotropy's
-directionality and strength are controlled by rotation and scale
-parameters that are applied to the original coordinates, creating a
-transformed set of coordinates whose spatial covariance is geometrically
-isotropic. Geometrically anisotropic models are fit by specifying
-\texttt{anisotropy}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> sppois <- spglm(
-+   formula = count ~ elev + strat + elev:strat,
-+   family = poisson,
-+   data = moose,
-+   spcov_type = "gaussian",
-+   anisotropy = TRUE
-+ )
-R> 
-R> spnbin <- update(sppois, family = nbinomial)
-\end{CodeInput}
-\end{CodeChunk}
-
-Because the models have the same support (i.e., both non-negative count
-models), we can use likelihood-based statistics to compare them:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> glances(sppois, spnbin, sort_by = "AIC") |>
-+   subset(select = c(model, npar, AIC, AICc, BIC))
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 2 x 5
-  model   npar   AIC  AICc   BIC
-  <chr>  <int> <dbl> <dbl> <dbl>
-1 spnbin     6 1318. 1319. 1339.
-2 sppois     5 1320. 1321. 1337.
-\end{CodeOutput}
-\end{CodeChunk}
-
-The negative binomial model has a slightly lower AIC and AICc, while the
-Poisson model has a slightly lower BIC. This is reasonable given the BIC
-penalizes additional parameters (here, an overdispersion parameter) more
-heavily than AIC and AICc. The leave-out-out MSPE prefers the negative
-binomial model:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> loocv(sppois) |>
-+   apply(2, round, digits = 4)
-\end{CodeInput}
-\begin{CodeOutput}
-   bias    MSPE   RMSPE 
- 1.2882 31.7882  5.6381 
-\end{CodeOutput}
-\begin{CodeInput}
-R> loocv(spnbin) |>
-+   apply(2, round, digits = 4)
-\end{CodeInput}
-\begin{CodeOutput}
-   bias    MSPE   RMSPE 
- 0.3485 28.3760  5.3269 
-\end{CodeOutput}
-\end{CodeChunk}
-
-A likelihood-based comparison between the negative binomial anisotropic
-model and the negative binomial isotropic model suggests that the
-anisotropic model is preferred (lower AIC, AICc, and BIC):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> spnbin_iso <- update(spnbin, anisotropy = FALSE)
-R> glances(spnbin_iso, spnbin, sort_by = "AIC") |>
-+   subset(select = c(model, npar, AIC, AICc, BIC))
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 2 x 5
-  model       npar   AIC  AICc   BIC
-  <chr>      <int> <dbl> <dbl> <dbl>
-1 spnbin         6 1318. 1319. 1339.
-2 spnbin_iso     4 1333. 1333. 1346.
-\end{CodeOutput}
-\end{CodeChunk}
-
-{plot} returns the spatial covariance as a function of direction
-(Figure\(~\)\ref{fig-tropy}):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot(spnbin_iso, which = 8)
-\end{CodeInput}
-
-\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-23-1} \end{center}
-
-\begin{CodeInput}
-R> plot(spnbin, which = 8)
-\end{CodeInput}
-
-\begin{center}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-23-2} \end{center}
-
-\end{CodeChunk}
-
-\begin{figure}
-\centering
-\includegraphics[width = 1\linewidth]{figures/figure-7.png}
-\caption{Level curves of equal autocorrelation for the negative binomial moose count models.
-The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of autocorrelation.
-In the isotropic level curve (left), spatial covariance decays equally in all directions.
-In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts).}
-\label{fig-tropy}
-\end{figure}
-
-Earlier we used {tidy} to tidy the model's fixed effects, but we can
-also use tidy to tidy the spatial covariance parameters:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> tidy(spnbin, effects = "spcov")
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 5 x 3
-  term    estimate is_known
-  <chr>      <dbl> <lgl>   
-1 de         4.65  FALSE   
-2 ie         0.148 FALSE   
-3 range  84013.    FALSE   
-4 rotate     2.74  FALSE   
-5 scale      0.259 FALSE   
-\end{CodeOutput}
-\end{CodeChunk}
-
-The \texttt{rotate} parameter is the number of radians in \([0, \pi]\)
-the ellipse is rotated and the \texttt{scale} parameter is the ratio in
-\((0, 1]\) of the minor axis length to the major axis length. The
-\texttt{is\_known} column indicates whether the parameter was assumed
-known during optimization, controlled by specifying the
-\texttt{spcov\_initial} argument.
-
-\subsection{Skewed data}\label{skewed-data}
-
-The \texttt{seal} data in {spmodel} is an \texttt{sf} object with data
-on harbor seal trends in Alaska. The \texttt{log\_trend} variable is the
-logarithm of a seal abundance temporal trend measure at the site (based
-on historical data), and the \texttt{stock} variable is a factor with
-two levels, \texttt{8} and \texttt{10}, where each level represents one
-of twelve seal stocks (i.e., breeds) in Alaska. The \texttt{seal}
-geometry is an areal polygon geometry and hence, spatial autoregressive
-models based on neighborhood distance are appropriate.
-
-SPGLMs for areal data are are fit in {spmodel} using {spgautor}, which
-has similar syntax as {spglm} but contains arguments to control the
-weight matrix (\(\mathbf{W}\) in Equation\(~\)\ref{eq-spcov-sar}) and
-whether or not row-standardization \citep{ver2018spatial} is applied. By
-default, polygons are neighbors if they share a boundary (i.e., Queen's
-contiguity; see \citet{pebesma2023spatial}) and row standardization is
-assumed. Weight matrices may be provided via the \texttt{W} argument and
-row standardization may be ignored via the \texttt{row\_st} arguent.
-Following \citet{ver2018spatial}, polygons without a neighbor are given
-their own (independent) variance parameter called \texttt{extra}.
-
-The trend data were originally logged to remove skew
-\citep{ver2018spatial}, but we will exponentiate \texttt{log\_trend}
-(Figure\(~\)\ref{fig-seal}) and model this skew directly using a SPGLM:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> seal$trend <- exp(seal$log_trend)
-\end{CodeInput}
-\end{CodeChunk}
-
-The \texttt{trend} variable has several missing (\texttt{NA}) values,
-which represent polygons at which predictions of \texttt{trend} are
-desired. To make predictions using spatial autoregressive models, the
-prediction locations must be known prior to model fitting because these
-locations affect the neighborhood structure of the observed data
-\citep{ver2018spatial}. This restriction is notably different than
-SPGLMs for point-referenced data (i.e., geostatistical models), which
-completely separates the estimation and prediction steps.
-
-\begin{figure}
-\centering
-\includegraphics[width = 1\linewidth]{figures/figure-8.png}
-\caption{Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed seal trends (right).}
-\label{fig-seal}
-\end{figure}
-
-{spmodel} supports the gamma and inverse Gaussian families for modeling
-skewed, positive response variables. {spmodel} also supports nonspatial
-random effects specified via the \texttt{random} argument, which uses a
-similar formula syntax as {nlme} \citep{pinheiro2006mixed} and {lme4}
-\citep{bates2015lme4}. Using likelihood-based statistics, we compare two
-models fit using the simultaneous autoregressive covariance and the
-Gamma and inverse Gaussian families. Both models have a random effect
-for seal stock, which builds additional correlation into the model for
-two polygons from the same stock:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> spgam <- spgautor(
-+   formula = trend ~ 1,
-+   family = Gamma,
-+   data = seal,
-+   spcov_type = "sar",
-+   random = ~ stock
-+ )
-R> spinvg <- update(spgam, family = inverse.gaussian)
-R> glances(spgam, spinvg, sort_by = "AIC") |>
-+   subset(select = c(model, npar, AIC, AICc, BIC))
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 2 x 5
-  model   npar   AIC  AICc   BIC
-  <chr>  <int> <dbl> <dbl> <dbl>
-1 spinvg     5  108.  109.  121.
-2 spgam      5  114.  115.  127.
-\end{CodeOutput}
-\end{CodeChunk}
-
-The inverse Gaussian model has a lower AIC, AICc, and BIC, which
-indicates it is a better fit than the gamma model.
-
-We may {tidy} the estimated stock random effect variance:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> tidy(spinvg, effects = "randcov")
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 1 x 3
-  term      estimate is_known
-  <chr>        <dbl> <lgl>   
-1 1 | stock  0.00411 FALSE   
-\end{CodeOutput}
-\end{CodeChunk}
-
-The locations to predict \texttt{trend} (\texttt{NA} values) are stored
-in the \texttt{newdata} element of \texttt{spinvg} and used for
-prediction:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> predict(spinvg, type = "response", interval = "prediction")
-\end{CodeInput}
-\end{CodeChunk}
-
-If using {augment} for prediction, \texttt{newdata} must be specified:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> head(augment(
-+   spinvg,
-+   newdata = spinvg$newdata,
-+   type.predict = "response",
-+   interval = "prediction"
-+ ))
-\end{CodeInput}
-\begin{CodeOutput}
-Simple feature collection with 6 features and 6 fields
-Geometry type: POLYGON
-Dimension:     XY
-Bounding box:  xmin: 1030504 ymin: 1012786 xmax: 1115097 ymax: 1057579
-Projected CRS: NAD83 / Alaska Albers
-# A tibble: 6 x 7
-  log_trend stock trend .fitted .lower .upper                           geometry
-      <dbl> <fct> <dbl>   <dbl>  <dbl>  <dbl>                      <POLYGON [m]>
-1        NA 8        NA   0.942  0.893  0.993 ((1035002 1054710, 1035002 105454~
-2        NA 8        NA   0.942  0.893  0.993 ((1043093 1020553, 1043097 102055~
-3        NA 8        NA   0.942  0.893  0.993 ((1099737 1054310, 1099752 105426~
-4        NA 8        NA   0.942  0.893  0.993 ((1099002 1036542, 1099134 103646~
-5        NA 8        NA   0.942  0.893  0.993 ((1076902 1053189, 1076912 105317~
-6        NA 8        NA   0.942  0.893  0.993 ((1070501 1046969, 1070317 104659~
-\end{CodeOutput}
-\end{CodeChunk}
-
-\subsection{Proportion data}\label{proportion-data}
-
-We end with two examples of beta regression for proportion data
-\citep{ferrari2004beta}. First, we model the nitrogen percentage in a
-caribou foraging experiment. Second, we model the proportion of voter
-turnout by Texas county in the United States (US) 1980 presidential
-election.
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-9.png}
-\caption{Caribou data from a spatial experimental design measuring the percentage of nitrogen (z) in soil and testing two factors: tarp type and water presence.}
-\label{fig-caribou}
-\end{figure}
-
-The \texttt{caribou} data in {spmodel} are a data frame from a caribou
-foraging experiment in Alaska meant to study the impact of water and
-tarp cover on the percentage of nitrogen in surrounding plants
-(Figure\(~\)\ref{fig-caribou}). \citet{lenart2002climate} studied these
-data treating nitrogen percentage as a continuous variable, but here we
-treat nitrogen percentage (\texttt{z}) as a proportion:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> spbeta <- spglm(
-+     formula = z/100 ~ water + tarp + water:tarp,
-+     family = "beta",
-+     data = caribou,
-+     spcov_type = "matern",
-+     xcoord = x,
-+     ycoord = y
-+ )
-\end{CodeInput}
-\end{CodeChunk}
-
-The nitrogen percentage is dynamically scaled in \texttt{formula} from
-(0, 100) to (0, 1) so that it is a proportion. Nitrogen percentage is
-modeled as a function of \texttt{water} (two levels: \texttt{"Y"} for
-water and \texttt{"N"} for no water), \texttt{tarp} (three levels:
-\texttt{"clear"} for a clear tarp, \texttt{"none"} for no tarp, and
-\texttt{"shade"} for a shaded tarp), and their interaction, which lets
-the effect of water presence vary across tarp type. \texttt{caribou} is
-a data frame (not an \texttt{sf} object), so we supply the x-coordinate
-and y-coordinate directly via \texttt{xcoord} and \texttt{ycoord}, and,
-consistent with the tidyverse approach \citep{wickham2019welcome},
-column names in \texttt{data} do not need to be quoted when referenced
-(but can be quoted).
-
-A summary of \texttt{spbeta} returns a coefficients table that provides
-parameter estimates relative to a reference group. When factors have
-more than two levels, it is not straightforward to use these contrasts
-to determine overall significance of the factor. The \texttt{anova()}
-function tests marginal (i.e., Type III sums of squares) significance of
-factors using the general linear hypothesis test for spatial (i.e.,
-correlated) data \citep{schabenberger2017statistical}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> tidy(anova(spbeta))
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 4 x 4
-  effects        df statistic       p.value
-  <chr>       <int>     <dbl>         <dbl>
-1 (Intercept)     1  35783.   0            
-2 water           1      1.52 0.218        
-3 tarp            2     38.4  0.00000000468
-4 water:tarp      2      5.74 0.0566       
-\end{CodeOutput}
-\end{CodeChunk}
-
-These results suggest there is some evidence that the effect of water on
-nitrogen percentage differ depending on the type of tarp used (0.01
-\textless{} \(p~\)value \textless{} 0.1).
-
-Sometimes averages or contrasts between factor levels that are not in
-the reference group are of interest. We again leverage {spmodel}'s
-built-in support for {emmeans} and use it to obtain the averages of each
-factor combination on the link (here, logit) scale:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> spemm <- emmeans(spbeta, ~ water + tarp)
-R> spemm
-\end{CodeInput}
-\begin{CodeOutput}
- water tarp  emmean     SE  df asymp.LCL asymp.UCL
- N     clear  -3.94 0.0208 Inf     -3.98     -3.90
- Y     clear  -3.90 0.0206 Inf     -3.94     -3.86
- N     none   -3.88 0.0204 Inf     -3.92     -3.84
- Y     none   -3.91 0.0206 Inf     -3.95     -3.87
- N     shade  -3.77 0.0196 Inf     -3.81     -3.73
- Y     shade  -3.83 0.0200 Inf     -3.87     -3.79
-
-Degrees-of-freedom method: asymptotic 
-Results are given on the logit (not the response) scale. 
-Confidence level used: 0.95 
-\end{CodeOutput}
-\end{CodeChunk}
-
-Delta method \citep{ver2012invented} standard errors are used when
-averages on the response (here, proportion) scale are desired:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> update(spemm, type = "response")
-\end{CodeInput}
-\begin{CodeOutput}
- water tarp  response       SE  df asymp.LCL asymp.UCL
- N     clear   0.0191 0.000390 Inf    0.0183    0.0198
- Y     clear   0.0198 0.000398 Inf    0.0190    0.0205
- N     none    0.0202 0.000404 Inf    0.0194    0.0210
- Y     none    0.0197 0.000398 Inf    0.0189    0.0205
- N     shade   0.0226 0.000434 Inf    0.0218    0.0235
- Y     shade   0.0213 0.000418 Inf    0.0205    0.0221
-
-Degrees-of-freedom method: asymptotic 
-Confidence level used: 0.95 
-Intervals are back-transformed from the logit scale 
-\end{CodeOutput}
-\end{CodeChunk}
-
-Pairwise contrasts use a Tukey \(p~\)value adjustment
-\citep{tukey1949comparing} by default. Here, we request no \(p~\)value
-adjustment:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> head(pairs(spemm, adjust = "none"))
-\end{CodeInput}
-\begin{CodeOutput}
- contrast          estimate     SE  df z.ratio p.value
- N clear - Y clear  -0.0361 0.0293 Inf  -1.233  0.2177
- N clear - N none   -0.0601 0.0292 Inf  -2.063  0.0391
- N clear - Y none   -0.0327 0.0293 Inf  -1.118  0.2637
- N clear - N shade  -0.1735 0.0286 Inf  -6.061  <.0001
- N clear - Y shade  -0.1136 0.0289 Inf  -3.932  0.0001
- Y clear - N none   -0.0241 0.0290 Inf  -0.830  0.4064
-
-Degrees-of-freedom method: asymptotic 
-Results are given on the log odds ratio (not the response) scale. 
-\end{CodeOutput}
-\end{CodeChunk}
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-10.png}
-\caption{Proportion of voter turnout in Texas for the 1980 presidential election.
-Circles represent voter turnout (based on color) and triangles represent locations at which voter turnout predictions are desired.}
-\label{fig-texas}
-\end{figure}
-
-We now model the \texttt{elect80} data in {spData}
-\citep{bivand2024spdata}, which contains voter turnout data by county in
-the 1980 US Presidential election \citep{pace1997quick}. The
-\texttt{texas} data in {spmodel} contains a subset of these data in the
-state of Texas. These data are point-referenced, but we may still use
-autoregressive models if neighborhood distance is determined using
-county centroids (i.e., counties whose centroid distance is less than
-some cutoff are defined as neighbors). The response variable of
-interest, \texttt{turnout}, is the proportion of registered voters in
-the county who voted in the election (Figure\(~\)\ref{fig-texas}).
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> spgautor_mods <- spgautor(
-+   formula = turnout ~ log_income,
-+   family = beta,
-+   data = texas,
-+   spcov_type = c("car", "sar"),
-+   cutoff = 2e5,
-+   estmethod = "ml"
-+ ) 
-\end{CodeInput}
-\end{CodeChunk}
-
-We model voter turnout as a function of log income using both the
-conditional and simultaneous autoregressive models with a neighbor
-distance cutoff of 200 kilometers and the maximum likelihood estimation
-method. When a vector is provided to \texttt{spcov\_type} in {spgautor}
-(or {spglm}), a model is fit for each spatial covariance type and stored
-in a list with name equal to the respective type. Then it is simple to
-glance at each model fit:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> glances(spgautor_mods) |>
-+   subset(select = c(model, npar, AIC, AICc, BIC))
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 2 x 5
-  model  npar   AIC  AICc   BIC
-  <chr> <int> <dbl> <dbl> <dbl>
-1 car       3 -38.3 -38.1 -21.2
-2 sar       3 -35.7 -35.5 -18.6
-\end{CodeOutput}
-\end{CodeChunk}
-
-The conditional autoregressive model has the best fit (in terms of AIC,
-AICc, and BIC). In this model, there is significant evidence
-\texttt{log\_income} is positively related to average voter turnout
-(\(p~\)value \textless{} 0.001):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> tidy(spgautor_mods$car)
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 2 x 5
-  term        estimate std.error statistic    p.value
-  <chr>          <dbl>     <dbl>     <dbl>      <dbl>
-1 (Intercept)   -4.74      1.06      -4.47 0.00000783
-2 log_income     0.532     0.117      4.55 0.00000531
-\end{CodeOutput}
-\end{CodeChunk}
-
-Another way to assess the impact of \texttt{log\_income} on
-\texttt{turnout} is a likelihood ratio test:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> reduced_car <- update(spgautor_mods$car, formula = turnout ~ 1)
-R> tidy(anova(reduced_car, spgautor_mods$car))
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 1 x 5
-  full              reduced        df statistic     p.value
-  <chr>             <chr>       <dbl>     <dbl>       <dbl>
-1 spgautor_mods$car reduced_car     1      24.1 0.000000930
-\end{CodeOutput}
-\end{CodeChunk}
-
-Likelihood ratio tests compare the fit of a ``full'' model compared to a
-``reduced'' model that is completely nested within the full model. If
-there is evidence the full model explains significantly more information
-than the reduced model, the likelihood ratio test \(p~\)value will be
-small. Similar to the summary output from the general linear hypothesis
-test, the likelihood ratio suggests \texttt{log\_income} is related to
-average voter turnout (\(p~\)value \textless{} 0.001).
-
-The default estimation method in {spmodel} is REML, but note that these
-models used maximum likelihood (ML). ML is very similar to REML -- the
-difference is that for ML, the fixed effects \(\boldsymbol{\beta}\) are
-not integrated out of Equation\(~\)\eqref{eq-marginal} but are rather
-back-substituted. While REML typically performs better for fixed effect
-estimation and prediction \citep{zimmerman2024spatial}, ML allows
-likelihood-based comparisons (e.g., AIC) for models with simultaneously
-varying fixed effect and covariance structures, while REML
-likelihood-based comparisons are only valid for models sharing the same
-fixed effect structure (though \citet{gurka2006selecting} provides some
-evidence that this restriction may be unnecessary).
-
-The point-referenced \texttt{texas} and \texttt{caribou} data may be
-analyzed using {spgautor} or {spglm} and comparisons across these
-structures can be made using likelihood-based statistics (as long as the
-supports of the response distribution are the same). Put another way,
-likelihood-based statistics can be used to determine whether
-geostatistical (distance-based) or autoregressive (neighbor-based)
-structures perform best when the data are point-referenced.
-
-\section{Discussion}\label{sec-discussion}
-
-SPGLMs are fit in {spmodel} using a novel application of the Laplace
-approximation that marginalizes over the latent (i.e., unobserved) mean,
-\(\mathbf{w}\), and the fixed effects, \(\boldsymbol{\beta}\). The
-approach is quite flexible and accommodates any general response
-distribution and covariance structure. \citet{ver2024marginal} show that
-the approach, as implemented in {spmodel}, generally yields unbiased
-estimators with proper interval coverage and often outperforms the
-Bayesian approach from {spBayes}, the INLA approach from {R-INLA}, and
-the automatic differentiation approach from {glmmTMB}.
-
-{spmodel}'s {spglm} and {spgautor} functions are similar in structure
-and syntax as the base-{R} {glm} function, easing the transition from
-GLMs to SPGLMs. These functions support six response distributions
-(Table\(~\)\ref{table-links}) and 20 spatial covariance functions.
-{spmodel} provides several additional features that accommodate
-geometric anisotropy, nonspatial random effects, fixing spatial
-covariance parameters at known values, data having thousands of
-observations (following \citet{ver2023indexing}), incorporating spatial
-dependence in machine learning (e.g., random forests;
-\citet{breiman2001random}), simulating spatially dependent data (e.g.,
-{sprbinom}, {sprpois}), and several others. Learn more at
-\url{https://CRAN.R-project.org/package=spmodel} and links therein.
-
-\section*{Computational details}\label{computational-details}
-\addcontentsline{toc}{section}{Computational details}
-
-The results in this paper were obtained using {R} 4.4.0 with the
-{spmodel} 0.9.0 package. Figures were created using the {ggplot2} 3.5.1
-package \citep{wickham2016ggplot2} and base {R}.
-
-\section*{Data and code availability}\label{data-and-code-availability}
-\addcontentsline{toc}{section}{Data and code availability}
-
-All writing and code associated with this manuscript is available for
-viewing and download on GitHub at
-\url{https://github.com/USEPA/spmodel.glm.manuscript}. All data used are
-part of the {spmodel} {R} package available for download from CRAN at
-\url{https://CRAN.R-project.org/package=spmodel}.
-
-\section*{Acknowledgments}\label{acknowledgments}
-\addcontentsline{toc}{section}{Acknowledgments}
-
-We would like to thank initial reviewers and editors for feedback that
-has greatly improved the manuscript.
-
-The views expressed in this manuscript are those of the authors and do
-not necessarily represent the views or policies of the U.S.
-Environmental Protection Agency or the National Oceanic and Atmospheric
-Administration. Any mention of trade names, products, or services does
-not imply an endorsement by the U.S. government, the U.S. Environmental
-Protection Agency, or the National Oceanic and Atmospheric
-Administration. The U.S. Environmental Protection Agency and the
-National Oceanic and Atmospheric Administration do not endorse any
-commercial products, services or enterprises.
+\subsection{Modeling lake conductivity in Southwest,
+USA}\label{modeling-lake-conductivity-in-southwest-usa}
 

### 2025-08-22 inst/manuscript/extras.R

@@ -1,0 +1,442 @@
+
+
+
+
+
+
+
+
+The \code{spglm()} (for point-referenced data) and \code{spgautor()} (for areal data) functions in \pkg{spmodel} fit SPGLMs using the Laplace approximation outlined in Section$~$\ref{sec-spglm}. Both \code{spglm()} and \code{spgautor()} generally require the following four arguments: `formula`, a formula that describes the relationship between the response variable and explanatory variables; `family`, the response distribution (which can be `binomial`, `poisson`, `nbinomial`, `Gamma`, `inverse.gaussian`, or `beta`); `data`, the data frame that holds the variables in `formula` as well as spatial locations; and `spcov_type`, the spatial covariance type. The first three arguments are shared by [glm]{.fct}; thus, the transition from GLMs to SPGLMs requires only one additional argument: `spcov_type`. The \code{spglm()} spatial covariance types measure dependence as a function of Euclidean distance among observations; an example is the exponential spatial covariance:
+  \begin{equation}\label{eq-spcov-exp}
+\boldsymbol{\Sigma} = \sigma^2_{de} \exp(-\mathbf{H}/\phi) + \sigma^2_{ie}\mathbf{I},
+\end{equation}
+where $\mathbf{H}$ is a matrix of pairwise distances among all observations. \code{spglm()} currently supports 18 distinct spatial covariance functions.
+
+The \code{spgautor()} spatial covariance types measure dependence as a function of neighborhood distance among observations; an example is the simultaneous autoregressive covariance matrix:
+  \begin{equation}\label{eq-spcov-sar}
+\boldsymbol{\Sigma} = \sigma^2_{de} [(\mathbf{I} - \phi \mathbf{W})(\mathbf{I} - \phi \mathbf{W})^T]^{-1} + \sigma^2_{ie}\mathbf{I},
+\end{equation}
+where $\mathbf{W}$ is a matrix that represents the neighborhood structure among all observations. \code{spgautor()} currently supports two distinct spatial covariance functions.
+
[…]

### 2025-08-22 inst/manuscript/figures/figures.R

@@ -5,6 +5,54 @@ library(patchwork)
 library(here)
 fig_path <- here("inst", "manuscript", "figures")
 
+
+#######################
+######## Figure 1
+#######################
+h <- seq(0, 1, length.out = 1000)
+range1 <- 0.2
+dat1 <- data.frame(Distance = h, Correlation = exp(-h/range1), Range = "0.2")
+range2 <- 0.5
+dat2 <- data.frame(Distance = h, Correlation = exp(-h/range2), Range = "0.5")
+range3 <- 0.8
+dat3 <- data.frame(Distance = h, Correlation = exp(-h/range3), Range = "0.8")
+dat <- bind_rows(dat1, dat2, dat3)
+cor_range <- ggplot(dat, aes(x = Distance, y = Correlation, color = Range, linetype = Range)) +
+  geom_line(linewidth = 1.2) +
+  scale_color_viridis_d() +
+  theme_bw(base_size = 14)
+cor_range
+ggsave(
+  filename = paste0(fig_path, "/figure-01.png"),
+  plot = cor_range,
+  dpi = 300,
+  height = 6.07,
+  width = 6.49
+)
+
+#######################
+######## Figure 1
+#######################
+h <- seq(0, 1, length.out = 1000)
+range <- 0.5
+dat1 <- data.frame(Distance = h, Correlation = exp(-h/range), Type = "Exponential")
+dat2 <- data.frame(Distance = h, Correlation = exp(-(h/range)^2), Type = "Gaussian")
+dat3 <- data.frame(Distance = h, Correlation = (1 - 1.5 * h/range + 0.5 * (h/range)^3) * (h <= range), Type = "Spherical")
+dat <- bind_rows(dat1, dat2, dat3)
+cor_type <- ggplot(dat, aes(x = Distance, y = Correlation, color = Type, linetype = Type)) +
+  geom_line(linewidth = 1.2) +
+  scale_color_viridis_d() +
+  theme_bw(base_size = 14)
+cor_type
+ggsave(
+  filename = paste0(fig_path, "/figure-02.png"),
+  plot = cor_type,
+  dpi = 300,
+  height = 6.07,
+  width = 6.49
+)
+
+
 #######################
 ######## Figure 1
 #######################

### 2025-08-26 inst/manuscript/manuscript.Rmd
some updates

@@
 
-@ver2024marginal show that the fixed effect estimator is $\hat{\boldsymbol{\beta}} = (\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{X})^{-1}\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{w}$, where $\hat{\boldsymbol{\Sigma}}$ is the covariance matrix $\boldsymbol{\Sigma}$ evaluated at $\hat{\boldsymbol{\theta}}$.
+Though the fixed effects are integrated out of the likelihood, we can still estimate them using generalized least squares (GLS) principles, a common practice for linear models estimated using restricted maximum likelihood methods.
+Had we observed $\mathbf{w}$, a GLS estimator for $\boldsymbol{\beta}$ is given by
+\begin{equation}\label{eq-gls1}
+  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
+\end{equation}
+where $\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}$.
+However, we only observe $\hat{\mathbf{w}}$, so it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$.
+Thus, to derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$.
+To do so, we must condition on $\mathbf{w}$ as if it were observed and invoke properties of the laws of total expectation and variance.
+Because $\hat{\mathbf{w}}$ was optimized via the likelihood, we assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance approximately equal to $-\mathbf{H}^{-1}$ (the inverse Hessian).
+It follows that $\text{E}(\hat{\mathbf{w}})$ is given by
+\begin{equation}
+  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
+   = \text{E}(\mathbf{w}) 
+   = \mathbf{X}\boldsymbol{\beta}
+\end{equation}
+and $\text{Var}(\hat{\mathbf{w}})$ is given by
+\begin{align}
+  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
+  & = \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
+  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
+\end{align}
+Putting this all together, it follows that
+\begin{equation}
+  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
+  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}
+\end{equation}
+and
+\begin{align}
+  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
+  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
+  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
+  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
+  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}
+\end{align}
+In practice, $\text{Var}(\hat{\boldsymbol{\beta}})$ is estimated by evaluating $\boldsymbol{\Sigma}$ at $\hat{\boldsymbol{\theta}}$, the estimated covariance parameter vector.
+
+These results are important because they justify closed-form solutions for $\hat{\boldsymbol{\beta}}$ and its associated variance.
+Closed-form solutions are useful because they bypass the need for computationally expensive sampling-based strategies to evaluate the mean and variance of $\hat{\boldsymbol{\beta}}$ -- a common technique for other approaches to SPGLMs like Bayesian MCMC.
 
@@
 
-@ver2024marginal also derive the covariance matrix of $\hat{\boldsymbol{\beta}}$ and show how to predict $\mathbf{w}$ at unobserved locations and quantify their uncertainties.
+We may also predict values of the latent mean (on the link scale) at new locations by leveraging the spatial covariance between observed locations and new locations.
+Again suppose that we observed $\mathbf{w}$ and we want to make predictions at $\mathbf{u}$, a vector of latent means at the new locations that follows the same SPGLM from Equation~\ref{eq-spglm} with fixed effects design matrix, $\mathbf{X}_{\mathbf{u}}$.
+The vector $(\mathbf{w}, \mathbf{u})^\top$ has the following properties:
+\begin{align}
+  \text{E}(\mathbf{w}, \mathbf{u})^\top & = (\text{E}(\mathbf{w}), \text{E}(\mathbf{u}))^\top = (\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top \\
+  \text{Var}(\mathbf{w}, \mathbf{u})^\top & = \begin{bmatrix} \text{Var}(\mathbf{w}, \mathbf{w}) & \text{Var}(\mathbf{w}, \mathbf{u}) \\ \text{Var}(\mathbf{u}, \mathbf{w}) & \text{Var}(\mathbf{u}, \mathbf{u}) \end{bmatrix} = \begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}
+\end{align}
+Because we have observed $\mathbf{w}$, we may derive the conditional distribution of $\mathbf{u}|\mathbf{w}$, which has the following properties:
+\begin{align}
+  \text{E}(\mathbf{w} | \mathbf{u}) & = \mathbf{X}_{\mathbf{u}} \boldsymbol{\beta} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{w} - \mathbf{X}\boldsymbol{\beta}) \\
+  \text{E}(\mathbf{w} | \mathbf{u}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
+\end{align}
+Two more sources of uncertainty, $\hat{\boldsymbol{\beta}}$ and $\hat{\mathbf{w}}$.
+Spatial prediction is often also called Kriging [@cressie1990origins].
 

### 2025-08-26 inst/manuscript/manuscript.tex
some updates

@@
 
-\citet{ver2024marginal} show that the fixed effect estimator is
-\(\hat{\boldsymbol{\beta}} = (\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{X})^{-1}\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{w}\),
-where \(\hat{\boldsymbol{\Sigma}}\) is the covariance matrix
-\(\boldsymbol{\Sigma}\) evaluated at \(\hat{\boldsymbol{\theta}}\).
+Though the fixed effects are integrated out of the likelihood, we can
+still estimate them using generalized least squares (GLS) principles, a
+common practice for linear models estimated using restricted maximum
+likelihood methods. Had we observed \(\mathbf{w}\), a GLS estimator for
+\(\boldsymbol{\beta}\) is given by \begin{equation}\label{eq-gls1}
+  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
+\end{equation} where
+\(\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\).
+However, we only observe \(\hat{\mathbf{w}}\), so it is reasonable to
+define \(\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}\). Thus,
+to derive properties of \(\hat{\boldsymbol{\beta}}\) like expectation
+and variance, we must derive these properties for \(\hat{\mathbf{w}}\).
+To do so, we must condition on \(\mathbf{w}\) as if it were observed and
+invoke properties of the laws of total expectation and variance. Because
+\(\hat{\mathbf{w}}\) was optimized via the likelihood, we assume that
+given \(\mathbf{w}\), \(\hat{\mathbf{w}}\) has mean \(\mathbf{w}\) and
+variance approximately equal to \(-\mathbf{H}^{-1}\) (the inverse
+Hessian). It follows that \(\text{E}(\hat{\mathbf{w}})\) is given by
+\begin{equation}
+  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
+   = \text{E}(\mathbf{w}) 
+   = \mathbf{X}\boldsymbol{\beta}
+\end{equation} and \(\text{Var}(\hat{\mathbf{w}})\) is given by
+\begin{align}
+  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
+  & = \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
+  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
+\end{align} Putting this all together, it follows that \begin{equation}
+  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
+  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}
+\end{equation} and \begin{align}
+  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
+  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
+  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
+  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
+  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}
+\end{align} In practice, \(\text{Var}(\hat{\boldsymbol{\beta}})\) is
+estimated by evaluating \(\boldsymbol{\Sigma}\) at
+\(\hat{\boldsymbol{\theta}}\), the estimated covariance parameter
+vector.
+
+These results are important because they justify closed-form solutions
+for \(\hat{\boldsymbol{\beta}}\) and its associated variance.
+Closed-form solutions are useful because they bypass the need for
+computationally expensive sampling-based strategies to evaluate the mean
+and variance of \(\hat{\boldsymbol{\beta}}\) -- a common technique for
+other approaches to SPGLMs like Bayesian MCMC.
 
@@
 
-\citet{ver2024marginal} also derive the covariance matrix of
-\(\hat{\boldsymbol{\beta}}\) and show how to predict \(\mathbf{w}\) at
-unobserved locations and quantify their uncertainties.
+We may also predict values of the latent mean (on the link scale) at new
+locations by leveraging the spatial covariance between observed
+locations and new locations. Again suppose that we observed
+\(\mathbf{w}\) and we want to make predictions at \(\mathbf{u}\), a
+vector of latent means at the new locations that follows the same SPGLM
+from Equation\textasciitilde{}\ref{eq-spglm} with fixed effects design
+matrix, \(\mathbf{X}_{\mathbf{u}}\). The vector
+\((\mathbf{w}, \mathbf{u})^\top\) has the following properties:
+\begin{align}
+  \text{E}(\mathbf{w}, \mathbf{u})^\top & = (\text{E}(\mathbf{w}), \text{E}(\mathbf{u}))^\top = (\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top \\
+  \text{Var}(\mathbf{w}, \mathbf{u})^\top & = \begin{bmatrix} \text{Var}(\mathbf{w}, \mathbf{w}) & \text{Var}(\mathbf{w}, \mathbf{u}) \\ \text{Var}(\mathbf{u}, \mathbf{w}) & \text{Var}(\mathbf{u}, \mathbf{u}) \end{bmatrix} = \begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}
+\end{align} Because we have observed \(\mathbf{w}\), we may derive the
+conditional distribution of \(\mathbf{u}|\mathbf{w}\), which has the
+following properties: \begin{align}
+  \text{E}(\mathbf{w} | \mathbf{u}) & = \mathbf{X}_{\mathbf{u}} \boldsymbol{\beta} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{w} - \mathbf{X}\boldsymbol{\beta}) \\
+  \text{E}(\mathbf{w} | \mathbf{u}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
+\end{align} Two more sources of uncertainty,
+\(\hat{\boldsymbol{\beta}}\) and \(\hat{\mathbf{w}}\). Spatial
+prediction is often also called Kriging \citep{cressie1990origins}.
 

### 2025-08-27 inst/manuscript/manuscript.Rmd
some updates

@@
 where $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ is the density for the appropriate response distribution of $\mathbf{y}$ (e.g., binomial, Poisson) given the latent $\mathbf{w}$ and dispersion parameter ($\varphi$), and $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ is the multivariate Gaussian density for $\mathbf{w}$ given the explanatory variables ($\mathbf{X}$), fixed effects ($\boldsymbol{\beta}$), and spatial covariance parameters ($\boldsymbol{\theta}$).
-Following @@harville1977maximum, we can integrate $\boldsymbol{\beta}$ out of Equation$~$\ref{eq-spglm}, which yields
+The elements of $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ are conditionally independent (given $\mathbf{w}$), but the elements of $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ share spatial covariance.
+Following @harville1977maximum, we can integrate $\boldsymbol{\beta}$ out of Equation$~$\ref{eq-spglm}, which yields
 \begin{equation}\label{eq-marginal2}
@@
 
+## Inspecting model diagnostics
+
+Inspecting model diagnostics is an important step of the modeling process that can yield valuable insights into model behavior and unusual observations.
+@montgomery2021introduction contextualize three components of unusual observations: outliers, leverage, and influence.
+An observation is an outlier if it has an unusual response value relative to expectation.
+The response GLM residuals simply compare the observation to its fitted latent mean:
+\begin{equation}
+  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
+\end{equation}
+Because observations often have a unique support in a GLM (e.g., only two possible response values for binary data) and the variance of an observation generally depends on its mean, response residuals lack some utility.
+Deviance residuals are a function of response residuals that are appropriately scaled to behave more like response residuals in a standard linear model.
+Deviance residuals are given by
+\begin{equation}
+  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
+\end{equation}
+where $\mathbf{d}$ is a vector of individual deviances.
+The sum of the squared deviance residuals equals the sum of $\mathbf{d}$.
+The sum of $\mathbf{d}$ is the deviance of the model fit, which quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)$ for all $i$) and the fitted model.
+Deviance is often used as a fit statistic; lower values of deviance imply a better model fit.
+Pearson and standardized residuals are other types of GLM residuals that involve some scaling of the response residuals; the Pearson residuals scale $\mathbf{r}_{r}$ by the squrae root of $\mathbf{V}$, while the standardized residuals scale the deviance residuals by $\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}$, where $\mathbf{L}_{ii}$ is the $i$th diagonal element of the leverage matrix, which we discuss next.
+An observation has high leverage if its combination of explanatory variables is far away from other observations.
+In a linear model, the leverage values are the diagonal of the leverage (i.e., projection, hat) matrix, $\mathbf{L} = \mathbf{X}(\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top$.
+In a GLM, the leverage matrix is given by
+\begin{equation}
+  \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2},
+\end{equation}
+where $\mathbf{V}$ is a diagonal matrix with $i$th diagonal element equal to the variance of the response distribution evaluated at $f^{-1}(\mathbf{w}_i)$ [@faraway2016extending].
+The matrix $\mathbf{V}$ is sometimes called the GLM weight matrix.
+The larger the $i$th diagonal element of the hat matrix, the more severe the leverage from the $i$th observation.
+An observation is influential if it has a sizeable impact on model fit. Influence is measured using Cook's distance [@cook1979influential; @cook1982residuals], which is given for a GLM by
+\begin{equation}
+  \mathbf{c} = \mathbf{r}^2_{s} \frac{diag(\mathbf{L})}{\text{tr}(\mathbf{L})(\mathbf{1} - diag(\mathbf{L}))},
+\end{equation}
+where $\mathbf{r}^2_{s}$ are the standardized residuals and $diag(\mathbf{L})$ indicates the diagonal elements of the leverage matrix.
+The larger the $i$th diagonal element of the hat matrix, the more severe the influence from the $i$th observation.
+@montgomery2021introduction provide guidance for interpreting these types of statistics, including cutoffs to consider when identifying unusual residual, leverage, or influence values.
+
+In a linear model, the $R^2$ (R-squared) statistic quantifies the proportion of variability in the data captured by the explanatory variables and is calculated as one minus the ratio of the error sum of squares to the total sum of squares [@rencher2008linear].
+In a GLM, there are many ways to define such a statistic [@smith2013comparison].
+One such approach is to use one minus the deviance ratio:
+\begin{equation}
+  PR^2 = 1 - \frac{deviance_{fit}}{deviance_{null}},
+\end{equation}
+where $deviance_{fit}$ is the deviance of the fitted model (sometimes called the residual deviance) and $deviance_{null}$ is the deviance of the model taking $\mathbf{X} \equiv \mathbf{1}$, a column of all ones (i.e., an intercept-only model).
+In practice, $deviance_{null}$ is derived by computing $\hat{\mathbf{w}}$ when $\mathbf{X} \equiv \mathbf{1}$ given $\hat{\boldsymbol{\theta}}$ and $\hat{\varphi}$ from the fitted model.
+Like the standard $R^2$, this statistic attempts to capture variability (i.e., deviance) attributable to the explanatory variables.
+Because the $deviance_{null}$ denominator changes across fitted models (as the values of $\hat{\boldsymbol{\theta}}$ and $\hat{\varphi}$ change), this statistic should not be used as a model comparison tool.
+Instead, it should be used as an informative diagnostic tool unique to each model fit.
 ## Predicting at new locations
 
-We may also predict values of the latent mean (on the link scale) at new locations by leveraging the spatial covariance between observed locations and new locations.
+We may also predict values of the latent mean (on the link scale) at new locations by leveraging the spatial covariance between observed locations and new locations (spatial prediction is also called Kriging; see @cressie1990origins).
 Again suppose that we observed $\mathbf{w}$ and we want to make predictions at $\mathbf{u}$, a vector of latent means at the new locations that follows the same SPGLM from Equation~\ref{eq-spglm} with fixed effects design matrix, $\mathbf{X}_{\mathbf{u}}$.
@@
 \end{align}
-Two more sources of uncertainty, $\hat{\boldsymbol{\beta}}$ and $\hat{\mathbf{w}}$.
-Spatial prediction is often also called Kriging [@cressie1990origins].
+@ver2024marginal show how these equations are adjusted to reflect uncertainty in both $\hat{\boldsymbol{\beta}}$ and $\hat{\mathbf{w}}$ while leveraging the laws of total expectation and variance yet again.
+They derive the predictor of $\mathbf{u}$, $\hat{\mathbf{u}}$, and its associated variance, given by:
+\begin{align}
+  \hat{\mathbf{u}} & = \mathbf{X}_{\mathbf{u}} \hat{\boldsymbol{\beta}} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\hat{\mathbf{w}} - \mathbf{X}\hat{\boldsymbol{\beta}}) \\
+  \text{Var}(\hat{\mathbf{u}}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}} + \mathbf{K}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{K}^\top + \boldsymbol{\Lambda}(-\mathbf{H})^{-1}\boldsymbol{\Lambda}^\top,
+\end{align}
+where $\mathbf{K} = \mathbf{X}_{\mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \mathbf{X}$ and $\boldsymbol{\Lambda} = \mathbf{X}_{\mathbf{u}}\mathbf{B} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{1} - \mathbf{X}\mathbf{B})$ for a vector of ones, $\mathbf{1}$.
+
+As with $\hat{\boldsymbol{\beta}}$, in practice these covariance matrices are evaluated at $\hat{\boldsymbol{\theta}}$.
+Moreover, these closed-form solutions provided enhance computational efficiency and clarity of the predictor's behavior.
 
@@
 
-## Additional applications {#sec-applications2}
+The \code{moose} data in \pkg{spmodel} contain information on moose (Alces Alces) presence in the Togiak region of Alaska, USA.
+\code{moose}is an \code{sf} object, a special data frame that is supplemented with spatial information using the \pkg{sf} package in \proglang{R} [@pebesma2018sf].
+The first few rows of \code{moose} look like:
+
+```{r}
+head(moose)
+```
+
+There are five columns: \code{elev}, the numeric site elevation (meters); \code{strat} a stratification variable for sampling with two levels, \code{"L"} and \code{"M"}, which are categorized by landscape metrics at each site; \code{count}, the number of moose at each site; \code{presence}, a factor that indicates whether at least one moose was observed at each site (\code{0} implies no moose; \code{1} implies at least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG: 3338) projected coordinate of each site (these data are point-referenced because each observation occurs at point coordinates and are represented by a \code{POINT} geometry.
+The \code{moose_preds} data in \pkg{spmodel} contain spatial locations at which predictions of moose presence are desired (and is also point-referenced).
+\code{moose_preds} is also an \code{sf} object with measurements for \code{elev} and \code{strat} and the same projection system.
+Figure$~$\ref{fig-moose-data} shows the \code{presence} variable in \code{moose} as well as the spatial locations of both \code{moose} and \code{moose_preds}.
+Moose are most commonly present in the southwestern and eastern parts of the domain and least commonly present in the northwest (Figure$~$\ref{fig-moose-data}).
+Next we show how to use \pkg{spmodel} to study the effect of elevation and strata on moose presence while accounting for spatial covariance and to make predictions of moose presence at new locations.
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-1.png}
+\caption{Moose presence in Alaska. Circles represent moose presence or absence (based on color) and triangles represent locations at which moose presence probability predictions are desired.}
+\label{fig-moose-data}
+\end{figure}
+
+## Model Fitting
+
+SPGLMs in \pkg{spmodel} are fit using the \code{spglm()} function.
+The \code{spglm()} function requires four arguments: `formula`, the relationship between the response and explanatory variables; `family`, the response distribution assumed for the repsonse variable; `data`, the data frame that contains the variables in `formula`, and `spcov_type`, the type of spatial covariance.
+These first three arguments are the three required arguments to \code{glm()} for nonspatial GLMs.
+So, the transition from \code{glm()} to \code{spglm()}
+simply requires one additional argument: \code{spcov_type}.
+When \code{data} is not an \code{sf} object, \code{spglm()} also requires the \code{xcoord} and \code{ycoord} arguments, which indicate the columns in \code{data} that represent the x- and y-coordinates, respectively (it is assumed these coordinates are already projected).
+
+We use \code{spglm()} to fit a spatial logistic regression model quantifying the effect of elevation and strata on moose presence:
+```{r}
+spbin <- spglm(
+  formula = presence ~ elev + strat,
+  family = binomial,
+  data = moose,
+  spcov_type = "exponential"
+)
+```
+
+The \code{summary()} function returns a model summary that returns relevant information like the function call, deviance residuals, a coefficients table of fixed effects, the pseudo R-squared, spatial covariance parameter coefficient estimates, and the GLM dispersion parameter (fixed at one in logistic regression):
+```{r}
+summary(spbin)
+```
+
+Based on this model, there is some evidence that elevation is associated with higher probabilities of moose presence ($p$-value $\approx$ 0.087) but strong evidence that moose are more prevalent in the \code{"M"} strata than the \code{"L"} strata ($p$-value < 0.001).
+The fixed effects coefficients table from \code{summary()} is often of primary practical interest, but it is not easily usable when printed directly to the \proglang{R} console.
+The \code{tidy()} function tidies this table, turning it into a data frame (i.e., a tibble) with standard column names:
+```{r}
+tidy(spbin, conf.int = TRUE)
+```
+
+## Model Comparison
+
+The strength of spatial covarinace in the data affects how beneficial a SPGLM is relative to a GLM.
+When the spatial covariance is strong, the SPGLM should notably outperform the GLM.
+When the spatial covariance is weak, the SPGLM and GLM should perform similarly.
+We can quantify the benefits of incorporating spatial covariance for a particular data set by comparing the fit of a SPGLM to a GLM.
+We can fit a GLM in \code{spmodel} by specifying \code{spcov_type = "none"}:
+
+```{r}
+bin <- spglm(
+  formula = presence ~ elev + strat,
+  family = binomial,
+  data = moose,
+  spcov_type = "none"
+)
+```
+
+While the \code{spglm()} approach evaluates the HGLMM likelhood with $\sigma^2_{de} = 0$ and $\sigma^2_{ie} \approx 0$ instead of just the GLM likelihood, the parameter estimates and their standard errors are the same:
+
+```{r}
+bin_glm <- glm(
+  formula = presence ~ elev + strat,
+  family = binomial,
+  data = moose,
+)
+round(coef(bin), digits = 4)
+round(coef(bin_glm), digits = 4)
+round(sqrt(diag(vcov(bin))), digits = 4)
+round(sqrt(diag(vcov(bin_glm))), digits = 4)
+```
+
+However, using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel} helper functions are available and that each of the \code{spglm()} models uses the same likelihood:
+```{r}
+glance(spbin)
+glance(bin)
+```
+
+The likelihood-based statistics AIC, AICc, BIC, and deviance are much lower for the SPGLM, indicating a better fit relative to the GLM.
+We may also perform a likelihood ratio test (LRT) between the two models, as the GLM is a special case of the SPGLM (i.e., is nested within the SPGLM):
+```{r}
+tidy(anova(spbin, bin))
+```
+
+The LRT test statistic is $\approx$ 31.5 with three degrees of freedom (the difference in covariance parameters), yielding a $p$-value $< 0.001$ that indicates preference for the full model (SPGLM) relative to the reduced model (GLM).
+
+An alternative approach to model comparison is to use a cross-validation procedure [@james2013introduction].
+The \code{loocv()} function performs leave-one-out cross validation, comparing the predicted mean (on the response scale) to the observed response variable for each hold-out observation, recomputing estimates of $\boldsymbol{\beta}$ each time.
+Then, statistics like bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) can be used to evaluate models:
+
+```{r}
+loocv(spbin)
+loocv(bin)
+```
+
+Both models have negligible bias, but the SPGLM has much lower MSPE and RMSPE than the GLM, indicating the SPGLM predictions are far more efficient.
+Three separate metrics (likelihood-based statistics, likelihood-ratio test, and leave-one-out cross validation) prefer the SPGLM to the GLM.
+
+We can compare two SPGLMs with different spatial covariance functions using likelihood-based statistics and leave-one-out cross validation, but we can't use the LRT because generally, the spatial covariance functions aren't nested:
+```{r}
+spbin2 <- update(spbin, spcov_type = "spherical")
+glances(spbin, spbin2)
+loocv(spbin)
+loocv(spbin2)
+```
+
+The \code{"exponential"} spatial covariance (\code{spbin}) has a slightly lower deviance but slightly higher AIC, AICc, and BIC than 
+the \code{"spherical"} spatial covariance (\code{spbin2}).
+Both spatial covariance functions nearly identical leave-one-out cross validation metrics.
+For practical purposes, these models fit quite similarly.
+
+## Model Diagnostics
+
+\code{spmodel} provides a suite of tools for model diagnostics.
+
+## Prediction
+
+# Additional applications {#sec-applications2}
 
@@
 
+# Discussion
+
 \bibliography{references.bib}

### 2025-08-27 inst/manuscript/manuscript.tex
some updates

@@
 (\(\mathbf{X}\)), fixed effects (\(\boldsymbol{\beta}\)), and spatial
-covariance parameters (\(\boldsymbol{\theta}\)). Following
-@\citet{harville1977maximum}, we can integrate \(\boldsymbol{\beta}\)
-out of Equation\(~\)\ref{eq-spglm}, which yields
+covariance parameters (\(\boldsymbol{\theta}\)). The elements of
+\([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) are conditionally
+independent (given \(\mathbf{w}\)), but the elements of
+\([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\) share spatial
+covariance. Following \citet{harville1977maximum}, we can integrate
+\(\boldsymbol{\beta}\) out of Equation\(~\)\ref{eq-spglm}, which yields
 \begin{equation}\label{eq-marginal2}
@@
 
-\subsection{Predicting at new
-locations}\label{predicting-at-new-locations}
+\subsection{Inspecting model
+diagnostics}\label{inspecting-model-diagnostics}
+
+Inspecting model diagnostics is an important step of the modeling
+process that can yield valuable insights into model behavior and unusual
+observations. \citet{montgomery2021introduction} contextualize three
+components of unusual observations: outliers, leverage, and influence.
+An observation is an outlier if it has an unusual response value
+relative to expectation. The response GLM residuals simply compare the
+observation to its fitted latent mean: \begin{equation}
+  \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
+\end{equation} Because observations often have a unique support in a GLM
+(e.g., only two possible response values for binary data) and the
+variance of an observation generally depends on its mean, response
+residuals lack some utility. Deviance residuals are a function of
+response residuals that are appropriately scaled to behave more like
+response residuals in a standard linear model. Deviance residuals are
+given by \begin{equation}
+  \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
+\end{equation} where \(\mathbf{d}\) is a vector of individual deviances.
+The sum of the squared deviance residuals equals the sum of
+\(\mathbf{d}\). The sum of \(\mathbf{d}\) is the deviance of the model
+fit, which quantifies twice the difference in log likelihoods between
+the a saturated model that fits every observation perfectly (i.e.,
+\(\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)\) for all \(i\)) and the
+fitted model. Deviance is often used as a fit statistic; lower values of
+deviance imply a better model fit. Pearson and standardized residuals
+are other types of GLM residuals that involve some scaling of the
+response residuals; the Pearson residuals scale \(\mathbf{r}_{r}\) by
+the squrae root of \(\mathbf{V}\), while the standardized residuals
+scale the deviance residuals by
+\(\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}\), where \(\mathbf{L}_{ii}\) is
+the \(i\)th diagonal element of the leverage matrix, which we discuss
+next. An observation has high leverage if its combination of explanatory
+variables is far away from other observations. In a linear model, the
+leverage values are the diagonal of the leverage (i.e., projection, hat)
+matrix,
+\(\mathbf{L} = \mathbf{X}(\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top\).
+In a GLM, the leverage matrix is given by \begin{equation}
+  \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2},
+\end{equation} where \(\mathbf{V}\) is a diagonal matrix with \(i\)th
+diagonal element equal to the variance of the response distribution
+evaluated at \(f^{-1}(\mathbf{w}_i)\) \citep{faraway2016extending}. The
+matrix \(\mathbf{V}\) is sometimes called the GLM weight matrix. The
+larger the \(i\)th diagonal element of the hat matrix, the more severe
+the leverage from the \(i\)th observation. An observation is influential
+if it has a sizeable impact on model fit. Influence is measured using
+Cook's distance \citep{cook1979influential, cook1982residuals}, which is
+given for a GLM by \begin{equation}
+  \mathbf{c} = \mathbf{r}^2_{s} \frac{diag(\mathbf{L})}{\text{tr}(\mathbf{L})(\mathbf{1} - diag(\mathbf{L}))},
+\end{equation} where \(\mathbf{r}^2_{s}\) are the standardized residuals
+and \(diag(\mathbf{L})\) indicates the diagonal elements of the leverage
+matrix. The larger the \(i\)th diagonal element of the hat matrix, the
+more severe the influence from the \(i\)th observation.
+\citet{montgomery2021introduction} provide guidance for interpreting
+these types of statistics, including cutoffs to consider when
+identifying unusual residual, leverage, or influence values.
+
+In a linear model, the \(R^2\) (R-squared) statistic quantifies the
+proportion of variability in the data captured by the explanatory
+variables and is calculated as one minus the ratio of the error sum of
+squares to the total sum of squares \citep{rencher2008linear}. In a GLM,
+there are many ways to define such a statistic
+\citep{smith2013comparison}. One such approach is to use one minus the
+deviance ratio: \begin{equation}
+  PR^2 = 1 - \frac{deviance_{fit}}{deviance_{null}},
+\end{equation} where \(deviance_{fit}\) is the deviance of the fitted
+model (sometimes called the residual deviance) and \(deviance_{null}\)
+is the deviance of the model taking \(\mathbf{X} \equiv \mathbf{1}\), a
+column of all ones (i.e., an intercept-only model). In practice,
+\(deviance_{null}\) is derived by computing \(\hat{\mathbf{w}}\) when
+\(\mathbf{X} \equiv \mathbf{1}\) given \(\hat{\boldsymbol{\theta}}\) and
+\(\hat{\varphi}\) from the fitted model. Like the standard \(R^2\), this
+statistic attempts to capture variability (i.e., deviance) attributable
+to the explanatory variables. Because the \(deviance_{null}\)
+denominator changes across fitted models (as the values of
+\(\hat{\boldsymbol{\theta}}\) and \(\hat{\varphi}\) change), this
+statistic should not be used as a model comparison tool. Instead, it
+should be used as an informative diagnostic tool unique to each model
+fit. \#\# Predicting at new locations
 
@@
 locations by leveraging the spatial covariance between observed
-locations and new locations. Again suppose that we observed
+locations and new locations (spatial prediction is also called Kriging;
+see \citet{cressie1990origins}). Again suppose that we observed
 \(\mathbf{w}\) and we want to make predictions at \(\mathbf{u}\), a
@@
   \text{E}(\mathbf{w} | \mathbf{u}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
-\end{align} Two more sources of uncertainty,
-\(\hat{\boldsymbol{\beta}}\) and \(\hat{\mathbf{w}}\). Spatial
-prediction is often also called Kriging \citep{cressie1990origins}.
+\end{align} \citet{ver2024marginal} show how these equations are
+adjusted to reflect uncertainty in both \(\hat{\boldsymbol{\beta}}\) and
+\(\hat{\mathbf{w}}\) while leveraging the laws of total expectation and
+variance yet again. They derive the predictor of \(\mathbf{u}\),
+\(\hat{\mathbf{u}}\), and its associated variance, given by:
+\begin{align}
+  \hat{\mathbf{u}} & = \mathbf{X}_{\mathbf{u}} \hat{\boldsymbol{\beta}} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\hat{\mathbf{w}} - \mathbf{X}\hat{\boldsymbol{\beta}}) \\
+  \text{Var}(\hat{\mathbf{u}}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}} + \mathbf{K}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{K}^\top + \boldsymbol{\Lambda}(-\mathbf{H})^{-1}\boldsymbol{\Lambda}^\top,
+\end{align} where
+\(\mathbf{K} = \mathbf{X}_{\mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \mathbf{X}\)
+and
+\(\boldsymbol{\Lambda} = \mathbf{X}_{\mathbf{u}}\mathbf{B} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{1} - \mathbf{X}\mathbf{B})\)
+for a vector of ones, \(\mathbf{1}\).
+
+As with \(\hat{\boldsymbol{\beta}}\), in practice these covariance
+matrices are evaluated at \(\hat{\boldsymbol{\theta}}\). Moreover, these
+closed-form solutions provided enhance computational efficiency and
+clarity of the predictor's behavior.
 
@@
 
-\subsection{Additional applications}\label{sec-applications2}
+The \code{moose} data in \pkg{spmodel} contain information on moose
+(Alces Alces) presence in the Togiak region of Alaska, USA.
+\code{moose}is an \code{sf} object, a special data frame that is
+supplemented with spatial information using the \pkg{sf} package in
+\proglang{R} \citep{pebesma2018sf}. The first few rows of \code{moose}
+look like:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> head(moose)
+\end{CodeInput}
+\begin{CodeOutput}
+Simple feature collection with 6 features and 4 fields
+Geometry type: POINT
+Dimension:     XY
+Bounding box:  xmin: 281896.4 ymin: 1518398 xmax: 311325.3 ymax: 1541016
+Projected CRS: NAD83 / Alaska Albers
+# A tibble: 6 x 5
+   elev strat count presence           geometry
+  <dbl> <chr> <dbl> <fct>           <POINT [m]>
+1  469. L         0 0        (293542.6 1541016)
+2  362. L         0 0        (298313.1 1533972)
+3  173. M         0 0        (281896.4 1532516)
+4  280. L         0 0        (298651.3 1530264)
+5  620. L         0 0        (311325.3 1527705)
+6  164. M         0 0        (291421.5 1518398)
+\end{CodeOutput}
+\end{CodeChunk}
+
+There are five columns: \code{elev}, the numeric site elevation
+(meters); \code{strat} a stratification variable for sampling with two
+levels, \code{"L"} and \code{"M"}, which are categorized by landscape
+metrics at each site; \code{count}, the number of moose at each site;
+\code{presence}, a factor that indicates whether at least one moose was
+observed at each site (\code{0} implies no moose; \code{1} implies at
+least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG:
+3338) projected coordinate of each site (these data are point-referenced
+because each observation occurs at point coordinates and are represented
+by a \code{POINT} geometry. The \code{moose_preds} data in \pkg{spmodel}
+contain spatial locations at which predictions of moose presence are
+desired (and is also point-referenced). \code{moose_preds} is also an
+\code{sf} object with measurements for \code{elev} and \code{strat} and
+the same projection system. Figure\(~\)\ref{fig-moose-data} shows the
+\code{presence} variable in \code{moose} as well as the spatial
+locations of both \code{moose} and \code{moose_preds}. Moose are most
+commonly present in the southwestern and eastern parts of the domain and
+least commonly present in the northwest
+(Figure\(~\)\ref{fig-moose-data}). Next we show how to use \pkg{spmodel}
+to study the effect of elevation and strata on moose presence while
+accounting for spatial covariance and to make predictions of moose
+presence at new locations.
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-1.png}
+\caption{Moose presence in Alaska. Circles represent moose presence or absence (based on color) and triangles represent locations at which moose presence probability predictions are desired.}
+\label{fig-moose-data}
+\end{figure}
+
+\subsection{Model Fitting}\label{model-fitting}
+
+SPGLMs in \pkg{spmodel} are fit using the \code{spglm()} function. The
+\code{spglm()} function requires four arguments: \texttt{formula}, the
+relationship between the response and explanatory variables;
+\texttt{family}, the response distribution assumed for the repsonse
+variable; \texttt{data}, the data frame that contains the variables in
+\texttt{formula}, and \texttt{spcov\_type}, the type of spatial
+covariance. These first three arguments are the three required arguments
+to \code{glm()} for nonspatial GLMs. So, the transition from
+\code{glm()} to \code{spglm()} simply requires one additional argument:
+\code{spcov_type}. When \code{data} is not an \code{sf} object,
+\code{spglm()} also requires the \code{xcoord} and \code{ycoord}
+arguments, which indicate the columns in \code{data} that represent the
+x- and y-coordinates, respectively (it is assumed these coordinates are
+already projected).
+
+We use \code{spglm()} to fit a spatial logistic regression model
+quantifying the effect of elevation and strata on moose presence:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> spbin <- spglm(
++   formula = presence ~ elev + strat,
++   family = binomial,
++   data = moose,
++   spcov_type = "exponential"
++ )
+\end{CodeInput}
+\end{CodeChunk}
+
+The \code{summary()} function returns a model summary that returns
+relevant information like the function call, deviance residuals, a
+coefficients table of fixed effects, the pseudo R-squared, spatial
+covariance parameter coefficient estimates, and the GLM dispersion
+parameter (fixed at one in logistic regression):
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> summary(spbin)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Call:
+spglm(formula = presence ~ elev + strat, family = binomial, data = moose, 
+    spcov_type = "exponential")
+
+Deviance Residuals:
+    Min      1Q  Median      3Q     Max 
+-1.7535 -0.8005  0.3484  0.7893  1.5797
+
+Coefficients (fixed):
+             Estimate Std. Error z value Pr(>|z|)    
+(Intercept) -2.465713   1.486212  -1.659 0.097104 .  
+elev         0.006036   0.003525   1.712 0.086861 .  
+stratM       1.439273   0.420591   3.422 0.000622 ***
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+Pseudo R-squared: 0.06275
+
+Coefficients (exponential spatial covariance):
+       de        ie     range 
+5.145e+00 1.294e-03 4.199e+04
+
+Coefficients (Dispersion for binomial family):
+dispersion 
+         1 
+\end{CodeOutput}
+\end{CodeChunk}
+
+Based on this model, there is some evidence that elevation is associated
+with higher probabilities of moose presence (\(p\)-value \(\approx\)
+0.087) but strong evidence that moose are more prevalent in the
+\code{"M"} strata than the \code{"L"} strata (\(p\)-value \textless{}
+0.001). The fixed effects coefficients table from \code{summary()} is
+often of primary practical interest, but it is not easily usable when
+printed directly to the \proglang{R} console. The \code{tidy()} function
+tidies this table, turning it into a data frame (i.e., a tibble) with
+standard column names:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> tidy(spbin, conf.int = TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 3 x 7
+  term        estimate std.error statistic  p.value  conf.low conf.high
+  <chr>          <dbl>     <dbl>     <dbl>    <dbl>     <dbl>     <dbl>
+1 (Intercept) -2.47      1.49        -1.66 0.0971   -5.38        0.447 
+2 elev         0.00604   0.00353      1.71 0.0869   -0.000873    0.0129
+3 stratM       1.44      0.421        3.42 0.000622  0.615       2.26  
+\end{CodeOutput}
+\end{CodeChunk}
+
+\subsection{Model Comparison}\label{model-comparison}
+
+The strength of spatial covarinace in the data affects how beneficial a
+SPGLM is relative to a GLM. When the spatial covariance is strong, the
+SPGLM should notably outperform the GLM. When the spatial covariance is
+weak, the SPGLM and GLM should perform similarly. We can quantify the
+benefits of incorporating spatial covariance for a particular data set
+by comparing the fit of a SPGLM to a GLM. We can fit a GLM in
+\code{spmodel} by specifying \code{spcov_type = "none"}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> bin <- spglm(
++   formula = presence ~ elev + strat,
++   family = binomial,
++   data = moose,
++   spcov_type = "none"
++ )
+\end{CodeInput}
+\end{CodeChunk}
+
+While the \code{spglm()} approach evaluates the HGLMM likelhood with
+\(\sigma^2_{de} = 0\) and \(\sigma^2_{ie} \approx 0\) instead of just
+the GLM likelihood, the parameter estimates and their standard errors
+are the same:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> bin_glm <- glm(
++   formula = presence ~ elev + strat,
++   family = binomial,
++   data = moose,
++ )
+R> round(coef(bin), digits = 4)
+\end{CodeInput}
+\begin{CodeOutput}
+(Intercept)        elev      stratM 
+    -0.4247     -0.0003      0.8070 
+\end{CodeOutput}
+\begin{CodeInput}
+R> round(coef(bin_glm), digits = 4)
+\end{CodeInput}
+\begin{CodeOutput}
+(Intercept)        elev      stratM 
+    -0.4247     -0.0003      0.8070 
+\end{CodeOutput}
+\begin{CodeInput}
+R> round(sqrt(diag(vcov(bin))), digits = 4)
+\end{CodeInput}
+\begin{CodeOutput}
+(Intercept)        elev      stratM 
+     0.4208      0.0019      0.2906 
+\end{CodeOutput}
+\begin{CodeInput}
+R> round(sqrt(diag(vcov(bin_glm))), digits = 4)
+\end{CodeInput}
+\begin{CodeOutput}
+(Intercept)        elev      stratM 
+     0.4208      0.0019      0.2906 
+\end{CodeOutput}
+\end{CodeChunk}
+
+However, using \code{spglm()} instead of \code{glm()} ensures that
+\pkg{spmodel} helper functions are available and that each of the
+\code{spglm()} models uses the same likelihood:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> glance(spbin)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 1 x 10
+      n     p  npar value   AIC  AICc   BIC logLik deviance pseudo.r.squared
+  <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>            <dbl>
+1   218     3     3  676.  682.  683.  693.  -338.     176.           0.0627
+\end{CodeOutput}
+\begin{CodeInput}
+R> glance(bin)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 1 x 10
+      n     p  npar value   AIC  AICc   BIC logLik deviance pseudo.r.squared
+  <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>            <dbl>
+1   218     3     0  708.  708.  708.  708.  -354.     294.           0.0280
+\end{CodeOutput}
+\end{CodeChunk}
+
+The likelihood-based statistics AIC, AICc, BIC, and deviance are much
+lower for the SPGLM, indicating a better fit relative to the GLM. We may
+also perform a likelihood ratio test (LRT) between the two models, as
+the GLM is a special case of the SPGLM (i.e., is nested within the
+SPGLM):
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> tidy(anova(spbin, bin))
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 1 x 5
+  full  reduced    df statistic     p.value
+  <chr> <chr>   <int>     <dbl>       <dbl>
+1 spbin bin         3      31.5 0.000000652
+\end{CodeOutput}
+\end{CodeChunk}
+
+The LRT test statistic is \(\approx\) 31.5 with three degrees of freedom
+(the difference in covariance parameters), yielding a \(p\)-value
+\(< 0.001\) that indicates preference for the full model (SPGLM)
+relative to the reduced model (GLM).
+
+An alternative approach to model comparison is to use a cross-validation
+procedure \citep{james2013introduction}. The \code{loocv()} function
+performs leave-one-out cross validation, comparing the predicted mean
+(on the response scale) to the observed response variable for each
+hold-out observation, recomputing estimates of \(\boldsymbol{\beta}\)
+each time. Then, statistics like bias, mean-squared-prediction error
+(MSPE), and the square root of MSPE (RMSPE) can be used to evaluate
+models:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> loocv(spbin)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 1 x 3
+       bias  MSPE RMSPE
+      <dbl> <dbl> <dbl>
+1 0.0000206 0.156 0.394
+\end{CodeOutput}
+\begin{CodeInput}
+R> loocv(bin)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 1 x 3
+      bias  MSPE RMSPE
+     <dbl> <dbl> <dbl>
+1 -1.23e-9 0.240 0.490
+\end{CodeOutput}
+\end{CodeChunk}
+
+Both models have negligible bias, but the SPGLM has much lower MSPE and
+RMSPE than the GLM, indicating the SPGLM predictions are far more
+efficient. Three separate metrics (likelihood-based statistics,
+likelihood-ratio test, and leave-one-out cross validation) prefer the
+SPGLM to the GLM.
+
+We can compare two SPGLMs with different spatial covariance functions
+using likelihood-based statistics and leave-one-out cross validation,
+but we can't use the LRT because generally, the spatial covariance
+functions aren't nested:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> spbin2 <- update(spbin, spcov_type = "spherical")
+R> glances(spbin, spbin2)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 2 x 11
+  model      n     p  npar value   AIC  AICc   BIC logLik deviance
+  <chr>  <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>
+1 spbin2   218     3     3  675.  681.  681.  691.  -338.     180.
+2 spbin    218     3     3  676.  682.  683.  693.  -338.     176.
+# i 1 more variable: pseudo.r.squared <dbl>
+\end{CodeOutput}
+\begin{CodeInput}
+R> loocv(spbin)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 1 x 3
+       bias  MSPE RMSPE
+      <dbl> <dbl> <dbl>
+1 0.0000206 0.156 0.394
+\end{CodeOutput}
+\begin{CodeInput}
+R> loocv(spbin2)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 1 x 3
+      bias  MSPE RMSPE
+     <dbl> <dbl> <dbl>
+1 0.000121 0.155 0.394
+\end{CodeOutput}
+\end{CodeChunk}
+
+The \code{"exponential"} spatial covariance (\code{spbin}) has a
+slightly lower deviance but slightly higher AIC, AICc, and BIC than the
+\code{"spherical"} spatial covariance (\code{spbin2}). Both spatial
+covariance functions nearly identical leave-one-out cross validation
+metrics. For practical purposes, these models fit quite similarly.
+
+\subsection{Model Diagnostics}\label{model-diagnostics}
+
+\code{spmodel} provides a suite of tools for model diagnostics.
+
+\subsection{Prediction}\label{prediction}
+
+\section{Additional applications}\label{sec-applications2}
 
@@
 
+\section{Discussion}\label{discussion}
+
 \bibliography{references.bib}

### 2025-08-27 inst/manuscript/manuscript.Rmd
more updates

@@
 library(dplyr)
-library(emmeans)
+#library(emmeans)
+#library(car)
 library(here)
@@
 \pkg{spmodel} (version 0.11.0) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in terms of scope and feel.
-Both packages use similar syntax as \code{glm()}, accommodate flexible \code{formula} arguments (e.g., offsets, splines), handle spatial covariance that decays at different rates in different rates (i.e., geometric anisotropy), incorporate nonspatial random effects, support other \proglang{R} packages for modeling like \pkg{broom} [@robinson2021broom; @kuhn2022tidy] and \pkg{emmeans} [@lenth2024emmeans], and have tools for model summaries, prediction, and simulating data.
+Both packages use similar syntax as \code{glm()}, accommodate flexible \code{formula} arguments (e.g., offsets, splines), handle spatial covariance that decays at different rates in different rates (i.e., geometric anisotropy), incorporate nonspatial random effects, support other \proglang{R} packages for modeling like \pkg{broom} [@robinson2021broom; @kuhn2022tidy], \pkg{emmeans} [@lenth2024emmeans], and \pkg{car} [@fox2019car], and have tools for model summaries, prediction, and simulating data.
 There are some notable differences between the two packages, however.
@@
 Instead, it should be used as an informative diagnostic tool unique to each model fit.
+
 ## Predicting at new locations
@@
 
-\code{spmodel} provides a suite of tools for model diagnostics.
+\code{spmodel} provides a suite of tools for model diagnostics. The \code{augment()} function augments the model data with diagnostics:
+```{r}
+augment(spbin)
+```
+
+The fitted values (\code{.fitted}) can be returned on either the link ($\hat{\mathbf{w}}$) or response ($f^{-1}\hat{\mathbf{w}}$) scale and the residuals (\code{.resid}) can deviance, pearson, or response residuals.
+The defaults are the link scale and deviance residuals, respectively.
+Also returned by \code{augment()} are the leverage (\code{.hat}), Cook's distance (\code{.cooksd}), and standardized residuals \code{.std.resid}.
+A benefit of using \code{augment()} when \code{data} is an \code{sf} object is that the output is also an \code{sf} object, which makes it straightforward to create spatial diagnostic plots (Figure$~$\ref{fig-sp-diagnostic}).
+Standard \proglang{R} helpers (e.g., \code{fitted()}, \code{residuals()}) are available to alternatively extract model diagnostics from the model object.
+
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-2.png}
+\caption{Spatial logistic regression model diagnostics from [augment]{.fct}. The leverage (i.e., hat) values (left) and standardized residuals (right).}
+\label{fig-sp-diagnostic}
+\end{figure}
+
+The \code{plot()} function can also be used to return similar diagnostics as from \code{lm()} and \code{glm()} with additional tools for spatial covariance.
+For example, we can inspect Cook's distance values and the empirical spatial covariance as a function (Figure$~$\ref{fig-sp-diagnostic2}) with
+```{r, eval = FALSE}
+plot(spbin, which = c(4, 7))
+```
+
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-3.png}
+\caption{Spatial logistic regression model diagnostics from [plot]{.fct}. The Cook's distance values (left) and the fitted spatial covariance as a function of distance (right).}
+\label{fig-sp-diagnostic2}
+\end{figure}
+
+The \code{varcomp()} function partitions model variability into several different components:
+```{r}
+varcomp(spbin)
+```
+
+The pseudo R-squared ($PR^2$), the proportion of variability attributable to the explanatory variables, is reported in the first row.
+The remaining variability ($1 - PR^2$) is allocated proportionally to \code{de} and \code{ie} according to $\sigma^2_{de}$ and $\sigma^2_{ie}$.
+This variability partitioning is a useful that helps quantify how much the explanatory variables, residual spatial variance, and residual nonspatial variance contribute to model fit, but as with $PR^2$, should not be used as a model comparison tool.
 
@@
 
+We can predict the probability of moose presence using \code{predict()}:
+```{r}
+predict(spbin, newdata = moose_preds)[1:5]
+```
+
+By default, predictions are returned on the link scale, but this can be changed to the response scale via \code{type}:
+```{r}
+predict(spbin, newdata = moose_preds, type = "response")[1:5]
+```
+
+Predictions on the response scale are visualized alongside the fitted values ($f^{-1}\hat{\mathbf{w}}$) in Figure$~$\ref{fig-moose-fit}.
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-4.png}
+\caption{Moose presence probability fitted values and predictions. Fitted values are represeneted by circles and predictions by triangles.}
+\label{fig-moose-fit}
+\end{figure}
+
+Prediction intervals are returned via \code{interval}
+```{r}
+predict(spbin, newdata = moose_preds, interval = "prediction")[1:5, ]
+```
+
+We can alternatively use \code{augment()} to augment the prediction data with predictions. Arguments to \code{predict()} can also be passed to \code{augment()}:
+```{r}
+augment(spbin, newdata = moose_preds, interval = "prediction")
+```
+
+By using \code{augment()} when \code{newdata} is an \code{sf} object, predictions and their corresponding uncertainties are readily available for spatial mapping (Figure$~$\ref{fig-moose-int}).
+
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-5.png}
+\caption{Moose presence probability prediction intervals. 95\% prediction interval lower bound (left) and 95\% prediction interval upper bound (right).}
+\label{fig-moose-int}
+\end{figure}
+
 # Additional applications {#sec-applications2}
 
-## Modeling moose counts in Alaska, USA
+Throughout the remainder of this section, we briefly highlight some additional \pkg{spmodel} capabilities for SPGLMs.
+In Section$~$\ref{sec-moose-count}, we fit Poisson and negative binomial models with and without geometric anisotropy for the point-referenced moose count data.
+In Section$~$\ref{sec-seal}, we fit a binomial model to the areal seal trend data with a nonspatial random effect.
+In Section$~$\ref{sec-texas}, we fit beta models to Texas voter turnout data.
+We explore how the Texas data can be treated as point-referenced or areal and compare the two spatial covariance structures empirically.
+We also highlight how the maximum likelihood estimation method is useful when comparing two models with different explanatory variables.
+Finally, in Section$~$\ref{sec-lake}, we fit a Gamma model to the point-referenced lake conductivity data.
+We show how to perform a spatial analysis of variance (ANOVA) and leverage modeling functions from other \proglang{R} packages like \pkg{emmeans} and \pkg{car}.
 
-## Modeling harbor seal trends in Alaska, USA
+## Modeling moose counts in Alaska, USA {#sec-moose-count}
 
-## Modeling voter turnout in Texas, USA
+```{r}
+sppois <- spglm(
+  formula = count ~ elev + strat,
+  family = poisson,
+  data = moose,
+  spcov_type = "gaussian"
+)
+spnb <- update(sppois, family = nbinomial)
+sppois_anis <- update(sppois, anisotropy = TRUE)
+spnb_anis <- update(sppois_anis, family = nbinomial)
+```
 
-## Modeling lake conductivity in Southwest, USA
+```{r}
+BIC(sppois, spnb, sppois_anis, spnb_anis)
+```
+
+```{r, eval = FALSE}
+plot(spnb, which = 8)
+plot(spnb_anis, which = 8)
+```
+
+## Modeling harbor seal trends in Alaska, USA {#sec-seal}
+
+```{r}
+spbin <- spgautor(
+  formula = log_trend > 0 ~ 1,
+  family = binomial,
+  data = seal,
+  spcov_type = "car",
+  random = ~ stock
+)
+tidy(spbin, conf.int = TRUE)
+```
+
+## Modeling voter turnout in Texas, USA {#sec-texas}
+
+```{r}
+spbeta_geo <- spglm(
+  formula = turnout ~ log_income,
+  family = "beta", 
+  data = texas,
+  spcov_type = "matern"
+)
+
+spbeta_auto <- spgautor(
+  formula = turnout ~ log_income,
+  family = "beta", 
+  data = texas,
+  spcov_type = "car",
+  cutoff = 1e5
+)
+```
+
+```{r}
+AIC(spbeta_geo, spbeta_auto)
+```
+
+```{r}
+tidy(spbeta_geo)
+spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
+spbeta_red_ml <- update(spbeta_full_ml, formula = turnout ~ 1)
+tidy(anova(spbeta_full_ml, spbeta_red_ml))
+```
+
+## Modeling lake conductivity in Southwest, USA {#sec-lake}
+
+```{r}
+spgam <- spglm(
+  formula = exp(log_cond) ~ temp * state + origin, 
+  family = "Gamma",
+  data = lake,
+  spcov_type = "cauchy",
+  partition_factor = ~ year
+)
+```
+
+```{r}
+anova(spgam)
+```
+
+```{r}
+car::vif(spgam)
+```
+
+```{r}
+pairs(emmeans::emmeans(spgam, ~ state | temp))
+emmeans::emtrends(spgam, ~ state, var = "temp")
+```
 

### 2025-08-27 inst/manuscript/manuscript.tex
more updates

@@
 packages for modeling like \pkg{broom}
-\citep{robinson2021broom, kuhn2022tidy} and \pkg{emmeans}
-\citep{lenth2024emmeans}, and have tools for model summaries,
-prediction, and simulating data. There are some notable differences
-between the two packages, however. \pkg{sdmTMB} supports several
-additional GLM distributions like the Tweedie, supports Hurdle models,
-and can incorporate prior information through Bayesian applications.
-\pkg{sdmTMB} also provides tools for working with temporal data and
-enhanced visualizations of marginal effects. \pkg{sdmTMB} does require a
-preprocessing step of constructing a mesh for the stochastic partial
-differential equation approach, and the density of the mesh can affect
-model results and computational complexity. \pkg{spmodel} does not
-require the construction of a mesh prior to modeling. \pkg{spmodel}
+\citep{robinson2021broom, kuhn2022tidy}, \pkg{emmeans}
+\citep{lenth2024emmeans}, and \pkg{car} \citep{fox2019car}, and have
+tools for model summaries, prediction, and simulating data. There are
+some notable differences between the two packages, however. \pkg{sdmTMB}
+supports several additional GLM distributions like the Tweedie, supports
+Hurdle models, and can incorporate prior information through Bayesian
+applications. \pkg{sdmTMB} also provides tools for working with temporal
+data and enhanced visualizations of marginal effects. \pkg{sdmTMB} does
+require a preprocessing step of constructing a mesh for the stochastic
+partial differential equation approach, and the density of the mesh can
+affect model results and computational complexity. \pkg{spmodel} does
+not require the construction of a mesh prior to modeling. \pkg{spmodel}
 supports 20 different spatial covariances and models them directly,
@@
 should be used as an informative diagnostic tool unique to each model
-fit. \#\# Predicting at new locations
+fit.
+
+\subsection{Predicting at new
+locations}\label{predicting-at-new-locations}
 
@@
 
-\code{spmodel} provides a suite of tools for model diagnostics.
+\code{spmodel} provides a suite of tools for model diagnostics. The
+\code{augment()} function augments the model data with diagnostics:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> augment(spbin)
+\end{CodeInput}
+\begin{CodeOutput}
+Simple feature collection with 218 features and 8 fields
+Geometry type: POINT
+Dimension:     XY
+Bounding box:  xmin: 269085 ymin: 1416151 xmax: 419057.4 ymax: 1541016
+Projected CRS: NAD83 / Alaska Albers
+# A tibble: 218 x 9
+   presence  elev strat .fitted .resid    .hat  .cooksd .std.resid
+ * <fct>    <dbl> <chr>   <dbl>  <dbl>   <dbl>    <dbl>      <dbl>
+ 1 0         469. L       -1.95 -0.516 0.0476  0.00465      -0.528
+ 2 0         362. L       -2.70 -0.361 0.0123  0.000548     -0.363
+ 3 0         173. M       -1.96 -0.514 0.00455 0.000405     -0.516
+ 4 0         280. L       -3.15 -0.290 0.00413 0.000117     -0.291
+ 5 0         620. L       -1.19 -0.728 0.168   0.0427       -0.798
+ 6 0         164. M       -1.71 -0.576 0.00534 0.000598     -0.578
+ 7 0         164. M       -1.60 -0.606 0.00576 0.000714     -0.608
+ 8 0         186. L       -2.50 -0.397 0.00439 0.000233     -0.398
+ 9 0         362. L       -1.88 -0.532 0.0239  0.00237      -0.539
+10 0         430. L       -1.54 -0.623 0.0497  0.00713      -0.639
+# i 208 more rows
+# i 1 more variable: geometry <POINT [m]>
+\end{CodeOutput}
+\end{CodeChunk}
+
+The fitted values (\code{.fitted}) can be returned on either the link
+(\(\hat{\mathbf{w}}\)) or response (\(f^{-1}\hat{\mathbf{w}}\)) scale
+and the residuals (\code{.resid}) can deviance, pearson, or response
+residuals. The defaults are the link scale and deviance residuals,
+respectively. Also returned by \code{augment()} are the leverage
+(\code{.hat}), Cook's distance (\code{.cooksd}), and standardized
+residuals \code{.std.resid}. A benefit of using \code{augment()} when
+\code{data} is an \code{sf} object is that the output is also an
+\code{sf} object, which makes it straightforward to create spatial
+diagnostic plots (Figure\(~\)\ref{fig-sp-diagnostic}). Standard
+\proglang{R} helpers (e.g., \code{fitted()}, \code{residuals()}) are
+available to alternatively extract model diagnostics from the model
+object.
+
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-2.png}
+\caption{Spatial logistic regression model diagnostics from [augment]{.fct}. The leverage (i.e., hat) values (left) and standardized residuals (right).}
+\label{fig-sp-diagnostic}
+\end{figure}
+
+The \code{plot()} function can also be used to return similar
+diagnostics as from \code{lm()} and \code{glm()} with additional tools
+for spatial covariance. For example, we can inspect Cook's distance
+values and the empirical spatial covariance as a function
+(Figure\(~\)\ref{fig-sp-diagnostic2}) with
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot(spbin, which = c(4, 7))
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-3.png}
+\caption{Spatial logistic regression model diagnostics from [plot]{.fct}. The Cook's distance values (left) and the fitted spatial covariance as a function of distance (right).}
+\label{fig-sp-diagnostic2}
+\end{figure}
+
+The \code{varcomp()} function partitions model variability into several
+different components:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> varcomp(spbin)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 3 x 2
+  varcomp            proportion
+  <chr>                   <dbl>
+1 Covariates (PR-sq)   0.0627  
+2 de                   0.937   
+3 ie                   0.000236
+\end{CodeOutput}
+\end{CodeChunk}
+
+The pseudo R-squared (\(PR^2\)), the proportion of variability
+attributable to the explanatory variables, is reported in the first row.
+The remaining variability (\(1 - PR^2\)) is allocated proportionally to
+\code{de} and \code{ie} according to \(\sigma^2_{de}\) and
+\(\sigma^2_{ie}\). This variability partitioning is a useful that helps
+quantify how much the explanatory variables, residual spatial variance,
+and residual nonspatial variance contribute to model fit, but as with
+\(PR^2\), should not be used as a model comparison tool.
 
@@
 
+We can predict the probability of moose presence using \code{predict()}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> predict(spbin, newdata = moose_preds)[1:5]
+\end{CodeInput}
+\begin{CodeOutput}
+          1           2           3           4           5 
+ 0.06664165 -0.79069107 -1.60387940 -0.83159357  1.38183928 
+\end{CodeOutput}
+\end{CodeChunk}
+
+By default, predictions are returned on the link scale, but this can be
+changed to the response scale via \code{type}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> predict(spbin, newdata = moose_preds, type = "response")[1:5]
+\end{CodeInput}
+\begin{CodeOutput}
+        1         2         3         4         5 
+0.5166542 0.3120203 0.1674401 0.3033082 0.7992862 
+\end{CodeOutput}
+\end{CodeChunk}
+
+Predictions on the response scale are visualized alongside the fitted
+values (\(f^{-1}\hat{\mathbf{w}}\)) in Figure\(~\)\ref{fig-moose-fit}.
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-4.png}
+\caption{Moose presence probability fitted values and predictions. Fitted values are represeneted by circles and predictions by triangles.}
+\label{fig-moose-fit}
+\end{figure}
+
+Prediction intervals are returned via \code{interval}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> predict(spbin, newdata = moose_preds, interval = "prediction")[1:5, ]
+\end{CodeInput}
+\begin{CodeOutput}
+          fit        lwr       upr
+1  0.06664165 -2.0374370 2.1707203
+2 -0.79069107 -3.4758514 1.8944692
+3 -1.60387940 -4.0953329 0.8875741
+4 -0.83159357 -3.0704818 1.4072947
+5  1.38183928 -0.7692107 3.5328893
+\end{CodeOutput}
+\end{CodeChunk}
+
+We can alternatively use \code{augment()} to augment the prediction data
+with predictions. Arguments to \code{predict()} can also be passed to
+\code{augment()}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> augment(spbin, newdata = moose_preds, interval = "prediction")
+\end{CodeInput}
+\begin{CodeOutput}
+Simple feature collection with 100 features and 5 fields
+Geometry type: POINT
+Dimension:     XY
+Bounding box:  xmin: 269386.2 ymin: 1418453 xmax: 419976.2 ymax: 1541763
+Projected CRS: NAD83 / Alaska Albers
+# A tibble: 100 x 6
+    elev strat .fitted .lower  .upper           geometry
+ * <dbl> <chr>   <dbl>  <dbl>   <dbl>        <POINT [m]>
+ 1  143. L      0.0666 -2.04   2.17   (401239.6 1436192)
+ 2  324. L     -0.791  -3.48   1.89   (352640.6 1490695)
+ 3  158. L     -1.60   -4.10   0.888  (360954.9 1491590)
+ 4  221. M     -0.832  -3.07   1.41   (291839.8 1466091)
+ 5  209. M      1.38   -0.769  3.53   (310991.9 1441630)
+ 6  218. L     -2.59   -5.20   0.0177 (304473.8 1512103)
+ 7  127. L     -2.73   -5.24  -0.220  (339011.1 1459318)
+ 8  122. L     -2.32   -4.74   0.0920 (342827.3 1463452)
+ 9  191  L     -1.17   -4.01   1.66   (284453.8 1502837)
+10  105. L     -0.905  -3.05   1.24   (391343.9 1483791)
+# i 90 more rows
+\end{CodeOutput}
+\end{CodeChunk}
+
+By using \code{augment()} when \code{newdata} is an \code{sf} object,
+predictions and their corresponding uncertainties are readily available
+for spatial mapping (Figure\(~\)\ref{fig-moose-int}).
+
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-5.png}
+\caption{Moose presence probability prediction intervals. 95\% prediction interval lower bound (left) and 95\% prediction interval upper bound (right).}
+\label{fig-moose-int}
+\end{figure}
+
 \section{Additional applications}\label{sec-applications2}
 
-\subsection{Modeling moose counts in Alaska,
-USA}\label{modeling-moose-counts-in-alaska-usa}
-
-\subsection{Modeling harbor seal trends in Alaska,
-USA}\label{modeling-harbor-seal-trends-in-alaska-usa}
-
-\subsection{Modeling voter turnout in Texas,
-USA}\label{modeling-voter-turnout-in-texas-usa}
+Throughout the remainder of this section, we briefly highlight some
+additional \pkg{spmodel} capabilities for SPGLMs. In
+Section\(~\)\ref{sec-moose-count}, we fit Poisson and negative binomial
+models with and without geometric anisotropy for the point-referenced
+moose count data. In Section\(~\)\ref{sec-seal}, we fit a binomial model
+to the areal seal trend data with a nonspatial random effect. In
+Section\(~\)\ref{sec-texas}, we fit beta models to Texas voter turnout
+data. We explore how the Texas data can be treated as point-referenced
+or areal and compare the two spatial covariance structures empirically.
+We also highlight how the maximum likelihood estimation method is useful
+when comparing two models with different explanatory variables. Finally,
+in Section\(~\)\ref{sec-lake}, we fit a Gamma model to the
+point-referenced lake conductivity data. We show how to perform a
+spatial analysis of variance (ANOVA) and leverage modeling functions
+from other \proglang{R} packages like \pkg{emmeans} and \pkg{car}.
+
+\subsection{Modeling moose counts in Alaska, USA}\label{sec-moose-count}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> sppois <- spglm(
++   formula = count ~ elev + strat,
++   family = poisson,
++   data = moose,
++   spcov_type = "gaussian"
++ )
+R> spnb <- update(sppois, family = nbinomial)
+R> sppois_anis <- update(sppois, anisotropy = TRUE)
+R> spnb_anis <- update(sppois_anis, family = nbinomial)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> BIC(sppois, spnb, sppois_anis, spnb_anis)
+\end{CodeInput}
+\begin{CodeOutput}
+            df      BIC
+sppois       3 1343.892
+spnb         4 1340.706
+sppois_anis  5 1337.610
+spnb_anis    6 1335.051
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot(spnb, which = 8)
+R> plot(spnb_anis, which = 8)
+\end{CodeInput}
+\end{CodeChunk}
+
+\subsection{Modeling harbor seal trends in Alaska, USA}\label{sec-seal}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> spbin <- spgautor(
++   formula = log_trend > 0 ~ 1,
++   family = binomial,
++   data = seal,
++   spcov_type = "car",
++   random = ~ stock
++ )
+R> tidy(spbin, conf.int = TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 1 x 7
+  term        estimate std.error statistic p.value conf.low conf.high
+  <chr>          <dbl>     <dbl>     <dbl>   <dbl>    <dbl>     <dbl>
+1 (Intercept)   -0.340     0.673    -0.506   0.613    -1.66     0.979
+\end{CodeOutput}
+\end{CodeChunk}
+
+\subsection{Modeling voter turnout in Texas, USA}\label{sec-texas}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> spbeta_geo <- spglm(
++   formula = turnout ~ log_income,
++   family = "beta", 
++   data = texas,
++   spcov_type = "matern"
++ )
+R> 
+R> spbeta_auto <- spgautor(
++   formula = turnout ~ log_income,
++   family = "beta", 
++   data = texas,
++   spcov_type = "car",
++   cutoff = 1e5
++ )
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> AIC(spbeta_geo, spbeta_auto)
+\end{CodeInput}
+\begin{CodeOutput}
+            df       AIC
+spbeta_geo   5 -44.53113
+spbeta_auto  3 -22.46104
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> tidy(spbeta_geo)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 2 x 5
+  term        estimate std.error statistic    p.value
+  <chr>          <dbl>     <dbl>     <dbl>      <dbl>
+1 (Intercept)   -5.20      1.11      -4.70 0.00000260
+2 log_income     0.579     0.122      4.76 0.00000194
+\end{CodeOutput}
+\begin{CodeInput}
+R> spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
+R> spbeta_red_ml <- update(spbeta_full_ml, formula = turnout ~ 1)
+R> tidy(anova(spbeta_full_ml, spbeta_red_ml))
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 1 x 5
+  full          reduced           df statistic    p.value
+  <chr>         <chr>          <dbl>     <dbl>      <dbl>
+1 spbeta_red_ml spbeta_full_ml     1      23.2 0.00000149
+\end{CodeOutput}
+\end{CodeChunk}
 
 \subsection{Modeling lake conductivity in Southwest,
-USA}\label{modeling-lake-conductivity-in-southwest-usa}
+USA}\label{sec-lake}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> spgam <- spglm(
++   formula = exp(log_cond) ~ temp * state + origin, 
++   family = "Gamma",
++   data = lake,
++   spcov_type = "cauchy",
++   partition_factor = ~ year
++ )
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> anova(spgam)
+\end{CodeInput}
+\begin{CodeOutput}
+Analysis of Variance Table
+
+Response: exp(log_cond)
+            Df    Chi2 Pr(>Chi2)    
+(Intercept)  1 51.5270 7.062e-13 ***
+temp         1 25.5146 4.390e-07 ***
+state        3  3.0747 0.3802528    
+origin       1  0.1429 0.7053819    
+temp:state   3 19.7668 0.0001897 ***
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> car::vif(spgam)
+\end{CodeInput}
+\begin{CodeOutput}
+                 GVIF Df GVIF^(1/(2*Df))
+temp         4.691914  1        2.166083
+state      127.082397  3        2.242234
+origin       1.264940  1        1.124695
+temp:state  76.387383  3        2.059856
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> pairs(emmeans::emmeans(spgam, ~ state | temp))
+\end{CodeInput}
+\begin{CodeOutput}
+temp = 7.63:
+ contrast estimate    SE  df z.ratio p.value
+ AZ - CO    -1.012 0.337 Inf  -3.004  0.0142
+ AZ - NV    -0.900 0.348 Inf  -2.584  0.0480
+ AZ - UT    -1.331 0.326 Inf  -4.082  0.0003
+ CO - NV     0.112 0.258 Inf   0.434  0.9727
+ CO - UT    -0.319 0.223 Inf  -1.427  0.4822
+ NV - UT    -0.431 0.244 Inf  -1.763  0.2915
+
+Results are averaged over the levels of: origin 
+Degrees-of-freedom method: asymptotic 
+Results are given on the log (not the response) scale. 
+P value adjustment: tukey method for comparing a family of 4 estimates 
+\end{CodeOutput}
+\begin{CodeInput}
+R> emmeans::emtrends(spgam, ~ state, var = "temp")
+\end{CodeInput}
+\begin{CodeOutput}
+ state temp.trend     SE  df asymp.LCL asymp.UCL
+ AZ         0.152 0.0301 Inf    0.0929     0.211
+ CO         0.289 0.0370 Inf    0.2161     0.361
+ NV         0.171 0.0504 Inf    0.0718     0.270
+ UT         0.352 0.0372 Inf    0.2791     0.425
+
+Results are averaged over the levels of: origin 
+Degrees-of-freedom method: asymptotic 
+Results are given on the exp (not the response) scale. 
+Confidence level used: 0.95 
+\end{CodeOutput}
+\end{CodeChunk}
 

### 2025-09-08 inst/manuscript/manuscript.Rmd
some updates

@@
 Non-Gaussian that belong to an exponential family data can be naturally modeled using a generalized linear model (GLM) regression framework [@nelder1972generalized; @mccullagh1989generalized; @myers2012generalized; @faraway2016extending].
-In a GLM, an $n \times 1$ response variable $\mathbf{y}$ belongs to a a statistical distribution (e.g., Poisson, Binomial) with some mean and variance.
+In a GLM, an $n \times 1$ response variable $\mathbf{y}$ belongs to a statistical distribution (e.g., Poisson, Binomial) with some mean and variance.
 Often, the analysis goal is to study the impact of a linear function of several explanatory variables on $\text{y}$ through a GLM.
@@
 While the mean is typically constrained in some way (e.g., between zero and one if a probability), the link function generally makes $\mathbf{w}$ is unconstrained.
-Common link functions inlude the log odds (i.e., logit) link for binary and proprtion data and the log link count and skewed data.
+Common link functions inlude the log odds (i.e., logit) link for binary and proportion data and the log link count and skewed data.
 Equation$~$\ref{eq-glm} can also be written in terms of the inverse link function, $f^{-1}(\cdot)$:
@@
 We assume that $\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}$ and $\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}$, where $\text{E}(\cdot)$ and $\text{Cov}(\cdot)$ denote expectation and covariance, respectively.
-The variance parameter $\sigma^2_\tau$ controls the magnitude of spatial covariance and is often called a partial sill, while the matrix $\mathbf{R}$ is an $n \times n$ spatial correlation matrix that depends on a range parameter controls the distance-decay rate of the spatial correlation.
+The variance parameter $\sigma^2_\tau$ controls the magnitude of spatial covariance and is often called a partial sill, while the matrix $\mathbf{R}$ is an $n \times n$ spatial correlation matrix that depends on a range parameter which controls the distance-decay rate of the spatial correlation.
 One example of a spatial covariance matrix is the "exponential", which is given by
@@
 \end{equation}
-To better align with intuition, we henceforth $\sigma^2_{\tau}$ as $\sigma^2_{de}$ (for spatial error variance) and $\sigma^2_{\epsilon}$ as $\sigma^2_{ie}$ (for independent error variance).
+To better align with intuition, we henceforth $\sigma^2_{\tau}$ as $\sigma^2_{de}$ (for spatially dependent error variance) and $\sigma^2_{\epsilon}$ as $\sigma^2_{ie}$ (for independent error variance).
 The parameters $\sigma^2_{de}$, $\sigma^2_{ie}$, the range parameter $\phi$ in $\mathbf{R}$, and any other parameters in $\mathbf{R}$ compose $\boldsymbol{\theta}$, the covariance parameter vector.
@@
 \pkg{spmodel} also provides tools to predict $\mathbf{w}$ at new locations and quantify uncertainty in those prediction using \code{predict()}.
-This core functionality, combined with several advanced features we describe throughout the manuscript, enable \pkg{spmodel} to provide some novel and important capabilities previously missing from the existing SPGLM ecosystem in \proglang{R}.
+This core functionality, combined with several advanced features we describe throughout the manuscript, enables \pkg{spmodel} to provide some novel and important capabilities previously missing from the existing SPGLM ecosystem in \proglang{R}.
 
@@
 \pkg{spmodel} (version 0.11.0) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in terms of scope and feel.
-Both packages use similar syntax as \code{glm()}, accommodate flexible \code{formula} arguments (e.g., offsets, splines), handle spatial covariance that decays at different rates in different rates (i.e., geometric anisotropy), incorporate nonspatial random effects, support other \proglang{R} packages for modeling like \pkg{broom} [@robinson2021broom; @kuhn2022tidy], \pkg{emmeans} [@lenth2024emmeans], and \pkg{car} [@fox2019car], and have tools for model summaries, prediction, and simulating data.
+Both packages use similar syntax as \code{glm()}, accommodate flexible \code{formula} arguments (e.g., offsets, splines), handle spatial covariance that decays at different rates in different directions (i.e., geometric anisotropy), incorporate nonspatial random effects, support other \proglang{R} packages for modeling like \pkg{broom} [@robinson2021broom; @kuhn2022tidy], \pkg{emmeans} [@lenth2024emmeans], and \pkg{car} [@fox2019car], and have tools for model summaries, prediction, and simulating data.
 There are some notable differences between the two packages, however.
 \pkg{sdmTMB} supports several additional GLM distributions like the Tweedie, supports Hurdle models, and can incorporate prior information through Bayesian applications.
-\pkg{sdmTMB} also provides tools for working with temporal data and enhanced visualizations of marginal effects.
+\pkg{sdmTMB} also provides tools for working with temporal data and spatiotemporal data and provides enhanced visualizations of the model's marginal effects.
 \pkg{sdmTMB} does require a preprocessing step of constructing a mesh for the stochastic partial differential equation approach, and the density of the mesh can affect model results and computational complexity.
-\pkg{spmodel} does not require the construction of a mesh prior to modeling.
+On the other hand, \pkg{spmodel} does not require the construction of a mesh prior to modeling.
 \pkg{spmodel} supports 20 different spatial covariances and models them directly, rather than using a precision matrix approximation to the Matérn spatial covariance as in \pkg{sdmTMB}.
+\pkg{spmodel} can also model areal data directly using neighborhood distance and autoregressive models rather than relying on the polygon centroid (as in \pkg{sdmTMB}).
 \pkg{spmodel} also provides experimental design tools (e.g., analysis of variance, contrasts), supports \pkg{sf} objects in modeling and prediction functions [@pebesma2018sf], has several specialized model diagnostics like leverage values and Cook's distances, and has analytic solutions for prediction standard errors.
@@
 
-Prediction intervals are returned via \code{interval}
+Prediction intervals for the probability of moose presence (on the link scale) are returned via \code{interval}
 ```{r}
@@
 
+In addition to moose presence, moose counts are also recorded in `moose` (Figure$~$\ref{fig-moose-data-count}.
+The Poisson and negative binomial response distributions can be used to model SPGLMs for count data.
+Using a Gaussian spatial covariance function, we may fit both a Poisson and negative binomial SPGLM changing the `family` argument:
+
 ```{r}
@@
 spnb <- update(sppois, family = nbinomial)
-sppois_anis <- update(sppois, anisotropy = TRUE)
-spnb_anis <- update(sppois_anis, family = nbinomial)
 ```
 
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-6.png}
+\caption{Moose counts in Alaska. Circles represent moose counts (based on color) and triangles represent locations at which mean count predictions are desired.}
+\label{fig-moose-data-count}
+\end{figure}
+
+Because the Poisson and negative binomial distributions have the same support (nonnegative integers), we can compare them using  AIC, AICc, or BIC:
+```{r}
+BIC(sppois, spnb)
+```
+
+Implicit in our spatial covariance functions thus far has been an assumption of geometric isotropy.
+A spatial covariance function is geometrically isotropic if it decays with distance at the same rate in all directions (Figure$~$\ref{fig-tropy}; left).
+A spatial covariance is geometrically isotropic if it decays with distance at different rates in different directions (Figure$~$\ref{fig-tropy}; right).
+Geometric anisotropy is formally incorporated by rotating and scaling original coordinates, yielding transformed coordinates that are geometrically isotropic:
+\begin{equation}
+  \begin{bmatrix}
+    x^* \\
+    y^*
+  \end{bmatrix} = 
+  \begin{bmatrix}
+    1 & 0 \\
+    0 & 1 / \omega
+  \end{bmatrix}
+  \begin{bmatrix}
+    \cos(\alpha) & \sin(\alpha) \\
+    -\sin(\alpha) & \cos(\alpha)
+  \end{bmatrix}  
+  \begin{bmatrix}
+    x \\
+    y
+  \end{bmatrix}.
+\end{equation} 
+The parameter $\alpha$ controls the rotation angle and the parameter $\omega$ controls the scaling...
+Then using these transformed coordinates, the partial sill ($\sigma^2_{de}$), nugget ($\sigma^2_{ie}$), and range ($\phi$) parameters are estimated. 
+```{r}
+sppois_anis <- update(sppois, anisotropy = TRUE)
+spnb_anis <- update(spnb, anisotropy = TRUE)
+```
+
+According to BIC, the spatial negative binomial model performs best.
+This model simultaneously accounts for anisotropy and overdispersion.
+The spatial correlation is strongest in a northwest-southeast direction and weakest in the northeast-southwest direction (Figure$~$\ref{fig-tropy}), which is intuitive given the similar patterns in moose counts from Figure$~$\ref{fig-moose-data-count}.
 ```{r}
@@
 
+The `plot()` function can be used to visualize the anisotropy  (Figure$~$\ref{fig-tropy}):
 ```{r, eval = FALSE}
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-7.png}
+\caption{Level curves of equal correlation for the negative binomial moose count models.
+The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of correlation.
+In the isotropic level curve (left), spatial covariance decays equally in all directions.
+In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts).}
+\label{fig-tropy}
+\end{figure}
+
+## Modeling lake conductivity in Southwest, USA {#sec-lake}
+
+The \code{lake} data in \code{spmodel} contains climate and chemical data for several lakes in four southwestern states in the United States: Arizona, Colorado, Nevada, and Utah.
+We desire an SPGLM that characterizes the effect of temperature, state, and lake origin (whether the lake is naturally occurring or human made) on lake conductivity.
+Conductivity is a measure of dissolved ions (measured here in water), which is important for various physical, chemical, and biological processes.
+Chemical data is often heavily right-skewed, so we model it using an SPGLM assuming a Gamma distribution for the response.
+The \code{log_cond} variable in \code{lake} is the logarithm of conductivity, and we can dynamically exponentiate within \code{formula} so that it is on the original scale:
+```{r}
+spgam <- spglm(
+  formula = exp(log_cond) ~ temp + state + temp:state + origin, 
+  family = "Gamma",
+  data = lake,
+  spcov_type = "cauchy",
+  partition_factor = ~ year
+)
+```
+
+We model conductivity as a function of temperature, state, and lake origin, and we allow the effect of temperature to vary by state (\code{temp:state}).
+The partition factor restricts spatial covariance to apply only for two observations sampled during the same year.
+Data were collected in 2012 and 2017, so the \code{year} partition factor assumes independence between observations in 2012 and 2017.
+While we used the partition factor here illustratively, more generally, the utility of partition factors can be highly context dependent.
+
+When categorical variables have more than two levels, the default reference group contrasts are not well-suited to assess the variable's overall significance:
+```{r}
+summary(spgam)
+```
+
+A more effective approach is to use an analysis of variance (ANOVA), which is well-suited to asses the overall significance of each variable:
+```{r}
+anova(spgam)
+```
+
+The main effect for temperature and the temperature by state interaction are highly significant ($p$-value < 0.001), while the main effects for state and lake origin are not significant.
+The ANOVA table can be tidied using \code{tidy()}.
+
+Variance inflation factors assess the degree to which standard errors $\hat{\boldsymbol{\beta}}$ are inflated due to covariance among the columns of $\mathbf{X}$.
+Generalized variance inflation factors can capture the variance inflation for subsets of $\mathbf{X}$ that may include categorical variables with more than two levels [@fox1992generalized]:
+
+```{r}
+library(car)
+vif(spgam)
+```
+
+The GVIF$^{1/2df}$ values for \code{temp}, \code{state}, and \code{temp:state} are just greater than two, which suggests moderate multicollinearity for these terms -- unsurprising given the \code{temp:state} interaction in the model.
+The GVIF$^{1/2df}$ for \code{origin} is close to one, which suggests little to no multicollinearity for this term.
+
+Because of the interaction between \code{temp} and \code{state}, contrasts that assess mean differences among states should condition upon a specific temperature value.
+By default, \pkg{emmeans} uses the mean temperature value to assess contrasts:
+```{r}
+library(emmeans)
+pairs(emmeans(spgam, ~ state | temp))
+```
+
+Again because of the interaction between \code{temp} and \code{state}, we should assess temperature trends separately for each state:
+```{r}
+emtrends(spgam, ~ state, var = "temp")
+```
+
 ## Modeling harbor seal trends in Alaska, USA {#sec-seal}
 
+The \code{seal} data in \pkg{spmodel} contains harbor seal abundance trends for two different harbor seal stocks (genetically distinct populations).
+While the \code{moose} and \code{lake} data were point-referenced, the \code{seal} data are areal.
+Each polygon in the \code{seal} data represents a distinct harbor seal haulout location (Figure$~$\ref{fig-seal}).
+A haulout location is a spot on coastal rocks that harbor seals go to rest, molt, and give birth, among other tasks.
+
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-8.png}
+\caption{Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed log seal trends (right).}
+\label{fig-seal}
+\end{figure}
+
+For each polygon, a Poisson regression was used to quantify the mean trend in abundance over approximately 30 years.
+If the logarithm of seal abundance (\code{log_trend}) is negative (positive), it means abundance is decreasing (increasing).
+We use a binomial SPGLM to quantify how likely it is for abundance trends to be decreasing:
 ```{r}
+is_decreasing <- seal$log_trend < 0
 spbin <- spgautor(
-  formula = log_trend > 0 ~ 1,
+  formula = is_decreasing ~ 1,
   family = binomial,
@@
 )
+```
+
+To model spatial dependence, we used a conditional autoregressive function.
+Conditional and simultaneous autoregressive functions characterize spatial distance through neighborhood relationships and have \code{spcov_type} values of \code{"car"} and \code{"sar"}, respectively.
+By default, Queen's distance is used to determine whether two sites are neighbors, though custom neighborhood matrices are passed via \code{W}.
+Row standardization is also assumed by default, though this can be changed via \code{row_st}.
+Using \code{random}, we also specified a nonspatial random effect for seal stock.
+The \code{random} argument uses similar syntax as \pkg{lme4} [@bates2015lme4] and \pkg{nlme} [@pinheiro2006mixed] to specify nonspatial random effects.
+
+Tidying the model reveals the mean probability that trends are decreasing is between -0.979 and 1.66 (on the logit scale):
+```{r}
 tidy(spbin, conf.int = TRUE)
@@
 
+Back-transforming this interval to the probability scale yields:
+```{r}
+emmeans(spbin, ~ 1, type = "response")
+```
+
+The \code{SE} column is the standard error on the response scale obtained from the delta method [@oehlert1992note].
+
+For areal data, the prediction locations must be specified at the time of model fitting, as they affect the spatial covariance function's neighborhood structure.
+Prediction locations whose response values have an \code{NA} (i.e., missing) value are converted into a \code{newdata} object that is stored in the model output.
+For example, rows one and nine are locations without seal trends, meaning they are not used in model fitting but are desired for prediction:
+```{r}
+seal
+```
+
+Then, \code{predict()} can be called without having to specify \code{newdata}:
+
+```{r}
+predict(spbin, type = "response", interval = "prediction")[1:5, ]
+```
+
+We could have alternatively used a (geostatistical) SPGLM via \code{spglm()}.
+When areal data are used with \code{spglm()}, the centroids of each polygon are used as the point-referenced coordinates.
+We further explore comparisons between point-referenced and aeral data next.
+
 ## Modeling voter turnout in Texas, USA {#sec-texas}
 
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-10.png}
+\caption{Proportion of voter turnout in Texas for the 1980 presidential election.
+Circles represent voter turnout (based on color) and triangles represent locations at which voter turnout predictions are desired.}
+\label{fig-texas}
+\end{figure}
+
+The \code{texas} data in \pkg{spmodel} contains voter turnout data for Texas counties in the 1980 United States Presidential Election [@bivand2024spdata].
+The data are point-referenced, with polygon centroids representing the spatial location of each county (Figure$~$\ref{fig-texas}).
+Beta regression is a GLM used to model rate and proportion data in (0, 1) [@ferrari2004beta, @cribari2010beta].
+We model voter turnout rates as a function of mean log income of county residents using an SPGLM assuming a beta distributed response variable:
 ```{r}
@@
 )
+```
 
+Alternatively, we could use an autoregressive model to fit the model, constructing a neighborhood matrix by assuming centroids within \code{cutoff} of one another are neighbors:
+```{r}
 spbeta_auto <- spgautor(
@@
 
+According to AIC, the SPGLM for point-referenced data is preferred:
 ```{r}
@@
 
+The default estimation method in \pkg{spmodel} is restricted maximum likelihood (REML).
+One drawback of REML is that likelihood-based statistics are only valid for model comparison when the models have the same explanatory variable and fixed effect structure.
+This is because the error contrasts used to construct the REML likelihood change based on $\mathbf{X}$ and $\boldsymbol{\beta}$.
+An alternative is to use maximum likelihood, which can use likelihood-based statistics to compare models with different explanatory variable and fixed effect structures.
+While we could assess the significance of log turnout with the \code{spbeta_geo} model fit with REML (e.g., via \code{tidy()}), an alternative approach is to use a likelihood ratio test as long as the two models are nested.
+We refit the point-referenced model using maximum likelihood and compare it to a similar model without log income using the likelihood ratio test, accessible via \code{anova()} when providing two models:
 ```{r}
-tidy(spbeta_geo)
 spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
 spbeta_red_ml <- update(spbeta_full_ml, formula = turnout ~ 1)
-tidy(anova(spbeta_full_ml, spbeta_red_ml))
+anova(spbeta_full_ml, spbeta_red_ml)
 ```
 
-## Modeling lake conductivity in Southwest, USA {#sec-lake}
+The likelihood ratio test suggests that log income is significantly related to voter turnout. The \code{anova()} output can also be tidied using \code{tidy()}.
 
-```{r}
-spgam <- spglm(
-  formula = exp(log_cond) ~ temp * state + origin, 
-  family = "Gamma",
-  data = lake,
-  spcov_type = "cauchy",
-  partition_factor = ~ year
-)
-```
+SAY SOMETHING ABOUT PREDICT.
 
-```{r}
-anova(spgam)
-```
+# Discussion {#sec-discussion}
 
-```{r}
-car::vif(spgam)
-```
+SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace approximation that marginalizes over the latent (i.e., unobserved) mean, $\mathbf{w}$, and the fixed effects, $\boldsymbol{\beta}$.
+The approach is quite flexible and accommodates any general response distribution and covariance structure.
+\pkg{spmodel}'s \code{spglm()} and \code{spgautor()} functions are similar in structure and syntax as base \proglang{R}'s \code{glm()} function, easing the transition for practitioners from GLMs to SPGLMs.
+These functions support six response distributions for binary, count, and skewed data and support 20 spatial covariance functions.
+\pkg{spmodel} provides several additional features that are not covered here, including fitting multiple models simultaneously, fixing spatial covariance parameters at known values, fitting models to large non-Gaussian data having thousands of observations via spatial indexing [@ver2023indexing], incorporating spatial dependence in machine learning (e.g., random forests; @breiman2001random), simulating spatially dependent data (e.g., \code{spbinom()}, \code{sprpois()}, etc.), and more.
+Further details are provided by [https://CRAN.R-project.org/package=spmodel](https://CRAN.R-project.org/package=spmodel) and links therein.
 
-```{r}
-pairs(emmeans::emmeans(spgam, ~ state | temp))
-emmeans::emtrends(spgam, ~ state, var = "temp")
-```
+# Computational details {.unnumbered}
 
-# Discussion
+The results in this paper were obtained using [R]{.proglang} 4.4.0 with the
+\pkg{spmodel} 0.9.0 package. Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.
+
+# Data and code availability {.unnumbered}
+
+All writing and code associated with this manuscript is available for viewing and download on GitHub at [https://github.com/USEPA/spmodel.glm.manuscript](https://github.com/USEPA/spmodel.glm.manuscript).
+All data used are part of the \pkg{spmodel} [R]{.proglang} package available for download from CRAN at [https://CRAN.R-project.org/package=spmodel](https://CRAN.R-project.org/package=spmodel).
+
+# Acknowledgments {.unnumbered}
+
+We would like to thank initial reviewers and editors for feedback that has greatly improved the manuscript.
+
+The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of the U.S.
+Environmental Protection Agency or the National Oceanic and Atmospheric Administration.
+Any mention of trade names, products, or services does not imply
+an endorsement by the U.S. government, the U.S. Environmental Protection Agency, or the National Oceanic and Atmospheric
+Administration. The U.S. Environmental Protection Agency and the National Oceanic and Atmospheric Administration do not endorse
+any commercial products, services or enterprises.
 

### 2025-09-08 inst/manuscript/manuscript.tex
some updates

@@
 In a GLM, an \(n \times 1\) response variable \(\mathbf{y}\) belongs to
-a a statistical distribution (e.g., Poisson, Binomial) with some mean
-and variance. Often, the analysis goal is to study the impact of a
-linear function of several explanatory variables on \(\text{y}\) through
-a GLM. In this context, the latent (i.e., unobserved) mean of
-\(\mathbf{y}\), \(\boldsymbol{\mu}\), is linked to these explanatory
-variables via a link function: \begin{equation}\label{eq-glm}
+a statistical distribution (e.g., Poisson, Binomial) with some mean and
+variance. Often, the analysis goal is to study the impact of a linear
+function of several explanatory variables on \(\text{y}\) through a GLM.
+In this context, the latent (i.e., unobserved) mean of \(\mathbf{y}\),
+\(\boldsymbol{\mu}\), is linked to these explanatory variables via a
+link function: \begin{equation}\label{eq-glm}
 f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
@@
 generally makes \(\mathbf{w}\) is unconstrained. Common link functions
-inlude the log odds (i.e., logit) link for binary and proprtion data and
-the log link count and skewed data. Equation\(~\)\ref{eq-glm} can also
-be written in terms of the inverse link function, \(f^{-1}(\cdot)\):
-\begin{equation}\label{eq-glm2}
+inlude the log odds (i.e., logit) link for binary and proportion data
+and the log link count and skewed data. Equation\(~\)\ref{eq-glm} can
+also be written in terms of the inverse link function,
+\(f^{-1}(\cdot)\): \begin{equation}\label{eq-glm2}
 \boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}),
@@
 partial sill, while the matrix \(\mathbf{R}\) is an \(n \times n\)
-spatial correlation matrix that depends on a range parameter controls
-the distance-decay rate of the spatial correlation. One example of a
-spatial covariance matrix is the ``exponential'', which is given by
+spatial correlation matrix that depends on a range parameter which
+controls the distance-decay rate of the spatial correlation. One example
+of a spatial covariance matrix is the ``exponential'', which is given by
 \begin{equation}\label{eq-spcov-exp}
@@
 \end{equation} To better align with intuition, we henceforth
-\(\sigma^2_{\tau}\) as \(\sigma^2_{de}\) (for spatial error variance)
-and \(\sigma^2_{\epsilon}\) as \(\sigma^2_{ie}\) (for independent error
-variance). The parameters \(\sigma^2_{de}\), \(\sigma^2_{ie}\), the
-range parameter \(\phi\) in \(\mathbf{R}\), and any other parameters in
-\(\mathbf{R}\) compose \(\boldsymbol{\theta}\), the covariance parameter
-vector.
+\(\sigma^2_{\tau}\) as \(\sigma^2_{de}\) (for spatially dependent error
+variance) and \(\sigma^2_{\epsilon}\) as \(\sigma^2_{ie}\) (for
+independent error variance). The parameters \(\sigma^2_{de}\),
+\(\sigma^2_{ie}\), the range parameter \(\phi\) in \(\mathbf{R}\), and
+any other parameters in \(\mathbf{R}\) compose \(\boldsymbol{\theta}\),
+the covariance parameter vector.
 
@@
 functionality, combined with several advanced features we describe
-throughout the manuscript, enable \pkg{spmodel} to provide some novel
+throughout the manuscript, enables \pkg{spmodel} to provide some novel
 and important capabilities previously missing from the existing SPGLM
@@
 (e.g., offsets, splines), handle spatial covariance that decays at
-different rates in different rates (i.e., geometric anisotropy),
+different rates in different directions (i.e., geometric anisotropy),
 incorporate nonspatial random effects, support other \proglang{R}
@@
 applications. \pkg{sdmTMB} also provides tools for working with temporal
-data and enhanced visualizations of marginal effects. \pkg{sdmTMB} does
-require a preprocessing step of constructing a mesh for the stochastic
-partial differential equation approach, and the density of the mesh can
-affect model results and computational complexity. \pkg{spmodel} does
-not require the construction of a mesh prior to modeling. \pkg{spmodel}
+data and spatiotemporal data and provides enhanced visualizations of the
+model's marginal effects. \pkg{sdmTMB} does require a preprocessing step
+of constructing a mesh for the stochastic partial differential equation
+approach, and the density of the mesh can affect model results and
+computational complexity. On the other hand, \pkg{spmodel} does not
+require the construction of a mesh prior to modeling. \pkg{spmodel}
 supports 20 different spatial covariances and models them directly,
 rather than using a precision matrix approximation to the Matérn spatial
-covariance as in \pkg{sdmTMB}. \pkg{spmodel} also provides experimental
-design tools (e.g., analysis of variance, contrasts), supports \pkg{sf}
-objects in modeling and prediction functions \citep{pebesma2018sf}, has
-several specialized model diagnostics like leverage values and Cook's
-distances, and has analytic solutions for prediction standard errors.
-Other similarities and differences do exist between \pkg{sdmTMB} and
-\pkg{spmodel}, and both packages continue to evolve. Overall, we believe
-that these packages are complementary and enhance the suite of SPGLM
-tools accessible to practitioners.
+covariance as in \pkg{sdmTMB}. \pkg{spmodel} can also model areal data
+directly using neighborhood distance and autoregressive models rather
+than relying on the polygon centroid (as in \pkg{sdmTMB}). \pkg{spmodel}
+also provides experimental design tools (e.g., analysis of variance,
+contrasts), supports \pkg{sf} objects in modeling and prediction
+functions \citep{pebesma2018sf}, has several specialized model
+diagnostics like leverage values and Cook's distances, and has analytic
+solutions for prediction standard errors. Other similarities and
+differences do exist between \pkg{sdmTMB} and \pkg{spmodel}, and both
+packages continue to evolve. Overall, we believe that these packages are
+complementary and enhance the suite of SPGLM tools accessible to
+practitioners.
 
@@
 
-Prediction intervals are returned via \code{interval}
+Prediction intervals for the probability of moose presence (on the link
+scale) are returned via \code{interval}
 
@@
 
+In addition to moose presence, moose counts are also recorded in
+\texttt{moose} (Figure\(~\)\ref{fig-moose-data-count}. The Poisson and
+negative binomial response distributions can be used to model SPGLMs for
+count data. Using a Gaussian spatial covariance function, we may fit
+both a Poisson and negative binomial SPGLM changing the \texttt{family}
+argument:
+
 \begin{CodeChunk}
@@
 R> spnb <- update(sppois, family = nbinomial)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-6.png}
+\caption{Moose counts in Alaska. Circles represent moose counts (based on color) and triangles represent locations at which mean count predictions are desired.}
+\label{fig-moose-data-count}
+\end{figure}
+
+Because the Poisson and negative binomial distributions have the same
+support (nonnegative integers), we can compare them using AIC, AICc, or
+BIC:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> BIC(sppois, spnb)
+\end{CodeInput}
+\begin{CodeOutput}
+       df      BIC
+sppois  3 1343.892
+spnb    4 1340.706
+\end{CodeOutput}
+\end{CodeChunk}
+
+Implicit in our spatial covariance functions thus far has been an
+assumption of geometric isotropy. A spatial covariance function is
+geometrically isotropic if it decays with distance at the same rate in
+all directions (Figure\(~\)\ref{fig-tropy}; left). A spatial covariance
+is geometrically isotropic if it decays with distance at different rates
+in different directions (Figure\(~\)\ref{fig-tropy}; right). Geometric
+anisotropy is formally incorporated by rotating and scaling original
+coordinates, yielding transformed coordinates that are geometrically
+isotropic: \begin{equation}
+  \begin{bmatrix}
+    x^* \\
+    y^*
+  \end{bmatrix} = 
+  \begin{bmatrix}
+    1 & 0 \\
+    0 & 1 / \omega
+  \end{bmatrix}
+  \begin{bmatrix}
+    \cos(\alpha) & \sin(\alpha) \\
+    -\sin(\alpha) & \cos(\alpha)
+  \end{bmatrix}  
+  \begin{bmatrix}
+    x \\
+    y
+  \end{bmatrix}.
+\end{equation} The parameter \(\alpha\) controls the rotation angle and
+the parameter \(\omega\) controls the scaling\ldots{} Then using these
+transformed coordinates, the partial sill (\(\sigma^2_{de}\)), nugget
+(\(\sigma^2_{ie}\)), and range (\(\phi\)) parameters are estimated.
+
+\begin{CodeChunk}
+\begin{CodeInput}
 R> sppois_anis <- update(sppois, anisotropy = TRUE)
-R> spnb_anis <- update(sppois_anis, family = nbinomial)
-\end{CodeInput}
-\end{CodeChunk}
+R> spnb_anis <- update(spnb, anisotropy = TRUE)
+\end{CodeInput}
+\end{CodeChunk}
+
+According to BIC, the spatial negative binomial model performs best.
+This model simultaneously accounts for anisotropy and overdispersion.
+The spatial correlation is strongest in a northwest-southeast direction
+and weakest in the northeast-southwest direction
+(Figure\(~\)\ref{fig-tropy}), which is intuitive given the similar
+patterns in moose counts from Figure\(~\)\ref{fig-moose-data-count}.
 
@@
 
+The \texttt{plot()} function can be used to visualize the anisotropy
+(Figure\(~\)\ref{fig-tropy}):
+
 \begin{CodeChunk}
@@
 
-\subsection{Modeling harbor seal trends in Alaska, USA}\label{sec-seal}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> spbin <- spgautor(
-+   formula = log_trend > 0 ~ 1,
-+   family = binomial,
-+   data = seal,
-+   spcov_type = "car",
-+   random = ~ stock
-+ )
-R> tidy(spbin, conf.int = TRUE)
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 1 x 7
-  term        estimate std.error statistic p.value conf.low conf.high
-  <chr>          <dbl>     <dbl>     <dbl>   <dbl>    <dbl>     <dbl>
-1 (Intercept)   -0.340     0.673    -0.506   0.613    -1.66     0.979
-\end{CodeOutput}
-\end{CodeChunk}
-
-\subsection{Modeling voter turnout in Texas, USA}\label{sec-texas}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> spbeta_geo <- spglm(
-+   formula = turnout ~ log_income,
-+   family = "beta", 
-+   data = texas,
-+   spcov_type = "matern"
-+ )
-R> 
-R> spbeta_auto <- spgautor(
-+   formula = turnout ~ log_income,
-+   family = "beta", 
-+   data = texas,
-+   spcov_type = "car",
-+   cutoff = 1e5
-+ )
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> AIC(spbeta_geo, spbeta_auto)
-\end{CodeInput}
-\begin{CodeOutput}
-            df       AIC
-spbeta_geo   5 -44.53113
-spbeta_auto  3 -22.46104
-\end{CodeOutput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> tidy(spbeta_geo)
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 2 x 5
-  term        estimate std.error statistic    p.value
-  <chr>          <dbl>     <dbl>     <dbl>      <dbl>
-1 (Intercept)   -5.20      1.11      -4.70 0.00000260
-2 log_income     0.579     0.122      4.76 0.00000194
-\end{CodeOutput}
-\begin{CodeInput}
-R> spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
-R> spbeta_red_ml <- update(spbeta_full_ml, formula = turnout ~ 1)
-R> tidy(anova(spbeta_full_ml, spbeta_red_ml))
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 1 x 5
-  full          reduced           df statistic    p.value
-  <chr>         <chr>          <dbl>     <dbl>      <dbl>
-1 spbeta_red_ml spbeta_full_ml     1      23.2 0.00000149
-\end{CodeOutput}
-\end{CodeChunk}
+\begin{figure}[h]
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-7.png}
+\caption{Level curves of equal correlation for the negative binomial moose count models.
+The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of correlation.
+In the isotropic level curve (left), spatial covariance decays equally in all directions.
+In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts).}
+\label{fig-tropy}
+\end{figure}
 
@@
 
+The \code{lake} data in \code{spmodel} contains climate and chemical
+data for several lakes in four southwestern states in the United States:
+Arizona, Colorado, Nevada, and Utah. We desire an SPGLM that
+characterizes the effect of temperature, state, and lake origin (whether
+the lake is naturally occurring or human made) on lake conductivity.
+Conductivity is a measure of dissolved ions (measured here in water),
+which is important for various physical, chemical, and biological
+processes. Chemical data is often heavily right-skewed, so we model it
+using an SPGLM assuming a Gamma distribution for the response. The
+\code{log_cond} variable in \code{lake} is the logarithm of
+conductivity, and we can dynamically exponentiate within \code{formula}
+so that it is on the original scale:
+
 \begin{CodeChunk}
@@
 R> spgam <- spglm(
-+   formula = exp(log_cond) ~ temp * state + origin, 
++   formula = exp(log_cond) ~ temp + state + temp:state + origin, 
 +   family = "Gamma",
@@
 \end{CodeChunk}
+
+We model conductivity as a function of temperature, state, and lake
+origin, and we allow the effect of temperature to vary by state
+(\code{temp:state}). The partition factor restricts spatial covariance
+to apply only for two observations sampled during the same year. Data
+were collected in 2012 and 2017, so the \code{year} partition factor
+assumes independence between observations in 2012 and 2017. While we
+used the partition factor here illustratively, more generally, the
+utility of partition factors can be highly context dependent.
+
+When categorical variables have more than two levels, the default
+reference group contrasts are not well-suited to assess the variable's
+overall significance:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> summary(spgam)
+\end{CodeInput}
+\begin{CodeOutput}
+
+Call:
+spglm(formula = exp(log_cond) ~ temp + state + temp:state + origin, 
+    family = "Gamma", data = lake, spcov_type = "cauchy", partition_factor = ~year)
+
+Deviance Residuals:
+     Min       1Q   Median       3Q      Max 
+-1.35762 -0.20796 -0.03706  0.17869  1.10616
+
+Coefficients (fixed):
+              Estimate Std. Error z value Pr(>|z|)    
+(Intercept)    3.59325    0.50058   7.178 7.06e-13 ***
+temp           0.15182    0.03006   5.051 4.39e-07 ***
+stateCO       -0.03214    0.56098  -0.057  0.95432    
+stateNV        0.75664    0.66851   1.132  0.25771    
+stateUT       -0.19696    0.55916  -0.352  0.72466    
+originNATURAL  0.08313    0.21988   0.378  0.70538    
+temp:stateCO   0.13679    0.04808   2.845  0.00444 ** 
+temp:stateNV   0.01882    0.05820   0.323  0.74645    
+temp:stateUT   0.20015    0.04846   4.131 3.62e-05 ***
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+Pseudo R-squared: 0.7061
+
+Coefficients (cauchy spatial covariance):
+       de        ie     range     extra 
+2.069e-02 2.952e-01 4.119e+06 5.645e-01
+
+Coefficients (Dispersion for Gamma family):
+dispersion 
+     3.761 
+\end{CodeOutput}
+\end{CodeChunk}
+
+A more effective approach is to use an analysis of variance (ANOVA),
+which is well-suited to asses the overall significance of each variable:
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> car::vif(spgam)
+The main effect for temperature and the temperature by state interaction
+are highly significant (\(p\)-value \textless{} 0.001), while the main
+effects for state and lake origin are not significant. The ANOVA table
+can be tidied using \code{tidy()}.
+
+Variance inflation factors assess the degree to which standard errors
+\(\hat{\boldsymbol{\beta}}\) are inflated due to covariance among the
+columns of \(\mathbf{X}\). Generalized variance inflation factors can
+capture the variance inflation for subsets of \(\mathbf{X}\) that may
+include categorical variables with more than two levels
+\citep{fox1992generalized}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library(car)
+\end{CodeInput}
+\begin{CodeOutput}
+Warning: package 'car' was built under R version 4.4.3
+\end{CodeOutput}
+\begin{CodeOutput}
+Loading required package: carData
+\end{CodeOutput}
+\begin{CodeOutput}
+Warning: package 'carData' was built under R version 4.4.3
+\end{CodeOutput}
+\begin{CodeOutput}
+
+Attaching package: 'car'
+\end{CodeOutput}
+\begin{CodeOutput}
+The following object is masked from 'package:dplyr':
+
+    recode
+\end{CodeOutput}
+\begin{CodeInput}
+R> vif(spgam)
 \end{CodeInput}
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> pairs(emmeans::emmeans(spgam, ~ state | temp))
+The GVIF\(^{1/2df}\) values for \code{temp}, \code{state}, and
+\code{temp:state} are just greater than two, which suggests moderate
+multicollinearity for these terms -- unsurprising given the
+\code{temp:state} interaction in the model. The GVIF\(^{1/2df}\) for
+\code{origin} is close to one, which suggests little to no
+multicollinearity for this term.
+
+Because of the interaction between \code{temp} and \code{state},
+contrasts that assess mean differences among states should condition
+upon a specific temperature value. By default, \pkg{emmeans} uses the
+mean temperature value to assess contrasts:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library(emmeans)
+\end{CodeInput}
+\begin{CodeOutput}
+Welcome to emmeans.
+Caution: You lose important information if you filter this package's results.
+See '? untidy'
+\end{CodeOutput}
+\begin{CodeInput}
+R> pairs(emmeans(spgam, ~ state | temp))
 \end{CodeInput}
@@
 \end{CodeOutput}
-\begin{CodeInput}
-R> emmeans::emtrends(spgam, ~ state, var = "temp")
+\end{CodeChunk}
+
+Again because of the interaction between \code{temp} and \code{state},
+we should assess temperature trends separately for each state:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> emtrends(spgam, ~ state, var = "temp")
 \end{CodeInput}
@@
 
-\section{Discussion}\label{discussion}
+\subsection{Modeling harbor seal trends in Alaska, USA}\label{sec-seal}
+
+The \code{seal} data in \pkg{spmodel} contains harbor seal abundance
+trends for two different harbor seal stocks (genetically distinct
+populations). While the \code{moose} and \code{lake} data were
+point-referenced, the \code{seal} data are areal. Each polygon in the
+\code{seal} data represents a distinct harbor seal haulout location
+(Figure\(~\)\ref{fig-seal}). A haulout location is a spot on coastal
+rocks that harbor seals go to rest, molt, and give birth, among other
+tasks.
+
+\begin{figure}
+\centering
+\includegraphics[width = 1\linewidth]{figures/figure-8.png}
+\caption{Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed log seal trends (right).}
+\label{fig-seal}
+\end{figure}
+
+For each polygon, a Poisson regression was used to quantify the mean
+trend in abundance over approximately 30 years. If the logarithm of seal
+abundance (\code{log_trend}) is negative (positive), it means abundance
+is decreasing (increasing). We use a binomial SPGLM to quantify how
+likely it is for abundance trends to be decreasing:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> is_decreasing <- seal$log_trend < 0
+R> spbin <- spgautor(
++   formula = is_decreasing ~ 1,
++   family = binomial,
++   data = seal,
++   spcov_type = "car",
++   random = ~ stock
++ )
+\end{CodeInput}
+\end{CodeChunk}
+
+To model spatial dependence, we used a conditional autoregressive
+function. Conditional and simultaneous autoregressive functions
+characterize spatial distance through neighborhood relationships and
+have \code{spcov_type} values of \code{"car"} and \code{"sar"},
+respectively. By default, Queen's distance is used to determine whether
+two sites are neighbors, though custom neighborhood matrices are passed
+via \code{W}. Row standardization is also assumed by default, though
+this can be changed via \code{row_st}. Using \code{random}, we also
+specified a nonspatial random effect for seal stock. The \code{random}
+argument uses similar syntax as \pkg{lme4} \citep{bates2015lme4} and
+\pkg{nlme} \citep{pinheiro2006mixed} to specify nonspatial random
+effects.
+
+Tidying the model reveals the mean probability that trends are
+decreasing is between -0.979 and 1.66 (on the logit scale):
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> tidy(spbin, conf.int = TRUE)
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 1 x 7
+  term        estimate std.error statistic p.value conf.low conf.high
+  <chr>          <dbl>     <dbl>     <dbl>   <dbl>    <dbl>     <dbl>
+1 (Intercept)    0.340     0.673     0.506   0.613   -0.979      1.66
+\end{CodeOutput}
+\end{CodeChunk}
+
+Back-transforming this interval to the probability scale yields:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> emmeans(spbin, ~ 1, type = "response")
+\end{CodeInput}
+\begin{CodeOutput}
+ 1        prob    SE  df asymp.LCL asymp.UCL
+ overall 0.584 0.164 Inf     0.273      0.84
+
+Degrees-of-freedom method: asymptotic 
+Confidence level used: 0.95 
+Intervals are back-transformed from the logit scale 
+\end{CodeOutput}
+\end{CodeChunk}
+
+The \code{SE} column is the standard error on the response scale
+obtained from the delta method \citep{oehlert1992note}.
+
+For areal data, the prediction locations must be specified at the time
+of model fitting, as they affect the spatial covariance function's
+neighborhood structure. Prediction locations whose response values have
+an \code{NA} (i.e., missing) value are converted into a \code{newdata}
+object that is stored in the model output. For example, rows one and
+nine are locations without seal trends, meaning they are not used in
+model fitting but are desired for prediction:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> seal
+\end{CodeInput}
+\begin{CodeOutput}
+Simple feature collection with 149 features and 2 fields
+Geometry type: POLYGON
+Dimension:     XY
+Bounding box:  xmin: 913618.8 ymin: 855730.2 xmax: 1221859 ymax: 1145054
+Projected CRS: NAD83 / Alaska Albers
+# A tibble: 149 x 3
+   log_trend stock                                                      geometry
+ *     <dbl> <fct>                                                 <POLYGON [m]>
+ 1  NA       8     ((1035002 1054710, 1035002 1054542, 1035002 1053542, 1035002~
+ 2  -0.282   8     ((1037002 1039492, 1037006 1039490, 1037017 1039492, 1037035~
+ 3  -0.00121 8     ((1070158 1030216, 1070185 1030207, 1070187 1030207, 1070211~
+ 4   0.0354  8     ((1054906 1034826, 1054931 1034821, 1054936 1034822, 1055001~
+ 5  -0.0160  8     ((1025142 1056940, 1025184 1056889, 1025222 1056836, 1025256~
+ 6   0.0872  8     ((1026035 1044623, 1026037 1044605, 1026072 1044610, 1026083~
+ 7  -0.266   8     ((1100345 1060709, 1100287 1060706, 1100228 1060706, 1100170~
+ 8   0.0743  8     ((1030247 1029637, 1030248 1029637, 1030265 1029642, 1030328~
+ 9  NA       8     ((1043093 1020553, 1043097 1020550, 1043101 1020550, 1043166~
+10  -0.00961 8     ((1116002 1024542, 1116002 1023542, 1116002 1022542, 1116002~
+# i 139 more rows
+\end{CodeOutput}
+\end{CodeChunk}
+
+Then, \code{predict()} can be called without having to specify
+\code{newdata}:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> predict(spbin, type = "response", interval = "prediction")[1:5, ]
+\end{CodeInput}
+\begin{CodeOutput}
+         fit       lwr       upr
+1  0.6807677 0.3863736 0.8783808
+9  0.5945680 0.2467634 0.8678078
+13 0.6189055 0.2974432 0.8616799
+15 0.6040102 0.2921802 0.8493132
+18 0.6375700 0.3356282 0.8596641
+\end{CodeOutput}
+\end{CodeChunk}
+
+We could have alternatively used a (geostatistical) SPGLM via
+\code{spglm()}. When areal data are used with \code{spglm()}, the
+centroids of each polygon are used as the point-referenced coordinates.
+We further explore comparisons between point-referenced and aeral data
+next.
+
+\subsection{Modeling voter turnout in Texas, USA}\label{sec-texas}
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.70\linewidth]{figures/figure-10.png}
+\caption{Proportion of voter turnout in Texas for the 1980 presidential election.
+Circles represent voter turnout (based on color) and triangles represent locations at which voter turnout predictions are desired.}
+\label{fig-texas}
+\end{figure}
+
+The \code{texas} data in \pkg{spmodel} contains voter turnout data for
+Texas counties in the 1980 United States Presidential Election
+\citep{bivand2024spdata}. The data are point-referenced, with polygon
+centroids representing the spatial location of each county
+(Figure\(~\)\ref{fig-texas}). Beta regression is a GLM used to model
+rate and proportion data in (0, 1)
+\citep[\citet{cribari2010beta}]{ferrari2004beta}. We model voter turnout
+rates as a function of mean log income of county residents using an
+SPGLM assuming a beta distributed response variable:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> spbeta_geo <- spglm(
++   formula = turnout ~ log_income,
++   family = "beta", 
++   data = texas,
++   spcov_type = "matern"
++ )
+\end{CodeInput}
+\end{CodeChunk}
+
+Alternatively, we could use an autoregressive model to fit the model,
+constructing a neighborhood matrix by assuming centroids within
+\code{cutoff} of one another are neighbors:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> spbeta_auto <- spgautor(
++   formula = turnout ~ log_income,
++   family = "beta", 
++   data = texas,
++   spcov_type = "car",
++   cutoff = 1e5
++ )
+\end{CodeInput}
+\end{CodeChunk}
+
+According to AIC, the SPGLM for point-referenced data is preferred:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> AIC(spbeta_geo, spbeta_auto)
+\end{CodeInput}
+\begin{CodeOutput}
+            df       AIC
+spbeta_geo   5 -44.53113
+spbeta_auto  3 -22.46104
+\end{CodeOutput}
+\end{CodeChunk}
+
+The default estimation method in \pkg{spmodel} is restricted maximum
+likelihood (REML). One drawback of REML is that likelihood-based
+statistics are only valid for model comparison when the models have the
+same explanatory variable and fixed effect structure. This is because
+the error contrasts used to construct the REML likelihood change based
+on \(\mathbf{X}\) and \(\boldsymbol{\beta}\). An alternative is to use
+maximum likelihood, which can use likelihood-based statistics to compare
+models with different explanatory variable and fixed effect structures.
+While we could assess the significance of log turnout with the
+\code{spbeta_geo} model fit with REML (e.g., via \code{tidy()}), an
+alternative approach is to use a likelihood ratio test as long as the
+two models are nested. We refit the point-referenced model using maximum
+likelihood and compare it to a similar model without log income using
+the likelihood ratio test, accessible via \code{anova()} when providing
+two models:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
+R> spbeta_red_ml <- update(spbeta_full_ml, formula = turnout ~ 1)
+R> anova(spbeta_full_ml, spbeta_red_ml)
+\end{CodeInput}
+\begin{CodeOutput}
+Likelihood Ratio Test
+
+Response: turnout
+                                Df   Chi2 Pr(>Chi2)    
+spbeta_red_ml vs spbeta_full_ml  1 23.155 1.494e-06 ***
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{CodeOutput}
+\end{CodeChunk}
+
+The likelihood ratio test suggests that log income is significantly
+related to voter turnout. The \code{anova()} output can also be tidied
+using \code{tidy()}.
+
+\section{Discussion}\label{sec-discussion}
+
+SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace
+approximation that marginalizes over the latent (i.e., unobserved) mean,
+\(\mathbf{w}\), and the fixed effects, \(\boldsymbol{\beta}\). The
+approach is quite flexible and accommodates any general response
+distribution and covariance structure. \pkg{spmodel}'s \code{spglm()}
+and \code{spgautor()} functions are similar in structure and syntax as
+base \proglang{R}'s \code{glm()} function, easing the transition for
+practitioners from GLMs to SPGLMs. These functions support six response
+distributions for binary, count, and skewed data and support 20 spatial
+covariance functions. \pkg{spmodel} provides several additional features
+that are not covered here, including fitting multiple models
+simultaneously, fixing spatial covariance parameters at known values,
+fitting models to large non-Gaussian data having thousands of
+observations via spatial indexing \citep{ver2023indexing}, incorporating
+spatial dependence in machine learning (e.g., random forests;
+\citet{breiman2001random}), simulating spatially dependent data (e.g.,
+\code{spbinom()}, \code{sprpois()}, etc.), and more. Further details are
+provided by \url{https://CRAN.R-project.org/package=spmodel} and links
+therein.
+
+\section*{Computational details}\label{computational-details}
+\addcontentsline{toc}{section}{Computational details}
+
+The results in this paper were obtained using {R} 4.4.0 with the
+\pkg{spmodel} 0.9.0 package. Figures were created using the {ggplot2}
+3.5.1 package \citep{wickham2016ggplot2} and base {R}.
+
+\section*{Data and code availability}\label{data-and-code-availability}
+\addcontentsline{toc}{section}{Data and code availability}
+
+All writing and code associated with this manuscript is available for
+viewing and download on GitHub at
+\url{https://github.com/USEPA/spmodel.glm.manuscript}. All data used are
+part of the \pkg{spmodel} {R} package available for download from CRAN
+at \url{https://CRAN.R-project.org/package=spmodel}.
+
+\section*{Acknowledgments}\label{acknowledgments}
+\addcontentsline{toc}{section}{Acknowledgments}
+
+We would like to thank initial reviewers and editors for feedback that
+has greatly improved the manuscript.
+
+The views expressed in this manuscript are those of the authors and do
+not necessarily represent the views or policies of the U.S.
+Environmental Protection Agency or the National Oceanic and Atmospheric
+Administration. Any mention of trade names, products, or services does
+not imply an endorsement by the U.S. government, the U.S. Environmental
+Protection Agency, or the National Oceanic and Atmospheric
+Administration. The U.S. Environmental Protection Agency and the
+National Oceanic and Atmospheric Administration do not endorse any
+commercial products, services or enterprises.
 

### 2025-09-08 inst/manuscript/figures/figures.R

@@ -229,7 +229,7 @@ dev.off()
 ########################
 ####### Figure 8
 ########################
-seal$trend <- exp(seal$log_trend)
+seal$trend <- seal$log_trend
 seal$sampled <- factor(if_else(is.na(seal$trend), "no", "yes"), levels = c("yes", "no"))
 p1 <- ggplot(seal, aes(fill = sampled, color = stock)) +
   geom_sf(size = 2) +
@@ -242,7 +242,7 @@ p1 <- ggplot(seal, aes(fill = sampled, color = stock)) +
 p2 <- ggplot(seal, aes(fill = trend)) +
   geom_sf(size = 2) +
   theme_bw(base_size = 14) +
-  scale_fill_viridis_c(option = "D", begin = 0.5, limits = c(0, max(seal$trend))) +
+  scale_fill_viridis_c(option = "D", begin = 0.5, limits = c(min(seal$trend), max(seal$trend))) +
   scale_y_continuous(breaks = seq(57, 59, length.out = 5)) +
   scale_x_continuous(breaks = seq(-139, -134, length.out = 3))
 

### 2025-09-10 inst/manuscript/manuscript.Rmd
some updates

@@
 abstract: |
-  Generalized linear models (GLMs) describe a non-normal response variable that may be binary, count, skewed, or a proportion.
-Typically, observations in a GLM are assumed independent of one another.
-For spatial data, this independence assumption is impractical, as nearby locations tend to be more similar than locations far apart.
-The \pkg{spmodel} \proglang{R} package provides tools to fit GLMs that incorporate spatial correlation (i.e., spatial generalized linear models, or SPGLMs).
-SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace approximation via \code{spglm()} for point-referenced data or \code{spgautor()} for areal (i.e., lattice), data.
-\code{spglm()} and \code{spgautor()} closely resemble [glm]{.fct} from base \proglang{R} but include arguments that control the spatial correlation structure.
-\pkg{spmodel} has many helper functions for model inspection and diagnostics, some of which leverage other \proglang{R} packages like [broom]{.pkg} and [emmeans]{.pkg}.
-\pkg{spmodel} has tools to make predictions of the latent spatial-mean process at unobserved locations.
-\pkg{spmodel} also provides many advanced features like accommodating geometric anisotropy and nonspatial random effects, simulating spatially autocorrelated data, and more.
-Here we use \pkg{spmodel} to illustrate the modeling of binary, count, skewed and proportion response variables from several point-referenced and areal data sets.
+  Non-Gaussian data are common in practice and include binary, count, skewed, and proprtion data types.
+Often, non-Gaussian data are modeled using a generalized linear model (GLM).
+GLMs typically assume that observations are independent of one another.
+This is an impractical assumption for spatial data, as nearby observations tend to be more similar than distant ones.
+The \pkg{spmodel} package in \proglang{R} provides a suite of tools for fitting spatial generalized linear models (SPGLMs) to non-Gaussian data and making spatial predictions (i.e., Kriging).
+SPGLMs for point-referenced data (x- and y-coordinates) are fit using the \code{spglm()} function, while SPGLMs for areal (lattice, polygon) data are fit using the \code{spgautor()} function.
+Both \code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood which marginalizes over the model's fixed effects and latent mean while formally incorporating spatial covariance (i.e., spatial correlation).
+The inputs and outputs of \code{spglm()} and \code{spgautor()} closely resemble the \code{glm()} function from base \proglang{R}, easing the transition from GLMs to SPGLMs.
+\pkg{spmodel} provides and builds upon several commonly used helper functions for model building like \code{summary()}, \code{plot()}, \code{fitted()}, and \code{tidy()}, among others.
+Spatial predictions of the latent mean at unobserved locations are obtained using \code{predict()} or \code{augment()}.
+\pkg{spmodel} accommodates myriad advanced modeling features like geometric anisotrpoy, nonspatial random effects, analysis of variance, and more.
+Throughout, we use \pkg{spmodel} to fit SPGLMs to moose presence and counts in Alaska, United States (US), skewed conductivity data in the Southwestern US, harbor seal abundance trends in Alaska, US, and voter turnout rates in Texas, US.
 keywords:
@@
   \usepackage{caption, subcaption}
+  \usepackage{autonum}
 output: rticles::jss_article
@@
 ```{r, setup, include=FALSE}
-options(prompt = 'R> ', continue = '+ ')
+options(prompt = 'R> ', continue = '+ ', width = 70)
 library(spmodel)
@@
 
+\linenumbers
+
 \newpage
@@
 
-In practice, non-Gaussian data are ubiquitous.
-Non-Gaussian that belong to an exponential family data can be naturally modeled using a generalized linear model (GLM) regression framework [@nelder1972generalized; @mccullagh1989generalized; @myers2012generalized; @faraway2016extending].
-In a GLM, an $n \times 1$ response variable $\mathbf{y}$ belongs to a statistical distribution (e.g., Poisson, Binomial) with some mean and variance.
-Often, the analysis goal is to study the impact of a linear function of several explanatory variables on $\text{y}$ through a GLM.
+In practice, non-Gaussian data (e.g., binary, count, skewed, and proportion data) are ubiquitous.
+Non-Gaussian that belong to an exponential family data can be naturally modeled using a generalized linear model (GLM) regression framework [@nelder1972generalized; @mccullagh1989generalized].
+In a GLM, an $n \times 1$ response variable $\mathbf{y}$ belongs to a statistical distribution (e.g., Binomial, Poisson) with some mean and variance.
+Often, the analysis goal is to study the impact of a linear function of several explanatory variables on $\mathbf{y}$ through a GLM.
 In this context, the latent (i.e., unobserved) mean of $\mathbf{y}$, $\boldsymbol{\mu}$, is linked to these explanatory variables via a link function:
-\begin{equation}\label{eq-glm}
+\begin{align}\label{eq-glm}
 f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
-\end{equation}
+\end{align}
 where for a sample size $n$, $f(\cdot)$ is a link function that connects $\boldsymbol{\mu}$ to $\mathbf{w}$, $\mathbf{X}$ is the $n \times p$ design matrix of explanatory variables, and $\boldsymbol{\beta}$ is the $p \times 1$ vector of fixed effects.
-While the mean is typically constrained in some way (e.g., between zero and one if a probability), the link function generally makes $\mathbf{w}$ is unconstrained.
-Common link functions inlude the log odds (i.e., logit) link for binary and proportion data and the log link count and skewed data.
+While the mean is typically constrained in some way (e.g., if a probability, between zero and one), the link function generally makes $\mathbf{w}$ unconstrained.
+Common link functions include the log odds (i.e., logit) link for binary and proportion data and the log link for count and skewed data.
 Equation$~$\ref{eq-glm} can also be written in terms of the inverse link function, $f^{-1}(\cdot)$:
-\begin{equation}\label{eq-glm2}
-\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}),
-\end{equation}
+\begin{align}\label{eq-glm2}
+\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}).
+\end{align}
 
@@
 This independence assumption is typically impractical for spatial data.
-In spatial data, nearby observations tend to be more similar than distant observations [@tobler1970computer], leading to positive spatial covariance among observations.
+For spatial data, nearby observations tend to be more similar than distant observations [@tobler1970computer], which leads to positive spatial covariance.
 The consequences of ignoring spatial covariance in statistical models for spatial data can be severe and include imprecise parameter estimates as well as misleading standard errors that inflate Type-I error rates and decrease power [@zimmerman2024spatial].
 
-An approach for handling spatial data using a GLM is to assume $\mathbf{w}$ has spatial covariance.
+An approach for handling spatial data using a GLM is to assume the elements of $\mathbf{w}$ share covariance that varies spatially.
 This is achieved by adding to Equation$~$\ref{eq-glm} two random effects, $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$.
@@
 We assume that $\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}$ and $\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}$, where $\text{E}(\cdot)$ and $\text{Cov}(\cdot)$ denote expectation and covariance, respectively.
-The variance parameter $\sigma^2_\tau$ controls the magnitude of spatial covariance and is often called a partial sill, while the matrix $\mathbf{R}$ is an $n \times n$ spatial correlation matrix that depends on a range parameter which controls the distance-decay rate of the spatial correlation.
+The variance parameter $\sigma^2_\tau$ controls the magnitude of spatial covariance and is often called a partial sill.
+The matrix $\mathbf{R}$ is an $n \times n$ spatial correlation matrix that depends on a range parameter controlling the distance-decay rate of the spatial correlation.
 One example of a spatial covariance matrix is the "exponential", which is given by
-\begin{equation}\label{eq-spcov-exp}
-  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de} \exp(-\mathbf{H}/\phi),
-\end{equation}
-where $\mathbf{H}$ is a matrix of pairwise distances among the elements of $\mathbf{y}$ and $\phi$ is a range parameter.
+\begin{align}\label{eq-spcov-exp}
+  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de} \mathbf{R}_{exp} =  \sigma^2_{de} \exp(-\mathbf{H}/\phi),
+\end{align}
+where $\mathbf{H}$ is a matrix of pairwise distances among the elements of $\mathbf{y}$ and $\phi$ is the range parameter.
 From Equation$~$\ref{eq-spcov-exp}, as the distance between two elements of $\mathbf{y}$ increases, the spatial covariance decreases, which reflects intuition.
@@
 The variance parameter $\sigma^2_\epsilon$ controls the magnitude of nonspatial variability (i.e., fine-scale variation) and is often called a nugget.
+Often in spatial statistics, quantities are explicitly referenced with respect to $\mathbf{s}$, a vector of coordinates indexing the observation [@cressie1993statistics].
+For example, $\mathbf{y}$ and $\mathbf{X}$ may instead be written $\mathbf{y}(\mathbf{s})$ and $\mathbf{X}$, respectively.
+We acknowledge the utility of this nomenclature but drop the explicit dependence on $\mathbf{s}$ for simplicity of notation.
 
@@
 Through inclusion of $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$, the spatial GLM (SPGLM) can be written as
-\begin{equation}\label{eq-spglm}
+\begin{align}\label{eq-spglm}
 f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
-\end{equation}
-Often in spatial statistics, quantities are explicitly referenced with respect to $\mathbf{s}$, a vector of coordinates indexing the observation [@cressie1993statistics].
-For example, $\mathbf{y}$ and $\mathbf{X}$ may instead be written $\mathbf{y}(\mathbf{s})$ and $\mathbf{X}$, respectively.
-We acknowledge the utility of this nomenclature but drop the explicit dependence on $\mathbf{s}$ for simplicity of notation.
+\end{align}
 Assuming independence among $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$, it follows that 
-\begin{equation}\label{eq-spcov}
+\begin{align}\label{eq-spcov}
  \text{Cov}(\boldsymbol{\tau} + \boldsymbol{\epsilon}) = \text{Cov}(\boldsymbol{\tau}) + \text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{\tau}\mathbf{R} + \sigma^2_{\epsilon} \mathbf{I}.
-\end{equation}
-To better align with intuition, we henceforth $\sigma^2_{\tau}$ as $\sigma^2_{de}$ (for spatially dependent error variance) and $\sigma^2_{\epsilon}$ as $\sigma^2_{ie}$ (for independent error variance).
-The parameters $\sigma^2_{de}$, $\sigma^2_{ie}$, the range parameter $\phi$ in $\mathbf{R}$, and any other parameters in $\mathbf{R}$ compose $\boldsymbol{\theta}$, the covariance parameter vector.
+\end{align}
+Henceforth, we refer to $\sigma^2_{\tau}$ as $\sigma^2_{de}$ (for spatially dependent error variance) and $\sigma^2_{\epsilon}$ as $\sigma^2_{ie}$ (for independent error variance).
+The parameters $\sigma^2_{de}$, $\sigma^2_{ie}$, and $\phi$, in addition to any other parameters in $\mathbf{R}$, compose $\boldsymbol{\theta}$, the covariance parameter vector.
 
@@
 Working with precision matrices, however, can be more restrictive and less intuitive than working directly with the covariance matrix.
-The [FRK]{.pkg} [@sainsbury2024modeling], \pkg{glmmTMB} [@brooks2017glmmtmb], \pkg{hglm} [@ronnegard2010hglm], \pkg{mgcv} [@wood2017generalized], and \pkg{spaMM} [@rousset2014spamm] packages directly use Laplace, quasi-likelihood, or reduced-rank approaches to estimate parameters.
+The \pkg{FRK} [@sainsbury2024modeling], \pkg{glmmTMB} [@brooks2017glmmtmb], \pkg{hglm} [@ronnegard2010hglm], \pkg{mgcv} [@wood2017generalized], and \pkg{spaMM} [@rousset2014spamm] packages directly use Laplace, quasi-likelihood, or reduced-rank approaches to estimate parameters.
 These direct approaches tend to be computationally efficient, as they don't rely on MCMC sampling.
 In contrast to the Bayesian approach, a drawback of these direct approaches is that prior information cannot be formally incorporated and covariance parameter uncertainty is more challenging to quantify.
-The \pkg{sdmTMB} [@anderson2024sdmtmb] package combines elements of \pkg{R-INLA}, \pkg{glmmTMB}, and properties of Gaussian Markov random fields to fit a wide variety of SPGLMs, and \pkg{tinyVAST} [@thorson2025tinyVAST] extends some of these models to multivariate or (dynamic) structural equation models.
+The \pkg{sdmTMB} [@anderson2024sdmtmb] package combines elements of \pkg{R-INLA}, \pkg{glmmTMB}, and Gaussian Markov random fields to fit a wide variety of SPGLMs, while \pkg{tinyVAST} [@thorson2025tinyVAST] extends some of these models to multivariate or (dynamic) structural equation models.
 
-@ver2024marginal proposed a novel approach to fitting SPGLMs that leverages the Laplace approximation while marginalizing over both the latent $\mathbf{w}$ and the fixed effects ($\boldsymbol{\beta}$) and accommodating spatial covariance.
-@ver2024marginal showed that this approach performed efficiently in a variety of simulation settings, generally having appropriate confidence interval coverage for the fixed effects and prediction interval coverage for new $\mathbf{w}$.
+Building from @evangelou2011estimation and @bonat2016practical, @ver2024marginal proposed a novel approach for fitting SPGLMs that leverages the Laplace approximation while marginalizing over both the latent $\mathbf{w}$ and the fixed effects ($\boldsymbol{\beta}$) and accommodating spatial covariance.
+This approach performed efficiently in a variety of simulation settings, generally having appropriate confidence interval coverage for the fixed effects and prediction interval coverage for $\mathbf{w}$ at new locations.
 The approach performed similarly to the Bayesian SPGLM approach in \pkg{spBayes} and the automatic differentiation SPGLM approach in \pkg{glmmTMB} but was much faster.
@@
 For moderate sample sizes, it performed similarly to \pkg{R-INLA}, though \pkg{R-INLA} was faster.
-This novel approach is particularly attractive for two reasons.
+The novel Laplace approach is particularly attractive for two reasons.
 First, it is general enough that can be applied to any covariance structure (not just spatial).
 Second, after estimating the covariance parameters, analytical solutions exist for the fixed effects (and their standard errors) as well as predictions of the latent $\mathbf{w}$ at new locations (and their standard errors).
-The \pkg{spmodel} \proglang{R} package [@dumelle2023spmodel] recently provided full support for the methods in @ver2024marginal applied to binary, count, skewed, and proportion data for over 20 different spatial covariance types.
 
-The \pkg{spmodel} \proglang{R} package [@dumelle2023spmodel] recently provided a full set of modeling tools for SPGLMs fit using the methods described in @ver2024marginal.
+The \pkg{spmodel} \proglang{R} package [@dumelle2023spmodel] recently provided a full set of modeling tools for SPGLMs fit using the novel Laplace approach described by @ver2024marginal.
 These modeling tools are approachable and mirror the familiar \code{glm()} syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs relatively seamless.
-The \code{spglm()} function fits SPGLMs for point-referenced data (e.g., x-coordinates and y-coordinates representing point locations in a field), while the \code{spgautor()} function fits SPGLMs for areal data (e.g., polygon boundaries representing geographic subsets of a region).
+The \code{spglm()} function fits SPGLMs for point-referenced data (e.g., x-coordinates and y-coordinates representing point locations in a field; models are sometimes called "geostatistical" models), while the \code{spgautor()} function fits SPGLMs for areal data (e.g., polygon boundaries representing geographic subsets of a region; models are sometimes called "autoregressive" models).
 \pkg{spmodel} supports the binomial distribution for binary data, Poisson and negative binomial distributions for count data, Gamma and inverse Gaussian distributions for skewed data, and the beta distribution for proportion data.
 There are 20 different spatial covariance structures available including the exponential, Gaussian, and spherical for point-referenced data (Figure$~$\ref{fig-type}) and the conditional autoregressive, and simultaneous autoregressive structures for areal data.
-\pkg{spmodel} provides tools for commonly used model summaries, visualizations, and diagnostics (e.g., Cook's distance) using standard \proglang{R} helper functions like \code{summary()}, \code{plot()}, and \code{cooks.distance()}.
-\pkg{spmodel} also provides tools to predict $\mathbf{w}$ at new locations and quantify uncertainty in those prediction using \code{predict()}.
-This core functionality, combined with several advanced features we describe throughout the manuscript, enables \pkg{spmodel} to provide some novel and important capabilities previously missing from the existing SPGLM ecosystem in \proglang{R}.
+\pkg{spmodel} provides tools for commonly used model summaries, visualizations, and diagnostics (e.g., Cook's distance) using standard \proglang{R} helper functions like \code{summary()}, \code{plot()}, \code{fitted()}, and \code{tidy()}, among othres.
+\pkg{spmodel} also provides tools to predict $\mathbf{w}$ at new locations and quantify uncertainty in those prediction using \code{predict()} and \code{augment()}.
+This core functionality, combined with several advanced features we describe throughout the manuscript, enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools previously missing from the existing \proglang{R} ecosystem.
 
@@
 On the other hand, \pkg{spmodel} does not require the construction of a mesh prior to modeling.
-\pkg{spmodel} supports 20 different spatial covariances and models them directly, rather than using a precision matrix approximation to the Matérn spatial covariance as in \pkg{sdmTMB}.
-\pkg{spmodel} can also model areal data directly using neighborhood distance and autoregressive models rather than relying on the polygon centroid (as in \pkg{sdmTMB}).
-\pkg{spmodel} also provides experimental design tools (e.g., analysis of variance, contrasts), supports \pkg{sf} objects in modeling and prediction functions [@pebesma2018sf], has several specialized model diagnostics like leverage values and Cook's distances, and has analytic solutions for prediction standard errors.
+\pkg{spmodel} also supports 20 different spatial covariances and models them directly, rather than using a precision matrix approximation to the Matérn spatial covariance as in \pkg{sdmTMB}.
+\pkg{spmodel} can also model areal data directly using neighborhood distance and autoregressive models rather than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not be within the polygon's boundaries.
+\pkg{spmodel} also provides experimental design tools (e.g., analysis of variance, contrasts), supports \pkg{sf} objects in modeling and prediction functions [@pebesma2018sf], has several specialized model diagnostics like leverage values and Cook's distances, and has analytic solutions for fixed effect and prediction standard errors.
 Other similarities and differences do exist between \pkg{sdmTMB} and \pkg{spmodel}, and both packages continue to evolve.
@@
 In Section$~$\ref{sec-spglm}, we provide some background for the SPGLM fitting and prediction routines in \pkg{spmodel}.
-In Section$~$\ref{sec-applications}, we provide several applications of \pkg{spmodel} to spatial binary, count, skewed and proportion data with both point-referenced and areal supports.
+In Section$~$\ref{sec-applications}, we provide an overview of core SPGLM functionaility in \pkg{spmodel} by modeling moose presence in Alaska, United States (US).
+In Section$~$\ref{sec-applications2}, we model moose counts in Alaska, US; skewed lake conductivity in the Southwestern US; harbor seal abundance trend behavior in Alaska, US; and voter turnout rates in Texas, US.
 And in Section$~$\ref{sec-discussion}, we end with a discussion synthesizing \pkg{spmodel}'s contributions to the analysis of SPGLMs in \proglang{R}.
 
-# The spatial generalized linear model and marginalizatoin {#sec-spglm}
+# The spatial generalized linear model and marginalization {#sec-spglm}
 
-\pkg{spmodel} implements the novel methods described in @ver2024marginal to fit SPGLMs, which leverages the Laplace approximation and marginalizes over both the latent $\mathbf{w}$ and the fixed effects while accommodating spatial covariance.
-A beneficial aspect of this approach is that it formally maximizes a hierarchical GLM likelihood [@lee1996hierarchical; @wood2017generalized].
-This makes likelihood-based statistics for model comparison like AIC [@akaike1974new], AICc [@hoeting2006model], BIC [@schwarz1978estimating], deviance [@mccullagh1989generalized], and likelihood ratio tests available.
+The novel Laplace approach implemented in \pkg{spmodel} formally maximizes a hierarchical GLM likelihood [@lee1996hierarchical; @wood2017generalized], making likelihood-based statistics for model comparison like AIC [@akaike1974new], AICc [@hoeting2006model], BIC [@schwarz1978estimating], deviance [@mccullagh1989generalized], and likelihood ratio tests available.
 These types of statistics are not available for quasi-likelihood [@wedderburn1974quasi; @breslow1993approximate] or pseudo-likelihood approaches [@wolfinger1993generalized], which only specify the first two moments of a distribution.
-@ver2024marginal provides thorough details regarding the method and contextualizes its development which built upon similar methods [@evangelou2011estimation, @bonat2016practical].
-Next, we describe a brief overview of the approach and how it can be used for parameter estimation, inference, and prediction.
+Next, we describe a brief overview of the approach and how it can be used for several primary data analysis tasks [@tredennick2021practical] like model comparison, parameter estimation, inference, model diagnostics, and prediction.
 
@@
 We can write the SPGLM likelihood hierarchically as
-\begin{equation}\label{eq-marginal}
+\begin{align}\label{eq-marginal}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w},
-\end{equation}
+\end{align}
 where $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ is the density for the appropriate response distribution of $\mathbf{y}$ (e.g., binomial, Poisson) given the latent $\mathbf{w}$ and dispersion parameter ($\varphi$), and $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ is the multivariate Gaussian density for $\mathbf{w}$ given the explanatory variables ($\mathbf{X}$), fixed effects ($\boldsymbol{\beta}$), and spatial covariance parameters ($\boldsymbol{\theta}$).
 The elements of $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ are conditionally independent (given $\mathbf{w}$), but the elements of $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ share spatial covariance.
-Following @harville1977maximum, we can integrate $\boldsymbol{\beta}$ out of Equation$~$\ref{eq-spglm}, which yields
-\begin{equation}\label{eq-marginal2}
+Following @harville1977maximum, we can integrate $\boldsymbol{\beta}$ out of Equation$~$\ref{eq-marginal}, which yields
+\begin{align}\label{eq-marginal2}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\mathbf{w},
-\end{equation}
+\end{align}
 where $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ is the restricted (i.e., residual) multivariate Gaussian density [@patterson1971recovery] for $\mathbf{w}$ given the explanatory variables and covariance parameters.
-Equation$~$\ref{eq-marginal2} can be synonymous written after profiling the overall variance out of $\boldsymbol{\Sigma}$, which reduces the dimension of $\boldsymbol{\theta}$ by one for optimization [@wolfinger1994computing].
+Equation$~$\ref{eq-marginal2} can synonymously be written after profiling the overall variance out of $\boldsymbol{\Sigma}$, which reduces the dimension of $\boldsymbol{\theta}$ by one for optimization [@wolfinger1994computing].
 The restricted multivariate Gaussian density is given by
-\begin{equation}\label{eq-reml-def}
+\begin{align}\label{eq-reml-def}
 [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] = \frac{\exp(-\frac{1}{2}(\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}}) \boldsymbol{\Sigma}^{-1} (\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}})^T)}{(2 \pi)^{(n - p)/2} |\boldsymbol{\Sigma}|^{1/2}|\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X}|^{1/2}},
-\end{equation}
+\end{align}
 where $\tilde{\boldsymbol{\beta}} = (\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{w}$ and $|\cdot|$ denotes the determinant. Next, let 
-\begin{equation}\label{eq-marginal03}
+\begin{align}\label{eq-marginal03}
   \ell_\mathbf{w} = \log([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}])
-\end{equation}
+\end{align}
 and rewrite Equation$~$\ref{eq-marginal2} as 
-\begin{equation}\label{eq-marginal3}
+\begin{align}\label{eq-marginal3}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \exp(\ell_\mathbf{w}) d\mathbf{w}.
-\end{equation}
-A second-order Taylor series expansion of $\ell_\mathbf{w}$ around $\hat{\mathbf{w}}$ yields
-\begin{equation}\label{eq-marginal4}
+\end{align}
+A second-order Taylor series expansion of $\ell_\mathbf{w}$ around a point $\hat{\mathbf{w}}$ yields
+\begin{align}\label{eq-marginal4}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\hat{\mathbf{w}}} + \mathbf{g}^T(\mathbf{w} - \hat{\mathbf{w}}) + \frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T \mathbf{G} (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w},
-\end{equation}
+\end{align}
 where $\mathbf{g}$ and $\mathbf{G}$ are the gradient and Hessian, respectively, of $\ell_\mathbf{w}$ with respect to $\mathbf{w}$. If $\hat{\mathbf{w}}$ is a value for which $\mathbf{g} = \mathbf{0}$,
-\begin{equation}\label{eq-marginal5}
+\begin{align}\label{eq-marginal5}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\hat{\mathbf{w}}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T (-\mathbf{G}) (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w}.
-\end{equation}
+\end{align}
 The integral in Equation$~$\ref{eq-marginal5} can be solved by leveraging properties of the normalizing constant of a multivariate Gaussian distribution.
 Thus, rewriting $\exp(\ell_{\hat{\mathbf{w}}})$ yields
-\begin{equation}\label{eq-marginal6}
+\begin{align}\label{eq-marginal6}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\hat{\mathbf{w}}), \varphi] [\hat{\mathbf{w}} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\hat{\mathbf{w}}}|^{-1/2}.
-\end{equation}
+\end{align}
 
@@
 Had we observed $\mathbf{w}$, a GLS estimator for $\boldsymbol{\beta}$ is given by
-\begin{equation}\label{eq-gls1}
+\begin{align}\label{eq-gls1}
   \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
-\end{equation}
+\end{align}
 where $\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}$.
@@
 It follows that $\text{E}(\hat{\mathbf{w}})$ is given by
-\begin{equation}
+\begin{align}
   \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
@@
    = \mathbf{X}\boldsymbol{\beta}
-\end{equation}
+\end{align}
 and $\text{Var}(\hat{\mathbf{w}})$ is given by
@@
 Putting this all together, it follows that
-\begin{equation}
+\begin{align}
   \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
   = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}
-\end{equation}
+\end{align}
 and
@@
   & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
-  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
-  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}
+  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
+  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}
 \end{align}
@@
 These results are important because they justify closed-form solutions for $\hat{\boldsymbol{\beta}}$ and its associated variance.
-Closed-form solutions are useful because they bypass the need for computationally expensive sampling-based strategies to evaluate the mean and variance of $\hat{\boldsymbol{\beta}}$ -- a common technique for other approaches to SPGLMs like Bayesian MCMC.
+Closed-form solutions are useful because they bypass the need for sampling-based strategies to evaluate the mean and variance of $\hat{\boldsymbol{\beta}}$, a common technique for other approaches to SPGLMs like Bayesian MCMC.
 
@@
 The response GLM residuals simply compare the observation to its fitted latent mean:
-\begin{equation}
+\begin{align}
   \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
-\end{equation}
+\end{align}
 Because observations often have a unique support in a GLM (e.g., only two possible response values for binary data) and the variance of an observation generally depends on its mean, response residuals lack some utility.
@@
 Deviance residuals are given by
-\begin{equation}
+\begin{align}
   \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
-\end{equation}
+\end{align}
 where $\mathbf{d}$ is a vector of individual deviances.
 The sum of the squared deviance residuals equals the sum of $\mathbf{d}$.
-The sum of $\mathbf{d}$ is the deviance of the model fit, which quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)$ for all $i$) and the fitted model.
+The sum of $\mathbf{d}$ is the deviance of the model fit, which quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized].
 Deviance is often used as a fit statistic; lower values of deviance imply a better model fit.
-Pearson and standardized residuals are other types of GLM residuals that involve some scaling of the response residuals; the Pearson residuals scale $\mathbf{r}_{r}$ by the squrae root of $\mathbf{V}$, while the standardized residuals scale the deviance residuals by $\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}$, where $\mathbf{L}_{ii}$ is the $i$th diagonal element of the leverage matrix, which we discuss next.
+Pearson and standardized residuals are other types of GLM residuals that involve a scaling of the response residuals; the Pearson residuals scale $\mathbf{r}_{r}$ by the square root of $\mathbf{V}$, while the standardized residuals scale the deviance residuals by $\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}$, where $\mathbf{L}_{ii}$ is the $i$th diagonal element of the leverage matrix, which we discuss next.
+
 An observation has high leverage if its combination of explanatory variables is far away from other observations.
-In a linear model, the leverage values are the diagonal of the leverage (i.e., projection, hat) matrix, $\mathbf{L} = \mathbf{X}(\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top$.
+In a linear model, the leverage (i.e., hat) values are the diagonal of the leverage (i.e., projection, hat) matrix, $\mathbf{L} = \mathbf{X}(\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top$.
 In a GLM, the leverage matrix is given by
-\begin{equation}
+\begin{align}
   \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2},
-\end{equation}
-where $\mathbf{V}$ is a diagonal matrix with $i$th diagonal element equal to the variance of the response distribution evaluated at $f^{-1}(\mathbf{w}_i)$ [@faraway2016extending].
-The matrix $\mathbf{V}$ is sometimes called the GLM weight matrix.
-The larger the $i$th diagonal element of the hat matrix, the more severe the leverage from the $i$th observation.
-An observation is influential if it has a sizeable impact on model fit. Influence is measured using Cook's distance [@cook1979influential; @cook1982residuals], which is given for a GLM by
-\begin{equation}
-  \mathbf{c} = \mathbf{r}^2_{s} \frac{diag(\mathbf{L})}{\text{tr}(\mathbf{L})(\mathbf{1} - diag(\mathbf{L}))},
-\end{equation}
+\end{align}
+where $\mathbf{V}$ is a diagonal matrix with $i$th diagonal element equal to the variance of the response distribution evaluated at $f^{-1}(\mathbf{w}_i)$ [@faraway2016extending]; $\mathbf{V}$ is sometimes called the GLM weight matrix.
+The larger the value of $\mathbf{L}_{ii}$, the more severe the leverage from the $i$th observation.
+
+An observation is influential if it has a sizable impact on model fit. Influence is measured using Cook's distance [@cook1979influential; @cook1982residuals], which is given for a GLM by
+\begin{align}
+  \mathbf{c} = \frac{\mathbf{r}^2_{s}}{\text{tr}(\mathbf{L})} \frac{diag(\mathbf{L})}{(\mathbf{1} - diag(\mathbf{L}))},
+\end{align}
 where $\mathbf{r}^2_{s}$ are the standardized residuals and $diag(\mathbf{L})$ indicates the diagonal elements of the leverage matrix.
-The larger the $i$th diagonal element of the hat matrix, the more severe the influence from the $i$th observation.
+The larger the value of $\mathbf{c}_i$, the more severe the influence from the $i$th observation.
 @montgomery2021introduction provide guidance for interpreting these types of statistics, including cutoffs to consider when identifying unusual residual, leverage, or influence values.
 
-In a linear model, the $R^2$ (R-squared) statistic quantifies the proportion of variability in the data captured by the explanatory variables and is calculated as one minus the ratio of the error sum of squares to the total sum of squares [@rencher2008linear].
-In a GLM, there are many ways to define such a statistic [@smith2013comparison].
-One such approach is to use one minus the deviance ratio:
-\begin{equation}
+In a linear model, the $R^2$ (R-squared) statistic quantifies the proportion of variability in the data captured by the explanatory variables.
+It is calculated as one minus the ratio of the error sum of squares to the total sum of squares [@rencher2008linear].
+In a GLM, there are many ways to define a statistic that emulates the aforementioned meaning of $R^2$ from the linear model [@smith2013comparison].
+This statistic is called a pseudo R-squared ($PR^2$).
+One $PR^2$ for GLMs simply replaces the sums of squares ratio from the linear model with the deviance ratio:
+\begin{align}
   PR^2 = 1 - \frac{deviance_{fit}}{deviance_{null}},
-\end{equation}
+\end{align}
 where $deviance_{fit}$ is the deviance of the fitted model (sometimes called the residual deviance) and $deviance_{null}$ is the deviance of the model taking $\mathbf{X} \equiv \mathbf{1}$, a column of all ones (i.e., an intercept-only model).
 In practice, $deviance_{null}$ is derived by computing $\hat{\mathbf{w}}$ when $\mathbf{X} \equiv \mathbf{1}$ given $\hat{\boldsymbol{\theta}}$ and $\hat{\varphi}$ from the fitted model.
-Like the standard $R^2$, this statistic attempts to capture variability (i.e., deviance) attributable to the explanatory variables.
+Like $R^2$, $PR^2$ can be adjusted to account for the numbers of parameters estimated in a model.
 Because the $deviance_{null}$ denominator changes across fitted models (as the values of $\hat{\boldsymbol{\theta}}$ and $\hat{\varphi}$ change), this statistic should not be used as a model comparison tool.
-Instead, it should be used as an informative diagnostic tool unique to each model fit.
+Rather, it should be used as an informative diagnostic tool that unique to each model fit and describes how much variability from that model is attributable to the explanatory variables.
 
@@
 We may also predict values of the latent mean (on the link scale) at new locations by leveraging the spatial covariance between observed locations and new locations (spatial prediction is also called Kriging; see @cressie1990origins).
-Again suppose that we observed $\mathbf{w}$ and we want to make predictions at $\mathbf{u}$, a vector of latent means at the new locations that follows the same SPGLM from Equation~\ref{eq-spglm} with fixed effects design matrix, $\mathbf{X}_{\mathbf{u}}$.
-The vector $(\mathbf{w}, \mathbf{u})^\top$ has the following properties:
+Again suppose that we observed $\mathbf{w}$ and we want to make predictions at $\mathbf{u}$, a vector of latent means at the new locations that follows the same SPGLM from Equation$~$\ref{eq-spglm} and having fixed effects design matrix, $\mathbf{X}_{\mathbf{u}}$.
+The vector $(\mathbf{w}, \mathbf{u})^\top$ has expectation $(\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top$ and covariance matrix $\begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}$, where $\boldsymbol{\Sigma} = \text{Var}(\mathbf{w}, \mathbf{w})$, $\boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}} = \text{Var}(\mathbf{w}, \mathbf{u})$, $\boldsymbol{\Sigma}_{\mathbf{u} \mathbf{w}} = \boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}}^\top$ and $\boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} = \text{Var}(\mathbf{u}, \mathbf{u})$
+By assumption, we have observed $\mathbf{w}$, so we may derive the conditional distribution of $\mathbf{u}|\mathbf{w}$, which has the following properties:
 \begin{align}
-  \text{E}(\mathbf{w}, \mathbf{u})^\top & = (\text{E}(\mathbf{w}), \text{E}(\mathbf{u}))^\top = (\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top \\
-  \text{Var}(\mathbf{w}, \mathbf{u})^\top & = \begin{bmatrix} \text{Var}(\mathbf{w}, \mathbf{w}) & \text{Var}(\mathbf{w}, \mathbf{u}) \\ \text{Var}(\mathbf{u}, \mathbf{w}) & \text{Var}(\mathbf{u}, \mathbf{u}) \end{bmatrix} = \begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}
-\end{align}
-Because we have observed $\mathbf{w}$, we may derive the conditional distribution of $\mathbf{u}|\mathbf{w}$, which has the following properties:
-\begin{align}
-  \text{E}(\mathbf{w} | \mathbf{u}) & = \mathbf{X}_{\mathbf{u}} \boldsymbol{\beta} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{w} - \mathbf{X}\boldsymbol{\beta}) \\
-  \text{E}(\mathbf{w} | \mathbf{u}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
+  \text{E}(\mathbf{u} | \mathbf{w}) & = \mathbf{X}_{\mathbf{u}} \boldsymbol{\beta} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{w} - \mathbf{X}\boldsymbol{\beta}) \\
+  \text{E}(\mathbf{u} | \mathbf{w}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
 \end{align}
@@
 where $\mathbf{K} = \mathbf{X}_{\mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \mathbf{X}$ and $\boldsymbol{\Lambda} = \mathbf{X}_{\mathbf{u}}\mathbf{B} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{1} - \mathbf{X}\mathbf{B})$ for a vector of ones, $\mathbf{1}$.
-
 As with $\hat{\boldsymbol{\beta}}$, in practice these covariance matrices are evaluated at $\hat{\boldsymbol{\theta}}$.
-Moreover, these closed-form solutions provided enhance computational efficiency and clarity of the predictor's behavior.
 
@@
 The \code{moose} data in \pkg{spmodel} contain information on moose (Alces Alces) presence in the Togiak region of Alaska, USA.
-\code{moose}is an \code{sf} object, a special data frame that is supplemented with spatial information using the \pkg{sf} package in \proglang{R} [@pebesma2018sf].
+\code{moose} is an \code{sf} object, a special data frame that is supplemented with spatial information using the \pkg{sf} package in \proglang{R} [@pebesma2018sf].
 The first few rows of \code{moose} look like:
@@
 There are five columns: \code{elev}, the numeric site elevation (meters); \code{strat} a stratification variable for sampling with two levels, \code{"L"} and \code{"M"}, which are categorized by landscape metrics at each site; \code{count}, the number of moose at each site; \code{presence}, a factor that indicates whether at least one moose was observed at each site (\code{0} implies no moose; \code{1} implies at least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG: 3338) projected coordinate of each site (these data are point-referenced because each observation occurs at point coordinates and are represented by a \code{POINT} geometry.
-The \code{moose_preds} data in \pkg{spmodel} contain spatial locations at which predictions of moose presence are desired (and is also point-referenced).
-\code{moose_preds} is also an \code{sf} object with measurements for \code{elev} and \code{strat} and the same projection system.
-Figure$~$\ref{fig-moose-data} shows the \code{presence} variable in \code{moose} as well as the spatial locations of both \code{moose} and \code{moose_preds}.
-Moose are most commonly present in the southwestern and eastern parts of the domain and least commonly present in the northwest (Figure$~$\ref{fig-moose-data}).
-Next we show how to use \pkg{spmodel} to study the effect of elevation and strata on moose presence while accounting for spatial covariance and to make predictions of moose presence at new locations.
+Moose are most prevalent in the southwestern and eastern parts of the Togiak region (Figure$~$\ref{fig-moose-data}).
 
@@
 
+The \code{moose_preds} data in \pkg{spmodel} is an \code{sf} object with point locations at which moose presence predictions are desired.
+Like \code{moose}, \code{moose_preds} contains \code{elev} and \code{strat} for each site:
+```{r}
+head(moose_preds)
+```
+
 ## Model Fitting
@@
 SPGLMs in \pkg{spmodel} are fit using the \code{spglm()} function.
-The \code{spglm()} function requires four arguments: `formula`, the relationship between the response and explanatory variables; `family`, the response distribution assumed for the repsonse variable; `data`, the data frame that contains the variables in `formula`, and `spcov_type`, the type of spatial covariance.
-These first three arguments are the three required arguments to \code{glm()} for nonspatial GLMs.
+The \code{spglm()} function requires four arguments: \code{formula}, the relationship between the response and explanatory variables; \code{family}, the response distribution assumed for the repsonse variable; \code{data}, the data frame that contains the variables in \code{formula}, and \code{spcov_type}, the type of spatial covariance.
+The \code{formula}, \code{family}, and \code{data} arguments are the three required arguments to \code{glm()} for nonspatial GLMs.
 So, the transition from \code{glm()} to \code{spglm()}
 simply requires one additional argument: \code{spcov_type}.
-When \code{data} is not an \code{sf} object, \code{spglm()} also requires the \code{xcoord} and \code{ycoord} arguments, which indicate the columns in \code{data} that represent the x- and y-coordinates, respectively (it is assumed these coordinates are already projected).
+When \code{data} is not an \code{sf} object, \code{spglm()} also requires the \code{xcoord} and \code{ycoord} arguments, which indicate the columns in \code{data} that represent projected the x- and y-coordinates, respectively.
 
-We use \code{spglm()} to fit a spatial logistic regression model quantifying the effect of elevation and strata on moose presence:
+We use \code{spglm()} to fit a spatial GLM (i.e., here, a spatial logistic regression) quantifying the effect of elevation and strata on moose presence:
 ```{r}
@@
 
-Based on this model, there is some evidence that elevation is associated with higher probabilities of moose presence ($p$-value $\approx$ 0.087) but strong evidence that moose are more prevalent in the \code{"M"} strata than the \code{"L"} strata ($p$-value < 0.001).
-The fixed effects coefficients table from \code{summary()} is often of primary practical interest, but it is not easily usable when printed directly to the \proglang{R} console.
+The model suggests that elevation is positively associated with the log odds of moose presence ($p$-value $\approx$ 0.087), after controlling for strata.
+The model also suggests that moose have a higher log odds of presence in the \code{"M"} strata compared to the  \code{"L"} strata ($p$-value < 0.001), after controlling for elevation.
+
+The fixed effects coefficients table from \code{summary()} is often of primary scientific interest, but it is not immediately usable when printed directly to the \proglang{R} console.
 The \code{tidy()} function tidies this table, turning it into a data frame (i.e., a tibble) with standard column names:
@@
 ```{r}
-tidy(anova(spbin, bin))
+anova(spbin, bin)
 ```
 
-The LRT test statistic is $\approx$ 31.5 with three degrees of freedom (the difference in covariance parameters), yielding a $p$-value $< 0.001$ that indicates preference for the full model (SPGLM) relative to the reduced model (GLM).
+The LRT provides strong evidence that the SPGLM is preferred to the GLM ($p$-value < 0.001).
 
 An alternative approach to model comparison is to use a cross-validation procedure [@james2013introduction].
-The \code{loocv()} function performs leave-one-out cross validation, comparing the predicted mean (on the response scale) to the observed response variable for each hold-out observation, recomputing estimates of $\boldsymbol{\beta}$ each time.
-Then, statistics like bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) can be used to evaluate models:
+The \code{loocv()} function performs leave-one-out cross validation, comparing the predicted mean (on the response scale) to the observed response variable for each hold-out observation, recomputing estimates of $\boldsymbol{\beta}$ each iteration Then, statistics like bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) can be used to evaluate models:
 
@@
 ```{r}
-spbin2 <- update(spbin, spcov_type = "spherical")
+spbin2 <- update(spbin, spcov_type = "gaussian")
 glances(spbin, spbin2)
@@
 
-The \code{"exponential"} spatial covariance (\code{spbin}) has a slightly lower deviance but slightly higher AIC, AICc, and BIC than 
-the \code{"spherical"} spatial covariance (\code{spbin2}).
-Both spatial covariance functions nearly identical leave-one-out cross validation metrics.
-For practical purposes, these models fit quite similarly.
+The \code{"exponential"} spatial covariance (\code{spbin}) has a slightly lower (better) deviance but slightly higher (worse) AIC, AICc, and BIC than the \code{"gaussian"} spatial covariance (\code{spbin2}).
+Both spatial covariance functions have similar leave-one-out cross validation metrics, though the \code{"gaussian"} spatial covariance RMSPE is slightly lower (better).
+For practical purposes, these models fit similarly.
+
+Often times in spatial statistics, the difference in model fit between the best spatial model and worst spatial model is much smaller than the difference in model fit between the worst spatial model and the nonspatial model, implying that accounting for some form of spatial covariance is very beneficial.
+Two spatial covariance functions to consider starting with are the exponential and Gaussian, which have quite different origin behaviors (Figure$~$\ref{fig-type}), something @stein1999interpolation argues is important to characterize accurately.
 
@@
 
-\code{spmodel} provides a suite of tools for model diagnostics. The \code{augment()} function augments the model data with diagnostics:
+\code{spmodel} provides a suite of tools for model diagnostics. One is \code{augment()}, which augments the model data with diagnostics:
 ```{r}
@@
 
-The fitted values (\code{.fitted}) can be returned on either the link ($\hat{\mathbf{w}}$) or response ($f^{-1}\hat{\mathbf{w}}$) scale and the residuals (\code{.resid}) can deviance, pearson, or response residuals.
-The defaults are the link scale and deviance residuals, respectively.
+The fitted values (\code{.fitted}) can be returned on either the link ($\hat{\mathbf{w}}$) or response ($f^{-1}(\hat{\mathbf{w}})$) scale and the residuals (\code{.resid}) can deviance, Pearson, or response residuals.
+The default fitted values are the link scale and the default residuals are deviance residuals.
 Also returned by \code{augment()} are the leverage (\code{.hat}), Cook's distance (\code{.cooksd}), and standardized residuals \code{.std.resid}.
 A benefit of using \code{augment()} when \code{data} is an \code{sf} object is that the output is also an \code{sf} object, which makes it straightforward to create spatial diagnostic plots (Figure$~$\ref{fig-sp-diagnostic}).
-Standard \proglang{R} helpers (e.g., \code{fitted()}, \code{residuals()}) are available to alternatively extract model diagnostics from the model object.
+Standard \proglang{R} helpers (e.g., \code{fitted()}, \code{residuals()}) are also available to extract model diagnostics from the model object.
 
@@
 \includegraphics[width = 1\linewidth]{figures/figure-2.png}
-\caption{Spatial logistic regression model diagnostics from [augment]{.fct}. The leverage (i.e., hat) values (left) and standardized residuals (right).}
+\caption{Moose presence model diagnostics, including leverage values (left) and standardized residuals (right).}
 \label{fig-sp-diagnostic}
@@
 
-The \code{plot()} function can also be used to return similar diagnostics as from \code{lm()} and \code{glm()} with additional tools for spatial covariance.
-For example, we can inspect Cook's distance values and the empirical spatial covariance as a function (Figure$~$\ref{fig-sp-diagnostic2}) with
+The \code{plot()} function can also be used to return similar diagnostics as from \code{lm()} and \code{glm()} with additional tools for diagnosing spatial covariance.
+For example, we can inspect Cook's distance values and the empirical spatial covariance as a function of distance with (Figure$~$\ref{fig-sp-diagnostic2}):
 ```{r, eval = FALSE}
@@
 \includegraphics[width = 1\linewidth]{figures/figure-3.png}
-\caption{Spatial logistic regression model diagnostics from [plot]{.fct}. The Cook's distance values (left) and the fitted spatial covariance as a function of distance (right).}
+\caption{Moose presence model diagnostics, including Cook's distance and the fitted spatial covariance as a function of distance (right).}
 \label{fig-sp-diagnostic2}
@@
 
-The \code{varcomp()} function partitions model variability into several different components:
+The \code{varcomp()} function partitions model variability into several different components, helping to elucidate the model's structure:
 ```{r}
@@
 
-The pseudo R-squared ($PR^2$), the proportion of variability attributable to the explanatory variables, is reported in the first row.
+The pseudo R-squared ($PR^2$) is reported in the first row.
 The remaining variability ($1 - PR^2$) is allocated proportionally to \code{de} and \code{ie} according to $\sigma^2_{de}$ and $\sigma^2_{ie}$.
@@
 
-Predictions on the response scale are visualized alongside the fitted values ($f^{-1}\hat{\mathbf{w}}$) in Figure$~$\ref{fig-moose-fit}.
+Predictions on the response scale are visualized alongside the fitted values ($f^{-1}(\hat{\mathbf{w}})$) in Figure$~$\ref{fig-moose-fit}.
 
@@
 
-Prediction intervals for the probability of moose presence (on the link scale) are returned via \code{interval}
+Prediction intervals for the probability of moose presence (on the link scale) are returned by suppplying \code{interval}:
 ```{r}
@@
 In Section$~$\ref{sec-moose-count}, we fit Poisson and negative binomial models with and without geometric anisotropy for the point-referenced moose count data.
-In Section$~$\ref{sec-seal}, we fit a binomial model to the areal seal trend data with a nonspatial random effect.
-In Section$~$\ref{sec-texas}, we fit beta models to Texas voter turnout data.
-We explore how the Texas data can be treated as point-referenced or areal and compare the two spatial covariance structures empirically.
-We also highlight how the maximum likelihood estimation method is useful when comparing two models with different explanatory variables.
-Finally, in Section$~$\ref{sec-lake}, we fit a Gamma model to the point-referenced lake conductivity data.
-We show how to perform a spatial analysis of variance (ANOVA) and leverage modeling functions from other \proglang{R} packages like \pkg{emmeans} and \pkg{car}.
+In Section$~$\ref{sec-lake}, we fit a Gamma model to the point-referenced lake conductivity data, showing how to perform a spatial analysis of variance (ANOVA), fit contrasts for models with interactions and use maximum likelihood to compare two models with different explanatory variables.
+In Section$~$\ref{sec-seal}, we fit a binomial model to the areal harbor seal trend data with a nonspatial random effect.
+Finally in Section$~$\ref{sec-texas}, we fit beta models to Texas voter turnout data, which can be treated as point-referenced or areal.
 
@@
 
-In addition to moose presence, moose counts are also recorded in `moose` (Figure$~$\ref{fig-moose-data-count}.
+In addition to moose presence, moose counts are also recorded in \code{moose} (Figure$~$\ref{fig-moose-data-count}.
 The Poisson and negative binomial response distributions can be used to model SPGLMs for count data.
-Using a Gaussian spatial covariance function, we may fit both a Poisson and negative binomial SPGLM changing the `family` argument:
+The Poisson distribution mean is equal to its variance, while the negative binomial has an extra parameter to accommodate overdispersion (where the variance is larger than the mean).
+Using a spherical spatial covariance function, we may fit both a Poisson and negative binomial SPGLM changing the \code{family} argument:
 
@@
   data = moose,
-  spcov_type = "gaussian"
+  spcov_type = "spherical"
 )
@@
 Geometric anisotropy is formally incorporated by rotating and scaling original coordinates, yielding transformed coordinates that are geometrically isotropic:
-\begin{equation}
+\begin{align}
   \begin{bmatrix}
@@
   \end{bmatrix}.
-\end{equation} 
-The parameter $\alpha$ controls the rotation angle and the parameter $\omega$ controls the scaling...
-Then using these transformed coordinates, the partial sill ($\sigma^2_{de}$), nugget ($\sigma^2_{ie}$), and range ($\phi$) parameters are estimated. 
+\end{align} 
+The parameters $\omega$ and $\alpha$ controls the scaling and rotation, respectively, of the major and minor axes of a level curve of equal correlation (Figure$~$\ref{fig-tropy}).
+Using these transformed coordinates, the partial sill ($\sigma^2_{de}$), nugget ($\sigma^2_{ie}$), and range ($\phi$) parameters are estimated.
+We accommodate geometric anisotropy by supplying \code{anisotropy}:
 ```{r}
@@
 
-According to BIC, the spatial negative binomial model performs best.
-This model simultaneously accounts for anisotropy and overdispersion.
-The spatial correlation is strongest in a northwest-southeast direction and weakest in the northeast-southwest direction (Figure$~$\ref{fig-tropy}), which is intuitive given the similar patterns in moose counts from Figure$~$\ref{fig-moose-data-count}.
+According to BIC, the spatial negative binomial model with geometric anisotropy performs best:
 ```{r}
@@
 
-The `plot()` function can be used to visualize the anisotropy  (Figure$~$\ref{fig-tropy}):
+The \code{plot()} function can be used to visualize the anisotropy (Figure$~$\ref{fig-tropy}):
 ```{r, eval = FALSE}
@@
 ```
+
+The spatial correlation is strongest in a northwest-southeast direction and weakest in the northeast-southwest direction (Figure$~$\ref{fig-tropy}), which is intuitive given the similar patterns in moose counts from Figure$~$\ref{fig-moose-data-count}.
 
@@
 Chemical data is often heavily right-skewed, so we model it using an SPGLM assuming a Gamma distribution for the response.
-The \code{log_cond} variable in \code{lake} is the logarithm of conductivity, and we can dynamically exponentiate within \code{formula} so that it is on the original scale:
+The \code{log_cond} variable in \code{lake} is the logarithm of conductivity, which we dynamically exponentiate within \code{formula} so that it is on the original scale:
 ```{r}
@@
 We model conductivity as a function of temperature, state, and lake origin, and we allow the effect of temperature to vary by state (\code{temp:state}).
-The partition factor restricts spatial covariance to apply only for two observations sampled during the same year.
-Data were collected in 2012 and 2017, so the \code{year} partition factor assumes independence between observations in 2012 and 2017.
+The \code{year} partition factor (specified via \code{partition_factor}) restricts spatial covariance to be nonzero only for observations sampled during the same year.
+Data were collected in 2012 and 2017, so this partition factor assumes independence between observations in 2012 and 2017.
 While we used the partition factor here illustratively, more generally, the utility of partition factors can be highly context dependent.
@@
 
+```{r, warning=FALSE, message=FALSE}
+library(car)
+```
+
 ```{r}
-library(car)
 vif(spgam)
@@
 Because of the interaction between \code{temp} and \code{state}, contrasts that assess mean differences among states should condition upon a specific temperature value.
-By default, \pkg{emmeans} uses the mean temperature value to assess contrasts:
+By default, \pkg{emmeans} uses the mean temperature value (here, 7.63) to assess contrasts:
+```{r, warning=FALSE, message=FALSE}
+library(emmeans)
+```
+
 ```{r}
-library(emmeans)
 pairs(emmeans(spgam, ~ state | temp))
@@
 
-Again because of the interaction between \code{temp} and \code{state}, we should assess temperature trends separately for each state:
+Again, because of the interaction between \code{temp} and \code{state}, we should assess temperature trends separately for each state:
 ```{r}
@@
 Each polygon in the \code{seal} data represents a distinct harbor seal haulout location (Figure$~$\ref{fig-seal}).
-A haulout location is a spot on coastal rocks that harbor seals go to rest, molt, and give birth, among other tasks.
+A haulout location is a spot on coastal rocks that harbor seals go to rest, molt, and give birth.
 
@@
 
-For each polygon, a Poisson regression was used to quantify the mean trend in abundance over approximately 30 years.
-If the logarithm of seal abundance (\code{log_trend}) is negative (positive), it means abundance is decreasing (increasing).
-We use a binomial SPGLM to quantify how likely it is for abundance trends to be decreasing:
+For each polygon, a Poisson regression was used to quantify the mean trend in abundance over approximately 30 years [@ver2018spatial].
+If the logarithm of mean abundance trends (\code{log_trend}) is negative (positive), it means abundance is decreasing (increasing).
+We use a binomial SPGLM to quantify the likelihood that mean abundance trends are decreasing:
 ```{r}
@@
 To model spatial dependence, we used a conditional autoregressive function.
-Conditional and simultaneous autoregressive functions characterize spatial distance through neighborhood relationships and have \code{spcov_type} values of \code{"car"} and \code{"sar"}, respectively.
+Conditional and simultaneous autoregressive functions characterize spatial distance through neighborhood relationships (rather than Euclidean distance) and have \code{spcov_type} values of \code{"car"} and \code{"sar"}, respectively.
 By default, Queen's distance is used to determine whether two sites are neighbors, though custom neighborhood matrices are passed via \code{W}.
-Row standardization is also assumed by default, though this can be changed via \code{row_st}.
+Row standardization is also assumed by default; this can be changed via \code{row_st}.
 Using \code{random}, we also specified a nonspatial random effect for seal stock.
@@
 
-Tidying the model reveals the mean probability that trends are decreasing is between -0.979 and 1.66 (on the logit scale):
+Tidying the model reveals the estimates and confidence intervals on the log odds scale:
 ```{r}
@@
 
-Back-transforming this interval to the probability scale yields:
+Back-transforming the confidence interval to the probability scale yields:
 ```{r}
@@
 
-For areal data, the prediction locations must be specified at the time of model fitting, as they affect the spatial covariance function's neighborhood structure.
+In contrast to point-referenced data, prediction locations for areal data must be specified at the time of model fitting, as they affect the spatial covariance function's neighborhood structure.
 Prediction locations whose response values have an \code{NA} (i.e., missing) value are converted into a \code{newdata} object that is stored in the model output.
@@
 When areal data are used with \code{spglm()}, the centroids of each polygon are used as the point-referenced coordinates.
-We further explore comparisons between point-referenced and aeral data next.
+We further explore comparisons between point-referenced and aeral data in the next example.
 
@@
 The data are point-referenced, with polygon centroids representing the spatial location of each county (Figure$~$\ref{fig-texas}).
-Beta regression is a GLM used to model rate and proportion data in (0, 1) [@ferrari2004beta, @cribari2010beta].
+Beta regression is a GLM used to model rate and proportion data in the (0, 1) interval [@ferrari2004beta; @cribari2010beta].
 We model voter turnout rates as a function of mean log income of county residents using an SPGLM assuming a beta distributed response variable:
@@
 An alternative is to use maximum likelihood, which can use likelihood-based statistics to compare models with different explanatory variable and fixed effect structures.
-While we could assess the significance of log turnout with the \code{spbeta_geo} model fit with REML (e.g., via \code{tidy()}), an alternative approach is to use a likelihood ratio test as long as the two models are nested.
-We refit the point-referenced model using maximum likelihood and compare it to a similar model without log income using the likelihood ratio test, accessible via \code{anova()} when providing two models:
+While we could assess the significance of log turnout with the \code{spbeta_geo} model fit with REML (e.g., via \code{tidy()}), an alternative approach is to use a likelihood ratio test:
 ```{r}
 spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
-spbeta_red_ml <- update(spbeta_full_ml, formula = turnout ~ 1)
+spbeta_red_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
 anova(spbeta_full_ml, spbeta_red_ml)
@@
 
-The likelihood ratio test suggests that log income is significantly related to voter turnout. The \code{anova()} output can also be tidied using \code{tidy()}.
+The likelihood ratio test suggests that log income is significantly related to voter turnout ($p$-value < 0.001). We could instead use a different likelihood-based statistic like AIC:
+```{r}
+AIC(spbeta_full_ml, spbeta_red_ml)
+```
 
-SAY SOMETHING ABOUT PREDICT.
+The AIC prefers the full model, which suggests that log income is important for predicting voter turnout.
 
@@
 
-SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace approximation that marginalizes over the latent (i.e., unobserved) mean, $\mathbf{w}$, and the fixed effects, $\boldsymbol{\beta}$.
-The approach is quite flexible and accommodates any general response distribution and covariance structure.
-\pkg{spmodel}'s \code{spglm()} and \code{spgautor()} functions are similar in structure and syntax as base \proglang{R}'s \code{glm()} function, easing the transition for practitioners from GLMs to SPGLMs.
-These functions support six response distributions for binary, count, and skewed data and support 20 spatial covariance functions.
-\pkg{spmodel} provides several additional features that are not covered here, including fitting multiple models simultaneously, fixing spatial covariance parameters at known values, fitting models to large non-Gaussian data having thousands of observations via spatial indexing [@ver2023indexing], incorporating spatial dependence in machine learning (e.g., random forests; @breiman2001random), simulating spatially dependent data (e.g., \code{spbinom()}, \code{sprpois()}, etc.), and more.
+SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace approximation that simultaneously marginalizes over the latent (i.e., unobserved) random effects and the fixed effects, $\boldsymbol{\beta}$.
+The approach is very flexible and accommodates general response distributions and covariance structures, though here we focus on spatial applications.
+\pkg{spmodel}'s \code{spglm()} and \code{spgautor()} fit SPGLMs that are similar in structure and syntax as base \proglang{R}'s \code{glm()} function, easing the transition from GLMs to SPGLMs for practitioners.
+The \code{spglm()} and \code{spgautor()} functions support six response distributions for binary, count, and skewed data and 20 spatial covariance functions.
+\pkg{spmodel} also provides a suite of tools for data visualization, inference, model diagnostics, and prediction, providing a framework that can be used for all stages of a data analysis.
+There are many additional \pkg{spmodel} features that are not covered here, including fitting multiple models simultaneously, fixing spatial covariance and dispersion parameters at known values, fitting models to large non-Gaussian data having thousands of observations via spatial indexing [@ver2023indexing], incorporating spatial dependence in machine learning (e.g., random forests; @breiman2001random), simulating spatially dependent data (e.g., \code{spbinom()}, \code{sprpois()}, etc.), and more.
 Further details are provided by [https://CRAN.R-project.org/package=spmodel](https://CRAN.R-project.org/package=spmodel) and links therein.
@@
 The results in this paper were obtained using [R]{.proglang} 4.4.0 with the
-\pkg{spmodel} 0.9.0 package. Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.
+\pkg{spmodel} 0.11.0 package. Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.
 

### 2025-09-10 inst/manuscript/manuscript.tex
some updates

@@
 \Abstract{
-Generalized linear models (GLMs) describe a non-normal response variable
-that may be binary, count, skewed, or a proportion. Typically,
-observations in a GLM are assumed independent of one another. For
-spatial data, this independence assumption is impractical, as nearby
-locations tend to be more similar than locations far apart. The
-\pkg{spmodel} \proglang{R} package provides tools to fit GLMs that
-incorporate spatial correlation (i.e., spatial generalized linear
-models, or SPGLMs). SPGLMs are fit in \pkg{spmodel} using a novel
-application of the Laplace approximation via \code{spglm()} for
-point-referenced data or \code{spgautor()} for areal (i.e., lattice),
-data. \code{spglm()} and \code{spgautor()} closely resemble {glm} from
-base \proglang{R} but include arguments that control the spatial
-correlation structure. \pkg{spmodel} has many helper functions for model
-inspection and diagnostics, some of which leverage other \proglang{R}
-packages like {broom} and {emmeans}. \pkg{spmodel} has tools to make
-predictions of the latent spatial-mean process at unobserved locations.
-\pkg{spmodel} also provides many advanced features like accommodating
-geometric anisotropy and nonspatial random effects, simulating spatially
-autocorrelated data, and more. Here we use \pkg{spmodel} to illustrate
-the modeling of binary, count, skewed and proportion response variables
-from several point-referenced and areal data sets.
+Non-Gaussian data are common in practice and include binary, count,
+skewed, and proprtion data types. Often, non-Gaussian data are modeled
+using a generalized linear model (GLM). GLMs typically assume that
+observations are independent of one another. This is an impractical
+assumption for spatial data, as nearby observations tend to be more
+similar than distant ones. The \pkg{spmodel} package in \proglang{R}
+provides a suite of tools for fitting spatial generalized linear models
+(SPGLMs) to non-Gaussian data and making spatial predictions (i.e.,
+Kriging). SPGLMs for point-referenced data (x- and y-coordinates) are
+fit using the \code{spglm()} function, while SPGLMs for areal (lattice,
+polygon) data are fit using the \code{spgautor()} function. Both
+\code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood
+which marginalizes over the model's fixed effects and latent mean while
+formally incorporating spatial covariance (i.e., spatial correlation).
+The inputs and outputs of \code{spglm()} and \code{spgautor()} closely
+resemble the \code{glm()} function from base \proglang{R}, easing the
+transition from GLMs to SPGLMs. \pkg{spmodel} provides and builds upon
+several commonly used helper functions for model building like
+\code{summary()}, \code{plot()}, \code{fitted()}, and \code{tidy()},
+among others. Spatial predictions of the latent mean at unobserved
+locations are obtained using \code{predict()} or \code{augment()}.
+\pkg{spmodel} accommodates myriad advanced modeling features like
+geometric anisotrpoy, nonspatial random effects, analysis of variance,
+and more. Throughout, we use \pkg{spmodel} to fit SPGLMs to moose
+presence and counts in Alaska, United States (US), skewed conductivity
+data in the Southwestern US, harbor seal abundance trends in Alaska, US,
+and voter turnout rates in Texas, US.
 }
@@
 \usepackage{caption, subcaption}
+\usepackage{autonum}
 
@@
 
-In practice, non-Gaussian data are ubiquitous. Non-Gaussian that belong
-to an exponential family data can be naturally modeled using a
-generalized linear model (GLM) regression framework
-\citep{nelder1972generalized, mccullagh1989generalized, myers2012generalized, faraway2016extending}.
-In a GLM, an \(n \times 1\) response variable \(\mathbf{y}\) belongs to
-a statistical distribution (e.g., Poisson, Binomial) with some mean and
-variance. Often, the analysis goal is to study the impact of a linear
-function of several explanatory variables on \(\text{y}\) through a GLM.
-In this context, the latent (i.e., unobserved) mean of \(\mathbf{y}\),
+In practice, non-Gaussian data (e.g., binary, count, skewed, and
+proportion data) are ubiquitous. Non-Gaussian that belong to an
+exponential family data can be naturally modeled using a generalized
+linear model (GLM) regression framework
+\citep{nelder1972generalized, mccullagh1989generalized}. In a GLM, an
+\(n \times 1\) response variable \(\mathbf{y}\) belongs to a statistical
+distribution (e.g., Binomial, Poisson) with some mean and variance.
+Often, the analysis goal is to study the impact of a linear function of
+several explanatory variables on \(\mathbf{y}\) through a GLM. In this
+context, the latent (i.e., unobserved) mean of \(\mathbf{y}\),
 \(\boldsymbol{\mu}\), is linked to these explanatory variables via a
-link function: \begin{equation}\label{eq-glm}
+link function: \begin{align}\label{eq-glm}
 f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
-\end{equation} where for a sample size \(n\), \(f(\cdot)\) is a link
+\end{align} where for a sample size \(n\), \(f(\cdot)\) is a link
 function that connects \(\boldsymbol{\mu}\) to \(\mathbf{w}\),
@@
 fixed effects. While the mean is typically constrained in some way
-(e.g., between zero and one if a probability), the link function
-generally makes \(\mathbf{w}\) is unconstrained. Common link functions
-inlude the log odds (i.e., logit) link for binary and proportion data
-and the log link count and skewed data. Equation\(~\)\ref{eq-glm} can
-also be written in terms of the inverse link function,
-\(f^{-1}(\cdot)\): \begin{equation}\label{eq-glm2}
-\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}),
-\end{equation}
+(e.g., if a probability, between zero and one), the link function
+generally makes \(\mathbf{w}\) unconstrained. Common link functions
+include the log odds (i.e., logit) link for binary and proportion data
+and the log link for count and skewed data. Equation\(~\)\ref{eq-glm}
+can also be written in terms of the inverse link function,
+\(f^{-1}(\cdot)\): \begin{align}\label{eq-glm2}
+\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta} \equiv f^{-1}(\mathbf{w}) = f^{-1}(\mathbf{X} \boldsymbol{\beta}).
+\end{align}
 
@@
 This independence assumption is typically impractical for spatial data.
-In spatial data, nearby observations tend to be more similar than
-distant observations \citep{tobler1970computer}, leading to positive
-spatial covariance among observations. The consequences of ignoring
-spatial covariance in statistical models for spatial data can be severe
-and include imprecise parameter estimates as well as misleading standard
-errors that inflate Type-I error rates and decrease power
-\citep{zimmerman2024spatial}.
-
-An approach for handling spatial data using a GLM is to assume
-\(\mathbf{w}\) has spatial covariance. This is achieved by adding to
-Equation\(~\)\ref{eq-glm} two random effects, \(\boldsymbol{\tau}\) and
-\(\boldsymbol{\epsilon}\). The random effect \(\boldsymbol{\tau}\) is an
-\(n \times 1\) column vector of spatially dependent random errors. We
-assume that \(\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}\) and
+For spatial data, nearby observations tend to be more similar than
+distant observations \citep{tobler1970computer}, which leads to positive
+spatial covariance. The consequences of ignoring spatial covariance in
+statistical models for spatial data can be severe and include imprecise
+parameter estimates as well as misleading standard errors that inflate
+Type-I error rates and decrease power \citep{zimmerman2024spatial}.
+
+An approach for handling spatial data using a GLM is to assume the
+elements of \(\mathbf{w}\) share covariance that varies spatially. This
+is achieved by adding to Equation\(~\)\ref{eq-glm} two random effects,
+\(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\). The random effect
+\(\boldsymbol{\tau}\) is an \(n \times 1\) column vector of spatially
+dependent random errors. We assume that
+\(\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}\) and
 \(\text{Cov}(\boldsymbol{\tau}) = \sigma^2_\tau \mathbf{R}\), where
@@
 controls the magnitude of spatial covariance and is often called a
-partial sill, while the matrix \(\mathbf{R}\) is an \(n \times n\)
-spatial correlation matrix that depends on a range parameter which
-controls the distance-decay rate of the spatial correlation. One example
-of a spatial covariance matrix is the ``exponential'', which is given by
-\begin{equation}\label{eq-spcov-exp}
-  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de} \exp(-\mathbf{H}/\phi),
-\end{equation} where \(\mathbf{H}\) is a matrix of pairwise distances
-among the elements of \(\mathbf{y}\) and \(\phi\) is a range parameter.
-From Equation\(~\)\ref{eq-spcov-exp}, as the distance between two
-elements of \(\mathbf{y}\) increases, the spatial covariance decreases,
-which reflects intuition. Moreover, as the range parameter, \(\phi\),
+partial sill. The matrix \(\mathbf{R}\) is an \(n \times n\) spatial
+correlation matrix that depends on a range parameter controlling the
+distance-decay rate of the spatial correlation. One example of a spatial
+covariance matrix is the ``exponential'', which is given by
+\begin{align}\label{eq-spcov-exp}
+  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de} \mathbf{R}_{exp} =  \sigma^2_{de} \exp(-\mathbf{H}/\phi),
+\end{align} where \(\mathbf{H}\) is a matrix of pairwise distances among
+the elements of \(\mathbf{y}\) and \(\phi\) is the range parameter. From
+Equation\(~\)\ref{eq-spcov-exp}, as the distance between two elements of
+\(\mathbf{y}\) increases, the spatial covariance decreases, which
+reflects intuition. Moreover, as the range parameter, \(\phi\),
 increases, the strength of spatial dependence increases
@@
 variability (i.e., fine-scale variation) and is often called a nugget.
+Often in spatial statistics, quantities are explicitly referenced with
+respect to \(\mathbf{s}\), a vector of coordinates indexing the
+observation \citep{cressie1993statistics}. For example, \(\mathbf{y}\)
+and \(\mathbf{X}\) may instead be written \(\mathbf{y}(\mathbf{s})\) and
+\(\mathbf{X}\), respectively. We acknowledge the utility of this
+nomenclature but drop the explicit dependence on \(\mathbf{s}\) for
+simplicity of notation.
 
@@
 \(\boldsymbol{\epsilon}\), the spatial GLM (SPGLM) can be written as
-\begin{equation}\label{eq-spglm}
+\begin{align}\label{eq-spglm}
 f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
-\end{equation} Often in spatial statistics, quantities are explicitly
-referenced with respect to \(\mathbf{s}\), a vector of coordinates
-indexing the observation \citep{cressie1993statistics}. For example,
-\(\mathbf{y}\) and \(\mathbf{X}\) may instead be written
-\(\mathbf{y}(\mathbf{s})\) and \(\mathbf{X}\), respectively. We
-acknowledge the utility of this nomenclature but drop the explicit
-dependence on \(\mathbf{s}\) for simplicity of notation. Assuming
-independence among \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\),
-it follows that \begin{equation}\label{eq-spcov}
+\end{align} Assuming independence among \(\boldsymbol{\tau}\) and
+\(\boldsymbol{\epsilon}\), it follows that \begin{align}\label{eq-spcov}
  \text{Cov}(\boldsymbol{\tau} + \boldsymbol{\epsilon}) = \text{Cov}(\boldsymbol{\tau}) + \text{Cov}(\boldsymbol{\epsilon}) = \sigma^2_{\tau}\mathbf{R} + \sigma^2_{\epsilon} \mathbf{I}.
-\end{equation} To better align with intuition, we henceforth
-\(\sigma^2_{\tau}\) as \(\sigma^2_{de}\) (for spatially dependent error
-variance) and \(\sigma^2_{\epsilon}\) as \(\sigma^2_{ie}\) (for
-independent error variance). The parameters \(\sigma^2_{de}\),
-\(\sigma^2_{ie}\), the range parameter \(\phi\) in \(\mathbf{R}\), and
-any other parameters in \(\mathbf{R}\) compose \(\boldsymbol{\theta}\),
-the covariance parameter vector.
+\end{align} Henceforth, we refer to \(\sigma^2_{\tau}\) as
+\(\sigma^2_{de}\) (for spatially dependent error variance) and
+\(\sigma^2_{\epsilon}\) as \(\sigma^2_{ie}\) (for independent error
+variance). The parameters \(\sigma^2_{de}\), \(\sigma^2_{ie}\), and
+\(\phi\), in addition to any other parameters in \(\mathbf{R}\), compose
+\(\boldsymbol{\theta}\), the covariance parameter vector.
 
@@
 less intuitive than working directly with the covariance matrix. The
-{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
+\pkg{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
 \citep{brooks2017glmmtmb}, \pkg{hglm} \citep{ronnegard2010hglm},
@@
 challenging to quantify. The \pkg{sdmTMB} \citep{anderson2024sdmtmb}
-package combines elements of \pkg{R-INLA}, \pkg{glmmTMB}, and properties
-of Gaussian Markov random fields to fit a wide variety of SPGLMs, and
+package combines elements of \pkg{R-INLA}, \pkg{glmmTMB}, and Gaussian
+Markov random fields to fit a wide variety of SPGLMs, while
 \pkg{tinyVAST} \citep{thorson2025tinyVAST} extends some of these models
@@
 
-\citet{ver2024marginal} proposed a novel approach to fitting SPGLMs that
-leverages the Laplace approximation while marginalizing over both the
-latent \(\mathbf{w}\) and the fixed effects (\(\boldsymbol{\beta}\)) and
-accommodating spatial covariance. \citet{ver2024marginal} showed that
-this approach performed efficiently in a variety of simulation settings,
+Building from \citet{evangelou2011estimation} and
+\citet{bonat2016practical}, \citet{ver2024marginal} proposed a novel
+approach for fitting SPGLMs that leverages the Laplace approximation
+while marginalizing over both the latent \(\mathbf{w}\) and the fixed
+effects (\(\boldsymbol{\beta}\)) and accommodating spatial covariance.
+This approach performed efficiently in a variety of simulation settings,
 generally having appropriate confidence interval coverage for the fixed
-effects and prediction interval coverage for new \(\mathbf{w}\). The
-approach performed similarly to the Bayesian SPGLM approach in
-\pkg{spBayes} and the automatic differentiation SPGLM approach in
-\pkg{glmmTMB} but was much faster. At small sample sizes, the approach
-outperformed the approximate Bayesian SPGLM approach in \pkg{R-INLA} and
-had similar computational times. For moderate sample sizes, it performed
-similarly to \pkg{R-INLA}, though \pkg{R-INLA} was faster. This novel
-approach is particularly attractive for two reasons. First, it is
-general enough that can be applied to any covariance structure (not just
-spatial). Second, after estimating the covariance parameters, analytical
-solutions exist for the fixed effects (and their standard errors) as
-well as predictions of the latent \(\mathbf{w}\) at new locations (and
-their standard errors). The \pkg{spmodel} \proglang{R} package
-\citep{dumelle2023spmodel} recently provided full support for the
-methods in \citet{ver2024marginal} applied to binary, count, skewed, and
-proportion data for over 20 different spatial covariance types.
+effects and prediction interval coverage for \(\mathbf{w}\) at new
+locations. The approach performed similarly to the Bayesian SPGLM
+approach in \pkg{spBayes} and the automatic differentiation SPGLM
+approach in \pkg{glmmTMB} but was much faster. At small sample sizes,
+the approach outperformed the approximate Bayesian SPGLM approach in
+\pkg{R-INLA} and had similar computational times. For moderate sample
+sizes, it performed similarly to \pkg{R-INLA}, though \pkg{R-INLA} was
+faster. The novel Laplace approach is particularly attractive for two
+reasons. First, it is general enough that can be applied to any
+covariance structure (not just spatial). Second, after estimating the
+covariance parameters, analytical solutions exist for the fixed effects
+(and their standard errors) as well as predictions of the latent
+\(\mathbf{w}\) at new locations (and their standard errors).
 
@@
 recently provided a full set of modeling tools for SPGLMs fit using the
-methods described in \citet{ver2024marginal}. These modeling tools are
-approachable and mirror the familiar \code{glm()} syntax from
-base-\proglang{R}, making the transition from GLMs to SPGLMs relatively
-seamless. The \code{spglm()} function fits SPGLMs for point-referenced
-data (e.g., x-coordinates and y-coordinates representing point locations
-in a field), while the \code{spgautor()} function fits SPGLMs for areal
-data (e.g., polygon boundaries representing geographic subsets of a
-region). \pkg{spmodel} supports the binomial distribution for binary
+novel Laplace approach described by \citet{ver2024marginal}. These
+modeling tools are approachable and mirror the familiar \code{glm()}
+syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs
+relatively seamless. The \code{spglm()} function fits SPGLMs for
+point-referenced data (e.g., x-coordinates and y-coordinates
+representing point locations in a field; models are sometimes called
+``geostatistical'' models), while the \code{spgautor()} function fits
+SPGLMs for areal data (e.g., polygon boundaries representing geographic
+subsets of a region; models are sometimes called ``autoregressive''
+models). \pkg{spmodel} supports the binomial distribution for binary
 data, Poisson and negative binomial distributions for count data, Gamma
@@
 standard \proglang{R} helper functions like \code{summary()},
-\code{plot()}, and \code{cooks.distance()}. \pkg{spmodel} also provides
-tools to predict \(\mathbf{w}\) at new locations and quantify
-uncertainty in those prediction using \code{predict()}. This core
-functionality, combined with several advanced features we describe
-throughout the manuscript, enables \pkg{spmodel} to provide some novel
-and important capabilities previously missing from the existing SPGLM
-ecosystem in \proglang{R}.
+\code{plot()}, \code{fitted()}, and \code{tidy()}, among othres.
+\pkg{spmodel} also provides tools to predict \(\mathbf{w}\) at new
+locations and quantify uncertainty in those prediction using
+\code{predict()} and \code{augment()}. This core functionality, combined
+with several advanced features we describe throughout the manuscript,
+enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools
+previously missing from the existing \proglang{R} ecosystem.
 
@@
 computational complexity. On the other hand, \pkg{spmodel} does not
-require the construction of a mesh prior to modeling. \pkg{spmodel}
+require the construction of a mesh prior to modeling. \pkg{spmodel} also
 supports 20 different spatial covariances and models them directly,
@@
 directly using neighborhood distance and autoregressive models rather
-than relying on the polygon centroid (as in \pkg{sdmTMB}). \pkg{spmodel}
-also provides experimental design tools (e.g., analysis of variance,
-contrasts), supports \pkg{sf} objects in modeling and prediction
-functions \citep{pebesma2018sf}, has several specialized model
-diagnostics like leverage values and Cook's distances, and has analytic
-solutions for prediction standard errors. Other similarities and
+than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not
+be within the polygon's boundaries. \pkg{spmodel} also provides
+experimental design tools (e.g., analysis of variance, contrasts),
+supports \pkg{sf} objects in modeling and prediction functions
+\citep{pebesma2018sf}, has several specialized model diagnostics like
+leverage values and Cook's distances, and has analytic solutions for
+fixed effect and prediction standard errors. Other similarities and
 differences do exist between \pkg{sdmTMB} and \pkg{spmodel}, and both
@@
 fitting and prediction routines in \pkg{spmodel}. In
-Section\(~\)\ref{sec-applications}, we provide several applications of
-\pkg{spmodel} to spatial binary, count, skewed and proportion data with
-both point-referenced and areal supports. And in
-Section\(~\)\ref{sec-discussion}, we end with a discussion synthesizing
-\pkg{spmodel}'s contributions to the analysis of SPGLMs in \proglang{R}.
+Section\(~\)\ref{sec-applications}, we provide an overview of core SPGLM
+functionaility in \pkg{spmodel} by modeling moose presence in Alaska,
+United States (US). In Section\(~\)\ref{sec-applications2}, we model
+moose counts in Alaska, US; skewed lake conductivity in the Southwestern
+US; harbor seal abundance trend behavior in Alaska, US; and voter
+turnout rates in Texas, US. And in Section\(~\)\ref{sec-discussion}, we
+end with a discussion synthesizing \pkg{spmodel}'s contributions to the
+analysis of SPGLMs in \proglang{R}.
 
 \section{The spatial generalized linear model and
-marginalizatoin}\label{sec-spglm}
-
-\pkg{spmodel} implements the novel methods described in
-\citet{ver2024marginal} to fit SPGLMs, which leverages the Laplace
-approximation and marginalizes over both the latent \(\mathbf{w}\) and
-the fixed effects while accommodating spatial covariance. A beneficial
-aspect of this approach is that it formally maximizes a hierarchical GLM
-likelihood \citep{lee1996hierarchical, wood2017generalized}. This makes
+marginalization}\label{sec-spglm}
+
+The novel Laplace approach implemented in \pkg{spmodel} formally
+maximizes a hierarchical GLM likelihood
+\citep{lee1996hierarchical, wood2017generalized}, making
 likelihood-based statistics for model comparison like AIC
@@
 approaches \citep{wolfinger1993generalized}, which only specify the
-first two moments of a distribution. \citet{ver2024marginal} provides
-thorough details regarding the method and contextualizes its development
-which built upon similar methods
-\citep[\citet{bonat2016practical}]{evangelou2011estimation}. Next, we
-describe a brief overview of the approach and how it can be used for
-parameter estimation, inference, and prediction.
+first two moments of a distribution. Next, we describe a brief overview
+of the approach and how it can be used for several primary data analysis
+tasks \citep{tredennick2021practical} like model comparison, parameter
+estimation, inference, model diagnostics, and prediction.
 
@@
 We can write the SPGLM likelihood hierarchically as
-\begin{equation}\label{eq-marginal}
+\begin{align}\label{eq-marginal}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w},
-\end{equation} where \([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) is
-the density for the appropriate response distribution of \(\mathbf{y}\)
+\end{align} where \([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) is the
+density for the appropriate response distribution of \(\mathbf{y}\)
 (e.g., binomial, Poisson) given the latent \(\mathbf{w}\) and dispersion
@@
 covariance. Following \citet{harville1977maximum}, we can integrate
-\(\boldsymbol{\beta}\) out of Equation\(~\)\ref{eq-spglm}, which yields
-\begin{equation}\label{eq-marginal2}
+\(\boldsymbol{\beta}\) out of Equation\(~\)\ref{eq-marginal}, which
+yields \begin{align}\label{eq-marginal2}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\mathbf{w},
-\end{equation} where \([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\)
-is the restricted (i.e., residual) multivariate Gaussian density
+\end{align} where \([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\) is
+the restricted (i.e., residual) multivariate Gaussian density
 \citep{patterson1971recovery} for \(\mathbf{w}\) given the explanatory
 variables and covariance parameters. Equation\(~\)\ref{eq-marginal2} can
-be synonymous written after profiling the overall variance out of
+synonymously be written after profiling the overall variance out of
 \(\boldsymbol{\Sigma}\), which reduces the dimension of
@@
 \citep{wolfinger1994computing}. The restricted multivariate Gaussian
-density is given by \begin{equation}\label{eq-reml-def}
+density is given by \begin{align}\label{eq-reml-def}
 [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] = \frac{\exp(-\frac{1}{2}(\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}}) \boldsymbol{\Sigma}^{-1} (\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}})^T)}{(2 \pi)^{(n - p)/2} |\boldsymbol{\Sigma}|^{1/2}|\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X}|^{1/2}},
-\end{equation} where
+\end{align} where
 \(\tilde{\boldsymbol{\beta}} = (\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{w}\)
 and \(|\cdot|\) denotes the determinant. Next, let
-\begin{equation}\label{eq-marginal03}
+\begin{align}\label{eq-marginal03}
   \ell_\mathbf{w} = \log([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}])
-\end{equation} and rewrite Equation\(~\)\ref{eq-marginal2} as
-\begin{equation}\label{eq-marginal3}
+\end{align} and rewrite Equation\(~\)\ref{eq-marginal2} as
+\begin{align}\label{eq-marginal3}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \exp(\ell_\mathbf{w}) d\mathbf{w}.
-\end{equation} A second-order Taylor series expansion of
-\(\ell_\mathbf{w}\) around \(\hat{\mathbf{w}}\) yields
-\begin{equation}\label{eq-marginal4}
+\end{align} A second-order Taylor series expansion of
+\(\ell_\mathbf{w}\) around a point \(\hat{\mathbf{w}}\) yields
+\begin{align}\label{eq-marginal4}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\hat{\mathbf{w}}} + \mathbf{g}^T(\mathbf{w} - \hat{\mathbf{w}}) + \frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T \mathbf{G} (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w},
-\end{equation} where \(\mathbf{g}\) and \(\mathbf{G}\) are the gradient
-and Hessian, respectively, of \(\ell_\mathbf{w}\) with respect to
+\end{align} where \(\mathbf{g}\) and \(\mathbf{G}\) are the gradient and
+Hessian, respectively, of \(\ell_\mathbf{w}\) with respect to
 \(\mathbf{w}\). If \(\hat{\mathbf{w}}\) is a value for which
-\(\mathbf{g} = \mathbf{0}\), \begin{equation}\label{eq-marginal5}
+\(\mathbf{g} = \mathbf{0}\), \begin{align}\label{eq-marginal5}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\hat{\mathbf{w}}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T (-\mathbf{G}) (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w}.
-\end{equation} The integral in Equation\(~\)\ref{eq-marginal5} can be
+\end{align} The integral in Equation\(~\)\ref{eq-marginal5} can be
 solved by leveraging properties of the normalizing constant of a
@@
 \(\exp(\ell_{\hat{\mathbf{w}}})\) yields
-\begin{equation}\label{eq-marginal6}
+\begin{align}\label{eq-marginal6}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\hat{\mathbf{w}}), \varphi] [\hat{\mathbf{w}} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\hat{\mathbf{w}}}|^{-1/2}.
-\end{equation}
+\end{align}
 
@@
 likelihood methods. Had we observed \(\mathbf{w}\), a GLS estimator for
-\(\boldsymbol{\beta}\) is given by \begin{equation}\label{eq-gls1}
+\(\boldsymbol{\beta}\) is given by \begin{align}\label{eq-gls1}
   \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
-\end{equation} where
+\end{align} where
 \(\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\).
@@
 Hessian). It follows that \(\text{E}(\hat{\mathbf{w}})\) is given by
-\begin{equation}
+\begin{align}
   \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
@@
    = \mathbf{X}\boldsymbol{\beta}
-\end{equation} and \(\text{Var}(\hat{\mathbf{w}})\) is given by
+\end{align} and \(\text{Var}(\hat{\mathbf{w}})\) is given by
 \begin{align}
@@
   & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
-\end{align} Putting this all together, it follows that \begin{equation}
+\end{align} Putting this all together, it follows that \begin{align}
   \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
   = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}
-\end{equation} and \begin{align}
+\end{align} and \begin{align}
   \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
@@
   & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
-  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
-  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}
+  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
+  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}
 \end{align} In practice, \(\text{Var}(\hat{\boldsymbol{\beta}})\) is
@@
 Closed-form solutions are useful because they bypass the need for
-computationally expensive sampling-based strategies to evaluate the mean
-and variance of \(\hat{\boldsymbol{\beta}}\) -- a common technique for
-other approaches to SPGLMs like Bayesian MCMC.
+sampling-based strategies to evaluate the mean and variance of
+\(\hat{\boldsymbol{\beta}}\), a common technique for other approaches to
+SPGLMs like Bayesian MCMC.
 
@@
 relative to expectation. The response GLM residuals simply compare the
-observation to its fitted latent mean: \begin{equation}
+observation to its fitted latent mean: \begin{align}
   \mathbf{r}_{r} = \mathbf{y} - f^{-1}(\hat{\mathbf{w}})
-\end{equation} Because observations often have a unique support in a GLM
+\end{align} Because observations often have a unique support in a GLM
 (e.g., only two possible response values for binary data) and the
@@
 response residuals in a standard linear model. Deviance residuals are
-given by \begin{equation}
+given by \begin{align}
   \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
-\end{equation} where \(\mathbf{d}\) is a vector of individual deviances.
+\end{align} where \(\mathbf{d}\) is a vector of individual deviances.
 The sum of the squared deviance residuals equals the sum of
@@
 \(\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)\) for all \(i\)) and the
-fitted model. Deviance is often used as a fit statistic; lower values of
-deviance imply a better model fit. Pearson and standardized residuals
-are other types of GLM residuals that involve some scaling of the
-response residuals; the Pearson residuals scale \(\mathbf{r}_{r}\) by
-the squrae root of \(\mathbf{V}\), while the standardized residuals
-scale the deviance residuals by
+fitted model \citep{myers2012generalized}. Deviance is often used as a
+fit statistic; lower values of deviance imply a better model fit.
+Pearson and standardized residuals are other types of GLM residuals that
+involve a scaling of the response residuals; the Pearson residuals scale
+\(\mathbf{r}_{r}\) by the square root of \(\mathbf{V}\), while the
+standardized residuals scale the deviance residuals by
 \(\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}\), where \(\mathbf{L}_{ii}\) is
 the \(i\)th diagonal element of the leverage matrix, which we discuss
-next. An observation has high leverage if its combination of explanatory
+next.
+
+An observation has high leverage if its combination of explanatory
 variables is far away from other observations. In a linear model, the
-leverage values are the diagonal of the leverage (i.e., projection, hat)
-matrix,
+leverage (i.e., hat) values are the diagonal of the leverage (i.e.,
+projection, hat) matrix,
 \(\mathbf{L} = \mathbf{X}(\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top\).
-In a GLM, the leverage matrix is given by \begin{equation}
+In a GLM, the leverage matrix is given by \begin{align}
   \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2},
-\end{equation} where \(\mathbf{V}\) is a diagonal matrix with \(i\)th
+\end{align} where \(\mathbf{V}\) is a diagonal matrix with \(i\)th
 diagonal element equal to the variance of the response distribution
-evaluated at \(f^{-1}(\mathbf{w}_i)\) \citep{faraway2016extending}. The
-matrix \(\mathbf{V}\) is sometimes called the GLM weight matrix. The
-larger the \(i\)th diagonal element of the hat matrix, the more severe
-the leverage from the \(i\)th observation. An observation is influential
-if it has a sizeable impact on model fit. Influence is measured using
-Cook's distance \citep{cook1979influential, cook1982residuals}, which is
-given for a GLM by \begin{equation}
-  \mathbf{c} = \mathbf{r}^2_{s} \frac{diag(\mathbf{L})}{\text{tr}(\mathbf{L})(\mathbf{1} - diag(\mathbf{L}))},
-\end{equation} where \(\mathbf{r}^2_{s}\) are the standardized residuals
+evaluated at \(f^{-1}(\mathbf{w}_i)\) \citep{faraway2016extending};
+\(\mathbf{V}\) is sometimes called the GLM weight matrix. The larger the
+value of \(\mathbf{L}_{ii}\), the more severe the leverage from the
+\(i\)th observation.
+
+An observation is influential if it has a sizable impact on model fit.
+Influence is measured using Cook's distance
+\citep{cook1979influential, cook1982residuals}, which is given for a GLM
+by \begin{align}
+  \mathbf{c} = \frac{\mathbf{r}^2_{s}}{\text{tr}(\mathbf{L})} \frac{diag(\mathbf{L})}{(\mathbf{1} - diag(\mathbf{L}))},
+\end{align} where \(\mathbf{r}^2_{s}\) are the standardized residuals
 and \(diag(\mathbf{L})\) indicates the diagonal elements of the leverage
-matrix. The larger the \(i\)th diagonal element of the hat matrix, the
-more severe the influence from the \(i\)th observation.
+matrix. The larger the value of \(\mathbf{c}_i\), the more severe the
+influence from the \(i\)th observation.
 \citet{montgomery2021introduction} provide guidance for interpreting
@@
 proportion of variability in the data captured by the explanatory
-variables and is calculated as one minus the ratio of the error sum of
+variables. It is calculated as one minus the ratio of the error sum of
 squares to the total sum of squares \citep{rencher2008linear}. In a GLM,
-there are many ways to define such a statistic
-\citep{smith2013comparison}. One such approach is to use one minus the
-deviance ratio: \begin{equation}
+there are many ways to define a statistic that emulates the
+aforementioned meaning of \(R^2\) from the linear model
+\citep{smith2013comparison}. This statistic is called a pseudo R-squared
+(\(PR^2\)). One \(PR^2\) for GLMs simply replaces the sums of squares
+ratio from the linear model with the deviance ratio: \begin{align}
   PR^2 = 1 - \frac{deviance_{fit}}{deviance_{null}},
-\end{equation} where \(deviance_{fit}\) is the deviance of the fitted
-model (sometimes called the residual deviance) and \(deviance_{null}\)
-is the deviance of the model taking \(\mathbf{X} \equiv \mathbf{1}\), a
-column of all ones (i.e., an intercept-only model). In practice,
+\end{align} where \(deviance_{fit}\) is the deviance of the fitted model
+(sometimes called the residual deviance) and \(deviance_{null}\) is the
+deviance of the model taking \(\mathbf{X} \equiv \mathbf{1}\), a column
+of all ones (i.e., an intercept-only model). In practice,
 \(deviance_{null}\) is derived by computing \(\hat{\mathbf{w}}\) when
 \(\mathbf{X} \equiv \mathbf{1}\) given \(\hat{\boldsymbol{\theta}}\) and
-\(\hat{\varphi}\) from the fitted model. Like the standard \(R^2\), this
-statistic attempts to capture variability (i.e., deviance) attributable
-to the explanatory variables. Because the \(deviance_{null}\)
-denominator changes across fitted models (as the values of
-\(\hat{\boldsymbol{\theta}}\) and \(\hat{\varphi}\) change), this
-statistic should not be used as a model comparison tool. Instead, it
-should be used as an informative diagnostic tool unique to each model
-fit.
+\(\hat{\varphi}\) from the fitted model. Like \(R^2\), \(PR^2\) can be
+adjusted to account for the numbers of parameters estimated in a model.
+Because the \(deviance_{null}\) denominator changes across fitted models
+(as the values of \(\hat{\boldsymbol{\theta}}\) and \(\hat{\varphi}\)
+change), this statistic should not be used as a model comparison tool.
+Rather, it should be used as an informative diagnostic tool that unique
+to each model fit and describes how much variability from that model is
+attributable to the explanatory variables.
 
@@
 vector of latent means at the new locations that follows the same SPGLM
-from Equation\textasciitilde{}\ref{eq-spglm} with fixed effects design
-matrix, \(\mathbf{X}_{\mathbf{u}}\). The vector
-\((\mathbf{w}, \mathbf{u})^\top\) has the following properties:
-\begin{align}
-  \text{E}(\mathbf{w}, \mathbf{u})^\top & = (\text{E}(\mathbf{w}), \text{E}(\mathbf{u}))^\top = (\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top \\
-  \text{Var}(\mathbf{w}, \mathbf{u})^\top & = \begin{bmatrix} \text{Var}(\mathbf{w}, \mathbf{w}) & \text{Var}(\mathbf{w}, \mathbf{u}) \\ \text{Var}(\mathbf{u}, \mathbf{w}) & \text{Var}(\mathbf{u}, \mathbf{u}) \end{bmatrix} = \begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}
-\end{align} Because we have observed \(\mathbf{w}\), we may derive the
+from Equation\(~\)\ref{eq-spglm} and having fixed effects design matrix,
+\(\mathbf{X}_{\mathbf{u}}\). The vector
+\((\mathbf{w}, \mathbf{u})^\top\) has expectation
+\((\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top\)
+and covariance matrix
+\(\begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}\),
+where \(\boldsymbol{\Sigma} = \text{Var}(\mathbf{w}, \mathbf{w})\),
+\(\boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}} = \text{Var}(\mathbf{w}, \mathbf{u})\),
+\(\boldsymbol{\Sigma}_{\mathbf{u} \mathbf{w}} = \boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}}^\top\)
+and
+\(\boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} = \text{Var}(\mathbf{u}, \mathbf{u})\)
+By assumption, we have observed \(\mathbf{w}\), so we may derive the
 conditional distribution of \(\mathbf{u}|\mathbf{w}\), which has the
 following properties: \begin{align}
-  \text{E}(\mathbf{w} | \mathbf{u}) & = \mathbf{X}_{\mathbf{u}} \boldsymbol{\beta} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{w} - \mathbf{X}\boldsymbol{\beta}) \\
-  \text{E}(\mathbf{w} | \mathbf{u}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
+  \text{E}(\mathbf{u} | \mathbf{w}) & = \mathbf{X}_{\mathbf{u}} \boldsymbol{\beta} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{w} - \mathbf{X}\boldsymbol{\beta}) \\
+  \text{E}(\mathbf{u} | \mathbf{w}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
 \end{align} \citet{ver2024marginal} show how these equations are
@@
 \(\boldsymbol{\Lambda} = \mathbf{X}_{\mathbf{u}}\mathbf{B} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{1} - \mathbf{X}\mathbf{B})\)
-for a vector of ones, \(\mathbf{1}\).
-
-As with \(\hat{\boldsymbol{\beta}}\), in practice these covariance
-matrices are evaluated at \(\hat{\boldsymbol{\theta}}\). Moreover, these
-closed-form solutions provided enhance computational efficiency and
-clarity of the predictor's behavior.
+for a vector of ones, \(\mathbf{1}\). As with
+\(\hat{\boldsymbol{\beta}}\), in practice these covariance matrices are
+evaluated at \(\hat{\boldsymbol{\theta}}\).
 
@@
 The \code{moose} data in \pkg{spmodel} contain information on moose
-(Alces Alces) presence in the Togiak region of Alaska, USA.
-\code{moose}is an \code{sf} object, a special data frame that is
-supplemented with spatial information using the \pkg{sf} package in
-\proglang{R} \citep{pebesma2018sf}. The first few rows of \code{moose}
-look like:
+(Alces Alces) presence in the Togiak region of Alaska, USA. \code{moose}
+is an \code{sf} object, a special data frame that is supplemented with
+spatial information using the \pkg{sf} package in \proglang{R}
+\citep{pebesma2018sf}. The first few rows of \code{moose} look like:
 
@@
 because each observation occurs at point coordinates and are represented
-by a \code{POINT} geometry. The \code{moose_preds} data in \pkg{spmodel}
-contain spatial locations at which predictions of moose presence are
-desired (and is also point-referenced). \code{moose_preds} is also an
-\code{sf} object with measurements for \code{elev} and \code{strat} and
-the same projection system. Figure\(~\)\ref{fig-moose-data} shows the
-\code{presence} variable in \code{moose} as well as the spatial
-locations of both \code{moose} and \code{moose_preds}. Moose are most
-commonly present in the southwestern and eastern parts of the domain and
-least commonly present in the northwest
-(Figure\(~\)\ref{fig-moose-data}). Next we show how to use \pkg{spmodel}
-to study the effect of elevation and strata on moose presence while
-accounting for spatial covariance and to make predictions of moose
-presence at new locations.
+by a \code{POINT} geometry. Moose are most prevalent in the southwestern
+and eastern parts of the Togiak region
+(Figure\(~\)\ref{fig-moose-data}).
 
@@
 
+The \code{moose_preds} data in \pkg{spmodel} is an \code{sf} object with
+point locations at which moose presence predictions are desired. Like
+\code{moose}, \code{moose_preds} contains \code{elev} and \code{strat}
+for each site:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> head(moose_preds)
+\end{CodeInput}
+\begin{CodeOutput}
+Simple feature collection with 6 features and 2 fields
+Geometry type: POINT
+Dimension:     XY
+Bounding box:  xmin: 291839.8 ymin: 1436192 xmax: 401239.6 ymax: 1512103
+Projected CRS: NAD83 / Alaska Albers
+# A tibble: 6 x 3
+   elev strat           geometry
+  <dbl> <chr>        <POINT [m]>
+1  143. L     (401239.6 1436192)
+2  324. L     (352640.6 1490695)
+3  158. L     (360954.9 1491590)
+4  221. M     (291839.8 1466091)
+5  209. M     (310991.9 1441630)
+6  218. L     (304473.8 1512103)
+\end{CodeOutput}
+\end{CodeChunk}
+
 \subsection{Model Fitting}\label{model-fitting}
@@
 SPGLMs in \pkg{spmodel} are fit using the \code{spglm()} function. The
-\code{spglm()} function requires four arguments: \texttt{formula}, the
+\code{spglm()} function requires four arguments: \code{formula}, the
 relationship between the response and explanatory variables;
-\texttt{family}, the response distribution assumed for the repsonse
-variable; \texttt{data}, the data frame that contains the variables in
-\texttt{formula}, and \texttt{spcov\_type}, the type of spatial
-covariance. These first three arguments are the three required arguments
-to \code{glm()} for nonspatial GLMs. So, the transition from
-\code{glm()} to \code{spglm()} simply requires one additional argument:
-\code{spcov_type}. When \code{data} is not an \code{sf} object,
-\code{spglm()} also requires the \code{xcoord} and \code{ycoord}
-arguments, which indicate the columns in \code{data} that represent the
-x- and y-coordinates, respectively (it is assumed these coordinates are
-already projected).
-
-We use \code{spglm()} to fit a spatial logistic regression model
-quantifying the effect of elevation and strata on moose presence:
+\code{family}, the response distribution assumed for the repsonse
+variable; \code{data}, the data frame that contains the variables in
+\code{formula}, and \code{spcov_type}, the type of spatial covariance.
+The \code{formula}, \code{family}, and \code{data} arguments are the
+three required arguments to \code{glm()} for nonspatial GLMs. So, the
+transition from \code{glm()} to \code{spglm()} simply requires one
+additional argument: \code{spcov_type}. When \code{data} is not an
+\code{sf} object, \code{spglm()} also requires the \code{xcoord} and
+\code{ycoord} arguments, which indicate the columns in \code{data} that
+represent projected the x- and y-coordinates, respectively.
+
+We use \code{spglm()} to fit a spatial GLM (i.e., here, a spatial
+logistic regression) quantifying the effect of elevation and strata on
+moose presence:
 
@@
 
-Based on this model, there is some evidence that elevation is associated
-with higher probabilities of moose presence (\(p\)-value \(\approx\)
-0.087) but strong evidence that moose are more prevalent in the
-\code{"M"} strata than the \code{"L"} strata (\(p\)-value \textless{}
-0.001). The fixed effects coefficients table from \code{summary()} is
-often of primary practical interest, but it is not easily usable when
+The model suggests that elevation is positively associated with the log
+odds of moose presence (\(p\)-value \(\approx\) 0.087), after
+controlling for strata. The model also suggests that moose have a higher
+log odds of presence in the \code{"M"} strata compared to the \code{"L"}
+strata (\(p\)-value \textless{} 0.001), after controlling for elevation.
+
+The fixed effects coefficients table from \code{summary()} is often of
+primary scientific interest, but it is not immediately usable when
 printed directly to the \proglang{R} console. The \code{tidy()} function
@@
   <chr>          <dbl>     <dbl>     <dbl>    <dbl>    <dbl>     <dbl>
-1 (Intercept) -2.47      1.49        -1.66 0.0971   -5.38        0.447 
-2 elev         0.00604   0.00353      1.71 0.0869   -0.000873    0.0129
-3 stratM       1.44      0.421        3.42 0.000622  0.615       2.26  
+1 (Intercept) -2.47      1.49        -1.66 0.0971   -5.38e+0    0.447 
+2 elev         0.00604   0.00353      1.71 0.0869   -8.73e-4    0.0129
+3 stratM       1.44      0.421        3.42 0.000622  6.15e-1    2.26  
 \end{CodeOutput}
@@
 # A tibble: 1 x 10
-      n     p  npar value   AIC  AICc   BIC logLik deviance pseudo.r.squared
-  <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>            <dbl>
-1   218     3     3  676.  682.  683.  693.  -338.     176.           0.0627
+      n     p  npar value   AIC  AICc   BIC logLik deviance
+  <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>
+1   218     3     3  676.  682.  683.  693.  -338.     176.
+# i 1 more variable: pseudo.r.squared <dbl>
 \end{CodeOutput}
@@
 # A tibble: 1 x 10
-      n     p  npar value   AIC  AICc   BIC logLik deviance pseudo.r.squared
-  <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>            <dbl>
-1   218     3     0  708.  708.  708.  708.  -354.     294.           0.0280
+      n     p  npar value   AIC  AICc   BIC logLik deviance
+  <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>
+1   218     3     0  708.  708.  708.  708.  -354.     294.
+# i 1 more variable: pseudo.r.squared <dbl>
 \end{CodeOutput}
@@
 \begin{CodeInput}
-R> tidy(anova(spbin, bin))
-\end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 1 x 5
-  full  reduced    df statistic     p.value
-  <chr> <chr>   <int>     <dbl>       <dbl>
-1 spbin bin         3      31.5 0.000000652
-\end{CodeOutput}
-\end{CodeChunk}
-
-The LRT test statistic is \(\approx\) 31.5 with three degrees of freedom
-(the difference in covariance parameters), yielding a \(p\)-value
-\(< 0.001\) that indicates preference for the full model (SPGLM)
-relative to the reduced model (GLM).
+R> anova(spbin, bin)
+\end{CodeInput}
+\begin{CodeOutput}
+Likelihood Ratio Test
+
+Response: presence
+             Df   Chi2 Pr(>Chi2)    
+spbin vs bin  3 31.546 6.525e-07 ***
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{CodeOutput}
+\end{CodeChunk}
+
+The LRT provides strong evidence that the SPGLM is preferred to the GLM
+(\(p\)-value \textless{} 0.001).
 
@@
 hold-out observation, recomputing estimates of \(\boldsymbol{\beta}\)
-each time. Then, statistics like bias, mean-squared-prediction error
+each iteration Then, statistics like bias, mean-squared-prediction error
 (MSPE), and the square root of MSPE (RMSPE) can be used to evaluate
@@
 \begin{CodeInput}
-R> spbin2 <- update(spbin, spcov_type = "spherical")
+R> spbin2 <- update(spbin, spcov_type = "gaussian")
 R> glances(spbin, spbin2)
@@
   <chr>  <int> <dbl> <int> <dbl> <dbl> <dbl> <dbl>  <dbl>    <dbl>
-1 spbin2   218     3     3  675.  681.  681.  691.  -338.     180.
+1 spbin2   218     3     3  674.  680.  680.  690.  -337.     198.
 2 spbin    218     3     3  676.  682.  683.  693.  -338.     176.
@@
       <dbl> <dbl> <dbl>
-1 0.000121 0.155 0.394
+1 -0.000261 0.146 0.382
 \end{CodeOutput}
@@
 The \code{"exponential"} spatial covariance (\code{spbin}) has a
-slightly lower deviance but slightly higher AIC, AICc, and BIC than the
-\code{"spherical"} spatial covariance (\code{spbin2}). Both spatial
-covariance functions nearly identical leave-one-out cross validation
-metrics. For practical purposes, these models fit quite similarly.
+slightly lower (better) deviance but slightly higher (worse) AIC, AICc,
+and BIC than the \code{"gaussian"} spatial covariance (\code{spbin2}).
+Both spatial covariance functions have similar leave-one-out cross
+validation metrics, though the \code{"gaussian"} spatial covariance
+RMSPE is slightly lower (better). For practical purposes, these models
+fit similarly.
+
+Often times in spatial statistics, the difference in model fit between
+the best spatial model and worst spatial model is much smaller than the
+difference in model fit between the worst spatial model and the
+nonspatial model, implying that accounting for some form of spatial
+covariance is very beneficial. Two spatial covariance functions to
+consider starting with are the exponential and Gaussian, which have
+quite different origin behaviors (Figure\(~\)\ref{fig-type}), something
+\citet{stein1999interpolation} argues is important to characterize
+accurately.
 
@@
 
-\code{spmodel} provides a suite of tools for model diagnostics. The
-\code{augment()} function augments the model data with diagnostics:
+\code{spmodel} provides a suite of tools for model diagnostics. One is
+\code{augment()}, which augments the model data with diagnostics:
 
@@
 The fitted values (\code{.fitted}) can be returned on either the link
-(\(\hat{\mathbf{w}}\)) or response (\(f^{-1}\hat{\mathbf{w}}\)) scale
-and the residuals (\code{.resid}) can deviance, pearson, or response
-residuals. The defaults are the link scale and deviance residuals,
-respectively. Also returned by \code{augment()} are the leverage
-(\code{.hat}), Cook's distance (\code{.cooksd}), and standardized
-residuals \code{.std.resid}. A benefit of using \code{augment()} when
-\code{data} is an \code{sf} object is that the output is also an
-\code{sf} object, which makes it straightforward to create spatial
-diagnostic plots (Figure\(~\)\ref{fig-sp-diagnostic}). Standard
-\proglang{R} helpers (e.g., \code{fitted()}, \code{residuals()}) are
-available to alternatively extract model diagnostics from the model
-object.
+(\(\hat{\mathbf{w}}\)) or response (\(f^{-1}(\hat{\mathbf{w}})\)) scale
+and the residuals (\code{.resid}) can deviance, Pearson, or response
+residuals. The default fitted values are the link scale and the default
+residuals are deviance residuals. Also returned by \code{augment()} are
+the leverage (\code{.hat}), Cook's distance (\code{.cooksd}), and
+standardized residuals \code{.std.resid}. A benefit of using
+\code{augment()} when \code{data} is an \code{sf} object is that the
+output is also an \code{sf} object, which makes it straightforward to
+create spatial diagnostic plots (Figure\(~\)\ref{fig-sp-diagnostic}).
+Standard \proglang{R} helpers (e.g., \code{fitted()},
+\code{residuals()}) are also available to extract model diagnostics from
+the model object.
 
@@
 diagnostics as from \code{lm()} and \code{glm()} with additional tools
-for spatial covariance. For example, we can inspect Cook's distance
-values and the empirical spatial covariance as a function
-(Figure\(~\)\ref{fig-sp-diagnostic2}) with
+for diagnosing spatial covariance. For example, we can inspect Cook's
+distance values and the empirical spatial covariance as a function of
+distance with (Figure\(~\)\ref{fig-sp-diagnostic2}):
 
@@
 The \code{varcomp()} function partitions model variability into several
-different components:
+different components, helping to elucidate the model's structure:
 
@@
 
-The pseudo R-squared (\(PR^2\)), the proportion of variability
-attributable to the explanatory variables, is reported in the first row.
-The remaining variability (\(1 - PR^2\)) is allocated proportionally to
+The pseudo R-squared (\(PR^2\)) is reported in the first row. The
+remaining variability (\(1 - PR^2\)) is allocated proportionally to
 \code{de} and \code{ie} according to \(\sigma^2_{de}\) and
@@
 Predictions on the response scale are visualized alongside the fitted
-values (\(f^{-1}\hat{\mathbf{w}}\)) in Figure\(~\)\ref{fig-moose-fit}.
+values (\(f^{-1}(\hat{\mathbf{w}})\)) in Figure\(~\)\ref{fig-moose-fit}.
 
@@
 Prediction intervals for the probability of moose presence (on the link
-scale) are returned via \code{interval}
+scale) are returned by suppplying \code{interval}:
 
@@
 models with and without geometric anisotropy for the point-referenced
-moose count data. In Section\(~\)\ref{sec-seal}, we fit a binomial model
-to the areal seal trend data with a nonspatial random effect. In
-Section\(~\)\ref{sec-texas}, we fit beta models to Texas voter turnout
-data. We explore how the Texas data can be treated as point-referenced
-or areal and compare the two spatial covariance structures empirically.
-We also highlight how the maximum likelihood estimation method is useful
-when comparing two models with different explanatory variables. Finally,
-in Section\(~\)\ref{sec-lake}, we fit a Gamma model to the
-point-referenced lake conductivity data. We show how to perform a
-spatial analysis of variance (ANOVA) and leverage modeling functions
-from other \proglang{R} packages like \pkg{emmeans} and \pkg{car}.
+moose count data. In Section\(~\)\ref{sec-lake}, we fit a Gamma model to
+the point-referenced lake conductivity data, showing how to perform a
+spatial analysis of variance (ANOVA), fit contrasts for models with
+interactions and use maximum likelihood to compare two models with
+different explanatory variables. In Section\(~\)\ref{sec-seal}, we fit a
+binomial model to the areal harbor seal trend data with a nonspatial
+random effect. Finally in Section\(~\)\ref{sec-texas}, we fit beta
+models to Texas voter turnout data, which can be treated as
+point-referenced or areal.
 
@@
 In addition to moose presence, moose counts are also recorded in
-\texttt{moose} (Figure\(~\)\ref{fig-moose-data-count}. The Poisson and
+\code{moose} (Figure\(~\)\ref{fig-moose-data-count}. The Poisson and
 negative binomial response distributions can be used to model SPGLMs for
-count data. Using a Gaussian spatial covariance function, we may fit
-both a Poisson and negative binomial SPGLM changing the \texttt{family}
-argument:
+count data. The Poisson distribution mean is equal to its variance,
+while the negative binomial has an extra parameter to accommodate
+overdispersion (where the variance is larger than the mean). Using a
+spherical spatial covariance function, we may fit both a Poisson and
+negative binomial SPGLM changing the \code{family} argument:
 
@@
 +   data = moose,
-+   spcov_type = "gaussian"
++   spcov_type = "spherical"
 + )
@@
        df      BIC
-sppois  3 1343.892
-spnb    4 1340.706
+sppois  3 1344.574
+spnb    4 1343.105
 \end{CodeOutput}
@@
 coordinates, yielding transformed coordinates that are geometrically
-isotropic: \begin{equation}
+isotropic: \begin{align}
   \begin{bmatrix}
@@
   \end{bmatrix}.
-\end{equation} The parameter \(\alpha\) controls the rotation angle and
-the parameter \(\omega\) controls the scaling\ldots{} Then using these
-transformed coordinates, the partial sill (\(\sigma^2_{de}\)), nugget
-(\(\sigma^2_{ie}\)), and range (\(\phi\)) parameters are estimated.
+\end{align} The parameters \(\omega\) and \(\alpha\) controls the
+scaling and rotation, respectively, of the major and minor axes of a
+level curve of equal correlation (Figure\(~\)\ref{fig-tropy}). Using
+these transformed coordinates, the partial sill (\(\sigma^2_{de}\)),
+nugget (\(\sigma^2_{ie}\)), and range (\(\phi\)) parameters are
+estimated. We accommodate geometric anisotropy by supplying
+\code{anisotropy}:
 
@@
 
-According to BIC, the spatial negative binomial model performs best.
-This model simultaneously accounts for anisotropy and overdispersion.
+According to BIC, the spatial negative binomial model with geometric
+anisotropy performs best:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> BIC(sppois, spnb, sppois_anis, spnb_anis)
+\end{CodeInput}
+\begin{CodeOutput}
+            df      BIC
+sppois       3 1344.574
+spnb         4 1343.105
+sppois_anis  5 1341.143
+spnb_anis    6 1339.714
+\end{CodeOutput}
+\end{CodeChunk}
+
+The \code{plot()} function can be used to visualize the anisotropy
+(Figure\(~\)\ref{fig-tropy}):
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> plot(spnb, which = 8)
+R> plot(spnb_anis, which = 8)
+\end{CodeInput}
+\end{CodeChunk}
+
 The spatial correlation is strongest in a northwest-southeast direction
@@
 patterns in moose counts from Figure\(~\)\ref{fig-moose-data-count}.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> BIC(sppois, spnb, sppois_anis, spnb_anis)
-\end{CodeInput}
-\begin{CodeOutput}
-            df      BIC
-sppois       3 1343.892
-spnb         4 1340.706
-sppois_anis  5 1337.610
-spnb_anis    6 1335.051
-\end{CodeOutput}
-\end{CodeChunk}
-
-The \texttt{plot()} function can be used to visualize the anisotropy
-(Figure\(~\)\ref{fig-tropy}):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> plot(spnb, which = 8)
-R> plot(spnb_anis, which = 8)
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 \code{log_cond} variable in \code{lake} is the logarithm of
-conductivity, and we can dynamically exponentiate within \code{formula}
-so that it is on the original scale:
+conductivity, which we dynamically exponentiate within \code{formula} so
+that it is on the original scale:
 
@@
 origin, and we allow the effect of temperature to vary by state
-(\code{temp:state}). The partition factor restricts spatial covariance
-to apply only for two observations sampled during the same year. Data
-were collected in 2012 and 2017, so the \code{year} partition factor
-assumes independence between observations in 2012 and 2017. While we
-used the partition factor here illustratively, more generally, the
-utility of partition factors can be highly context dependent.
+(\code{temp:state}). The \code{year} partition factor (specified via
+\code{partition_factor}) restricts spatial covariance to be nonzero only
+for observations sampled during the same year. Data were collected in
+2012 and 2017, so this partition factor assumes independence between
+observations in 2012 and 2017. While we used the partition factor here
+illustratively, more generally, the utility of partition factors can be
+highly context dependent.
 
@@
 \end{CodeInput}
-\begin{CodeOutput}
-Warning: package 'car' was built under R version 4.4.3
-\end{CodeOutput}
-\begin{CodeOutput}
-Loading required package: carData
-\end{CodeOutput}
-\begin{CodeOutput}
-Warning: package 'carData' was built under R version 4.4.3
-\end{CodeOutput}
-\begin{CodeOutput}
-
-Attaching package: 'car'
-\end{CodeOutput}
-\begin{CodeOutput}
-The following object is masked from 'package:dplyr':
-
-    recode
-\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
 \begin{CodeInput}
@@
 upon a specific temperature value. By default, \pkg{emmeans} uses the
-mean temperature value to assess contrasts:
+mean temperature value (here, 7.63) to assess contrasts:
 
@@
 \end{CodeInput}
-\begin{CodeOutput}
-Welcome to emmeans.
-Caution: You lose important information if you filter this package's results.
-See '? untidy'
-\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
 \begin{CodeInput}
@@
 
-Again because of the interaction between \code{temp} and \code{state},
+Again, because of the interaction between \code{temp} and \code{state},
 we should assess temperature trends separately for each state:
@@
 (Figure\(~\)\ref{fig-seal}). A haulout location is a spot on coastal
-rocks that harbor seals go to rest, molt, and give birth, among other
-tasks.
+rocks that harbor seals go to rest, molt, and give birth.
 
@@
 For each polygon, a Poisson regression was used to quantify the mean
-trend in abundance over approximately 30 years. If the logarithm of seal
-abundance (\code{log_trend}) is negative (positive), it means abundance
-is decreasing (increasing). We use a binomial SPGLM to quantify how
-likely it is for abundance trends to be decreasing:
+trend in abundance over approximately 30 years \citep{ver2018spatial}.
+If the logarithm of mean abundance trends (\code{log_trend}) is negative
+(positive), it means abundance is decreasing (increasing). We use a
+binomial SPGLM to quantify the likelihood that mean abundance trends are
+decreasing:
 
@@
 function. Conditional and simultaneous autoregressive functions
-characterize spatial distance through neighborhood relationships and
-have \code{spcov_type} values of \code{"car"} and \code{"sar"},
-respectively. By default, Queen's distance is used to determine whether
-two sites are neighbors, though custom neighborhood matrices are passed
-via \code{W}. Row standardization is also assumed by default, though
-this can be changed via \code{row_st}. Using \code{random}, we also
-specified a nonspatial random effect for seal stock. The \code{random}
-argument uses similar syntax as \pkg{lme4} \citep{bates2015lme4} and
-\pkg{nlme} \citep{pinheiro2006mixed} to specify nonspatial random
-effects.
-
-Tidying the model reveals the mean probability that trends are
-decreasing is between -0.979 and 1.66 (on the logit scale):
+characterize spatial distance through neighborhood relationships (rather
+than Euclidean distance) and have \code{spcov_type} values of
+\code{"car"} and \code{"sar"}, respectively. By default, Queen's
+distance is used to determine whether two sites are neighbors, though
+custom neighborhood matrices are passed via \code{W}. Row
+standardization is also assumed by default; this can be changed via
+\code{row_st}. Using \code{random}, we also specified a nonspatial
+random effect for seal stock. The \code{random} argument uses similar
+syntax as \pkg{lme4} \citep{bates2015lme4} and \pkg{nlme}
+\citep{pinheiro2006mixed} to specify nonspatial random effects.
+
+Tidying the model reveals the estimates and confidence intervals on the
+log odds scale:
 
@@
 
-Back-transforming this interval to the probability scale yields:
+Back-transforming the confidence interval to the probability scale
+yields:
 
@@
 
-For areal data, the prediction locations must be specified at the time
-of model fitting, as they affect the spatial covariance function's
-neighborhood structure. Prediction locations whose response values have
-an \code{NA} (i.e., missing) value are converted into a \code{newdata}
-object that is stored in the model output. For example, rows one and
-nine are locations without seal trends, meaning they are not used in
-model fitting but are desired for prediction:
+In contrast to point-referenced data, prediction locations for areal
+data must be specified at the time of model fitting, as they affect the
+spatial covariance function's neighborhood structure. Prediction
+locations whose response values have an \code{NA} (i.e., missing) value
+are converted into a \code{newdata} object that is stored in the model
+output. For example, rows one and nine are locations without seal
+trends, meaning they are not used in model fitting but are desired for
+prediction:
 
@@
  *     <dbl> <fct>                                       <POLYGON [m]>
- 1  NA       8     ((1035002 1054710, 1035002 1054542, 1035002 1053542, 1035002~
- 2  -0.282   8     ((1037002 1039492, 1037006 1039490, 1037017 1039492, 1037035~
- 3  -0.00121 8     ((1070158 1030216, 1070185 1030207, 1070187 1030207, 1070211~
- 4   0.0354  8     ((1054906 1034826, 1054931 1034821, 1054936 1034822, 1055001~
- 5  -0.0160  8     ((1025142 1056940, 1025184 1056889, 1025222 1056836, 1025256~
- 6   0.0872  8     ((1026035 1044623, 1026037 1044605, 1026072 1044610, 1026083~
- 7  -0.266   8     ((1100345 1060709, 1100287 1060706, 1100228 1060706, 1100170~
- 8   0.0743  8     ((1030247 1029637, 1030248 1029637, 1030265 1029642, 1030328~
- 9  NA       8     ((1043093 1020553, 1043097 1020550, 1043101 1020550, 1043166~
-10  -0.00961 8     ((1116002 1024542, 1116002 1023542, 1116002 1022542, 1116002~
+ 1  NA       8     ((1035002 1054710, 1035002 1054542, 1035002 105354~
+ 2  -0.282   8     ((1037002 1039492, 1037006 1039490, 1037017 103949~
+ 3  -0.00121 8     ((1070158 1030216, 1070185 1030207, 1070187 103020~
+ 4   0.0354  8     ((1054906 1034826, 1054931 1034821, 1054936 103482~
+ 5  -0.0160  8     ((1025142 1056940, 1025184 1056889, 1025222 105683~
+ 6   0.0872  8     ((1026035 1044623, 1026037 1044605, 1026072 104461~
+ 7  -0.266   8     ((1100345 1060709, 1100287 1060706, 1100228 106070~
+ 8   0.0743  8     ((1030247 1029637, 1030248 1029637, 1030265 102964~
+ 9  NA       8     ((1043093 1020553, 1043097 1020550, 1043101 102055~
+10  -0.00961 8     ((1116002 1024542, 1116002 1023542, 1116002 102254~
 # i 139 more rows
@@
 We further explore comparisons between point-referenced and aeral data
-next.
+in the next example.
 
@@
 (Figure\(~\)\ref{fig-texas}). Beta regression is a GLM used to model
-rate and proportion data in (0, 1)
-\citep[\citet{cribari2010beta}]{ferrari2004beta}. We model voter turnout
-rates as a function of mean log income of county residents using an
-SPGLM assuming a beta distributed response variable:
+rate and proportion data in the (0, 1) interval
+\citep{ferrari2004beta, cribari2010beta}. We model voter turnout rates
+as a function of mean log income of county residents using an SPGLM
+assuming a beta distributed response variable:
 
@@
 \code{spbeta_geo} model fit with REML (e.g., via \code{tidy()}), an
-alternative approach is to use a likelihood ratio test as long as the
-two models are nested. We refit the point-referenced model using maximum
-likelihood and compare it to a similar model without log income using
-the likelihood ratio test, accessible via \code{anova()} when providing
-two models:
+alternative approach is to use a likelihood ratio test:
 
@@
 R> spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
-R> spbeta_red_ml <- update(spbeta_full_ml, formula = turnout ~ 1)
+R> spbeta_red_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
 R> anova(spbeta_full_ml, spbeta_red_ml)
@@
 The likelihood ratio test suggests that log income is significantly
-related to voter turnout. The \code{anova()} output can also be tidied
-using \code{tidy()}.
+related to voter turnout (\(p\)-value \textless{} 0.001). We could
+instead use a different likelihood-based statistic like AIC:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> AIC(spbeta_full_ml, spbeta_red_ml)
+\end{CodeInput}
+\begin{CodeOutput}
+               df       AIC
+spbeta_full_ml  7 -31.25900
+spbeta_red_ml   6 -10.10354
+\end{CodeOutput}
+\end{CodeChunk}
+
+The AIC prefers the full model, which suggests that log income is
+important for predicting voter turnout.
 
@@
 SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace
-approximation that marginalizes over the latent (i.e., unobserved) mean,
-\(\mathbf{w}\), and the fixed effects, \(\boldsymbol{\beta}\). The
-approach is quite flexible and accommodates any general response
-distribution and covariance structure. \pkg{spmodel}'s \code{spglm()}
-and \code{spgautor()} functions are similar in structure and syntax as
-base \proglang{R}'s \code{glm()} function, easing the transition for
-practitioners from GLMs to SPGLMs. These functions support six response
-distributions for binary, count, and skewed data and support 20 spatial
-covariance functions. \pkg{spmodel} provides several additional features
-that are not covered here, including fitting multiple models
-simultaneously, fixing spatial covariance parameters at known values,
-fitting models to large non-Gaussian data having thousands of
-observations via spatial indexing \citep{ver2023indexing}, incorporating
-spatial dependence in machine learning (e.g., random forests;
-\citet{breiman2001random}), simulating spatially dependent data (e.g.,
-\code{spbinom()}, \code{sprpois()}, etc.), and more. Further details are
-provided by \url{https://CRAN.R-project.org/package=spmodel} and links
-therein.
+approximation that simultaneously marginalizes over the latent (i.e.,
+unobserved) random effects and the fixed effects,
+\(\boldsymbol{\beta}\). The approach is very flexible and accommodates
+general response distributions and covariance structures, though here we
+focus on spatial applications. \pkg{spmodel}'s \code{spglm()} and
+\code{spgautor()} fit SPGLMs that are similar in structure and syntax as
+base \proglang{R}'s \code{glm()} function, easing the transition from
+GLMs to SPGLMs for practitioners. The \code{spglm()} and
+\code{spgautor()} functions support six response distributions for
+binary, count, and skewed data and 20 spatial covariance functions.
+\pkg{spmodel} also provides a suite of tools for data visualization,
+inference, model diagnostics, and prediction, providing a framework that
+can be used for all stages of a data analysis. There are many additional
+\pkg{spmodel} features that are not covered here, including fitting
+multiple models simultaneously, fixing spatial covariance and dispersion
+parameters at known values, fitting models to large non-Gaussian data
+having thousands of observations via spatial indexing
+\citep{ver2023indexing}, incorporating spatial dependence in machine
+learning (e.g., random forests; \citet{breiman2001random}), simulating
+spatially dependent data (e.g., \code{spbinom()}, \code{sprpois()},
+etc.), and more. Further details are provided by
+\url{https://CRAN.R-project.org/package=spmodel} and links therein.
 
@@
 The results in this paper were obtained using {R} 4.4.0 with the
-\pkg{spmodel} 0.9.0 package. Figures were created using the {ggplot2}
+\pkg{spmodel} 0.11.0 package. Figures were created using the {ggplot2}
 3.5.1 package \citep{wickham2016ggplot2} and base {R}.

### 2025-09-11 inst/manuscript/manuscript.Rmd
some updates

@@
   \usepackage{autonum}
+  \usepackage{adjustbox}
 output: rticles::jss_article
@@
 library(dplyr)
-#library(emmeans)
-#library(car)
+library(emmeans)
+library(car)
 library(here)
@@
 In practice, non-Gaussian data (e.g., binary, count, skewed, and proportion data) are ubiquitous.
-Non-Gaussian that belong to an exponential family data can be naturally modeled using a generalized linear model (GLM) regression framework [@nelder1972generalized; @mccullagh1989generalized].
+Non-Gaussian data that belong to an exponential family can be naturally modeled using a generalized linear model (GLM) regression framework [@nelder1972generalized; @mccullagh1989generalized].
 In a GLM, an $n \times 1$ response variable $\mathbf{y}$ belongs to a statistical distribution (e.g., Binomial, Poisson) with some mean and variance.
@@
 It is often convenient to compute the maximum likelihood estimates using the iteratively reweighted least squares (IRWLS) algorithm [@wood2017generalized], which is the approach used by the  \code{glm()} function in the \proglang{R} programming language [@rcore2024].
-GLMs add an additional layer of complexity compared to linear regression models, as the left-hand size of Equation$~$\ref{eq-glm} is a function of the mean of $\mathbf{y}$ rather than $\mathbf{y}$ itself (as in linear regression models).
+GLMs add an additional layer of complexity compared to linear regression models, as the left-hand side of Equation$~$\ref{eq-glm} is a function of the mean of $\mathbf{y}$ rather than $\mathbf{y}$ itself (as in linear regression models).
 
@@
 The matrix $\mathbf{R}$ is an $n \times n$ spatial correlation matrix that depends on a range parameter controlling the distance-decay rate of the spatial correlation.
-One example of a spatial covariance matrix is the "exponential", which is given by
+One example of a spatial covariance matrix is the "exponential," which is given by
 \begin{align}\label{eq-spcov-exp}
-  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de} \mathbf{R}_{exp} =  \sigma^2_{de} \exp(-\mathbf{H}/\phi),
+  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{\tau} \mathbf{R}_{exp} =  \sigma^2_{\tau} \exp(-\mathbf{H}/\phi),
 \end{align}
@@
 Often in spatial statistics, quantities are explicitly referenced with respect to $\mathbf{s}$, a vector of coordinates indexing the observation [@cressie1993statistics].
-For example, $\mathbf{y}$ and $\mathbf{X}$ may instead be written $\mathbf{y}(\mathbf{s})$ and $\mathbf{X}$, respectively.
+For example, $\mathbf{y}$ and $\mathbf{X}$ may instead be written $\mathbf{y}(\mathbf{s})$ and $\mathbf{X}(\mathbf{s})$, respectively.
 We acknowledge the utility of this nomenclature but drop the explicit dependence on $\mathbf{s}$ for simplicity of notation.
@@
 Working with precision matrices, however, can be more restrictive and less intuitive than working directly with the covariance matrix.
+
 The \pkg{FRK} [@sainsbury2024modeling], \pkg{glmmTMB} [@brooks2017glmmtmb], \pkg{hglm} [@ronnegard2010hglm], \pkg{mgcv} [@wood2017generalized], and \pkg{spaMM} [@rousset2014spamm] packages directly use Laplace, quasi-likelihood, or reduced-rank approaches to estimate parameters.
@@
 The novel Laplace approach is particularly attractive for two reasons.
-First, it is general enough that can be applied to any covariance structure (not just spatial).
+First, it is general enough that it can be applied to any covariance structure (not just spatial).
 Second, after estimating the covariance parameters, analytical solutions exist for the fixed effects (and their standard errors) as well as predictions of the latent $\mathbf{w}$ at new locations (and their standard errors).
@@
 The \code{spglm()} function fits SPGLMs for point-referenced data (e.g., x-coordinates and y-coordinates representing point locations in a field; models are sometimes called "geostatistical" models), while the \code{spgautor()} function fits SPGLMs for areal data (e.g., polygon boundaries representing geographic subsets of a region; models are sometimes called "autoregressive" models).
-\pkg{spmodel} supports the binomial distribution for binary data, Poisson and negative binomial distributions for count data, Gamma and inverse Gaussian distributions for skewed data, and the beta distribution for proportion data.
+For both point-referenced data and areal data, \pkg{spmodel} supports the binomial distribution for binary data, Poisson and negative binomial distributions for count data, Gamma and inverse Gaussian distributions for skewed data, and the beta distribution for proportion data.
 There are 20 different spatial covariance structures available including the exponential, Gaussian, and spherical for point-referenced data (Figure$~$\ref{fig-type}) and the conditional autoregressive, and simultaneous autoregressive structures for areal data.
-\pkg{spmodel} provides tools for commonly used model summaries, visualizations, and diagnostics (e.g., Cook's distance) using standard \proglang{R} helper functions like \code{summary()}, \code{plot()}, \code{fitted()}, and \code{tidy()}, among othres.
+\pkg{spmodel} provides tools for commonly used model summaries, visualizations, and diagnostics (e.g., Cook's distance) using standard \proglang{R} helper functions like \code{summary()}, \code{plot()}, \code{fitted()}, and \code{tidy()}, among others.
 \pkg{spmodel} also provides tools to predict $\mathbf{w}$ at new locations and quantify uncertainty in those prediction using \code{predict()} and \code{augment()}.
@@
 
-\pkg{spmodel} (version 0.11.0) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in terms of scope and feel.
+Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version 0.11.0) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in terms of scope and feel.
 Both packages use similar syntax as \code{glm()}, accommodate flexible \code{formula} arguments (e.g., offsets, splines), handle spatial covariance that decays at different rates in different directions (i.e., geometric anisotropy), incorporate nonspatial random effects, support other \proglang{R} packages for modeling like \pkg{broom} [@robinson2021broom; @kuhn2022tidy], \pkg{emmeans} [@lenth2024emmeans], and \pkg{car} [@fox2019car], and have tools for model summaries, prediction, and simulating data.
@@
 \pkg{spmodel} also supports 20 different spatial covariances and models them directly, rather than using a precision matrix approximation to the Matérn spatial covariance as in \pkg{sdmTMB}.
-\pkg{spmodel} can also model areal data directly using neighborhood distance and autoregressive models rather than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not be within the polygon's boundaries.
-\pkg{spmodel} also provides experimental design tools (e.g., analysis of variance, contrasts), supports \pkg{sf} objects in modeling and prediction functions [@pebesma2018sf], has several specialized model diagnostics like leverage values and Cook's distances, and has analytic solutions for fixed effect and prediction standard errors.
+\pkg{spmodel} can model data directly using neighborhood distance and autoregressive models, rather than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not be within the polygon's boundaries.
+\pkg{spmodel} provides experimental design tools (e.g., analysis of variance, contrasts), supports \pkg{sf} objects in modeling and prediction functions [@pebesma2018sf], has several specialized model diagnostics like leverage values and Cook's distances, and has analytic solutions for fixed effect and prediction standard errors.
 Other similarities and differences do exist between \pkg{sdmTMB} and \pkg{spmodel}, and both packages continue to evolve.
@@
 Next, we describe a brief overview of the approach and how it can be used for several primary data analysis tasks [@tredennick2021practical] like model comparison, parameter estimation, inference, model diagnostics, and prediction.
+Then in Section$~$\ref{sec-applications} and Section$~$\ref{sec-applications2}, we show how to use \pkg{spmodel} to carry out these primary data analysis tasks for various data sets.
 
@@
 \begin{align}\label{eq-marginal}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w},
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w},
 \end{align}
-where $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ is the density for the appropriate response distribution of $\mathbf{y}$ (e.g., binomial, Poisson) given the latent $\mathbf{w}$ and dispersion parameter ($\varphi$), and $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ is the multivariate Gaussian density for $\mathbf{w}$ given the explanatory variables ($\mathbf{X}$), fixed effects ($\boldsymbol{\beta}$), and spatial covariance parameters ($\boldsymbol{\theta}$).
-The elements of $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ are conditionally independent (given $\mathbf{w}$), but the elements of $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ share spatial covariance.
+where $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ is the density for the appropriate response distribution of $\mathbf{y}$ (e.g., binomial, Poisson) given the latent $\mathbf{w}$ and dispersion parameter ($\varphi$), and $[\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}]$ is the multivariate Gaussian density for $\mathbf{w}$ given the explanatory variables ($\mathbf{X}$), fixed effects ($\boldsymbol{\beta}$), and spatial covariance parameters ($\boldsymbol{\theta}$).
+The elements of $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ are conditionally independent (given $\mathbf{w}$), but the elements of $[\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}]$ share spatial covariance.
 Following @harville1977maximum, we can integrate $\boldsymbol{\beta}$ out of Equation$~$\ref{eq-marginal}, which yields
@@
 Maximizing the natural logarithm of Equation$~$\ref{eq-marginal6} requires a doubly iterative process over $\boldsymbol{\theta}$ and $\varphi$ as well as $\mathbf{w}$, eventually yielding the the marginal restricted maximum likelihood estimators $\hat{\varphi}$ and $\hat{\boldsymbol{\theta}}$ and their corresponding values of $\hat{\mathbf{w}}$.
-Maximizing this log likelihood is a computationally expensive operation that involves repeatedly evaluating $\boldsymbol{\Sigma}^{-1}$, $\mathbf{g}$, and $\mathbf{G}$; see @ver2024marginal for more details and forms of $\mathbf{g}$ and $\mathbf{G}$ for various response distributions.
+Maximizing this log likelihood requires repeatedly evaluating $\boldsymbol{\Sigma}^{-1}$, $\mathbf{g}$, and $\mathbf{G}$; see @ver2024marginal for more details and comparisons to other approaches as well as and forms of $\mathbf{g}$ and $\mathbf{G}$ for various response distributions.
 
@@
 
-## Inspecting model diagnostics
+## Inspecting model diagnostics {#sec:diagnostics}
 
@@
 Again suppose that we observed $\mathbf{w}$ and we want to make predictions at $\mathbf{u}$, a vector of latent means at the new locations that follows the same SPGLM from Equation$~$\ref{eq-spglm} and having fixed effects design matrix, $\mathbf{X}_{\mathbf{u}}$.
-The vector $(\mathbf{w}, \mathbf{u})^\top$ has expectation $(\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top$ and covariance matrix $\begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}$, where $\boldsymbol{\Sigma} = \text{Var}(\mathbf{w}, \mathbf{w})$, $\boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}} = \text{Var}(\mathbf{w}, \mathbf{u})$, $\boldsymbol{\Sigma}_{\mathbf{u} \mathbf{w}} = \boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}}^\top$ and $\boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} = \text{Var}(\mathbf{u}, \mathbf{u})$
+The vector $(\mathbf{w}, \mathbf{u})^\top$ has expectation $(\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top$ and covariance matrix $\begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}$, where $\boldsymbol{\Sigma} = \text{Var}(\mathbf{w}, \mathbf{w})$, $\boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}} = \text{Var}(\mathbf{w}, \mathbf{u})$, $\boldsymbol{\Sigma}_{\mathbf{u} \mathbf{w}} = \boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}}^\top$ and $\boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} = \text{Var}(\mathbf{u}, \mathbf{u})$.
 By assumption, we have observed $\mathbf{w}$, so we may derive the conditional distribution of $\mathbf{u}|\mathbf{w}$, which has the following properties:
@@
   \text{E}(\mathbf{u} | \mathbf{w}) & = \mathbf{X}_{\mathbf{u}} \boldsymbol{\beta} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{w} - \mathbf{X}\boldsymbol{\beta}) \\
-  \text{E}(\mathbf{u} | \mathbf{w}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
+  \text{Var}(\mathbf{u} | \mathbf{w}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
 \end{align}
@@
 SPGLMs in \pkg{spmodel} are fit using the \code{spglm()} function.
-The \code{spglm()} function requires four arguments: \code{formula}, the relationship between the response and explanatory variables; \code{family}, the response distribution assumed for the repsonse variable; \code{data}, the data frame that contains the variables in \code{formula}, and \code{spcov_type}, the type of spatial covariance.
+The \code{spglm()} function requires four arguments: \code{formula}, the relationship between the response and explanatory variables; \code{family}, the response distribution assumed for the response variable; \code{data}, the data frame that contains the variables in \code{formula}, and \code{spcov_type}, the type of spatial covariance.
 The \code{formula}, \code{family}, and \code{data} arguments are the three required arguments to \code{glm()} for nonspatial GLMs.
@@
 simply requires one additional argument: \code{spcov_type}.
-When \code{data} is not an \code{sf} object, \code{spglm()} also requires the \code{xcoord} and \code{ycoord} arguments, which indicate the columns in \code{data} that represent projected the x- and y-coordinates, respectively.
+When \code{data} is not an \code{sf} object, \code{spglm()} also requires the \code{xcoord} and \code{ycoord} arguments, which indicate the columns in \code{data} that represent the projected x- and y-coordinates, respectively.
 
@@
 
-The \code{summary()} function returns a model summary that returns relevant information like the function call, deviance residuals, a coefficients table of fixed effects, the pseudo R-squared, spatial covariance parameter coefficient estimates, and the GLM dispersion parameter (fixed at one in logistic regression):
+The \code{summary()} function returns a model summary that returns relevant information like the function call, deviance residuals, a coefficients table of fixed effects, the pseudo R-squared, spatial covariance parameters, and the GLM dispersion parameter (fixed at one in logistic regression):
 ```{r}
@@
 
-The strength of spatial covarinace in the data affects how beneficial a SPGLM is relative to a GLM.
+The strength of spatial covariance in the data affects how beneficial an SPGLM is relative to a GLM.
 When the spatial covariance is strong, the SPGLM should notably outperform the GLM.
@@
 
-While the \code{spglm()} approach evaluates the HGLMM likelhood with $\sigma^2_{de} = 0$ and $\sigma^2_{ie} \approx 0$ instead of just the GLM likelihood, the parameter estimates and their standard errors are the same:
+While the \code{spglm()} approach evaluates the HGLMM likelihood with $\sigma^2_{de} = 0$ and $\sigma^2_{ie} \approx 0$ instead of just the GLM likelihood, the parameter estimates and their standard errors are the same:
 
@@
 An alternative approach to model comparison is to use a cross-validation procedure [@james2013introduction].
-The \code{loocv()} function performs leave-one-out cross validation, comparing the predicted mean (on the response scale) to the observed response variable for each hold-out observation, recomputing estimates of $\boldsymbol{\beta}$ each iteration Then, statistics like bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) can be used to evaluate models:
+The \code{loocv()} function performs leave-one-out cross validation, comparing the predicted mean (on the response scale) to the observed response variable for each hold-out observation, recomputing estimates of $\boldsymbol{\beta}$ in each iteration.
+Performing leave-one-out cross validation tends to be more computationally efficient than fitting the model, as leave-one-out cross validation requires only one set of products involving the inverse covariance matrix (the primary computational burden), while fitting traditional models requires these products for each optimization iteration.
+After performing leave-one-out cross validation, statistics like bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) can be used to evaluate models:
 
@@
 
-We can compare two SPGLMs with different spatial covariance functions using likelihood-based statistics and leave-one-out cross validation, but we can't use the LRT because generally, the spatial covariance functions aren't nested:
+We can compare two SPGLMs with different spatial covariance functions using likelihood-based statistics and leave-one-out cross validation, but we can't use the LRT because generally, the spatial covariance functions are not nested:
 ```{r}
@@
 
-Often times in spatial statistics, the difference in model fit between the best spatial model and worst spatial model is much smaller than the difference in model fit between the worst spatial model and the nonspatial model, implying that accounting for some form of spatial covariance is very beneficial.
+Frequently in spatial statistics, the difference in model fit between the best spatial model and worst spatial model is much smaller than the difference in model fit between the worst spatial model and the nonspatial model, implying that accounting for some form of spatial covariance is very beneficial.
 Two spatial covariance functions to consider starting with are the exponential and Gaussian, which have quite different origin behaviors (Figure$~$\ref{fig-type}), something @stein1999interpolation argues is important to characterize accurately.
@@
 
-\code{spmodel} provides a suite of tools for model diagnostics. One is \code{augment()}, which augments the model data with diagnostics:
+\code{spmodel} provides a suite of tools for model diagnostics.
+One is \code{augment()}, which augments the data used in the model with several model diagnostics (introduced in Section$~$\ref{sec:diagnostics}):
 ```{r}
@@
 
-The fitted values (\code{.fitted}) can be returned on either the link ($\hat{\mathbf{w}}$) or response ($f^{-1}(\hat{\mathbf{w}})$) scale and the residuals (\code{.resid}) can deviance, Pearson, or response residuals.
-The default fitted values are the link scale and the default residuals are deviance residuals.
-Also returned by \code{augment()} are the leverage (\code{.hat}), Cook's distance (\code{.cooksd}), and standardized residuals \code{.std.resid}.
+The fitted values (\code{.fitted}) can be returned on either the link ($\hat{\mathbf{w}}$) or response ($f^{-1}(\hat{\mathbf{w}})$) scale and the residuals (\code{.resid}) can be deviance, Pearson, or response residuals.
+The default fitted values are on the link scale and the default residuals are deviance residuals.
+Also returned by \code{augment()} are the leverage (\code{.hat}), Cook's distance (\code{.cooksd}), and standardized residuals \code{.std.resid} described in Section$~$\ref{sec:diagnostics}.
 A benefit of using \code{augment()} when \code{data} is an \code{sf} object is that the output is also an \code{sf} object, which makes it straightforward to create spatial diagnostic plots (Figure$~$\ref{fig-sp-diagnostic}).
@@
 
-The \code{plot()} function can also be used to return similar diagnostics as from \code{lm()} and \code{glm()} with additional tools for diagnosing spatial covariance.
+The \code{plot()} function can also be used to return similar diagnostics as from \code{lm()} and \code{glm()}, with additional tools for diagnosing spatial covariance.
 For example, we can inspect Cook's distance values and the empirical spatial covariance as a function of distance with (Figure$~$\ref{fig-sp-diagnostic2}):
@@
 \includegraphics[width = 1\linewidth]{figures/figure-3.png}
-\caption{Moose presence model diagnostics, including Cook's distance and the fitted spatial covariance as a function of distance (right).}
+\caption{Moose presence model diagnostics, including Cook's distance (left) and the fitted spatial covariance as a function of distance (right).}
 \label{fig-sp-diagnostic2}
@@
 The remaining variability ($1 - PR^2$) is allocated proportionally to \code{de} and \code{ie} according to $\sigma^2_{de}$ and $\sigma^2_{ie}$.
-This variability partitioning is a useful that helps quantify how much the explanatory variables, residual spatial variance, and residual nonspatial variance contribute to model fit, but as with $PR^2$, should not be used as a model comparison tool.
+This variability partitioning is a useful tool that helps quantify how much the explanatory variables, residual spatial variance, and residual nonspatial variance contribute to model fit, but as with $PR^2$, should not be used as a model comparison tool.
 
@@
 
-We can predict the probability of moose presence using \code{predict()}:
+We can predict the probability of moose presence at the locations in \code{moose_preds} using \code{predict()}:
 ```{r}
@@
 \includegraphics[width = 0.70\linewidth]{figures/figure-4.png}
-\caption{Moose presence probability fitted values and predictions. Fitted values are represeneted by circles and predictions by triangles.}
+\caption{Moose presence probability fitted values and predictions. Fitted values are represented by circles and predictions by triangles.}
 \label{fig-moose-fit}
@@
 
-Prediction intervals for the probability of moose presence (on the link scale) are returned by suppplying \code{interval}:
+Prediction intervals for the probability of moose presence (on the link scale) are returned by supplying \code{interval}:
 ```{r}
@@
 \includegraphics[width = 1\linewidth]{figures/figure-5.png}
-\caption{Moose presence probability prediction intervals. 95\% prediction interval lower bound (left) and 95\% prediction interval upper bound (right).}
+\caption{Moose presence 95\% prediction interval lower bounds (left) and upper bounds (right).}
 \label{fig-moose-int}
@@
 In Section$~$\ref{sec-moose-count}, we fit Poisson and negative binomial models with and without geometric anisotropy for the point-referenced moose count data.
-In Section$~$\ref{sec-lake}, we fit a Gamma model to the point-referenced lake conductivity data, showing how to perform a spatial analysis of variance (ANOVA), fit contrasts for models with interactions and use maximum likelihood to compare two models with different explanatory variables.
+In Section$~$\ref{sec-lake}, we fit a Gamma model to the point-referenced lake conductivity data, showing how to fit a model with a partition factor, perform a spatial analysis of variance (ANOVA), fit contrasts for models with interactions.
 In Section$~$\ref{sec-seal}, we fit a binomial model to the areal harbor seal trend data with a nonspatial random effect.
-Finally in Section$~$\ref{sec-texas}, we fit beta models to Texas voter turnout data, which can be treated as point-referenced or areal.
+Finally in Section$~$\ref{sec-texas}, we fit beta models to Texas voter turnout data, which can be treated as point-referenced or areal, and use maximum likelihood to compare two models with different explanatory variables.
+Table$~$\ref{tab:additional} outlines the section number, data set, family (i.e., response distribution), geometry type, and additional \pkg{spmodel} features for each application.
+
+\begin{table}
+    \centering
+    \begin{tabular}{|l|llll|}
+    \hline
+    Section & Data & Family & Geometry & Additional Features \\
+    \hline
+    \hline
+    4.1 & Moose Counts & Poisson & Point & Geometric Anisotropy  \\
+    &  & NBinomial &  &  \\
+    \hline
+    4.2 & Lake Conductivity & Gamma & Point & Partition Factor  \\
+    & & & & ANOVA \\
+    & & & & Contrasts \\
+    \hline
+    4.3 & Harbor Seals & Binomial & Areal & Nonspatial Random Effects  \\
+    \hline
+    4.4 & Texas Voter Turnout & Beta & Point & Likelihood-Ratio Test  \\
+    & & & Areal & \\
+    \hline 
+    \end{tabular}
+\caption{Section number, data set, family, geometry type, and additional features for each application.}
+\label{tab:additional}
+\end{table}
 
@@
 
-In addition to moose presence, moose counts are also recorded in \code{moose} (Figure$~$\ref{fig-moose-data-count}.
+In addition to moose presence, moose counts are also recorded in \code{moose} (Figure$~$\ref{fig-moose-data-count}).
 The Poisson and negative binomial response distributions can be used to model SPGLMs for count data.
@@
 A spatial covariance function is geometrically isotropic if it decays with distance at the same rate in all directions (Figure$~$\ref{fig-tropy}; left).
-A spatial covariance is geometrically isotropic if it decays with distance at different rates in different directions (Figure$~$\ref{fig-tropy}; right).
+A spatial covariance is geometrically anisotropic if it decays with distance at different rates in different directions (Figure$~$\ref{fig-tropy}; right).
 Geometric anisotropy is formally incorporated by rotating and scaling original coordinates, yielding transformed coordinates that are geometrically isotropic:
@@
 \end{align} 
-The parameters $\omega$ and $\alpha$ controls the scaling and rotation, respectively, of the major and minor axes of a level curve of equal correlation (Figure$~$\ref{fig-tropy}).
+The parameters $\omega$ and $\alpha$ controls the scaling and rotation, respectively, of the major and minor axes of a level curve of equal spatial covariance (Figure$~$\ref{fig-tropy}).
 Using these transformed coordinates, the partial sill ($\sigma^2_{de}$), nugget ($\sigma^2_{ie}$), and range ($\phi$) parameters are estimated.
@@
 
-The spatial correlation is strongest in a northwest-southeast direction and weakest in the northeast-southwest direction (Figure$~$\ref{fig-tropy}), which is intuitive given the similar patterns in moose counts from Figure$~$\ref{fig-moose-data-count}.
+The spatial covariance is strongest in a northwest-southeast direction and weakest in the northeast-southwest direction (Figure$~$\ref{fig-tropy}), which is intuitive given the similar patterns in moose counts from Figure$~$\ref{fig-moose-data-count}.
 
@@
 \includegraphics[width = 1\linewidth]{figures/figure-7.png}
-\caption{Level curves of equal correlation for the negative binomial moose count models.
-The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of correlation.
+\caption{Level curves of equal spatial covariance for the negative binomial moose count models.
+The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of spatial covariance.
 In the isotropic level curve (left), spatial covariance decays equally in all directions.
@@
 Conductivity is a measure of dissolved ions (measured here in water), which is important for various physical, chemical, and biological processes.
-Chemical data is often heavily right-skewed, so we model it using an SPGLM assuming a Gamma distribution for the response.
+Chemical data are often heavily right-skewed, so we model them using an SPGLM assuming a Gamma distribution for the response.
 The \code{log_cond} variable in \code{lake} is the logarithm of conductivity, which we dynamically exponentiate within \code{formula} so that it is on the original scale:
@@
 spgam <- spglm(
-  formula = exp(log_cond) ~ temp + state + temp:state + origin, 
+  formula = exp(log_cond) ~ temp * state + origin, 
   family = "Gamma",
@@
 
-A more effective approach is to use an analysis of variance (ANOVA), which is well-suited to asses the overall significance of each variable:
+A more effective approach is to use an analysis of variance (ANOVA), which is well-suited to assess the overall significance of each variable:
 ```{r}
@@
 The main effect for temperature and the temperature by state interaction are highly significant ($p$-value < 0.001), while the main effects for state and lake origin are not significant.
-The ANOVA table can be tidied using \code{tidy()}.
 
@@
 
-```{r, warning=FALSE, message=FALSE}
+```{r, eval=FALSE}
 library(car)
@@
 By default, \pkg{emmeans} uses the mean temperature value (here, 7.63) to assess contrasts:
-```{r, warning=FALSE, message=FALSE}
+```{r, eval=FALSE}
 library(emmeans)
@@
 While the \code{moose} and \code{lake} data were point-referenced, the \code{seal} data are areal.
-Each polygon in the \code{seal} data represents a distinct harbor seal haulout location (Figure$~$\ref{fig-seal}).
-A haulout location is a spot on coastal rocks that harbor seals go to rest, molt, and give birth.
+Each polygon in the \code{seal} data represents a distinct harbor seal haulout region (Figure$~$\ref{fig-seal}).
+A haulout region is an area of coastal rocks that harbor seals go to rest, molt, and give birth.
 
@@
 Conditional and simultaneous autoregressive functions characterize spatial distance through neighborhood relationships (rather than Euclidean distance) and have \code{spcov_type} values of \code{"car"} and \code{"sar"}, respectively.
-By default, Queen's distance is used to determine whether two sites are neighbors, though custom neighborhood matrices are passed via \code{W}.
+By default, Queen's distance is used to determine whether two sites are neighbors, though custom neighborhood matrices can are passed via \code{W}.
 Row standardization is also assumed by default; this can be changed via \code{row_st}.
@@
 
-The default estimation method in \pkg{spmodel} is restricted maximum likelihood (REML).
-One drawback of REML is that likelihood-based statistics are only valid for model comparison when the models have the same explanatory variable and fixed effect structure.
-This is because the error contrasts used to construct the REML likelihood change based on $\mathbf{X}$ and $\boldsymbol{\beta}$.
-An alternative is to use maximum likelihood, which can use likelihood-based statistics to compare models with different explanatory variable and fixed effect structures.
-While we could assess the significance of log turnout with the \code{spbeta_geo} model fit with REML (e.g., via \code{tidy()}), an alternative approach is to use a likelihood ratio test:
+The default estimation method in \pkg{spmodel} for SPGLMs is restricted maximum likelihood (REML), while maximum likelihood (ML) can also be used.
+A benefit of benefit of REML is that it can yield unbiased estimates of covariance parameters [@cressie1993asymptotic], but a drawback is that likelihood-based statistics are only valid for model comparison when the models have the same explanatory variable and fixed effect structure (because the error contrasts used to construct the REML likelihood change based on $\mathbf{X}$ and $\boldsymbol{\beta}$).
+In contrast, ML estimators are biased, though in practice this bias may be small.
+Moreover, when using ML, likelihood-based comparisons are valid for models having different explanatory variable and fixed effect structures.
+Using ML, we can evaluate the significance of log income on voter turnout with a likelihood ratio test:
 ```{r}
@@
 
-The likelihood ratio test suggests that log income is significantly related to voter turnout ($p$-value < 0.001). We could instead use a different likelihood-based statistic like AIC:
+The likelihood ratio test suggests that log income is significantly related to voter turnout ($p$-value < 0.001).
+Alternatively, we could have instead used a different likelihood-based statistic like AIC:
 ```{r}
@@
 
-The AIC prefers the full model, which suggests that log income is important for predicting voter turnout.
+The AIC also prefers the full model, suggesting that log income is important for predicting voter turnout.
 

### 2025-09-11 inst/manuscript/manuscript.tex
some updates

@@
 \usepackage{autonum}
+\usepackage{adjustbox}
 
@@
 In practice, non-Gaussian data (e.g., binary, count, skewed, and
-proportion data) are ubiquitous. Non-Gaussian that belong to an
-exponential family data can be naturally modeled using a generalized
-linear model (GLM) regression framework
+proportion data) are ubiquitous. Non-Gaussian data that belong to an
+exponential family can be naturally modeled using a generalized linear
+model (GLM) regression framework
 \citep{nelder1972generalized, mccullagh1989generalized}. In a GLM, an
@@
 additional layer of complexity compared to linear regression models, as
-the left-hand size of Equation\(~\)\ref{eq-glm} is a function of the
+the left-hand side of Equation\(~\)\ref{eq-glm} is a function of the
 mean of \(\mathbf{y}\) rather than \(\mathbf{y}\) itself (as in linear
@@
 distance-decay rate of the spatial correlation. One example of a spatial
-covariance matrix is the ``exponential'', which is given by
+covariance matrix is the ``exponential,'' which is given by
 \begin{align}\label{eq-spcov-exp}
-  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{de} \mathbf{R}_{exp} =  \sigma^2_{de} \exp(-\mathbf{H}/\phi),
+  \text{Cov}(\boldsymbol{\tau}) = \sigma^2_{\tau} \mathbf{R}_{exp} =  \sigma^2_{\tau} \exp(-\mathbf{H}/\phi),
 \end{align} where \(\mathbf{H}\) is a matrix of pairwise distances among
@@
 and \(\mathbf{X}\) may instead be written \(\mathbf{y}(\mathbf{s})\) and
-\(\mathbf{X}\), respectively. We acknowledge the utility of this
-nomenclature but drop the explicit dependence on \(\mathbf{s}\) for
+\(\mathbf{X}(\mathbf{s})\), respectively. We acknowledge the utility of
+this nomenclature but drop the explicit dependence on \(\mathbf{s}\) for
 simplicity of notation.
@@
 Working with precision matrices, however, can be more restrictive and
-less intuitive than working directly with the covariance matrix. The
-\pkg{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
+less intuitive than working directly with the covariance matrix.
+
+The \pkg{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
 \citep{brooks2017glmmtmb}, \pkg{hglm} \citep{ronnegard2010hglm},
@@
 faster. The novel Laplace approach is particularly attractive for two
-reasons. First, it is general enough that can be applied to any
+reasons. First, it is general enough that it can be applied to any
 covariance structure (not just spatial). Second, after estimating the
@@
 subsets of a region; models are sometimes called ``autoregressive''
-models). \pkg{spmodel} supports the binomial distribution for binary
-data, Poisson and negative binomial distributions for count data, Gamma
-and inverse Gaussian distributions for skewed data, and the beta
-distribution for proportion data. There are 20 different spatial
-covariance structures available including the exponential, Gaussian, and
-spherical for point-referenced data (Figure\(~\)\ref{fig-type}) and the
-conditional autoregressive, and simultaneous autoregressive structures
-for areal data. \pkg{spmodel} provides tools for commonly used model
-summaries, visualizations, and diagnostics (e.g., Cook's distance) using
-standard \proglang{R} helper functions like \code{summary()},
-\code{plot()}, \code{fitted()}, and \code{tidy()}, among othres.
-\pkg{spmodel} also provides tools to predict \(\mathbf{w}\) at new
-locations and quantify uncertainty in those prediction using
-\code{predict()} and \code{augment()}. This core functionality, combined
-with several advanced features we describe throughout the manuscript,
-enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools
-previously missing from the existing \proglang{R} ecosystem.
+models). For both point-referenced data and areal data, \pkg{spmodel}
+supports the binomial distribution for binary data, Poisson and negative
+binomial distributions for count data, Gamma and inverse Gaussian
+distributions for skewed data, and the beta distribution for proportion
+data. There are 20 different spatial covariance structures available
+including the exponential, Gaussian, and spherical for point-referenced
+data (Figure\(~\)\ref{fig-type}) and the conditional autoregressive, and
+simultaneous autoregressive structures for areal data. \pkg{spmodel}
+provides tools for commonly used model summaries, visualizations, and
+diagnostics (e.g., Cook's distance) using standard \proglang{R} helper
+functions like \code{summary()}, \code{plot()}, \code{fitted()}, and
+\code{tidy()}, among others. \pkg{spmodel} also provides tools to
+predict \(\mathbf{w}\) at new locations and quantify uncertainty in
+those prediction using \code{predict()} and \code{augment()}. This core
+functionality, combined with several advanced features we describe
+throughout the manuscript, enables \pkg{spmodel} to introduce novel,
+important SPGLM modeling tools previously missing from the existing
+\proglang{R} ecosystem.
 
@@
 
-\pkg{spmodel} (version 0.11.0) is arguably most similar to \pkg{sdmTMB}
-(version 0.7.4) in terms of scope and feel. Both packages use similar
-syntax as \code{glm()}, accommodate flexible \code{formula} arguments
-(e.g., offsets, splines), handle spatial covariance that decays at
-different rates in different directions (i.e., geometric anisotropy),
-incorporate nonspatial random effects, support other \proglang{R}
-packages for modeling like \pkg{broom}
-\citep{robinson2021broom, kuhn2022tidy}, \pkg{emmeans}
-\citep{lenth2024emmeans}, and \pkg{car} \citep{fox2019car}, and have
-tools for model summaries, prediction, and simulating data. There are
-some notable differences between the two packages, however. \pkg{sdmTMB}
-supports several additional GLM distributions like the Tweedie, supports
-Hurdle models, and can incorporate prior information through Bayesian
-applications. \pkg{sdmTMB} also provides tools for working with temporal
-data and spatiotemporal data and provides enhanced visualizations of the
-model's marginal effects. \pkg{sdmTMB} does require a preprocessing step
-of constructing a mesh for the stochastic partial differential equation
+Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version
+0.11.0) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in
+terms of scope and feel. Both packages use similar syntax as
+\code{glm()}, accommodate flexible \code{formula} arguments (e.g.,
+offsets, splines), handle spatial covariance that decays at different
+rates in different directions (i.e., geometric anisotropy), incorporate
+nonspatial random effects, support other \proglang{R} packages for
+modeling like \pkg{broom} \citep{robinson2021broom, kuhn2022tidy},
+\pkg{emmeans} \citep{lenth2024emmeans}, and \pkg{car}
+\citep{fox2019car}, and have tools for model summaries, prediction, and
+simulating data. There are some notable differences between the two
+packages, however. \pkg{sdmTMB} supports several additional GLM
+distributions like the Tweedie, supports Hurdle models, and can
+incorporate prior information through Bayesian applications.
+\pkg{sdmTMB} also provides tools for working with temporal data and
+spatiotemporal data and provides enhanced visualizations of the model's
+marginal effects. \pkg{sdmTMB} does require a preprocessing step of
+constructing a mesh for the stochastic partial differential equation
 approach, and the density of the mesh can affect model results and
@@
 rather than using a precision matrix approximation to the Matérn spatial
-covariance as in \pkg{sdmTMB}. \pkg{spmodel} can also model areal data
-directly using neighborhood distance and autoregressive models rather
-than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not
-be within the polygon's boundaries. \pkg{spmodel} also provides
-experimental design tools (e.g., analysis of variance, contrasts),
-supports \pkg{sf} objects in modeling and prediction functions
-\citep{pebesma2018sf}, has several specialized model diagnostics like
-leverage values and Cook's distances, and has analytic solutions for
-fixed effect and prediction standard errors. Other similarities and
-differences do exist between \pkg{sdmTMB} and \pkg{spmodel}, and both
-packages continue to evolve. Overall, we believe that these packages are
-complementary and enhance the suite of SPGLM tools accessible to
-practitioners.
+covariance as in \pkg{sdmTMB}. \pkg{spmodel} can model data directly
+using neighborhood distance and autoregressive models, rather than
+relying on the polygon centroid (as in \pkg{sdmTMB}), which may not be
+within the polygon's boundaries. \pkg{spmodel} provides experimental
+design tools (e.g., analysis of variance, contrasts), supports \pkg{sf}
+objects in modeling and prediction functions \citep{pebesma2018sf}, has
+several specialized model diagnostics like leverage values and Cook's
+distances, and has analytic solutions for fixed effect and prediction
+standard errors. Other similarities and differences do exist between
+\pkg{sdmTMB} and \pkg{spmodel}, and both packages continue to evolve.
+Overall, we believe that these packages are complementary and enhance
+the suite of SPGLM tools accessible to practitioners.
 
@@
 tasks \citep{tredennick2021practical} like model comparison, parameter
-estimation, inference, model diagnostics, and prediction.
+estimation, inference, model diagnostics, and prediction. Then in
+Section\(~\)\ref{sec-applications} and
+Section\(~\)\ref{sec-applications2}, we show how to use \pkg{spmodel} to
+carry out these primary data analysis tasks for various data sets.
 
@@
 \begin{align}\label{eq-marginal}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w},
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} \int_{\boldsymbol{\beta}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}] d\boldsymbol{\beta} d\mathbf{w},
 \end{align} where \([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) is the
@@
 parameter (\(\varphi\)), and
-\([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\) is the multivariate
-Gaussian density for \(\mathbf{w}\) given the explanatory variables
-(\(\mathbf{X}\)), fixed effects (\(\boldsymbol{\beta}\)), and spatial
-covariance parameters (\(\boldsymbol{\theta}\)). The elements of
+\([\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}]\)
+is the multivariate Gaussian density for \(\mathbf{w}\) given the
+explanatory variables (\(\mathbf{X}\)), fixed effects
+(\(\boldsymbol{\beta}\)), and spatial covariance parameters
+(\(\boldsymbol{\theta}\)). The elements of
 \([\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]\) are conditionally
 independent (given \(\mathbf{w}\)), but the elements of
-\([\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]\) share spatial
-covariance. Following \citet{harville1977maximum}, we can integrate
-\(\boldsymbol{\beta}\) out of Equation\(~\)\ref{eq-marginal}, which
-yields \begin{align}\label{eq-marginal2}
+\([\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}]\)
+share spatial covariance. Following \citet{harville1977maximum}, we can
+integrate \(\boldsymbol{\beta}\) out of Equation\(~\)\ref{eq-marginal},
+which yields \begin{align}\label{eq-marginal2}
   [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] = \int_{\mathbf{w}} [\mathbf{y} | f^{-1}(\mathbf{w}), \varphi] [\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] d\mathbf{w},
@@
 \(\hat{\boldsymbol{\theta}}\) and their corresponding values of
-\(\hat{\mathbf{w}}\). Maximizing this log likelihood is a
-computationally expensive operation that involves repeatedly evaluating
-\(\boldsymbol{\Sigma}^{-1}\), \(\mathbf{g}\), and \(\mathbf{G}\); see
-\citet{ver2024marginal} for more details and forms of \(\mathbf{g}\) and
-\(\mathbf{G}\) for various response distributions.
+\(\hat{\mathbf{w}}\). Maximizing this log likelihood requires repeatedly
+evaluating \(\boldsymbol{\Sigma}^{-1}\), \(\mathbf{g}\), and
+\(\mathbf{G}\); see \citet{ver2024marginal} for more details and
+comparisons to other approaches as well as and forms of \(\mathbf{g}\)
+and \(\mathbf{G}\) for various response distributions.
 
@@
 
-\subsection{Inspecting model
-diagnostics}\label{inspecting-model-diagnostics}
+\subsection{Inspecting model diagnostics}\label{sec:diagnostics}
 
@@
 and
-\(\boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} = \text{Var}(\mathbf{u}, \mathbf{u})\)
+\(\boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} = \text{Var}(\mathbf{u}, \mathbf{u})\).
 By assumption, we have observed \(\mathbf{w}\), so we may derive the
@@
   \text{E}(\mathbf{u} | \mathbf{w}) & = \mathbf{X}_{\mathbf{u}} \boldsymbol{\beta} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{w} - \mathbf{X}\boldsymbol{\beta}) \\
-  \text{E}(\mathbf{u} | \mathbf{w}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
+  \text{Var}(\mathbf{u} | \mathbf{w}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
 \end{align} \citet{ver2024marginal} show how these equations are
@@
 relationship between the response and explanatory variables;
-\code{family}, the response distribution assumed for the repsonse
+\code{family}, the response distribution assumed for the response
 variable; \code{data}, the data frame that contains the variables in
@@
 \code{ycoord} arguments, which indicate the columns in \code{data} that
-represent projected the x- and y-coordinates, respectively.
+represent the projected x- and y-coordinates, respectively.
 
@@
 coefficients table of fixed effects, the pseudo R-squared, spatial
-covariance parameter coefficient estimates, and the GLM dispersion
-parameter (fixed at one in logistic regression):
+covariance parameters, and the GLM dispersion parameter (fixed at one in
+logistic regression):
 
@@
 
-The strength of spatial covarinace in the data affects how beneficial a
+The strength of spatial covariance in the data affects how beneficial an
 SPGLM is relative to a GLM. When the spatial covariance is strong, the
@@
 
-While the \code{spglm()} approach evaluates the HGLMM likelhood with
+While the \code{spglm()} approach evaluates the HGLMM likelihood with
 \(\sigma^2_{de} = 0\) and \(\sigma^2_{ie} \approx 0\) instead of just
@@
 (on the response scale) to the observed response variable for each
-hold-out observation, recomputing estimates of \(\boldsymbol{\beta}\)
-each iteration Then, statistics like bias, mean-squared-prediction error
-(MSPE), and the square root of MSPE (RMSPE) can be used to evaluate
-models:
+hold-out observation, recomputing estimates of \(\boldsymbol{\beta}\) in
+each iteration. Performing leave-one-out cross validation tends to be
+more computationally efficient than fitting the model, as leave-one-out
+cross validation requires only one set of products involving the inverse
+covariance matrix (the primary computational burden), while fitting
+traditional models requires these products for each optimization
+iteration. After performing leave-one-out cross validation, statistics
+like bias, mean-squared-prediction error (MSPE), and the square root of
+MSPE (RMSPE) can be used to evaluate models:
 
@@
 but we can't use the LRT because generally, the spatial covariance
-functions aren't nested:
+functions are not nested:
 
@@
 
-Often times in spatial statistics, the difference in model fit between
+Frequently in spatial statistics, the difference in model fit between
 the best spatial model and worst spatial model is much smaller than the
@@
 \code{spmodel} provides a suite of tools for model diagnostics. One is
-\code{augment()}, which augments the model data with diagnostics:
+\code{augment()}, which augments the data used in the model with several
+model diagnostics (introduced in Section\(~\)\ref{sec:diagnostics}):
 
@@
 (\(\hat{\mathbf{w}}\)) or response (\(f^{-1}(\hat{\mathbf{w}})\)) scale
-and the residuals (\code{.resid}) can deviance, Pearson, or response
-residuals. The default fitted values are the link scale and the default
-residuals are deviance residuals. Also returned by \code{augment()} are
-the leverage (\code{.hat}), Cook's distance (\code{.cooksd}), and
-standardized residuals \code{.std.resid}. A benefit of using
+and the residuals (\code{.resid}) can be deviance, Pearson, or response
+residuals. The default fitted values are on the link scale and the
+default residuals are deviance residuals. Also returned by
+\code{augment()} are the leverage (\code{.hat}), Cook's distance
+(\code{.cooksd}), and standardized residuals \code{.std.resid} described
+in Section\(~\)\ref{sec:diagnostics}. A benefit of using
 \code{augment()} when \code{data} is an \code{sf} object is that the
@@
 The \code{plot()} function can also be used to return similar
-diagnostics as from \code{lm()} and \code{glm()} with additional tools
+diagnostics as from \code{lm()} and \code{glm()}, with additional tools
 for diagnosing spatial covariance. For example, we can inspect Cook's
@@
 \code{de} and \code{ie} according to \(\sigma^2_{de}\) and
-\(\sigma^2_{ie}\). This variability partitioning is a useful that helps
-quantify how much the explanatory variables, residual spatial variance,
-and residual nonspatial variance contribute to model fit, but as with
-\(PR^2\), should not be used as a model comparison tool.
+\(\sigma^2_{ie}\). This variability partitioning is a useful tool that
+helps quantify how much the explanatory variables, residual spatial
+variance, and residual nonspatial variance contribute to model fit, but
+as with \(PR^2\), should not be used as a model comparison tool.
 
@@
 
-We can predict the probability of moose presence using \code{predict()}:
+We can predict the probability of moose presence at the locations in
+\code{moose_preds} using \code{predict()}:
 
@@
 Prediction intervals for the probability of moose presence (on the link
-scale) are returned by suppplying \code{interval}:
+scale) are returned by supplying \code{interval}:
 
@@
 moose count data. In Section\(~\)\ref{sec-lake}, we fit a Gamma model to
-the point-referenced lake conductivity data, showing how to perform a
-spatial analysis of variance (ANOVA), fit contrasts for models with
-interactions and use maximum likelihood to compare two models with
-different explanatory variables. In Section\(~\)\ref{sec-seal}, we fit a
-binomial model to the areal harbor seal trend data with a nonspatial
-random effect. Finally in Section\(~\)\ref{sec-texas}, we fit beta
-models to Texas voter turnout data, which can be treated as
-point-referenced or areal.
+the point-referenced lake conductivity data, showing how to fit a model
+with a partition factor, perform a spatial analysis of variance (ANOVA),
+fit contrasts for models with interactions. In
+Section\(~\)\ref{sec-seal}, we fit a binomial model to the areal harbor
+seal trend data with a nonspatial random effect. Finally in
+Section\(~\)\ref{sec-texas}, we fit beta models to Texas voter turnout
+data, which can be treated as point-referenced or areal, and use maximum
+likelihood to compare two models with different explanatory variables.
+Table\(~\)\ref{tab:additional} outlines the section number, data set,
+family (i.e., response distribution), geometry type, and additional
+\pkg{spmodel} features for each application.
+
+\begin{table}
+    \centering
+    \begin{tabular}{|l|llll|}
+    \hline
+    Section & Data & Family & Geometry & Additional Features \\
+    \hline
+    \hline
+    4.1 & Moose Counts & Poisson & Point & Geometric Anisotropy  \\
+    &  & NBinomial &  &  \\
+    \hline
+    4.2 & Lake Conductivity & Gamma & Point & Partition Factor  \\
+    & & & & ANOVA \\
+    & & & & Contrasts \\
+    \hline
+    4.3 & Harbor Seals & Binomial & Areal & Nonspatial Random Effects  \\
+    \hline
+    4.4 & Texas Voter Turnout & Beta & Point & Likelihood-Ratio Test  \\
+    & & & Areal & \\
+    \hline 
+    \end{tabular}
+\caption{Section number, data set, family, geometry type, and additional features for each application.}
+\label{tab:additional}
+\end{table}
 
@@
 In addition to moose presence, moose counts are also recorded in
-\code{moose} (Figure\(~\)\ref{fig-moose-data-count}. The Poisson and
+\code{moose} (Figure\(~\)\ref{fig-moose-data-count}). The Poisson and
 negative binomial response distributions can be used to model SPGLMs for
@@
 all directions (Figure\(~\)\ref{fig-tropy}; left). A spatial covariance
-is geometrically isotropic if it decays with distance at different rates
-in different directions (Figure\(~\)\ref{fig-tropy}; right). Geometric
-anisotropy is formally incorporated by rotating and scaling original
-coordinates, yielding transformed coordinates that are geometrically
-isotropic: \begin{align}
+is geometrically anisotropic if it decays with distance at different
+rates in different directions (Figure\(~\)\ref{fig-tropy}; right).
+Geometric anisotropy is formally incorporated by rotating and scaling
+original coordinates, yielding transformed coordinates that are
+geometrically isotropic: \begin{align}
   \begin{bmatrix}
@@
 scaling and rotation, respectively, of the major and minor axes of a
-level curve of equal correlation (Figure\(~\)\ref{fig-tropy}). Using
-these transformed coordinates, the partial sill (\(\sigma^2_{de}\)),
-nugget (\(\sigma^2_{ie}\)), and range (\(\phi\)) parameters are
-estimated. We accommodate geometric anisotropy by supplying
-\code{anisotropy}:
+level curve of equal spatial covariance (Figure\(~\)\ref{fig-tropy}).
+Using these transformed coordinates, the partial sill
+(\(\sigma^2_{de}\)), nugget (\(\sigma^2_{ie}\)), and range (\(\phi\))
+parameters are estimated. We accommodate geometric anisotropy by
+supplying \code{anisotropy}:
 
@@
 
-The spatial correlation is strongest in a northwest-southeast direction
+The spatial covariance is strongest in a northwest-southeast direction
 and weakest in the northeast-southwest direction
@@
 which is important for various physical, chemical, and biological
-processes. Chemical data is often heavily right-skewed, so we model it
-using an SPGLM assuming a Gamma distribution for the response. The
+processes. Chemical data are often heavily right-skewed, so we model
+them using an SPGLM assuming a Gamma distribution for the response. The
 \code{log_cond} variable in \code{lake} is the logarithm of
@@
 R> spgam <- spglm(
-+   formula = exp(log_cond) ~ temp + state + temp:state + origin, 
++   formula = exp(log_cond) ~ temp * state + origin, 
 +   family = "Gamma",
@@
 Call:
-spglm(formula = exp(log_cond) ~ temp + state + temp:state + origin, 
-    family = "Gamma", data = lake, spcov_type = "cauchy", partition_factor = ~year)
+spglm(formula = exp(log_cond) ~ temp * state + origin, family = "Gamma", 
+    data = lake, spcov_type = "cauchy", partition_factor = ~year)
 
@@
 A more effective approach is to use an analysis of variance (ANOVA),
-which is well-suited to asses the overall significance of each variable:
+which is well-suited to assess the overall significance of each
+variable:
 
@@
 are highly significant (\(p\)-value \textless{} 0.001), while the main
-effects for state and lake origin are not significant. The ANOVA table
-can be tidied using \code{tidy()}.
+effects for state and lake origin are not significant.
 
@@
 point-referenced, the \code{seal} data are areal. Each polygon in the
-\code{seal} data represents a distinct harbor seal haulout location
-(Figure\(~\)\ref{fig-seal}). A haulout location is a spot on coastal
+\code{seal} data represents a distinct harbor seal haulout region
+(Figure\(~\)\ref{fig-seal}). A haulout region is an area of coastal
 rocks that harbor seals go to rest, molt, and give birth.
@@
 distance is used to determine whether two sites are neighbors, though
-custom neighborhood matrices are passed via \code{W}. Row
+custom neighborhood matrices can are passed via \code{W}. Row
 standardization is also assumed by default; this can be changed via
@@
 
-The default estimation method in \pkg{spmodel} is restricted maximum
-likelihood (REML). One drawback of REML is that likelihood-based
-statistics are only valid for model comparison when the models have the
-same explanatory variable and fixed effect structure. This is because
-the error contrasts used to construct the REML likelihood change based
-on \(\mathbf{X}\) and \(\boldsymbol{\beta}\). An alternative is to use
-maximum likelihood, which can use likelihood-based statistics to compare
-models with different explanatory variable and fixed effect structures.
-While we could assess the significance of log turnout with the
-\code{spbeta_geo} model fit with REML (e.g., via \code{tidy()}), an
-alternative approach is to use a likelihood ratio test:
+The default estimation method in \pkg{spmodel} for SPGLMs is restricted
+maximum likelihood (REML), while maximum likelihood (ML) can also be
+used. A benefit of benefit of REML is that it can yield unbiased
+estimates of covariance parameters \citep{cressie1993asymptotic}, but a
+drawback is that likelihood-based statistics are only valid for model
+comparison when the models have the same explanatory variable and fixed
+effect structure (because the error contrasts used to construct the REML
+likelihood change based on \(\mathbf{X}\) and \(\boldsymbol{\beta}\)).
+In contrast, ML estimators are biased, though in practice this bias may
+be small. Moreover, when using ML, likelihood-based comparisons are
+valid for models having different explanatory variable and fixed effect
+structures. Using ML, we can evaluate the significance of log income on
+voter turnout with a likelihood ratio test:
 
@@
 The likelihood ratio test suggests that log income is significantly
-related to voter turnout (\(p\)-value \textless{} 0.001). We could
-instead use a different likelihood-based statistic like AIC:
+related to voter turnout (\(p\)-value \textless{} 0.001). Alternatively,
+we could have instead used a different likelihood-based statistic like
+AIC:
 
@@
 
-The AIC prefers the full model, which suggests that log income is
+The AIC also prefers the full model, suggesting that log income is
 important for predicting voter turnout.

### 2025-09-11 inst/manuscript/code/code.R

@@ -3,86 +3,229 @@
 #=========================================================#
 
 library(spmodel)
-library(ggplot2)
-library(dplyr)
-library(emmeans)
+library(ggplot2, verbose = FALSE)
+library(dplyr, verbose = FALSE)
+library(emmeans, verbose = FALSE)
+library(car)
 
 #=========================================================#
-# Section 3.1 Binary Data ####
+# Section 3: Modeling moose presence in Alaska, USA ####
 #=========================================================#
 
 head(moose)
 
+head(moose_preds)
+
+#==========================================#
+# Section 3.1: Model Fitting ####
+#==========================================#
+
 spbin <- spglm(
-  formula = presence ~ elev + strat + elev:strat,
+  formula = presence ~ elev + strat,
   family = binomial,
   data = moose,
-  spcov_type = "spherical"
+  spcov_type = "exponential"
 )
 
 summary(spbin)
 
-tidy(spbin)
+tidy(spbin, conf.int = TRUE)
 
-moose$strat2 <- factor(moose$strat, levels = c("M", "L"))
-update(spbin, formula = presence ~ elev + strat2 + elev:strat2) |>
-  summary()
+#==========================================#
+# Section 3.2: Model Comparison ####
+#==========================================#
 
-emtrends(spbin, "strat", "elev")
-
-glance(spbin)
-
-head(augment(spbin))
-
-varcomp(spbin)
-
-head(predict(spbin, newdata = moose_preds))
-
-head(augment(
-  spbin,
-  newdata = moose_preds,
-  type.predict = "response",
-  interval = "prediction"
-))
-
-# fit nonspatial model using spglm()
 bin <- spglm(
-  formula = presence ~ elev + strat + elev:strat,
+  formula = presence ~ elev + strat,
   family = binomial,
   data = moose,
   spcov_type = "none"
 )
-# fit nonspatial model using glm()
+
 bin_glm <- glm(
-  formula = presence ~ elev + strat + elev:strat,
+  formula = presence ~ elev + strat,
   family = binomial,
-  data = moose
+  data = moose,
 )
-# compare fixed effect coefficients and standard errors up to four decimals
-data.frame(
-  est_none = coef(bin),
-  est_glm = coef(bin_glm),
[…]

### 2025-09-11 inst/manuscript/extras.R

@@ -1,442 +1,0 @@
-
-
-
-
-
-
-
-
-The \code{spglm()} (for point-referenced data) and \code{spgautor()} (for areal data) functions in \pkg{spmodel} fit SPGLMs using the Laplace approximation outlined in Section$~$\ref{sec-spglm}. Both \code{spglm()} and \code{spgautor()} generally require the following four arguments: `formula`, a formula that describes the relationship between the response variable and explanatory variables; `family`, the response distribution (which can be `binomial`, `poisson`, `nbinomial`, `Gamma`, `inverse.gaussian`, or `beta`); `data`, the data frame that holds the variables in `formula` as well as spatial locations; and `spcov_type`, the spatial covariance type. The first three arguments are shared by [glm]{.fct}; thus, the transition from GLMs to SPGLMs requires only one additional argument: `spcov_type`. The \code{spglm()} spatial covariance types measure dependence as a function of Euclidean distance among observations; an example is the exponential spatial covariance:
-  \begin{equation}\label{eq-spcov-exp}
-\boldsymbol{\Sigma} = \sigma^2_{de} \exp(-\mathbf{H}/\phi) + \sigma^2_{ie}\mathbf{I},
-\end{equation}
-where $\mathbf{H}$ is a matrix of pairwise distances among all observations. \code{spglm()} currently supports 18 distinct spatial covariance functions.
-
-The \code{spgautor()} spatial covariance types measure dependence as a function of neighborhood distance among observations; an example is the simultaneous autoregressive covariance matrix:
-  \begin{equation}\label{eq-spcov-sar}
-\boldsymbol{\Sigma} = \sigma^2_{de} [(\mathbf{I} - \phi \mathbf{W})(\mathbf{I} - \phi \mathbf{W})^T]^{-1} + \sigma^2_{ie}\mathbf{I},
-\end{equation}
-where $\mathbf{W}$ is a matrix that represents the neighborhood structure among all observations. \code{spgautor()} currently supports two distinct spatial covariance functions.
-
[…]

### 2025-09-11 inst/manuscript/figures/figures.R

@@ -17,6 +17,7 @@ dat2 <- data.frame(Distance = h, Correlation = exp(-h/range2), Range = "0.5")
 range3 <- 0.8
 dat3 <- data.frame(Distance = h, Correlation = exp(-h/range3), Range = "0.8")
 dat <- bind_rows(dat1, dat2, dat3)
+dat$Range <- factor(dat$Range, levels = c("0.8", "0.5", "0.2"))
 cor_range <- ggplot(dat, aes(x = Distance, y = Correlation, color = Range, linetype = Range)) +
   geom_line(linewidth = 1.2) +
   scale_color_viridis_d() +
@@ -86,10 +87,10 @@ ggsave(
 
 
 spbin <- spglm(
-  formula = presence ~ elev + strat + elev:strat,
+  formula = presence ~ elev + strat,
   family = binomial,
   data = moose,
-  spcov_type = "spherical"
+  spcov_type = "exponential"
 )
 spbin_aug <- augment(spbin)
 
@@ -210,10 +211,10 @@ ggsave(
 ########################
 
 spnbin <- spglm(
-  formula = count ~ elev + strat + elev:strat,
+  formula = count ~ elev + strat,
   family = nbinomial,
   data = moose,
-  spcov_type = "gaussian",
+  spcov_type = "spherical",
   anisotropy = TRUE
 )
 
@@ -257,32 +258,9 @@ ggsave(
 
 
 ########################
-####### Figure 9
-########################
-
-obs_prop <- ggplot(caribou, aes(x = x, y = y, color = z, shape = tarp, size = water)) +
-  geom_point() +
-  scale_color_viridis_c(option = "H") +
-  scale_size_discrete(range = c(2, 4)) +
-  theme_bw(base_size = 14) +
-  theme(
-    legend.title = element_text(size = 12),
-    legend.text = element_text(size = 12),
-    legend.key.height = unit(0.03, 'npc')
-  )
-
-ggsave(
-  filename = paste0(fig_path, "/figure-9.png"),
-  plot = obs_prop,
-  dpi = 300,
-  height = 5,
-  width = 7.68
-)
-
-
-########################
 ####### Figure 10
 ########################
+
 texas$samp <- factor(if_else(is.na(texas$turnout), "no", "yes"), levels = c("yes", "no"))
 obs_prop2 <- ggplot(texas, aes(color = turnout, shape = samp)) +
   geom_sf(size = 2) +

### 2025-09-12 inst/manuscript/manuscript.Rmd
some updates

@@
 SPGLMs for point-referenced data (x- and y-coordinates) are fit using the \code{spglm()} function, while SPGLMs for areal (lattice, polygon) data are fit using the \code{spgautor()} function.
-Both \code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood which marginalizes over the model's fixed effects and latent mean while formally incorporating spatial covariance (i.e., spatial correlation).
+Both \code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood which marginalizes over the model's fixed effects and latent mean while formally incorporating spatial covariance.
 The inputs and outputs of \code{spglm()} and \code{spgautor()} closely resemble the \code{glm()} function from base \proglang{R}, easing the transition from GLMs to SPGLMs.
@@
   # at least one keyword must be supplied
-  formatted: [autoregressive model, geostatistical model, spatial covariance, spatial correlation]
-  plain:     [autoregressive model, geostatistical model, spatial covariance, spatial correlation]
+  formatted: [autoregressive model, geostatistical model, inverse link function, Poisson regression, link function, logistic regression, spatial covariance, spatial correlation]
+  plain:     [autoregressive model, geostatistical model, inverse link function, Poisson regression, link function, logistic regression, spatial covariance, spatial correlation]
 preamble: |
@@
 In a GLM, an $n \times 1$ response variable $\mathbf{y}$ belongs to a statistical distribution (e.g., Binomial, Poisson) with some mean and variance.
-Often, the analysis goal is to study the impact of a linear function of several explanatory variables on $\mathbf{y}$ through a GLM.
+Often, the analysis goal is to study the impact of a linear function of several explanatory variables on the mean of $\mathbf{y}$ through a GLM.
 In this context, the latent (i.e., unobserved) mean of $\mathbf{y}$, $\boldsymbol{\mu}$, is linked to these explanatory variables via a link function:
@@
 
-An approach for handling spatial data using a GLM is to assume the elements of $\mathbf{w}$ share covariance that varies spatially.
+An approach for handling spatial data using a GLM is to assume the elements of $\mathbf{w}$ exhibit covariance that varies spatially and nonspatially.
 This is achieved by adding to Equation$~$\ref{eq-glm} two random effects, $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$.
@@
 The variance parameter $\sigma^2_\epsilon$ controls the magnitude of nonspatial variability (i.e., fine-scale variation) and is often called a nugget.
-Often in spatial statistics, quantities are explicitly referenced with respect to $\mathbf{s}$, a vector of coordinates indexing the observation [@cressie1993statistics].
+Often in spatial statistics, quantities are explicitly referenced with respect to $\mathbf{s}$, a vector of spatial coordinates indexing the observation [@cressie1993statistics].
 For example, $\mathbf{y}$ and $\mathbf{X}$ may instead be written $\mathbf{y}(\mathbf{s})$ and $\mathbf{X}(\mathbf{s})$, respectively.
-We acknowledge the utility of this nomenclature but drop the explicit dependence on $\mathbf{s}$ for simplicity of notation.
+We acknowledge the utility of this nomenclature but drop the explicit dependence on $\mathbf{s}$ for simplicity of notation moving forward.
 
@@
 Working with precision matrices, however, can be more restrictive and less intuitive than working directly with the covariance matrix.
-
 The \pkg{FRK} [@sainsbury2024modeling], \pkg{glmmTMB} [@brooks2017glmmtmb], \pkg{hglm} [@ronnegard2010hglm], \pkg{mgcv} [@wood2017generalized], and \pkg{spaMM} [@rousset2014spamm] packages directly use Laplace, quasi-likelihood, or reduced-rank approaches to estimate parameters.
@@
 
-Building from @evangelou2011estimation and @bonat2016practical, @ver2024marginal proposed a novel approach for fitting SPGLMs that leverages the Laplace approximation while marginalizing over both the latent $\mathbf{w}$ and the fixed effects ($\boldsymbol{\beta}$) and accommodating spatial covariance.
+Building from @evangelou2011estimation and @bonat2016practical, @ver2024marginal proposed a novel approach for fitting SPGLMs that leverages the Laplace approximation while marginalizing over both the latent $\mathbf{w}$ and the fixed effects ($\boldsymbol{\beta}$) and accommodating general covariance structures, including spatial ones.
 This approach performed efficiently in a variety of simulation settings, generally having appropriate confidence interval coverage for the fixed effects and prediction interval coverage for $\mathbf{w}$ at new locations.
@@
 
-The \pkg{spmodel} \proglang{R} package [@dumelle2023spmodel] recently provided a full set of modeling tools for SPGLMs fit using the novel Laplace approach described by @ver2024marginal.
+The \pkg{spmodel} \proglang{R} package [@dumelle2023spmodel] recently made public a full set of modeling tools for SPGLMs fit using the novel Laplace approach described by @ver2024marginal.
 These modeling tools are approachable and mirror the familiar \code{glm()} syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs relatively seamless.
-The \code{spglm()} function fits SPGLMs for point-referenced data (e.g., x-coordinates and y-coordinates representing point locations in a field; models are sometimes called "geostatistical" models), while the \code{spgautor()} function fits SPGLMs for areal data (e.g., polygon boundaries representing geographic subsets of a region; models are sometimes called "autoregressive" models).
-For both point-referenced data and areal data, \pkg{spmodel} supports the binomial distribution for binary data, Poisson and negative binomial distributions for count data, Gamma and inverse Gaussian distributions for skewed data, and the beta distribution for proportion data.
+The \code{spglm()} function fits SPGLMs for point-referenced data (e.g., x- and y-coordinates representing point locations in a field; these models are sometimes called "geostatistical" models), while the \code{spgautor()} function fits SPGLMs for areal data (e.g., polygon boundaries representing geographic subsets of a region; these models are sometimes called "autoregressive" models).
+For both point-referenced data and areal supports, \pkg{spmodel} supports the binomial distribution for binary data, Poisson and negative binomial distributions for count data, Gamma and inverse Gaussian distributions for skewed data, and the beta distribution for proportion data.
 There are 20 different spatial covariance structures available including the exponential, Gaussian, and spherical for point-referenced data (Figure$~$\ref{fig-type}) and the conditional autoregressive, and simultaneous autoregressive structures for areal data.
@@
 \pkg{sdmTMB} also provides tools for working with temporal data and spatiotemporal data and provides enhanced visualizations of the model's marginal effects.
-\pkg{sdmTMB} does require a preprocessing step of constructing a mesh for the stochastic partial differential equation approach, and the density of the mesh can affect model results and computational complexity.
-On the other hand, \pkg{spmodel} does not require the construction of a mesh prior to modeling.
-\pkg{spmodel} also supports 20 different spatial covariances and models them directly, rather than using a precision matrix approximation to the Matérn spatial covariance as in \pkg{sdmTMB}.
+\pkg{sdmTMB} does require a preprocessing step of constructing a mesh prior to model fitting (using the stochastic partial differential equation approach), and the density of the mesh can affect model results and computational complexity.
+On the other hand, \pkg{spmodel} does not require the construction of a mesh prior to model fitting.
+\pkg{spmodel} supports 20 different spatial covariances and models them directly, rather than using a precision matrix approximation to the Matérn spatial covariance as in \pkg{sdmTMB}.
 \pkg{spmodel} can model data directly using neighborhood distance and autoregressive models, rather than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not be within the polygon's boundaries.
@@
 In Section$~$\ref{sec-spglm}, we provide some background for the SPGLM fitting and prediction routines in \pkg{spmodel}.
-In Section$~$\ref{sec-applications}, we provide an overview of core SPGLM functionaility in \pkg{spmodel} by modeling moose presence in Alaska, United States (US).
+In Section$~$\ref{sec-applications}, we provide an overview of core SPGLM functionality in \pkg{spmodel} by modeling moose presence in Alaska, United States (US).
 In Section$~$\ref{sec-applications2}, we model moose counts in Alaska, US; skewed lake conductivity in the Southwestern US; harbor seal abundance trend behavior in Alaska, US; and voter turnout rates in Texas, US.
@@
 Next, we describe a brief overview of the approach and how it can be used for several primary data analysis tasks [@tredennick2021practical] like model comparison, parameter estimation, inference, model diagnostics, and prediction.
-Then in Section$~$\ref{sec-applications} and Section$~$\ref{sec-applications2}, we show how to use \pkg{spmodel} to carry out these primary data analysis tasks for various data sets.
+Then in Section$~$\ref{sec-applications} and Section$~$\ref{sec-applications2}, we show how to use \pkg{spmodel} to carry out these primary data analysis tasks with various data sets.
 
@@
 where $[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}]$ is the restricted (i.e., residual) multivariate Gaussian density [@patterson1971recovery] for $\mathbf{w}$ given the explanatory variables and covariance parameters.
-Equation$~$\ref{eq-marginal2} can synonymously be written after profiling the overall variance out of $\boldsymbol{\Sigma}$, which reduces the dimension of $\boldsymbol{\theta}$ by one for optimization [@wolfinger1994computing].
 The restricted multivariate Gaussian density is given by
 \begin{align}\label{eq-reml-def}
-[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] = \frac{\exp(-\frac{1}{2}(\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}}) \boldsymbol{\Sigma}^{-1} (\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}})^T)}{(2 \pi)^{(n - p)/2} |\boldsymbol{\Sigma}|^{1/2}|\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X}|^{1/2}},
+[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] = \frac{\exp(-\frac{1}{2}(\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}}) \boldsymbol{\Sigma}^{-1} (\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}})^\top)}{(2 \pi)^{(n - p)/2} |\boldsymbol{\Sigma}|^{1/2}|\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X}|^{1/2}},
 \end{align}
-where $\tilde{\boldsymbol{\beta}} = (\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{w}$ and $|\cdot|$ denotes the determinant. Next, let 
+where $\tilde{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{w}$, $\boldsymbol{\Sigma}$ denotes the covariance matrix (of $\mathbf{w}$), and $|\cdot|$ denotes the determinant.
+Equation$~$\ref{eq-marginal2} can synonymously be written after profiling the overall variance out of $\boldsymbol{\Sigma}$, which reduces the dimension of $\boldsymbol{\theta}$ by one for optimization [@wolfinger1994computing].
+Next, let
 \begin{align}\label{eq-marginal03}
@@
 \begin{align}\label{eq-marginal4}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\hat{\mathbf{w}}} + \mathbf{g}^T(\mathbf{w} - \hat{\mathbf{w}}) + \frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T \mathbf{G} (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w},
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\hat{\mathbf{w}}} + \mathbf{g}^\top(\mathbf{w} - \hat{\mathbf{w}}) + \frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^\top \mathbf{G} (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w},
 \end{align}
@@
 \begin{align}\label{eq-marginal5}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\hat{\mathbf{w}}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T (-\mathbf{G}) (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w}.
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\hat{\mathbf{w}}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^\top (-\mathbf{G}) (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w}.
 \end{align}
@@
 Maximizing the natural logarithm of Equation$~$\ref{eq-marginal6} requires a doubly iterative process over $\boldsymbol{\theta}$ and $\varphi$ as well as $\mathbf{w}$, eventually yielding the the marginal restricted maximum likelihood estimators $\hat{\varphi}$ and $\hat{\boldsymbol{\theta}}$ and their corresponding values of $\hat{\mathbf{w}}$.
-Maximizing this log likelihood requires repeatedly evaluating $\boldsymbol{\Sigma}^{-1}$, $\mathbf{g}$, and $\mathbf{G}$; see @ver2024marginal for more details and comparisons to other approaches as well as and forms of $\mathbf{g}$ and $\mathbf{G}$ for various response distributions.
+Maximizing this log likelihood requires repeatedly evaluating $\boldsymbol{\Sigma}^{-1}$, $\mathbf{g}$, and $\mathbf{G}$; see @ver2024marginal for more details and comparisons to other approaches as well as forms of $\mathbf{g}$ and $\mathbf{G}$ for various response distributions.
 
@@
 To do so, we must condition on $\mathbf{w}$ as if it were observed and invoke properties of the laws of total expectation and variance.
-Because $\hat{\mathbf{w}}$ was optimized via the likelihood, we assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance approximately equal to $-\mathbf{H}^{-1}$ (the inverse Hessian).
+Because $\hat{\mathbf{w}}$ was obtained by maximizing the likelihood, we may assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance approximately equal to $-\mathbf{H}^{-1}$ (the inverse Hessian).
 It follows that $\text{E}(\hat{\mathbf{w}})$ is given by
@@
 \end{align}
-Putting this all together, it follows that
+Putting this all together, it follows that $\hat{\boldsymbol{\beta}}$ is unbiased for $\boldsymbol{\beta}$:
 \begin{align}
   \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
-  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}
+  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}.
 \end{align}
-and
+Moreover, it follows that
 \begin{align}
@@
   & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
-  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}
+  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}.
 \end{align}
@@
 @montgomery2021introduction contextualize three components of unusual observations: outliers, leverage, and influence.
-An observation is an outlier if it has an unusual response value relative to expectation.
+An observation is an outlier if it has an extreme response value relative to expectation.
 The response GLM residuals simply compare the observation to its fitted latent mean:
@@
 The sum of $\mathbf{d}$ is the deviance of the model fit, which quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized].
-Deviance is often used as a fit statistic; lower values of deviance imply a better model fit.
+Deviance is often used as a fit statistic; lower values of deviance imply a better model fit (compared to the observed data).
 Pearson and standardized residuals are other types of GLM residuals that involve a scaling of the response residuals; the Pearson residuals scale $\mathbf{r}_{r}$ by the square root of $\mathbf{V}$, while the standardized residuals scale the deviance residuals by $\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}$, where $\mathbf{L}_{ii}$ is the $i$th diagonal element of the leverage matrix, which we discuss next.
@@
 The larger the value of $\mathbf{c}_i$, the more severe the influence from the $i$th observation.
-@montgomery2021introduction provide guidance for interpreting these types of statistics, including cutoffs to consider when identifying unusual residual, leverage, or influence values.
+@montgomery2021introduction provide guidance for interpreting these types of statistics, including cutoffs to consider when identifying extreme residual, leverage, or influence values.
 
@@
 \begin{align}
-  PR^2 = 1 - \frac{deviance_{fit}}{deviance_{null}},
+  PR^2 = 1 - \frac{deviance_{error}}{deviance_{total}},
 \end{align}
-where $deviance_{fit}$ is the deviance of the fitted model (sometimes called the residual deviance) and $deviance_{null}$ is the deviance of the model taking $\mathbf{X} \equiv \mathbf{1}$, a column of all ones (i.e., an intercept-only model).
-In practice, $deviance_{null}$ is derived by computing $\hat{\mathbf{w}}$ when $\mathbf{X} \equiv \mathbf{1}$ given $\hat{\boldsymbol{\theta}}$ and $\hat{\varphi}$ from the fitted model.
+where $deviance_{error}$ is the deviance of the fitted model (sometimes called the error or residual deviance) and $deviance_{total}$ is the deviance of the intercept-only model (sometimes called the total or null deviance).
+In practice, $deviance_{total}$ is derived by computing $\hat{\mathbf{w}}$ when $\mathbf{X} \equiv \mathbf{1}$ (a column of all ones), given $\hat{\boldsymbol{\theta}}$ and $\hat{\varphi}$ from the fitted model.
 Like $R^2$, $PR^2$ can be adjusted to account for the numbers of parameters estimated in a model.
-Because the $deviance_{null}$ denominator changes across fitted models (as the values of $\hat{\boldsymbol{\theta}}$ and $\hat{\varphi}$ change), this statistic should not be used as a model comparison tool.
-Rather, it should be used as an informative diagnostic tool that unique to each model fit and describes how much variability from that model is attributable to the explanatory variables.
+Like $R^2$, $PR^2$ can be adjusted to account for the numbers of parameters estimated in a model.
+Because the $deviance_{total}$ denominator changes across fitted models (as the values of $\hat{\boldsymbol{\theta}}$ and $\hat{\varphi}$ change), this statistic should not be used as a model comparison tool.
+Rather, it should be used as an informative diagnostic tool that is unique to each model fit and describes how much variability from that model is attributable to the explanatory variables.
 
@@
 We may also predict values of the latent mean (on the link scale) at new locations by leveraging the spatial covariance between observed locations and new locations (spatial prediction is also called Kriging; see @cressie1990origins).
-Again suppose that we observed $\mathbf{w}$ and we want to make predictions at $\mathbf{u}$, a vector of latent means at the new locations that follows the same SPGLM from Equation$~$\ref{eq-spglm} and having fixed effects design matrix, $\mathbf{X}_{\mathbf{u}}$.
+Again suppose that we observed $\mathbf{w}$ and we want to make predictions at $\mathbf{u}$, a vector of latent means at the new locations that follows the same SPGLM from Equation$~$\ref{eq-spglm} and has design matrix, $\mathbf{X}_{\mathbf{u}}$.
 The vector $(\mathbf{w}, \mathbf{u})^\top$ has expectation $(\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top$ and covariance matrix $\begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}$, where $\boldsymbol{\Sigma} = \text{Var}(\mathbf{w}, \mathbf{w})$, $\boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}} = \text{Var}(\mathbf{w}, \mathbf{u})$, $\boldsymbol{\Sigma}_{\mathbf{u} \mathbf{w}} = \boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}}^\top$ and $\boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} = \text{Var}(\mathbf{u}, \mathbf{u})$.
@@
 \code{moose} is an \code{sf} object, a special data frame that is supplemented with spatial information using the \pkg{sf} package in \proglang{R} [@pebesma2018sf].
-The first few rows of \code{moose} look like:
+After loading \pkg{spmodel}, the first few rows of \code{moose} look like:
+
+```{r, eval = FALSE}
+library("spmodel")
+```
 
@@
 
-There are five columns: \code{elev}, the numeric site elevation (meters); \code{strat} a stratification variable for sampling with two levels, \code{"L"} and \code{"M"}, which are categorized by landscape metrics at each site; \code{count}, the number of moose at each site; \code{presence}, a factor that indicates whether at least one moose was observed at each site (\code{0} implies no moose; \code{1} implies at least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG: 3338) projected coordinate of each site (these data are point-referenced because each observation occurs at point coordinates and are represented by a \code{POINT} geometry.
+There are five columns: \code{elev}, the numeric site elevation (meters); \code{strat} a stratification variable for sampling with two levels, \code{"L"} and \code{"M"}, which are categorized by landscape metrics at each site; \code{count}, the number of moose at each site; \code{presence}, a factor that indicates whether at least one moose was observed at each site (\code{0} implies no moose; \code{1} implies at least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG: 3338) projected coordinate of each site.
+These data are point-referenced because each observation occurs at point coordinates and are represented by a \code{POINT} geometry.
 Moose are most prevalent in the southwestern and eastern parts of the Togiak region (Figure$~$\ref{fig-moose-data}).
@@
 
-SPGLMs in \pkg{spmodel} are fit using the \code{spglm()} function.
+SPGLMs in \pkg{spmodel} for point-referenced data are fit using the \code{spglm()} function.
 The \code{spglm()} function requires four arguments: \code{formula}, the relationship between the response and explanatory variables; \code{family}, the response distribution assumed for the response variable; \code{data}, the data frame that contains the variables in \code{formula}, and \code{spcov_type}, the type of spatial covariance.
@@
 
-We use \code{spglm()} to fit a spatial GLM (i.e., here, a spatial logistic regression) quantifying the effect of elevation and strata on moose presence:
+We use \code{spglm()} to fit a SPGLM (i.e., here, a spatial logistic regression model) quantifying the effect of elevation and strata on moose presence:
 ```{r}
@@
 
-The \code{summary()} function returns a model summary that returns relevant information like the function call, deviance residuals, a coefficients table of fixed effects, the pseudo R-squared, spatial covariance parameters, and the GLM dispersion parameter (fixed at one in logistic regression):
+The \code{summary()} function returns a model summary with relevant information like the function call, deviance residuals, a coefficients table of fixed effects, the pseudo R-squared, spatial covariance parameters, and the GLM dispersion parameter (fixed at one in logistic regression):
 ```{r}
@@
 
-The model suggests that elevation is positively associated with the log odds of moose presence ($p$-value $\approx$ 0.087), after controlling for strata.
-The model also suggests that moose have a higher log odds of presence in the \code{"M"} strata compared to the  \code{"L"} strata ($p$-value < 0.001), after controlling for elevation.
+The model provides some evidence that elevation is positively associated with the log odds of moose presence ($p~$value $\approx$ 0.087), after controlling for strata.
+The model also provides strong evidence that moose have a higher log odds of presence in the \code{"M"} strata compared to the  \code{"L"} strata ($p~$value < 0.001), after controlling for elevation.
 
@@
 
-The LRT provides strong evidence that the SPGLM is preferred to the GLM ($p$-value < 0.001).
+The LRT provides strong evidence that the SPGLM is preferred to the GLM ($p~$value < 0.001).
 
@@
 The \code{loocv()} function performs leave-one-out cross validation, comparing the predicted mean (on the response scale) to the observed response variable for each hold-out observation, recomputing estimates of $\boldsymbol{\beta}$ in each iteration.
-Performing leave-one-out cross validation tends to be more computationally efficient than fitting the model, as leave-one-out cross validation requires only one set of products involving the inverse covariance matrix (the primary computational burden), while fitting traditional models requires these products for each optimization iteration.
+Performing leave-one-out cross validation tends to be more computationally efficient than fitting the model, as leave-one-out cross validation requires only one set of products involving the inverse covariance matrix (a primary computational burden), while fitting traditional models requires these products for each optimization iteration.
 After performing leave-one-out cross validation, statistics like bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) can be used to evaluate models:
@@
 The default fitted values are on the link scale and the default residuals are deviance residuals.
-Also returned by \code{augment()} are the leverage (\code{.hat}), Cook's distance (\code{.cooksd}), and standardized residuals \code{.std.resid} described in Section$~$\ref{sec:diagnostics}.
+Also returned by \code{augment()} are the leverage (\code{.hat}), Cook's distance (\code{.cooksd}), and standardized residuals (\code{.std.resid}) described in Section$~$\ref{sec:diagnostics}.
 A benefit of using \code{augment()} when \code{data} is an \code{sf} object is that the output is also an \code{sf} object, which makes it straightforward to create spatial diagnostic plots (Figure$~$\ref{fig-sp-diagnostic}).
@@
 The remaining variability ($1 - PR^2$) is allocated proportionally to \code{de} and \code{ie} according to $\sigma^2_{de}$ and $\sigma^2_{ie}$.
-This variability partitioning is a useful tool that helps quantify how much the explanatory variables, residual spatial variance, and residual nonspatial variance contribute to model fit, but as with $PR^2$, should not be used as a model comparison tool.
+This variability partitioning is a useful tool that helps quantify how much the explanatory variables, residual spatial variance, and residual nonspatial variance contribute to model fit; as with $PR^2$, it should not be used for model comparison, but rather as a helpful model diagnostic.
 
@@
 
-Predictions on the response scale are visualized alongside the fitted values ($f^{-1}(\hat{\mathbf{w}})$) in Figure$~$\ref{fig-moose-fit}.
-
 \begin{figure}
@@
 
+Predictions on the response scale are visualized alongside the fitted values ($f^{-1}(\hat{\mathbf{w}})$) in Figure$~$\ref{fig-moose-fit}.
 Prediction intervals for the probability of moose presence (on the link scale) are returned by supplying \code{interval}:
@@
 In Section$~$\ref{sec-moose-count}, we fit Poisson and negative binomial models with and without geometric anisotropy for the point-referenced moose count data.
-In Section$~$\ref{sec-lake}, we fit a Gamma model to the point-referenced lake conductivity data, showing how to fit a model with a partition factor, perform a spatial analysis of variance (ANOVA), fit contrasts for models with interactions.
+In Section$~$\ref{sec-lake}, we fit a Gamma model to the point-referenced lake conductivity data, showing how to fit a model with a partition factor, perform a spatial analysis of variance (ANOVA), and estimate contrasts for models with interactions.
 In Section$~$\ref{sec-seal}, we fit a binomial model to the areal harbor seal trend data with a nonspatial random effect.
 Finally in Section$~$\ref{sec-texas}, we fit beta models to Texas voter turnout data, which can be treated as point-referenced or areal, and use maximum likelihood to compare two models with different explanatory variables.
-Table$~$\ref{tab:additional} outlines the section number, data set, family (i.e., response distribution), geometry type, and additional \pkg{spmodel} features for each application.
+Table$~$\ref{tab:additional} outlines, for each application, the section number, data set, family (i.e., response distribution), geometry type (point-referenced or areal), and additional \pkg{spmodel} features highlighted.
 
@@
 
-Because the Poisson and negative binomial distributions have the same support (nonnegative integers), we can compare them using  AIC, AICc, or BIC:
+Because the Poisson and negative binomial distributions have the same response support (nonnegative integers), we can compare them using  AIC, AICc, or BIC:
 ```{r}
@@
 
-We model conductivity as a function of temperature, state, and lake origin, and we allow the effect of temperature to vary by state (\code{temp:state}).
+We model conductivity as a function of temperature, state, and lake origin, and we allow the effect of temperature to vary by state (\code{temp:state} interaction).
 The \code{year} partition factor (specified via \code{partition_factor}) restricts spatial covariance to be nonzero only for observations sampled during the same year.
@@
 
-The main effect for temperature and the temperature by state interaction are highly significant ($p$-value < 0.001), while the main effects for state and lake origin are not significant.
+The main effect for temperature and the temperature by state interaction are highly significant ($p~$value < 0.001), while the main effects for state and lake origin are not significant.
 
@@
 ```{r, eval=FALSE}
-library(car)
+library("car")
 ```
@@
 ```{r, eval=FALSE}
-library(emmeans)
+library("emmeans")
 ```
@@
 Conditional and simultaneous autoregressive functions characterize spatial distance through neighborhood relationships (rather than Euclidean distance) and have \code{spcov_type} values of \code{"car"} and \code{"sar"}, respectively.
-By default, Queen's distance is used to determine whether two sites are neighbors, though custom neighborhood matrices can are passed via \code{W}.
+By default, Queen's distance is used to determine whether two sites are neighbors, though custom neighborhood matrices can be passed via \code{W}.
 Row standardization is also assumed by default; this can be changed via \code{row_st}.
-Using \code{random}, we also specified a nonspatial random effect for seal stock.
+Using \code{random}, we also specified a nonspatial random effect for seal stock, which implies seals belonging to the same stock share extra covariance.
 The \code{random} argument uses similar syntax as \pkg{lme4} [@bates2015lme4] and \pkg{nlme} [@pinheiro2006mixed] to specify nonspatial random effects.
@@
 
-The \code{SE} column is the standard error on the response scale obtained from the delta method [@oehlert1992note].
+The \code{SE} column is the standard error on the response scale obtained from the delta method [@oehlert1992note; @ver2012invented].
 
@@
 When areal data are used with \code{spglm()}, the centroids of each polygon are used as the point-referenced coordinates.
-We further explore comparisons between point-referenced and aeral data in the next example.
+We further explore comparisons between point-referenced and areal data in the next example.
 
@@
 A benefit of benefit of REML is that it can yield unbiased estimates of covariance parameters [@cressie1993asymptotic], but a drawback is that likelihood-based statistics are only valid for model comparison when the models have the same explanatory variable and fixed effect structure (because the error contrasts used to construct the REML likelihood change based on $\mathbf{X}$ and $\boldsymbol{\beta}$).
-In contrast, ML estimators are biased, though in practice this bias may be small.
+In contrast, ML estimators are generally biased for covariance parameters, though in practice this bias tends to be small.
 Moreover, when using ML, likelihood-based comparisons are valid for models having different explanatory variable and fixed effect structures.
-Using ML, we can evaluate the significance of log income on voter turnout with a likelihood ratio test:
+Using ML, we can evaluate the significance of log income on voter turnout using a likelihood ratio test:
 ```{r}
@@
 
-The likelihood ratio test suggests that log income is significantly related to voter turnout ($p$-value < 0.001).
+The likelihood ratio test provides strong evidence that log income is significantly related to voter turnout ($p~$value < 0.001).
 Alternatively, we could have instead used a different likelihood-based statistic like AIC:
@@
 
-SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace approximation that simultaneously marginalizes over the latent (i.e., unobserved) random effects and the fixed effects, $\boldsymbol{\beta}$.
-The approach is very flexible and accommodates general response distributions and covariance structures, though here we focus on spatial applications.
-\pkg{spmodel}'s \code{spglm()} and \code{spgautor()} fit SPGLMs that are similar in structure and syntax as base \proglang{R}'s \code{glm()} function, easing the transition from GLMs to SPGLMs for practitioners.
+SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace approximation that simultaneously marginalizes over the latent (i.e., unobserved) random effects and the fixed effects.
+\pkg{spmodel}'s \code{spglm()} (for point-referenced data) and \code{spgautor()} (for areal data) fit SPGLMs that are similar in structure and syntax as base \proglang{R}'s \code{glm()} function, easing the transition from GLMs to SPGLMs for practitioners.
 The \code{spglm()} and \code{spgautor()} functions support six response distributions for binary, count, and skewed data and 20 spatial covariance functions.
-\pkg{spmodel} also provides a suite of tools for data visualization, inference, model diagnostics, and prediction, providing a framework that can be used for all stages of a data analysis.
+\pkg{spmodel} has a suite of tools for data visualization, inference, model diagnostics, and prediction, providing a framework that can be used for all stages of a data analysis.
 There are many additional \pkg{spmodel} features that are not covered here, including fitting multiple models simultaneously, fixing spatial covariance and dispersion parameters at known values, fitting models to large non-Gaussian data having thousands of observations via spatial indexing [@ver2023indexing], incorporating spatial dependence in machine learning (e.g., random forests; @breiman2001random), simulating spatially dependent data (e.g., \code{spbinom()}, \code{sprpois()}, etc.), and more.
@@
 
-# Computational details {.unnumbered}
+# Data and code availability {.unnumbered}
 
-The results in this paper were obtained using [R]{.proglang} 4.4.0 with the
+The results in this manuscript were obtained using [R]{.proglang} 4.4.0 with the
 \pkg{spmodel} 0.11.0 package. Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.
-
-# Data and code availability {.unnumbered}
 
@@
 All data used are part of the \pkg{spmodel} [R]{.proglang} package available for download from CRAN at [https://CRAN.R-project.org/package=spmodel](https://CRAN.R-project.org/package=spmodel).
+Results were obtained using [R]{.proglang} 4.4.0 with the \pkg{spmodel} 0.11.0 package.
+Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.
 
@@
 
-We would like to thank initial reviewers and editors for feedback that has greatly improved the manuscript.
+We would like to genuinely thank the associate editor, anonymous reviewers, and editorial staff for significant support and feedback that greatly improved the manuscript.
 
-The views expressed in this manuscript are those of the authors and do not necessarily represent the views or policies of the U.S.
+The views expressed in this article are those of the author(s) and do not necessarily represent the views or policies of the U.S.
+government, U.S.
 Environmental Protection Agency or the National Oceanic and Atmospheric Administration.
-Any mention of trade names, products, or services does not imply
-an endorsement by the U.S. government, the U.S. Environmental Protection Agency, or the National Oceanic and Atmospheric
-Administration. The U.S. Environmental Protection Agency and the National Oceanic and Atmospheric Administration do not endorse
-any commercial products, services or enterprises.
+Mention of trade names or commercial products does not constitute endorsement or recommendation for use.
 

### 2025-09-12 inst/manuscript/manuscript.tex
some updates

@@
 which marginalizes over the model's fixed effects and latent mean while
-formally incorporating spatial covariance (i.e., spatial correlation).
-The inputs and outputs of \code{spglm()} and \code{spgautor()} closely
-resemble the \code{glm()} function from base \proglang{R}, easing the
-transition from GLMs to SPGLMs. \pkg{spmodel} provides and builds upon
-several commonly used helper functions for model building like
-\code{summary()}, \code{plot()}, \code{fitted()}, and \code{tidy()},
-among others. Spatial predictions of the latent mean at unobserved
-locations are obtained using \code{predict()} or \code{augment()}.
-\pkg{spmodel} accommodates myriad advanced modeling features like
-geometric anisotrpoy, nonspatial random effects, analysis of variance,
-and more. Throughout, we use \pkg{spmodel} to fit SPGLMs to moose
-presence and counts in Alaska, United States (US), skewed conductivity
-data in the Southwestern US, harbor seal abundance trends in Alaska, US,
-and voter turnout rates in Texas, US.
+formally incorporating spatial covariance. The inputs and outputs of
+\code{spglm()} and \code{spgautor()} closely resemble the \code{glm()}
+function from base \proglang{R}, easing the transition from GLMs to
+SPGLMs. \pkg{spmodel} provides and builds upon several commonly used
+helper functions for model building like \code{summary()},
+\code{plot()}, \code{fitted()}, and \code{tidy()}, among others. Spatial
+predictions of the latent mean at unobserved locations are obtained
+using \code{predict()} or \code{augment()}. \pkg{spmodel} accommodates
+myriad advanced modeling features like geometric anisotrpoy, nonspatial
+random effects, analysis of variance, and more. Throughout, we use
+\pkg{spmodel} to fit SPGLMs to moose presence and counts in Alaska,
+United States (US), skewed conductivity data in the Southwestern US,
+harbor seal abundance trends in Alaska, US, and voter turnout rates in
+Texas, US.
 }
 
-\Keywords{autoregressive model, geostatistical model, spatial
-covariance, spatial correlation}
-\Plainkeywords{autoregressive model, geostatistical model, spatial
-covariance, spatial correlation}
+\Keywords{autoregressive model, geostatistical model, inverse link
+function, Poisson regression, link function, logistic
+regression, spatial covariance, spatial correlation}
+\Plainkeywords{autoregressive model, geostatistical model, inverse link
+function, Poisson regression, link function, logistic
+regression, spatial covariance, spatial correlation}
 
@@
 Often, the analysis goal is to study the impact of a linear function of
-several explanatory variables on \(\mathbf{y}\) through a GLM. In this
-context, the latent (i.e., unobserved) mean of \(\mathbf{y}\),
-\(\boldsymbol{\mu}\), is linked to these explanatory variables via a
-link function: \begin{align}\label{eq-glm}
+several explanatory variables on the mean of \(\mathbf{y}\) through a
+GLM. In this context, the latent (i.e., unobserved) mean of
+\(\mathbf{y}\), \(\boldsymbol{\mu}\), is linked to these explanatory
+variables via a link function: \begin{align}\label{eq-glm}
 f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
@@
 An approach for handling spatial data using a GLM is to assume the
-elements of \(\mathbf{w}\) share covariance that varies spatially. This
-is achieved by adding to Equation\(~\)\ref{eq-glm} two random effects,
-\(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\). The random effect
-\(\boldsymbol{\tau}\) is an \(n \times 1\) column vector of spatially
-dependent random errors. We assume that
+elements of \(\mathbf{w}\) exhibit covariance that varies spatially and
+nonspatially. This is achieved by adding to Equation\(~\)\ref{eq-glm}
+two random effects, \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\).
+The random effect \(\boldsymbol{\tau}\) is an \(n \times 1\) column
+vector of spatially dependent random errors. We assume that
 \(\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}\) and
@@
 Often in spatial statistics, quantities are explicitly referenced with
-respect to \(\mathbf{s}\), a vector of coordinates indexing the
+respect to \(\mathbf{s}\), a vector of spatial coordinates indexing the
 observation \citep{cressie1993statistics}. For example, \(\mathbf{y}\)
@@
 this nomenclature but drop the explicit dependence on \(\mathbf{s}\) for
-simplicity of notation.
+simplicity of notation moving forward.
 
@@
 Working with precision matrices, however, can be more restrictive and
-less intuitive than working directly with the covariance matrix.
-
-The \pkg{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
+less intuitive than working directly with the covariance matrix. The
+\pkg{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
 \citep{brooks2017glmmtmb}, \pkg{hglm} \citep{ronnegard2010hglm},
@@
 while marginalizing over both the latent \(\mathbf{w}\) and the fixed
-effects (\(\boldsymbol{\beta}\)) and accommodating spatial covariance.
-This approach performed efficiently in a variety of simulation settings,
-generally having appropriate confidence interval coverage for the fixed
-effects and prediction interval coverage for \(\mathbf{w}\) at new
-locations. The approach performed similarly to the Bayesian SPGLM
-approach in \pkg{spBayes} and the automatic differentiation SPGLM
-approach in \pkg{glmmTMB} but was much faster. At small sample sizes,
-the approach outperformed the approximate Bayesian SPGLM approach in
-\pkg{R-INLA} and had similar computational times. For moderate sample
-sizes, it performed similarly to \pkg{R-INLA}, though \pkg{R-INLA} was
-faster. The novel Laplace approach is particularly attractive for two
-reasons. First, it is general enough that it can be applied to any
-covariance structure (not just spatial). Second, after estimating the
-covariance parameters, analytical solutions exist for the fixed effects
-(and their standard errors) as well as predictions of the latent
-\(\mathbf{w}\) at new locations (and their standard errors).
+effects (\(\boldsymbol{\beta}\)) and accommodating general covariance
+structures, including spatial ones. This approach performed efficiently
+in a variety of simulation settings, generally having appropriate
+confidence interval coverage for the fixed effects and prediction
+interval coverage for \(\mathbf{w}\) at new locations. The approach
+performed similarly to the Bayesian SPGLM approach in \pkg{spBayes} and
+the automatic differentiation SPGLM approach in \pkg{glmmTMB} but was
+much faster. At small sample sizes, the approach outperformed the
+approximate Bayesian SPGLM approach in \pkg{R-INLA} and had similar
+computational times. For moderate sample sizes, it performed similarly
+to \pkg{R-INLA}, though \pkg{R-INLA} was faster. The novel Laplace
+approach is particularly attractive for two reasons. First, it is
+general enough that it can be applied to any covariance structure (not
+just spatial). Second, after estimating the covariance parameters,
+analytical solutions exist for the fixed effects (and their standard
+errors) as well as predictions of the latent \(\mathbf{w}\) at new
+locations (and their standard errors).
 
 The \pkg{spmodel} \proglang{R} package \citep{dumelle2023spmodel}
-recently provided a full set of modeling tools for SPGLMs fit using the
-novel Laplace approach described by \citet{ver2024marginal}. These
+recently made public a full set of modeling tools for SPGLMs fit using
+the novel Laplace approach described by \citet{ver2024marginal}. These
 modeling tools are approachable and mirror the familiar \code{glm()}
@@
 relatively seamless. The \code{spglm()} function fits SPGLMs for
-point-referenced data (e.g., x-coordinates and y-coordinates
-representing point locations in a field; models are sometimes called
+point-referenced data (e.g., x- and y-coordinates representing point
+locations in a field; these models are sometimes called
 ``geostatistical'' models), while the \code{spgautor()} function fits
 SPGLMs for areal data (e.g., polygon boundaries representing geographic
-subsets of a region; models are sometimes called ``autoregressive''
-models). For both point-referenced data and areal data, \pkg{spmodel}
-supports the binomial distribution for binary data, Poisson and negative
-binomial distributions for count data, Gamma and inverse Gaussian
-distributions for skewed data, and the beta distribution for proportion
-data. There are 20 different spatial covariance structures available
-including the exponential, Gaussian, and spherical for point-referenced
-data (Figure\(~\)\ref{fig-type}) and the conditional autoregressive, and
-simultaneous autoregressive structures for areal data. \pkg{spmodel}
-provides tools for commonly used model summaries, visualizations, and
-diagnostics (e.g., Cook's distance) using standard \proglang{R} helper
-functions like \code{summary()}, \code{plot()}, \code{fitted()}, and
-\code{tidy()}, among others. \pkg{spmodel} also provides tools to
-predict \(\mathbf{w}\) at new locations and quantify uncertainty in
-those prediction using \code{predict()} and \code{augment()}. This core
-functionality, combined with several advanced features we describe
-throughout the manuscript, enables \pkg{spmodel} to introduce novel,
-important SPGLM modeling tools previously missing from the existing
-\proglang{R} ecosystem.
+subsets of a region; these models are sometimes called
+``autoregressive'' models). For both point-referenced data and areal
+supports, \pkg{spmodel} supports the binomial distribution for binary
+data, Poisson and negative binomial distributions for count data, Gamma
+and inverse Gaussian distributions for skewed data, and the beta
+distribution for proportion data. There are 20 different spatial
+covariance structures available including the exponential, Gaussian, and
+spherical for point-referenced data (Figure\(~\)\ref{fig-type}) and the
+conditional autoregressive, and simultaneous autoregressive structures
+for areal data. \pkg{spmodel} provides tools for commonly used model
+summaries, visualizations, and diagnostics (e.g., Cook's distance) using
+standard \proglang{R} helper functions like \code{summary()},
+\code{plot()}, \code{fitted()}, and \code{tidy()}, among others.
+\pkg{spmodel} also provides tools to predict \(\mathbf{w}\) at new
+locations and quantify uncertainty in those prediction using
+\code{predict()} and \code{augment()}. This core functionality, combined
+with several advanced features we describe throughout the manuscript,
+enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools
+previously missing from the existing \proglang{R} ecosystem.
 
@@
 marginal effects. \pkg{sdmTMB} does require a preprocessing step of
-constructing a mesh for the stochastic partial differential equation
-approach, and the density of the mesh can affect model results and
-computational complexity. On the other hand, \pkg{spmodel} does not
-require the construction of a mesh prior to modeling. \pkg{spmodel} also
-supports 20 different spatial covariances and models them directly,
-rather than using a precision matrix approximation to the Matérn spatial
-covariance as in \pkg{sdmTMB}. \pkg{spmodel} can model data directly
-using neighborhood distance and autoregressive models, rather than
-relying on the polygon centroid (as in \pkg{sdmTMB}), which may not be
-within the polygon's boundaries. \pkg{spmodel} provides experimental
-design tools (e.g., analysis of variance, contrasts), supports \pkg{sf}
-objects in modeling and prediction functions \citep{pebesma2018sf}, has
-several specialized model diagnostics like leverage values and Cook's
-distances, and has analytic solutions for fixed effect and prediction
-standard errors. Other similarities and differences do exist between
-\pkg{sdmTMB} and \pkg{spmodel}, and both packages continue to evolve.
-Overall, we believe that these packages are complementary and enhance
-the suite of SPGLM tools accessible to practitioners.
+constructing a mesh prior to model fitting (using the stochastic partial
+differential equation approach), and the density of the mesh can affect
+model results and computational complexity. On the other hand,
+\pkg{spmodel} does not require the construction of a mesh prior to model
+fitting. \pkg{spmodel} supports 20 different spatial covariances and
+models them directly, rather than using a precision matrix approximation
+to the Matérn spatial covariance as in \pkg{sdmTMB}. \pkg{spmodel} can
+model data directly using neighborhood distance and autoregressive
+models, rather than relying on the polygon centroid (as in
+\pkg{sdmTMB}), which may not be within the polygon's boundaries.
+\pkg{spmodel} provides experimental design tools (e.g., analysis of
+variance, contrasts), supports \pkg{sf} objects in modeling and
+prediction functions \citep{pebesma2018sf}, has several specialized
+model diagnostics like leverage values and Cook's distances, and has
+analytic solutions for fixed effect and prediction standard errors.
+Other similarities and differences do exist between \pkg{sdmTMB} and
+\pkg{spmodel}, and both packages continue to evolve. Overall, we believe
+that these packages are complementary and enhance the suite of SPGLM
+tools accessible to practitioners.
 
@@
 Section\(~\)\ref{sec-applications}, we provide an overview of core SPGLM
-functionaility in \pkg{spmodel} by modeling moose presence in Alaska,
+functionality in \pkg{spmodel} by modeling moose presence in Alaska,
 United States (US). In Section\(~\)\ref{sec-applications2}, we model
@@
 Section\(~\)\ref{sec-applications2}, we show how to use \pkg{spmodel} to
-carry out these primary data analysis tasks for various data sets.
+carry out these primary data analysis tasks with various data sets.
 
@@
 \citep{patterson1971recovery} for \(\mathbf{w}\) given the explanatory
-variables and covariance parameters. Equation\(~\)\ref{eq-marginal2} can
-synonymously be written after profiling the overall variance out of
-\(\boldsymbol{\Sigma}\), which reduces the dimension of
-\(\boldsymbol{\theta}\) by one for optimization
-\citep{wolfinger1994computing}. The restricted multivariate Gaussian
-density is given by \begin{align}\label{eq-reml-def}
-[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] = \frac{\exp(-\frac{1}{2}(\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}}) \boldsymbol{\Sigma}^{-1} (\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}})^T)}{(2 \pi)^{(n - p)/2} |\boldsymbol{\Sigma}|^{1/2}|\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X}|^{1/2}},
+variables and covariance parameters. The restricted multivariate
+Gaussian density is given by \begin{align}\label{eq-reml-def}
+[\mathbf{w} | \mathbf{X}, \boldsymbol{\theta}] = \frac{\exp(-\frac{1}{2}(\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}}) \boldsymbol{\Sigma}^{-1} (\mathbf{y} - \mathbf{X}\tilde{\boldsymbol{\beta}})^\top)}{(2 \pi)^{(n - p)/2} |\boldsymbol{\Sigma}|^{1/2}|\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X}|^{1/2}},
 \end{align} where
-\(\tilde{\boldsymbol{\beta}} = (\mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^T \boldsymbol{\Sigma}^{-1} \mathbf{w}\)
-and \(|\cdot|\) denotes the determinant. Next, let
+\(\tilde{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})^{-1} \mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{w}\),
+\(\boldsymbol{\Sigma}\) denotes the covariance matrix (of
+\(\mathbf{w}\)), and \(|\cdot|\) denotes the determinant.
+Equation\(~\)\ref{eq-marginal2} can synonymously be written after
+profiling the overall variance out of \(\boldsymbol{\Sigma}\), which
+reduces the dimension of \(\boldsymbol{\theta}\) by one for optimization
+\citep{wolfinger1994computing}. Next, let
 \begin{align}\label{eq-marginal03}
@@
 \begin{align}\label{eq-marginal4}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\hat{\mathbf{w}}} + \mathbf{g}^T(\mathbf{w} - \hat{\mathbf{w}}) + \frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T \mathbf{G} (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w},
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\hat{\mathbf{w}}} + \mathbf{g}^\top(\mathbf{w} - \hat{\mathbf{w}}) + \frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^\top \mathbf{G} (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w},
 \end{align} where \(\mathbf{g}\) and \(\mathbf{G}\) are the gradient and
@@
 \(\mathbf{g} = \mathbf{0}\), \begin{align}\label{eq-marginal5}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\hat{\mathbf{w}}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^T (-\mathbf{G}) (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w}.
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\hat{\mathbf{w}}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^\top (-\mathbf{G}) (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w}.
 \end{align} The integral in Equation\(~\)\ref{eq-marginal5} can be
@@
 \(\mathbf{G}\); see \citet{ver2024marginal} for more details and
-comparisons to other approaches as well as and forms of \(\mathbf{g}\)
-and \(\mathbf{G}\) for various response distributions.
+comparisons to other approaches as well as forms of \(\mathbf{g}\) and
+\(\mathbf{G}\) for various response distributions.
 
@@
 invoke properties of the laws of total expectation and variance. Because
-\(\hat{\mathbf{w}}\) was optimized via the likelihood, we assume that
-given \(\mathbf{w}\), \(\hat{\mathbf{w}}\) has mean \(\mathbf{w}\) and
-variance approximately equal to \(-\mathbf{H}^{-1}\) (the inverse
-Hessian). It follows that \(\text{E}(\hat{\mathbf{w}})\) is given by
-\begin{align}
+\(\hat{\mathbf{w}}\) was obtained by maximizing the likelihood, we may
+assume that given \(\mathbf{w}\), \(\hat{\mathbf{w}}\) has mean
+\(\mathbf{w}\) and variance approximately equal to \(-\mathbf{H}^{-1}\)
+(the inverse Hessian). It follows that \(\text{E}(\hat{\mathbf{w}})\) is
+given by \begin{align}
   \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
@@
   & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
-\end{align} Putting this all together, it follows that \begin{align}
+\end{align} Putting this all together, it follows that
+\(\hat{\boldsymbol{\beta}}\) is unbiased for \(\boldsymbol{\beta}\):
+\begin{align}
   \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
-  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}
-\end{align} and \begin{align}
+  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}.
+\end{align} Moreover, it follows that \begin{align}
   \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
@@
   & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
-  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}
+  & = \mathbf{B}(-\mathbf{H})^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}.
 \end{align} In practice, \(\text{Var}(\hat{\boldsymbol{\beta}})\) is
@@
 components of unusual observations: outliers, leverage, and influence.
-An observation is an outlier if it has an unusual response value
+An observation is an outlier if it has an extreme response value
 relative to expectation. The response GLM residuals simply compare the
@@
 fitted model \citep{myers2012generalized}. Deviance is often used as a
-fit statistic; lower values of deviance imply a better model fit.
-Pearson and standardized residuals are other types of GLM residuals that
-involve a scaling of the response residuals; the Pearson residuals scale
-\(\mathbf{r}_{r}\) by the square root of \(\mathbf{V}\), while the
-standardized residuals scale the deviance residuals by
-\(\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}\), where \(\mathbf{L}_{ii}\) is
-the \(i\)th diagonal element of the leverage matrix, which we discuss
-next.
+fit statistic; lower values of deviance imply a better model fit
+(compared to the observed data). Pearson and standardized residuals are
+other types of GLM residuals that involve a scaling of the response
+residuals; the Pearson residuals scale \(\mathbf{r}_{r}\) by the square
+root of \(\mathbf{V}\), while the standardized residuals scale the
+deviance residuals by \(\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}\), where
+\(\mathbf{L}_{ii}\) is the \(i\)th diagonal element of the leverage
+matrix, which we discuss next.
 
@@
 these types of statistics, including cutoffs to consider when
-identifying unusual residual, leverage, or influence values.
+identifying extreme residual, leverage, or influence values.
 
@@
 ratio from the linear model with the deviance ratio: \begin{align}
-  PR^2 = 1 - \frac{deviance_{fit}}{deviance_{null}},
-\end{align} where \(deviance_{fit}\) is the deviance of the fitted model
-(sometimes called the residual deviance) and \(deviance_{null}\) is the
-deviance of the model taking \(\mathbf{X} \equiv \mathbf{1}\), a column
-of all ones (i.e., an intercept-only model). In practice,
-\(deviance_{null}\) is derived by computing \(\hat{\mathbf{w}}\) when
-\(\mathbf{X} \equiv \mathbf{1}\) given \(\hat{\boldsymbol{\theta}}\) and
-\(\hat{\varphi}\) from the fitted model. Like \(R^2\), \(PR^2\) can be
+  PR^2 = 1 - \frac{deviance_{error}}{deviance_{total}},
+\end{align} where \(deviance_{error}\) is the deviance of the fitted
+model (sometimes called the error or residual deviance) and
+\(deviance_{total}\) is the deviance of the intercept-only model
+(sometimes called the total or null deviance). In practice,
+\(deviance_{total}\) is derived by computing \(\hat{\mathbf{w}}\) when
+\(\mathbf{X} \equiv \mathbf{1}\) (a column of all ones), given
+\(\hat{\boldsymbol{\theta}}\) and \(\hat{\varphi}\) from the fitted
+model. Like \(R^2\), \(PR^2\) can be adjusted to account for the numbers
+of parameters estimated in a model. Like \(R^2\), \(PR^2\) can be
 adjusted to account for the numbers of parameters estimated in a model.
-Because the \(deviance_{null}\) denominator changes across fitted models
-(as the values of \(\hat{\boldsymbol{\theta}}\) and \(\hat{\varphi}\)
-change), this statistic should not be used as a model comparison tool.
-Rather, it should be used as an informative diagnostic tool that unique
-to each model fit and describes how much variability from that model is
-attributable to the explanatory variables.
+Because the \(deviance_{total}\) denominator changes across fitted
+models (as the values of \(\hat{\boldsymbol{\theta}}\) and
+\(\hat{\varphi}\) change), this statistic should not be used as a model
+comparison tool. Rather, it should be used as an informative diagnostic
+tool that is unique to each model fit and describes how much variability
+from that model is attributable to the explanatory variables.
 
@@
 vector of latent means at the new locations that follows the same SPGLM
-from Equation\(~\)\ref{eq-spglm} and having fixed effects design matrix,
+from Equation\(~\)\ref{eq-spglm} and has design matrix,
 \(\mathbf{X}_{\mathbf{u}}\). The vector
@@
 spatial information using the \pkg{sf} package in \proglang{R}
-\citep{pebesma2018sf}. The first few rows of \code{moose} look like:
+\citep{pebesma2018sf}. After loading \pkg{spmodel}, the first few rows
+of \code{moose} look like:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library("spmodel")
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG:
-3338) projected coordinate of each site (these data are point-referenced
+3338) projected coordinate of each site. These data are point-referenced
 because each observation occurs at point coordinates and are represented
@@
 
-SPGLMs in \pkg{spmodel} are fit using the \code{spglm()} function. The
-\code{spglm()} function requires four arguments: \code{formula}, the
-relationship between the response and explanatory variables;
-\code{family}, the response distribution assumed for the response
-variable; \code{data}, the data frame that contains the variables in
-\code{formula}, and \code{spcov_type}, the type of spatial covariance.
-The \code{formula}, \code{family}, and \code{data} arguments are the
-three required arguments to \code{glm()} for nonspatial GLMs. So, the
-transition from \code{glm()} to \code{spglm()} simply requires one
-additional argument: \code{spcov_type}. When \code{data} is not an
+SPGLMs in \pkg{spmodel} for point-referenced data are fit using the
+\code{spglm()} function. The \code{spglm()} function requires four
+arguments: \code{formula}, the relationship between the response and
+explanatory variables; \code{family}, the response distribution assumed
+for the response variable; \code{data}, the data frame that contains the
+variables in \code{formula}, and \code{spcov_type}, the type of spatial
+covariance. The \code{formula}, \code{family}, and \code{data} arguments
+are the three required arguments to \code{glm()} for nonspatial GLMs.
+So, the transition from \code{glm()} to \code{spglm()} simply requires
+one additional argument: \code{spcov_type}. When \code{data} is not an
 \code{sf} object, \code{spglm()} also requires the \code{xcoord} and
@@
 
-We use \code{spglm()} to fit a spatial GLM (i.e., here, a spatial
-logistic regression) quantifying the effect of elevation and strata on
+We use \code{spglm()} to fit a SPGLM (i.e., here, a spatial logistic
+regression model) quantifying the effect of elevation and strata on
 moose presence:
@@
 
-The \code{summary()} function returns a model summary that returns
-relevant information like the function call, deviance residuals, a
-coefficients table of fixed effects, the pseudo R-squared, spatial
-covariance parameters, and the GLM dispersion parameter (fixed at one in
-logistic regression):
+The \code{summary()} function returns a model summary with relevant
+information like the function call, deviance residuals, a coefficients
+table of fixed effects, the pseudo R-squared, spatial covariance
+parameters, and the GLM dispersion parameter (fixed at one in logistic
+regression):
 
@@
 
-The model suggests that elevation is positively associated with the log
-odds of moose presence (\(p\)-value \(\approx\) 0.087), after
-controlling for strata. The model also suggests that moose have a higher
-log odds of presence in the \code{"M"} strata compared to the \code{"L"}
-strata (\(p\)-value \textless{} 0.001), after controlling for elevation.
+The model provides some evidence that elevation is positively associated
+with the log odds of moose presence (\(p~\)value \(\approx\) 0.087),
+after controlling for strata. The model also provides strong evidence
+that moose have a higher log odds of presence in the \code{"M"} strata
+compared to the \code{"L"} strata (\(p~\)value \textless{} 0.001), after
+controlling for elevation.
 
@@
 The LRT provides strong evidence that the SPGLM is preferred to the GLM
-(\(p\)-value \textless{} 0.001).
+(\(p~\)value \textless{} 0.001).
 
@@
 cross validation requires only one set of products involving the inverse
-covariance matrix (the primary computational burden), while fitting
+covariance matrix (a primary computational burden), while fitting
 traditional models requires these products for each optimization
@@
 \code{augment()} are the leverage (\code{.hat}), Cook's distance
-(\code{.cooksd}), and standardized residuals \code{.std.resid} described
-in Section\(~\)\ref{sec:diagnostics}. A benefit of using
+(\code{.cooksd}), and standardized residuals (\code{.std.resid})
+described in Section\(~\)\ref{sec:diagnostics}. A benefit of using
 \code{augment()} when \code{data} is an \code{sf} object is that the
@@
 helps quantify how much the explanatory variables, residual spatial
-variance, and residual nonspatial variance contribute to model fit, but
-as with \(PR^2\), should not be used as a model comparison tool.
+variance, and residual nonspatial variance contribute to model fit; as
+with \(PR^2\), it should not be used for model comparison, but rather as
+a helpful model diagnostic.
 
@@
 \end{CodeChunk}
-
-Predictions on the response scale are visualized alongside the fitted
-values (\(f^{-1}(\hat{\mathbf{w}})\)) in Figure\(~\)\ref{fig-moose-fit}.
 
@@
 
+Predictions on the response scale are visualized alongside the fitted
+values (\(f^{-1}(\hat{\mathbf{w}})\)) in Figure\(~\)\ref{fig-moose-fit}.
 Prediction intervals for the probability of moose presence (on the link
@@
 with a partition factor, perform a spatial analysis of variance (ANOVA),
-fit contrasts for models with interactions. In
+and estimate contrasts for models with interactions. In
 Section\(~\)\ref{sec-seal}, we fit a binomial model to the areal harbor
@@
 likelihood to compare two models with different explanatory variables.
-Table\(~\)\ref{tab:additional} outlines the section number, data set,
-family (i.e., response distribution), geometry type, and additional
-\pkg{spmodel} features for each application.
+Table\(~\)\ref{tab:additional} outlines, for each application, the
+section number, data set, family (i.e., response distribution), geometry
+type (point-referenced or areal), and additional \pkg{spmodel} features
+highlighted.
 
@@
 Because the Poisson and negative binomial distributions have the same
-support (nonnegative integers), we can compare them using AIC, AICc, or
-BIC:
+response support (nonnegative integers), we can compare them using AIC,
+AICc, or BIC:
 
@@
 origin, and we allow the effect of temperature to vary by state
-(\code{temp:state}). The \code{year} partition factor (specified via
-\code{partition_factor}) restricts spatial covariance to be nonzero only
-for observations sampled during the same year. Data were collected in
-2012 and 2017, so this partition factor assumes independence between
-observations in 2012 and 2017. While we used the partition factor here
-illustratively, more generally, the utility of partition factors can be
-highly context dependent.
+(\code{temp:state} interaction). The \code{year} partition factor
+(specified via \code{partition_factor}) restricts spatial covariance to
+be nonzero only for observations sampled during the same year. Data were
+collected in 2012 and 2017, so this partition factor assumes
+independence between observations in 2012 and 2017. While we used the
+partition factor here illustratively, more generally, the utility of
+partition factors can be highly context dependent.
 
@@
 The main effect for temperature and the temperature by state interaction
-are highly significant (\(p\)-value \textless{} 0.001), while the main
+are highly significant (\(p~\)value \textless{} 0.001), while the main
 effects for state and lake origin are not significant.
@@
 \begin{CodeInput}
-R> library(car)
+R> library("car")
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> library(emmeans)
+R> library("emmeans")
 \end{CodeInput}
@@
 distance is used to determine whether two sites are neighbors, though
-custom neighborhood matrices can are passed via \code{W}. Row
+custom neighborhood matrices can be passed via \code{W}. Row
 standardization is also assumed by default; this can be changed via
 \code{row_st}. Using \code{random}, we also specified a nonspatial
-random effect for seal stock. The \code{random} argument uses similar
+random effect for seal stock, which implies seals belonging to the same
+stock share extra covariance. The \code{random} argument uses similar
 syntax as \pkg{lme4} \citep{bates2015lme4} and \pkg{nlme}
@@
 The \code{SE} column is the standard error on the response scale
-obtained from the delta method \citep{oehlert1992note}.
+obtained from the delta method \citep{oehlert1992note, ver2012invented}.
 
@@
 centroids of each polygon are used as the point-referenced coordinates.
-We further explore comparisons between point-referenced and aeral data
+We further explore comparisons between point-referenced and areal data
 in the next example.
@@
 likelihood change based on \(\mathbf{X}\) and \(\boldsymbol{\beta}\)).
-In contrast, ML estimators are biased, though in practice this bias may
-be small. Moreover, when using ML, likelihood-based comparisons are
-valid for models having different explanatory variable and fixed effect
-structures. Using ML, we can evaluate the significance of log income on
-voter turnout with a likelihood ratio test:
+In contrast, ML estimators are generally biased for covariance
+parameters, though in practice this bias tends to be small. Moreover,
+when using ML, likelihood-based comparisons are valid for models having
+different explanatory variable and fixed effect structures. Using ML, we
+can evaluate the significance of log income on voter turnout using a
+likelihood ratio test:
 
@@
 
-The likelihood ratio test suggests that log income is significantly
-related to voter turnout (\(p\)-value \textless{} 0.001). Alternatively,
-we could have instead used a different likelihood-based statistic like
-AIC:
+The likelihood ratio test provides strong evidence that log income is
+significantly related to voter turnout (\(p~\)value \textless{} 0.001).
+Alternatively, we could have instead used a different likelihood-based
+statistic like AIC:
 
@@
 approximation that simultaneously marginalizes over the latent (i.e.,
-unobserved) random effects and the fixed effects,
-\(\boldsymbol{\beta}\). The approach is very flexible and accommodates
-general response distributions and covariance structures, though here we
-focus on spatial applications. \pkg{spmodel}'s \code{spglm()} and
-\code{spgautor()} fit SPGLMs that are similar in structure and syntax as
-base \proglang{R}'s \code{glm()} function, easing the transition from
-GLMs to SPGLMs for practitioners. The \code{spglm()} and
-\code{spgautor()} functions support six response distributions for
-binary, count, and skewed data and 20 spatial covariance functions.
-\pkg{spmodel} also provides a suite of tools for data visualization,
-inference, model diagnostics, and prediction, providing a framework that
-can be used for all stages of a data analysis. There are many additional
-\pkg{spmodel} features that are not covered here, including fitting
-multiple models simultaneously, fixing spatial covariance and dispersion
-parameters at known values, fitting models to large non-Gaussian data
-having thousands of observations via spatial indexing
-\citep{ver2023indexing}, incorporating spatial dependence in machine
-learning (e.g., random forests; \citet{breiman2001random}), simulating
-spatially dependent data (e.g., \code{spbinom()}, \code{sprpois()},
-etc.), and more. Further details are provided by
-\url{https://CRAN.R-project.org/package=spmodel} and links therein.
-
-\section*{Computational details}\label{computational-details}
-\addcontentsline{toc}{section}{Computational details}
-
-The results in this paper were obtained using {R} 4.4.0 with the
+unobserved) random effects and the fixed effects. \pkg{spmodel}'s
+\code{spglm()} (for point-referenced data) and \code{spgautor()} (for
+areal data) fit SPGLMs that are similar in structure and syntax as base
+\proglang{R}'s \code{glm()} function, easing the transition from GLMs to
+SPGLMs for practitioners. The \code{spglm()} and \code{spgautor()}
+functions support six response distributions for binary, count, and
+skewed data and 20 spatial covariance functions. \pkg{spmodel} has a
+suite of tools for data visualization, inference, model diagnostics, and
+prediction, providing a framework that can be used for all stages of a
+data analysis. There are many additional \pkg{spmodel} features that are
+not covered here, including fitting multiple models simultaneously,
+fixing spatial covariance and dispersion parameters at known values,
+fitting models to large non-Gaussian data having thousands of
+observations via spatial indexing \citep{ver2023indexing}, incorporating
+spatial dependence in machine learning (e.g., random forests;
+\citet{breiman2001random}), simulating spatially dependent data (e.g.,
+\code{spbinom()}, \code{sprpois()}, etc.), and more. Further details are
+provided by \url{https://CRAN.R-project.org/package=spmodel} and links
+therein.
+
+\section*{Data and code availability}\label{data-and-code-availability}
+\addcontentsline{toc}{section}{Data and code availability}
+
+The results in this manuscript were obtained using {R} 4.4.0 with the
 \pkg{spmodel} 0.11.0 package. Figures were created using the {ggplot2}
 3.5.1 package \citep{wickham2016ggplot2} and base {R}.
-
-\section*{Data and code availability}\label{data-and-code-availability}
-\addcontentsline{toc}{section}{Data and code availability}
 
@@
 part of the \pkg{spmodel} {R} package available for download from CRAN
-at \url{https://CRAN.R-project.org/package=spmodel}.
+at \url{https://CRAN.R-project.org/package=spmodel}. Results were
+obtained using {R} 4.4.0 with the \pkg{spmodel} 0.11.0 package. Figures
+were created using the {ggplot2} 3.5.1 package
+\citep{wickham2016ggplot2} and base {R}.
 
@@
 
-We would like to thank initial reviewers and editors for feedback that
-has greatly improved the manuscript.
-
-The views expressed in this manuscript are those of the authors and do
-not necessarily represent the views or policies of the U.S.
-Environmental Protection Agency or the National Oceanic and Atmospheric
-Administration. Any mention of trade names, products, or services does
-not imply an endorsement by the U.S. government, the U.S. Environmental
-Protection Agency, or the National Oceanic and Atmospheric
-Administration. The U.S. Environmental Protection Agency and the
-National Oceanic and Atmospheric Administration do not endorse any
-commercial products, services or enterprises.
+We would like to the associate editor, anonymous reviewers, and
+editorial staff for significant support and feedback that greatly
+improved the manuscript.
+
+The views expressed in this article are those of the author(s) and do
+not necessarily represent the views or policies of the U.S. government,
+U.S. Environmental Protection Agency or the National Oceanic and
+Atmospheric Administration. Mention of trade names or commercial
+products does not constitute endorsement or recommendation for use.
 

### 2025-11-14 inst/manuscript/manuscript.Rmd
some updates

@@
 The elements of $[\mathbf{y} | f^{-1}(\mathbf{w}), \varphi]$ are conditionally independent (given $\mathbf{w}$), but the elements of $[\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}]$ share spatial covariance.
-Following @harville1977maximum, we can integrate $\boldsymbol{\beta}$ out of Equation$~$\ref{eq-marginal}, which yields
+Following @harville1974bayesian, we can integrate $\boldsymbol{\beta}$ out of Equation$~$\ref{eq-marginal}, which yields
 \begin{align}\label{eq-marginal2}
@@
 \end{align}
-A second-order Taylor series expansion of $\ell_\mathbf{w}$ around a point $\hat{\mathbf{w}}$ yields
+A second-order Taylor series expansion of $\ell_\mathbf{w}$ around a point $\mathbf{w}^{*}$ yields
 \begin{align}\label{eq-marginal4}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\hat{\mathbf{w}}} + \mathbf{g}^\top(\mathbf{w} - \hat{\mathbf{w}}) + \frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^\top \mathbf{G} (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w},
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\mathbf{w}^{*}} + \mathbf{g}^\top(\mathbf{w} - \mathbf{w}^{*}) + \frac{1}{2}(\mathbf{w} - \mathbf{w}^{*})^\top \mathbf{G} (\mathbf{w} - \mathbf{w}^{*}))d\mathbf{w},
 \end{align}
-where $\mathbf{g}$ and $\mathbf{G}$ are the gradient and Hessian, respectively, of $\ell_\mathbf{w}$ with respect to $\mathbf{w}$. If $\hat{\mathbf{w}}$ is a value for which $\mathbf{g} = \mathbf{0}$,
+where $\mathbf{g}$ and $\mathbf{G}$ are the gradient and Hessian, respectively, of $\ell_\mathbf{w}$ with respect to $\mathbf{w}$. If $\mathbf{w}^{*}$ is a value for which $\mathbf{g} = \mathbf{0}$,
 \begin{align}\label{eq-marginal5}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\hat{\mathbf{w}}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^\top (-\mathbf{G}) (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w}.
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\mathbf{w}^{*}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \mathbf{w}^{*})^\top (-\mathbf{G}) (\mathbf{w} - \mathbf{w}^{*}))d\mathbf{w}.
 \end{align}
 The integral in Equation$~$\ref{eq-marginal5} can be solved by leveraging properties of the normalizing constant of a multivariate Gaussian distribution.
-Thus, rewriting $\exp(\ell_{\hat{\mathbf{w}}})$ yields
+Thus, rewriting $\exp(\ell_{\mathbf{w}^{*}})$ yields
 \begin{align}\label{eq-marginal6}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\hat{\mathbf{w}}), \varphi] [\hat{\mathbf{w}} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\hat{\mathbf{w}}}|^{-1/2}.
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{w}^{*}), \varphi] [\mathbf{w}^{*} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{w}^{*}}|^{-1/2}.
 \end{align}
 
-Maximizing the natural logarithm of Equation$~$\ref{eq-marginal6} requires a doubly iterative process over $\boldsymbol{\theta}$ and $\varphi$ as well as $\mathbf{w}$, eventually yielding the the marginal restricted maximum likelihood estimators $\hat{\varphi}$ and $\hat{\boldsymbol{\theta}}$ and their corresponding values of $\hat{\mathbf{w}}$.
-Maximizing this log likelihood requires repeatedly evaluating $\boldsymbol{\Sigma}^{-1}$, $\mathbf{g}$, and $\mathbf{G}$; see @ver2024marginal for more details and comparisons to other approaches as well as forms of $\mathbf{g}$ and $\mathbf{G}$ for various response distributions.
+Maximizing the natural logarithm of Equation$~$\ref{eq-marginal6} requires a doubly iterative process over 1) $\boldsymbol{\theta}$ and $\varphi$ and 2) the latent $\mathbf{w}$ (to find each set of $\mathbf{w}^{*}$).
+This doubly iterative process eventually yields the marginal restricted maximum likelihood estimators $\hat{\varphi}$ and $\hat{\boldsymbol{\theta}}$ and their respective values of $\mathbf{w}^{*}$, which we call $\hat{\mathbf{w}}$.
+Maximizing the marginal restricted likelihood requires repeatedly evaluating $\boldsymbol{\Sigma}^{-1}$, $\mathbf{g}$, and $\mathbf{G}$; see @ver2024marginal for more details and comparisons to other approaches as well as forms of $\mathbf{g}$ and $\mathbf{G}$ for various response distributions.
 
@@
 where $\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}$.
-However, we only observe $\hat{\mathbf{w}}$, so it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$.
-Thus, to derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$.
+However, we do not observe $\mathbf{w}$ and instead estimate $\mathbf{w}$ via $\hat{\mathbf{w}}$.
+Thus, it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$.
+To derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$.
 To do so, we must condition on $\mathbf{w}$ as if it were observed and invoke properties of the laws of total expectation and variance.
@@
 where $\mathbf{d}$ is a vector of individual deviances.
-The sum of the squared deviance residuals equals the sum of $\mathbf{d}$.
-The sum of $\mathbf{d}$ is the deviance of the model fit, which quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized].
+The sum of the squared deviance residuals equals the sum of the elements of $\mathbf{d}$, known as the deviance of the model fit.
+The deviance of the model fit quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized].
 Deviance is often used as a fit statistic; lower values of deviance imply a better model fit (compared to the observed data).

### 2025-11-14 inst/manuscript/manuscript.tex
some updates

@@
 \([\mathbf{w} | \mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\theta}]\)
-share spatial covariance. Following \citet{harville1977maximum}, we can
+share spatial covariance. Following \citet{harville1974bayesian}, we can
 integrate \(\boldsymbol{\beta}\) out of Equation\(~\)\ref{eq-marginal},
@@
 \end{align} A second-order Taylor series expansion of
-\(\ell_\mathbf{w}\) around a point \(\hat{\mathbf{w}}\) yields
+\(\ell_\mathbf{w}\) around a point \(\mathbf{w}^{*}\) yields
 \begin{align}\label{eq-marginal4}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\hat{\mathbf{w}}} + \mathbf{g}^\top(\mathbf{w} - \hat{\mathbf{w}}) + \frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^\top \mathbf{G} (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w},
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \int_{\mathbf{w}} \exp(\ell_{\mathbf{w}^{*}} + \mathbf{g}^\top(\mathbf{w} - \mathbf{w}^{*}) + \frac{1}{2}(\mathbf{w} - \mathbf{w}^{*})^\top \mathbf{G} (\mathbf{w} - \mathbf{w}^{*}))d\mathbf{w},
 \end{align} where \(\mathbf{g}\) and \(\mathbf{G}\) are the gradient and
 Hessian, respectively, of \(\ell_\mathbf{w}\) with respect to
-\(\mathbf{w}\). If \(\hat{\mathbf{w}}\) is a value for which
+\(\mathbf{w}\). If \(\mathbf{w}^{*}\) is a value for which
 \(\mathbf{g} = \mathbf{0}\), \begin{align}\label{eq-marginal5}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\hat{\mathbf{w}}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \hat{\mathbf{w}})^\top (-\mathbf{G}) (\mathbf{w} - \hat{\mathbf{w}}))d\mathbf{w}.
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx \exp(\ell_{\mathbf{w}^{*}}) \int_{\mathbf{w}} \exp(-\frac{1}{2}(\mathbf{w} - \mathbf{w}^{*})^\top (-\mathbf{G}) (\mathbf{w} - \mathbf{w}^{*}))d\mathbf{w}.
 \end{align} The integral in Equation\(~\)\ref{eq-marginal5} can be
@@
 multivariate Gaussian distribution. Thus, rewriting
-\(\exp(\ell_{\hat{\mathbf{w}}})\) yields
-\begin{align}\label{eq-marginal6}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\hat{\mathbf{w}}), \varphi] [\hat{\mathbf{w}} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\hat{\mathbf{w}}}|^{-1/2}.
+\(\exp(\ell_{\mathbf{w}^{*}})\) yields \begin{align}\label{eq-marginal6}
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{w}^{*}), \varphi] [\mathbf{w}^{*} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{w}^{*}}|^{-1/2}.
 \end{align}
@@
 Maximizing the natural logarithm of Equation\(~\)\ref{eq-marginal6}
-requires a doubly iterative process over \(\boldsymbol{\theta}\) and
-\(\varphi\) as well as \(\mathbf{w}\), eventually yielding the the
+requires a doubly iterative process over 1) \(\boldsymbol{\theta}\) and
+\(\varphi\) and 2) the latent \(\mathbf{w}\) (to find each set of
+\(\mathbf{w}^{*}\)). This doubly iterative process eventually yields the
 marginal restricted maximum likelihood estimators \(\hat{\varphi}\) and
-\(\hat{\boldsymbol{\theta}}\) and their corresponding values of
-\(\hat{\mathbf{w}}\). Maximizing this log likelihood requires repeatedly
-evaluating \(\boldsymbol{\Sigma}^{-1}\), \(\mathbf{g}\), and
-\(\mathbf{G}\); see \citet{ver2024marginal} for more details and
-comparisons to other approaches as well as forms of \(\mathbf{g}\) and
-\(\mathbf{G}\) for various response distributions.
+\(\hat{\boldsymbol{\theta}}\) and their respective values of
+\(\mathbf{w}^{*}\), which we call \(\hat{\mathbf{w}}\). Maximizing the
+marginal restricted likelihood requires repeatedly evaluating
+\(\boldsymbol{\Sigma}^{-1}\), \(\mathbf{g}\), and \(\mathbf{G}\); see
+\citet{ver2024marginal} for more details and comparisons to other
+approaches as well as forms of \(\mathbf{g}\) and \(\mathbf{G}\) for
+various response distributions.
 
@@
 \(\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\).
-However, we only observe \(\hat{\mathbf{w}}\), so it is reasonable to
-define \(\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}\). Thus,
-to derive properties of \(\hat{\boldsymbol{\beta}}\) like expectation
-and variance, we must derive these properties for \(\hat{\mathbf{w}}\).
-To do so, we must condition on \(\mathbf{w}\) as if it were observed and
+However, we do not observe \(\mathbf{w}\) and instead estimate
+\(\mathbf{w}\) via \(\hat{\mathbf{w}}\). Thus, it is reasonable to
+define \(\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}\). To
+derive properties of \(\hat{\boldsymbol{\beta}}\) like expectation and
+variance, we must derive these properties for \(\hat{\mathbf{w}}\). To
+do so, we must condition on \(\mathbf{w}\) as if it were observed and
 invoke properties of the laws of total expectation and variance. Because
@@
 \end{align} where \(\mathbf{d}\) is a vector of individual deviances.
-The sum of the squared deviance residuals equals the sum of
-\(\mathbf{d}\). The sum of \(\mathbf{d}\) is the deviance of the model
-fit, which quantifies twice the difference in log likelihoods between
-the a saturated model that fits every observation perfectly (i.e.,
-\(\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)\) for all \(i\)) and the
-fitted model \citep{myers2012generalized}. Deviance is often used as a
-fit statistic; lower values of deviance imply a better model fit
+The sum of the squared deviance residuals equals the sum of the elements
+of \(\mathbf{d}\), known as the deviance of the model fit. The deviance
+of the model fit quantifies twice the difference in log likelihoods
+between the a saturated model that fits every observation perfectly
+(i.e., \(\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)\) for all \(i\)) and
+the fitted model \citep{myers2012generalized}. Deviance is often used as
+a fit statistic; lower values of deviance imply a better model fit
 (compared to the observed data). Pearson and standardized residuals are
@@
 
-We would like to the associate editor, anonymous reviewers, and
-editorial staff for significant support and feedback that greatly
-improved the manuscript.
+We would like to genuinely thank the associate editor, anonymous
+reviewers, and editorial staff for significant support and feedback that
+greatly improved the manuscript.
 

### 2025-11-26 inst/manuscript/manuscript.Rmd
some updates

@@
 abstract: |
-  Non-Gaussian data are common in practice and include binary, count, skewed, and proprtion data types.
+  Non-Gaussian data are common in practice and include binary, count, skewed, and proportion data types.
 Often, non-Gaussian data are modeled using a generalized linear model (GLM).
@@
 The \pkg{spmodel} package in \proglang{R} provides a suite of tools for fitting spatial generalized linear models (SPGLMs) to non-Gaussian data and making spatial predictions (i.e., Kriging).
-SPGLMs for point-referenced data (x- and y-coordinates) are fit using the \code{spglm()} function, while SPGLMs for areal (lattice, polygon) data are fit using the \code{spgautor()} function.
-Both \code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood which marginalizes over the model's fixed effects and latent mean while formally incorporating spatial covariance.
+SPGLMs for point-referenced (x- and y-coordinates) support are fit using the \code{spglm()} function, while SPGLMs for areal (lattice, polygon) support are fit using the \code{spgautor()} function.
+Both \code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood which marginalizes over the model's fixed effects and latent mean while formally incorporating spatial covariance among observations.
 The inputs and outputs of \code{spglm()} and \code{spgautor()} closely resemble the \code{glm()} function from base \proglang{R}, easing the transition from GLMs to SPGLMs.
-\pkg{spmodel} provides and builds upon several commonly used helper functions for model building like \code{summary()}, \code{plot()}, \code{fitted()}, and \code{tidy()}, among others.
+\pkg{spmodel} provides and builds upon several commonly used helper functions for model building like \code{summary()}, \code{plot()}, and \code{fitted()}, among others.
 Spatial predictions of the latent mean at unobserved locations are obtained using \code{predict()} or \code{augment()}.
@@
   # at least one keyword must be supplied
-  formatted: [autoregressive model, geostatistical model, inverse link function, Poisson regression, link function, logistic regression, spatial covariance, spatial correlation]
-  plain:     [autoregressive model, geostatistical model, inverse link function, Poisson regression, link function, logistic regression, spatial covariance, spatial correlation]
+  formatted: [autoregressive model, geostatistical model, Poisson regression, link function, logistic regression, overdispersion, spatial covariance, spatial dependence]
+  plain:     [autoregressive model, geostatistical model, Poisson regression, link function, logistic regression, overdispersion, spatial covariance, spatial dependence]
 preamble: |
@@
 Non-Gaussian data that belong to an exponential family can be naturally modeled using a generalized linear model (GLM) regression framework [@nelder1972generalized; @mccullagh1989generalized].
-In a GLM, an $n \times 1$ response variable $\mathbf{y}$ belongs to a statistical distribution (e.g., Binomial, Poisson) with some mean and variance.
+In a GLM, an $n \times 1$ response variable $\mathbf{y}$ belongs to a statistical distribution (e.g., binomial, Poisson) with some mean and variance.
 Often, the analysis goal is to study the impact of a linear function of several explanatory variables on the mean of $\mathbf{y}$ through a GLM.
@@
 The GLM fixed effects ($\boldsymbol{\beta}$) are typically estimated via maximum likelihood [@chambers1992S].
-It is often convenient to compute the maximum likelihood estimates using the iteratively reweighted least squares (IRWLS) algorithm [@wood2017generalized], which is the approach used by the  \code{glm()} function in the \proglang{R} programming language [@rcore2024].
+It is often convenient to compute the maximum likelihood estimates using the iteratively reweighted least squares (IRWLS) algorithm [@wood2017generalized], which is an approach used by the  \code{glm()} function in the \proglang{R} programming language [@rcore2024].
 GLMs add an additional layer of complexity compared to linear regression models, as the left-hand side of Equation$~$\ref{eq-glm} is a function of the mean of $\mathbf{y}$ rather than $\mathbf{y}$ itself (as in linear regression models).
@@
 This independence assumption is typically impractical for spatial data.
-For spatial data, nearby observations tend to be more similar than distant observations [@tobler1970computer], which leads to positive spatial covariance.
+For spatial data, nearby observations tend to be more similar than distant observations [@tobler1970computer], which leads to positive spatial covariance among observations.
 The consequences of ignoring spatial covariance in statistical models for spatial data can be severe and include imprecise parameter estimates as well as misleading standard errors that inflate Type-I error rates and decrease power [@zimmerman2024spatial].
 
-An approach for handling spatial data using a GLM is to assume the elements of $\mathbf{w}$ exhibit covariance that varies spatially and nonspatially.
+An approach for handling spatial data using a GLM is to assume the elements of $\mathbf{w}$ exhibit both spatial and nonspatial variability.
 This is achieved by adding to Equation$~$\ref{eq-glm} two random effects, $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$.
@@
 Recently, however, there have been numerous, significant advances in \proglang{R} software that have made these models more accessible to practitioners.
-The \pkg{brms} [@burkner2017brms], \pkg{carBayes} [@lee2013carbayes], \pkg{ngspatial} [@hughes2020ngspatial], \pkg{R-INLA} [@lindgren2015bayesian] and \pkg{inlabru} [@bachl2019inlabru], \pkg{spBayes} [@finley2007spbayes], \pkg{spOccupancy} [@doser2022spoccupancy], \pkg{spAbundance} [@doser2024spabundance], and \pkg{spNNGP} [@finley2002spnngp] packages take a Bayesian approach, either directly sampling from posterior distributions of parameters (e.g., using MCMC) or approximating them.
+The \pkg{brms} [@burkner2017brms], \pkg{carBayes} [@lee2013carbayes], \pkg{ngspatial} [@hughes2020ngspatial], \pkg{R-INLA} [@lindgren2015bayesian], \pkg{inlabru} [@bachl2019inlabru], \pkg{spBayes} [@finley2007spbayes], \pkg{spOccupancy} [@doser2022spoccupancy], \pkg{spAbundance} [@doser2024spabundance], and \pkg{spNNGP} [@finley2002spnngp] packages take a Bayesian approach, either directly sampling from posterior distributions of parameters (e.g., using MCMC) or approximating them.
 A benefit of Bayesian approaches is that prior information can be incorporated and uncertainty quantification of parameter estimates is straightforward.
 However, Bayesian approaches, especially those using MCMC, can be computationally expensive.
-In order to reduce computation time, many of these packages work with the precision matrix instead of the covariance matrix so that computationally expensive matrix inversion is not required.
-For example, \pkg{R-INLA} uses the precision matrix and tends to be very fast.
-Working with precision matrices, however, can be more restrictive and less intuitive than working directly with the covariance matrix.
+In order to reduce computation time, many of these packages (e.g., \pkg{R-INLA}) work with the precision matrix instead of the covariance matrix so that computationally expensive matrix inversion is not required.
+Working with precision matrices, however, can be more restrictive and less intuitive than working with covariance matrices.
 The \pkg{FRK} [@sainsbury2024modeling], \pkg{glmmTMB} [@brooks2017glmmtmb], \pkg{hglm} [@ronnegard2010hglm], \pkg{mgcv} [@wood2017generalized], and \pkg{spaMM} [@rousset2014spamm] packages directly use Laplace, quasi-likelihood, or reduced-rank approaches to estimate parameters.
@@
 
-Building from @evangelou2011estimation and @bonat2016practical, @ver2024marginal proposed a novel approach for fitting SPGLMs that leverages the Laplace approximation while marginalizing over both the latent $\mathbf{w}$ and the fixed effects ($\boldsymbol{\beta}$) and accommodating general covariance structures, including spatial ones.
-This approach performed efficiently in a variety of simulation settings, generally having appropriate confidence interval coverage for the fixed effects and prediction interval coverage for $\mathbf{w}$ at new locations.
-The approach performed similarly to the Bayesian SPGLM approach in \pkg{spBayes} and the automatic differentiation SPGLM approach in \pkg{glmmTMB} but was much faster.
+Building from @evangelou2011estimation and @bonat2016practical, @ver2024marginal proposed a novel approach for fitting SPGLMs that leverages the Laplace approximation while marginalizing over both the latent $\mathbf{w}$ and the fixed effects ($\boldsymbol{\beta}$).
+This approach performed well in a variety of simulation settings, generally having appropriate confidence interval coverage for the fixed effects and prediction interval coverage for $\mathbf{w}$ at new locations.
+It also performed similarly to the Bayesian SPGLM approach in \pkg{spBayes} and the automatic differentiation SPGLM approach in \pkg{glmmTMB} but was much faster.
 At small sample sizes, the approach outperformed the approximate Bayesian SPGLM approach in \pkg{R-INLA} and had similar computational times.
@@
 
-The \pkg{spmodel} \proglang{R} package [@dumelle2023spmodel] recently made public a full set of modeling tools for SPGLMs fit using the novel Laplace approach described by @ver2024marginal.
+The \pkg{spmodel} \proglang{R} package [@dumelle2023spmodel] recently released a full set of modeling tools for SPGLMs fit using the novel Laplace approach described by @ver2024marginal.
 These modeling tools are approachable and mirror the familiar \code{glm()} syntax from base-\proglang{R}, making the transition from GLMs to SPGLMs relatively seamless.
-The \code{spglm()} function fits SPGLMs for point-referenced data (e.g., x- and y-coordinates representing point locations in a field; these models are sometimes called "geostatistical" models), while the \code{spgautor()} function fits SPGLMs for areal data (e.g., polygon boundaries representing geographic subsets of a region; these models are sometimes called "autoregressive" models).
-For both point-referenced data and areal supports, \pkg{spmodel} supports the binomial distribution for binary data, Poisson and negative binomial distributions for count data, Gamma and inverse Gaussian distributions for skewed data, and the beta distribution for proportion data.
-There are 20 different spatial covariance structures available including the exponential, Gaussian, and spherical for point-referenced data (Figure$~$\ref{fig-type}) and the conditional autoregressive, and simultaneous autoregressive structures for areal data.
-\pkg{spmodel} provides tools for commonly used model summaries, visualizations, and diagnostics (e.g., Cook's distance) using standard \proglang{R} helper functions like \code{summary()}, \code{plot()}, \code{fitted()}, and \code{tidy()}, among others.
+The \code{spglm()} function fits SPGLMs for point-referenced support (e.g., x- and y-coordinates representing point locations in a field; these models are sometimes called "geostatistical" models), while the \code{spgautor()} function fits SPGLMs for areal support (e.g., polygon boundaries representing geographic subsets of a region; these models are sometimes called "autoregressive" models).
+For both point-referenced and areal supports, \pkg{spmodel} supports the binomial distribution for binary data, Poisson and negative binomial distributions for count data, Gamma and inverse Gaussian distributions for skewed data, and the beta distribution for proportion data.
+There are 20 different spatial covariance structures available including the exponential, Gaussian, and spherical for point-referenced support (Figure$~$\ref{fig-type}) and the conditional autoregressive, and simultaneous autoregressive structures for areal support \pkg{spmodel} provides tools for commonly used model summaries, visualizations, and diagnostics (e.g., fitted values) using standard \proglang{R} helper functions like \code{summary()}, \code{plot()}, and \code{fitted()}, among others.
 \pkg{spmodel} also provides tools to predict $\mathbf{w}$ at new locations and quantify uncertainty in those prediction using \code{predict()} and \code{augment()}.
@@
 
-Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version 0.11.0) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in terms of scope and feel.
+Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version 0.11.1) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in terms of scope and feel.
 Both packages use similar syntax as \code{glm()}, accommodate flexible \code{formula} arguments (e.g., offsets, splines), handle spatial covariance that decays at different rates in different directions (i.e., geometric anisotropy), incorporate nonspatial random effects, support other \proglang{R} packages for modeling like \pkg{broom} [@robinson2021broom; @kuhn2022tidy], \pkg{emmeans} [@lenth2024emmeans], and \pkg{car} [@fox2019car], and have tools for model summaries, prediction, and simulating data.
@@
 In Section$~$\ref{sec-applications2}, we model moose counts in Alaska, US; skewed lake conductivity in the Southwestern US; harbor seal abundance trend behavior in Alaska, US; and voter turnout rates in Texas, US.
-And in Section$~$\ref{sec-discussion}, we end with a discussion synthesizing \pkg{spmodel}'s contributions to the analysis of SPGLMs in \proglang{R}.
+And in Section$~$\ref{sec-discussion}, we end with a discussion synthesizing \pkg{spmodel}'s contributions to SPGLMs in \proglang{R}.
 
@@
 Next, we describe a brief overview of the approach and how it can be used for several primary data analysis tasks [@tredennick2021practical] like model comparison, parameter estimation, inference, model diagnostics, and prediction.
-Then in Section$~$\ref{sec-applications} and Section$~$\ref{sec-applications2}, we show how to use \pkg{spmodel} to carry out these primary data analysis tasks with various data sets.
 
@@
 \begin{align}\label{eq-marginal6}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{w}^{*}), \varphi] [\mathbf{w}^{*} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{w}^{*}}|^{-1/2}.
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{w}^{*}), \varphi] [\mathbf{w}^{*} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{w}^{*}}|^{-1/2},
 \end{align}
+which can be directly evaluated.
+This result suggests a doubly iterative optimization over 1) $\boldsymbol{\theta}$ and $\varphi$ and 2) the latent $\mathbf{w}$ (to find each set of $\mathbf{w}^{*}$), which ultimately yields the marginal restricted maximum
+likelihood estimators $\hat{\varphi}$ and $\hat{\boldsymbol{\theta}}$ and their respective values of $\mathbf{w}^{*}$, which we call $\hat{\mathbf{w}}$.
+@ver2024marginal provide further details, which includes explicit forms of $\mathbf{g}$ and $\mathbf{G}$ for various response distributions.
 
-Maximizing the natural logarithm of Equation$~$\ref{eq-marginal6} requires a doubly iterative process over 1) $\boldsymbol{\theta}$ and $\varphi$ and 2) the latent $\mathbf{w}$ (to find each set of $\mathbf{w}^{*}$).
-This doubly iterative process eventually yields the marginal restricted maximum likelihood estimators $\hat{\varphi}$ and $\hat{\boldsymbol{\theta}}$ and their respective values of $\mathbf{w}^{*}$, which we call $\hat{\mathbf{w}}$.
-Maximizing the marginal restricted likelihood requires repeatedly evaluating $\boldsymbol{\Sigma}^{-1}$, $\mathbf{g}$, and $\mathbf{G}$; see @ver2024marginal for more details and comparisons to other approaches as well as forms of $\mathbf{g}$ and $\mathbf{G}$ for various response distributions.
+## Estimating fixed effects {#sec:fixed}
 
-## Estimating fixed effects
-
-Though the fixed effects are integrated out of the likelihood, we can still estimate them using generalized least squares (GLS) principles, a common practice for linear models estimated using restricted maximum likelihood methods.
+We can estimate the fixed effects using generalized least squares (GLS) principles, a common practice for linear models estimated using restricted maximum likelihood methods.
 Had we observed $\mathbf{w}$, a GLS estimator for $\boldsymbol{\beta}$ is given by
@@
 where $\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}$.
-However, we do not observe $\mathbf{w}$ and instead estimate $\mathbf{w}$ via $\hat{\mathbf{w}}$.
-Thus, it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$.
-To derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$.
-To do so, we must condition on $\mathbf{w}$ as if it were observed and invoke properties of the laws of total expectation and variance.
-Because $\hat{\mathbf{w}}$ was obtained by maximizing the likelihood, we may assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance approximately equal to $-\mathbf{H}^{-1}$ (the inverse Hessian).
-It follows that $\text{E}(\hat{\mathbf{w}})$ is given by
+While we do not $\mathbf{w}$, we do estimate it via $\hat{\mathbf{w}}$, and thus it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$.
+To derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$ by conditioning on $\mathbf{w}$ as if it were observed and leveraging the laws of total expectation and variance.
+Based on asymptotic properties of (restricted) maximum likelihood estimators [@cressie1993asymptotic], we may assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance equal to $-\mathbf{H}^{-1}$, the negative inverse Hessian (i.e., the inverse observed information matrix).
+Thus it follows that $\text{E}(\hat{\mathbf{w}})$ is given by
 \begin{align}
@@
 
-These results are important because they justify closed-form solutions for $\hat{\boldsymbol{\beta}}$ and its associated variance.
-Closed-form solutions are useful because they bypass the need for sampling-based strategies to evaluate the mean and variance of $\hat{\boldsymbol{\beta}}$, a common technique for other approaches to SPGLMs like Bayesian MCMC.
+These results are important because they justify analytic (i.e., closed-form) solutions for $\hat{\boldsymbol{\beta}}$ and its associated variance.
+Analytic solutions are useful because they bypass the need for sampling-based strategies to evaluate the mean and variance of $\hat{\boldsymbol{\beta}}$, a common technique for other approaches to SPGLMs like Bayesian MCMC that can be computationally intensive.
 
@@
 The sum of the squared deviance residuals equals the sum of the elements of $\mathbf{d}$, known as the deviance of the model fit.
-The deviance of the model fit quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized].
+The deviance of the model fit quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\text{y}_i = f^{-1}(\hat{\text{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized].
 Deviance is often used as a fit statistic; lower values of deviance imply a better model fit (compared to the observed data).
-Pearson and standardized residuals are other types of GLM residuals that involve a scaling of the response residuals; the Pearson residuals scale $\mathbf{r}_{r}$ by the square root of $\mathbf{V}$, while the standardized residuals scale the deviance residuals by $\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}$, where $\mathbf{L}_{ii}$ is the $i$th diagonal element of the leverage matrix, which we discuss next.
+Pearson and standardized residuals are other types of GLM residuals that involve a scaling of the response residuals.
+The Pearson residuals scale $\mathbf{r}_{r}$ by the square root of $\mathbf{V}$, a diagonal matrix with $i$th diagonal element equal to the variance of the response distribution evaluated at $f^{-1}(\hat{\text{w}}_i)$ [@faraway2016extending]; $\mathbf{V}$ is sometimes called the GLM weight matrix.
+The standardized residuals scale the deviance residuals by $\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}$, where $\mathbf{L}_{ii}$ is the $i$th diagonal element of the leverage matrix, which we discuss next.
 
@@
 \begin{align}
-  \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2},
+  \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2}.
 \end{align}
-where $\mathbf{V}$ is a diagonal matrix with $i$th diagonal element equal to the variance of the response distribution evaluated at $f^{-1}(\mathbf{w}_i)$ [@faraway2016extending]; $\mathbf{V}$ is sometimes called the GLM weight matrix.
 The larger the value of $\mathbf{L}_{ii}$, the more severe the leverage from the $i$th observation.
@@
 where $\mathbf{r}^2_{s}$ are the standardized residuals and $diag(\mathbf{L})$ indicates the diagonal elements of the leverage matrix.
-The larger the value of $\mathbf{c}_i$, the more severe the influence from the $i$th observation.
+The larger the value of $\text{c}_i$, the more severe the influence from the $i$th observation.
 @montgomery2021introduction provide guidance for interpreting these types of statistics, including cutoffs to consider when identifying extreme residual, leverage, or influence values.
@@
 Like $R^2$, $PR^2$ can be adjusted to account for the numbers of parameters estimated in a model.
-Like $R^2$, $PR^2$ can be adjusted to account for the numbers of parameters estimated in a model.
 Because the $deviance_{total}$ denominator changes across fitted models (as the values of $\hat{\boldsymbol{\theta}}$ and $\hat{\varphi}$ change), this statistic should not be used as a model comparison tool.
@@
 We may also predict values of the latent mean (on the link scale) at new locations by leveraging the spatial covariance between observed locations and new locations (spatial prediction is also called Kriging; see @cressie1990origins).
-Again suppose that we observed $\mathbf{w}$ and we want to make predictions at $\mathbf{u}$, a vector of latent means at the new locations that follows the same SPGLM from Equation$~$\ref{eq-spglm} and has design matrix, $\mathbf{X}_{\mathbf{u}}$.
+Like in Section$~$\ref{sec:fixed}, suppose first that we observed $\mathbf{w}$ and we want to make predictions at $\mathbf{u}$, a vector of latent means at the new locations that follows the same SPGLM from Equation$~$\ref{eq-spglm} and has design matrix, $\mathbf{X}_{\mathbf{u}}$.
 The vector $(\mathbf{w}, \mathbf{u})^\top$ has expectation $(\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top$ and covariance matrix $\begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}$, where $\boldsymbol{\Sigma} = \text{Var}(\mathbf{w}, \mathbf{w})$, $\boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}} = \text{Var}(\mathbf{w}, \mathbf{u})$, $\boldsymbol{\Sigma}_{\mathbf{u} \mathbf{w}} = \boldsymbol{\Sigma}_{\mathbf{w} \mathbf{u}}^\top$ and $\boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} = \text{Var}(\mathbf{u}, \mathbf{u})$.
-By assumption, we have observed $\mathbf{w}$, so we may derive the conditional distribution of $\mathbf{u}|\mathbf{w}$, which has the following properties:
+Thus we may derive the conditional distribution of $\mathbf{u}|\mathbf{w}$, which has the following properties:
 \begin{align}
@@
 \end{align}
-@ver2024marginal show how these equations are adjusted to reflect uncertainty in both $\hat{\boldsymbol{\beta}}$ and $\hat{\mathbf{w}}$ while leveraging the laws of total expectation and variance yet again.
-They derive the predictor of $\mathbf{u}$, $\hat{\mathbf{u}}$, and its associated variance, given by:
+Recall, however, that we do not actually observe $\mathbf{w}$ and instead compute $\hat{\mathbf{w}}$; so, to predict $\mathbf{u}$ and quantify its uncertainty, we must again leverage the laws of total expectation and variance.
+@ver2024marginal show that $\hat{\mathbf{u}}$ and its associated variance are given by:
 \begin{align}
@@
 where $\mathbf{K} = \mathbf{X}_{\mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \mathbf{X}$ and $\boldsymbol{\Lambda} = \mathbf{X}_{\mathbf{u}}\mathbf{B} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{1} - \mathbf{X}\mathbf{B})$ for a vector of ones, $\mathbf{1}$.
-As with $\hat{\boldsymbol{\beta}}$, in practice these covariance matrices are evaluated at $\hat{\boldsymbol{\theta}}$.
+As with $\hat{\boldsymbol{\beta}}$, these covariance matrices are evaluated at $\hat{\boldsymbol{\theta}}$ in practice.
 
@@
 The results in this manuscript were obtained using [R]{.proglang} 4.4.0 with the
-\pkg{spmodel} 0.11.0 package. Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.
+\pkg{spmodel} 0.11.1 package. Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.
 
@@
 All data used are part of the \pkg{spmodel} [R]{.proglang} package available for download from CRAN at [https://CRAN.R-project.org/package=spmodel](https://CRAN.R-project.org/package=spmodel).
-Results were obtained using [R]{.proglang} 4.4.0 with the \pkg{spmodel} 0.11.0 package.
+Results were obtained using [R]{.proglang} 4.4.0 with the \pkg{spmodel} 0.11.1 package.
 Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.

### 2025-11-26 inst/manuscript/manuscript.tex
some updates

@@
 Non-Gaussian data are common in practice and include binary, count,
-skewed, and proprtion data types. Often, non-Gaussian data are modeled
+skewed, and proportion data types. Often, non-Gaussian data are modeled
 using a generalized linear model (GLM). GLMs typically assume that
@@
 (SPGLMs) to non-Gaussian data and making spatial predictions (i.e.,
-Kriging). SPGLMs for point-referenced data (x- and y-coordinates) are
+Kriging). SPGLMs for point-referenced (x- and y-coordinates) support are
 fit using the \code{spglm()} function, while SPGLMs for areal (lattice,
-polygon) data are fit using the \code{spgautor()} function. Both
+polygon) support are fit using the \code{spgautor()} function. Both
 \code{spglm()} and \code{spgautor()} maximize a novel Laplace likelihood
 which marginalizes over the model's fixed effects and latent mean while
-formally incorporating spatial covariance. The inputs and outputs of
-\code{spglm()} and \code{spgautor()} closely resemble the \code{glm()}
-function from base \proglang{R}, easing the transition from GLMs to
-SPGLMs. \pkg{spmodel} provides and builds upon several commonly used
-helper functions for model building like \code{summary()},
-\code{plot()}, \code{fitted()}, and \code{tidy()}, among others. Spatial
-predictions of the latent mean at unobserved locations are obtained
-using \code{predict()} or \code{augment()}. \pkg{spmodel} accommodates
-myriad advanced modeling features like geometric anisotrpoy, nonspatial
-random effects, analysis of variance, and more. Throughout, we use
+formally incorporating spatial covariance among observations. The inputs
+and outputs of \code{spglm()} and \code{spgautor()} closely resemble the
+\code{glm()} function from base \proglang{R}, easing the transition from
+GLMs to SPGLMs. \pkg{spmodel} provides and builds upon several commonly
+used helper functions for model building like \code{summary()},
+\code{plot()}, and \code{fitted()}, among others. Spatial predictions of
+the latent mean at unobserved locations are obtained using
+\code{predict()} or \code{augment()}. \pkg{spmodel} accommodates myriad
+advanced modeling features like geometric anisotrpoy, nonspatial random
+effects, analysis of variance, and more. Throughout, we use
 \pkg{spmodel} to fit SPGLMs to moose presence and counts in Alaska,
@@
 
-\Keywords{autoregressive model, geostatistical model, inverse link
-function, Poisson regression, link function, logistic
-regression, spatial covariance, spatial correlation}
-\Plainkeywords{autoregressive model, geostatistical model, inverse link
-function, Poisson regression, link function, logistic
-regression, spatial covariance, spatial correlation}
+\Keywords{autoregressive model, geostatistical model, Poisson
+regression, link function, logistic regression, overdispersion, spatial
+covariance, spatial dependence}
+\Plainkeywords{autoregressive model, geostatistical model, Poisson
+regression, link function, logistic regression, overdispersion, spatial
+covariance, spatial dependence}
 
@@
 \(n \times 1\) response variable \(\mathbf{y}\) belongs to a statistical
-distribution (e.g., Binomial, Poisson) with some mean and variance.
+distribution (e.g., binomial, Poisson) with some mean and variance.
 Often, the analysis goal is to study the impact of a linear function of
@@
 reweighted least squares (IRWLS) algorithm \citep{wood2017generalized},
-which is the approach used by the \code{glm()} function in the
+which is an approach used by the \code{glm()} function in the
 \proglang{R} programming language \citep{rcore2024}. GLMs add an
@@
 distant observations \citep{tobler1970computer}, which leads to positive
-spatial covariance. The consequences of ignoring spatial covariance in
-statistical models for spatial data can be severe and include imprecise
-parameter estimates as well as misleading standard errors that inflate
-Type-I error rates and decrease power \citep{zimmerman2024spatial}.
+spatial covariance among observations. The consequences of ignoring
+spatial covariance in statistical models for spatial data can be severe
+and include imprecise parameter estimates as well as misleading standard
+errors that inflate Type-I error rates and decrease power
+\citep{zimmerman2024spatial}.
 
 An approach for handling spatial data using a GLM is to assume the
-elements of \(\mathbf{w}\) exhibit covariance that varies spatially and
-nonspatially. This is achieved by adding to Equation\(~\)\ref{eq-glm}
-two random effects, \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\).
-The random effect \(\boldsymbol{\tau}\) is an \(n \times 1\) column
-vector of spatially dependent random errors. We assume that
+elements of \(\mathbf{w}\) exhibit both spatial and nonspatial
+variability. This is achieved by adding to Equation\(~\)\ref{eq-glm} two
+random effects, \(\boldsymbol{\tau}\) and \(\boldsymbol{\epsilon}\). The
+random effect \(\boldsymbol{\tau}\) is an \(n \times 1\) column vector
+of spatially dependent random errors. We assume that
 \(\text{E}(\boldsymbol{\tau}) = \boldsymbol{0}\) and
@@
 \pkg{ngspatial} \citep{hughes2020ngspatial}, \pkg{R-INLA}
-\citep{lindgren2015bayesian} and \pkg{inlabru} \citep{bachl2019inlabru},
+\citep{lindgren2015bayesian}, \pkg{inlabru} \citep{bachl2019inlabru},
 \pkg{spBayes} \citep{finley2007spbayes}, \pkg{spOccupancy}
@@
 especially those using MCMC, can be computationally expensive. In order
-to reduce computation time, many of these packages work with the
-precision matrix instead of the covariance matrix so that
-computationally expensive matrix inversion is not required. For example,
-\pkg{R-INLA} uses the precision matrix and tends to be very fast.
-Working with precision matrices, however, can be more restrictive and
-less intuitive than working directly with the covariance matrix. The
-\pkg{FRK} \citep{sainsbury2024modeling}, \pkg{glmmTMB}
-\citep{brooks2017glmmtmb}, \pkg{hglm} \citep{ronnegard2010hglm},
-\pkg{mgcv} \citep{wood2017generalized}, and \pkg{spaMM}
-\citep{rousset2014spamm} packages directly use Laplace,
-quasi-likelihood, or reduced-rank approaches to estimate parameters.
-These direct approaches tend to be computationally efficient, as they
-don't rely on MCMC sampling. In contrast to the Bayesian approach, a
-drawback of these direct approaches is that prior information cannot be
-formally incorporated and covariance parameter uncertainty is more
-challenging to quantify. The \pkg{sdmTMB} \citep{anderson2024sdmtmb}
-package combines elements of \pkg{R-INLA}, \pkg{glmmTMB}, and Gaussian
-Markov random fields to fit a wide variety of SPGLMs, while
-\pkg{tinyVAST} \citep{thorson2025tinyVAST} extends some of these models
-to multivariate or (dynamic) structural equation models.
+to reduce computation time, many of these packages (e.g., \pkg{R-INLA})
+work with the precision matrix instead of the covariance matrix so that
+computationally expensive matrix inversion is not required. Working with
+precision matrices, however, can be more restrictive and less intuitive
+than working with covariance matrices. The \pkg{FRK}
+\citep{sainsbury2024modeling}, \pkg{glmmTMB} \citep{brooks2017glmmtmb},
+\pkg{hglm} \citep{ronnegard2010hglm}, \pkg{mgcv}
+\citep{wood2017generalized}, and \pkg{spaMM} \citep{rousset2014spamm}
+packages directly use Laplace, quasi-likelihood, or reduced-rank
+approaches to estimate parameters. These direct approaches tend to be
+computationally efficient, as they don't rely on MCMC sampling. In
+contrast to the Bayesian approach, a drawback of these direct approaches
+is that prior information cannot be formally incorporated and covariance
+parameter uncertainty is more challenging to quantify. The \pkg{sdmTMB}
+\citep{anderson2024sdmtmb} package combines elements of \pkg{R-INLA},
+\pkg{glmmTMB}, and Gaussian Markov random fields to fit a wide variety
+of SPGLMs, while \pkg{tinyVAST} \citep{thorson2025tinyVAST} extends some
+of these models to multivariate or (dynamic) structural equation models.
 
@@
 while marginalizing over both the latent \(\mathbf{w}\) and the fixed
-effects (\(\boldsymbol{\beta}\)) and accommodating general covariance
-structures, including spatial ones. This approach performed efficiently
-in a variety of simulation settings, generally having appropriate
-confidence interval coverage for the fixed effects and prediction
-interval coverage for \(\mathbf{w}\) at new locations. The approach
-performed similarly to the Bayesian SPGLM approach in \pkg{spBayes} and
-the automatic differentiation SPGLM approach in \pkg{glmmTMB} but was
-much faster. At small sample sizes, the approach outperformed the
-approximate Bayesian SPGLM approach in \pkg{R-INLA} and had similar
-computational times. For moderate sample sizes, it performed similarly
-to \pkg{R-INLA}, though \pkg{R-INLA} was faster. The novel Laplace
-approach is particularly attractive for two reasons. First, it is
-general enough that it can be applied to any covariance structure (not
-just spatial). Second, after estimating the covariance parameters,
-analytical solutions exist for the fixed effects (and their standard
-errors) as well as predictions of the latent \(\mathbf{w}\) at new
-locations (and their standard errors).
+effects (\(\boldsymbol{\beta}\)). This approach performed well in a
+variety of simulation settings, generally having appropriate confidence
+interval coverage for the fixed effects and prediction interval coverage
+for \(\mathbf{w}\) at new locations. It also performed similarly to the
+Bayesian SPGLM approach in \pkg{spBayes} and the automatic
+differentiation SPGLM approach in \pkg{glmmTMB} but was much faster. At
+small sample sizes, the approach outperformed the approximate Bayesian
+SPGLM approach in \pkg{R-INLA} and had similar computational times. For
+moderate sample sizes, it performed similarly to \pkg{R-INLA}, though
+\pkg{R-INLA} was faster. The novel Laplace approach is particularly
+attractive for two reasons. First, it is general enough that it can be
+applied to any covariance structure (not just spatial). Second, after
+estimating the covariance parameters, analytical solutions exist for the
+fixed effects (and their standard errors) as well as predictions of the
+latent \(\mathbf{w}\) at new locations (and their standard errors).
 
 The \pkg{spmodel} \proglang{R} package \citep{dumelle2023spmodel}
-recently made public a full set of modeling tools for SPGLMs fit using
-the novel Laplace approach described by \citet{ver2024marginal}. These
+recently released a full set of modeling tools for SPGLMs fit using the
+novel Laplace approach described by \citet{ver2024marginal}. These
 modeling tools are approachable and mirror the familiar \code{glm()}
@@
 relatively seamless. The \code{spglm()} function fits SPGLMs for
-point-referenced data (e.g., x- and y-coordinates representing point
+point-referenced support (e.g., x- and y-coordinates representing point
 locations in a field; these models are sometimes called
 ``geostatistical'' models), while the \code{spgautor()} function fits
-SPGLMs for areal data (e.g., polygon boundaries representing geographic
-subsets of a region; these models are sometimes called
-``autoregressive'' models). For both point-referenced data and areal
+SPGLMs for areal support (e.g., polygon boundaries representing
+geographic subsets of a region; these models are sometimes called
+``autoregressive'' models). For both point-referenced and areal
 supports, \pkg{spmodel} supports the binomial distribution for binary
@@
 covariance structures available including the exponential, Gaussian, and
-spherical for point-referenced data (Figure\(~\)\ref{fig-type}) and the
-conditional autoregressive, and simultaneous autoregressive structures
-for areal data. \pkg{spmodel} provides tools for commonly used model
-summaries, visualizations, and diagnostics (e.g., Cook's distance) using
-standard \proglang{R} helper functions like \code{summary()},
-\code{plot()}, \code{fitted()}, and \code{tidy()}, among others.
+spherical for point-referenced support (Figure\(~\)\ref{fig-type}) and
+the conditional autoregressive, and simultaneous autoregressive
+structures for areal support \pkg{spmodel} provides tools for commonly
+used model summaries, visualizations, and diagnostics (e.g., fitted
+values) using standard \proglang{R} helper functions like
+\code{summary()}, \code{plot()}, and \code{fitted()}, among others.
 \pkg{spmodel} also provides tools to predict \(\mathbf{w}\) at new
@@
 Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version
-0.11.0) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in
+0.11.1) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in
 terms of scope and feel. Both packages use similar syntax as
@@
 turnout rates in Texas, US. And in Section\(~\)\ref{sec-discussion}, we
-end with a discussion synthesizing \pkg{spmodel}'s contributions to the
-analysis of SPGLMs in \proglang{R}.
+end with a discussion synthesizing \pkg{spmodel}'s contributions to
+SPGLMs in \proglang{R}.
 
@@
 tasks \citep{tredennick2021practical} like model comparison, parameter
-estimation, inference, model diagnostics, and prediction. Then in
-Section\(~\)\ref{sec-applications} and
-Section\(~\)\ref{sec-applications2}, we show how to use \pkg{spmodel} to
-carry out these primary data analysis tasks with various data sets.
+estimation, inference, model diagnostics, and prediction.
 
@@
 \(\exp(\ell_{\mathbf{w}^{*}})\) yields \begin{align}\label{eq-marginal6}
-  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{w}^{*}), \varphi] [\mathbf{w}^{*} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{w}^{*}}|^{-1/2}.
-\end{align}
-
-Maximizing the natural logarithm of Equation\(~\)\ref{eq-marginal6}
-requires a doubly iterative process over 1) \(\boldsymbol{\theta}\) and
+  [\mathbf{y}|\mathbf{X}, \varphi, \boldsymbol{\theta}] \approx [\mathbf{y} | f^{-1}(\mathbf{w}^{*}), \varphi] [\mathbf{w}^{*} | \mathbf{X}, \boldsymbol{\theta}] (2 \pi)^{n/2}|-\mathbf{G}_{\mathbf{w}^{*}}|^{-1/2},
+\end{align} which can be directly evaluated. This result suggests a
+doubly iterative optimization over 1) \(\boldsymbol{\theta}\) and
 \(\varphi\) and 2) the latent \(\mathbf{w}\) (to find each set of
-\(\mathbf{w}^{*}\)). This doubly iterative process eventually yields the
-marginal restricted maximum likelihood estimators \(\hat{\varphi}\) and
+\(\mathbf{w}^{*}\)), which ultimately yields the marginal restricted
+maximum likelihood estimators \(\hat{\varphi}\) and
 \(\hat{\boldsymbol{\theta}}\) and their respective values of
-\(\mathbf{w}^{*}\), which we call \(\hat{\mathbf{w}}\). Maximizing the
-marginal restricted likelihood requires repeatedly evaluating
-\(\boldsymbol{\Sigma}^{-1}\), \(\mathbf{g}\), and \(\mathbf{G}\); see
-\citet{ver2024marginal} for more details and comparisons to other
-approaches as well as forms of \(\mathbf{g}\) and \(\mathbf{G}\) for
-various response distributions.
-
-\subsection{Estimating fixed effects}\label{estimating-fixed-effects}
-
-Though the fixed effects are integrated out of the likelihood, we can
-still estimate them using generalized least squares (GLS) principles, a
-common practice for linear models estimated using restricted maximum
-likelihood methods. Had we observed \(\mathbf{w}\), a GLS estimator for
-\(\boldsymbol{\beta}\) is given by \begin{align}\label{eq-gls1}
+\(\mathbf{w}^{*}\), which we call \(\hat{\mathbf{w}}\).
+\citet{ver2024marginal} provide further details, which includes explicit
+forms of \(\mathbf{g}\) and \(\mathbf{G}\) for various response
+distributions.
+
+\subsection{Estimating fixed effects}\label{sec:fixed}
+
+We can estimate the fixed effects using generalized least squares (GLS)
+principles, a common practice for linear models estimated using
+restricted maximum likelihood methods. Had we observed \(\mathbf{w}\), a
+GLS estimator for \(\boldsymbol{\beta}\) is given by
+\begin{align}\label{eq-gls1}
   \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
@@
 \(\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\).
-However, we do not observe \(\mathbf{w}\) and instead estimate
-\(\mathbf{w}\) via \(\hat{\mathbf{w}}\). Thus, it is reasonable to
-define \(\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}\). To
-derive properties of \(\hat{\boldsymbol{\beta}}\) like expectation and
-variance, we must derive these properties for \(\hat{\mathbf{w}}\). To
-do so, we must condition on \(\mathbf{w}\) as if it were observed and
-invoke properties of the laws of total expectation and variance. Because
-\(\hat{\mathbf{w}}\) was obtained by maximizing the likelihood, we may
-assume that given \(\mathbf{w}\), \(\hat{\mathbf{w}}\) has mean
-\(\mathbf{w}\) and variance approximately equal to \(-\mathbf{H}^{-1}\)
-(the inverse Hessian). It follows that \(\text{E}(\hat{\mathbf{w}})\) is
-given by \begin{align}
+While we do not \(\mathbf{w}\), we do estimate it via
+\(\hat{\mathbf{w}}\), and thus it is reasonable to define
+\(\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}\). To derive
+properties of \(\hat{\boldsymbol{\beta}}\) like expectation and
+variance, we must derive these properties for \(\hat{\mathbf{w}}\) by
+conditioning on \(\mathbf{w}\) as if it were observed and leveraging the
+laws of total expectation and variance. Based on asymptotic properties
+of (restricted) maximum likelihood estimators
+\citep{cressie1993asymptotic}, we may assume that given \(\mathbf{w}\),
+\(\hat{\mathbf{w}}\) has mean \(\mathbf{w}\) and variance equal to
+\(-\mathbf{H}^{-1}\), the negative inverse Hessian (i.e., the inverse
+observed information matrix). Thus it follows that
+\(\text{E}(\hat{\mathbf{w}})\) is given by \begin{align}
   \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
@@
 
-These results are important because they justify closed-form solutions
-for \(\hat{\boldsymbol{\beta}}\) and its associated variance.
-Closed-form solutions are useful because they bypass the need for
-sampling-based strategies to evaluate the mean and variance of
-\(\hat{\boldsymbol{\beta}}\), a common technique for other approaches to
-SPGLMs like Bayesian MCMC.
+These results are important because they justify analytic (i.e.,
+closed-form) solutions for \(\hat{\boldsymbol{\beta}}\) and its
+associated variance. Analytic solutions are useful because they bypass
+the need for sampling-based strategies to evaluate the mean and variance
+of \(\hat{\boldsymbol{\beta}}\), a common technique for other approaches
+to SPGLMs like Bayesian MCMC that can be computationally intensive.
 
@@
 between the a saturated model that fits every observation perfectly
-(i.e., \(\mathbf{y} = f^{-1}(\hat{\mathbf{w}}_i)\) for all \(i\)) and
-the fitted model \citep{myers2012generalized}. Deviance is often used as
-a fit statistic; lower values of deviance imply a better model fit
+(i.e., \(\text{y}_i = f^{-1}(\hat{\text{w}}_i)\) for all \(i\)) and the
+fitted model \citep{myers2012generalized}. Deviance is often used as a
+fit statistic; lower values of deviance imply a better model fit
 (compared to the observed data). Pearson and standardized residuals are
 other types of GLM residuals that involve a scaling of the response
-residuals; the Pearson residuals scale \(\mathbf{r}_{r}\) by the square
-root of \(\mathbf{V}\), while the standardized residuals scale the
-deviance residuals by \(\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}\), where
-\(\mathbf{L}_{ii}\) is the \(i\)th diagonal element of the leverage
-matrix, which we discuss next.
+residuals. The Pearson residuals scale \(\mathbf{r}_{r}\) by the square
+root of \(\mathbf{V}\), a diagonal matrix with \(i\)th diagonal element
+equal to the variance of the response distribution evaluated at
+\(f^{-1}(\hat{\text{w}}_i)\) \citep{faraway2016extending};
+\(\mathbf{V}\) is sometimes called the GLM weight matrix. The
+standardized residuals scale the deviance residuals by
+\(\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}\), where \(\mathbf{L}_{ii}\) is
+the \(i\)th diagonal element of the leverage matrix, which we discuss
+next.
 
@@
 In a GLM, the leverage matrix is given by \begin{align}
-  \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2},
-\end{align} where \(\mathbf{V}\) is a diagonal matrix with \(i\)th
-diagonal element equal to the variance of the response distribution
-evaluated at \(f^{-1}(\mathbf{w}_i)\) \citep{faraway2016extending};
-\(\mathbf{V}\) is sometimes called the GLM weight matrix. The larger the
-value of \(\mathbf{L}_{ii}\), the more severe the leverage from the
-\(i\)th observation.
+  \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2}.
+\end{align} The larger the value of \(\mathbf{L}_{ii}\), the more severe
+the leverage from the \(i\)th observation.
 
@@
 and \(diag(\mathbf{L})\) indicates the diagonal elements of the leverage
-matrix. The larger the value of \(\mathbf{c}_i\), the more severe the
+matrix. The larger the value of \(\text{c}_i\), the more severe the
 influence from the \(i\)th observation.
@@
 model. Like \(R^2\), \(PR^2\) can be adjusted to account for the numbers
-of parameters estimated in a model. Like \(R^2\), \(PR^2\) can be
-adjusted to account for the numbers of parameters estimated in a model.
-Because the \(deviance_{total}\) denominator changes across fitted
-models (as the values of \(\hat{\boldsymbol{\theta}}\) and
-\(\hat{\varphi}\) change), this statistic should not be used as a model
-comparison tool. Rather, it should be used as an informative diagnostic
-tool that is unique to each model fit and describes how much variability
-from that model is attributable to the explanatory variables.
+of parameters estimated in a model. Because the \(deviance_{total}\)
+denominator changes across fitted models (as the values of
+\(\hat{\boldsymbol{\theta}}\) and \(\hat{\varphi}\) change), this
+statistic should not be used as a model comparison tool. Rather, it
+should be used as an informative diagnostic tool that is unique to each
+model fit and describes how much variability from that model is
+attributable to the explanatory variables.
 
@@
 locations and new locations (spatial prediction is also called Kriging;
-see \citet{cressie1990origins}). Again suppose that we observed
-\(\mathbf{w}\) and we want to make predictions at \(\mathbf{u}\), a
-vector of latent means at the new locations that follows the same SPGLM
-from Equation\(~\)\ref{eq-spglm} and has design matrix,
-\(\mathbf{X}_{\mathbf{u}}\). The vector
+see \citet{cressie1990origins}). Like in Section\(~\)\ref{sec:fixed},
+suppose first that we observed \(\mathbf{w}\) and we want to make
+predictions at \(\mathbf{u}\), a vector of latent means at the new
+locations that follows the same SPGLM from Equation\(~\)\ref{eq-spglm}
+and has design matrix, \(\mathbf{X}_{\mathbf{u}}\). The vector
 \((\mathbf{w}, \mathbf{u})^\top\) has expectation
@@
 \(\boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} = \text{Var}(\mathbf{u}, \mathbf{u})\).
-By assumption, we have observed \(\mathbf{w}\), so we may derive the
-conditional distribution of \(\mathbf{u}|\mathbf{w}\), which has the
-following properties: \begin{align}
+Thus we may derive the conditional distribution of
+\(\mathbf{u}|\mathbf{w}\), which has the following properties:
+\begin{align}
   \text{E}(\mathbf{u} | \mathbf{w}) & = \mathbf{X}_{\mathbf{u}} \boldsymbol{\beta} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{w} - \mathbf{X}\boldsymbol{\beta}) \\
   \text{Var}(\mathbf{u} | \mathbf{w}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
-\end{align} \citet{ver2024marginal} show how these equations are
-adjusted to reflect uncertainty in both \(\hat{\boldsymbol{\beta}}\) and
-\(\hat{\mathbf{w}}\) while leveraging the laws of total expectation and
-variance yet again. They derive the predictor of \(\mathbf{u}\),
-\(\hat{\mathbf{u}}\), and its associated variance, given by:
+\end{align} Recall, however, that we do not actually observe
+\(\mathbf{w}\) and instead compute \(\hat{\mathbf{w}}\); so, to predict
+\(\mathbf{u}\) and quantify its uncertainty, we must again leverage the
+laws of total expectation and variance. \citet{ver2024marginal} show
+that \(\hat{\mathbf{u}}\) and its associated variance are given by:
 \begin{align}
@@
 for a vector of ones, \(\mathbf{1}\). As with
-\(\hat{\boldsymbol{\beta}}\), in practice these covariance matrices are
-evaluated at \(\hat{\boldsymbol{\theta}}\).
+\(\hat{\boldsymbol{\beta}}\), these covariance matrices are evaluated at
+\(\hat{\boldsymbol{\theta}}\) in practice.
 
@@
 The results in this manuscript were obtained using {R} 4.4.0 with the
-\pkg{spmodel} 0.11.0 package. Figures were created using the {ggplot2}
+\pkg{spmodel} 0.11.1 package. Figures were created using the {ggplot2}
 3.5.1 package \citep{wickham2016ggplot2} and base {R}.
@@
 at \url{https://CRAN.R-project.org/package=spmodel}. Results were
-obtained using {R} 4.4.0 with the \pkg{spmodel} 0.11.0 package. Figures
+obtained using {R} 4.4.0 with the \pkg{spmodel} 0.11.1 package. Figures
 were created using the {ggplot2} 3.5.1 package

### 2025-11-27 inst/manuscript/manuscript.Rmd
some more updates

@@
 
-There are five columns: \code{elev}, the numeric site elevation (meters); \code{strat} a stratification variable for sampling with two levels, \code{"L"} and \code{"M"}, which are categorized by landscape metrics at each site; \code{count}, the number of moose at each site; \code{presence}, a factor that indicates whether at least one moose was observed at each site (\code{0} implies no moose; \code{1} implies at least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG: 3338) projected coordinate of each site.
-These data are point-referenced because each observation occurs at point coordinates and are represented by a \code{POINT} geometry.
+There are five columns in `moose`: \code{elev}, the numeric site elevation (meters); \code{strat} a stratification variable for sampling with two levels (\code{"L"} and \code{"M"}) categorized by landscape metrics at each site; \code{count}, the number of moose at each site; \code{presence}, a factor that indicates whether at least one moose was observed at each site (\code{0} implies no moose; \code{1} implies at least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG: 3338) projected coordinate of each site.
+These data have point-referenced support because each observation occurs at point coordinates represented by a \code{POINT} geometry.
 Moose are most prevalent in the southwestern and eastern parts of the Togiak region (Figure$~$\ref{fig-moose-data}).
@@
 
-SPGLMs in \pkg{spmodel} for point-referenced data are fit using the \code{spglm()} function.
+SPGLMs in \pkg{spmodel} for point-referenced support are fit using the \code{spglm()} function.
 The \code{spglm()} function requires four arguments: \code{formula}, the relationship between the response and explanatory variables; \code{family}, the response distribution assumed for the response variable; \code{data}, the data frame that contains the variables in \code{formula}, and \code{spcov_type}, the type of spatial covariance.
-The \code{formula}, \code{family}, and \code{data} arguments are the three required arguments to \code{glm()} for nonspatial GLMs.
-So, the transition from \code{glm()} to \code{spglm()}
+The \code{formula}, \code{family}, and \code{data} arguments are familiar because they are the three required arguments to \code{glm()} for nonspatial GLMs.
+So the transition from \code{glm()} to \code{spglm()}
 simply requires one additional argument: \code{spcov_type}.
@@
 )
-```
 
-While the \code{spglm()} approach evaluates the HGLMM likelihood with $\sigma^2_{de} = 0$ and $\sigma^2_{ie} \approx 0$ instead of just the GLM likelihood, the parameter estimates and their standard errors are the same:
-
-```{r}
 bin_glm <- glm(
@@
 
-However, using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel} helper functions are available and that each of the \code{spglm()} models uses the same likelihood:
+Using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel} helper functions are available and that each of the \code{spglm()} models uses the same hierarchical likelihood (Equation~\ref{eq-marginal}):
 ```{r}
@@
 The likelihood-based statistics AIC, AICc, BIC, and deviance are much lower for the SPGLM, indicating a better fit relative to the GLM.
-We may also perform a likelihood ratio test (LRT) between the two models, as the GLM is a special case of the SPGLM (i.e., is nested within the SPGLM):
-```{r}
-anova(spbin, bin)
-```
 
-The LRT provides strong evidence that the SPGLM is preferred to the GLM ($p~$value < 0.001).
-
-An alternative approach to model comparison is to use a cross-validation procedure [@james2013introduction].
+Instead of relying on likelihood-based statistics, models can be compared using a cross-validation procedure [@james2013introduction].
 The \code{loocv()} function performs leave-one-out cross validation, comparing the predicted mean (on the response scale) to the observed response variable for each hold-out observation, recomputing estimates of $\boldsymbol{\beta}$ in each iteration.
@@
 Both models have negligible bias, but the SPGLM has much lower MSPE and RMSPE than the GLM, indicating the SPGLM predictions are far more efficient.
-Three separate metrics (likelihood-based statistics, likelihood-ratio test, and leave-one-out cross validation) prefer the SPGLM to the GLM.
+Both the liklelihood-based leave-one-out cross validation metrics prefer the SPGLM to the GLM.
 
-We can compare two SPGLMs with different spatial covariance functions using likelihood-based statistics and leave-one-out cross validation, but we can't use the LRT because generally, the spatial covariance functions are not nested:
+We may also want to compare the fit of two SPGLMs with different spatial covariance structures.
+For example, we may want to compare the fit of an SPGLM with the `"exponential"` spatial covariance to the fit of a SPGLM with the `"gaussian"` spatial covariance:
 ```{r}
@@
 
-The \code{"exponential"} spatial covariance (\code{spbin}) has a slightly lower (better) deviance but slightly higher (worse) AIC, AICc, and BIC than the \code{"gaussian"} spatial covariance (\code{spbin2}).
-Both spatial covariance functions have similar leave-one-out cross validation metrics, though the \code{"gaussian"} spatial covariance RMSPE is slightly lower (better).
+The SPGLM with the \code{"exponential"} spatial covariance (\code{spbin}) has a slightly lower (better) deviance but slightly higher (worse) AIC, AICc, and BIC than the SPGLM with the \code{"gaussian"} spatial covariance (\code{spbin2}).
+Both SPGLMs have similar leave-one-out cross validation metrics, though the SPGLM with the \code{"gaussian"} spatial covariance has slightly lower (better) RMSPE.
 For practical purposes, these models fit similarly.

### 2025-11-27 inst/manuscript/manuscript.tex
some more updates

@@
 
-There are five columns: \code{elev}, the numeric site elevation
-(meters); \code{strat} a stratification variable for sampling with two
-levels, \code{"L"} and \code{"M"}, which are categorized by landscape
+There are five columns in \texttt{moose}: \code{elev}, the numeric site
+elevation (meters); \code{strat} a stratification variable for sampling
+with two levels (\code{"L"} and \code{"M"}) categorized by landscape
 metrics at each site; \code{count}, the number of moose at each site;
@@
 least one moose); and \code{geometry}, the NAD83/Alaska Albers (EPSG:
-3338) projected coordinate of each site. These data are point-referenced
-because each observation occurs at point coordinates and are represented
-by a \code{POINT} geometry. Moose are most prevalent in the southwestern
-and eastern parts of the Togiak region
+3338) projected coordinate of each site. These data have
+point-referenced support because each observation occurs at point
+coordinates represented by a \code{POINT} geometry. Moose are most
+prevalent in the southwestern and eastern parts of the Togiak region
 (Figure\(~\)\ref{fig-moose-data}).
@@
 
-SPGLMs in \pkg{spmodel} for point-referenced data are fit using the
+SPGLMs in \pkg{spmodel} for point-referenced support are fit using the
 \code{spglm()} function. The \code{spglm()} function requires four
@@
 covariance. The \code{formula}, \code{family}, and \code{data} arguments
-are the three required arguments to \code{glm()} for nonspatial GLMs.
-So, the transition from \code{glm()} to \code{spglm()} simply requires
-one additional argument: \code{spcov_type}. When \code{data} is not an
-\code{sf} object, \code{spglm()} also requires the \code{xcoord} and
-\code{ycoord} arguments, which indicate the columns in \code{data} that
-represent the projected x- and y-coordinates, respectively.
+are familiar because they are the three required arguments to
+\code{glm()} for nonspatial GLMs. So the transition from \code{glm()} to
+\code{spglm()} simply requires one additional argument:
+\code{spcov_type}. When \code{data} is not an \code{sf} object,
+\code{spglm()} also requires the \code{xcoord} and \code{ycoord}
+arguments, which indicate the columns in \code{data} that represent the
+projected x- and y-coordinates, respectively.
 
@@
 + )
-\end{CodeInput}
-\end{CodeChunk}
-
-While the \code{spglm()} approach evaluates the HGLMM likelihood with
-\(\sigma^2_{de} = 0\) and \(\sigma^2_{ie} \approx 0\) instead of just
-the GLM likelihood, the parameter estimates and their standard errors
-are the same:
-
-\begin{CodeChunk}
-\begin{CodeInput}
+R> 
 R> bin_glm <- glm(
@@
 
-However, using \code{spglm()} instead of \code{glm()} ensures that
-\pkg{spmodel} helper functions are available and that each of the
-\code{spglm()} models uses the same likelihood:
+Using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel}
+helper functions are available and that each of the \code{spglm()}
+models uses the same hierarchical likelihood
+(Equation\textasciitilde{}\ref{eq-marginal}):
 
@@
 The likelihood-based statistics AIC, AICc, BIC, and deviance are much
-lower for the SPGLM, indicating a better fit relative to the GLM. We may
-also perform a likelihood ratio test (LRT) between the two models, as
-the GLM is a special case of the SPGLM (i.e., is nested within the
-SPGLM):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> anova(spbin, bin)
-\end{CodeInput}
-\begin{CodeOutput}
-Likelihood Ratio Test
-
-Response: presence
-             Df   Chi2 Pr(>Chi2)    
-spbin vs bin  3 31.546 6.525e-07 ***
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{CodeOutput}
-\end{CodeChunk}
-
-The LRT provides strong evidence that the SPGLM is preferred to the GLM
-(\(p~\)value \textless{} 0.001).
-
-An alternative approach to model comparison is to use a cross-validation
-procedure \citep{james2013introduction}. The \code{loocv()} function
-performs leave-one-out cross validation, comparing the predicted mean
-(on the response scale) to the observed response variable for each
-hold-out observation, recomputing estimates of \(\boldsymbol{\beta}\) in
-each iteration. Performing leave-one-out cross validation tends to be
-more computationally efficient than fitting the model, as leave-one-out
-cross validation requires only one set of products involving the inverse
+lower for the SPGLM, indicating a better fit relative to the GLM.
+
+Instead of relying on likelihood-based statistics, models can be
+compared using a cross-validation procedure
+\citep{james2013introduction}. The \code{loocv()} function performs
+leave-one-out cross validation, comparing the predicted mean (on the
+response scale) to the observed response variable for each hold-out
+observation, recomputing estimates of \(\boldsymbol{\beta}\) in each
+iteration. Performing leave-one-out cross validation tends to be more
+computationally efficient than fitting the model, as leave-one-out cross
+validation requires only one set of products involving the inverse
 covariance matrix (a primary computational burden), while fitting
@@
 RMSPE than the GLM, indicating the SPGLM predictions are far more
-efficient. Three separate metrics (likelihood-based statistics,
-likelihood-ratio test, and leave-one-out cross validation) prefer the
-SPGLM to the GLM.
-
-We can compare two SPGLMs with different spatial covariance functions
-using likelihood-based statistics and leave-one-out cross validation,
-but we can't use the LRT because generally, the spatial covariance
-functions are not nested:
+efficient. Both the liklelihood-based leave-one-out cross validation
+metrics prefer the SPGLM to the GLM.
+
+We may also want to compare the fit of two SPGLMs with different spatial
+covariance structures. For example, we may want to compare the fit of an
+SPGLM with the \texttt{"exponential"} spatial covariance to the fit of a
+SPGLM with the \texttt{"gaussian"} spatial covariance:
 
@@
 
-The \code{"exponential"} spatial covariance (\code{spbin}) has a
-slightly lower (better) deviance but slightly higher (worse) AIC, AICc,
-and BIC than the \code{"gaussian"} spatial covariance (\code{spbin2}).
-Both spatial covariance functions have similar leave-one-out cross
-validation metrics, though the \code{"gaussian"} spatial covariance
-RMSPE is slightly lower (better). For practical purposes, these models
-fit similarly.
+The SPGLM with the \code{"exponential"} spatial covariance
+(\code{spbin}) has a slightly lower (better) deviance but slightly
+higher (worse) AIC, AICc, and BIC than the SPGLM with the
+\code{"gaussian"} spatial covariance (\code{spbin2}). Both SPGLMs have
+similar leave-one-out cross validation metrics, though the SPGLM with
+the \code{"gaussian"} spatial covariance has slightly lower (better)
+RMSPE. For practical purposes, these models fit similarly.
 

### 2025-11-28 inst/manuscript/manuscript.Rmd
more updates

@@
 \centering
-\includegraphics[width = 0.7\linewidth]{figures/figure-01.png}
+\includegraphics[width = 0.7\linewidth]{figures/figure-1.png}
 \caption{An exponential spatial correlation function with varying range parameters.}
@@
 \centering
-\includegraphics[width = 0.7\linewidth]{figures/figure-02.png}
+\includegraphics[width = 0.7\linewidth]{figures/figure-2.png}
 \caption{Exponential, Gaussian, and spherical spatial correlation functions all with range parameters equal to 0.5.}
@@
 The novel Laplace approach implemented in \pkg{spmodel} formally maximizes a hierarchical GLM likelihood [@lee1996hierarchical; @wood2017generalized], making likelihood-based statistics for model comparison like AIC [@akaike1974new], AICc [@hoeting2006model], BIC [@schwarz1978estimating], deviance [@mccullagh1989generalized], and likelihood ratio tests available.
-These types of statistics are not available for quasi-likelihood [@wedderburn1974quasi; @breslow1993approximate] or pseudo-likelihood approaches [@wolfinger1993generalized], which only specify the first two moments of a distribution.
+These types of statistics are not directly available for quasi-likelihood [@wedderburn1974quasi; @breslow1993approximate] or pseudo-likelihood approaches [@wolfinger1993generalized], which only specify the first two moments of a distribution.
 Next, we describe a brief overview of the approach and how it can be used for several primary data analysis tasks [@tredennick2021practical] like model comparison, parameter estimation, inference, model diagnostics, and prediction.
@@
 \centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-1.png}
+\includegraphics[width = 0.70\linewidth]{figures/figure-3.png}
 \caption{Moose presence in Alaska. Circles represent moose presence or absence (based on color) and triangles represent locations at which moose presence probability predictions are desired.}
@@
 The default fitted values are on the link scale and the default residuals are deviance residuals.
-Also returned by \code{augment()} are the leverage (\code{.hat}), Cook's distance (\code{.cooksd}), and standardized residuals (\code{.std.resid}) described in Section$~$\ref{sec:diagnostics}.
+Also returned by \code{augment()} are the leverage (\code{.hat}), Cook's distance (\code{.cooksd}), and standardized residuals (\code{.std.resid}).
 A benefit of using \code{augment()} when \code{data} is an \code{sf} object is that the output is also an \code{sf} object, which makes it straightforward to create spatial diagnostic plots (Figure$~$\ref{fig-sp-diagnostic}).
@@
 \centering
-\includegraphics[width = 1\linewidth]{figures/figure-2.png}
+\includegraphics[width = 1\linewidth]{figures/figure-4.png}
 \caption{Moose presence model diagnostics, including leverage values (left) and standardized residuals (right).}
@@
 \centering
-\includegraphics[width = 1\linewidth]{figures/figure-3.png}
+\includegraphics[width = 1\linewidth]{figures/figure-5.png}
 \caption{Moose presence model diagnostics, including Cook's distance (left) and the fitted spatial covariance as a function of distance (right).}
@@
 \centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-4.png}
-\caption{Moose presence probability fitted values and predictions. Fitted values are represented by circles and predictions by triangles.}
+\includegraphics[width = 0.70\linewidth]{figures/figure-6.png}
+\caption{Moose presence probability fitted values ($f^{-1}(\hat{\mathbf{w}})$) and predictions. Fitted values are represented by circles and predictions by triangles.}
 \label{fig-moose-fit}
@@
 
-Predictions on the response scale are visualized alongside the fitted values ($f^{-1}(\hat{\mathbf{w}})$) in Figure$~$\ref{fig-moose-fit}.
+Moose are far more common in the eastern and southwestern portion of the domain (Figure$~$\ref{fig-moose-fit}).
 Prediction intervals for the probability of moose presence (on the link scale) are returned by supplying \code{interval}:
@@
 \centering
-\includegraphics[width = 1\linewidth]{figures/figure-5.png}
+\includegraphics[width = 1\linewidth]{figures/figure-7.png}
 \caption{Moose presence 95\% prediction interval lower bounds (left) and upper bounds (right).}
@@
 In Section$~$\ref{sec-seal}, we fit a binomial model to the areal harbor seal trend data with a nonspatial random effect.
-Finally in Section$~$\ref{sec-texas}, we fit beta models to Texas voter turnout data, which can be treated as point-referenced or areal, and use maximum likelihood to compare two models with different explanatory variables.
-Table$~$\ref{tab:additional} outlines, for each application, the section number, data set, family (i.e., response distribution), geometry type (point-referenced or areal), and additional \pkg{spmodel} features highlighted.
+Finally in Section$~$\ref{sec-texas}, we fit beta models to Texas voter turnout data, which can be modeled using point-referenced or areal support, and use maximum likelihood to compare two models with different explanatory variables.
+Table$~$\ref{tab:additional} outlines, for each application, the section number, data set, family (i.e., response distribution), geometry type (point-referenced or areal support), and additional \pkg{spmodel} features highlighted.
 
@@
 \centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-6.png}
+\includegraphics[width = 0.70\linewidth]{figures/figure-8.png}
 \caption{Moose counts in Alaska. Circles represent moose counts (based on color) and triangles represent locations at which mean count predictions are desired.}
@@
 
-The spatial covariance is strongest in a northwest-southeast direction and weakest in the northeast-southwest direction (Figure$~$\ref{fig-tropy}), which is intuitive given the similar patterns in moose counts from Figure$~$\ref{fig-moose-data-count}.
+From Figure$~$\ref{fig-tropy}, the spatial covariance is strongest in a northwest-southeast direction (2.729 radians) and weakest in the northeast-southwest direction (0.413 radians), which is intuitive given the similar patterns in moose counts from Figure$~$\ref{fig-moose-data-count}.
 
@@
 \centering
-\includegraphics[width = 1\linewidth]{figures/figure-7.png}
+\includegraphics[width = 1\linewidth]{figures/figure-9.png}
 \caption{Level curves of equal spatial covariance for the negative binomial moose count models.
@@
 In the isotropic level curve (left), spatial covariance decays equally in all directions.
-In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts).}
+In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction (0.413 radians) and slowest in the northwest-southeast direction (2.729 radians).}
 \label{fig-tropy}
@@
 The \code{seal} data in \pkg{spmodel} contains harbor seal abundance trends for two different harbor seal stocks (genetically distinct populations).
-While the \code{moose} and \code{lake} data were point-referenced, the \code{seal} data are areal.
+While the \code{moose} and \code{lake} data have point-referenced support, the \code{seal} data have areal support.
 Each polygon in the \code{seal} data represents a distinct harbor seal haulout region (Figure$~$\ref{fig-seal}).
@@
 \centering
-\includegraphics[width = 1\linewidth]{figures/figure-8.png}
+\includegraphics[width = 1\linewidth]{figures/figure-10.png}
 \caption{Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed log seal trends (right).}
@@
 
-We could have alternatively used a (geostatistical) SPGLM via \code{spglm()}.
-When areal data are used with \code{spglm()}, the centroids of each polygon are used as the point-referenced coordinates.
-We further explore comparisons between point-referenced and areal data in the next example.
+We could have alternatively used a (geostatistical) SPGLM via \code{spglm()}. When areal data are used with \code{spglm()}, the centroids of each polygon are used as the point-referenced coordinates.
 
@@
 \centering
-\includegraphics[width = 0.70\linewidth]{figures/figure-10.png}
+\includegraphics[width = 0.70\linewidth]{figures/figure-11.png}
 \caption{Proportion of voter turnout in Texas for the 1980 presidential election.
@@
 The \code{texas} data in \pkg{spmodel} contains voter turnout data for Texas counties in the 1980 United States Presidential Election [@bivand2024spdata].
-The data are point-referenced, with polygon centroids representing the spatial location of each county (Figure$~$\ref{fig-texas}).
+The data have point-referenced support, with polygon centroids representing the spatial location of each county (Figure$~$\ref{fig-texas}).
 Beta regression is a GLM used to model rate and proportion data in the (0, 1) interval [@ferrari2004beta; @cribari2010beta].
-We model voter turnout rates as a function of mean log income of county residents using an SPGLM assuming a beta distributed response variable:
+We model voter turnout rates as a function of mean log income of county residents using a geostatistical SPGLM assuming a beta distributed response variable:
 ```{r}
@@
 
-Alternatively, we could use an autoregressive model to fit the model, constructing a neighborhood matrix by assuming centroids within \code{cutoff} of one another are neighbors:
+Alternatively, we could fit an autoregressive SPGLM by assuming the data have areal support and constructing a neighborhood matrix that treats counties as neighbors if the distance between their centroids is less than \code{cutoff}:
 ```{r}
@@
 
-According to AIC, the SPGLM for point-referenced data is preferred:
+According to AIC, the geostatistical SPGLM:
 ```{r}
@@
 The default estimation method in \pkg{spmodel} for SPGLMs is restricted maximum likelihood (REML), while maximum likelihood (ML) can also be used.
-A benefit of benefit of REML is that it can yield unbiased estimates of covariance parameters [@cressie1993asymptotic], but a drawback is that likelihood-based statistics are only valid for model comparison when the models have the same explanatory variable and fixed effect structure (because the error contrasts used to construct the REML likelihood change based on $\mathbf{X}$ and $\boldsymbol{\beta}$).
-In contrast, ML estimators are generally biased for covariance parameters, though in practice this bias tends to be small.
-Moreover, when using ML, likelihood-based comparisons are valid for models having different explanatory variable and fixed effect structures.
-Using ML, we can evaluate the significance of log income on voter turnout using a likelihood ratio test:
+A benefit of benefit of REML is that it can yield unbiased estimates of covariance parameters [@cressie1993asymptotic].
+A drawback of REML is that likelihood-based statistics are often considered invalid when the models have different explanatory variable or fixed effect structures [@wolfinger1993covariance], though @gurka2006selecting provides some evidence to the contrary.
+In contrast to REML estimators, ML estimators are generally biased for covariance parameters, though in practice this bias tends to be small.
+Moreover, when using ML, likelihood-based comparisons are valid for models with different explanatory variable or fixed effect structures.
+Using ML, we can evaluate the significance of log income on voter turnout using a likelihood ratio test (for nested models):
 ```{r}
 spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
-spbeta_red_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
-anova(spbeta_full_ml, spbeta_red_ml)
+spbeta_reduced_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
+anova(spbeta_full_ml, spbeta_reduced_ml)
 ```
@@
 ```{r}
-AIC(spbeta_full_ml, spbeta_red_ml)
+AIC(spbeta_full_ml, spbeta_reduced_ml)
 ```
@@
 SPGLMs are fit in \pkg{spmodel} using a novel application of the Laplace approximation that simultaneously marginalizes over the latent (i.e., unobserved) random effects and the fixed effects.
-\pkg{spmodel}'s \code{spglm()} (for point-referenced data) and \code{spgautor()} (for areal data) fit SPGLMs that are similar in structure and syntax as base \proglang{R}'s \code{glm()} function, easing the transition from GLMs to SPGLMs for practitioners.
+\pkg{spmodel}'s \code{spglm()} (for point-referenced support) and \code{spgautor()} (for areal support) fit SPGLMs that are similar in structure and syntax as base \proglang{R}'s \code{glm()} function, easing the transition for practitioners from GLMs to SPGLMs.
 The \code{spglm()} and \code{spgautor()} functions support six response distributions for binary, count, and skewed data and 20 spatial covariance functions.
@@
 All data used are part of the \pkg{spmodel} [R]{.proglang} package available for download from CRAN at [https://CRAN.R-project.org/package=spmodel](https://CRAN.R-project.org/package=spmodel).
-Results were obtained using [R]{.proglang} 4.4.0 with the \pkg{spmodel} 0.11.1 package.
-Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.
 

### 2025-11-28 inst/manuscript/manuscript.tex
more updates

@@
 \citep{mccullagh1989generalized}, and likelihood ratio tests available.
-These types of statistics are not available for quasi-likelihood
-\citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
-approaches \citep{wolfinger1993generalized}, which only specify the
-first two moments of a distribution. Next, we describe a brief overview
-of the approach and how it can be used for several primary data analysis
-tasks \citep{tredennick2021practical} like model comparison, parameter
-estimation, inference, model diagnostics, and prediction.
+These types of statistics are not directly available for
+quasi-likelihood \citep{wedderburn1974quasi, breslow1993approximate} or
+pseudo-likelihood approaches \citep{wolfinger1993generalized}, which
+only specify the first two moments of a distribution. Next, we describe
+a brief overview of the approach and how it can be used for several
+primary data analysis tasks \citep{tredennick2021practical} like model
+comparison, parameter estimation, inference, model diagnostics, and
+prediction.
 
@@
 \code{augment()} are the leverage (\code{.hat}), Cook's distance
-(\code{.cooksd}), and standardized residuals (\code{.std.resid})
-described in Section\(~\)\ref{sec:diagnostics}. A benefit of using
-\code{augment()} when \code{data} is an \code{sf} object is that the
-output is also an \code{sf} object, which makes it straightforward to
-create spatial diagnostic plots (Figure\(~\)\ref{fig-sp-diagnostic}).
-Standard \proglang{R} helpers (e.g., \code{fitted()},
-\code{residuals()}) are also available to extract model diagnostics from
-the model object.
+(\code{.cooksd}), and standardized residuals (\code{.std.resid}). A
+benefit of using \code{augment()} when \code{data} is an \code{sf}
+object is that the output is also an \code{sf} object, which makes it
+straightforward to create spatial diagnostic plots
+(Figure\(~\)\ref{fig-sp-diagnostic}). Standard \proglang{R} helpers
+(e.g., \code{fitted()}, \code{residuals()}) are also available to
+extract model diagnostics from the model object.
 
@@
 
-Predictions on the response scale are visualized alongside the fitted
-values (\(f^{-1}(\hat{\mathbf{w}})\)) in Figure\(~\)\ref{fig-moose-fit}.
-Prediction intervals for the probability of moose presence (on the link
-scale) are returned by supplying \code{interval}:
+Moose are far more common in the eastern and southwestern portion of the
+domain (Figure\(~\)\ref{fig-moose-fit}). Prediction intervals for the
+probability of moose presence (on the link scale) are returned by
+supplying \code{interval}:
 
@@
 Section\(~\)\ref{sec-texas}, we fit beta models to Texas voter turnout
-data, which can be treated as point-referenced or areal, and use maximum
-likelihood to compare two models with different explanatory variables.
-Table\(~\)\ref{tab:additional} outlines, for each application, the
-section number, data set, family (i.e., response distribution), geometry
-type (point-referenced or areal), and additional \pkg{spmodel} features
-highlighted.
+data, which can be modeled using point-referenced or areal support, and
+use maximum likelihood to compare two models with different explanatory
+variables. Table\(~\)\ref{tab:additional} outlines, for each
+application, the section number, data set, family (i.e., response
+distribution), geometry type (point-referenced or areal support), and
+additional \pkg{spmodel} features highlighted.
 
@@
 
-The spatial covariance is strongest in a northwest-southeast direction
-and weakest in the northeast-southwest direction
-(Figure\(~\)\ref{fig-tropy}), which is intuitive given the similar
-patterns in moose counts from Figure\(~\)\ref{fig-moose-data-count}.
+From Figure\(~\)\ref{fig-tropy}, the spatial covariance is strongest in
+a northwest-southeast direction (2.729 radians) and weakest in the
+northeast-southwest direction (0.413 radians), which is intuitive given
+the similar patterns in moose counts from
+Figure\(~\)\ref{fig-moose-data-count}.
 
@@
 trends for two different harbor seal stocks (genetically distinct
-populations). While the \code{moose} and \code{lake} data were
-point-referenced, the \code{seal} data are areal. Each polygon in the
-\code{seal} data represents a distinct harbor seal haulout region
-(Figure\(~\)\ref{fig-seal}). A haulout region is an area of coastal
-rocks that harbor seals go to rest, molt, and give birth.
+populations). While the \code{moose} and \code{lake} data have
+point-referenced support, the \code{seal} data have areal support. Each
+polygon in the \code{seal} data represents a distinct harbor seal
+haulout region (Figure\(~\)\ref{fig-seal}). A haulout region is an area
+of coastal rocks that harbor seals go to rest, molt, and give birth.
 
@@
 centroids of each polygon are used as the point-referenced coordinates.
-We further explore comparisons between point-referenced and areal data
-in the next example.
 
@@
 Texas counties in the 1980 United States Presidential Election
-\citep{bivand2024spdata}. The data are point-referenced, with polygon
-centroids representing the spatial location of each county
+\citep{bivand2024spdata}. The data have point-referenced support, with
+polygon centroids representing the spatial location of each county
 (Figure\(~\)\ref{fig-texas}). Beta regression is a GLM used to model
@@
 \citep{ferrari2004beta, cribari2010beta}. We model voter turnout rates
-as a function of mean log income of county residents using an SPGLM
-assuming a beta distributed response variable:
+as a function of mean log income of county residents using a
+geostatistical SPGLM assuming a beta distributed response variable:
 
@@
 
-Alternatively, we could use an autoregressive model to fit the model,
-constructing a neighborhood matrix by assuming centroids within
-\code{cutoff} of one another are neighbors:
+Alternatively, we could fit an autoregressive SPGLM by assuming the data
+have areal support and constructing a neighborhood matrix that treats
+counties as neighbors if the distance between their centroids is less
+than \code{cutoff}:
 
@@
 
-According to AIC, the SPGLM for point-referenced data is preferred:
+According to AIC, the geostatistical SPGLM:
 
@@
 used. A benefit of benefit of REML is that it can yield unbiased
-estimates of covariance parameters \citep{cressie1993asymptotic}, but a
-drawback is that likelihood-based statistics are only valid for model
-comparison when the models have the same explanatory variable and fixed
-effect structure (because the error contrasts used to construct the REML
-likelihood change based on \(\mathbf{X}\) and \(\boldsymbol{\beta}\)).
-In contrast, ML estimators are generally biased for covariance
-parameters, though in practice this bias tends to be small. Moreover,
-when using ML, likelihood-based comparisons are valid for models having
-different explanatory variable and fixed effect structures. Using ML, we
-can evaluate the significance of log income on voter turnout using a
-likelihood ratio test:
+estimates of covariance parameters \citep{cressie1993asymptotic}. A
+drawback of REML is that likelihood-based statistics are often
+considered invalid when the models have different explanatory variable
+or fixed effect structures \citep{wolfinger1993covariance}, though
+\citet{gurka2006selecting} provides some evidence to the contrary. In
+contrast to REML estimators, ML estimators are generally biased for
+covariance parameters, though in practice this bias tends to be small.
+Moreover, when using ML, likelihood-based comparisons are valid for
+models with different explanatory variable or fixed effect structures.
+Using ML, we can evaluate the significance of log income on voter
+turnout using a likelihood ratio test (for nested models):
 
@@
 R> spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
-R> spbeta_red_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
-R> anova(spbeta_full_ml, spbeta_red_ml)
+R> spbeta_reduced_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
+R> anova(spbeta_full_ml, spbeta_reduced_ml)
 \end{CodeInput}
@@
                                     Df   Chi2 Pr(>Chi2)    
-spbeta_red_ml vs spbeta_full_ml  1 23.155 1.494e-06 ***
+spbeta_reduced_ml vs spbeta_full_ml  1 23.155 1.494e-06 ***
 ---
@@
 \begin{CodeInput}
-R> AIC(spbeta_full_ml, spbeta_red_ml)
+R> AIC(spbeta_full_ml, spbeta_reduced_ml)
 \end{CodeInput}
@@
 spbeta_full_ml     7 -31.25900
-spbeta_red_ml   6 -10.10354
+spbeta_reduced_ml  6 -10.10354
 \end{CodeOutput}
@@
 unobserved) random effects and the fixed effects. \pkg{spmodel}'s
-\code{spglm()} (for point-referenced data) and \code{spgautor()} (for
-areal data) fit SPGLMs that are similar in structure and syntax as base
-\proglang{R}'s \code{glm()} function, easing the transition from GLMs to
-SPGLMs for practitioners. The \code{spglm()} and \code{spgautor()}
-functions support six response distributions for binary, count, and
-skewed data and 20 spatial covariance functions. \pkg{spmodel} has a
-suite of tools for data visualization, inference, model diagnostics, and
-prediction, providing a framework that can be used for all stages of a
-data analysis. There are many additional \pkg{spmodel} features that are
-not covered here, including fitting multiple models simultaneously,
-fixing spatial covariance and dispersion parameters at known values,
-fitting models to large non-Gaussian data having thousands of
-observations via spatial indexing \citep{ver2023indexing}, incorporating
-spatial dependence in machine learning (e.g., random forests;
-\citet{breiman2001random}), simulating spatially dependent data (e.g.,
-\code{spbinom()}, \code{sprpois()}, etc.), and more. Further details are
-provided by \url{https://CRAN.R-project.org/package=spmodel} and links
-therein.
+\code{spglm()} (for point-referenced support) and \code{spgautor()} (for
+areal support) fit SPGLMs that are similar in structure and syntax as
+base \proglang{R}'s \code{glm()} function, easing the transition for
+practitioners from GLMs to SPGLMs. The \code{spglm()} and
+\code{spgautor()} functions support six response distributions for
+binary, count, and skewed data and 20 spatial covariance functions.
+\pkg{spmodel} has a suite of tools for data visualization, inference,
+model diagnostics, and prediction, providing a framework that can be
+used for all stages of a data analysis. There are many additional
+\pkg{spmodel} features that are not covered here, including fitting
+multiple models simultaneously, fixing spatial covariance and dispersion
+parameters at known values, fitting models to large non-Gaussian data
+having thousands of observations via spatial indexing
+\citep{ver2023indexing}, incorporating spatial dependence in machine
+learning (e.g., random forests; \citet{breiman2001random}), simulating
+spatially dependent data (e.g., \code{spbinom()}, \code{sprpois()},
+etc.), and more. Further details are provided by
+\url{https://CRAN.R-project.org/package=spmodel} and links therein.
 
@@
 part of the \pkg{spmodel} {R} package available for download from CRAN
-at \url{https://CRAN.R-project.org/package=spmodel}. Results were
-obtained using {R} 4.4.0 with the \pkg{spmodel} 0.11.1 package. Figures
-were created using the {ggplot2} 3.5.1 package
-\citep{wickham2016ggplot2} and base {R}.
+at \url{https://CRAN.R-project.org/package=spmodel}.
 

### 2025-11-28 inst/manuscript/code/code.R

@@ -1,16 +1,8 @@
 #=========================================================#
-# Load Software ####
-#=========================================================#
-
-library(spmodel)
-library(ggplot2, verbose = FALSE)
-library(dplyr, verbose = FALSE)
-library(emmeans, verbose = FALSE)
-library(car)
-
-#=========================================================#
 # Section 3: Modeling moose presence in Alaska, USA ####
 #=========================================================#
+
+library("spmodel")
 
 head(moose)
 
@@ -55,8 +47,6 @@ round(sqrt(diag(vcov(bin_glm))), digits = 4)
 glance(spbin)
 glance(bin)
 
-anova(spbin, bin)
-
 loocv(spbin)
 loocv(bin)
 
@@ -74,7 +64,7 @@ augment(spbin)
 varcomp(spbin)
 
 #==========================================#
-# Section 3.4: Model Diagnostics ####
+# Section 3.4: Prediction ####
 #==========================================#
 
 predict(spbin, newdata = moose_preds)[1:5]
@@ -124,7 +114,11 @@ summary(spgam)
 
 anova(spgam)
 
+library("car", verbose = FALSE)
+
 vif(spgam)
+
+library("emmeans", verbose = FALSE)
 
 pairs(emmeans(spgam, ~ state | temp))
 
@@ -173,16 +167,19 @@ spbeta_auto <- spgautor(
 AIC(spbeta_geo, spbeta_auto)
 
 spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
-spbeta_red_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
-anova(spbeta_full_ml, spbeta_red_ml)
+spbeta_reduced_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
+anova(spbeta_full_ml, spbeta_reduced_ml)
 
-AIC(spbeta_full_ml, spbeta_red_ml)
+AIC(spbeta_full_ml, spbeta_reduced_ml)
 
 #=========================================================#
 # Figures ####
 #=========================================================#
 
-library(patchwork)
+library("ggplot2", verbose = FALSE)
+library("dplyr", verbose = FALSE)
+library("patchwork")
+
 okabe <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
 
[…]

### 2025-11-28 inst/manuscript/figures/figures.R

@@ -24,7 +24,7 @@ cor_range <- ggplot(dat, aes(x = Distance, y = Correlation, color = Range, linet
   theme_bw(base_size = 14)
 cor_range
 ggsave(
-  filename = paste0(fig_path, "/figure-01.png"),
+  filename = paste0(fig_path, "/figure-1.png"),
   plot = cor_range,
   dpi = 300,
   height = 6.07,
@@ -32,7 +32,7 @@ ggsave(
 )
 
 #######################
-######## Figure 1
+######## Figure 2
 #######################
 h <- seq(0, 1, length.out = 1000)
 range <- 0.5
@@ -46,7 +46,7 @@ cor_type <- ggplot(dat, aes(x = Distance, y = Correlation, color = Type, linetyp
   theme_bw(base_size = 14)
 cor_type
 ggsave(
-  filename = paste0(fig_path, "/figure-02.png"),
+  filename = paste0(fig_path, "/figure-2.png"),
   plot = cor_type,
   dpi = 300,
   height = 6.07,
@@ -55,7 +55,7 @@ ggsave(
 
 
 #######################
-######## Figure 1
+######## Figure 3
 #######################
 
 okabe <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
@@ -74,7 +74,7 @@ obs_pred_presence <- ggplot(moose_comb, aes(color = presence, shape = samp)) +
   geom_sf(size = 2, data = moose_comb %>% filter(samp == "no"), color = "red", shape = 17)
 obs_pred_presence
 ggsave(
-  filename = paste0(fig_path, "/figure-1.png"),
+  filename = paste0(fig_path, "/figure-3.png"),
   plot = obs_pred_presence,
   dpi = 300,
   height = 5,
@@ -82,7 +82,7 @@ ggsave(
 )
 
 #######################
-######## Figure 2
+######## Figure 4
 #######################
 
 
@@ -110,7 +110,7 @@ p2 <- ggplot(spbin_aug, aes(color = .std.resid)) +
 
 p1 + p2
 ggsave(
-  filename = paste0(fig_path, "/figure-2.png"),
+  filename = paste0(fig_path, "/figure-4.png"),
   plot = p1 + p2,
   dpi = 300,
   height = 3,
@@ -118,11 +118,11 @@ ggsave(
 )
 
 ########################
-####### Figure 3
+####### Figure 5
 ########################
 
[…]

### 2025-11-28 inst/manuscript_old/code/code.R

@@ -1,318 +1,0 @@
-#=========================================================#
-# Load Software ####
-#=========================================================#
-
-library(spmodel)
-library(ggplot2)
-library(dplyr)
-library(emmeans)
-
-#=========================================================#
-# Section 3.1 Binary Data ####
-#=========================================================#
-
-head(moose)
-
-spbin <- spglm(
-  formula = presence ~ elev + strat + elev:strat,
-  family = binomial,
-  data = moose,
-  spcov_type = "spherical"
-)
-
-summary(spbin)
-
-tidy(spbin)
-
-moose$strat2 <- factor(moose$strat, levels = c("M", "L"))
-update(spbin, formula = presence ~ elev + strat2 + elev:strat2) |>
-  summary()
-
-emtrends(spbin, "strat", "elev")
-
-glance(spbin)
-
-head(augment(spbin))
-
-varcomp(spbin)
-
-head(predict(spbin, newdata = moose_preds))
-
-head(augment(
-  spbin,
-  newdata = moose_preds,
-  type.predict = "response",
-  interval = "prediction"
-))
-
-# fit nonspatial model using spglm()
-bin <- spglm(
-  formula = presence ~ elev + strat + elev:strat,
-  family = binomial,
-  data = moose,
-  spcov_type = "none"
-)
-# fit nonspatial model using glm()
-bin_glm <- glm(
-  formula = presence ~ elev + strat + elev:strat,
-  family = binomial,
-  data = moose
-)
-# compare fixed effect coefficients and standard errors up to four decimals
-data.frame(
-  est_none = coef(bin),
-  est_glm = coef(bin_glm),
-  se_none = sqrt(diag(vcov(bin))),
-  se_glm = sqrt(diag(vcov(bin_glm)))
-) |>
-  apply(2, round, digits = 4)
-
-glances(spbin, bin)
-
-loocv(spbin) |>
-  apply(2, round, digits = 4)
-loocv(bin) |>
-  apply(2, round, digits = 4)
-
-AUROC(spbin)
-AUROC(bin)
-
-## Figures ####
-
-### Figure 1 ####
-okabe <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
-moose_comb <- bind_rows(moose |> mutate(samp = "yes"), moose_preds |> mutate(samp = "no"))
[…]

### 2025-11-28 inst/manuscript_old/figures/figures.R

@@ -1,253 +1,0 @@
-# Preliminaries
-library(tidyverse)
-library(spmodel)
-library(patchwork)
-library(here)
-fig_path <- here("inst", "jss", "manuscript", "figures")
-
-#######################
-######## Figure 1
-#######################
-
-okabe <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
-moose_comb <- bind_rows(moose |> mutate(samp = "yes"), moose_preds |> mutate(samp = "no"))
-obs_pred_presence <- ggplot(moose_comb, aes(color = presence, shape = samp)) +
-  geom_sf(size = 2) +
-  scale_color_manual(name = "moose", values = okabe[1:2], breaks = c(1, 0), labels = c("yes", "no")) +
-  scale_shape_manual(name = "sampled", values = c(19, 17), breaks = c("yes", "no"), labels = c("yes", "no")) +
-  theme_bw(base_size = 16) +
-  scale_y_continuous(breaks = seq(62.6, 63.6, length.out = 5)) +
-  scale_x_continuous(breaks = seq(-148.5, -146, length.out = 3)) +
-  guides(
-    color = guide_legend(order = 1),
-    shape = guide_legend(order = 2)
-  ) +
-  geom_sf(size = 2, data = moose_comb %>% filter(samp == "no"), color = "red", shape = 17)
-obs_pred_presence
-ggsave(
-  filename = paste0(fig_path, "/figure-1.png"),
-  plot = obs_pred_presence,
-  dpi = 300,
-  height = 5,
-  width = 7.68
-)
-
-#######################
-######## Figure 2
-#######################
-
-
-spbin <- spglm(
-  formula = presence ~ elev + strat + elev:strat,
-  family = binomial,
-  data = moose,
-  spcov_type = "spherical"
-)
-spbin_aug <- augment(spbin)
-
-p1 <- ggplot(spbin_aug, aes(color = .hat)) +
-  geom_sf(size = 1) +
-  scale_color_viridis_c(option = "E", name = ".hat") +
-  scale_y_continuous(breaks = seq(62.6, 63.6, length.out = 5)) +
-  scale_x_continuous(breaks = seq(-148.5, -146, length.out = 3)) +
-  theme_bw(base_size = 14)
-
-p2 <- ggplot(spbin_aug, aes(color = .std.resid)) +
-  geom_sf(size = 1) +
[…]

### 2025-11-28 inst/manuscript/manuscript.Rmd
some updates

@@
 To derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$ by conditioning on $\mathbf{w}$ as if it were observed and leveraging the laws of total expectation and variance.
-Based on asymptotic properties of (restricted) maximum likelihood estimators [@cressie1993asymptotic], we may assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance equal to $-\mathbf{H}^{-1}$, the negative inverse Hessian (i.e., the inverse observed information matrix).
+Based on asymptotic properties of (restricted) maximum likelihood estimators [@cressie1993asymptotic], we may assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance approximately equal to $-\mathbf{H}^{-1}$, the negative inverse Hessian (i.e., the inverse observed information matrix).
 Thus it follows that $\text{E}(\hat{\mathbf{w}})$ is given by
@@
   \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
-  & = \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
+  & \approx \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
   & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}

### 2025-11-28 inst/manuscript/manuscript.tex
some updates

@@
 \citep{cressie1993asymptotic}, we may assume that given \(\mathbf{w}\),
-\(\hat{\mathbf{w}}\) has mean \(\mathbf{w}\) and variance equal to
-\(-\mathbf{H}^{-1}\), the negative inverse Hessian (i.e., the inverse
-observed information matrix). Thus it follows that
+\(\hat{\mathbf{w}}\) has mean \(\mathbf{w}\) and variance approximately
+equal to \(-\mathbf{H}^{-1}\), the negative inverse Hessian (i.e., the
+inverse observed information matrix). Thus it follows that
 \(\text{E}(\hat{\mathbf{w}})\) is given by \begin{align}

### 2025-12-10 inst/manuscript/manuscript.Rmd
some upates

@@
     address: |
+      | Office of Research and Development
       | 200 SW 35th St
@@
 abstract: |
-  Non-Gaussian data are common in practice and include binary, count, skewed, and proportion data types.
+  Non-Gaussian data are common in practice and include binary, count, skewed, and proportion structures.
 Often, non-Gaussian data are modeled using a generalized linear model (GLM).
@@
 
-\linenumbers
+<!-- \linenumbers -->
 
@@
 For spatial data, nearby observations tend to be more similar than distant observations [@tobler1970computer], which leads to positive spatial covariance among observations.
-The consequences of ignoring spatial covariance in statistical models for spatial data can be severe and include imprecise parameter estimates as well as misleading standard errors that inflate Type-I error rates and decrease power [@zimmerman2024spatial].
+The consequences of ignoring spatial covariance in statistical models for spatial data can be severe and include imprecise parameter estimates as well as misleading standard errors that inflate Type-I error rates and decrease power [@legendre1993spatial; @zimmerman2024spatial].
 
@@
 \end{align}
-Assuming independence among $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$, it follows that 
+Assuming independence between $\boldsymbol{\tau}$ and $\boldsymbol{\epsilon}$, it follows that 
 \begin{align}\label{eq-spcov}
@@
 For both point-referenced and areal supports, \pkg{spmodel} supports the binomial distribution for binary data, Poisson and negative binomial distributions for count data, Gamma and inverse Gaussian distributions for skewed data, and the beta distribution for proportion data.
-There are 20 different spatial covariance structures available including the exponential, Gaussian, and spherical for point-referenced support (Figure$~$\ref{fig-type}) and the conditional autoregressive, and simultaneous autoregressive structures for areal support \pkg{spmodel} provides tools for commonly used model summaries, visualizations, and diagnostics (e.g., fitted values) using standard \proglang{R} helper functions like \code{summary()}, \code{plot()}, and \code{fitted()}, among others.
-\pkg{spmodel} also provides tools to predict $\mathbf{w}$ at new locations and quantify uncertainty in those prediction using \code{predict()} and \code{augment()}.
+There are 20 different spatial covariance structures available including the exponential, Gaussian, and spherical for point-referenced support (Figure$~$\ref{fig-type}) and the conditional autoregressive and simultaneous autoregressive structures for areal support.
+\pkg{spmodel} provides commonly used model summaries, visualizations, and diagnostics (e.g., fitted values) using standard \proglang{R} helper functions like \code{summary()}, \code{plot()}, and \code{fitted()}, among others.
+\pkg{spmodel} also has tools to predict $\mathbf{w}$ at new locations and quantify uncertainty in those prediction using \code{predict()} and \code{augment()}.
 This core functionality, combined with several advanced features we describe throughout the manuscript, enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools previously missing from the existing \proglang{R} ecosystem.
@@
 Of the existing \proglang{R} packages for SPGLMs, \pkg{spmodel} (version 0.11.1) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in terms of scope and feel.
-Both packages use similar syntax as \code{glm()}, accommodate flexible \code{formula} arguments (e.g., offsets, splines), handle spatial covariance that decays at different rates in different directions (i.e., geometric anisotropy), incorporate nonspatial random effects, support other \proglang{R} packages for modeling like \pkg{broom} [@robinson2021broom; @kuhn2022tidy], \pkg{emmeans} [@lenth2024emmeans], and \pkg{car} [@fox2019car], and have tools for model summaries, prediction, and simulating data.
+Both packages use similar syntax to \code{glm()}, accommodate flexible \code{formula} arguments (e.g., transformations, splines), handle spatial covariance that decays at different rates in different directions (i.e., geometric anisotropy), incorporate nonspatial random effects, support other \proglang{R} packages for modeling like \pkg{broom} [@robinson2021broom; @kuhn2022tidy], \pkg{emmeans} [@lenth2024emmeans], and \pkg{car} [@fox2019car], and have tools for model summaries, prediction, and simulating data.
 There are some notable differences between the two packages, however.
@@
 \pkg{spmodel} supports 20 different spatial covariances and models them directly, rather than using a precision matrix approximation to the Matérn spatial covariance as in \pkg{sdmTMB}.
-\pkg{spmodel} can model data directly using neighborhood distance and autoregressive models, rather than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not be within the polygon's boundaries.
+\pkg{spmodel} can model data using neighborhood distance and autoregressive models, rather than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not be within the polygon's boundaries.
 \pkg{spmodel} provides experimental design tools (e.g., analysis of variance, contrasts), supports \pkg{sf} objects in modeling and prediction functions [@pebesma2018sf], has several specialized model diagnostics like leverage values and Cook's distances, and has analytic solutions for fixed effect and prediction standard errors.
@@
 where $\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}$.
-While we do not $\mathbf{w}$, we do estimate it via $\hat{\mathbf{w}}$, and thus it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$.
+While we do not know $\mathbf{w}$, we do estimate it via $\hat{\mathbf{w}}$, and thus it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$.
 To derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$ by conditioning on $\mathbf{w}$ as if it were observed and leveraging the laws of total expectation and variance.
@@
 Because observations often have a unique support in a GLM (e.g., only two possible response values for binary data) and the variance of an observation generally depends on its mean, response residuals lack some utility.
-Deviance residuals are a function of response residuals that are appropriately scaled to behave more like response residuals in a standard linear model.
-Deviance residuals are given by
+Deviance, Pearson, and standardized residuals are all different function of responses residuals that are appropriately scaled to behave more like response residuals from a standard linear model.
+For example, deviance residuals are given by
 \begin{align}
@@
 The deviance of the model fit quantifies twice the difference in log likelihoods between the a saturated model that fits every observation perfectly (i.e., $\text{y}_i = f^{-1}(\hat{\text{w}}_i)$ for all $i$) and the fitted model [@myers2012generalized].
-Deviance is often used as a fit statistic; lower values of deviance imply a better model fit (compared to the observed data).
-Pearson and standardized residuals are other types of GLM residuals that involve a scaling of the response residuals.
-The Pearson residuals scale $\mathbf{r}_{r}$ by the square root of $\mathbf{V}$, a diagonal matrix with $i$th diagonal element equal to the variance of the response distribution evaluated at $f^{-1}(\hat{\text{w}}_i)$ [@faraway2016extending]; $\mathbf{V}$ is sometimes called the GLM weight matrix.
-The standardized residuals scale the deviance residuals by $\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}$, where $\mathbf{L}_{ii}$ is the $i$th diagonal element of the leverage matrix, which we discuss next.
+Deviance is often used as a fit statistic, whereby lower values of deviance imply a better model fit (given the observed data).
+Scaling the response residuals by the reciprocal square root of the variance function (for the appropriate response distribution) yields the Pearson residuals, and scaling the Pearson residuals by the reciprocal square root of one minus the leverage value (which we discuss next) yields the standardized residuals [@faraway2016extending].
 
@@
 \end{align}
+where $\mathbf{V}$ is a diagonal matrix with $i$th diagonal element equal to the variance function evaluated at $f^{-1}(\hat{\text{w}}_i)$.
 The larger the value of $\mathbf{L}_{ii}$, the more severe the leverage from the $i$th observation.
@@
 
-The \code{moose} data in \pkg{spmodel} contain information on moose (Alces Alces) presence in the Togiak region of Alaska, USA.
+The \code{moose} data in \pkg{spmodel} contain information on moose (Alces alces) presence in the Togiak region of Alaska, USA.
 \code{moose} is an \code{sf} object, a special data frame that is supplemented with spatial information using the \pkg{sf} package in \proglang{R} [@pebesma2018sf].
@@
 
-Using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel} helper functions are available and that each of the \code{spglm()} models uses the same hierarchical likelihood (Equation~\ref{eq-marginal}):
+Using \code{spglm()} instead of \code{glm()} ensures that \pkg{spmodel} helper functions are available and that each of the \code{spglm()} models uses the same hierarchical likelihood (Equation$~$\ref{eq-marginal}):
 ```{r}
@@
 Both models have negligible bias, but the SPGLM has much lower MSPE and RMSPE than the GLM, indicating the SPGLM predictions are far more efficient.
-Both the liklelihood-based leave-one-out cross validation metrics prefer the SPGLM to the GLM.
+Both the likelihood-based leave-one-out cross validation metrics prefer the SPGLM to the GLM.
 
@@
 The \code{plot()} function can also be used to return similar diagnostics as from \code{lm()} and \code{glm()}, with additional tools for diagnosing spatial covariance.
-For example, we can inspect Cook's distance values and the empirical spatial covariance as a function of distance with (Figure$~$\ref{fig-sp-diagnostic2}):
+For example, we can inspect Cook's distance values and the fitted spatial covariance as a function of distance with (Figure$~$\ref{fig-sp-diagnostic2}):
 ```{r, eval = FALSE}
@@
 The Poisson distribution mean is equal to its variance, while the negative binomial has an extra parameter to accommodate overdispersion (where the variance is larger than the mean).
-Using a spherical spatial covariance function, we may fit both a Poisson and negative binomial SPGLM changing the \code{family} argument:
+Using a spherical spatial covariance function, we may fit both a Poisson and negative binomial SPGLM by changing the \code{family} argument:
 
@@
 The \code{year} partition factor (specified via \code{partition_factor}) restricts spatial covariance to be nonzero only for observations sampled during the same year.
-Data were collected in 2012 and 2017, so this partition factor assumes independence between observations in 2012 and 2017.
+Data were collected in 2012 and 2017, so this partition factor assumes observations in 2012 are independent of observations in 2017.
 While we used the partition factor here illustratively, more generally, the utility of partition factors can be highly context dependent.
@@
 Conditional and simultaneous autoregressive functions characterize spatial distance through neighborhood relationships (rather than Euclidean distance) and have \code{spcov_type} values of \code{"car"} and \code{"sar"}, respectively.
-By default, Queen's distance is used to determine whether two sites are neighbors, though custom neighborhood matrices can be passed via \code{W}.
-Row standardization is also assumed by default; this can be changed via \code{row_st}.
-Using \code{random}, we also specified a nonspatial random effect for seal stock, which implies seals belonging to the same stock share extra covariance.
+By default, Queen's distance is used to determine whether two sites are neighbors, though custom neighborhood matrices can be passed via the \code{W} argument.
+Row standardization is also assumed by default; this can be changed via the \code{row_st} argument.
+Using the \code{random} argument, we also specified a nonspatial random effect for seal stock, which implies seals belonging to the same stock share extra covariance.
 The \code{random} argument uses similar syntax as \pkg{lme4} [@bates2015lme4] and \pkg{nlme} [@pinheiro2006mixed] to specify nonspatial random effects.
@@
 
-According to AIC, the geostatistical SPGLM:
+According to AIC, the geostatistical SPGLM is preferred:
 ```{r}
@@
 The default estimation method in \pkg{spmodel} for SPGLMs is restricted maximum likelihood (REML), while maximum likelihood (ML) can also be used.
-A benefit of benefit of REML is that it can yield unbiased estimates of covariance parameters [@cressie1993asymptotic].
+A benefit of REML is that it can yield unbiased estimates of covariance parameters [@cressie1993asymptotic].
 A drawback of REML is that likelihood-based statistics are often considered invalid when the models have different explanatory variable or fixed effect structures [@wolfinger1993covariance], though @gurka2006selecting provides some evidence to the contrary.
@@
 spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
-spbeta_reduced_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
+spbeta_reduced_ml <- update(spbeta_full_ml, formula = turnout ~ 1)
 anova(spbeta_full_ml, spbeta_reduced_ml)
@@
 
-The results in this manuscript were obtained using [R]{.proglang} 4.4.0 with the
-\pkg{spmodel} 0.11.1 package. Figures were created using the [ggplot2]{.pkg} 3.5.1 package [@wickham2016ggplot2] and base [R]{.proglang}.
+The results in this manuscript were obtained using [R]{.proglang} 4.4.1 with the
+\pkg{spmodel} 0.11.1 package. Figures were created using the [ggplot2]{.pkg} 4.0.0 package [@wickham2016ggplot2] and base [R]{.proglang}.
 

### 2025-12-10 inst/manuscript/manuscript.tex
some upates

@@
 Non-Gaussian data are common in practice and include binary, count,
-skewed, and proportion data types. Often, non-Gaussian data are modeled
+skewed, and proportion structures. Often, non-Gaussian data are modeled
 using a generalized linear model (GLM). GLMs typically assume that
@@
 Environmental Protection Agency\\
+    Office of Research and Development\\
 200 SW 35th St\\
@@
 \begin{document}
-
-\linenumbers
 
@@
 errors that inflate Type-I error rates and decrease power
-\citep{zimmerman2024spatial}.
+\citep{legendre1993spatial, zimmerman2024spatial}.
 
@@
 f(\boldsymbol{\mu}|\mathbf{X}, \boldsymbol{\beta}, \boldsymbol{\tau}, \boldsymbol{\epsilon}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\tau} + \boldsymbol{\epsilon}.
-\end{align} Assuming independence among \(\boldsymbol{\tau}\) and
+\end{align} Assuming independence between \(\boldsymbol{\tau}\) and
 \(\boldsymbol{\epsilon}\), it follows that \begin{align}\label{eq-spcov}
@@
 spherical for point-referenced support (Figure\(~\)\ref{fig-type}) and
-the conditional autoregressive, and simultaneous autoregressive
-structures for areal support \pkg{spmodel} provides tools for commonly
-used model summaries, visualizations, and diagnostics (e.g., fitted
-values) using standard \proglang{R} helper functions like
-\code{summary()}, \code{plot()}, and \code{fitted()}, among others.
-\pkg{spmodel} also provides tools to predict \(\mathbf{w}\) at new
-locations and quantify uncertainty in those prediction using
-\code{predict()} and \code{augment()}. This core functionality, combined
-with several advanced features we describe throughout the manuscript,
-enables \pkg{spmodel} to introduce novel, important SPGLM modeling tools
+the conditional autoregressive and simultaneous autoregressive
+structures for areal support. \pkg{spmodel} provides commonly used model
+summaries, visualizations, and diagnostics (e.g., fitted values) using
+standard \proglang{R} helper functions like \code{summary()},
+\code{plot()}, and \code{fitted()}, among others. \pkg{spmodel} also has
+tools to predict \(\mathbf{w}\) at new locations and quantify
+uncertainty in those prediction using \code{predict()} and
+\code{augment()}. This core functionality, combined with several
+advanced features we describe throughout the manuscript, enables
+\pkg{spmodel} to introduce novel, important SPGLM modeling tools
 previously missing from the existing \proglang{R} ecosystem.
@@
 0.11.1) is arguably most similar to \pkg{sdmTMB} (version 0.7.4) in
-terms of scope and feel. Both packages use similar syntax as
+terms of scope and feel. Both packages use similar syntax to
 \code{glm()}, accommodate flexible \code{formula} arguments (e.g.,
-offsets, splines), handle spatial covariance that decays at different
-rates in different directions (i.e., geometric anisotropy), incorporate
-nonspatial random effects, support other \proglang{R} packages for
-modeling like \pkg{broom} \citep{robinson2021broom, kuhn2022tidy},
-\pkg{emmeans} \citep{lenth2024emmeans}, and \pkg{car}
-\citep{fox2019car}, and have tools for model summaries, prediction, and
-simulating data. There are some notable differences between the two
-packages, however. \pkg{sdmTMB} supports several additional GLM
-distributions like the Tweedie, supports Hurdle models, and can
-incorporate prior information through Bayesian applications.
-\pkg{sdmTMB} also provides tools for working with temporal data and
-spatiotemporal data and provides enhanced visualizations of the model's
-marginal effects. \pkg{sdmTMB} does require a preprocessing step of
-constructing a mesh prior to model fitting (using the stochastic partial
-differential equation approach), and the density of the mesh can affect
-model results and computational complexity. On the other hand,
+transformations, splines), handle spatial covariance that decays at
+different rates in different directions (i.e., geometric anisotropy),
+incorporate nonspatial random effects, support other \proglang{R}
+packages for modeling like \pkg{broom}
+\citep{robinson2021broom, kuhn2022tidy}, \pkg{emmeans}
+\citep{lenth2024emmeans}, and \pkg{car} \citep{fox2019car}, and have
+tools for model summaries, prediction, and simulating data. There are
+some notable differences between the two packages, however. \pkg{sdmTMB}
+supports several additional GLM distributions like the Tweedie, supports
+Hurdle models, and can incorporate prior information through Bayesian
+applications. \pkg{sdmTMB} also provides tools for working with temporal
+data and spatiotemporal data and provides enhanced visualizations of the
+model's marginal effects. \pkg{sdmTMB} does require a preprocessing step
+of constructing a mesh prior to model fitting (using the stochastic
+partial differential equation approach), and the density of the mesh can
+affect model results and computational complexity. On the other hand,
 \pkg{spmodel} does not require the construction of a mesh prior to model
@@
 to the Matérn spatial covariance as in \pkg{sdmTMB}. \pkg{spmodel} can
-model data directly using neighborhood distance and autoregressive
-models, rather than relying on the polygon centroid (as in
-\pkg{sdmTMB}), which may not be within the polygon's boundaries.
-\pkg{spmodel} provides experimental design tools (e.g., analysis of
-variance, contrasts), supports \pkg{sf} objects in modeling and
-prediction functions \citep{pebesma2018sf}, has several specialized
-model diagnostics like leverage values and Cook's distances, and has
-analytic solutions for fixed effect and prediction standard errors.
-Other similarities and differences do exist between \pkg{sdmTMB} and
-\pkg{spmodel}, and both packages continue to evolve. Overall, we believe
-that these packages are complementary and enhance the suite of SPGLM
-tools accessible to practitioners.
+model data using neighborhood distance and autoregressive models, rather
+than relying on the polygon centroid (as in \pkg{sdmTMB}), which may not
+be within the polygon's boundaries. \pkg{spmodel} provides experimental
+design tools (e.g., analysis of variance, contrasts), supports \pkg{sf}
+objects in modeling and prediction functions \citep{pebesma2018sf}, has
+several specialized model diagnostics like leverage values and Cook's
+distances, and has analytic solutions for fixed effect and prediction
+standard errors. Other similarities and differences do exist between
+\pkg{sdmTMB} and \pkg{spmodel}, and both packages continue to evolve.
+Overall, we believe that these packages are complementary and enhance
+the suite of SPGLM tools accessible to practitioners.
 
@@
 \(\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\).
-While we do not \(\mathbf{w}\), we do estimate it via
+While we do not know \(\mathbf{w}\), we do estimate it via
 \(\hat{\mathbf{w}}\), and thus it is reasonable to define
@@
 variance of an observation generally depends on its mean, response
-residuals lack some utility. Deviance residuals are a function of
-response residuals that are appropriately scaled to behave more like
-response residuals in a standard linear model. Deviance residuals are
-given by \begin{align}
+residuals lack some utility. Deviance, Pearson, and standardized
+residuals are all different function of responses residuals that are
+appropriately scaled to behave more like response residuals from a
+standard linear model. For example, deviance residuals are given by
+\begin{align}
   \mathbf{r}_{d} = sign(\mathbf{r}_{r})\sqrt{\mathbf{d}},
@@
 fitted model \citep{myers2012generalized}. Deviance is often used as a
-fit statistic; lower values of deviance imply a better model fit
-(compared to the observed data). Pearson and standardized residuals are
-other types of GLM residuals that involve a scaling of the response
-residuals. The Pearson residuals scale \(\mathbf{r}_{r}\) by the square
-root of \(\mathbf{V}\), a diagonal matrix with \(i\)th diagonal element
-equal to the variance of the response distribution evaluated at
-\(f^{-1}(\hat{\text{w}}_i)\) \citep{faraway2016extending};
-\(\mathbf{V}\) is sometimes called the GLM weight matrix. The
-standardized residuals scale the deviance residuals by
-\(\frac{1}{\sqrt{(1 - \mathbf{L}_{ii})}}\), where \(\mathbf{L}_{ii}\) is
-the \(i\)th diagonal element of the leverage matrix, which we discuss
-next.
+fit statistic, whereby lower values of deviance imply a better model fit
+(given the observed data). Scaling the response residuals by the
+reciprocal square root of the variance function (for the appropriate
+response distribution) yields the Pearson residuals, and scaling the
+Pearson residuals by the reciprocal square root of one minus the
+leverage value (which we discuss next) yields the standardized residuals
+\citep{faraway2016extending}.
 
@@
   \mathbf{L} = \mathbf{V}^{1/2} \mathbf{X} (\mathbf{X}^\top \mathbf{V} \mathbf{X}) \mathbf{X}^\top \mathbf{V}^{1/2}.
-\end{align} The larger the value of \(\mathbf{L}_{ii}\), the more severe
-the leverage from the \(i\)th observation.
+\end{align} where \(\mathbf{V}\) is a diagonal matrix with \(i\)th
+diagonal element equal to the variance function evaluated at
+\(f^{-1}(\hat{\text{w}}_i)\). The larger the value of
+\(\mathbf{L}_{ii}\), the more severe the leverage from the \(i\)th
+observation.
 
@@
 The \code{moose} data in \pkg{spmodel} contain information on moose
-(Alces Alces) presence in the Togiak region of Alaska, USA. \code{moose}
+(Alces alces) presence in the Togiak region of Alaska, USA. \code{moose}
 is an \code{sf} object, a special data frame that is supplemented with
@@
 models uses the same hierarchical likelihood
-(Equation\textasciitilde{}\ref{eq-marginal}):
+(Equation\(~\)\ref{eq-marginal}):
 
@@
 RMSPE than the GLM, indicating the SPGLM predictions are far more
-efficient. Both the liklelihood-based leave-one-out cross validation
+efficient. Both the likelihood-based leave-one-out cross validation
 metrics prefer the SPGLM to the GLM.
@@
 for diagnosing spatial covariance. For example, we can inspect Cook's
-distance values and the empirical spatial covariance as a function of
+distance values and the fitted spatial covariance as a function of
 distance with (Figure\(~\)\ref{fig-sp-diagnostic2}):
@@
 spherical spatial covariance function, we may fit both a Poisson and
-negative binomial SPGLM changing the \code{family} argument:
+negative binomial SPGLM by changing the \code{family} argument:
 
@@
 collected in 2012 and 2017, so this partition factor assumes
-independence between observations in 2012 and 2017. While we used the
-partition factor here illustratively, more generally, the utility of
-partition factors can be highly context dependent.
+observations in 2012 are independent of observations in 2017. While we
+used the partition factor here illustratively, more generally, the
+utility of partition factors can be highly context dependent.
 
@@
 distance is used to determine whether two sites are neighbors, though
-custom neighborhood matrices can be passed via \code{W}. Row
-standardization is also assumed by default; this can be changed via
-\code{row_st}. Using \code{random}, we also specified a nonspatial
-random effect for seal stock, which implies seals belonging to the same
-stock share extra covariance. The \code{random} argument uses similar
-syntax as \pkg{lme4} \citep{bates2015lme4} and \pkg{nlme}
-\citep{pinheiro2006mixed} to specify nonspatial random effects.
+custom neighborhood matrices can be passed via the \code{W} argument.
+Row standardization is also assumed by default; this can be changed via
+the \code{row_st} argument. Using the \code{random} argument, we also
+specified a nonspatial random effect for seal stock, which implies seals
+belonging to the same stock share extra covariance. The \code{random}
+argument uses similar syntax as \pkg{lme4} \citep{bates2015lme4} and
+\pkg{nlme} \citep{pinheiro2006mixed} to specify nonspatial random
+effects.
 
@@
 
-According to AIC, the geostatistical SPGLM:
+According to AIC, the geostatistical SPGLM is preferred:
 
@@
 maximum likelihood (REML), while maximum likelihood (ML) can also be
-used. A benefit of benefit of REML is that it can yield unbiased
-estimates of covariance parameters \citep{cressie1993asymptotic}. A
-drawback of REML is that likelihood-based statistics are often
-considered invalid when the models have different explanatory variable
-or fixed effect structures \citep{wolfinger1993covariance}, though
+used. A benefit of REML is that it can yield unbiased estimates of
+covariance parameters \citep{cressie1993asymptotic}. A drawback of REML
+is that likelihood-based statistics are often considered invalid when
+the models have different explanatory variable or fixed effect
+structures \citep{wolfinger1993covariance}, though
 \citet{gurka2006selecting} provides some evidence to the contrary. In
@@
 R> spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
-R> spbeta_reduced_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
+R> spbeta_reduced_ml <- update(spbeta_full_ml, formula = turnout ~ 1)
 R> anova(spbeta_full_ml, spbeta_reduced_ml)
@@
 
-The results in this manuscript were obtained using {R} 4.4.0 with the
+The results in this manuscript were obtained using {R} 4.4.1 with the
 \pkg{spmodel} 0.11.1 package. Figures were created using the {ggplot2}
-3.5.1 package \citep{wickham2016ggplot2} and base {R}.
+4.0.0 package \citep{wickham2016ggplot2} and base {R}.
 

### 2025-12-10 inst/manuscript/code/code.R

@@ -167,7 +167,7 @@ spbeta_auto <- spgautor(
 AIC(spbeta_geo, spbeta_auto)
 
 spbeta_full_ml <- update(spbeta_geo, estmethod = "ml")
-spbeta_reduced_ml <- update(spbeta_geo, estmethod = "ml", formula = turnout ~ 1)
+spbeta_reduced_ml <- update(spbeta_full_ml, formula = turnout ~ 1)
 anova(spbeta_full_ml, spbeta_reduced_ml)
 
 AIC(spbeta_full_ml, spbeta_reduced_ml)
@@ -249,20 +249,21 @@ spbin <- spglm(
 spbin_aug <- augment(spbin)
 
 p1 <- ggplot(spbin_aug, aes(color = .hat)) +
-  geom_sf(size = 1) +
+  geom_sf(size = 2) +
   scale_color_viridis_c(option = "E", name = ".hat") +
   scale_y_continuous(breaks = seq(62.6, 63.6, length.out = 5)) +
   scale_x_continuous(breaks = seq(-148.5, -146, length.out = 3)) +
   theme_bw(base_size = 14)
 
 p2 <- ggplot(spbin_aug, aes(color = .std.resid)) +
-  geom_sf(size = 1) +
+  geom_sf(size = 2) +
   scale_color_viridis_c(option = "A", limits = c(-2, 2)) +
   scale_y_continuous(breaks = seq(62.6, 63.6, length.out = 5)) +
   scale_x_continuous(breaks = seq(-148.5, -146, length.out = 3)) +
   theme_bw(base_size = 14)
 
-p1 + p2
+p1
+p2
 
 #==========================================#
 # Figure 5 ####
@@ -299,20 +300,21 @@ figure6
 #==========================================#
 
 p1 <- ggplot(preds_aug, aes(color = .lower)) +
-  geom_sf(pch = 17) +
+  geom_sf(size = 2, pch = 17) +
   scale_color_viridis_c(option = "H", name = "Lower", limits = c(0, 1)) +
   scale_y_continuous(breaks = seq(62.6, 63.6, length.out = 5)) +
   scale_x_continuous(breaks = seq(-148.5, -146, length.out = 3)) +
   theme_bw(base_size = 14)
 
 p2 <- ggplot(preds_aug, aes(color = .upper)) +
-  geom_sf(pch = 17) +
+  geom_sf(size = 2, pch = 17) +
   scale_color_viridis_c(option = "H", name = "Upper", limits = c(0, 1)) +
[…]

### 2025-12-11 inst/manuscript/manuscript.Rmd
typo

@@
 Because observations often have a unique support in a GLM (e.g., only two possible response values for binary data) and the variance of an observation generally depends on its mean, response residuals lack some utility.
-Deviance, Pearson, and standardized residuals are all different function of responses residuals that are appropriately scaled to behave more like response residuals from a standard linear model.
+Deviance, Pearson, and standardized residuals are all different functions of response residuals that are appropriately scaled to behave more like response residuals from a standard linear model.
 For example, deviance residuals are given by

### 2025-12-11 inst/manuscript/manuscript.tex
typo

@@
 residuals lack some utility. Deviance, Pearson, and standardized
-residuals are all different function of responses residuals that are
+residuals are all different functions of response residuals that are
 appropriately scaled to behave more like response residuals from a
